\addchap{\lsAcknowledgementTitle}

First and foremost, we thank the speakers of Iranian Armenian who shared their speech and culture with us. 

Afsheen extends his heartfelt gratitude to his parents, Khalil and Simin Sharifzadeh, whose support allowed him to acquire his first Armenian primer as a teenager, and years later to journey freely throughout Persia and Armenia conducting fieldwork. Afsheen thanks his brother, Arya Sharifzadeh, who acted as a sounding board for sociological and linguistic discussions, and without whom his research on the Iranian Armenian dialect would not have happened. To Dr. Ina Baghdiantz-McCabe, a world expert on New Julfa who inspired his passion for Iranian Armenian topics, Afsheen says \armenian{մի աշխարհ շնորհակալութիւն}. Afsheen extends a special thanks to his dear friends Sevada, Tina and Lernik Yedgarian for their undying hospitality and becoming something of an adoptive Armenian family in Tehran, Yerevan, and Los Angeles.

 
Bert thanks Armen, Anna Gevorkyan, Ara Ghazarians, Arsineh Artounians Hovannisian, Bavrina Bigjahan, Hagop Hachikian, Narineh Hacopian,  Anahit Keshishian, Armineh Mirzabegian, and Alla Petrosyan. He would especially like to thank  Karine Megerdoomian for her extensive insights into Iranian Armenian lexis, syntax, and sociolinguistics over the years, and Richard Hovannisian for introducing him to many members of the Iranian Armenian community in the Los Angeles area over the years, which first made him aware of how distinctive and widespread the dialect is, and for organising the 2004 UCLA conference on Armenians in Iran that forced him to begin crystallizing his fieldwork on the dialect.



Hossep thanks Hannah Cox and  Vartan Haghverdi  for teaching him what Iranian Armenian is in the first place.   Hossep also thanks Karine Megerdoomian for continuing fieldwork together. 

Among Hossep's consultants, much of the material in this grammar wouldn't have been possible without Hossep's main consultant and language teacher, Nicole Khachikian. Hossep thanks her from the bottom of his heart for all the Zoom meetings.  He thanks Arevik Torosyan for providing Eastern recordings, and for bearing with him over Facebook. Vahagn Petrosyan is owed special thanks for organizing the Eastern Armenian paradigms on Wiktionary in such a way that it was relatively easy to go about our morphology fieldwork.  We thank the many Eastern speakers who helped us with elicitations and Eastern syntax: Mariam Asatryan, Harut Hayrapetyan, Katherine Hodgson, Victoria Khurshudyan.  



We  thank Beaina Amirian, James Barry, Maryam Ghiasian, Shushan Karapetian, Varand Nikolaian,  Shakeh Amirian Petrossian,     Hakimeh Rezayi, Claris Sarkissian,  and Hamo Vassilian for help in tracking down Iranian Armenian sources. 

We thank the comedians who wrote and acted out the sample text on Instagram:\footnote{\url{https://www.instagram.com/tv/COWtIvUn4KA/}}  Tiffany Alice, Ryan Ebrahamian,   Helen Kalognomos,  Loucineh Mardirossian,  Shant Nazarian,  and Gilbert Sinanian. 
	
For discussion of various aspects of Armenian dialectology, we thank Nikita Bezrukov,  Peter Cowe, Garoun Engström,  Hagop Gulludjian,  Tereza Hovhannisyan,   Hrach Martirosyan, and Anooshik Melikian. For Turkic discussions, we thank   Stephen Nichols and Jonathan North Washington. 


For discussion of various aspects of Persian phonology and syntax, we thank Koorosh Ariyaee, Reza Falahati, Hamed Rahmani,  Nima Sadat-Tehrani, and Scott Seyfarth. We especially thank  Nazila Shafiei for syntax elicitations. 

For discussion of  the liquid deletion process, we thank Arto Anttila,  Canaan Breiss, Larry Hyman,  Mark Liberman, Kate Lindsey,  Nicholas Rolle, Katherine Russell, Hannah Sande, and Shanti Ulfsbjorninn. 


For discussions of the past perfective paradigms, we thank  Sargis Avetyan, Matthew Carter, Borja Herce, Brian Joseph, Laura Kalin, Shuan Karim, and Jordan Kodner. 

We thank Diana Forker, Felix Kopecky, and Sebastian Nordhoff for their patience and advice as we went through the publishing process. We owe a special thanks to our reviewers (Katherine Hodgson and Donald Stilo) and proofreaders for making this book better. 

Finally, we thank the sender of the anonymized email in \S\ref{section: intro: socio}. If it wasn't for that email, Hossep would not have bothered to do any of the synthesis or replication work. For better or worse, this grammar was done out of resistance against linguistic discrimination. 

