\title{A grammar of Rapa Nui}  %look no further, you can change those things right here.
%\subtitle{The language of Easter Island}
\BackTitle{A grammar of Rapa Nui} % Change if BackTitle != Title
\BackBody{This book is a comprehensive description of the grammar of Rapa Nui, the Polynesian language spoken on Easter Island. After an introductory chapter, the grammar deals with phonology, word classes, the noun phrase, possession, the verb phrase, verbal and nonverbal clauses, mood and negation, and clause combinations. 

The phonology of Rapa Nui reveals certain issues of typological interest, such as the existence of strict conditions on the phonological shape of words, word-final devoicing, and reduplication patterns motivated by metrical constraints. 

For Polynesian languages, the distinction between nouns and verbs in the lexicon has often been denied; in this grammar it is argued that this distinction is needed for Rapa Nui. Rapa Nui has sometimes been characterised as an ergative language; this grammar shows that it is unambiguously accusative. Subject and object marking depend on an interplay of syntactic, semantic and pragmatic factors. 

Other distinctive features of the language include the existence of a ‘neutral’ aspect marker, a serial verb construction, the emergence of copula verbs, a possessive-relative construction, and a tendency to maximise the use of the nominal domain. Rapa Nui’s relationship to the other Polynesian languages is a recurring theme in this grammar; the relationship to Tahitian (which has profoundly influenced Rapa Nui) especially deserves attention. 

The grammar is supplemented with a number of interlinear texts, two maps and a number of indices. }
%\dedication{Change dedication in localmetadata.tex}
\typesetter{Paulus Kieviet, Sebastian Nordhoff}
\proofreader{
Ahmet  Özdemir,
Andreas Hölzl,
Annie Zaenen,
Christian Döhler,
Eitan Grossman,
Elizabeth Zeitoun,
Ezekiel Bolaji,
Guylaine Brun-Trigaud,
Hella Olbertz,
Jean Nitzke,
Julien Heurdier, 
Klara Kim,
Matti Miestamo,  
Myke Brinkerhoff,
Natsuko Nakagawa,
Oliver \v{C}ulo,
Prisca Jerono
}
\author{Paulus Kieviet}
\BookDOI{10.5281/zenodo.235525}%ask coordinator for DOI
\renewcommand{\lsISBNdigital}{978-3-946234-75-3}
\renewcommand{\lsISBNhardcover}{978-3-946234-76-0}
\renewcommand{\lsISBNsoftcover}{978-1-542544-79-5 }
% \renewcommand{\lsISBNsoftcoverus}{000-0-000000-00-0}
\renewcommand{\lsSeries}{sidl} % use lowercase acronym, e.g. sidl, eotms, tgdi
\renewcommand{\lsSeriesNumber}{12} %will be assigned when the book enters the proofreading stage
\renewcommand{\lsURL}{http://langsci-press.org/catalog/book/124} % contact the coordinator for the right number

 
%<*coverdimen>

\setlength{\csspine}{37.98mm} % Please calculate: Total Page Number (excluding cover, usually (Total Page - 3)) * 0.0572008 mm
\setlength{\bodspine}{59.2mm} % Please calculate: Total Page Number (excluding cover) * BODFACTOR + BODABS

%</coverdimen> 

