\addchap{Preface}
\largerpage
\textit{Advances in Formal Slavic Linguistics 2022} brings together a collection of 22 articles presenting the latest developments in formal approaches to Slavic linguistics. Covering a broad spectrum of topics across all branches of linguistics, %\footnote{We could also list them (phonology, morphology, syntax, semantics and pragmatics)} 
the contributions explore various phenomena in BCMS, Slovenian, Russian, Ukrainian, Polish, Czech, Upper Sorbian, Bulgarian, and Macedonian, along with their regional dialects.

The authors use a wide range of methodologies, including fieldwork, acoustic analysis, corpus studies, psycholinguistic experiments, judgment studies, and theoretical modelling. The studies examine topics such as clitics, nominalizations, l-participles, the dual, verbal prefixes, assibilation, verbal and adjectival morphology, lexical stress, vowel reduction, focus particles, aspect, multiple wh-fronting, definiteness, polar questions, negation words, and argument structure. %\footnote{In random order...} 
Analytical frameworks include Metrical Phonology, Distributed Morphology, Nanosyntax, Minimalist syntax, and formal semantics.

Early versions of these articles were presented at the conference on Formal Description of Slavic Languages 15, held in Berlin on 5–7 October 2022, with a special session on formal approaches to Ukrainian. Each article underwent a thorough, double-blind peer review process. We would like to extend our sincere thanks to the reviewers, as well as to the community proofreaders. We are also grateful to Mihaela Chirpanlieva for her assistance with Bulgarian transliteration and to the Language Science Press editorial team, Sebastian Nordhoff and Felix Kopecky, as well as Radek Šimík, for their invaluable help in producing this volume.

We hope this book provides a valuable resource for linguists interested in the complexities of Slavic languages and their implications for linguistic theory.\\

\hfill Berlin, \today   \\
{}\hfill Berit Gehrke\\
{}\hfill Denisa Lenertová\\
{}\hfill Roland Meyer\\
{}\hfill Daria Seres\\
{}\hfill Luka Szucsich\\
{}\hfill Joanna Zaleska

