\addchap{Preface}

Although languages around the world display an overwhelming variety in
ways to describe colours, most of the research in the domain of colour
has focussed on the use of single colour terms. This approach has
allowed researchers in a wide range of fields to tackle interesting
questions, such as the extent to which colour categories are innate or
learned. In the field of artificial language evolution, the focus on
single colour terms has enabled researchers to build computational
models in which populations of linguistic agents can construct and
coordinate their own colour category system so that they become
successful in communication.

A few descriptive studies report on describing colours beyond the
restriction of using a single colour term. The results of these
studies seem conclusive: only a small minority (around 15\%) of all
colour samples would be described using a single colour term. Most
samples are described using more elaborate expressions, for example by
using modifiers or combinations of colour terms.

In this book, I show how the current models in artificial language
evolution can be extended to allow for richer descriptions of colour
samples. In order to do so, I deploy two powerful formalisms that have
been developed to support this kind of experiments: Incremental
Recruitment Language (IRL) to represent the semantics, or meaning, of
linguistic utterances and Fluid Construction Grammar (FCG) to
transform these meanings into linguistic utterances and back.

Four different language strategies are explored: the basic colour
strategy\linebreak (``blue''), the graded membership strategy (``greenish''),
the category combination strategy (``blue-green'') and the basic
modification strategy (``dark blue''). Each of these strategies is
realised in different languages around the world and some studies
reported on the most prototypical colour samples that are associated
with these expressions. For each strategy, I propose a semantic
template which captures the general cognitive operations required to
use that particular strategy and syntactic templates which represent
general grammatical rules that can express semantic templates in
language. I pursue a compositional approach, focussing on the re-use
of semantic primitives and syntactic templates as much as possible. I
show that more complicated language strategies can be conceived as
minor extensions of the basic colour strategy and that only a few
syntactic templates suffice to express all these strategies. Once
these strategies have been operationalised, I compare their naming
behaviour to human data reported in the literature. The performance of the
strategies can be compared in a baseline experiment in which simulated
language users engage in linguistic interactions, the
difficulty of which is carefully controlled.

The implementation of a language strategy can be completed by adding
learning operators which allow an agent to pick up the language system
of another agent and to extend the current language system whenever
the communicative need arises. The performance of these operators is
tested in an acquisition and a formation experiment. In an acquisition
experiment, one agent knows a predefined language system and acts as a
teacher. The goal of the learner agent is to acquire the predefined
language system and to become as successful in communication as two
agents which share perfect knowledge of the predefined language
system. In a formation experiment, a population of agents need to
invent and coordinate their own language system based on a particular
language strategy. I present results for both the basic colour
strategy and the graded membership strategy.

Once the implementation of a language strategy is completed, in-depth
studies can be carried out. I show the results of three different
studies using the basic colour strategy: (a) the positive impact of
the statistical distribution of colours in the environment on the
similarity between simulated and human basic colour systems (b) the
coordinating role of language on simulated language systems and the
positive impact of language on the similarity between simulated and
human basic colour systems (c) the impact of embodiment on the
performance of different learning operators. In embodied experiments,
two robots perceive a shared environment through their vision
systems. Although this introduces a certain level of noise as both
robots perceive the world from a different perspective, the data
contain a high level of structure as it is based on the colours of
the objects presented to the robots. Overall, embodiment has a
positive effect on the performance of the proposed learning operators.

In the history of a language, a competition between two strategies on
how to express a particular domain might arise. In the domain of
colour, this has been observed in a vast number of languages which
shift from being brightness based to being hue based. The colour term
\textit{yellow} used to reflect the meaning `to shine' in Old English but
shifted to a hue sense in Middle English and could be used to refer to
the colour of yolk or discoloured paper. I present a model in which a
population of agents successfully aligns on which language strategy
they use based on linguistic interactions. I show that this model is \enlargethispage{\baselineskip}
capable of reproducing the meaning shifts similar to those reported in
literature.

Finally, I address some questions on the origins of new language
strategies. New semantic templates can be generated through a
combinatorial search process in which semantic primitives are combined
to form complex semantic templates. I show that each of the proposed
language strategies for the domain of colour can be the outcome of
such a search process. The syntactic templates that have been
introduced to express these templates in language can be incorporated
in repair strategies which allow agents to invent, acquire and align
their own set of grammatical rules. I demonstrate how these repair
strategies allow a population of agents to form their own hierarchical
language that includes some recursive rules. These recursive rules
have the benefit of being able to express more complex meaning without
the cost of alignment in the population.

Even though the examples in this book are limited to the domain of
colour, the proposed templates can easily be extended to richer
examples and deployed in other continuous domains. The proposed
transformation processes could be used to name the colours of concepts
that vary in colour, like for example the colours used to describe
wine. Other possible domains include the spatial domain, in which
spatial categories, such as near and far, also exhibit properties of
graded membership which can be made explicit in language (eg. \textit{very
near}).  The results reported in this book should hence not be
thought of as final but rather as in interesting starting
point for a whole line of research on the origins and evolution of
natural languages.