\title{Clitics in the wild}
\subtitle{Empirical studies on the microvariation of the pronominal, reflexive and verbal clitics in Bosnian, Croatian and Serbian}
\author{Zrinka Kolaković and Edyta Jurkiewicz-Rohrbacher and Björn Hansen and Dušica Filipović Đurđević and Nataša Fritz}
\renewcommand{\lsSeries}{osl}
\renewcommand{\lsSeriesNumber}{7}
\BackBody{This collective monograph is the first data-oriented, empirical in-depth study of the system of clitics on Bosnian, Croatian and Serbian. It fills the gap between the theoretical and normative literature by including solid data on variation found in dialects and spoken language and obtained from massive Web Corpora and speakers’ acceptability judgements. The authors investigate three primary sources of variation: inventory, placement and morphonological processes. A separate part of the book is dedicated to the phenomenon of clitic climbing, the major challenge for any syntactic theory. The theory of complexity serves as the explanation for the very diverse constraints on clitic climbing established in the empirical studies. It allows to construct a series of hierarchies where the factors relevant for predicting clitic climbing interact with each other. Thus, the study pushes our understanding of clitics away from fine-grained descriptions and syntactic generalisations towards a probabilistic modelling of syntax.}

\renewcommand{\lsISBNdigital}{978-3-96110-336-2}
\renewcommand{\lsISBNhardcover}{978-3-98554-032-7}
\BookDOI{10.5281/zenodo.5792972}

\renewcommand{\lsID}{339}

\renewcommand{\lsImpressumCitationText}{Zrinka Kolaković, Edyta Jurkiewicz-Rohrbacher, Björn Hansen,
Dušica Filipović Đurđević \& Nataša Fritz.
2022.
\textit{Clitics in the wild: Empirical studies on the microvariation of the pronominal, reflexive and verbal clitics in Bosnian, Croatian and Serbian} (Open Slavic Linguistics 7).
Berlin: Language Science Press.
}

\proofreader{Alexandr Rosen,
Amir Ghorbanpour,
Amy Amoakuh,
Brett Reynolds,
Cesar Perez Guarda,
Christopher Straughn,
Jeroen van de Weijer,
Krystyna Kupiszewska,
Marten Stelling,
Jean Nitzke
}

\typesetter{Eberhard Gade, Edyta Jurkiewicz-Rohrbacher}
