\author{Prashant Parikh}
\title{Communication and content}  
%\subtitle{Change your subtitle in localmetadata.tex}
%\BackTitle{Change your backtitle in localmetadata.tex} % Change if BackTitle is different from Title

\renewcommand{\lsSeries}{tgdi} % use lowercase acronym, e.g. sidl, eotms, tgdi
\renewcommand{\lsSeriesNumber}{4} %will be assigned when the book enters the proofreading stage

\BackBody{\textit{Communication and content} presents a comprehensive and foundational account of meaning based
on new versions of situation theory and game theory. The literal and implied meanings of an utterance 
are derived from first principles assuming little more than the partial rationality of interacting
agents. New analyses of a number of diverse phenomena -- a wide notion of ambiguity and content
encompassing phonetics, syntax, semantics, pragmatics, and beyond, vagueness, convention and
conventional meaning, indeterminacy, universality, the role of truth in communication, semantic
change, translation, Frege’s puzzle of informative identities -- are developed. 
Communication, speaker meaning, and reference are defined. Frege’s context and compositional principles
are generalized and reconciled in a fixed-point principle, and a detailed critique of Grice, several
aspects of Lewis, and some aspects of the Romantic conception of meaning are offered. Connections
with other branches of linguistics, especially psycholinguistics, sociolinguistics, historical linguistics,
and natural language processing, are explored.

The book will be of interest to scholars in philosophy, linguistics, artificial intelligence, and cognitive science. 
It should also interest readers in related fields like literary and cultural theory and the social sciences.}

\dedication{for john perry and ennio stacchetti\\
	
	and\\
	
	for avani and neal}
\typesetter{Felix Kopecky}
\proofreader{Ahmet Bilal Özdemir, Brett Reynolds, Catherine Rudin, Christian Döhler, Eran Asoulin, Gerald Delahunty, Jeroen van de Weijer, Joshua Phillips, Ludger Paschen, Prisca Jerono, Sebastian Nordhoff, Trinka D'Cunha, Tom Bossuyt, Umesh Patil, Vadim Kimmelman}

\renewcommand{\lsID}{248} % contact the coordinator for the right number
\BookDOI{10.5281/zenodo.3243924}
\renewcommand{\lsISBNdigital}{978-3-96110-198-6}
\renewcommand{\lsISBNhardcover}{978-3-96110-199-3}
