\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.15394162}
%\bibliography{localbibliography}

\author{Svitlana Antonyuk\orcid{}\affiliation{University of Graz}}

\SetupAffiliations{mark style=none}


\title[From scope freezing to, well, everything]{From scope freezing to, well, everything: Investigations into the syntax of Instrumentals in Ukrainian}

\abstract{I present arguments in favor of a particular view of argument hierarchies in alternations that results from taking a deductive approach to scope freezing (\citealt{Antonyuk2015,Antonyuk2020, Antonyuk2023, AntonyukUnderReview})
%\todo{Svitlana, please, provide all the bibliographical references that are now in bold} 
and especially from relying on the Scope Freezing Diagnostic (SFD) as a source of insight in this domain. The primary focus of the paper is on several causative constructions in Ukrainian that feature an Instrumental case-marked argument. It is argued that in all constructions under consideration Instrumentals are merged in a structurally more prominent position than either the Accusative or the Dative argument, resulting in the V Instrumental $>>$ Accusative $>>$ Dative relative argument ordering at Merge. Next, I follow \citet{Lavine2022, Lavine2023} in arguing against the Undifferentiated Initiator view (\citealt{ant:Ramchand2008}; \citealt{Bruening2013}; \citealt{Legate2014}; \citealt{Wood2017}, i.a.) and provide a new argument in support of Lavine’s structural differentiation between the Instrumental Agents of passives and Instrumental Inanimate Causers.

\keywords{scope freezing; argument structure; argument structure alternations; causers; passives; non-agreeing accusatives; Ukrainian}}

\begin{document}
\maketitle

\section{Introduction}\label{sec:antonyuk:1}
While research into argument structure alternations (ASAs) has made great advances, yielding important results over more than six decades, it has also been characterized by the vastly different, indeed, often diametrically opposing conclusions that have been reached based on largely the same types of evidence. Limiting our attention to the Slavic languages for the purposes of this discussion, we notice that the debate around the Ditransitive Alternation (exemplified in \REF{ex:antonyuk:1} and \REF{ex:antonyuk:2} for English and Ukrainian) has witnessed accounts that posit a derivational relation between the two frames as well as those that posit independent projection of the two structures. Within these two large groups, further differences are observed based on the posited Merge position of the internal arguments, with accounts differing on, for example, whether the Dative argument is merged in the Specifier of V \citep{GreenbergFranks1991,Franks1995,Richardson2007}, the complement of V (\citealt{Bailyn1995,Bailyn2010, Bailyn2012}; \citealt{Antonyuk2015,Antonyuk2020}; \citealt{Titov2017}), the Specifier of an Applicative head \citep{Dyakonova2009}, as well as accounts that posit Datives base-merged at different heights (\citealt{BonehNash2017}). 

\ea%1
    \label{ex:antonyuk:1}
    The Ditransitive Alternation
\ea   Mike gave a toy to his cat.        \hfill Prepositional Dative (PPD)
\ex  Mike gave his cat a toy.       \hfill Double  Object Construction (DOC)
\z
\z

\ea%2
    \label{ex:antonyuk:2}
\ea \gll   Myxajlo   po-daruvav   \minsp{(} jakus’) igrašku   svojij kišci.    \\
	Myxajlo.\textsc{nom}     \textsc{po}-gift.\textsc{pst} {} some toy.\textsc{acc}    his cat.\textsc{dat}\\\hfill PPD
\glt ‘Myxajlo gifted (some)/a toy to his cat.’

\ex
\gll Myxajlo   po-daruvav     svojij kišci   \minsp{(} jakus’)  igrašku.    \\
  Myxajlo.\textsc{nom}     \textsc{po}-gift.\textsc{pst} his cat.\textsc{dat}  {}  some toy.\textsc{acc}\\\hfill DOC
\glt `Myxajlo gifted his cat (some)/a toy.’\hfill (Ukrainian)
\z
\z

\noindent A rather underinvestigated property of the Ditransitive Alternation, familiar since \citet{Larson1990}, is the “frozen” surface scope relation between the internal arguments in the DOC, which contrasts with scope fluidity/scope ambiguity of the PPD frame, cf. \REF{ex:antonyuk:3a}--\REF{ex:antonyuk:3b}.

\ea%3
    \label{ex:antonyuk:3}

\ea \label{ex:antonyuk:3a}    Mike gave some toy to every cat.        \\
\textit{Surface scope}: For some toy \textit{x}, for every cat \textit{y}, Mike gave \textit{x} to \textit{y }(e.g., this week).

\textit{Inverse Scope}: For every cat \textit{x}, for some toy \textit{y}, Mike gave  \textit{x} \textit{y} (different cat--toy pairings possible).

\ex \label{ex:antonyuk:3b}  Mike gave some cat every toy.        \hfill (frozen surface scope)

\textit{Surface scope}: For some cat\textit{ x}, for every toy \textit{y}, \textit{x} received \textit{y} from Mike (i.e., ‘one happy cat’ scenario).

\textit{Unavailable inverse scope}: For every toy \textit{x}, for some cat \textit{y}, \textit{x} was given to \textit{y} (different toy-cat pairings possible).
    \z
\z

\noindent As noted in \citet{Larson1990}, \citet{Schneider-Zioga1988} observes the arguably identical pattern of scope freezing in the `with'{}-variant of the \textit{Spray-Load} Alternation \REF{ex:antonyuk:4}, thus the scope ambiguity-scope freezing patterns found in the two alternations have come to be known as their characteristic, albeit fairly understudied, property.\footnote{\textrm{Scope freezing is the only property of the ditransitive alternation (DA) where the mirror-image behavior of the two frames breaks down with respect to the \citeauthor{BarssLasnik1986}'s (\citeyear{BarssLasnik1986}) diagnostics. Yet, as noted in \citet{HarleyMiyagawa2017}, it continues to be underinvestigated and poorly understood.}}
\ea%4
    \label{ex:antonyuk:4}
    The \textit{Spray-Load} Alternation
    \ea    Mike planted the flowers in the garden.       \hfill locative frame\\
    \ex Mike planted the garden with flowers.       \hfill the `with'-frame
\z
\z

\ea%5
    \label{ex:antonyuk:5}
\ea   Mike planted some sort of flowers in every garden.  \\
\textit{Surface scope}: For some sort of flowers \textit{x}, for every garden \textit{y}, Mike planted \textit{x} in \textit{y}.

\textit{Inverse scope}: For every garden \textit{x}, for some sort of flowers \textit{y}, Mike planted  \textit{x} with \textit{y}.

\ex    Mike planted some garden with every sort of flowers. \\  \hfill (frozen surface scope)

\textit{Surface scope}: For some garden \textit{x}, for every sort of flowers \textit{y}, Mike planted \textit{x} with \textit{y}.

\textit{Unavailable inverse scope}: For every sort of flowers \textit{x}, for some garden \textit{y}, Mike planted \textit{x} in \textit{y}.
\z
\z

\noindent While the above scope patterns and the scope freezing in particular have come into focus in \citet{Bruening2001}, the phenomenon has effectively been explained away: \citeauthor{Bruening2001} takes scope freezing to be epiphenomenal, that is, a mere consequence of the purportedly Superiority-obeying nature of the covert movement operation Quantifier Raising (QR), which, combined with \citeauthor{Richards1997}'s (\citeyear{Richards1997}) \textsc{Tucking in}, is taken to result in situations where the structurally more prominent QP takes obligatory wide scope upon QR due to the lower one obligatorily tucking in and thus scoping below it.\footnote{\citeauthor{Bruening2001}'s (\citeyear{Bruening2001}) treatment of scope freezing is critically evaluated and ultimately rejected in \citet{LarsonEtAl2019} for English and \citet{AntonyukMykhaylyk2022} for Ukrainian. See \citet{AbelsGrabska2022} for a recent account that adopts (a modified version of) \citet{Bruening2001} to model scope relations in Polish ditransitives and \citet{Hallman2018, Hallman2024}  for the causative ditransitive alternation in Syrian Arabic.} Assuming the treatment in \citet{Antonyuk2015, Antonyuk2020, Antonyuk2023, AntonyukUnderReview}, where scope freezing is viewed as an empirical phenomenon in its own right and thus in need of a principled theoretical explanation, the present paper will take without justification the need for the deductive approach to scope freezing underlying the above works and will proceed to rely on the \textsc{Scope Freezing Diagnostic} \citep{Antonyuk2015} yielded by this approach in order to probe underlying argument structure relations.\footnote{An anonymous reviewer challenges the terminology, pressing about what makes this approach deductive rather than inductive. Of course, the reviewer is correct to press on it, as the present approach does in fact combine elements of both. Inductively, this approach follows the data to form generalizations to form broader theory of scope freezing. The approach is however deductive in that, relying on the general minimalist set of assumptions and the newly formed theory (or, barring that, a set of strong generalizations about scope freezing), it allows us to test a range of properties under discussion, from properties of scope freezing itself, to argument hierarchy in argument structure alternations and syntactic properties of particular constructions to, e.g., information structural phenomena manifested in such constructions, such as specificity, discourse neutrality, etc. (see esp. \citealt{Antonyuk2023}). To the extent that any such insights are derived from the theory of scope freezing advocated for here, they are all derived by deductive reasoning. I am grateful to the reviewer for the chance to clarify this point.} As will hopefully become apparent, this approach to scope freezing and especially the diagnostic contributed by it is highly valuable as far as diagnostics developed for Slavic languages go, able to provide non-trivial insights into the syntax of argument structure alternations and remaining remarkably consistent where other diagnostics fail to be so. Among the findings reported here is that across a number of causative constructions involving an Instrumental NP, the latter is consistently implicated by the Scope Freezing Diagnostic (SFD) to be merged in a structurally superior position relative to its VP-internal co-argument(s). Furthermore, I show that an Instrumental Inanimate Causer NP is distinguished from the (morphologically indistinguishable) Instrumental Agent NP in passives in terms of scope behavior, which supports \citeauthor{Lavine2022}'s (\citeyear{Lavine2022,Lavine2023}) broader argument for the need to distinguish between agent/animate causers, which are arguments of Voice, and a variety of inanimate causers, argued to merge lower in the structure. 

The remainder of the paper is structured as follows. \sectref{sec:antonyuk:2} presents empirical data from Ukrainian that demonstrate the Scope Freezing Generalization and introduces the Scope Freezing Diagnostic that is based on it. \sectref{sec:antonyuk:3} is concerned with mapping out the verb phrase by applying the SFD to a variety of constructions involving Accusative, Dative and Instrumental arguments. It shows, i.a., that the relative ordering of internal arguments at Merge differs significantly from that often considered ``standard" in the field (\sectref{sec:antonyuk:3.1}). The rest of \sectref{sec:antonyuk:3} then goes deeper into the syntax of Instrumentals by contrasting their scope behavior with that of the Instrumentally case-marked subjects of passives, providing new evidence against the Undifferentiated Initiator view and thus offering support to \citeauthor{Lavine2022}'s (\citeyear{Lavine2022,Lavine2023}) position (\sectref{sec:antonyuk:3.2.1}) according to which Inanimate Causer Instrumentals are merged lower than Agents. \sectref{sec:antonyuk:3.2.2} argues that the Merge position of Inanimate Causer Instrumentals is at the same time much higher than usually assumed, specifically, Spec,\textit{v}Cause (rather than the often assumed Complement of V position). \sectref{sec:antonyuk:4} offers my conclusions.

\section{Scope freezing in Ukrainian: The empirical domain}\label{sec:antonyuk:2}

While taking such an approach to scope freezing (i.e., one deducing what scope freezing is from patterns of alternating word order-QP scope distribution) makes good scientific sense in general, it is arguably especially important as far as Slavic languages are concerned, for, as shown in \citet{Antonyuk2015} and elaborated in \citet{Antonyuk2023,AntonyukUnderReview}, unlike English, where there appear to be only two constructions that exhibit frozen surface scope, the phenomenon of frozen surface scope is not only found in (East) Slavic as well, but it is found in a considerably broader range of constructions.\footnote{ \textrm{The phenomena discussed here are observed in many Slavic languages across the subgroups. See \citet{Antonyuk2015, Antonyuk2020,Antonyuk2023} and \citet{BonehNash2017} on Russian. See \citet{AntonyukMykhaylyk2022} on the interaction of quantification and Object Shift in Ukrainian. See also \citet{MarvinStegovec2012} for a brief discussion of scope freezing in Slovenian DOCs and \citet{AbelsGrabska2022} for a detailed discussion of scope distribution in Polish ditransitives.} } What all these constructions have in common is the pattern, schematized in \REF{ex:antonyuk:6}, whereby one linearization of internal arguments is scopally ambiguous while the opposite order of arguments in the postverbal field is surface scope frozen, allowing only the QP scope interpretation that corresponds to overt c-command relations. 

\ea%6
    \label{ex:antonyuk:6}

  \ea{}  [\textsubscript{TP} EA [ <EA>  V$+$v  [\textsubscript{VP} QP2  <V>  QP1 ]]]      \hfill scope ambiguous\\
\ex{} [\textsubscript{TP} EA [ <EA>  V$+$v [\textsubscript{XP} QP1  <V> [\textsubscript{VP} QP2  <V>  QP1 ]]]] \\  \hfill scope frozen*\textsuperscript{,}\footnote{\textrm{When QP1 $=$ PP}\textrm{\textsubscript{LOCATIVE/DIRECTIONAL} }\textrm{the result of argument permutation is surface scope bias, not surface scope freezing. See \citet{Antonyuk2020, AntonyukUnderReview} for a relevant discussion.} }\\
*where QP1 ${\neq}~$PP\textsubscript{LOCATIVE/DIRECTIONAL}
\z
\z


\noindent According to \REF{ex:antonyuk:6}, the conditions under which scope freezing arises are quite specific and limited, that is, scope freezing arises whenever the structurally lower internal argument raises overtly across its structurally superordinate co-argument to a c-commanding position in the postverbal field. The word order alternations below provide some of the empirical basis for \REF{ex:antonyuk:6}.\footnote{See \citet{Dyakonova2009} and \citet{Bailyn2010, Bailyn2012} on Russian ditransitives and the discussion of the relation between the two linearizations. Both authors agree they instantiate the two frames of the Ditransitive Alternation.}

\ea%7
The Ditransitive Alternation
    \label{ex:antonyuk:7}
\ea
\gll   Myxajlo   po-daruvav   jakus’ igrašku     kožnij kišci.    \\
Myxajlo.\textsc{nom}     \textsc{po}-gift.\textsc{pst} some.\textsc{acc} toy.\textsc{acc}    every.\textsc{dat} cat.\textsc{dat}\\
\glt ‘Myxajlo gifted some toy to every cat.’

\textit{Surface scope}: For some toy \textit{x}, for every cat \textit{y}, Myxajlo gifted \textit{x} to \textit{y} (i.e., the same toy).

\textit{Inverse scope}: For every cat \textit{x}, for some toy \textit{y}, Myxajlo gifted \textit{x y} (i.e., different cat--toy pairs).

\ex
\gll Myxajlo   po-daruvav     jakijs’ kišci   kožnu igrašku.    \\
  Myxajlo.\textsc{nom}     \textsc{po}-gift.\textsc{pst} some.\textsc{dat} cat.\textsc{dat}   every.\textsc{acc} toy.\textsc{acc}\\
\glt `Myxajlo gifted some cat every toy.’

\textit{Surface scope}: For some cat \textit{x}, for every toy \textit{y}, Myxajlo gifted \textit{x y}.

\textit{Unavailable inverse scope}: For every toy \textit{x}, for some cat \textit{y}, Myxajlo gifted \textit{x} to \textit{y}.
    \z
\z


\ea%8
    \label{ex:antonyuk:8}
The \textit{Spray-Load} Alternation

 \ea \label{ex:antonyuk:8a}
\gll    Myxajlo   za-lyv     \minsp{[} jakyjs’ vyd pal’noho]   \minsp{[} v kožen bak].\\
  Myxajlo.\textsc{nom}   \textsc{za}-fill.\textsc{pst} {} some.\textsc{acc} type.\textsc{acc} gas.\textsc{gen}  {} into every.\textsc{acc} tank.\textsc{acc}\\
\glt `Myxajlo filled some type of gas into every tank.’

\textit{Surface scope}: For some type of gas \textit{x}, for every tank \textit{y}, Myxajlo filled \textit{x} into \textit{y}.

\textit{Inverse scope}: For every tank \textit{x}, for some type of gas \textit{y}, Myxajlo filled \textit{x} with \textit{y} (i.e., possibly different type of gas for each tank).

\ex \label{ex:antonyuk:8b}
\gll    Myxajlo   za-lyv     \minsp{[} jakyjs’ bak]     \minsp{[} kožnym vydom pal’noho].\\
  Myxajlo.\textsc{nom}   \textsc{za}-fill.\textsc{pst} {} some.\textsc{acc} tank.\textsc{acc}   {} every.\textsc{ins} type.\textsc{ins} gas.\textsc{gen}\\
\glt `Myxajlo filled some tank with every type of gas.’

\textit{Surface scope}: For some tank \textit{x}, for every type of gas \textit{y}, Myxajlo filled \textit{x} with \textit{y}.

\textit{Unavailable inverse scope}:  For every type of gas \textit{x}, for some tank \textit{y}, Myxajlo filled \textit{x} into \textit{y}.

    \z
\z

\newpage
\ea%9
    \label{ex:antonyuk:9}
Reflexive Causatives

 \ea
\gll   Likar   infikuvav-sja   jakojus’ xvoroboju   vid kožnoho pacijenta.\\
  doctor.\textsc{nom}   infect.\textsc{pst-refl}  some.\textsc{ins} illness.\textsc{ins}   from every.\textsc{gen} patient.\textsc{gen}\\
\glt `The doctor got infected from every patient with some illness.’

\textit{Surface scope}: For some illness \textit{x}, for every patient \textit{y}, the doctor got infected with \textit{x} from \textit{y}.

\textit{Inverse scope}: For every patient \textit{x}, for some illness \textit{y}, the doctor got inflected by \textit{x} with \textit{y}.

\ex
\gll Likar   infikuvav-sja   vid jakohos’  pacijenta   kožnoju xvoroboju.\\
  doctor.\textsc{nom}   infect.\textsc{pst-refl} from some.\textsc{gen} patient.\textsc{gen}    every.\textsc{ins} illness.\textsc{ins}\\
\glt `The doctor got infected with some illness from every patient.’

\textit{Surface scope}: For some patient \textit{x}, for every illness \textit{y}, the doctor got infected by \textit{x} with \textit{y}.

\textit{Unavailable inverse scope}: For every illness \textit{x}, for some patient \textit{y}, the doctor got infected with \textit{x} by \textit{y}.
    \z
\z

\ea%10
    \label{ex:antonyuk:10}
Causative verbs
\ea \label{ex:antonyuk:10a}
\gll   Likar   infikuvav   jakojus’ xvoroboju   kožnoho pacijenta.\\
  doctor.\textsc{nom} infect.\textsc{pst}  some.\textsc{ins} illness.\textsc{ins}  every.\textsc{acc} patient.\textsc{acc}\\
\glt `The doctor infected every patient with some illness.’

\textit{Surface scope}: For some illness \textit{x}, for every patient \textit{y}, the doctor infected \textit{y} with \textit{x}.

\textit{Inverse scope}: For every patient \textit{x}, for some illness \textit{y}, the doctor infected \textit{x} with \textit{y}.

\ex \label{ex:antonyuk:10b}
\gll   Likar   infikuvav   jakohos’ pacijenta   kožnoju xvoroboju.\\
  doctor.\textsc{nom} infect.\textsc{pst}  some.\textsc{acc} patient.\textsc{acc}  every.\textsc{ins} illness.\textsc{ins}\\
\glt `The doctor infected some patient with every illness.’

\textit{Surface scope}: For some patient \textit{x}, for every illness \textit{y}, the doctor infected \textit{x} with \textit{y}.

\textit{Unavailable inverse scope}: For every illness \textit{x}, for some patient \textit{y}, the doctor infected \textit{y} with \textit{x}.
    \z
\z

\noindent The key observation to note here is, of course, that the change in scope interpretations in each pair (i.e., the appearance of scope freezing in the (b) examples) tracks the change in overt word order, which constitutes the foundation for the \textsc{Scope Freezing Generalization} in \REF{ex:antonyuk:11}, according to which scope freezing is an inherently derivational, in fact, \textit{a derivation-by-movement} phenomenon.\footnote{\textrm{The importance of this conclusion should be immediately clear: It implicates a much greater degree of derivationality inherent in the derivation of numerous structures and argument structure alternations than is currently assumed.}} 


\eanoraggedright%11
    \label{ex:antonyuk:11}

           \textsc{The Scope Freezing Generalization} (SFG)  \citep{Antonyuk2015}


Scope freezing obtains when one QP raises overtly across another to a \mbox{c-commanding}´ position as a result of a single instance of movement within the \textit{v}P/VP.
    \z

\noindent That scope freezing must be limited to the domain of \textit{v}P/VP (i.e., below VoiceP in more modern terminology) becomes obvious once the above examples are compared to those involving a subject QP: Here we observe that the interaction between an object QP and a subject QP never amounts to a frozen scope relation.\footnote{Building on the original observation in \citet{Larson1990}, \citet{Bruening2001} provides independent evidence in favor of scope freezing being limited to object QPs, never an object and a subject QP. Thus, conclusions based on the richer East Slavic empirical data are fully aligned with the conclusions based on the English data. As argued especially in \citet{AntonyukUnderReview}, since the scope freezing found in Ukrainian and Russian is arguably identical to the English case, the conclusions reached on the basis of the former should be viewed as naturally extending to the latter.}

\ea%12
    \label{ex:antonyuk:12}
Simple SVO transitives: No scope freezing
%\ea 

\gll          Jakas’   divčynka   na-hoduvala     kožnu kišku.    \\
  some.\textsc{nom}   girl.\textsc{nom} \textsc{na}-feed.\textsc{pst.f}   every.\textsc{acc} cat.\textsc{acc}\\
\glt `Some girl fed every cat.’

\textit{Surface scope}: For some girl \textit{x}, for every cat \textit{y}, \textit{x} fed \textit{y}.

\textit{Inverse scope}: For every cat \textit{x}, for some girl \textit{y}, \textit{x} was fed by \textit{y}.
\z


%OSV (locally scrambled) clauses: No scope freezing


\ea%13
\label{ex:antonyuk:13}
OSV (locally scrambled) clauses: No scope freezing

\gll           Jakus’ kišku     kožna divčynka   na-hoduvala.  \\
  some.\textsc{acc} cat.\textsc{acc}     every.\textsc{nom.f} girl.\textsc{nom.f}  \textsc{na}-feed.\textsc{pst.f}\\
\glt `Some cat, every girl fed.’

\textit{Surface scope}: For some cat \textit{x}, for every girl \textit{y}, \textit{x} was fed by \textit{y}.

\textit{Inverse scope}: For every girl \textit{x}, for some cat \textit{y}, \textit{x} fed \textit{y}.
\z

%OVS clauses: no scope freezing

\largerpage[2]
\ea%14
    \label{ex:antonyuk:14}
OVS clauses: No scope freezing

\gll           Jakus’ kišku     na-hoduvala     kožna divčynka.\\
  some.\textsc{acc} cat.\textsc{acc}   \textsc{na}-feed.\textsc{pst.f} every.\textsc{nom.f} girl.\textsc{nom.f}\\
\glt `Some cat was fed by every girl.’

\textit{Surface scope}: For some cat \textit{x}, for every girl \textit{y}, \textit{x} was fed by \textit{y}.

\textit{Inverse scope}: For every girl \textit{x}, for some cat \textit{y}, \textit{x} fed \textit{y}.
\z\clearpage

\noindent While the absence of scope freezing in OVS is important, showing that object QP–subject QP pairs do not result in scope freezing under any circumstances, much will depend on one’s analysis of OVS, which is a hotly contested issue in Slavic linguistics.\footnote{\textrm{For the sake of clarity, I assume the analysis of OVS proposed in \citet{Antonyuk2021}.}} Whatever derivation of OVS one may assume, however, the basic empirical observation is that subject QP $>>$ object QP pairs as well as object QP $>>$  subject QP pairs do not result in scope freezing. Coupled with the observation that direct objects readily participate in frozen surface scope configurations, the conclusion this invites is that VoiceP, the layer that on most assumptions introduces the external argument, is outside the domain within which scope freezing can be established. On the assumption that OVS clauses are derived by overtly raising the object phrase to a position preceding and c-commanding the subject, the SVO--OVS examples are particularly telling in this respect when compared with, e.g., the ditransitive alternation in \REF{ex:antonyuk:7} or, e.g., \REF{ex:antonyuk:10}, where the two orders are arguably also derived by an overt instance of movement raising the lower object to a position preceding and c-commanding the higher one. Whereas in the former cases the overt instance of movement still yields scope ambiguity, in the latter case what results in surface scope freezing.


Another argument in favor of the domain of scope freezing to exclude the subject layer comes from nominalizations (see esp. \citealt{Alexiadou2009, Alexiadou2017a}; \citealt{Borer2014}; see also \citealt{Chomsky1970}; \citealt{Grimshaw1990}; \citealt{Harley2009}; \citealt{Marantz1997}, i.a.). Specifically, the nominalizations in \REF{ex:antonyuk:15} and \REF{ex:antonyuk:16}, which exclude the external argument, still preserve the argument structure of the corresponding verbal layer of full sentences, with the scope relations between the internal arguments being preserved as well. Thus, once again, we see scope freezing reappear in one of the two possible linearizations of internal arguments in the absence of (an overt) subject:

%Nominalizations \todo{this should be sectioned somehow}
\ea%15
    \label{ex:antonyuk:15}

\ea
\gll  za-lyv-annja     jakohos’ vydu pal’noho   v kožen bak    \\
  \textsc{za}-pour-\textit{annja}.\textsc{nom}   some.\textsc{gen}   type.\textsc{gen}   gas.\textsc{gen}  into every.\textsc{acc} tank.\textsc{acc}\\
\glt `the pouring of some type of gas into every tank’\hfill (cf. \REF{ex:antonyuk:8a})

\textit{Surface scope}: For some type of gas \textit{x}, for every tank \textit{y}, there is a filling of \textit{x} into \textit{y}.

\textit{Inverse scope}: For every tank \textit{x}, for some type of gas \textit{y}, there is a filling of \textit{x} with \textit{y} (i.e., possibly different type of gas for each tank).

\ex
\gll    za-lyv-annja     jakohos’ baku     kožnym vydom pal’noho   \\
  \textsc{za}-pour-\textit{annja}.\textsc{nom}  some.\textsc{gen} tank.\textsc{gen}   every.\textsc{ins} type.\textsc{ins} gas.\textsc{gen}\\ 
\glt `the pouring of some tank with every type of gas’\hfill (cf. \REF{ex:antonyuk:8b})

\textit{Surface scope}: For some tank \textit{x}, for every type of gas \textit{y}, there is a filling of \textit{x} with \textit{y}.

\textit{Unavailable inverse scope}:  For every type of gas \textit{x}, for some tank \textit{y}, there is a filling of \textit{x} into \textit{y}.
    \z
\z

\ea%16
    \label{ex:antonyuk:16}

\ea
\gll   infikuv-annja     jakojus’ xvoroboju   kožnoho pacijenta     \\
infect-\textit{annja}.\textsc{nom}  some.\textsc{ins} illness.\textsc{ins}   every.\textsc{acc} patient.\textsc{acc}\\
\glt `infecting of every patient with some illness’\hfill (cf. \REF{ex:antonyuk:10a})

  (Lit.: the infecting with some illness of every patient)

\textit{Surface scope}: For some illness \textit{x}, for every patient \textit{y}, there is infecting of \textit{y} with \textit{x}.

\textit{Inverse scope}: For every patient \textit{x}, for some illness \textit{y}, there is infecting of \textit{x} with \textit{y}.

\ex
\gll    infikuv-annja     jakohos’ pacijenta   kožnoju xvoroboju    \\
infect-\textit{annja}.\textsc{nom}  some.\textsc{acc} patient.\textsc{acc}   every.\textsc{ins} illness.\textsc{ins}\\
\glt `the infecting of some patient with  every illness’\hfill (cf. \REF{ex:antonyuk:10b})

\textit{Surface scope}: For some patient \textit{x}, for every illness \textit{y}, there is infecting of \textit{x} with \textit{y}.

\textit{Unavailable inverse scope}: For every illness \textit{x}, for some patient \textit{y}, there is infecting of  \textit{y} with \textit{x}.
    \z
\z

\noindent Of course, \textsc{Complex Event Nominals} (CENs), such as the examples in \REF{ex:antonyuk:15} and \REF{ex:antonyuk:16}, probably do not really “exclude” the subject layer, as the data above may superficially suggest: the subject is implicit, as evidenced by the ability of such nominalizations to be modified by agentive modifiers such as \textit{navmysne} ‘deliberate’, as in \textit{navmysne infikuvannja} ‘deliberate infecting’. Thus, especially within DM, on various instantiations of the “Phrasal Layering” analysis of Complex Event Nominals, the verbal structure, including the layer that introduces the external argument, is included in nominalizations, with the nominalizer head little \textit{n} attaching on top of fully projected verbal structure (\citealt{Alexiadou2001,Alexiadou2017a}; \citealt{Bruening2013}; \citealt{Borer2014}; \citealt{Iordachioaia2020}; \citealt{McGinnis2020}, i.a.).\footnote{\textrm{See also \citet{Wood2023} for a detailed theoretical overview as well as an alternative proposal based on Icelandic data that CENs can be formed in syntax without nominalizing full verbal structure.}} 

This means that the examples in \REF{ex:antonyuk:15} and \REF{ex:antonyuk:16} do not really provide bullet-proof evidence that the subject is “excluded” is any real sense, merely that it appears to be inert and not participating in whatever processes result in the establishment of the surface scope freezing relation. Minimally, the account of scope freezing adopted here (spelled out in more detail in later sections) is committed to the following conclusion: the preservation of the scope freezing relation means that the layer of structure arguably (and crucially) implicated in scope freezing, \mbox{ApplP} (located between \textit{v}P and VoiceP), must be included in the verbal structure nominalized by \textit{n} (cf. \citealt{Wood2023}).

To summarize, we see that the SFG points to a derivational nature of scope freezing, resulting from an overt instance of movement I will henceforth refer to as \textsc{Argument Inversion} (following \citealt{AntonyukMykhaylyk2022}). Taking place in the postverbal field, Argument Inversion (AI) constitutes a local instance of overt syntactic movement, and, as examples such as the anaphor binding data below suggest, AI involves A-movement:\footnote{\textrm{The original examples are due to \citet{Asarina2005}, cited in \citet{Bailyn2012}.}}

\ea%17
    \label{ex:antonyuk:17}

\ea \label{ex:antonyuk:17a}
\gll   Dolja     po-daruvala     nas\textsubscript{i}    odyn odnomu\textsubscript{i}.     \\
Fate.\textsc{nom}   \textsc{po}-gift.\textsc{pst}    us.\textsc{acc}     each other.\textsc{dat}\\
\glt ‘Fate gifted us to each other.’

 \ex \label{ex:antonyuk:17b}
\gll   Dolja     po-daruvala     nam\textsubscript{i}    odyn odnoho\textsubscript{i}.\\
  fate.\textsc{nom}  \textsc{po}-gift.\textsc{pst}    us.\textsc{dat}    each other.\textsc{acc}\\
\glt `Fate gifted us each other.’
    \z
\z

\noindent The anaphor binding data provides clear evidence that A-movement is involved. Considering A-movement leads to new binding relations, however, it can be argued that the directionality of the derivation in fact goes in the opposite direction. That is, rather than assuming that \REF{ex:antonyuk:17a} represents the base order and \REF{ex:antonyuk:17b} is derived from it by overt movement, it has been argued that \REF{ex:antonyuk:17b} represents the base order, and \REF{ex:antonyuk:17a} obtains from overtly raising the direct object across the Dative-marked anaphor, thus establishing a new binding relation. 

While this type of argument has indeed been made many times, let me point out that it works much better with cases involving binding than it does with cases involving scope relations. This is so since the binding data in \REF{ex:antonyuk:17} is symmetric, i.e., both \REF{ex:antonyuk:17a} and \REF{ex:antonyuk:17b} can be explained on either type of account. With QP scope, things are very different. On the widely held assumption that the Dative argument in ditransitives is merged in a structurally higher position than the Accusative argument, the scope freezing familiar from the DAT $>>$ ACC ditransitives must be seen as an inherent property of the construction or, e.g., on \citeauthor{Bruening2001}'s (\citeyear{Bruening2001}) Superiority account, as an entirely epiphenomenal property. As pointed out in \citet{LarsonEtAl2019}, \citeauthor{Bruening2001}’s account thus predicts that scope freezing should be found in English in many more constructions, namely all of those where the base relation between the two QPs can be plausibly analyzed (as in \citealt{Bruening2001}) as involving asymmetric c-command in the base structure, -- contrary to fact. Since this prediction is thoroughly falsified, we are left with a conclusion that scope freezing is, perhaps, just an inherent property of certain constructions, at least, as far as English is concerned. That is, we are left with no way of predicting where else we might find scope freezing. And perhaps one could accept it on the grounds that this property is somehow “exceptional”, i.e., found in these two ditransitive constructions only. 

Drawing now on our insights from Ukrainian/East Slavic, it becomes clear that our inability to model scope freezing, to predict where else in the language we may find it, is, in fact, a problem, as the phenomenon is found in a significant number of constructions, thus posing a challenge for syntactic theory. On the view adopted here, on the other hand, not only do we have an extensive list of environments where scope freezing has already been found to obtain, the Scope Freezing Generalization allows us to predict potential further environments where scope can be expected to ``freeze'' in this way as well. While this result is significant, the SFG also provides us with a powerful diagnostic tool for probing argument structure relations and allows for other non-trivial insights into the derivation of the extended verbal domain in East Slavic. Thus, I conclude that the below schematization accurately describes the conditions on scope freezing for the language under discussion, Ukrainian:

\ea%18
    \label{ex:antonyuk:18}

          \textsc{Surface scope freezing}, schematized% (the domain of scope freezing highlighted)
          :{}



[\textsubscript{TP} ExtA\textbf{ }[\textsubscript{VoiceP} <ExtA> V$+$v [\textsubscript{XP} QP1 [\textsubscript{VP} QP2  <V>\textbf{  }<QP1>]]]]
    \z



\noindent In what follows then, I will be relying on the \textsc{Scope} \textsc{Freezing} \textsc{Diagnostic} (SFD), which allows us to probe relative argument structure relations at Merge \citep{Antonyuk2015}:

\eanoraggedright%19
    \label{ex:antonyuk:19}

           \textsc{The Scope Freezing Diagnostic}

Frozen surface scope implicates a derived structure resulting from Argument Inversion.
    \z

\noindent The methodological goal of this paper is to demonstrate the remarkable internal consistency of the diagnostic and the non-trivial insights into the syntax of argument structure alternations that can be gained by applying it. I will not attempt to demonstrate all the insights already derived from the application of the SFD to the East Slavic (Ukrainian and Russian) data for reasons of space, directing the reader to the original research papers where these results are presented. Here I will only summarize and briefly exemplify some of the findings so we can build on them in this paper.\footnote{See \citet{Antonyuk2015, Antonyuk2020, Antonyuk2023,AntonyukUnderReview} and \citet{AntonyukMykhaylyk2022} for the original data and findings. See \citet{AbelsGrabska2022} for an experimental confirmation of the empirical claims in Antonyuk’s work based on Polish QP scope data as well as for a critical engagement with the account of scope freezing assumed here.}

\section{Mapping out the verb phrase}\label{sec:antonyuk:3}

\subsection{The relative ordering of internal arguments at Merge}
\label{sec:antonyuk:3.1}

As should already be clear from our preliminary discussion, the SFD is a simple diagnostic, one which can (at most) point to a derived structure among the alternating frames under investigation (if such exists), as well as point to the relative ordering of arguments at Merge. The SFD cannot tell us much about the exact Merge position of arguments; what it does give us is a heuristic according to which the linearization that is surface-scope frozen must be the derived one, and moreover that it must be derived by raising the structurally lower of the two internal arguments overtly across the structurally higher one in the postverbal field, thus gaining c-command over it.\footnote{\citet{AntonyukUnderReview} argues that, according to the SFD (and contra \citealt{Antonyuk2015}), the \textit{Spray-Load} Alternation is, in fact, \textit{not} a case of a derivational relation between the two alternating frames. Crucially, she argues that the `with'-variant (i.e., the Instrumental case-marked frame in Slavic) is, nevertheless, derived, as suggested by the fact that it exhibits scope freezing, just not from the locative frame, but from the Instrumental frame via Argument Inversion of the Accusative argument across NP\textsubscript{INS}. That is, scope freezing in the `with'-variant is indeed a marker of its derived status, but what it is derived from is not what is traditionally viewed as its “alternating” frame. This, of course, is good news, as the two frames differ not only in the morphosyntactic marking on their two internal arguments, but also in their theta roles, making a derivational account problematic in this case.}\textsuperscript{,}\footnote{Significantly, argument structure alternations, in Slavic and elsewhere, appear to never exhibit scope freezing on both possible orders/alternating frames. On accounts which posit independent projection of the two alternating frames, this should in principle be a logical possibility. On \citeauthor{Antonyuk2015}'s (\citeyear{Antonyuk2015}) treatment of scope freezing, assumed here, such a situation is impossible in principle, for obvious reasons. The only conceivable exception to this would be if what were (mistakenly) considered to be an alternation consisted of two structures, each of which would be derivationally related to another structure that is not viewed as part of the alternation. The closest case to this hypothetical situation would in fact be the \textit{Spray-Load} Alternation, where (as described in the previous footnote) the frozen frame (V NP\textsubscript{ACC} $>>$ NP\textsubscript{INS}) is derivationally related to another structure, (i.e., V NP\textsubscript{INS} $>>$ NP\textsubscript{ACC}), rather than to the locative frame. Apart from this scenario, I will venture a prediction that such a situation should be impossible.\label{ft:antonyuk:14}} In other words, what the SFD gives us is a mere ordering of arguments at Merge. For the constructions discussed earlier, these relative ordering statements are as follows:

\begin{table}
\caption{Relative ordering statements}
\label{tab:relative ordering}

\begin{tabularx}{.9\textwidth}{lll}
 \lsptoprule
& Frozen order &   Base\\
 \midrule
 The Ditransitive Alternation: &   DAT $>>$ ACC  &    ACC $>>$ DAT\\
The \textit{Spray-Load} Alternation: &    ACC $>>$ INS &   INS $>>$ ACC \\

Reflexive Causatives:  &   PP\textsubscript{FROM} $>>$ INS  &  INS $>>$ PP\textsubscript{FROM}\\

Causative verbs:   &   ACC $>>$ INS &   INS $>>$ ACC\\
\lspbottomrule
        
\end{tabularx}
\end{table}


Now, to some working on argument structure alternations and verbal argument structure more generally these insights may be surprising, as they go against a lot of what has arguably been assumed to be settled in the literature. My goal in this paper is not to provide conclusive evidence in favor of a particular structure for a particular alternation, but rather to demonstrate the general insights afforded by the SFD, especially in what concerns the relative Merge position of Instrumental arguments. A larger point, hinted at here and developed in detail elsewhere \citep{Antonyukinpreparation} is that, taken together, these insights suggest a rather interesting alternative view of how the derivation of the verb phrase in Discourse Configurational Slavic languages may proceed and what confounding factors have thus far prevented us from seeing this picture. 

  With respect to the Instrumentals specifically, the suggestion that they might be merged higher than the Accusative argument is certainly surprising, both because it is common to assume a low Merge/complement of V position for the Instrumental NP (as is standard for the Oblique case-marked arguments), as well as because the alternative would also suggest the Accusative NP must be merged low (in any case, lower than NP\textsubscript{INS}). As already suggested in Table \ref{tab:relative ordering}, the indication that the Instrumental NP is merged higher than either the Accusative-marked co-argument or the PP co-argument remains consistent even once we significantly expand the range of the constructions under investigation. 

  \largerpage
Consider the following data (which include examples presented earlier, for convenience):

\ea%21
    \label{ex:antonyuk:21}
Causative verbs
\ea \gll   Likar   infikuvav   jakojus’ xvoroboju   kožnoho pacijenta.\\
  doctor.\textsc{nom} infect.\textsc{pst}  some.\textsc{ins} illness.\textsc{ins}  every.\textsc{acc} patient.\textsc{acc}\\
\glt `The doctor infected every patient with some illness.’

\textit{Surface scope}: For some illness \textit{x}, for every patient \textit{y}, the doctor infected \textit{y} with \textit{x}.

\textit{Inverse scope}: For every patient \textit{x}, for some illness \textit{y}, the doctor infected \textit{x} with \textit{y}.

\ex
\gll   Likar   infikuvav   jakohos’ pacijenta   kožnoju xvoroboju.\\
  doctor.\textsc{nom} infect.\textsc{pst}  some.\textsc{acc} patient.\textsc{acc}  every.\textsc{ins} illness.\textsc{ins}\\
\glt `The doctor infected some patient with every illness’

\textit{Surface scope}: For some patient \textit{x}, for every illness \textit{y}, the doctor infected \textit{x} with \textit{y}.

\textit{Unavailable inverse scope}: For every illness \textit{x}, for some patient \textit{y}, the doctor infected \textit{y} with \textit{x}.

\z 
$\Rightarrow$ \textsc{Base order}:  V NP\textsubscript{INS} $>>$ NP\textsubscript{ACC}
\z

\noindent According to the SFD, the V NP\textsubscript{INS} $>>$ NP\textsubscript{ACC} linearization represents the relative order of arguments at Merge. The same conclusion is reached for the `with'-variant of the \textit{Spray-Load} Alternation:



\ea%22
    \label{ex:antonyuk:22}
Instrumental/`with'-frame of the Spray-Load Alternation
\ea
\gll    Myxailo   za-lyv     jakyjs’ bak   kožnym vydom pal’noho.\\
  Myxailo.\textsc{nom}   \textsc{za}-fill.\textsc{pst}  some.\textsc{acc} tank.\textsc{acc}   every.\textsc{ins} type.\textsc{ins} gas.\textsc{gen}\\
 \glt Lit.: ‘Myxailo filled some tank with every type of gas.’

\textit{Surface scope} (${\exists}$>${\forall}$):  for some tank \textit{x}, for every type of gas \textit{y}, Myxailo filled \textit{x} with \textit{y}.

\textit{Unavailable inverse scope} (${\forall}$>${\exists}$):  for every type of gas \textit{x}, for some tank \textit{y}, Myxailo filled \textit{x} into \textit{y}.

\ex
\gll   Myxailo   za-lyv     jakymos’ vydom pal’noho     kožen bak.\\
  Myxailo.\textsc{nom}   \textsc{za}-fill.\textsc{pst}  some.\textsc{ins} type.\textsc{ins} gas.\textsc{gen}     every.\textsc{acc} tank.\textsc{acc}\\
  \glt Lit.: ‘Myxailo filled with some type of gas every tank.’

\textit{Surface scope} (${\exists}$>${\forall}$): For some type of gas \textit{x}, for every tank \textit{y}, Myxailo filled \textit{x} into \textit{y}.

\textit{Inverse scope }(${\forall}$>${\exists}$): For every tank \textit{x}, for some type of gas \textit{y}, Myxailo filled \textit{x} with \textit{y}.
    

\z
$\Rightarrow$ NP\textsubscript{INS} $>>$ NP\textsubscript{ACC} is the order at Merge within the Instrumental/`with' frame.
\z


\noindent The passive-like \textit{-no/-to} construction, taken up in detail in \citet{Lavine2022, Lavine2023} and illustrated in \REF{ex:noto}, is useful for our purposes in that it involves two internal arguments, hence it can be subject to the SFD as well. 

\ea%23
\label{ex:antonyuk:23}
    The -\textit{no}/\textit{{}-to} construction (\citealt{LavineFreidin2002})
    
\gll   Cerkvu   bulo spaleno     blyskavkoju.    \\
  church.\textsc{acc}  was.\textsc{non-agr} burned.down.\textsc{non-agr}  lightning.\textsc{ins}\\
\glt `The church was burned down by lightning.’\label{ex:noto}
    \z

\noindent What we see is that, as before, the two arguments are permutable, and overtly permuting/crossing the arguments in the postverbal field results in scope freezing of the NP\textsubscript{ACC} $>>$ NP\textsubscript{INS} order.\footnote{\label{ft:antonyuk:15}\textrm{I assume that the surface word order in such structures is derived by ultimately raising either of the arguments to Spec,TP. In the scopally frozen sentence, Argument Inversion first raises the lower argument, NP}\textrm{\textsubscript{ACC}}\textrm{, to a position to the left of its co-argument, NP}\textrm{\textsubscript{INS}}\textrm{, still in the postverbal field, which results in scope freezing. The thus inverted NP}\textrm{\textsubscript{ACC} }\textrm{then undergoes raising into Spec,TP. See \citet{AntonyukMykhaylyk2022} for evidence that scope freezing, once established by AI in the postverbal field, cannot be disturbed/“unfrozen” by further syntactic movement.}}

\ea%24
    \label{ex:antonyuk:24}

\ea \label{ex:antonyuk:24a}
\gll   Jakus’ cerkvu     bulo spaleno     kožnoju blyskavkoju / z blyskavok.\\
  some.\textsc{acc} church.\textsc{acc}   was.\textsc{non-agr} burned.down.\textsc{non-agr}  every.\textsc{ins} lightning.\textsc{ins} {} from lightning.\textsc{gen.pl}\\
\glt ‘Some church was burned down by every lightning.’

  \textit{Surface scope}: For some church \textit{x}, for every lightning \textit{y}, \textit{x} was burned down by \textit{y}.

  \textit{Unavailable inverse scope}: For every lightning \textit{x}, for some church \textit{y}, \textit{x} burned down \textit{y}.

\ex
\gll   Jakojus' blyskavkoju     bulo spaleno     kožnu cerkvu.\\
  some.\textsc{ins} lightning.\textsc{ins}    was.\textsc{non-agr} burned.down.\textsc{non-agr}   every.\textsc{acc} church.\textsc{acc}\\
\glt `Some church was burned down by every lightning.’

  \textit{Surface scope}: For some lightning \textit{x}, for every church \textit{y}, \textit{x} burned down \textit{y}.

  \textit{Inverse scope}: For every church \textit{x}, for some lightning \textit{y}, \textit{x} was burned down by \textit{y}.
 \z
 $\Rightarrow$ \textsc{Base order}: V NP\textsubscript{INS} $>>$ NP\textsubscript{ACC}
\z

\noindent Something worth pointing out is that the surface scope interpretation for \REF{ex:antonyuk:24a} is infelicitous in that it entails a situation where one and the same church was burned down by every one of the lightning strikes under consideration. Thus, if there were five lightning strikes during the night, every single one must have hit the church and contributed to its burning down for the sentence to be true on its surface scope interpretation. Yet, despite its real-world incongruent surface scope interpretation, which should have really facilitated the inverse scope reading, if it were available, to come through, the latter is nevertheless excluded for this sentence. This helps demonstrate one of the core insights about scope freezing: it is an all-or-nothing, categorical phenomenon, as opposed to surface scope bias, familiar from other contexts, which can be manipulated by the choice of lexical items, syntactic contexts, information-structural properties of the sentence, etc. This property is what arguably makes the Scope Freezing Diagnostic an incredibly reliable, internally consistent diagnostic (see esp. \citealt{Antonyuk2023}).

%Non-agreeing accusatives:\todo{section?}

The exact same conclusion about the higher Merge position of the Instrumental argument (relative to the Accusative) can be reached by applying the SFD to the so-called \textsc{non-agreeing accusatives} (\citealt{LavineFreidin2002}; \citealt{Lavine2022,Lavine2023}, i.a.). For clarity, the derivational path for the derived (\REF{ex:antonyuk:25a} and \REF{ex:antonyuk:26a}) is assumed to be largely identical to that described for the \textit{{}-no/-to} constructions in footnote \ref{ft:antonyuk:15}:

\ea%25
    \label{ex:antonyuk:25}
%Non-agreeing accusatives
\ea \label{ex:antonyuk:25a}
\gll   Soldata   po-ranylo     kuleju.           \\
  soldier.\textsc{acc}  \textsc{po}-wound.\textsc{non-agr}   bullet.\textsc{ins}\\
\glt `A soldier was wounded by a bullet.’\\
\hfill (modeled on \citealt{LavineFreidin2002})

\ex
\gll   Kuleju     po-ranylo     soldata.\\
  bullet.\textsc{ins}  \textsc{po}-wound.\textsc{non-agr}  soldier.\textsc{acc}\\
\glt `A soldier was wounded by a bullet.’
    \z
\z

\ea%26
    \label{ex:antonyuk:26}
\ea \label{ex:antonyuk:26a}
\gll   Jakohos’ soldata   po-ranylo     kožnoju kuleju.    \\
  some.\textsc{acc} soldier.\textsc{acc}  \textsc{po}-wound.\textsc{non-agr}  every.\textsc{ins} bullet.\textsc{ins}\\
\glt `Some soldier was wounded by every bullet.’

\textit{Surface scope}: For some soldier \textit{x}, for every bullet \textit{y}, \textit{x} was wounded with \textit{y}.

\textit{Unavailable inverse scope}: For every bullet \textit{x}, for some soldier\textit{y}, \textit{x} wounded \textit{y}.
\ex
\gll   Jakojus’ kuleju   po-ranylo     kožnoho soldata.\\
  some.\textsc{ins} bullet.\textsc{ins}  \textsc{po}-wound.\textsc{non-agr}  every.\textsc{acc} soldier.\textsc{acc}\\
\glt `Every soldier was wounded by a bullet.’

\textit{Surface scope}: For some bullet \textit{x}, for every soldier \textit{y}, \textit{x} wounded \textit{y}.

\textit{Inverse scope}: For every soldier \textit{x}, for some bullet \textit{y}, \textit{x} was wounded with \textit{y}.

\z
$\Rightarrow$ \textsc{Base order}: V NP\textsubscript{INS} $>>$ NP\textsubscript{ACC}
\z


\noindent The non-agreeing accusatives involve two internal arguments, marked for Dative and Accusative case, allowing us to test whether the relative ordering V NP\textsubscript{ACC} $>>$ NP\textsubscript{DAT} obtained for the DOC will be replicated. As shown in \REF{ex:antonyuk:28}, this is indeed the case:

\ea%27
    \label{ex:antonyuk:27}
\ea
\gll   Xlopcevi   vidrizalo    palec’    na ruci. \\
  boy.\textsc{dat}    severed.\textsc{non-agr} finger.\textsc{acc} on hand\\
\glt `The boy’s finger was severed.’
\ex
\gll   Palec’     na ruci     vidrizalo     xlopcevi.\\
  finger.\textsc{acc}  on hand  severed.\textsc{non-agr}  boy.\textsc{dat}\\
\glt `A finger on hand was severed from a guy’s hand.’
\z
\z

\ea%28
\label{ex:antonyuk:28}
\ea
\gll  Jakomus’ xlopcevi   vidrizalo     kožen palec’     \minsp{(} na ruci).\\
some.\textsc{dat} guy.\textsc{dat} severed.\textsc{non-agr}  every.\textsc{acc} finger.\textsc{acc} {} on hand\\
\glt ‘Some boy got every one of his fingers severed.’\\
\textit{Surface scope}: For some boy \textit{x}, for every finger \textit{y}, \textit{x} had \textit{y} severed from \textit{x}’s hand.\\
\textit{Unavailable inverse scope}: For every finger \textit{x}, for some boy \textit{y}, \textit{x} was severed from \textit{y}’s hand.
\ex
\gll Jakyjs’ palec’     \minsp{(} na ruci)   vidrizalo    kožnomu xlopcevi.\\
some.\textsc{acc} finger.\textsc{acc} {} on hand   severed.\textsc{non-agr}  every.\textsc{dat} boy.\textsc{dat}\\
\glt  ‘Some finger was severed from every boy’s hand.’\\
\textit{Unavailable surface scope}: For some finger \textit{x}, for every boy \textit{y}, \textit{x} was severed from \textit{y}’s hand.\\
\textit{ Inverse scope}: For every boy \textit{x}, for some finger \textit{y}, \textit{x}’s y was severed.
\z
$\Rightarrow$ \textsc{Base order}: V NP\textsubscript{ACC} $>>$ NP\textsubscript{DAT}
\z

\noindent Now, it is interesting that up to now we have seen two of the relevant “Merge orderings”, that is, V NP\textsubscript{ACC} $>>$ NP\textsubscript{DAT} and V NP\textsubscript{INS} $>>$ NP\textsubscript{ACC,} established and replicated over several types of constructions here and elsewhere. By transitivity, NP\textsubscript{INS} should precede NP\textsubscript{DAT} as well: V NP\textsubscript{INS} $>>$ NP\textsubscript{DAT}. Since the example in \REF{ex:antonyuk:27} can be spelled out more fully, to include an implicit Instrumental Inanimate Causer argument, this new relative ordering, due to transitivity, can be put to the test by applying the SFD to the Instrumental and the Dative-marked quantificational arguments. Doing so confirms that V NP\textsubscript{INS} $>>$ NP\textsubscript{DAT} is, indeed, the correct base structure ordering:

\ea%29
    \label{ex:antonyuk:29}
\gll            Xlopcevi   vidrizalo    palec’    elektryčnoju pyloju.\\
boy.\textsc{dat}   severed.\textsc{non-agr} finger.\textsc{acc} electric.\textsc{ins} saw.\textsc{ins}\\
\glt ‘The boy’s finger got severed by an electric saw.’
    \z


\ea%30
    \label{ex:antonyuk:30}

\ea
\gll   Jakomus’ xlopcevi   vidrizalo    palec’    kožnym instrumentom.\\
some.\textsc{dat} guy.\textsc{dat} severed.\textsc{non-agr}  finger.\textsc{acc} every.\textsc{ins} instrument.\textsc{ins}\\
\glt ‘Some boy got a finger severed by every instrument.’ \hfill (frozen)


\ex
\gll   Jakymos’ instrumentom   vidrizalo     palec’     kožnomu xlopcevi.\\
some.\textsc{ins} instrument.\textsc{ins}    severed.\textsc{non-agr} finger.\textsc{acc} every.\textsc{dat} boy.\textsc{dat}\\
\glt ‘With some instrument (or other), every boy’s finger was severed.’ \\\hfill (ambiguous)
\z
$\Rightarrow$ \textsc{Base order}: V NP\textsubscript{INS} $>>$ NP\textsubscript{DAT}
\z

\noindent We have thus arrived at the relative “Merge ordering” for 3-argument NPs, summarized in \REF{ex:antonyuk:31}. The SFD makes abundantly clear that the NP\textsubscript{INS} argument cannot be equated with other obliques in occupying the lowermost, complement position (cf. \citealt{Pesetsky1995}, i.a.). In fact, according to the SFD, NP\textsubscript{INS} is merged the highest of the three internal arguments.


\ea \label{ex:antonyuk:31}
  V NP\textsubscript{ACC} $>>$ NP\textsubscript{DAT}\\
V NP\textsubscript{INS} $>>$ NP\textsubscript{ACC} \\
V NP\textsubscript{INS} $>>$ NP\textsubscript{DAT}
\z


\noindent Thus, whatever the actual Merge positions (something the SFD does not and cannot provide an answer to), the relative structure representation we arrive at looks like the following:


\ea%32
    \label{ex:antonyuk:32}

          V NP\textsubscript{INS} $>>$ NP\textsubscript{ACC} $>>$ NP\textsubscript{DAT}

\z


\subsection{The syntax of Instrumentals in Ukrainian}
\label{sec:antonyuk:3.2}


While the above relative ordering is all the SFD can give us, it is plenty, of course. Supplementing with other types of evidence (see \citealt{AntonyukMykhaylyk2022}; \citealt{AntonyukUnderReview} for arguments based on Ukrainian data; see also \citealt{Bailyn2010,Bailyn2012}; \citealt{Antonyuk2015,Antonyuk2020,Antonyuk2023}; \citealt{Titov2017} for related data from Russian; \citealt{Kovacevic2020} for Serbo-Croatian, i.a.), we get a fairly clear idea of what the actual Merge positions must be. Most relevantly for our purposes, the SFD evidence provided here complements the argumentation and the analysis of Inanimate Causers and the Split Voice structure (see \citealt{Pylkkänen2002,Pylkkänen2008}). In fact, we can adduce additional supporting evidence, also due to QP scope data, that strongly supports \citeauthor{Lavine2022}'s (\citeyear{Lavine2022, Lavine2023}) argumentation for the differentiated structural representation of the Instrumental case-marked Agents of passives and the Inanimate Causer Instrumentals. 



\subsubsection{On passives vs passive-like: Against the “Undifferentiated Initiator” view}
\label{sec:antonyuk:3.2.1}


\citet{Lavine2022, Lavine2023} argues against the “Undifferentiated Initiator” idea advanced in \citet{ant:Ramchand2008}; \citet{Bruening2013}; \citet{Legate2014} and \citet{Wood2017}. Examining a range of causative constructions in the crosslinguistic perspective, Lavine argues that the “Initiator” argument in the constructions under investigation is an Oblique Causer (Natural Force) argument that originates in the VP. This Natural Force Instrumental crucially licenses the presence of \textit{v}CauseP in the structure, which licenses Accusative case on the direct object in the absence of an external argument. While I will come back to the question of the precise Merge position for the Instrumental Inanimate Causer, as this is the question where the analysis proposed here differs from \citeauthor{Lavine2023}'s, let us for now focus on the latter point, namely the argument against the Undifferentiated Initiator position of \citet{ant:Ramchand2008} and subsequent work. Here, \citeauthor{Lavine2023}'s conclusions, which he reaches based on independent types of evidence, receive strong support from the quantifier scope data presented here.\footnote{\textrm{\citet{Lavine2023} takes on the theoretical question of whether Agents, Instruments and Natural Forces are all realizations of the “macro-Initiator role” and provides evidence that crosslinguistically, causer arguments are differentially realized in the syntax. Some of the strongest evidence to this effect comes from the differences in the distribution of the Inanimate (Instrumental) Causer arguments in Ukrainian, Polish, and Icelandic that sets them apart from the Agents of passives in these languages.}} Perhaps the strongest argument for limiting the domain of application of the Scope Freezing Generalization to \textit{v}P (i.e., crucially the layer below VoiceP) has been the absence of scope freezing in doubly quantified transitive SVO sentences (i.e., subject QP $>>$ V $>>$ object QP structures) as well as any other constructions involving a subject QP. Crucially, passive sentences do not exhibit scope freezing either (cf. \REF{ex:antonyuk:33} and the related OVS structure in \REF{ex:antonyuk:14}).

\ea%33
    \label{ex:antonyuk:33}
\gll           Jakas’ kiška     bula nahodovana     kožnoju divčynkoju.\\
  some.\textsc{nom.f} cat.\textsc{nom.f}   was.\textsc{f} fed.\textsc{pass.f} every.\textsc{ins} girl.\textsc{ins}\\
\glt `Some cat was fed by every girl.’

\textit{Surface scope}: For some cat \textit{x}, for every girl \textit{y}, \textit{x} was fed by \textit{y}.

\textit{Inverse scope}: For every girl \textit{x}, for some cat \textit{y}, \textit{x} fed \textit{y}.
    \z

\noindent \citeauthor{Lavine2023}'s analysis of Transitive Impersonals (which in Ukrainian happen to be morphologically indistinguishable from the Instrumental Agent of passives) makes a strong prediction regarding QP scope. It predicts that a passive structure and a superficially similar impersonal construction will differ with respect to the availability of scope freezing. This is so since on \citeauthor{Lavine2023}'s analysis the NP\textsubscript{INS} in an impersonal construction is merged lower than the merge position of an Instrumental Agent NP (assumed to be an argument of Voice). Hence, the former, but not the latter, is predicted to show scope freezing. Moreover, based on the data we have already seen, we can predict that transitive impersonals will show scope freezing in the ACC $>>$ INS order of arguments. Both predictions are correct (cf. \REF{ex:antonyuk:35a} and \REF{ex:antonyuk:36a}):\footnote{\textrm{As detailed in \citet{Lavine2022, Lavine2023}, a passive and a transitive impersonal differ with respect to the type of Instrumental NP they require: thus, a passive structure is infelicitous with an Inanimate Causer argument while a transitive impersonal is ungrammatical with an animate/Agent argument. This fact provides one of the arguments in favor of a differentiated approach to Instrumental NPs in Ukrainian.}}

\ea%34
    \label{ex:antonyuk:34}

\ea
Passive\\
\gll    Cerkv-a   bul-a spalen-a       \minsp{\{\#} blyskavk-oju / okupant-amy\}. \\
  church.\textsc{f}-\textsc{nom}   was-\textsc{agr.f} burned.down.\textsc{pass-agr.f} {} lightning-\textsc{ins} {} invaders-\textsc{ins}\\
\glt (Intended:) ‘The church was burned down by \{the lightning / the invaders\}.’

 \ex
 Trans. impers.\\
\gll   Cerkv-u                           bul-o                spalen-o                  \minsp{\{} blyskavk-oju           /  \minsp{\#}~okupant-amy\}.  \\
        church.\textsc{f}-\textsc{acc}   was-\textsc{non-agr} burned.down-\textsc{non-agr}  {}     lightning-\textsc{ins} {} ~~invaders-\textsc{ins}\\
\glt (Intended:) ‘The church was burned down by \{the lightning / the invaders\}.’
    \z
    \z


\ea%35
    \label{ex:antonyuk:35}

\ea \label{ex:antonyuk:35a}
Passive\\
\gll    Jakas’ cerkv-a     bul-a spalen-a       kožnym okupantom.\\
  some.\textsc{nom.sg.f} church.\textsc{f}-\textsc{nom}   was-\textsc{agr.f} burned.down.\textsc{pass-agr.f}  every.\textsc{ins} invader.\textsc{ins}\\
\glt ‘Some church was burned down by every invader.’    \hfill   (scope ambiguous)

\ex
\gll    Jakymos’ okupantom   bul-a spalen-a       kožn-a cerkv-a.\\
some.\textsc{ins.sg} invader.\textsc{ins}  was-\textsc{agr.f} burned.down.\textsc{pass-agr.f} every-\textsc{f.nom} church.\textsc{f-nom}\\
\glt ‘Every church was burned down by some invader.’  \hfill     (scope ambiguous)\\
    \z
    \z

\ea%36
    \label{ex:antonyuk:36}

\ea \label{ex:antonyuk:36a}
 Trans. impers.\\
\gll    Jak-us’ cerkv-u   bul-o spalen-o       kožn-oju blyskavk-oju. \\
  some-\textsc{acc.sg.f} church.\textsc{f-acc}  was-\textsc{non-agr} burned.down-\textsc{non-agr}  every-\textsc{ins} lightning-\textsc{ins}\\
\glt `Some church was burned by every lightning.’     \hfill (scope frozen)

\ex
\gll   Jak-ojus’ blyskavk-oju   bul-o spalen-o       kožn-u cerkv-u.\\
  some-\textsc{ins.sg} lightning-\textsc{ins}    was-\textsc{non-agr} burned.down-\textsc{non-agr}  every-\textsc{acc.f} church.\textsc{f-acc}\\
\glt `Every church was burned down by some lightning.’\\  \hfill    (scope ambiguous)
    \z
    \z


\noindent Of course, we want to ask whether differences regarding scope freezing availability in the two structures are indeed due to different Merge positions for the Instrumental Inanimate Causer argument and the Agent argument of a passive sentence. Needless to say, if we could attribute the lack of scope freezing in passives to some other factor, the above argument would disappear or be significantly weakened. Consider the facts again. Both the subjects of active SVO sentences and the subjects of passives exhibit lack of scope freezing, as do all other constructions in which one of the two interacting QPs is the subject QP. In other words, while there are indeed non-trivial structural, semantic, and information-structural differences between, e.g., OSV, OVS, and passives, conspicuously, the absence of scope freezing is a property they all share. Furthermore, all internal arguments irrespective of case marking and structural height (i.e., ACC/DAT/INSTR) invariably participate in the scope ambiguity-scope freezing distribution patterns under AI, which is to say that the lack of scope freezing cannot be attributed to some confound that is due to the presence of any of these arguments.\footnote{\textrm{As mentioned earlier, the only exception here is the locative/directional PPs, which consistently show surface scope bias, but not the categorical scope freezing. See \citet{Antonyuk2020, AntonyukUnderReview} for detailed discussions and the argument that the existence of this seeming exception to the Scope Freezing Generalization is not a problem for the SFG, but a source of additional insight in the search for an adequate, sufficiently restrictive account of scope freezing as a grammatical phenomenon.}} Thus, in a passive structure (which, again, differs in terms of scope freezing from the superficially similar transitive impersonal), the difference can hardly be due to anything other than the status of the Agent QP.



One may also wonder whether subjects fail to participate in scope freezing relations due to any of the semantic properties that a subject QP may be associated with by virtue of its syntactic and semantic prominence (e.g., givenness, specificity, topicality, etc). I believe this is unlikely: research on the interaction of specificity and quantification in Ukrainian shows that specificity and QP scope diverge, i.e., a(n object) QP can be specific and take either high or low scope; more generally, further syntactic movement and specificity of either of the two or both objects can neither perturb existing scope freezing nor establish a new scope freezing relation once a QP raises above low temporal/manner adverbs, generally taken to mark the \textit{v}P edge (in current terms: VoiceP edge).\footnote{\textrm{See \citet{AntonyukMykhaylyk2022} for details.}} This strongly suggests that the status of a subject QP as specific would likewise not interfere with its ability to participate in a scope freezing relation. From what I can tell, the same concerns givenness and topicality, though of course the status of the structurally higher QP as topical/given/specific can lead to some surface scope bias, meaning that the wide scope for the topical/given/specific subject QP may be the preferred interpretation in such a case.



On the Undifferentiated Initiator view all causer arguments, including the non-volitional ones, are arguments of Voice (\citealt{Kallulli2006}, \citealt{ant:Ramchand2008}, \citealt{Bruening2013}, \citealt{Legate2014}, \citealt{Wood2017}, i.a.). Thus, on these accounts, we expect that the Instrumental case-marked agents of passives and the Instrumental inanimate/ non-volitional causers of Transitive Impersonal constructions would be syntactically identical, precisely because of their being generated in the same position, Spec,VoiceP. As we have just observed, their behavior is certainly not identical as far as QP scope relations are concerned. Recall that on the account adopted here, the scope freezing relation obtains as a result of a single instance of movement of the structurally lower internal argument QP to a position above the structurally higher one, but, crucially, below the Merge position of the external argument. Thus, in the impersonal \textit{{}-no/-to} constructions, scope freezing obtains when the lower NP\textsubscript{ACC} raises overtly above NP\textsubscript{INS}. If we assume the generalization regarding scope freezing is correct, then on the Undifferentiated Initiator view, which has all initiators merged as arguments of Voice, \textit{all} structures involving any type of initiator argument are then predicted to lack scope freezing, precisely because, as we have shown, scope freezing emerges in a layer of structure that crucially excludes VoiceP. 



Needless to say, of course, much will ultimately depend on the account of passives one adopts, and this is certainly a research area characterized by continued debate as well as significant differences in the general approach of individual researchers. Let us focus here on the structural, configurational accounts in which the agent phrase of a passive is base-generated in the same position as in the active, in accordance with \citeauthor{Baker1988}'s (\citeyear{Baker1988}) UTAH (see esp. \citealt{Collins2005}; cf. \citealt{Bruening2013}; \citealt{Hallman2021}, i.a.). On \citeauthor{Collins2005}' (\citeyear{Collins2005}) influential account, a participial phase containing the VP fronts around the agent in its Merge position, thus inverting the hierarchical relation between the agent and the object phrase. On an account of passives roughly such as this one, the lack of scope freezing in passives follows naturally. This is so since the agent is merged in its usual position, Spec,VoiceP (assumed to be Spec,\textit{v}P in \citealt{Collins2005}), and we have seen extensive evidence that external arguments never participate in the surface scope freezing relation, thus we expect the same in passives. On the present account, which follows \citet{Lavine2022, Lavine2023}, what sets the non-volitional/inanimate causer Instrumentals apart from the agent Instrumentals of passives is precisely the structural difference in Merge positions, with the former being merged in a structurally lower position than the latter, hence the correctly predicted differences in scope behavior.\footnote{\textrm{A question remains, of course, why the nature of a QP’s Merge position should be relevant for establishing a scope freezing relation. While I cannot do this question justice here, in my related work (\citealt{AntonyukUnderReview}; \citealt{Antonyukinpreparation}) I argue that scope freezing is a first-phase syntax phenomenon, which arises when the lower QP raises into the specifier of ApplP (which I argue is located in Slavic between VoiceP and} \textrm{\textit{v}}\textrm{P). The overt movement of the lower argument into the specifier of ApplP is argued to define and lead to the Spell-out of the first phase, the domain in which the scope freezing relation is set. Note that irrespective of whether this account is correct, what we are crucially relying on here is the empirically grounded generalization that the domain of scope freezing excludes the subject QP. Thus, the external argument, taken to be generated in Spec,VoiceP, is always going to be literally too high, being outside the domain in which this scope relation can be established.} } 



Finally, note that I have provided no independent evidence that the Agent of passives is generated in the same position as the Agent of active transitives, though, arguably, the scope data (i.e., the same scope-taking ability of passives as other structures involving the Agent QP) do provide indirect evidence to this effect. Crucially, however, to argue against the Undifferentiated Initiator view, it is enough to show that the Instrumentals in Transitive Impersonals are generated in a position other than Spec,VoiceP.  I conclude that the scope difference between a passive \REF{ex:antonyuk:35} and a transitive impersonal \REF{ex:antonyuk:36} is indeed due to a difference in the structural position of the Instrumental argument in the two types of constructions under consideration. Thus, we gain a novel argument against the Undifferentiated Initiator view (\citealt{ant:Ramchand2008}; \citealt{Legate2014}; \citealt{Wood2017}, i.a.) and in favor of making a distinction between a higher verbal layer, VoiceP, which introduces the external argument and the lower verbal layer, \textit{v}P, which introduces causative semantics and an inanimate causer argument in the absence of Voice (\citealt{Lavine2022,Lavine2023}).  


\subsubsection{On the VP-external Merge position of argument Instrumentals}
\label{sec:antonyuk:3.2.2}

Going back to the previous point, i.e., the exact nature of the Inanimate Causer’s Merge position, as mentioned earlier, and as should be clear from the SFD insights already reviewed, I argue that the Merge position of Inanimate Causer Instrumentals cannot be VP-internal. \citet{Lavine2022, Lavine2023}, on the other hand, posits a low VP-internal position in the complement of V, which is traditionally reserved for Oblique arguments. 

As a reminder, the argument hierarchy deduced from applying the SFD to a variety of constructions is the following: V NP\textsubscript{INS} $>>$ NP\textsubscript{ACC} $>>$ NP\textsubscript{DAT}. Thus, while NP\textsubscript{DAT} is shown to pattern with the Obliques (i.e., arguably sharing the same Merge position with directional and locative PPs), the Inanimate Causer Instrumentals are in fact shown to have the highest Merge position of the three arguments. Hence, to the extent the SFD is accepted as being correct, the complement of V position for NP\textsubscript{INS} becomes ruled out. Spec,V, on the other hand, might be somewhat more likely. Specifically, in dyadic constructions involving two internal arguments, NP\textsubscript{ACC} and NP\textsubscript{INS}, the former is consistently shown by the SFD to be merged lower than NP\textsubscript{INS}, hence it is conceivable that the Merge position for the two arguments could simply be the reverse of what is commonly assumed.\footnote{\textrm{This is precisely what \citet{Landau2010} proposes as far as the position of the Accusatives in object experiencer constructions (i.e., an Accusative as a concealed low Oblique), which in Ukrainian involve Accusative and Instrumental argument NPs.} } This conclusion appears equally unavailable, however, once we consider again examples involving three internal arguments (ex. \REF{ex:antonyuk:29} and \REF{ex:antonyuk:30} repeated here for convenience as \REF{ex:antonyuk:29rep} and \REF{ex:antonyuk:30rep}).\footnote{\textrm{The \# sign next to \REF{ex:antonyuk:37a} is meant to indicate that this example (just like \REF{ex:antonyuk:24a}) is infelicitous on its surface scope interpretation in that it describes a situation where the same finger is severed with every instrument in some contextually salient set of instruments. Our world knowledge tells us this situation is impossible or improbable, but in terms of logico-semantic properties of the sentence, this interpretation indeed obtains and is in fact the only interpretation available for this sentence.}}

\ea%29
    \label{ex:antonyuk:29rep}
\gll            Xlopcevi   vidrizalo    palec’    elektryčnoju pyloju.\\
boy.\textsc{dat}   severed.\textsc{non-agr} finger.\textsc{acc} electric.\textsc{ins} saw.\textsc{ins}\\
\glt ‘The boy’s finger got severed by an electric saw.’
    \z


\ea%30
    \label{ex:antonyuk:30rep}

\ea
\gll   Jakomus’ xlopcevi   vidrizalo    palec’    kožnym instrumentom.\\
some.\textsc{dat} guy.\textsc{dat} severed.\textsc{non-agr}  finger.\textsc{acc} every.\textsc{ins} instrument.\textsc{ins}\\
\glt ‘Some boy got a finger severed by every instrument.’    \hfill   (frozen)



\ex
\gll   Jakymos’ instrumentom   vidrizalo     palec’     kožnomu xlopcevi.\\
some.\textsc{ins} instrument.\textsc{ins}     severed.\textsc{non-agr}  finger.\textsc{acc}  every.\textsc{dat} boy.\textsc{dat}\\
\glt ‘With some instrument (or other), every boy’s finger was severed.’\\ \hfill    (ambiguous)
\z 
$\Rightarrow$ \textsc{Base order}: V NP\textsubscript{INS} $>>$ NP\textsubscript{DAT}
\z


\ea%37
   \label{ex:antonyuk:37}

\ea[\#]{\gll   Xlopcevi   vidrizalo    jakyjs’ palec’    kožnym instrumentom. \\
boy.\textsc{dat}  severed.\textsc{non-agr} some finger.\textsc{acc} every.\textsc{ins} instrument.\textsc{ins}\\
\glt ‘Some boy got a finger severed by every instrument.’    \hfill    (frozen) \label{ex:antonyuk:37a}}



\ex
\gll   Xlopcevi   vidrizalo    jakymos’ instrumentom kožen palec’.\\
boy.\textsc{dat}    severed.\textsc{non-agr} some.\textsc{ins} finger.\textsc{ins} every.\textsc{acc} finger.\textsc{acc}\\
\glt ‘Some boy got every finger severed by some instrument.’    \hfill  (ambiguous)
\z
$\Rightarrow$ \textsc{Base order}: V NP\textsubscript{INS}  $>>$ NP\textsubscript{ACC}
\z


\noindent What we see here is that when NP\textsubscript{DAT}, NP\textsubscript{ACC} and NP\textsubscript{INS} are all arguments of the same verb, the SFD points to the same relative ordering of arguments, i.e., \mbox{V NP\textsubscript{INS}} $>>$ NP\textsubscript{ACC} $>>$ NP\textsubscript{DAT}. The above examples, while not producing any new evidence beyond that discussed earlier in the paper, nevertheless drive home the point that there is simply no place left in the VP for the Inanimate Causer argument. Hence, my proposal (in accordance with \citealt{Lavine2022,Lavine2023} and \citealt{Pylkkänen2002,Pylkkänen2008}), developed in detail in related work, is that the data considered in this paper are to be interpreted as evidence in favor of the Unbundled Voice projection in Ukrainian, the lower of which, call it \textit{v}CauseP, following \citet{Lavine2023}, hosts the Inanimate Causer Argument while the higher one, VoiceP \citep{ant:Kratzer1996}, hosts the Agent external argument. Without going into the specifics of the larger proposal here, I will point out that semantically, it makes good sense for a Causer argument to be merged higher than the other VP-internal arguments (NP\textsubscript{ACC}, NP\textsubscript{DAT}, PP); syntactically, apart from the purely theoretical reasons for favoring NP\textsubscript{INS} as being merged in Spec,\textit{v}CauseP, we now have solid empirical evidence that supports precisely this view.\footnote{As mentioned earlier, \citet{Lavine2023} treats the inanimate causer arguments in Transitive Impersonal constructions as the Natural Force argument, which is arguably well aligned with the standard view of Instrumental arguments as being merged low in the VP (cf. \citealt{Szucsich2007}; \citealt{Schäfer2008}; \citealt{JunghannsEtAl2017}; \citealt{Wood2017} i.a. for other theoretical solutions).}

\section{Conclusions}\label{sec:antonyuk:4}
\largerpage
In this paper I have provided a novel argument against the Undifferentiated Initiator view (\citealt{ant:Ramchand2008}; \citealt{Bruening2013}; \citealt{Legate2014} and \citealt{Wood2017}), which posits distinct structural Merge positions for the Instrumental Agent of passives and the morphologically indistinguishable Instrumental Inanimate Causer argument of a range of “Transitive Impersonal” constructions in Ukrainian, thus providing independent support for the main theoretical claim in \citet{Lavine2022, Lavine2023}. The novel evidence presented in this paper shows that the Instrumental Agents of passives and the Instrumental Inanimate Causers (e.g., in Ukrainian non-agreeing accusative constructions) behave differently with respect to QP scope, with the latter, but not the former, being able to participate in establishing a surface scope freezing relation. In this regard passives behave very much in line with a range of transitive constructions involving a subject QP: as shown in the paper, subject QPs categorically resist participating in scope freezing (\citealt{Larson1990}; \citealt{Bruening2001}). Thus, while no novel evidence bearing on the Merge position of the subject of passives is provided here, the paper does provide indirect support for configurational, structural accounts of passives such as \citet{Collins2005}, by showing strong similarities with respect to non-participation in scope freezing on the part of subject QPs in SVO, OSV, SOV, OVS and passive sentences. In other words, subject QPs are outside the domain in which scope freezing can be established, and the subjects of passives in this respect behave similarly to all other external arguments of transitive sentences.

I have also provided evidence in favor of a higher Merge position than is commonly assumed for a variety of Instrumental case-marked arguments. I suggest that their Inanimate Causer semantics ensures their being merged above all other internal arguments. Specifically, I argue that the Inanimate Causer Instrumentals must be merged outside the core VP (cf. \citealt{Lavine2022,Lavine2023} for a low complement of V position for all inanimate cause Instrumentals), the most likely candidate for the Merge position being \textit{v}CauseP (i.e., the lower Voice projection in \citeauthor{Pylkkänen2002}'s \citeyear{Pylkkänen2002, Pylkkänen2008} Unbundled Voice proposal), which introduces causative semantics in the absence of Voice, thus very much being in the spirit of Lavine’s work (see also esp. \citealt{Harley2013}).

  Finally, it should be pointed out that the above arguments were all made based on the insights afforded by the Scope Freezing Diagnostic (\citealt{Antonyuk2015,Antonyuk2020, Antonyuk2023, AntonyukUnderReview}), which is a testament to its diagnostic utility. Here as elsewhere, the SFD is shown to be a remarkably consistent diagnostic tool, one that is especially well suited for the Discourse Configurational Languages such as Ukrainian and the rest of Slavic where overt syntactic movement correlates with semantic interpretation as well as discourse-related properties of the sentence.


\section*{Abbreviations}

\begin{tabularx}{.45\textwidth}{lQ}
\textsc{acc} &accusative \\
\textsc{agr} & agreement \\
\textsc{dat} & dative \\
\textsc{f}    & feminine \\
\textsc{gen} & genitive \\             
\textsc{ins} & instrumental               \\
\textsc{nom} & nominative \\
        
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\textsc{non-agr} &  non-agreeing form\\
\textsc{pass}   & passive          \\
\textsc{pl}   & plural             \\
\textsc{pst}    &past              \\
\textsc{refl} &   reflexive        \\
\textsc{sg} & singular \\
&\\ % this dummy row achieves correct vertical alignment of both tables
\end{tabularx}
\section*{Acknowledgments}


I am grateful to the audiences at FDSL-15 in Berlin and the Theoretical Linguistics Lab at the University of Graz for questions and comments, and especially to Željko Bošković, James Lavine, and Edgar Onea for helpful and generous discussion of this material at various points. My sincere gratitude also goes to the anonymous reviewers of this paper for providing insightful questions that have helped significantly improve its quality. This research is supported by the Austrian Science Fund (FWF) Lise Meitner grant M 3361-G.

\printbibliography[heading=subbibliography,notkeyword=this]

\end{document}
