\title{A tale of two dialect regions}   
\subtitle{Sranan's 17th-century English input}
\BackBody{This book traces the precise origin of the early English lexical and lexico-phonetic influences in Sranan, an English-based creole spoken in Suriname. Sranan contains "fossilised" linguistic remnants of an early English colonial period. The book discusses whether Sranan’s English influence(s) originated from a single dialect from the general London area, as proposed by Norval Smith in 1987, or whether we are dealing with a composite of dialectal features from all over England. The book introduces a novel replicable methodology for linguistic reconstructions, which combines statistics (in the form of binomial probability), English dialect geography (via use of Orton’s et. al., 1962--1971, \textit{Survey of English Dialects}, which focuses on traditional regional English dialects across England and Wales), and 17th-century English migration history (compiled from \textit{The Complete Book of Emigrants: 1607–1660, The Bristol Registers of Servants Sent to Foreign Plantations, 1654–1686, Virtual Jamestown, Virginia Center for Digital History}, and Colonial State Papers secured from the \textit{British History Online} databases, among other relevant historical sources).}
\dedication{I would like to dedicate this work to my grandmother, Dulcie Mae Roper, God rest your soul. I have never forgotten you, nor will I ever. You taught me to never give up, but to always give my all, no matter the obstacles that might block my path. Thank you.}
\typesetter{André Ché Sherriah, Felix Kopecky, Sebastian Nordhoff}
\proofreader{%
Alexis Michaud,
Amir Ghorbanpour,
Andreas Hölzl,
Bev Erasmus,
Eran Asoulin,
Ivica Jeđud,
Jeroen van de Weijer,
Marijana Janjić,
Sandra Auderset
}
\author{André Ché Sherriah}
\renewcommand{\lsImpressionCitationAuthor}{André Ché Sherriah}
\BookDOI{10.5281/zenodo.2625403}%ask coordinator for DOI
\renewcommand{\lsISBNdigital}{978-3-96110-155-9}
\renewcommand{\lsISBNhardcover}{978-3-96110-156-6} 
\renewcommand{\lsSeries}{scl}  
\renewcommand{\lsSeriesNumber}{3}  
\renewcommand{\lsID}{179}

  
 
  
