\addchap{\lsAcknowledgementTitle} 

This book is a revised and improved version of the doctoral dissertation I submitted in July 2022. The original acknowledgments are reprinted below with minor updates.

I could not have completed this work without the people around me. First and foremost, I would like to thank my supervisors, Nikolaus P. Himmelmann and Kees van Deemter, for their great support and valuable guidance. Thank you, Nikolaus, for always being there for me and guiding me through my PhD project with your deep linguistic understanding. Thank you, Kees, for the great collaboration, the many joyful meetings, and the thoughtful ideas you always have. I really enjoy working with you and learning from you. I can not thank you both enough.

I would like to thank SFB1252 “Prominence in Language” for providing me with a great working environment and giving me the opportunity to work on my PhD. I would like to thank all the wonderful colleagues I have had the pleasure to work with over the years at the University of Cologne, especially in our great House of Prominence (HoP). To name just a few: Thank you to Alessia Cassarà, Barbara Zeyer, Eric Engel, Max Hörl, Constantijn Kaland, and Sara Meuser. Many thanks to everyone from the second floor of the HoP (for coffee, cookies and nice, happy conversations) and to the SFB lunch crew for the funniest (and sometimes absurd) conversations over the years. My very special thanks to Maria Bardaji i Farré (thank you, Maria, for your cheerful and positive presence in the very last phase of my PhD), Christoph Bracks (thank you, Chris, for being a great friend and helping me a lot in the last weeks of writing my PhD), and Guanyi Chen (thank you, Guanyi, for being such a great colleague and coauthor. I really enjoy working with you and have learned so much from you). My very special thanks to T. Mark Ellison (thank you, Mark, for always being there for me, teaching me so much, reading and commenting on my dissertation, and for all the fun conversations.)

I would like to thank my dearest friends, Arefeh (aka Fafafe), Golnoush, Mahsa, and Adineh for keeping me sane during the craziest days of my life. The phase of finishing up a PhD can get pretty lonely. It gets even lonelier when you have to do it during Covid. The best thing is to have great friends to support you, cheer you up, and dream with you about fun life after the PhD. Thank you all for being there for me. You're the best! I also want to give a special thanks to Rob, for being such a great (and funny) friend and also proofreading parts of my dissertation.

My deepest gratitude goes to my family. Thank you, Zohreh and Mamad, for being the most amazing siblings anyone could ask for, for always having my back and for always believing in me. And my deepest thanks goes to my mom (Maman). Thank you, Maman, for your unconditional love and support. I love you so much!

And a very big “thank you” to my partner, Christian. Thanks, Chris, for granting me full “Narrenfreiheit” over the past few months, for cooking delicious food, and for reading my dissertation more than once. Thanks for being such a wonderful partner.
