\addchap{\lsAcknowledgementTitle} 

The editors of this book are grateful to the Contemporary African Linguistics series editors -- Michael Marlo and Laura Downing -- for believing in this volume. We appreciate their encouragement and support and the very efficient editorial process that they facilitated.

A volume like this would not be possible without the careful and critical, but always constructive, input from reviewers. Each chapter in this volume was evaluated by at least two, and in some instances three, expert reviewers. Where appropriate, one of the reviews was conducted internally by another contributor to the volume, but in most instances, reviews were provided by external reviewers who were not otherwise involved in the project. We gratefully acknowledge the contributions of all reviewers: (external) Joe Amoako, Kelly Berkson, Lee Bickmore, Michael Cahill, Aline Cunia, Stuart Davis, Mark Dingemanse, Katherine Franich, Emily Hannink, Larry Hyman, Kyle Jerro, Eyo Mensah, Winfred Mkochi, David Odden, Martin Orwin, Augustina Owusu, Kwabena Sarfo-Kantankah, Yvonne Treis, H. Ekkehard Wolff, and Ben Yamsuan; (internal) Matthew Ajibade, Victor Alabi, Kofi Agyekum, Leonardo Contreras Roa, Kenneth de Jong, Abbie Hantgan, and Ron Schaefer.

Lastly, we gratefully acknowledge Aline Cunia, who set the references in BibTex, and Gary Mullin II, who provided expert technical assistance in formatting figures and other graphics. We are also indebted to the editorial team at Language Science Press -- Sebastian Nordhoff for his invaluable guidance during the proofreading stage and Felix Kopecky for technical help with matters related to formatting and indexing.