The \textit{Yu ingong} `fire dance' story is a parable about two older couples. The two men -- Piasorosoro and an unnamed man -- go into the middle of the garden while their wives set the garden ablaze. The men dance and play a drum. One man leaves the fire unscathed, but his friend, who copies him, is not. The moral of this story is not to imitate other people. In some versions of the story (including the main text below), Piasorosoro is the name of the man who was saved, and in some versions (including the song below), Piasorosoro is the name of the man who was burned.

This story mentions the \textit{alläp} `kundu drum', \textit{pite} `grass skirt', and \textit{banggu} `headdress.’ Figures \ref{fig:dance1} and \ref{fig:dance2} show  \name{Sali}{Goge (Wik)} and other dancers pictured wearing traditional cassowary feather headdresses, armbands, leg bands, and traditional grass skirts. They have also decorated their skin with white clay. \figref{fig:dance3} shows \WPPfull{} carrying a \textit{Kundu} drum, with singers and dancers in the background.

One unique term used in \textit{Yu ingong} `Fire Dance' is \textit{omad}, which signifies a close relationship between two women who went through initiation together (see Line \ref{ex:text14-4-3}). \textit{Omad} is one of many friendship and \isi{kinship} terms that form a special class of words in Ende. One unique property of kinship terms is that they have nonsingular forms, formed via \isi{reduplication}, a property shared only with adjectives and some rare nouns. 

Friendship is a common theme in Ende stories, as Ende people tend to move around the village and surrounding areas in pairs or groups for safety and company. When two friends want to signal an escalation in their friendship, they can choose between a number of special friendship terms that they will then use to call one another. One such term is \textit{nadum}, which is used to call anyone with whom you share a name. Giving someone one of your names is an honor. Friendship is depicted in almost every story in this collection.