\chapter{Sherbro songs}
\label{app:e}\hypertarget{Toc115517831}{}
Both of these songs were recorded in 2016 and form part of the archive of the project documenting the Sherbro language and culture.

\section{Kagboro Anthem}
\hypertarget{Toc115517832}{}
This anthem was composed by the 13\textsuperscript{th} Paramount Chief of Kagboro Chiefdom (1956-1999), the Hon. Paramount Chief Honoria Bailor Caulker\is{Caulker, Honoria Bailor}. It is sung to the tune of the South African national anthem, sometimes known as “Nkosi Sikelel iAfrika”.\\

Yin gbi ka yiɛ apuma Kagbɔɛ\\
All of us here are children of Kagboro\\

Lonthibul yi bi ha poin pɔk hiɛ\\
Unity is what we have to wake our country\\

Lonthibul yi bi ha yɛhi kath\\
Unity is what we have to hold firm\\

Ma gbentrɛɛ chɔygba\\
Don't delay us forever\\

Ma gbentrɛɛ chɔygba\\
Don't delay us forever\\

(Chorus)\\

Ha po, hɔɛ keni, Kagbɔ! Kagbɔ!\\
Get up, the day has broken, Kagboro! Kagboro!\\

Ha po, hɔɛ keni, Kagbɔ! Kagbɔ!\\
Get up, the day has broken, Kagboro! Kagboro!\\

Hɔbatok che rubani, Kagbɔai chɔygba!\\
God does not bless Himself, Kagboro forever!\\

Thasɔ, Bɛndu, Thumba, ni Manɔ\\
Tasso, Bendu, Thumba and Manor\\

Bompɛtok, Mɔkebe, Mɔpelɛ\\
Bumpetok, Morkebe, Morpeleh\\

Kɔndɔlɔ, mɔm bɛ ma lee ndɔi ai\\
Kondoloh, don't be left in slumber\\

(Chorus)\\

Ha po, hɔɛ keni\\
Get up, the day has broken\\

Ha po, hɔɛ keni\\
Get up, the day has broken.\\

Poepoe hɔɛ fɔn ra, Kagbɔ! Kagbɔ!\\
Early is the secret of farming, Kagboro! Kagboro!\\

Poepoe hɔɛ fɔn ra, Kagbɔ! Kagbɔ!\\
Early is the secret of farming, Kagboro! Kagboro!\\

Hɔbatok che rubani, Kagbɔai chɔygba!\\
God does not bless Himself, Kagboro forever!\\

che rubani, Kagbɔai chɔygba!\\
Does not bless Himself, Kagboro forever!\\

Mokandɔ Mɔchobo Mɔyibo\\
Morkandor, Morkobo, Moryibo\\

Ngiehun Gbualel ni mɔmbɛ\\
Ngiehun, Gbualel and yourself\\

Mofus, Mambo, Mɔya, Yɔndu, Rɛmbi\\
Morfoss, Mambo, Morya, Yondu, Rembi\\

(Chorus repeats.)\\

Hane gbi ha ha mpanth ma chɛkɛ\\
All those that are doing the farm work\\

Hane gbi ha hɔth hɛlɛ koɛ\\
All those who are fishing in the sea\\

Hane gbi ha ramil anya nakaɛ\\
All those who cure sick people\\

(Chorus repeats.)\\

Hane gbi ha kaŋ apumalɛ\\
All those who teach the children\\

Hane gbi ha ha muyu bil li haiɛ\\
All those that are patient in their marital homes\\

(Chorus repeats)\\

Hane gbi ha ha mpanth ma thoɛ\\
All those that are doing the bush work\\

Hane gbi ha ha mpanth ma chol lɛ\\
All those that are doing the skilled work\\

Nted ni Nkafo, Kɔlma, Kɔbɔ Hɛthɛ\\
Nted and Nkafo, Korlma, Korbor Hetheh\\

(Chorus repeats.)\\

Hane gbi ha chala Kɛnɛma koɛ\\
All those who live in Kenema\\

Hane gbi ha chala Tɔŋgɔɛ\\
All those who live in Tongor\\

Hane gbi ha chala Kiamp koɛ\\
All those that live in Freetown\\

(Chorus repeats.)\\

Hane gbi ha kɔ kaŋ Inglandɛ\\
All those that went to study in England\\

Hane gbi ha kɔ kaŋ Rusiaɛ\\
All those who went to study in Russia\\

Hane gbi ha kɔ kaŋ amɛrikaɛ\\
All those who went to study in America\\

(Chorus repeats)\\

Hane gbi ha chala gbes koɛ\\
All those who live in the east\\

Hane gbi ha chala pal li chɛthɛ\\
All those who live in the west\\

Hane gbi ha chala hɔlɔoailɛ\\
All those who live in this world\\

Ha po tama dukɔ\\
Get up for it not to fall\\

Ha po tama dukɔ\\
Get up for them not to fall\\

\newpage
Ha po ha chɔ ha pɔki nɔɛ, Kagbɔ! Kagbɔ!\\
Get up and fight for our country, Kagboro! Kagboro!\\

Ha po ha chɔ ha pɔki nɔɛ, Kagbɔ! Kagbɔ!\\
Get up and fight for your country, Kagboro! Kagboro!\\

Pɔk nɔ Kenaɛ, anyan ha ma kɔ kwe hin fio\\
It's your country, for people not to take it from our hands\\

Ha ma kɔ kwe hin fio.\\
For them not to take it from our hands.\\

\section{Hymn: Yaŋ ya si (Me I know)}
\hypertarget{Toc115517833}{}
Note: This is one of twenty hymns performed by the Shenge\is{Shenge} Youth Choir of the Methodist church in Shenge in February of 2016 (003a Shenge Youth Choir, Hymns 1-20). Abdul Rahman Bendu leads the group, sings, and plays one of the drums. He has also written many of the hymns.\\

Yaŋ ya si yɛ o\\
Me I know now-o\\

Yaŋ ya si\\
Me I know\\

Bahin chɔŋ mi len\\
My Lord loves me\\

Rɔŋ fili-i-i-i\\
Indeed it's the truth\\

Rɔŋ fili wɔ mi lɛli atok\\
Indeed it's the truth that he watches over me\\

Yɛ ya lɔl ya po yɛ wɔɛ ŋɔ keni we\\
When I sleep and wake up early in the morning\\

Yaŋ ya siɛ a bi len ŋa wɔ we\\
Me I know I have something to say\\

Yɛ ya wokɔ tikomiko a kɔni yena livil we\\
If I leave home to go anywhere far away\\

Ya bi ŋa wɛ a chɔŋɔ mɔ sɛkɛ Bahin\\
I have to say I give thanks to Our Father\\

A yiyɛ Bahin ŋa toŋi mi nai wɛ we\\
I ask the Lord to show me the way\\

Bahin wɛ wɔlɔ naɛ we\\
The Lord said he is the only way\\

Chaŋbo a thɔni ka Min Charaŋ dɛ we\\
Unless I am cleansed by the Holy Spirit\\

A che bi ŋa lɔɛ Arijana\\
I shall never enter Heaven\\

Bahin, a bi ŋa che gbɛŋ Mɔɛ ni cheche Mɔ kunɛ we\\
Lord, I shall be within Your glory and Your light\\

Ni jali tilaŋ gbi labiŋa bɛrɛ lɔ ni.\\
So there is abundance in all other things.\\

