\label{ch:exp-infinitival}

The results of the previous corpus and experimental studies show that extractions out of subject NP are not islands. More precisely, the constraints we can observe are not specific to subjects or extractions out of subjects.

What about infinitival and sentential subjects? \citegen{Ross.1967} original definition of subject island only targeted extractions out of clauses, not out of NPs. \citet{Ross.1967} furthermore acknowledges that the subject island constraint may be English-specific, and that Japanese, for example, seems to allow extraction out of sentential subjects.

Sentential and infinitival subjects are hard to target by a corpus study, therefore this chapter presents two experiments on infinitival subjects, the last ones in this book. Both are acceptability judgment studies. The organization of the chapter is the following:

\begin{description}

\item[Experiment 15:] In this acceptability judgment study, we crossed extraction type (extraction\slash non-extraction) with syntactic function (subject\slash object) and tested subextraction out of infinitivals. The extracted element was the locative complement of the verb of the embedded clause, the filler was  \emph{où}. Extractions out of infinitival subjects were significantly less acceptable than extractions out of infinitival objects (impersonal construction), but the factors did not interact.

\item[Experiment 16:] In this acceptability judgment study, we crossed extraction type (extraction\slash non-extraction) with syntactic function (subject\slash object) and test subextraction out of infinitivals. The extracted element was the direct object of the verb of the embedded clause, the filler was \emph{que}. Extractions out of infinitival subjects received significantly lower mean acceptability ratings than extractions out of infinitival objects (impersonal construction), but this time there was also a significant interaction such that extractions out of the subject had lower ratings than all other conditions.

\end{description}
