\addchap{\lsAcknowledgementTitle} 
\largerpage
It's hard to remember what exactly motivated me to write this translation. A lot has happened in my life since I first start drafted the translation as an ugly Google Doc in Oakland. 

But let's try to go down memory lane. After doing a grammar of Iranian Armenian in 2021, I realized that I'm deficient in my knowledge of dialectal Armenian. Being a perfectionist/completionist, I wanted to cover that gap the best way I know how -- by reading something. Nikita and Hrach (I think?) recommended that I try reading Adjarian 1911. Ironically, I also hate reading, and especially reading Armenian texts that aren't OCR-ed or have a clickable table of contents. I decided to read that book, but I wanted to only read it once and never again. And so I decided to kill two birds with one stone by translating the book while reading. I didn't realize that I would then have to read the book multiple times to correct and doublecheck my translation. Ironic, isn't it? 


I spent many an evening hour writing out this translation in 2022. It was a largely solo endeavour -- me alone at my desk, waiting for bedtime. Along the way, people came around sometime (online) and gave moral support -- especially Nikita Bezrukov, Hrach Martirosyan, and Bert Vaux. 

During the translation process, I sometimes had a dead-end or two for the optimal Armenian translation. And when I would see a word that was not Armenian... Well, I had no choice but to bother people on the internet. I thank the many people who helped me out in that regard. Their names are distributed in footnotes and lists across the translation. I would try to collect them all into an itemized list here, but then I'd fear missing a name. 


Eventually the draft was ready-ish for the public, and got to be presented at UCLA 2023, where further evidence was provided to me that the translation was not a waste of my time. Thank you to Peter Cowe, Hrayr Khanjian, and Anooshik Melikian for that. 

Lo and behold when it even got reviewed by Robin Meyer who helped me see how my Classical Armenian was deficient. He and Nikita Bezrukov where then helpful comrades in tracking down further references for Adjarian's bibliography.


Well, I suppose I'm rambling at this point. Thank you friends from 2022 for helping me write this book. May I never have to translate a beast again. 
