\addchap{\lsAcknowledgementTitle} 


This book is a version of a PhD dissertation, written between 2016–2021, and would not have been possible without the insightful comments and patience of my supervisors Professor Fernando Zúñiga, University of Berne, and Professor Isabelle Bril of the LACITO-CNRS. Thank you for taking the time to read and think, and for your patience. I came to you too rarely.

Florian Matter, Dominique Knuchel, Sarah Dopierala, Joël Schregenberger, für ds Läse vo Kapitle u natürlich für ds Mitdänke. Am Flo none bsungere Dank, für ds unermüedliche Hingrfraage vo myne Aanahme, für ds Daate-wöue-gseh, u ständig mit \LaTeX{}, Chaarte, u angerne Sache cho häuffe.  

Hao Cakeo, e holeke i wîî-je! Jacob Ganap Oué, Jean-Philippe Oué, André Nigai Kalène, Christophe Keela Pei, holeke thuan nyakovwe i vaaya fine thoatit, li délires ka meeka li been aman! Nyasi Baptiste Ucian, Elise Kalène, e holeke i hun-titabwi ko i ape-thapoke kan, Jethro ma Hedwige Delly ko i hun-wago, ka meeka li apuli xa-moo pala-vwe ko i hun-moo-le! Un deuxième merci encore au groupe de travail, qui sont les deuxièmes auteurs de ce livre.

Daniel Pei, pour m'avoir acceuilli, introduit, et logé, Cécile Pei pour m'avoir nourri pendant si longtemps, Richard Pei, Eugénie Pei, Kafeyat Pei, Roger Pei, Kukwe Pei, et toutes les mamans et papas qui m'ont élevé pendant ce temps-là, et Pwiya pour son amour maternel. Merci à Alain Mindia Pei et sa famille pour le jardin, l'acceuil et l'amitié.

Doriane Pei, Béty Gohoupe pour avoir été des soeurs, et pour m'avoir montré un côté du pays qui me reste chaud au coeur. Aouh bwaau, les jacks et les soirées à la plage avec vous me manquent!

My thanks to Mandana Seyfeddinipur, Sophie Salffner, and Stephanie Petit of ELAR for their patient work.

My family, Olivia Schneider, and Ghada ben Yahia, for support and patience along the way. It was a long one, and I appreciate your help more than I'll ever be able to say.

A heap of gratitude belongs to Martin Haspelmath and Sebastian Nordhoff for the comments and help with putting the thesis into the shape of a published grammar. Thanks for their valuable insight and attentive lecture also go to Ezekiel Bolaji, Benjamin Brosig, Christian	Döhler, Philip	Duncan, Carrie Dyck, Lachlan Mackenzie, Jean Nitzke, Tim Ongenae, Elliott Pearl, Katja Politt, Tabea	Reiner, Yvonne Treis, Jeroen van de Wijer, Mary Ann	Walter, as well as two anonymous reviewers. Thank you for your help!
