\addchap{\lsAcknowledgementTitle} 

Pulling together an edited volume is a surprisingly complicated task that requires significant coordination among people who probably don't receive enough credit. They all have our most profound thanks.

We are extremely grateful to the entire team at Language Science Press. Sebastian Nordhoff and Felix Kopecky have been incredibly responsive in assisting us in getting this volume to production. They have never hesitated to answer our stupidest questions about \LaTeX\ and have helped us debug this document every time we've introduced a crash-inducing formatting command. Our thanks to the devoted team of community proof-readers (David Carrasco Coquillat, Dmitry Nikolaev, Elen Le Foll, Elisa Roma, Elliott Pearl, Eva Schultze Berndt, Jean Nitzke, Jeroen van de Weijer, Katja Politt, Lachlan Mackenzie, Mary Ann Walter, Matthew Korte, Nicoletta Romeo, Rainer Schulze, Rebecca Madlener, Sebastian Nordhoff, and Silvie Strauß) for catching our mistakes and to the production team at the press. Any remaining errors are our own. 

%LANGUAGE SCIENCE PRESS FOLK: Please add names as appropriate to the above paragraph if we've missed anyone. 

We offer a huge thank you to all of our authors, all of whom were exceptionally amenable to our requests and patient as well - and without whom this volume would not have been possible. We would also like to thank all the other presenters at the Foundational Approaches to Celtic Linguistics series who didn't contribute to this volume, but who also helped make the series so successful: David Adger, Joe Eska, S.J. Hannahs, Holly Kennard, Will Lamb, Christopher Lewin, Kerry McCullough, Bernadette O'ROurke, Claire Nance, Kevin Rottet, Kevin Scannell, Sylvia Schreiner, Cassie Smith-Christmas, Thomas Stewart, John Walsh, Peredur Webb-Davies, and David Willis. We'd also like to thank the audiences who attended these talks and provided such a rich environment for discussion and debate.

We owe particular gratitude to Kerry McCullough, who was a critical force in organizing the speaker series. She helped select and schedule speakers, organized students to introduce them, made our logo, and handled our advertising. Other students who also assisted us include Jes\'us Gonz\'alez Franco, Andrea Fulgham, Brandon Gauthier, Luis Irizarry, Jennifer Medina, Alex Tubens Blanco, Colleen Patton, Sam Prins, John Powell, Dee Hunter, and Andrew Zupon.

%%%% FACL EDITORS PLEASE ADD NAMES ABOVE %%%%%%%%

High quality chapters require high quality reviewers, we relied on the following people to give detailed feedback and reviews on the papers. Thank you to: Ryan Bennett, Sonya Bird, Andrew Carnie, Ian Clayton, Joe Eska, Michael Hammond, S. J. Hannahs, Heidi Harley, Dee Hunter, Luis Irizarry, Jim McCloskey, Diane Ohala, John Powell, Samantha Prins, Natasha Warner, David Willis, and Andrew Zupon.

Go raibh maith agaibh, tapadh leibh, diolch yn fawr, trugarez deoc'ch, gura mie orroo, meur ras, merci and thank you!\bigskip\\
\noindent --- Andrew, Dee, Diane, Luis, Mike and Sam \hfill
Tucson, March 2025
