\addchap{Abbreviations}
% \addchap{Abbreviations and symbols}
 


\begin{tabularx}{.48\textwidth}{lQ} 
A					&	 transitive subject \\
\textsc{abst}		&	 abstractive \\
\textsc{acc}        &    accusative \\
\textsc{adj.pl}		&	 \isi{adjectival} plural \\
adv.				&	 \isi{adverb} \\
\textsc{and}		&	 an\isi{dative} \\
\textsc{anaph}		&	 \isi{anaphoric} \\
\textsc{anticip}	&	 anticipative \\
\isi{ATR}					&	 advanced tongue root \\
\textsc{aux}		&	 auxiliary \\
\textsc{bhvr}		&	 behaviorative \\
C					&	 consonant \\
\textsc{caus}		&	 \isi{causative} \\
CC					&	 \isi{copula} \isi{complement} \\
\textsc{comp}		&	 completive \\
compl.				&	 \isi{complementizer} \\
\textsc{cond}		&	 conditional \\
\textsc{conf}		& 	 confirmational \\
coordconn.			&	 coordinating connective \\
\textsc{cop}		&	 copulative \\
CS					&	 \isi{copula} subject \\
\textsc{dat}        &    \isi{dative} \\
dem.				&	 demonstrative \\
\textsc{dist}		&	 distal \\
\textsc{distr}		&	 distributive \\
\textsc{dp}			&	 \isi{dummy pronoun} \\
\textsc{dur}		&	 durative \\
E					&	 extended object \\
\textsc{exc}		&	 exclusive \\
FF					&	 final form (pre-pause)\\
\end{tabularx} 
 %
\begin{tabularx}{.45\textwidth}{lQ} 
\textsc{gen}        &    genitive \\
H					&	 high tone \\
\textsc{hypo}		&	 hypothetical \\
ideo.				&	 \isi{ideophone} \\
\textsc{imp}		&	 \isi{imperative} \\
\textsc{inc}		&	 inclusive \\
\textsc{inch}		&	 \isi{inchoative} \\
\textsc{inf}		&	 \isi{infinitive} \\
\textsc{infr}		&	 inferential \\
\textsc{ins}        &    instrumental \\
\textsc{int}		&	 \isi{intentional} \\
interj.				&	 \isi{interjection} \\
\textsc{ipfv}		&	 \isi{imperfective} \\
\textsc{ips}		&	 impersonal \isi{passive} \\
\textsc{irr}		&	 irrealis \\
L					&	 low tone \\
lit.				&	 literal \\
M					&	 mid tone \\
\textsc{med}		&	 medial \\
\textsc{mid}		&	 middle \\
n.					&	 noun \\
NF					&	 non-final form \\
\textsc{nom}        &    nominative \\
num.				&	 numeral \\
nurs.				&	 nursery word \\
O					&	 object \\
\textsc{obl}        &    oblique \\
\textsc{opt}		&	 optative \\
\textsc{pass}		&	 \isi{passive} \\
\textsc{pat}		&	 patientive \\
\textsc{phys}		&	 physical property \isi{adverb} \\
\end{tabularx}

\begin{tabularx}{.48\textwidth}{lQ} 
pl.					&	 plural \\
\textsc{plur}		&	 pluractional \\
prep.				&	 \isi{preposition} \\
\textsc{prf}		&	 present perfect \\
pro.				&	 pronoun \\
\textsc{prox}		&	 proximal \\
\textsc{pssm}		&	 impersonal possessum \\ 
\textsc{pst1}		&	 recent past \isi{tense} \\
\textsc{pst2}		&	 removed past \isi{tense} \\
\textsc{pst3}		&	 remote past \isi{tense} \\
\textsc{pst4}		&	 remotest past \isi{tense} \\
quant.				&	 \isi{quantifier} \\
\textsc{real}		&	 realis \\
\textsc{recent}		&	 recentive \\
\textsc{recip}		&	 \isi{reciprocal} \\
\textsc{refl}		&	 \isi{reflexive} \\
rel.				&	 relativizer \\
S					&	 \isi{intransitive} subject \\
\textsc{seq}		&	 sequential \\
sg.					&	 singular \\
\textsc{sim}		&	 simultaneous \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ} 
\textsc{stat}		&	 stative \\
\textsc{subj}		&	 subjunctive \\
subordconn.			&	 subordinating connective \\
v.					&	 verb \\
\textsc{ven}		&	 venitive \\
1					&	 first person \\
2					&	 second person \\
3					&	 third person \\
Ø					&	 zero realization \\
-					&	 morpheme boundary \\
=					&	 \isi{clitic} boundary \\{}
[{\dots}]			&	 phonetic form\\{}
/{\dots}/			&	 phonemic form\\{}
{{\dots}}			&	 morphemic form\\{}
[\'{}]				&	 high tone \\{}
[\`{}]				&	 low tone \\{}
[\^{}]				&	 high-falling tone \\
†					&	 deceased \\
§					&	 section \\ 
\end{tabularx}

  