This is the story about the origin of the bat. There once was a man who had a wife. She, the wife, she had a brother.

One day, the man cut down a sago tree and said to his wife: ``In this place, you have relatives and you have a brother. You can make sago without me.'' Then he went off to the Awin region. He went off and she had to stay alone with her brother. There was the sago which they processed and brought home. After they made the sago, they stored it and she would announce: ``Oh, you just wait, yeah? You all just wait for him to come back. Let him come back first before we eat it''. And the adults would withhold the sago, they never gave the sago to the children.

The parents themselves never gave it to them. But the children begged: ``Oh, come on, please give us some sago to eat!'' The woman and her brother would go and chop sago. They would bring it back home. They would take out the `Ombenep' and would they share any sago? - No, they never gave it to the children! They collected the sago to safe it for later. And they kept doing it. 

One day when the adults were off to make sago, the children had a discussion. ``Guys, what shall we do, what shall we do?'', the older brother said. There were three brothers, the older brother became a big bat in the end. One of them became a `Kudumbop' bat, another became a `Dimin' bat and the last one became a `Tinggulut' bat. They actually were four children.

They took some sticks of a Genemon tree. Then they took some bars of a Taro tree. Finally, they picked some Taro leaves. After picking the taro leaves, one of them asked: ``Guys, how is this done correctly?'' - ``Oh guys, we make it like this!'' So they took some sticks, connected them and then they stood and looked at their faces and their toes, they looked at their feet and hands. ``How can we make it?'' - ``Oh, guys, look here'', so the older brother taught them how it was done. They connected Genemon sticks with the Taro bar. They took a Taro bar and inserted the sticks. They just took the Taro bar and there were some sticks, they would insert those Genemon sticks, stuck it in and just repeated this until they had the claws, like a hook. There is a kind of rattan called `Nong tengge'. The top of Nong tengge is pointy like a hook. They took some of them and they stuck them on the Taro bars. There were five on one side and five on the other side, so it was enough.

It was a Taro bar. But they stuck in it with the Genemon sticks. They stuck and they formed all body parts from that: The shoulder bones, then the ribs, these ribs and the spine. But at this bone here, that was the thighbone.\footnote{In this part, the speaker points at his own body to illustrate.} They set it here and after they set it, they made wings and fur, and then went on to the head and everything around. Then the ears, nose, mouth - they made them all. ``Guys, what about the teeth?'' - ``Oh guys, the teeths should be ...'' They made them and put them on. After they had put them on, they broke it and then they kept making all the body parts the whole day. They made them all and finally, they were standing there. ``Oh wow! Wow, guys!'', one of them said. ``Oh guys, it is working!'' They others were still standing, but he was flying already. The bat was flying. ``Oh! My brothers, it is working'', he said.

Then it was someone else's turn. ``You there, just try to fly!'' - ``Hey, yeah! It is working'' and he was flying. ``Oh, come on boy, it's your turn! It's your turn to fly out now!'' - ``Oh, it's working!'' And again, ``It's working'', that's it. They flew out and everything was completed very well, that's it.

But now let's come back to the parents. I told you about them just recently. It was the story where the brother-in-law went to the Awin region. That was about the origin of the Moon.\footnote{The story and characters seems to be overlapping with the story \textit{The origin of the moon} (\chapref{text:moon})} In the story about the origin of the bat, the married couple had four children. The children did what I have been telling you because of the sago. They stayed at home quietly until it was getting dark but then they started. They began to hang themselves on the roof, the rafter and the shelf. They hang themselves up and stayed quietly. They stayed quietly, they did not talk. But the sago, ``Psh!'', when they kicked it, it went like this ``Pfbrrr!'' It went down like this, it bursted to dust like this. It went up and down to the floor, up and down, and they dropped the sago, and it went like this: ``Pssss!'', the sago. ``Oh my godness!''

When the parents came back home, they did not notice any sign from their children. ``Did the children leave?'' - ``Well, the door is closed firmly! What does this mean? Children, are you in here? Oh, it is quiet in there.'' - ``Did they go out? Oh no, they did not go out!'' They opened the door and ``Pung!'', they kicked it open. ``Oh no, stupid children, what did you do to the sago?'' - ``Oh friend, the sago, they tore it into pieces!'' - ``Oh! Mother! Those damn children!'' That's how their scolding went on and then ``Prrrrr!'', the bats went out through the door. They went out one by one. The Kudumbop bat flew around and the parents said: ``Oh! Let's take the fire tongs that are lying here.'' They tried to hit it exactly here at its spine and so it got a small waist. So, from that day on the Kudumbop bat has a lean waist. And they went out just like that.

The one who flew away first was the Yom bat. But the one she hit with a big fire tongue was the Kudumbop bat. At night, when it flies, it makes a sound like: ``Wauk-wauk wauk-wauk''. When it flies ``Wuk-wuk'' and ``Pak-pak'', that is it. That is the Kudumbop bat. But the Dimin bat usually comes in the evening and makes a sound like: ``Peet-peet peet-peet.'' It is called the Dimin bat in our language. Finally, the Tinggulut bat, when it becomes evening these small bats usually keep flying around here. It's the Tinggulut bat. And that's all of them. They became bats because of the sago. They were children and turned into bats.

Until this day, we do not eat sago when hunting bats. Our ancestors never ate sago. Do not eat sago! Do not drink water! When planning to hunt bats, they usually go without all this. When they go to hunt, they would leave the women and children at home. They would throw a magic spell. And when they find some bats on a Mudut tree, they would cast a certain magic spell. The magic spell goes like this: They first go and urinate. It's for staying calm while shooting. (Nowadays, the foreigners have a gun. A foreign gun, an air soft gun.) They would urinate and whisper the magic spell: ``Lady, lady Tom, lady Tomkon Kayon, you sit here and you sit and stretch your legs like a cross and give milk to your child because I am the one who comes to you.'' He speaks his name and then he says: ``Let me polish it'' and he urinates and then he rubs it at the tree and then
he would shoot his arrows.

It's like that. So, do not eat sago before hunting. They would cast the spell with urine and rub it at the tree. That's how they used to hunt bats. Oh, but nowadays, we live in the modern era. It's the modern era, so it is much easier. They can just go, shoot and bring it back home. In the old days, you would not shoot anything if you had eaten sago before you went hunting. The one over here and the one over there, they would already have flown away. ``Oh, why is this? They were up there, the bats! They became bats because of the problem with the sago! Therefore, the sago is very harmful'', they said. It was the ancestors who said so. The ancestors told it to their children. From their children to their children and from them to us. My friend Yakob's father would usually tell this story. He would always said: ``My children! The bats are not here for no reason.''  And when he told this story we listened to him astonished. Okay, so that's it. That is the origin of the bat.