This is about kid's fights. About their urge to fight. I want to tell this story about their fight.\footnote{It's not fully clear what the speaker meant by this introduction. The story is a narrative about a man outwitting a violent giant. Maybe he meant here that the story is told to violent kids who are provoking fights?}

A long time ago, there was a man who had two wives. One day they went to find sago. He cut down a sago tree and tapped it, while his two wives made the sago.

When he finished tapping, he said to his wives: ``Since I have chopped the sago, you go on squeezing it. You squeeze the sago, while I want to go hunting.'' They agreed and off he went. He was hunting for pig. And indeed, he killed a pig. But while the two women were squeezing the sago, a giant came along and heard them. ``Oh, sounds like someone is making sago. What is it?'' He approached quietly but he saw only the two women. He called them: ``What is going on down there?'' - ``Oh, we are chopping sago.'' So he came and sat down on the trunk of the sago palm. While he was sitting, one of the women squeezed sago and then left her place to take some new sago from where it had been tapped. But the sitting giant lured her:
``Oh friend! Please, look at this here! My string bag here is a very nice one. Come and see it.'' - ``Hmm ... where is it?'', she went to see it but suddenly he grabbed her and put her directly into the bag. Thus he killed her immediately and put her into the bag. The bag was large, very large. 

After that, he kept sitting there on the trunk. While he was sitting there, the other woman was kneading sago. Then she threw the finished sago away and came to take a new hunk. She came and asked: ``Oh! But where is the other woman?'' The giant answered: ``I don't know. She went upstream. Come on, just take that raw sago and squeeze it!'' So she took some raw sago and put it into the midrib of a palm leaf. After doing that she wanted to go. ``Oh, friend! Look here, into my bag! Look at this one, oh, it's a very nice one! It is colorful and very nice so please come here to see it'', the giant said. ``Okay, where is it?'', she went to see it. And suddenly he grabbed her, killed her and put her into the bag. His bag was large, so he could put two women into it. Then he carried them away.

It was beginning to get dark and the giant was carrying the two women whom he had put into the bag. He went on until he passed a slope. Meanwhile, the husband came back and wondered why the sago has not been finished. ``They have not finished squeezing this!'' So he called them repeatedly but did not get any response. They were gone, it was quiet. He quickly put the sago flour into the bag. Then he went on carrying a pig that he had killed. He carried the pig on one shoulder, the sago on the other and thus he went downstream. 

The man went until he faced the giant. He met the giant right there. ``Hey, where did you come from?'', he asked. - ``How about you?'' - ``It has become dark already, where do you want to go?'', he wanted to know. - ``Oh friend!'', the giant said. ``It has become dark now, so here are buttress roots of a Yom tree. On the other side over there, you make your hut and light a fire to sleep; and I will use this side here.'' And so they did. But the man had a bad feeling. ``Oh, he surely hit the two women'', he said. ``He killed them.'' He brought the Yok\footnote{A \textit{Yok} is a piece of pork wrapped in bark. Sometimes vegetables and sago are added too.}, some woods and other stuff and started cooking there. He roasted the pig while the giant was on the other side of the buttress root. While the man was cooking the Yok, he kept the fire lit so that it illuminated the area.
So he could watch the giant taking one of the women out of the bag and swallowing her as a whole. ``Oh, you did it. It was you!'', he thought to himself.

The man went to sleep. The fire was lit to roast the Yok all night long. But the giant who was sitting swallowed another human, like he did before. Then the giant slept very noisily. He slept restlessly and suddenly farted very noisily. ``Wuuu!'', he yelled out repeatedly. But the man saw the giant despite the darkness. ``It is you! You killed my family, my wives!''  He roasted the Yok until the morning and when dawn broke, he opened the Yok and took the pork out. Oh, but he re-tied that Yok. He first defecated into the Yok and then re-tied the Yok and lit the fire again. He put the Yok on the fire again. He raised it ever higher because he was worried that the fire will light up and burn the Yok. Then he fanned the flames and it was lighting up.

Meanwhile, it had become morning, so the giant secretly cast his eye on it. ``Oh, the Yok has not been unwrapped yet.'' It was still on the fire so he fell asleep again. But what the giant did not know: The man actually had opened the Yok. He had put the pork and everything into his bags and went away. He went on until he saw a big river, the Kowo river. There was a small canoe. He took it and also took his bags and stuff and sago like that, all the food. He put everything into the canoe and quickly rowed to the other side. As he arrived at the side, he hid the canoe with some grass, rattan and the wood and left it there. Then he sat down at river bank and some sago and his pork from the roasted Yok.

The giant looked around. ``Oh come on, he did not open the Yok yet!'' He called for the man. ``Hey, what's up? This Yok has not been opened here.'' He kept calling but it was quiet. ``Hey you, where did you go?'', he said then started, he opened the Yok. He opened it and looked inside. There was no pork. ``Oh, this person cheated on me and took off!'' He recognized footprints, there were footprints. He traced him over a long distance. He walked on and on and on until he arrived at the river bank. There he saw the man. ``Oh! He is sitting over there at the other side. Hey you! Where did you cross the river?'', he asked. The man answered: ``Oh, you can just come! Just come like that, come on! There is a bridge\footnote{What is meant with `bridge' here and subsequently is a ford, i.e. a shallow spot in the water where he can wade through the water.} over there. That's how I got here.'' The giant came and said: ``Oh! I think you are cheating on me!'' - ``Just come like that, careful. Come, come, come!'' He talked like that so the giant came until the water reached up to his waist. Then he turned around and saved himself back to the shore. ``Why do you run away? Come again! It's ok, you just come here like that, come! There's a tree you can step on'', he said. So the giant tried again. He went on until the water reached his chest. ``You are lying to me again!'' - ``Why should I keep on cheating you? It is the truth!'', the man said.
He moved along the shore and talked like that: ``Come on! Just come a little bit closer to the middle!'' He went on carrying his belongings. The water rose higher and higher. ``How did you cross the river?'', he said and his mouth was large so he drank water that entered while he was speaking. Now everything happened quickly. The giant sank into the water. He did like that and that's it.

So, that's what they tell to the children. The lesson is this: When we do bad things to other people, then we are cursed. The same will happen to us. It will happen just like that. I usually tell this to the children.
