This is about when we spent the afternoon catching a fish. I want to tell about it.

One afternoon, we were sitting around and wanted to cook some young Cassava leaves. We were just about to start cooking, when suddenly another man came. It was brother Kosmas. ``What are you doing in this afternoon?'', he asked. - ``Oh nothing'', we answered. ``We were about to cook these young Cassava leaves'' - ``Listen. When I was travelling yesterday, I saw a fish, down at Nimi river.'' Immediately, we set the Cassava leaves aside and the three of us went down to Nimi river. We went to the road and we picked some strands of Tenong rattan and took them with us.

We took them down to the river, where we divided the strands of the Tenong rattan and cut off the branch of a tree. Then we tied the rattan to the top of the branch. Having done all that we took it to the cave there, there was mud. We applied some mud to the wood and then took it to the cave where the fish was living. We stuck it into the cave so that a fog would cover the eyes of the fish and make it come out. We stubbed it and when we were done, we took it out again.

After taking it out, one of the guys dived down. Brother Kosmas dived first. He dived and searched, and searched, and searched but he did not find it. He was cold, so he came back out. Then another man, Ansel, he went in next. He searched, and searched, and searched, but he did not find it either. They were both cold so they came out and warmed up in the sun. They were sunbathing and so it was my turn. I went down and searched and searched: ``Oh, it is too difficult.'' I was cold too, so I came back out. I came back out and we were all sunbathing. The sun also shone really well and we could warm up our bodies.

As they were sunbathing, I said: ``You just go on sunbathing, I want to go down again. Let me dive and have another look.'' So I dived down again. I searched it and looked among the stones, among the woods - I searched, but nothing. Then I went to the side of the river and there, from underneath of the stones, I saw it. ``Here it is!'' It was here, so I came out and told them. I said: ``Please hand me that big spear.''
And they immediately concluded: ``If you want a spear, you must have seen it.'' - ``Yes, I have seen it'', I said. ``Beneath the stones, it's lying under the stones.'' They passed the spear on to me and then I went down.

When I went down and shoot at it, it wriggled strongly. It was hazy and wriggled, so I thought I had shot it already. But the next moment I saw it slip away. And I searched again, and searched, and searched. ``Oh, it slipped away'', I said. I searched it, and searched it but there was nothing.
I thought it had gone to another place and so I searched it in other places but did not find it. When I came back to the original place, I noticed it was lying upside down. This time I shot more focused and quickly, the spear went in directly and pierced the fish. Then I held it with the top of the spear and finally took it out.

I took the fish out from the river and put it on the land. As it was lying there, we sat down. After having warmed up a little more in the sun, I turned to brother Kosmas: ``How can we do with the fish? Shall we divide it in halves?'' - ``Oh, not like that. Let's bring it as a whole and when you have cooked it, then you give (some of it) to me'', he said. I agreed. We took the fish in one piece and went to a rib of palm tree, in which we wrapped it. Wrapped in this rib, then we brought it back home. At home we cooked it and then we shared it. We did like that.
