 Bahi looked at the sun. ``Wait.''

 Bahi went into the forest. He went away. Then he came upon the sound of people cutting down a \textit{hawa} palm, \textit{teo, teo}. His future father-in-law was cutting down a \textit{hawa} palm, to mix the fruits with manioc juice. Fish were killed, \textit{ote} fish. The \textit{ote} fish were in the makeshift basket on the ground. They had felt for the fish, and they had killed the fish. They were in the makeshift basket.

 ``Are those fish?''

 ``They are fish. The fish that I killed are in there. Hoo, let me cut down the \textit{hawa} palm, to add the fruits to my wife's manioc juice.'' He picked up the ax and went into the forest. He was cutting down the \textit{hawa} palm.

 ``I'm going to cut down my father-in-law's \textit{hawa} palm.''

 He went there. He went there to cut down his father-in-law's \textit{hawa} palm. He went there, and grabbed the ax. ``Father-in-law, give the ax to me.'' He chopped at the \textit{hawa} palm. He chopped at it really hard. The \textit{hawa} palm fell.

 The two women who were going to be his wives were standing there. ``You two go bathe. Go bathe. I'm going to cut down a \textit{hawa} palm,'' he said.
He went and cut down the \textit{hawa} palm. When the \textit{hawa} palm fell, he came back. He went to the two women who were bathing.

 ``Are you bathing?''

 ``We are bathing.''

 ``I'm going to bathe, too. I'm sweating.'' He bathed. After they bathed, they put their breechcloths back on.

 The two women went back. ``Let's go back.'' They went back.

 He went back after them. ``Come on, I want to drink \textit{hawa} drink. I want to eat fish. I'm hungry.''

 ``You go ahead with him. I will follow,'' their father said.

 ``I'm going to marry you. You are going to be my wives.'' He wanted them, to marry them. He took them back. He arrived back with them.

 ``Cook the fish quickly. I want to eat fish.'' So they cooked the fish. When the cooking of the fish was done, they called him.

 They toasted manioc mash. Her mother mashed the manioc juice in the mortar. She put the \textit{hawa} fruits in water. She mashed the \textit{hawa} fruits in the mortar. She mixed the manioc juice with her hands. ``Well, is he full?''

 ``Yes, I'm full, Mother-in-law. I'm full. I ate fish.''

 ``Aren't you going to drink manioc juice?''

 ``Wait, my food is almost gone.''

 His mother-in-law called him for him to drink manioc juice. He dipped out some manioc juice with a cup. ``Wait, I will dip out a little. I will try a little. Is it sweet?''

 ``It is sweet.''

 He got some manioc juice, and he drank the manioc juice. ``Yumm, it's good. \textit{Hawa} fruits are good with manioc juice.''

 So he came back. He was ready to leave. ``I want to leave, but you won't let me. I'm going to leave.''

 ``We won't let you leave because of what you said, when you said you would marry us.''

 ``Yes, I will marry you.''

 ``Don't deceive us.''

 ``I will marry you. He is my father-in-law.'' He was going to marry them, and he slept with them. He didn't sleep with them, he had sex with them all night.

 ``I'm going now. Where is there forest that has game?''

 ``There is a path going that way that has game. There are woolly monkeys living there.''

 ``Come on, let's go together. You will carry the woolly monkeys.'' He was ready to leave, and his two wives were ready to leave. They came on behind.

 He went into the forest, carrying a blowgun on his shoulder, and holding arrows. He went into the forest. As he went, the voices of woolly monkeys could be heard. ``\textit{Siya, siya.}'' The voices of woolly monkeys could be heard. As he listened to their voices, he twirled darts in his mouth. As he twirled the darts in his mouth, he went along planning against the monkeys.

 ``You stay here. I will come back quickly if I hear your voices.''

 He went into the forest. He went, and he shot some monkeys, \textit{fai saki}.

 ``\textit{Aaa, aaa}.'' A female monkey cried out.

 The monkeys got excited. ``\textit{Aa, aa, aa}.'' He was going to hit them with darts.

 He shot monkeys with the blowgun. He was picking up the dead monkeys. The monkeys were falling onto the ground.

 He came back. ``Hoo, are you still here?''

 ``We are still here.''

 ``Aren't there any dangerous animals?''

 ``There aren't any dangerous animals, and we are still here.''

 ``I guess there isn't a jaguar. If there were a jaguar, you would have cried out. Come on, let's go get the game.''

 They went to get the game. They went with him. He went there and picked up the monkeys.

 They picked up the monkeys. His one wife carried some of the monkeys, his other wife carried some of them, and he carried some of them.

 They came back, and arrived. ``I'm going to tie up the monkeys. You will carry them.''

 ``Aren't you going to carry any of them? Ours should be less.''

 ``You are the ones who are going to eat them.''

 He tied up the animals. He tied up the monkeys. His one wife carried some, and his other wife carried some, and he carried two. He came back. They came back.

 He had sex with them where he had shot the monkeys. He had sex with them. ``Here, I will make someone whose game will be eaten.'' He had sex with them.

 When the sex with them was over, he came back. They came back. ``Come on, you two carry them, the ones who made you suffer.\footnote{That is, the monkeys were the cause of the women having had to have sex with Bahi.} It's so you can eat them.'' He was ready to leave. They were ready to leave.

 They came back, and arrived back. When they came back, his father-in-law singed the hair off the monkeys. He cut up the monkeys. He made a roasting rack, too. He threw them onto the rack. Then they ate, the monkeys' livers, their intestines. They ate, his brothers-in-law. They ate, and they finished eating. They took snuff, and slept.

 He was ready to go again the next morning. ``I'm going out hunting.''

 ``Go hunting. You have seen the forest. You have seen where we go.''

 ``I want my brothers-in-law to go with me.'' His brothers-in-law went with him as he went out into the forest.

 As he went, it [a tapir] started making noise [brushing against leaves].

 ``It's a tapir. Be quiet. I will hit it with an arrow. Wait, watch me. I'm a hunter.''

 They crept toward it as it stood, breaking twigs. One of them inserted an arrow in his bow. [His brother-in-law] was going to shoot quickly, but he went first. \textit{Takai wa}.

 The tapir jumped up. It went away, hurt, and soon fell. He had hit it right in the armpit with the arrow.

 ``Come on, I killed it as it was eating. I really drew the bow hard! The arrow point went inside, and turned up.''

 His brother-in-law pulled out the arrow. ``The arrow point stayed in. I hit you with an arrow. My penis went in all the way in your younger sisters.'' He was talking naughty.

 They cut up the tapir. They cut up the fat tapir. They carried the tapir.

 ``The head will stay here on the ground. Let its head stay here on the ground. My mother-in-law will eat its head. You carry its head.''

 ``I will carry its head. You will see me, wait.''

 He put the tapir's head on top of the full basket. They all carried the tapir.

 ``Don't let the feet stay there on the ground. The feet will be eaten. Let's take it all.''

 They took it. They brought it, and arrived with it.

 ``They are coming back carrying game. While he is here, we are eating the game he kills.''

 He had impregnated his wives, they were pregnant. He spoke to his pregnant wives. ``What are you going to do? Shall I take you away?''

 ``I don't know.''

 ``I will take you, wait. When I hear what my mother says, when I go there and hear what my mother says, if my mother agrees, I will come back to get you. When I have come back to get you, and we have gone...I will sleep here [before leaving]. I will know what day it will be [that I will take you away]. If the babies are growing up when I come back, I will not leave. When the children are finished growing up, we will leave. I will ask my mother properly about it. When I ask my mother properly about it, I will also ask my father.''

 ``Is it scary?''

 ``It's not scary, it's just like where we are now. I went out [hunting] again and came out at another place, and then I came back here. My mother lives there. I came back here.''

 ``It's scary. I don't think I want to go there.''

 ``It's just like where we are now. You will see it. You don't believe me. It's just like where we are now.''

 He was ready to leave, but he turned back and slept there. He turned back because her labor had started. The baby was born.

 ``Is it a boy?''

 ``It's a boy.''

 ``What do you know, it's a boy. You will eat his game [that he will kill when he grows up]. Didn't you see me? You will eat his game. You have seen me, his father. I have killed a tapir with bow and arrow, and have killed woolly monkeys with a blowgun.''

 So he turned back. He went out hunting. ``I'm going hunting, wait. My son has been born.''

 ``Since my son hasn't been born yet, you shouldn't go now. Wait for me,'' [said his other wife].

 ``Yes, I will wait for you. The labor of the other baby will start. When the other baby is born, I will go to speak to my mother.''

 He was going to be ready to leave. When the other baby was born, he was ready to leave. ``The other baby who was born is a boy. What do you know, there are two boys. I will go and speak to my mother, Mother-in-law. Mother-in-law, I will go and speak to my mother, and my father.''

 He was ready to go, and he went. He didn't stay.

 ``I will come back here. I will come back. When the baby is nursing well, I will come back. When I have spoken to my mother...''

 So he went. He went, and disappeared into the forest.

 ``That village of Bahi is scary. I'm not going along.''

 ``He said it's the same as here. Didn't you hear what he said? He said it's just like the place we live in here,'' their older brother said. ``We will go together, so I can have sex with his younger sister, too. Wait, I'll ask him. Let him come back.''

 He was ready to go, and he disappeared into the forest.

 He waited for him. They waited for him. After three full moons, he finally appeared again. ``Bahi has come back. Bahi has come back.''

 He came back, and arrived. ``\textit{Hoo}, the children have grown. Now I finally have children. My sons' grandmother said I should take them there. What are you going to do? Do you want to stay with me?''

 ``I don't know. Bahi, isn't it scary?''

 ``It's not scary. It's the same as where we are now. It's just like here. It's just like where we are now.''

 ``Bahi, is there plenty of game there?''

 ``She is saying there isn't much game, but the game there are the same as where we live,'' [said Bahi].

 They were ready to leave. When they were about ready to leave, they slept there some more.

 ``Wait, I will kill some game for my mother-in-law with my blowgun. When I kill game with my blowgun, my mother-in-law and father-in-law will roast them on a rack.''

 ``Are you intending to take my daughters away? Isn't it scary there?''

 ``Father-in-law, it's not scary. It's the same as the place where we are. The places are the same. It's the same as the place we're living.''

 ``When we are ready to leave, and we go, and we arrive somewhere, we will go and go and will arrive. If we get up early, it won't be far. When we have slept in the house I slept in when I went before, when we have gotten up early and gone and gone, we will arrive,'' he said.

 ``Bahi really talks.''

 He listened to his father-in-law, and his father-in-law stopped speaking.

 ``Yes, Father-in-law, I'm going to take my wives there. I will come back and see you again.''

 ``If you take my daughters to a place that is not my village....''

 ``It's my village. Because it is my village there, I have come here to get my wives. It is my father and mother's village.'' So he took them away.

 ``Let him take them away,'' [said his father-in-law].

 ``He is taking them so their grandmother can see them. They won't die,'' [said his mother-in-law].

 ``I don't know if we will come back soon or not. I don't know if my wives will like the place when they see it.''

 So they went. They slept on the way, and went on again in the morning. He wanted to ask his brother-in-law. ``You said you wanted to marry my younger sister.''

 ``I want you to pay for my sisters to live here.''

 ``No, I don't want her to live over there. I want her to live in my village. You live in my village.''

 ``It's scary.''

 ``It's not scary. You heard me, it's the same forest, for walking. It's the same trails.''

 ``You said the trail goes up.''

 ``It doesn't go up. It's good to walk there. The trees are not high, they are like steps. There are tall trees. It's the same [whether the trees are tall or short].''

 ``I will slide back down.''\footnote{This is a reference to the ability of shamans to walk in the sky.}

 ``He won't slide back down. He crawls because he is a baby. Adults don't crawl.''

 They went. His brother-in-law made him go along. They went and went, and arrived at sundown. He had become Bahi's brother-in-law. He came. They arrived at sundown.

 They slept. He called his younger sister. ``Sister, Sister.''

 ``Huh?''

 ``Sister, come here a minute. Listen to me. Look at what I have in my hand. I am holding something in my hand that I bought.''

 ``Brother, what is it?''

 ``Look at what I am holding in my hand.''

 ``He [the baby] is fat. The one my brother is holding looks just like him. Brother, whose daughters' hands are you holding? Whose daughters did you marry?''

 ``I don't know his name. I haven't heard his name. Wait, I will hear his name now. I will go and see them again.''

 ``Brother, can we go together?''

 ``Yes, we will go together.''

 ``Your partner is sitting there, holding his hard penis [to hide it]. Sister, he is your future husband.''

 ``Are you going to stay there?'' [said Bahi].

 ``I don't know. I don't want to go to his village and stay there.''
