\ea\label{ex:02-a1}
Bai aawa Bai.\\
    \gll bahi a-awa Bahi\\
    sun(\textsc{m}) \textsc{dup}-see Bahi(\textsc{m})\\
    \glt `Bahi looked at the sun.'
\z

\ea\label{ex:02-a2}
Hiba.\\
    \gll hiba\\
    wait\\
    \glt `{``}Wait.'''
\z

\ea\label{ex:02-a3}
Bai tokematamonaka.\\
    \gll Bahi to-ka-himata-mona-ka\\
    Bahi(\textsc{m}) away-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `Bahi went into the forest.'
\z

\ea\label{ex:02-a4}
Toke, tokematamonaka.\\
    \gll to-ke to-ka-himata-mona-ka\\
    away-go.\textsc{m} away-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went away.'
\z

\ea\label{ex:02-a5}
Toke toke ya toke ya, toke me, me moni towasiwitibano, hawa me ka nani mati, teo teo.\\
    \gll to-ke to-ke ya to-ke ya to-ke me me moni to-wasi-witi-hiba-no hawa me kaa na-hani mati teo teo\\
    away-go.\textsc{m} away-go.\textsc{nom}.\textsc{m} \textsc{adju} away-go.\textsc{nom}.\textsc{m} \textsc{adju} away-go.\textsc{m} 3\textsc{pl} 3\textsc{pl}.\textsc{poss} sound away-find-out-\textsc{fut}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m} palm.sp(\textsc{f}) 3\textsc{pl}.\textsc{s} chop \textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{s} (sound\_of\_hitting) (sound\_of\_hitting)\\
    \glt `He went. Then he came upon the sound of people cutting down a \textit{hawa} palm, \textit{teo}, \textit{teo}.'
\z

\ea\label{ex:02-a6}
Kokoriba, kokoribanoho, kokoribano, hawa ka nematamona honofene tehenebona honofene.\\
    \gll kokori-ba kokori-ba-no-ho kokori-ba-no hawa kaa na-himata-mona honofene tehene-bona honofene\\
    3\textsc{sg}.\textsc{poss}.father\_in\_law(\textsc{m})-\textsc{fut} 3\textsc{sg}.\textsc{poss}.father\_in\_law(\textsc{m})-\textsc{fut}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup} 3\textsc{sg}.\textsc{poss}.father\_in\_law(\textsc{m})-\textsc{fut}-\textsc{ip}.\textsc{n}.\textsc{m} palm.sp(\textsc{f}) chop \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} manioc\_juice(\textsc{m}) additive.\textsc{m}-\textsc{int}.\textsc{m} manioc\_juice(\textsc{m})\\
    \glt `His future father-in-law was cutting down a \textit{hawa} palm, to mix the fruits with manioc juice.'
\z

\ea\label{ex:02-a7}
Aba me, aba me waka nemetemoneke oterawaha.\\
    \gll aba me aba me waka na-hemete-mone-ke ote-rawa-ha\\
    fish(\textsc{m}) 3\textsc{pl} fish(\textsc{m}) 3\textsc{pl}.\textsc{s} be\_killed \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} fish.sp(\textsc{f})-\textsc{fpl}-\textsc{dup}\\
    \glt `Fish were killed, \textit{ote} fish.'
\z

\ea\label{ex:02-a8}
Ote ya, tabasi ya ote warawemetemoneke ahi.\\
    \gll ote ya tabasi ya ote waa-rawa-hemete-mone-ke ahi\\
    fish.sp(\textsc{f}) \textsc{adju} makeshift\_basket(\textsc{f}) \textsc{adju} fish.sp(\textsc{f}) stand-\textsc{fpl}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `The \textit{ote} fish were in the makeshift basket on the ground.'
\z

\ea\label{ex:02-a9}
Aba me me boyo nani me aba me me waka nani me, aba me me waka nani aba me, tabasi ya me ihemetemoneke.\\
    \gll aba me me boyo na-hani me aba me me waka na-hani me aba me me waka na-hani aba me tabasi ya me iha-hemete-mone-ke\\
    fish(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} feel \textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{s} fish(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} kill \textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{s} fish(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} kill \textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{f} fish(\textsc{m}) 3\textsc{pl} makeshift\_basket(\textsc{f}) \textsc{adju} 3\textsc{pl}.\textsc{s} be\_located-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They had felt for the fish, and they had killed the fish. They were in the makeshift basket.'
\z

\ea\label{ex:02-a10}
Haa aba me amani?\\
    \gll haa aba me ama-ni\\
    huh? fish(\textsc{m}) 3\textsc{pl}.\textsc{s} be-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Are those fish?'''
\z


\ea\label{ex:02-a11}
Aba me amani.\\
    \gll aba me ama-ni\\
    fish(\textsc{m}) 3\textsc{pl}.\textsc{s} be-\textsc{bkg}.\textsc{f}\\
    \glt `{``}They are fish.'''
\z

\ea\label{ex:02-a12}
Aba me waka ona me waine.\\
    \gll aba me waka o-na me waa-ne\\
    fish(\textsc{m}) 3\textsc{pl}.\textsc{o} kill 1\textsc{sg}.\textsc{s}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} stand-\textsc{cont}.\textsc{f}\\
    \glt `{``}The fish that I killed are in there.'''
\z

\ea\label{ex:02-a13}
Oof, hawa ka okomate, oka yibote ka honofene tehenebonaha.\\
    \gll oof hawa kaa o-ka-mate o-ka yibote ka honofene tehene-bona-ha\\
    hoo palm.sp(\textsc{f}) chop 1\textsc{sg}.\textsc{s}-\textsc{comit}-short\_time.\textsc{imp}.\textsc{f} 1\textsc{sg}.\textsc{poss}-\textsc{poss} wife(\textsc{f}) \textsc{poss} manioc\_juice(\textsc{m}) additive.\textsc{m}-\textsc{int}.\textsc{m}-\textsc{dup}\\
    \glt `{``}Hoo, let me cut down the \textit{hawa} palm, to add the fruits to my wife's manioc juice.'''
\z

\ea\label{ex:02-a14}
Oka yibote ka honofene tehenebona ka okomate mata.\\
    \gll o-ka yibote ka honofene tehene-bona kaa o-ka-mate mata\\
    1\textsc{sg}.\textsc{poss}-\textsc{poss} wife(\textsc{f}) \textsc{poss} manioc\_juice(\textsc{m}) additive.\textsc{m}-\textsc{int}.\textsc{m} chop 1\textsc{sg}.\textsc{s}-\textsc{comit}-short\_time.\textsc{imp}.\textsc{f} for\_now\\
    \glt `{``}Let me cut down the \textit{hawa} palm, to add the fruits to my wife's manioc juice.'''
\z

\ea\label{ex:02-a15}
Bari iti toke, tokematamonaka.\\
    \gll bari iti to-ke to-ka-himata-mona-ka\\
    ax pick\_up.\textsc{m} away-go.\textsc{m} away-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He picked up the ax and went into the forest.'
\z

\ea\label{ex:02-a16}
Hawa ka ne hawa ka ne, Oko koko ka hawa ka okomate.\\
    \gll hawa kaa ne hawa kaa ne o-ka koko ka hawa kaa o-ka-mate\\
    palm.sp(\textsc{f}) chop \textsc{aux}.\textsc{m} palm.sp(\textsc{f}) chop \textsc{aux}.\textsc{m} 1\textsc{sg}.\textsc{poss}-\textsc{poss} father\_in\_law(\textsc{m}) \textsc{poss} palm.sp(\textsc{f}) chop 1\textsc{sg}.\textsc{s}-\textsc{comit}-short\_time.\textsc{imp}.\textsc{f}\\
    \glt `He was cutting down the \textit{hawa} palm. ``I'm going to cut down my father-in-law's \textit{hawa} palm.'''
\z

\ea\label{ex:02-a17}
Tokomakematamonaka.\\
    \gll to-ka-make-himata-mona-ka\\
    away-go-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went there.'
\z

\ea\label{ex:02-a18}
Kokori ka hawa ka kanebona tokomake tokomakematamonaka.\\
    \gll kokori ka hawa kaa ka-na-hibona to-ka-make to-ka-make-himata-mona-ka\\
    3\textsc{sg}.\textsc{poss}.father\_in\_law(\textsc{m}) \textsc{poss} palm.sp(\textsc{f}) chop \textsc{comit}-\textsc{aux}-\textsc{int}.\textsc{m} away-go-following away-go-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went there to cut down his father-in-law's \textit{hawa} palm.'
\z

\ea\label{ex:02-a19}
Tokomake bari wara tone Koko bari owa tiwakawa, ha owa ta.\\
    \gll to-ka-make bari wara to-ne koko bari owa ti-ka-kawa haa owa taa\\
    away-go-following ax grab \textsc{ch}-\textsc{aux}.\textsc{m} father\_in\_law(\textsc{m}) ax 1\textsc{sg}.\textsc{o} 2\textsc{sg}.\textsc{s}-\textsc{comit}-give\_something\_to \textsc{dem} 1\textsc{sg} contrast\\
    \glt `He went there, and grabbed the ax. ``Father-in-law, give the axe to me.'''
\z

\ea\label{ex:02-a20}
Hawa ka kanematamonaka.\\
    \gll hawa kaa ka-na-himata-mona-ka\\
    palm.sp(\textsc{f}) chop \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He chopped at the \textit{hawa} palm.'
\z

\ea\label{ex:02-a21}
Ka nemata kasiro ya ahi.\\
    \gll kaa na-himata kasiro ya ahi\\
    chop \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m} a\_lot \textsc{adju} then\\
    \glt `He chopped at it really hard.'
\z

\ea\label{ex:02-a22}
Hawa towakemetemoneke ahi.\\
    \gll hawa to-ka-ka-hemete-mone-ke ahi\\
    palm.sp(\textsc{f}) away-\textsc{comit}-go-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `The \textit{hawa} palm fell.'
\z

\ea\label{ex:02-a23}
Towakemetemoneke.\\
    \gll to-ka-ka-hemete-mone-ke\\
    away-\textsc{comit}-go-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `It fell.'
\z

\ea\label{ex:02-a24}
Fatiterawabani yoro narawa, Te afi kanahi.\\
    \gll fati-tee-rawa-ba-ni yoro na-rawa tee afi ka-na-hi\\
    3\textsc{sg}.\textsc{poss}.wife(\textsc{f})-\textsc{hab}-\textsc{fpl}-\textsc{fut}-\textsc{ip}.\textsc{n}.\textsc{f} stand \textsc{aux}-\textsc{fpl}.\textsc{f} 2\textsc{pl}.\textsc{s} bathe \textsc{comit}-\textsc{aux}-\textsc{imp}.\textsc{f}\\
    \glt `The two women who were going to be his wives were standing there. ``You two go bathe.'''
\z

\ea\label{ex:02-a25}
Te afi kana.\\
    \gll tee afi ka-na\\
    2\textsc{pl}.\textsc{s} bathe \textsc{comit}-\textsc{aux}\\
    \glt `{``}Go bathe.'''
\z

\ea\label{ex:02-a26}
Hawa ka okomatibe, ati ne hawa ka tokanamaki hawa towaki ya kame, me afi kana me ni ya tokomake tokomakematamonaka.\\
    \gll hawa kaa o-ka-mata-be ati ne hawa kaa to-ka-na-maki hawa to-ka-ki ya ka-me me afi ka-na me ni ya to-ka-make to-ka-make-himata-mona-ka\\
    palm.sp(\textsc{f}) chop 1\textsc{sg}.\textsc{s}-\textsc{comit}-short\_time-\textsc{immed}.\textsc{f} say \textsc{aux}.\textsc{m} palm.sp(\textsc{f}) chop away-\textsc{comit}-\textsc{aux}-following.\textsc{m} palm.sp(\textsc{f}) away-\textsc{comit}-go.\textsc{nom}.\textsc{f} \textsc{adju} come-back.\textsc{m} 3\textsc{pl}.\textsc{s} bathe \textsc{comit}-\textsc{aux}.\textsc{f} 3\textsc{pl} to \textsc{adju} away-go-following away-go-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `{``}I'm going to cut down the \textit{hawa} palm,'' he said. He went and cut down the \textit{hawa} palm. When the \textit{hawa} palm fell, he came back. He went to the two women who were bathing.'
\z

\ea\label{ex:02-a27}
Te afi kini?\\
    \gll tee afi ka-ni\\
    2\textsc{pl}.\textsc{s} bathe \textsc{comit}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Are you bathing?'''
\z

\ea\label{ex:02-a28}
Ota afi kana otani.\\
    \gll otaa afi ka-na otaa-ni\\
    1\textsc{ex}.\textsc{s} bathe \textsc{comit}-\textsc{aux}.\textsc{f} 1\textsc{ex}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}We are bathing.'''
\z


\ea\label{ex:02-a29}
Afi obisamateya.\\
    \gll afi o-bisa-mate-ya\\
    bathe 1\textsc{sg}.\textsc{s}-also-short\_time.\textsc{imp}.\textsc{f}-now\\
    \glt `{``}I'm going to bathe, too.'''
\z

\ea\label{ex:02-a30}
Oka tanakone tokake.\\
    \gll o-ka tanakone to-ka-ke\\
    1\textsc{sg}.\textsc{poss}-\textsc{poss} sweat away-go-\textsc{decl}.\textsc{f}\\
    \glt `{``}I'm sweating.'''
\z

\ea\label{ex:02-a31}
Faya afi ne, afi ne me afi naro, me afi na me, yayafa me akama me, me akama me towakama me, me towakamemetemoneke.\\
    \gll faya afi ne afi ne me afi na-haaro me afi na me yayafa me aka-ma me me aka-ma me to-ka-ka-ma me me to-ka-ka-ma-hemete-mone-ke\\
    so bathe \textsc{aux}.\textsc{m} bathe \textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{s} bathe \textsc{aux}-\textsc{pfut}.\textsc{f} 3\textsc{pl}.\textsc{s} bathe \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} breach\_cloth(\textsc{m}) 3\textsc{pl}.\textsc{s} wear-back.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} wear-back.\textsc{f} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He bathed. After they bathed, they put their breechcloths back on. The two women went back.'
\z

\ea\label{ex:02-a32}
E towakamahi.\\
    \gll ee to-ka-ka-ma-hi\\
    1\textsc{in}.\textsc{s} away-\textsc{comit}-go-back-\textsc{imp}.\textsc{f}\\
    \glt `{``}Let's go back.'''
\z

\ea\label{ex:02-a33}
Faya me, towakama me nowati ya tokomake tokomakematamonaka.\\
    \gll faya me to-ka-ka-ma me nowati ya to-ka-make to-ka-make-himata-mona-ka\\
    so 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back.\textsc{f} 3\textsc{pl}.\textsc{poss} behind \textsc{adju} away-go-following away-go-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `They went back. He went back after them.'
\z

\ea\label{ex:02-a34}
Tokomake, tokomakematamonaka.\\
    \gll to-ka-make to-ka-make-himata-mona-ka\\
    away-go-following away-go-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went back.'
\z

\ea\label{ex:02-a35}
Tokomake, Hima hawa hawa ofibeya.\\
    \gll to-ka-make hima hawa hawa o-fawa-be-ya\\
    away-go-following let's\_go palm.sp(\textsc{f}) palm.sp(\textsc{f}) 1\textsc{sg}.\textsc{s}-drink-\textsc{immed}.\textsc{f}-now\\
    \glt `He went back. ``Come on, I want to drink \textit{hawa} drink.'''
\z

\ea\label{ex:02-a36}
Aba me okobibeya.\\
    \gll aba me o-kaba-be-ya\\
    fish(\textsc{m}) 3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-eat-\textsc{immed}.\textsc{f}-now\\
    \glt `{``}I want to eat fish.'''
\z

\ea\label{ex:02-a37}
Ofimini.\\
    \gll o-fimi-ni\\
    1\textsc{sg}.\textsc{s}-be\_hungry-\textsc{bkg}.\textsc{f}\\
    \glt `{``}I'm hungry.'''
\z

\ea\label{ex:02-a38}
Te tai tokahamaho.\\
    \gll tee tai to-ka-ha-ma-ho\\
    2\textsc{pl}.\textsc{s} be\_ahead \textsc{ch}-\textsc{comit}-\textsc{aux}-back-\textsc{imp}.\textsc{m}\\
    \glt `{``}You go ahead with him.'''
\z

\ea\label{ex:02-a39}
Oyotobeya, me ka abi ati nematamona.\\
    \gll o-yoto-be-ya me ka abi ati na-himata-mona\\
    1\textsc{sg}.\textsc{s}-follow-\textsc{immed}.\textsc{f}-now 3\textsc{pl} \textsc{poss} father(\textsc{m}) say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `{``}I will follow,'' their father said.'
\z

\ea\label{ex:02-a40}
Te oyababana oke.\\
    \gll tee o-yaba-habana o-ke\\
    2\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-marry-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I'm going to marry you.'''
\z

\ea\label{ex:02-a41}
Oka yiboterawaba te amake.\\
    \gll o-ka yibote-rawa-ba tee ama-ke\\
    1\textsc{sg}.\textsc{poss}-\textsc{poss} wife(\textsc{f})-\textsc{fpl}-\textsc{fut} 2\textsc{pl}.\textsc{s} be-\textsc{decl}.\textsc{f}\\
    \glt `{``}You are going to be my wives.'''
\z

\ea\label{ex:02-a42}
Faya mera yabebona mera kayawe mera towakame, mera towakame mera kobo tokaname, Aba me te kawarikabote na.\\
    \gll faya mera yaba-hibona mera ka-yawe mera to-ka-ka-me mera to-ka-ka-me mera kobo to-ka-na-me aba me tee ka-waa-ri-kabote na\\
    so 3\textsc{pl}.\textsc{o} marry-\textsc{int}.\textsc{m} 3\textsc{pl}.\textsc{o} \textsc{comit}-be\_upset.\textsc{m} 3\textsc{pl}.\textsc{o} away-\textsc{comit}-go-back.\textsc{m} 3\textsc{pl}.\textsc{o} away-\textsc{comit}-go-back.\textsc{m} 3\textsc{pl}.\textsc{o} arrive away-\textsc{comit}-\textsc{aux}-back.\textsc{m} fish(\textsc{m}) 3\textsc{pl}.\textsc{o} 2\textsc{pl}.\textsc{s} \textsc{comit}-cook-raised\_surface-quickly \textsc{aux}\\
    \glt `He wanted them, to marry them. He took them back. He arrived back with them. ``Cook the fish quickly.'''
\z

\ea\label{ex:02-a43}
Aba me okobibeya.\\
    \gll aba me o-kaba-be-ya\\
    fish(\textsc{m}) 3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-eat-\textsc{immed}.\textsc{f}-now\\
    \glt `{``}I want to eat fish.'''
\z

\newpage
\ea\label{ex:02-a44}
Faya aba me me kawariemetemone fahi.\\
    \gll faya aba me me ka-waa-ri-hemete-mone fahi\\
    so fish(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{comit}-cook-raised\_surface-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `So they cooked the fish.'
\z

\ea\label{ex:02-a45}
Aba me me kawaria me, aba me kawari ahabi ya, me ha hine iyawa me sira, iyawa me siremetemoneke.\\
    \gll aba me me ka-waa-ria me aba me ka-waa-ri ahabi ya me haa hi-ne iyawa me sira iyawa me sira-hemete-mone-ke\\
    fish(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{comit}-cook-raised\_surface.\textsc{f} 3\textsc{pl}.\textsc{s} fish(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{comit}-be\_cooked-raised\_surface.\textsc{nfin} end.\textsc{nom}.\textsc{f} \textsc{adju} 3\textsc{pl}.\textsc{s} call \textsc{oc}-\textsc{aux}.\textsc{m} grated\_mash(\textsc{f}) 3\textsc{pl}.\textsc{s} toast.\textsc{f} grated\_mash(\textsc{f}) 3\textsc{pl}.\textsc{s} toast-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They cooked the fish. When the cooking of the fish was done, they called him. They toasted manioc mash.'
\z

\ea\label{ex:02-a46}
Mati honofene kawa nati toemetemone.\\
    \gll mati honofene kawa nati to-ha-hemete-mone\\
    3\textsc{sg}.\textsc{poss}.mother(\textsc{f}) manioc\_juice(\textsc{m}) thrust only\_one \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `Her mother mashed the manioc juice in the mortar.'
\z

\ea\label{ex:02-a47}
Hawa hawa kabofa, hawa kawa nisa honofene, tabo na, Fera akare awane?\\
    \gll hawa hawa ka-ibofa hawa kawa na-risa honofene tabo na fee-ra akare awa-ne\\
    palm.sp(\textsc{f}) palm.sp(\textsc{f}) \textsc{comit}-put\_in\_water.\textsc{f} palm.sp(\textsc{f}) thrust \textsc{aux}-down.\textsc{f} manioc\_juice(\textsc{m}) mash \textsc{aux}.\textsc{f} that.\textsc{m}-\textsc{npq} be\_satiated.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `She put the \textit{hawa} fruits in water. She mashed the \textit{hawa} fruits in the mortar. She mixed the manioc juice with her hands. ``Well, is he full?'''
\z

\ea\label{ex:02-a48}
Ee owakara oke aso.\\
    \gll ee o-akara o-ke aso\\
    yes 1\textsc{sg}.\textsc{s}-be\_satiated 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} mother\_in\_law(\textsc{f})\\
    \glt `{``}Yes, I'm full, Mother-in-law.'''
\z

\ea\label{ex:02-a49}
Owakara oke.\\
    \gll o-akara o-ke\\
    1\textsc{sg}.\textsc{s}-be\_satiated 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I'm full.'''
\z

\ea\label{ex:02-a50}
Aba me okoba oke.\\
    \gll aba me o-kaba o-ke\\
    fish(\textsc{m}) 3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-eat 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I ate fish.'''
\z

\ea\label{ex:02-a51}
Aba me okobini.\\
    \gll aba me o-kaba-ni\\
    fish(\textsc{m}) 3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-eat-\textsc{bkg}.\textsc{f}\\
    \glt `{``}I ate fish.'''
\z

\ea\label{ex:02-a52}
Honofene tifaribana?\\
    \gll honofene ti-fawa-ri-bana\\
    manioc\_juice(\textsc{m}) 2\textsc{sg}.\textsc{s}-drink-\textsc{neg}.\textsc{f}-\textsc{fut}\\
    \glt `{``}Aren't you going to drink manioc juice?'''
\z

\ea\label{ex:02-a53}
Hiba, otefe ahabake haro.\\
    \gll hiba o-tefe ahaba-ke haaro\\
    wait 1\textsc{sg}.\textsc{poss}-food end-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}Wait, my food is almost gone.'''
\z

\ea\label{ex:02-a54}
Honofene fawebona hinakaso ha hine, honofene kabi kanematamonaka.\\
    \gll honofene fawa-hibona hinakaso haa hi-ne honofene kabi ka-na-himata-mona-ka\\
    manioc\_juice(\textsc{m}) drink-\textsc{int}.\textsc{m} 3\textsc{sg}.\textsc{poss}.mother\_in\_law(\textsc{f}) call \textsc{oc}-\textsc{aux}.\textsc{m} manioc\_juice(\textsc{m}) dip\_out \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `His mother-in-law called him for him to drink manioc juice. He dipped out some manioc juice with a cup.'
\z

\ea\label{ex:02-a55}
Hiba, kakabi kana omataba.\\
    \gll hiba ka-kabi ka-na o-mata-ba\\
    wait \textsc{dup}-dip\_out \textsc{comit}-\textsc{aux} 1\textsc{sg}.\textsc{s}-short\_time-\textsc{immed}.\textsc{m}\\
    \glt `{``}Wait, I will dip out a little.'''
\z


\newpage
\ea\label{ex:02-a56}
Totomi omataba.\\
    \gll to-tomi o-mata-ba\\
    \textsc{dup}-try 1\textsc{sg}.\textsc{s}-short\_time-\textsc{immed}.\textsc{m}\\
    \glt `{``}I will try a little.'''
\z

\ea\label{ex:02-a57}
Sinatehino ama?\\
    \gll sina-tee-hino ama\\
    be\_sweet-\textsc{hab}-\textsc{ip}.\textsc{n}.\textsc{m} \textsc{sec}\\
    \glt `{``}Is it sweet?'''
\z

\ea\label{ex:02-a58}
Sinate amane.\\
    \gll sina-tee ama-ne\\
    be\_sweet-\textsc{hab} \textsc{sec}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}It is sweet.'''
\z

\ea\label{ex:02-a59}
Honofene kabi kaniha honofene fawa nematamonaka.\\
    \gll honofene kabi ka-niha honofene fawa na-himata-mona-ka\\
    manioc\_juice(\textsc{m}) dip\_out \textsc{comit}-\textsc{caus}.\textsc{list} manioc\_juice(\textsc{m}) drink.\textsc{list} \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He got some manioc juice, and he drank the manioc juice.'
\z

\ea\label{ex:02-a60}
Inaa, amosake.\\
    \gll inaa amosa-ke\\
    hmmm be\_good-\textsc{decl}.\textsc{f}\\
    \glt `{``}Yumm, it's good.'''
\z

\ea\label{ex:02-a61}
Hawa amosini honofene karo.\\
    \gll hawa amosa-ni honofene karo\\
    palm.sp(\textsc{f}) be\_good-\textsc{bkg}.\textsc{f} manioc\_juice(\textsc{m}) \textsc{loc}\\
    \glt `{``}\textit{Hawa} fruits are good with manioc juice.'''
\z

\ea\label{ex:02-a62}
Faya, kame, kamematamonaka.\\
    \gll faya ka-me ka-ma-himata-mona-ka\\
    so come-back.\textsc{m} come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `So he came back.'
\z

\newpage
\ea\label{ex:02-a63}
Kame, yana tone Okabone ona, owa te kayawa teke.\\
    \gll ka-me yana to-ne o-ka-habone o-na owa tee ka-yawa tee-ke\\
    come-back.\textsc{m} start \textsc{ch}-\textsc{aux}.\textsc{m} 1\textsc{sg}.\textsc{s}-go-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{s}-say.\textsc{f} 1\textsc{sg}.\textsc{o} 2\textsc{pl}.\textsc{s} \textsc{comit}-be\_upset.\textsc{f} 2\textsc{pl}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `He came back. He was ready to leave. ``I want to leave, but you won't let me.'''\footnote{Although the word \emph{ona} means `I said' (I said I will leave = I want to leave), the morpheme \emph{na} is an auxiliary, not a verb root meaning `say'. The verb is \emph{ati na} `say', but there is a morphophonological rule that deletes the root \emph{ati} and leaves just the auxiliary, when there is a person prefix (\cite[117]{Dixon:2004xr}). I have glossed it as `say' to make things easier for the reader. When the rule does not apply, i.e. when there is no person prefix, then the root \emph{ati} appears, as in (\ref{ex:02-a26}).}
\z

\ea\label{ex:02-a64}
Okabone onara oke.\\
    \gll o-ka-habone o-na-hara o-ke\\
    1\textsc{sg}.\textsc{s}-go-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{s}-say-\textsc{ip}.\textsc{e}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I'm going to leave.'''
\z

\ea\label{ex:02-a65}
Fara tiwati nima tiwa ota kayawi nineni ota tiyababone tina tiwa.\\
    \gll fara ti-ati ni-ma tiwa otaa ka-yawi na-ne-ni otaa ti-yaba-habone ti-na tiwa\\
    same.\textsc{f} 2\textsc{sg}.\textsc{poss}-voice \textsc{aux}.\textsc{nom}-like 2\textsc{sg}.\textsc{o} 1\textsc{ex}.\textsc{s} \textsc{comit}-be\_upset.\textsc{nfin} \textsc{aux}-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f} 1\textsc{ex}.\textsc{o} 2\textsc{sg}.\textsc{s}-marry-\textsc{int}.\textsc{f} 2\textsc{sg}.\textsc{s}-say.\textsc{f} 2\textsc{sg}.\textsc{s}\\
    \glt `{``}We won't let you leave because of what you said, when you said you would marry us.'''
\z

\ea\label{ex:02-a66}
Ee, te oyababana oke.\\
    \gll ee tee o-yaba-habana o-ke\\
    yes 2\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-marry-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Yes, I will marry you.'''
\z

\ea\label{ex:02-a67}
Ota tikeyeriyahi.\\
    \gll otaa ti-keyeha-ra-yahi\\
    1\textsc{ex}.\textsc{o} 2\textsc{sg}.\textsc{s}-deceive-\textsc{neg}-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `{``}Don't deceive us.'''
\z

\ea\label{ex:02-a68}
Ota tikeyeriyahi.\\
    \gll otaa ti-keyeha-ra-yahi\\
    1\textsc{ex}.\textsc{o} 2\textsc{sg}.\textsc{s}-deceive-\textsc{neg}-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `{``}Don't deceive us.'''
\z

\ea\label{ex:02-a69}
Ota, Te oyababana oke.\\
    \gll otaa tee o-yaba-habana o-ke\\
    1\textsc{ex} 2\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-marry-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I will marry you.'''\footnote{\emph{Ota} is a false start.}
\z

\ea\label{ex:02-a70}
Oka koko oka koko amaka hari.\\
    \gll o-ka koko o-ka koko ama-ka haari\\
    1\textsc{sg}.\textsc{poss}-\textsc{poss} father\_in\_law(\textsc{m}) 1\textsc{sg}.\textsc{poss}-\textsc{poss} father\_in\_law(\textsc{m}) be-\textsc{decl}.\textsc{m} that\_one.\textsc{m}\\
    \glt `{``}He is my father-in-law.'''
\z

\ea\label{ex:02-a71}
Faya me, mera yabebona mera amo kane mera, amo karehibanoho mera ahi nabebana.\\
    \gll faya me mera yaba-hibona mera amo ka-ne mera amo ka-ra-hiba-no-ho mera ahi nahaba-hibana\\
    so 3\textsc{pl} 3\textsc{pl}.\textsc{o} marry-\textsc{int}.\textsc{m} 3\textsc{pl}.\textsc{o} sleep \textsc{comit}-\textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{o} sleep \textsc{comit}-\textsc{neg}-\textsc{fut}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup} 3\textsc{pl}.\textsc{o} have\_sex\_with at\_night-\textsc{fut}.\textsc{m}\\
    \glt `He was going to marry them, and he slept with them. He didn't sleep with them, he had sex with them all night.'
\z

\ea\label{ex:02-a72}
Mera ai nabematamonaka fahi.\\
    \gll mera ahi nahaba-himata-mona-ka fahi\\
    3\textsc{pl}.\textsc{o} have\_sex\_with at\_night-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `He had sex with them all night.'
\z


\ea\label{ex:02-a73}
Mera ahi nabe, Totoka onahabone onara oke.\\
    \gll mera ahi nahabe to-to-ka o-na-habone o-na-hara o-ke\\
    3\textsc{pl}.\textsc{o} have\_sex\_with at\_night.\textsc{m} \textsc{dup}-away-go 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `He had sex with them all night. ``I'm going now.'''
\z

\ea\label{ex:02-a74}
Hibaka, yama abatehani tohari?\\
    \gll hibaka yama ahaba-tee-hani to-ha-ri\\
    where thing(\textsc{f}) die-\textsc{hab}-\textsc{ip}.\textsc{n}.\textsc{f} \textsc{ch}-be-\textsc{cq}.\textsc{f}\\
    \glt `{``}Where is there forest that has game?'''
\z

\ea\label{ex:02-a75}
Yama abate hawi tokine haro.\\
    \gll yama ahaba-tee hawi to-ka-ne haaro\\
    thing(\textsc{f}) die-\textsc{hab} trail.\textsc{f} away-go-\textsc{cont}.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}There is a path going that way that has game.'''
\z

\ea\label{ex:02-a76}
Wafa me winine haro.\\
    \gll wafa me wina-ne haaro\\
    woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{s} live-\textsc{cont}.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}There are woolly monkeys living there.'''
\z

\ea\label{ex:02-a77}
Hima, e tabahi.\\
    \gll hima ee taba-hi\\
    let's\_go 1\textsc{in}.\textsc{s} be\_together-\textsc{imp}.\textsc{f}\\
    \glt `{``}Come on, let's go together.'''
\z

\ea\label{ex:02-a78}
Wafa me te weye nibe.\\
    \gll wafa me tee weye na-be\\
    woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} 2\textsc{pl}.\textsc{s} carry \textsc{aux}-\textsc{immed}.\textsc{f}\\
    \glt `{``}You will carry the woolly monkeys.'''
\z

\ea\label{ex:02-a79}
Faya yana tonematamonaka.\\
    \gll faya yana to-na-himata-mona-ka\\
    so start \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He was ready to leave.'
\z


\ea\label{ex:02-a80}
Yana tone, fati yana tokarawa kayotorawemetemoneke.\\
    \gll yana to-ne fati yana to-ka-rawa ka-yoto-rawa-hemete-mone-ke\\
    start \textsc{ch}-\textsc{aux}.\textsc{m} 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) start \textsc{ch}-\textsc{comit}-\textsc{fpl}.\textsc{f} \textsc{comit}-follow-\textsc{fpl}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He was ready to leave, and his two wives were ready to leave. They came on behind.'
\z

\ea\label{ex:02-a81}
Karaboa tiwa kana wati tama kana, tokematamonaka.\\
    \gll karaboha tiwa ka-na wati tama ka-na to-ka-himata-mona-ka\\
    blowgun(\textsc{f}) carry \textsc{comit}-\textsc{aux}.\textsc{list} arrow(\textsc{m}) hold\_onto \textsc{comit}-\textsc{aux}.\textsc{list} away-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went into the forest, carrying a blowgun on his shoulder, and holding onto arrows.'
\z

\ea\label{ex:02-a82}
Toke, toke toke toke ya wafa me ati saimetemoneke ahi Siya siya siya siya siya siya.\\
    \gll to-ke to-ke to-ke to-ke ya wafa me ati sai-hemete-mone-ke ahi siya siya siya siya siya siya\\
    away-go.\textsc{m} away-go.\textsc{m} away-go.\textsc{m} away-go.\textsc{nom}.\textsc{m} \textsc{adju} woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{poss} voice be\_heard-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey)\\
    \glt `He went into the forest. As he went, the voices of woolly monkeys could be heard. ``\textit{Siya}, \textit{siya}.'''
\z

\largerpage
\ea\label{ex:02-a83}
Wafa me ati saihani me ati mite, sare sare rofi nematamonane.\\
    \gll wafa me ati sai-hani me ati mite sarehe sarehe rofi na-himata-mona-ne\\
    woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{poss} voice be\_heard-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{poss} voice hear.\textsc{m} blowgun\_dart(\textsc{f}) blowgun\_dart(\textsc{f}) twirl\_in\_mouth \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `The voices of woolly monkeys could be heard. As he listened to their voices, he twirled darts in his mouth.'
\z

\ea\label{ex:02-a84}
Sare sare rofi nematamonaka.\\
    \gll sarehe sarehe rofi na-himata-mona-ka\\
    blowgun\_dart(\textsc{f}) blowgun\_dart(\textsc{f}) twirl\_in\_mouth \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He twirled darts in his mouth.'
\z

\ea\label{ex:02-a85}
Sare rofi ne, wafa me wati kane wafa me, tokematamonaka.\\
    \gll sarehe rofi ne wafa me wati ka-ne wafa me to-ka-himata-mona-ka\\
    blowgun\_dart(\textsc{f}) twirl\_in\_mouth \textsc{aux}.\textsc{m} woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} plan\_against \textsc{comit}-\textsc{aux}.\textsc{m} woolly\_monkey(\textsc{m}) 3\textsc{pl} away-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `As he twirled the darts in his mouth, he went along planning against the monkeys.'\footnote{The second occurrence of the string \emph{wafa me} is a false start.}
\z

\ea\label{ex:02-a86}
Te yoyoro karahama nahi.\\
    \gll tee yo-yoro karahama na-hi\\
    2\textsc{pl}.\textsc{s} \textsc{dup}-stay only\_do \textsc{aux}-\textsc{imp}.\textsc{f}\\
    \glt `{``}You stay here.'''
\z

\ea\label{ex:02-a87}
Te yoyoro karahama niyahi.\\
    \gll tee yo-yoro karahama na-yahi\\
    2\textsc{pl}.\textsc{s} \textsc{dup}-stay only\_do \textsc{aux}-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `{``}You stay here.'''
\z

\ea\label{ex:02-a88}
Fito onamabana oke te ati omiti ya.\\
    \gll fito o-na-ma-habana o-ke tee ati o-miti ya\\
    rush 1\textsc{sg}.\textsc{s}-\textsc{aux}-back-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} 2\textsc{pl}.\textsc{poss} voice 1\textsc{sg}.\textsc{s}-hear.\textsc{nom}.\textsc{f} \textsc{adju}\\
    \glt `{``}I will come back quickly if I hear your voices.'''
\z


\ea\label{ex:02-a89}
Tokehimatamonaka.\\
    \gll to-ka-himata-mona-ka\\
    away-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went into the forest.'
\z

\ea\label{ex:02-a90}
Toke wafa me karaboematamonaka ahi, fai saki.\\
    \gll to-ke wafa me karaboha-himata-mona-ka ahi fai\_saki\\
    away-go.\textsc{m} woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} shoot\_with\_blowgun-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then (sound\_of\_blowgun\_dart)\\
    \glt `He went, and he shot some monkeys, \textit{fai} \textit{saki}.'
\z


\newpage
\ea\label{ex:02-a91}
Aaa aaa, wafa kiya kanemetemoneke ahi.\\
    \gll aaa aaa wafa kiya ka-na-hemete-mone-ke ahi\\
    (cry\_of\_pain) (cry\_of\_pain) woolly\_monkey(\textsc{m}) yell \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `{``}\textit{Aaa}, \textit{aaa}.'' A female monkey cried out.'
\z

\ea\label{ex:02-a92}
Wafa me kayawemetemoneke ahi aa aa aa aa aa aa aa aa me fora hinaba mati ahi.\\
    \gll wafa me ka-yawa-hemete-mone-ke ahi aa aa aa aa aa aa aa aa me fora hi-na-haba mati ahi\\
    woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{comit}-be\_excited-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey) (call\_of\_woolly\_monkey) 3\textsc{pl}.\textsc{o} shoot\_with\_blowgun \textsc{oc}-\textsc{aux}-\textsc{fut}.\textsc{f} 3\textsc{pl}.\textsc{o} then\\
    \glt `The monkeys got excited. ``\textit{Aa}, \textit{aa}, \textit{aa}.'' He was going to hit them with darts.'
\z


\ea\label{ex:02-a93}
Wafa me fora nematamonaka ahi.\\
    \gll wafa me fora na-himata-mona-ka ahi\\
    woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} shoot\_with\_blowgun \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `He shot monkeys with the blowgun.'
\z

\ea\label{ex:02-a94}
Wafa me fora ne ya, wafa me fora nematamonaka.\\
    \gll wafa me fora ne ya wafa me fora na-himata-mona-ka\\
    woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} shoot\_with\_blowgun \textsc{aux}.\textsc{nom}.\textsc{m} \textsc{adju} woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} shoot\_with\_blowgun \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He shot monkeys with the blowgun.'
\z

\ea\label{ex:02-a95}
Wafa me itibase, wafa me foro nemetemoneke.\\
    \gll wafa me iti-base wafa me foro na-hemete-mone-ke\\
    woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} pick\_up-edge.\textsc{m} woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{s} fall \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He was picking up the dead monkeys. The monkeys were falling onto the ground.'
\z

\ea\label{ex:02-a96}
Kame kamematamonane.\\
    \gll ka-me ka-ma-himata-mona-ne\\
    come-back.\textsc{m} come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `He came back.'
\z

\ea\label{ex:02-a97}
Kame, Oof te yoro nini?\\
    \gll ka-me oof tee yoro na-ni\\
    come-back.\textsc{m} hoo 2\textsc{pl}.\textsc{s} stay \textsc{aux}-\textsc{bkg}.\textsc{f}\\
    \glt `He came back. ``Hoo, are you still here?'''
\z

\ea\label{ex:02-a98}
Ota yoro nini.\\
    \gll otaa yoro na-ni\\
    1\textsc{ex}.\textsc{s} stay \textsc{aux}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}We are still here.'''
\z

\ea\label{ex:02-a99}
Yama honarini?\\
    \gll yama hona-ra-ni\\
    thing(\textsc{f}) have\_something-\textsc{neg}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Aren't there any dangerous animals?'''
\z

\ea\label{ex:02-a100}
Yama honararani.\\
    \gll yama hona-ra-hara-ni\\
    thing(\textsc{f}) have\_something-\textsc{neg}-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}There aren't any dangerous animals.'''
\z

\ea\label{ex:02-a101}
Yama honara ota, yoro nineni.\\
    \gll yama hona-ra otaa yoro na-ne-ni\\
    thing(\textsc{f}) have\_something-\textsc{neg}.\textsc{f} 1\textsc{ex}.\textsc{s} stay \textsc{aux}-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}There aren't any dangerous animals, and we are still here.'''
\z


\newpage
\ea\label{ex:02-a102}
Yome watare awineke.\\
    \gll yomee wata-re awine-ke\\
    jaguar(\textsc{m}) exist-\textsc{neg}.\textsc{m} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I guess there isn't a jaguar.'''
\z

\ea\label{ex:02-a103}
Yomahi yomahi, yome wehimanaha, kiya tikanene ama tike.\\
    \gll yomahi yomahi yomee waa-himanaha kiya ti-ka-na-hene ama ti-ke\\
    jaguar(\textsc{m}) jaguar(\textsc{m}) jaguar(\textsc{m}) stand-\textsc{hypoth}.\textsc{m} yell 2\textsc{sg}.\textsc{s}-\textsc{comit}-\textsc{aux}-\textsc{irr}.\textsc{f} \textsc{sec} 2\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}If there were a jaguar, you would have cried out.'''
\z

\ea\label{ex:02-a104}
Faya me yama, Hima, bani me e tonakama mata.\\
    \gll faya me yama hima bani me ee to-na-ka-ma mata\\
    so 3\textsc{pl} thing(\textsc{f}) let's\_go animal(\textsc{m}) 3\textsc{pl}.\textsc{o} 1\textsc{in}.\textsc{s} away-\textsc{caus}-go-back for\_now\\
    \glt `{``}Come on, let's go get the game.'''
\z

\ea\label{ex:02-a105}
Bani me me tonakama, me ya tabe tokomematamonaka.\\
    \gll bani me me to-na-ka-ma me ya tabe to-ka-ma-himata-mona-ka\\
    animal(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} away-\textsc{caus}-go-back.\textsc{f} 3\textsc{pl} \textsc{adju} be\_together.\textsc{m} away-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `They went to get the game. They went with him.'
\z

\ea\label{ex:02-a106}
Tokome, tokome wafa me yabematamonaka.\\
    \gll to-ka-me to-ka-me wafa me yaba-himata-mona-ka\\
    away-go-back.\textsc{m} away-go-back.\textsc{m} woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} pick\_up-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went there and picked up the monkeys.'
\z

\ea\label{ex:02-a107}
Wafa me wafa me me yabemetemone.\\
    \gll wafa me wafa me me yaba-hemete-mone\\
    woolly\_monkey(\textsc{m}) 3\textsc{pl} woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} pick\_up-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `They picked up the monkeys.'
\z

\ea\label{ex:02-a108}
Wafa me me yaba me wafa, fati one me tiwa hikana fati one me tiwa hikana ha me tiwa hikana, me kamemetemoneke.\\
    \gll wafa me me yaba me wafa fati one me tiwa hi-ka-na fati one me tiwa hi-ka-na haa me tiwa hi-ka-na me ka-ma-hemete-mone-ke\\
    woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} pick\_up.\textsc{f} 3\textsc{pl}.\textsc{s} woolly\_monkey(\textsc{m}) 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) one.\textsc{f} 3\textsc{pl}.\textsc{o} carry \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{list} 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) other.\textsc{f} 3\textsc{pl}.\textsc{o} carry \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{list} \textsc{dem} 3\textsc{pl}.\textsc{o} carry \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{list} 3\textsc{pl}.\textsc{s} come-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They picked up the monkeys. His one wife carried some of the monkeys, his other wife carried some of them, and he carried some of them. They came back.'
\z

\ea\label{ex:02-a109}
Me kama me, me kamemetemoneke.\\
    \gll me ka-ma me me ka-ma-hemete-mone-ke\\
    3\textsc{pl}.\textsc{s} come-back.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} come-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They came back.'
\z

\ea\label{ex:02-a110}
Me kama me, kobo nama Bani me wete omatibe.\\
    \gll me ka-ma me kobo na-ma bani me wete o-mata-be\\
    3\textsc{pl}.\textsc{s} come-back.\textsc{f} 3\textsc{pl}.\textsc{s} arrive \textsc{aux}-back.\textsc{f} animal(\textsc{m}) 3\textsc{pl}.\textsc{o} tie 1\textsc{sg}.\textsc{s}-short\_time-\textsc{immed}.\textsc{f}\\
    \glt `They came back, and arrived. ``I'm going to tie up the monkeys.'''
\z

\ea\label{ex:02-a111}
Me te weye nikimibe.\\
    \gll me tee weye na-kima-be\\
    3\textsc{pl}.\textsc{o} 2\textsc{pl}.\textsc{s} carry \textsc{aux}-two-\textsc{immed}.\textsc{f}\\
    \glt `{``}You will carry them.'''
\z

\ea\label{ex:02-a112}
Me one weweye kanahirini.\\
    \gll me one we-weye ka-na-hira-ni\\
    3\textsc{pl} other.\textsc{f} \textsc{dup}-be\_carried \textsc{comit}-\textsc{aux}-\textsc{neg}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Aren't you going to carry any of them?'''
\z

\ea\label{ex:02-a113}
Ota ka me tatama rama nahi.\\
    \gll otaa ka me ta-tama rama na-hi\\
    1\textsc{ex} \textsc{poss} 3\textsc{pl}.\textsc{s} \textsc{dup}-be\_many not\_much \textsc{aux}-\textsc{imp}.\textsc{f}\\
    \glt `{``}Ours should be less.'''
\z

\ea\label{ex:02-a114}
Me me hikabarahaba me awine?\\
    \gll me me hi-kaba-ra-haba me awine\\
    3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{oc}-eat-\textsc{neg}-\textsc{fut}.\textsc{f} 3\textsc{pl}.\textsc{o} seem.\textsc{f}\\
    \glt `{``}You are the ones who are going to eat them.'''
\z

\ea\label{ex:02-a115}
Bani me wete nematamonaka.\\
    \gll bani me wete na-himata-mona-ka\\
    animal(\textsc{m}) 3\textsc{pl}.\textsc{o} tie \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He tied up the animals.'
\z

\ea\label{ex:02-a116}
Wafa me wete ne, fati one me weye, fati one me weye, ha, me fama me weye hikana, kame, kame kame kame me kamemetemoneke.\\
    \gll wafa me wete ne fati one me weye fati one me weye haa me fama me weye hi-ka-na ka-me ka-me ka-me ka-me me ka-ma-hemete-mone-ke\\
    woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} tie \textsc{aux}.\textsc{m} 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) one.\textsc{f} 3\textsc{pl}.\textsc{o} carry.\textsc{list} 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) other.\textsc{f} 3\textsc{pl}.\textsc{o} carry.\textsc{list} \textsc{dem} 3\textsc{pl}.\textsc{s} be\_two.\textsc{f} 3\textsc{pl}.\textsc{o} carry \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{list} come-back.\textsc{m} come-back.\textsc{m} come-back.\textsc{m} come-back.\textsc{m} 3\textsc{pl}.\textsc{s} come-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He tied up the monkeys. His one wife carried some, and his other wife carried some, and he carried two. He came back. They came back.'
\z


\ea\label{ex:02-a117}
Me kama me, faya mera mera yori ne mera yori nematamonaka fahi wafa me fora ne ya.\\
    \gll me ka-ma me faya mera mera yori ne mera yori na-himata-mona-ka fahi wafa me fora ne ya\\
    3\textsc{pl}.\textsc{s} come-back.\textsc{f} 3\textsc{pl}.\textsc{s} so 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{o} have\_sex\_with \textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{o} have\_sex\_with \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} shoot\_with\_blowgun \textsc{aux}.\textsc{nom}.\textsc{m} \textsc{adju}\\
    \glt `They came back. He had sex with them where he had shot the monkeys.'
\z

\ea\label{ex:02-a118}
Mera yori nematamonaka.\\
    \gll mera yori na-himata-mona-ka\\
    3\textsc{pl}.\textsc{o} have\_sex\_with \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He had sex with them.'
\z

\newpage
\ea\label{ex:02-a119}
Hiya, ehene kabateba yama omateya.\\
    \gll hiya ehene kaba-tee-hiba yama o-mate-ya\\
    come\_on victim\_of.\textsc{m} be\_eaten-\textsc{hab}-\textsc{fut}.\textsc{m} make 1\textsc{sg}.\textsc{s}-short\_time.\textsc{imp}.\textsc{f}-now\\
    \glt `{``}Here, I will make someone whose game will be eaten.'''
\z

\ea\label{ex:02-a120}
Faya mera yori ne me yori ni abi ya kamematamona fahi.\\
    \gll faya mera yori ne me yori ni ahabi ya ka-ma-himata-mona fahi\\
    so 3\textsc{pl}.\textsc{o} have\_sex\_with \textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{s} be\_laid \textsc{aux}.\textsc{nfin} end.\textsc{nom}.\textsc{f} \textsc{adju} come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `He had sex with them. When the sex with them was over, he came back.'
\z

\ea\label{ex:02-a121}
Me kakama me, Hima me te weye nikima, te me yokoa mati.\\
    \gll me ka-ka-ma me hima me tee weye na-kima tee me yokoa mati\\
    3\textsc{pl}.\textsc{s} \textsc{comit}-come-back.\textsc{f} 3\textsc{pl}.\textsc{s} let's\_go 3\textsc{pl}.\textsc{o} 2\textsc{pl}.\textsc{s} carry \textsc{aux}-two 2\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} bring\_misfortune\_on.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `They came back. ``Come on, you two carry them, the ones who made you suffer.'''
\z


\ea\label{ex:02-a122}
Me te kabibeya.\\
    \gll me tee kaba-be-ya\\
    3\textsc{pl}.\textsc{o} 2\textsc{pl}.\textsc{s} eat-\textsc{immed}.\textsc{f}-now\\
    \glt `{``}It's so you can eat them.'''
\z

\ea\label{ex:02-a123}
Faya me, yana tone me yana tona me kama me, me kama me kobo namemetemoneke.\\
    \gll faya me yana to-ne me yana to-na me ka-ma me me ka-ma me kobo na-ma-hemete-mone-ke\\
    so 3\textsc{pl} start \textsc{ch}-\textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{s} start \textsc{ch}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} come-back.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} come-back.\textsc{f} 3\textsc{pl}.\textsc{s} arrive \textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He was ready to leave. They were ready to leave. They came back, and arrived back.'
\z

\newpage
\ea\label{ex:02-a124}
Tiwahiye tokomake fari.\\
    \gll ti-wahiye to-ka-make faari\\
    2\textsc{sg}.\textsc{s}-hide\_from.\textsc{m} away-go-following that\_one.\textsc{m}\\
    \glt `The [child] you hid from went away.'\footnote{Siko was talking to his wife. She had hid from a child, and the child couldn't find her and left.}
\z

\ea\label{ex:02-a125}
Me kama me kokori kokori wafa me tokematamonaka ahi.\\
    \gll me ka-ma me kokori kokori wafa me toka-himata-mona-ka ahi\\
    3\textsc{pl}.\textsc{s} come-back.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{sg}.\textsc{poss}.father\_in\_law(\textsc{m}) 3\textsc{sg}.\textsc{poss}.father\_in\_law(\textsc{m}) woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} partially\_burn-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `When they came back, his father-in-law singed the hair off the monkeys.'
\z

\ea\label{ex:02-a126}
Me toke wafa me toke wafa me, bobi ne wafa me tana, yama niha, tana ya me were hikanemetemoneke ahi.\\
    \gll me toke wafa me toke wafa me bobi ne wafa me tana yama niha tana ya me were hi-ka-na-hemete-mone-ke ahi\\
    3\textsc{pl}.\textsc{o} partially\_burn.\textsc{m} woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} partially\_burn.\textsc{m} woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} slit \textsc{aux}.\textsc{m} woolly\_monkey(\textsc{m}) 3\textsc{pl} roasting\_rack(\textsc{f}) make \textsc{caus}.\textsc{list} roasting\_rack(\textsc{f}) \textsc{adju} 3\textsc{pl}.\textsc{o} throw \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `He singed the hair off them. He singed the hair off the monkeys. He cut up the monkeys. He made a roasting rack, too. He threw them onto the rack.'
\z

\ea\label{ex:02-a127}
Manakobote me, tafemetemone fahi wafa me, wahati me yoti.\\
    \gll manakobote me tafa-hemete-mone fahi wafa me wahati me yoti\\
    then 3\textsc{pl}.\textsc{s} eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{poss} liver 3\textsc{pl}.\textsc{poss} intestines.\textsc{f}\\
    \glt `Then they ate, the monkeys' livers, their intestines.'
\z

\ea\label{ex:02-a128}
Me tafemetemoneke ahi wabori mati.\\
    \gll me tafa-hemete-mone-ke ahi wabori mati\\
    3\textsc{pl}.\textsc{s} eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) 3\textsc{pl}\\
    \glt `They ate, his brothers-in-law.'
\z

\ea\label{ex:02-a129}
Faya me tafa me, me tafi abemetemoneke fahi.\\
    \gll faya me tafa me me tafi ahaba-hemete-mone-ke fahi\\
    so 3\textsc{pl}.\textsc{s} eat.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} eat.\textsc{nfin} end-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `They ate, and they finished eating.'
\z

\ea\label{ex:02-a130}
Sina me hisi na, sina me hisi na me amo na me, me amo nemetemoneke.\\
    \gll sina me hisi na sina me hisi na me amo na me me amo na-hemete-mone-ke\\
    snuff(\textsc{f}) 3\textsc{pl}.\textsc{s} sniff \textsc{aux}.\textsc{f} snuff(\textsc{f}) 3\textsc{pl}.\textsc{s} sniff \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} sleep \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} sleep \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They took snuff, and slept.'
\z

\ea\label{ex:02-a131}
Me amo na me, yana tonaminatase, Aa yama yete onabone oke.\\
    \gll me amo na me yana to-na-mina-tase aa yama yete o-na-habone o-ke\\
    3\textsc{pl}.\textsc{s} sleep \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} start \textsc{ch}-\textsc{aux}-morning-again.\textsc{m} (pause) thing(\textsc{f}) hunt 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `They slept. He was ready to go again the next morning. ``I'm going out hunting.'''
\z


\ea\label{ex:02-a132}
Yama yete tiyahi.\\
    \gll yama yete ti-yahi\\
    thing(\textsc{f}) hunt 2\textsc{sg}.\textsc{s}-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `{``}Go hunting.'''
\z

\ea\label{ex:02-a133}
Yama tiwahara tini.\\
    \gll yama ti-awa-hara ti-ni\\
    thing(\textsc{f}) 2\textsc{sg}.\textsc{s}-see-\textsc{ip}.\textsc{e}.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}You have seen the forest.'''
\z

\ea\label{ex:02-a134}
Fara e toki ka yama tiwahara tini.\\
    \gll fara ee to-ki ka yama ti-awa-hara ti-ni\\
    same.\textsc{f} 1\textsc{in}.\textsc{s} away-go.\textsc{nom}.\textsc{f} which place(\textsc{f}) 2\textsc{sg}.\textsc{s}-see-\textsc{ip}.\textsc{e}.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}You have seen where we go.'''
\z

\ea\label{ex:02-a135}
Oko wabo me, oko wabo me ya otababana oke.\\
    \gll o-ka wabo me o-ka wabo me ya o-taba-habana o-ke\\
    1\textsc{sg}.\textsc{poss}-\textsc{poss} brother\_in\_law(\textsc{m}) 3\textsc{pl} 1\textsc{sg}.\textsc{poss}-\textsc{poss} brother\_in\_law(\textsc{m}) 3\textsc{pl} \textsc{adju} 1\textsc{sg}.\textsc{s}-be\_together-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I want my brothers-in-law to go with me.'''
\z

\ea\label{ex:02-a136}
Wabori me ya tabematamonane wabori mati.\\
    \gll wabori me ya taba-himata-mona-ne wabori mati\\
    3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) 3\textsc{pl} \textsc{adju} be\_together-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m} 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) 3\textsc{pl}\\
    \glt `His brothers-in-law went with him.'
\z

\ea\label{ex:02-a137}
Wabori me ya tabe toke, tokematamona.\\
    \gll wabori me ya tabe to-ke to-ka-himata-mona\\
    3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) 3\textsc{pl} \textsc{adju} be\_together.\textsc{m} away-go.\textsc{m} away-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `His brothers-in-law went with him as he went out into the forest.'
\z

\ea\label{ex:02-a138}
Toke, moni yana tonematamona.\\
    \gll to-ke moni yana to-na-himata-mona\\
    away-go.\textsc{m} sound start \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `As he went, it started making noise.'
\z

\ea\label{ex:02-a139}
Awi amaka.\\
    \gll awi ama-ka\\
    tapir(\textsc{m}) be-\textsc{decl}.\textsc{m}\\
    \glt `{``}It's a tapir.'''
\z

\ea\label{ex:02-a140}
Te awaraba nahi.\\
    \gll tee awaraba na-hi\\
    2\textsc{pl}.\textsc{s} leave\_something\_alone \textsc{aux}-\textsc{imp}.\textsc{f}\\
    \glt `{``}Be quiet.'''
\z

\ea\label{ex:02-a141}
Te awaraba nahi.\\
    \gll tee awaraba na-hi\\
    2\textsc{pl}.\textsc{s} leave\_something\_alone \textsc{aux}-\textsc{imp}.\textsc{f}\\
    \glt `{``}Be quiet.'''
\z

\ea\label{ex:02-a142}
Sa okobaya ha owa ta.\\
    \gll saa o-ka-ba-ya haa owa taa\\
    shoot\_with\_arrow 1\textsc{sg}.\textsc{s}-\textsc{comit}-\textsc{immed}.\textsc{m}-now \textsc{dem} 1\textsc{sg} contrast\\
    \glt `{``}I will hit it with an arrow.'''
\z

\ea\label{ex:02-a143}
Hibare mata owa te kakatomamatahi.\\
    \gll hibare mata owa tee ka-katoma-mata-hi\\
    wait for\_now 1\textsc{sg}.\textsc{o} 2\textsc{pl}.\textsc{s} \textsc{comit}-look\_at-short\_time-\textsc{imp}.\textsc{f}\\
    \glt `{``}Wait, watch me.'''
\z

\ea\label{ex:02-a144}
Yama tani oha owa awine oke.\\
    \gll yama tani o-ha owa awine o-ke\\
    thing(\textsc{f}) killer 1\textsc{sg}.\textsc{s}-be.\textsc{f} 1\textsc{sg}.\textsc{s} seem.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I'm a hunter.'''
\z

\ea\label{ex:02-a145}
Awa baka ne wahi me yete hikane, owa wati yori tonematamona.\\
    \gll awa baka ne wahi me yete hi-ka-ne owa wati yori to-na-himata-mona\\
    wood(\textsc{f}) break \textsc{aux}.\textsc{m} stand.\textsc{m} 3\textsc{pl}.\textsc{s} hunt \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} one.\textsc{m} arrow(\textsc{m}) insert \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `They crept toward it as it stood, breaking twigs. One of them inserted an arrow in his bow.'
\z


\ea\label{ex:02-a146}
Hoka kabote nebonaha ra yama nimatamonane.\\
    \gll hoka kabote na-hibona-ha ra yama niha-himata-mona-ne\\
    yank quickly \textsc{aux}-\textsc{int}.\textsc{m}-\textsc{dup} \textsc{o} do\_before \textsc{caus}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `[His brother-in-law] was going to shoot quickly, but he went first.'
\z

\ea\label{ex:02-a147}
Takai wa awi yora kamisematamonaka ahi.\\
    \gll takai\_waa awi yora ka-misa-himata-mona-ka ahi\\
    (sound\_of\_arrow\_hitting) tapir(\textsc{m}) jump \textsc{comit}-up-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `\textit{Takai} \textit{wa}. The tapir jumped up.'
\z

\ea\label{ex:02-a148}
Towakame, soni kerewe rebana.\\
    \gll to-ka-ka-me soni kerewe ra-hibana\\
    away-\textsc{comit}-go-back.\textsc{m} fall.\textsc{nfin} be\_slow \textsc{neg}-\textsc{fut}.\textsc{m}\\
    \glt `It went away, hurt, and soon fell.'
\z

\ea\label{ex:02-a149}
Fariboti sa hikane rama ahi.\\
    \gll fariboti saa hi-ka-ne rama ahi\\
    armpit shoot\_with\_arrow \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} unusual then\\
    \glt `He hit it right in the armpit with the arrow.'
\z

\ea\label{ex:02-a150}
Hima, okatomeba ama tafenoho.\\
    \gll hima o-katoma-hiba ama tafa-hino-ho\\
    let's\_go 1\textsc{sg}.\textsc{s}-damage-\textsc{fut}.\textsc{m} \textsc{sec} eat-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `{``}Come on, I killed it as it was eating.'''
\z

\ea\label{ex:02-a151}
Hoka karaba omako amane.\\
    \gll hoka ka-raba o-mako ama-ne\\
    yank \textsc{comit}-a\_bit 1\textsc{sg}.\textsc{s}-for\_this\_reason.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}I really drew the bow hard on it!'''
\z

\ea\label{ex:02-a152}
Wati ino foyahari hari, hari ino nokomisamari hari.\\
    \gll wati ino foya-haari haari haari ino noko-misa-ma-haari haari\\
    arrow(\textsc{m}) point.\textsc{m} be\_inside-look.\textsc{m} that\_one.\textsc{m} that\_one.\textsc{m} point.\textsc{m} point-up-back-\textsc{ip}.\textsc{e}.\textsc{m} that\_one.\textsc{m}\\
    \glt `{``}The arrow point went inside, and turned up.'''
\z

\ea\label{ex:02-a153}
Wati kote, wabori bete name, Wati, ino fawa tonaka.\\
    \gll wati kote wabori bete na-me wati ino fawa to-na-ka\\
    arrow(\textsc{m}) piece 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) pull \textsc{aux}-back.\textsc{m} arrow(\textsc{m}) point.\textsc{m} disappear \textsc{ch}-\textsc{aux}-\textsc{decl}.\textsc{m}\\
    \glt `His brother-in-law pulled out the arrow. ``The arrow point stayed in.'''
\z

\ea\label{ex:02-a154}
Tiwa sa okana, ayata oko sowi foya ama o asima ya.\\
    \gll tiwa saa o-ka-na ayata o-ka sowi foya ama o asima ya\\
    2\textsc{sg}.\textsc{o} shoot\_with\_arrow 1\textsc{sg}.\textsc{s}-\textsc{comit}-\textsc{aux}.\textsc{f} near(\textsc{f}) 1\textsc{sg}.\textsc{poss}-\textsc{poss} penis be\_inside.\textsc{f} \textsc{sec} 1\textsc{sg}.\textsc{s} 2\textsc{sg}.\textsc{poss}.younger\_sister(\textsc{f}) \textsc{adju}\\
    \glt `{``}I hit you with an arrow. My penis went in all the way in your younger sisters.'''
\z

\ea\label{ex:02-a155}
Ati kakasiro raba re ati ama.\\
    \gll ati ka-kasiro raba re ati ama\\
    voice \textsc{dup}-persist a\_bit \textsc{neg}.\textsc{m} voice be\\
    \glt `{``}He is talking naughty.'''
\z

\ea\label{ex:02-a156}
Me, awi me bobi nemetemoneke.\\
    \gll me awi me bobi na-hemete-mone-ke\\
    3\textsc{pl} tapir(\textsc{m}) 3\textsc{pl}.\textsc{s} slit \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They cut up the tapir.'
\z

\ea\label{ex:02-a157}
Awi me bobi na awi, watakeno me bobi na awi me weye nemetemoneke ahi.\\
    \gll awi me bobi na awi wataka-hino me bobi na awi me weye na-hemete-mone-ke ahi\\
    tapir(\textsc{m}) 3\textsc{pl}.\textsc{s} slit \textsc{aux}.\textsc{f} tapir(\textsc{m}) be\_fatty-\textsc{ip}.\textsc{n}.\textsc{m} 3\textsc{pl}.\textsc{s} slit \textsc{aux}.\textsc{f} tapir(\textsc{m}) 3\textsc{pl}.\textsc{s} carry \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `They cut up the tapir. They cut up the fat tapir. They carried the tapir.'
\z

\ea\label{ex:02-a158}
Haha homa haro.\\
    \gll haaha homa haaro\\
    this.\textsc{f} lie\_on\_ground.\textsc{list} that\_one.\textsc{f}\\
    \glt `{``}The head will stay here on the ground.'''
\z


\ea\label{ex:02-a159}
Haa tati homarama naho.\\
    \gll haa tati homa-rama na-ho\\
    \textsc{dem} head lie\_on\_ground-unusual \textsc{aux}-\textsc{imp}.\textsc{m}\\
    \glt `{``}Let its head stay here on the ground.'''
\z

\ea\label{ex:02-a160}
Tati okoso hikabeba amane.\\
    \gll tati okoso hi-kaba-hiba ama-ne\\
    head 1\textsc{sg}.\textsc{poss}.mother\_in\_law(\textsc{f}) \textsc{oc}-eat-\textsc{fut}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}My mother-in-law will eat its head.'''
\z

\ea\label{ex:02-a161}
Tati weye tinaho.\\
    \gll tati weye ti-na-ho\\
    head carry 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{m}\\
    \glt `{``}You carry its head.'''
\z

\ea\label{ex:02-a162}
Tati weye oreba ama.\\
    \gll tati weye o-re-hiba ama\\
    head carry 1\textsc{sg}.\textsc{s}-raised\_surface-\textsc{fut}.\textsc{m} \textsc{sec}\\
    \glt `{``}I will carry its head.'''
\z

\newpage
\ea\label{ex:02-a163}
Owa tiwibeya hiba.\\
    \gll owa ti-awa-be-ya hiba\\
    1\textsc{sg}.\textsc{o} 2\textsc{sg}.\textsc{s}-see-\textsc{immed}.\textsc{f}-now wait\\
    \glt `{``}You will see me, wait.'''
\z

\ea\label{ex:02-a164}
Awi tati nemesematamonane.\\
    \gll awi tati na-meseha-himata-mona-ne\\
    tapir(\textsc{m}) head \textsc{caus}-lie\_on\_top-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `He put the tapir's head on top of the full basket.'
\z

\ea\label{ex:02-a165}
Awi me nafi weye nemetemoneke ahi.\\
    \gll awi me nafi weye na-hemete-mone-ke ahi\\
    tapir(\textsc{m}) 3\textsc{pl}.\textsc{poss} all carry \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `They all carried the tapir.'
\z

\ea\label{ex:02-a166}
Teme kaomarayaho.\\
    \gll teme ka-homa-ra-yaho\\
    foot.\textsc{m} \textsc{comit}-lie\_on\_ground-\textsc{neg}-\textsc{dist}.\textsc{imp}.\textsc{m}\\
    \glt `{``}Don't let the feet stay there on the ground.'''
\z

\ea\label{ex:02-a167}
Teme kabeba amaka.\\
    \gll teme kaba-hiba ama-ka\\
    foot.\textsc{m} be\_eaten-\textsc{fut}.\textsc{m} \textsc{sec}-\textsc{decl}.\textsc{m}\\
    \glt `{``}The feet will be eaten.'''
\z

\ea\label{ex:02-a168}
Teme kabeba amane.\\
    \gll teme kaba-hiba ama-ne\\
    foot.\textsc{m} be\_eaten-\textsc{fut}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}The feet will be eaten.'''
\z

\ea\label{ex:02-a169}
Me, nafi, nafi e towakamaho.\\
    \gll me nafi nafi ee to-ka-ka-ma-ho\\
    3\textsc{pl} all all 1\textsc{in}.\textsc{s} away-\textsc{comit}-go-back-\textsc{dup}\\
    \glt `{``}Let's take it all.'''
\z

\ea\label{ex:02-a170}
Me hiwakamematamonaka.\\
    \gll me hi-ka-ka-ma-himata-mona-ka\\
    3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `They took it.'
\z

\ea\label{ex:02-a171}
Me hiwakame, kobo hikaname, Bani me me weye na me kamini.\\
    \gll me hi-ka-ka-me kobo hi-ka-na-me bani me me weye na me ka-ma-ni\\
    3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-go-back.\textsc{m} arrive \textsc{oc}-\textsc{comit}-\textsc{aux}-back.\textsc{m} animal(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} carry \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} come-back-\textsc{bkg}.\textsc{f}\\
    \glt `They brought it, and arrived with it. ``They are coming back carrying game.'''
\z

\ea\label{ex:02-a172}
Faya, Sawihi ene kakaba tohemate.\\
    \gll faya sawihi ehene ka-kaba to-ha-mate\\
    so be\_present.\textsc{m} victim\_of.\textsc{m} \textsc{dup}-be\_eaten \textsc{ch}-\textsc{aux}-short\_time.\textsc{m}\\
    \glt `{``}While he is here, we are eating the game he kills.'''
\z

\ea\label{ex:02-a173}
Fati naba toniharawe, naba toharawemetemoneke.\\
    \gll fati naba to-niha-rawe naba to-ha-rawa-hemete-mone-ke\\
    3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) pregnant \textsc{ch}-\textsc{caus}-\textsc{fpl}.\textsc{m} pregnant \textsc{ch}-be-\textsc{fpl}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He had impregnated his wives, they were pregnant.'
\z

\ea\label{ex:02-a174}
Me nabati weye ni ya, faya ati yana tonematamona fahi.\\
    \gll me nabati weye ni ya faya ati yana to-na-himata-mona fahi\\
    3\textsc{pl}.\textsc{poss} stomach carry \textsc{aux}.\textsc{nom}.\textsc{f} \textsc{adju} so voice start \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `He spoke to his pregnant wives.'
\z

\ea\label{ex:02-a175}
Himata nima te naba te amari?\\
    \gll himata ni-ma tee na-haba tee ama-ri\\
    what \textsc{aux}.\textsc{nom}-like 2\textsc{pl}.\textsc{s} \textsc{aux}-\textsc{fut}.\textsc{f} 2\textsc{pl}.\textsc{s} \textsc{sec}-\textsc{cq}.\textsc{f}\\
    \glt `{``}What are you going to do?'''
\z

\ea\label{ex:02-a176}
Tera owakamibana?\\
    \gll tera o-ka-ka-mi-bana\\
    2\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-\textsc{comit}-go-back.\textsc{f}-\textsc{fut}\\
    \glt `{``}Shall I take you away?'''
\z

\ea\label{ex:02-a177}
E awarini.\\
    \gll ee awa-ra-ni\\
    1\textsc{in}.\textsc{s} see-\textsc{neg}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}I don't know.'''
\z

\ea\label{ex:02-a178}
Te owakamaba hiba.\\
    \gll tee o-ka-ka-ma-haba hiba\\
    2\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-\textsc{comit}-go-back-\textsc{fut}.\textsc{f} wait\\
    \glt `{``}I will take you, wait.'''
\z

\ea\label{ex:02-a179}
Okomi okomi ati omiti ya, okomi ati omiti ya okomi ati omitami ya okomi ati amosi ya, tera onakamabana oke.\\
    \gll okomi okomi ati o-miti ya okomi ati o-miti ya okomi ati o-mita-mi ya okomi ati amosi ya tera o-na-ka-ma-habana o-ke\\
    1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) 1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) voice 1\textsc{sg}.\textsc{s}-hear.\textsc{nom}.\textsc{f} \textsc{adju} 1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) voice 1\textsc{sg}.\textsc{s}-hear.\textsc{nom}.\textsc{f} \textsc{adju} 1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) voice 1\textsc{sg}.\textsc{s}-hear-back.\textsc{nom}.\textsc{f} \textsc{adju} 1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) voice be\_good.\textsc{nom}.\textsc{f} \textsc{adju} 2\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-\textsc{caus}-come-back-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}When I hear what my mother says, when I go there and hear what my mother says, if my mother agrees, I will come back to get you.'''
\z

\ea\label{ex:02-a180}
Tera onakamabana oni.\\
    \gll tera o-na-ka-ma-habana o-ni\\
    2\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-\textsc{caus}-come-back-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}I will come back to get you.'''
\z

\ea\label{ex:02-a181}
Tera onakamaharo, e tokomaro, amo onamiba ama ahi.\\
    \gll tera o-na-ka-ma-haaro ee to-ka-ma-haaro amo o-na-mi-ba ama ahi\\
    2\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-\textsc{caus}-come-back-\textsc{pfut}.\textsc{f} 1\textsc{in}.\textsc{s} away-go-back-\textsc{pfut}.\textsc{f} sleep 1\textsc{sg}.\textsc{s}-\textsc{aux}-back.\textsc{nom}.\textsc{f}-\textsc{fut} be then\\
    \glt `{``}When I have come back to get you, and we have gone...I will sleep here.'''
\z

\ea\label{ex:02-a182}
Oko amo ni, owatoahaba owa awine oke.\\
    \gll o-ka amo ni o-wato-haba owa awine o-ke\\
    1\textsc{sg}.\textsc{poss}-\textsc{poss} sleep \textsc{aux}.\textsc{nfin} 1\textsc{sg}.\textsc{s}-know-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s} seem.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I will know what day it will be.'''
\z

\newpage
\ea\label{ex:02-a183}
Mate ihaha ihi owasimi ya, ma owahama, onabana oke mata.\\
    \gll matehe ihaha ihi o-wasi-mi ya maa o-waha-ma o-na-habana o-ke mata\\
    children(\textsc{f}) be\_born.\textsc{f} result\_of.\textsc{f} 1\textsc{sg}.\textsc{s}-find-back.\textsc{nom}.\textsc{f} \textsc{adju} stop 1\textsc{sg}.\textsc{s}-change-back.\textsc{list} 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} for\_now\\
    \glt `{``}If the babies are growing up when I come back, I will not leave.'''
\z

\ea\label{ex:02-a184}
Mate fotamisi forima ni ya e tokomawahaba ereke.\\
    \gll matehe fota-misi forima ni ya ee to-ka-ma-waha-haba ere-ke\\
    children(\textsc{f}) be\_big-up.\textsc{nfin} do\_properly \textsc{aux}.\textsc{nom}.\textsc{f} \textsc{adju} 1\textsc{in}.\textsc{s} away-go-back-change-\textsc{fut}.\textsc{f} 1\textsc{in}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}When the children are finished growing up, we will leave.'''
\z

\ea\label{ex:02-a185}
Okomi aate kani forima, onabana oke ahi.\\
    \gll okomi a-ate ka-ni forima o-na-habana o-ke ahi\\
    1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) \textsc{dup}-ask \textsc{comit}-\textsc{aux}.\textsc{nfin} do\_properly 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} then\\
    \glt `{``}I will ask my mother properly about it.'''
\z

\ea\label{ex:02-a186}
Okomi aate kani forima oni ya okobi, okobi aate.\\
    \gll okomi a-ate ka-ni forima o-ni ya okobi okobi a-ate\\
    1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) \textsc{dup}-ask \textsc{comit}-\textsc{aux}.\textsc{nfin} do\_properly 1\textsc{sg}.\textsc{s}-\textsc{aux}.\textsc{nom}.\textsc{f} \textsc{adju} 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) \textsc{dup}-ask.\textsc{list}\\
    \glt `{``}When I ask my mother properly about it, I will also ask my father.'''
\z

\ea\label{ex:02-a187}
Yama yofinara awineni?\\
    \gll yama yofina-ra awine-ni\\
    place(\textsc{f}) cause\_fear-\textsc{neg}.\textsc{f} seem.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Is it scary?'''
\z

\ea\label{ex:02-a188}
Yama yofinara fara nima na ya e naoneke.\\
    \gll yama yofina-ra fara ni-ma na ya ee naho-ne-ke\\
    place(\textsc{f}) cause\_fear-\textsc{neg}.\textsc{f} same.\textsc{f} \textsc{aux}.\textsc{nom}-like \textsc{aux}.\textsc{f} \textsc{adju} 1\textsc{in}.\textsc{s} stay-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}It's not scary, it's just like where we are now.'''
\z

\ea\label{ex:02-a189}
Fara nima na ya e naoneni.\\
    \gll fara ni-ma na ya ee naho-ne-ni\\
    same.\textsc{f} \textsc{aux}.\textsc{nom}-like \textsc{aux}.\textsc{f} \textsc{adju} 1\textsc{in}.\textsc{s} stay-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}It's just like where we are now.'''
\z

\ea\label{ex:02-a190}
Okoma ahi tasi owitia, okoma ama o ahi.\\
    \gll o-ka-ma ahi tasi o-witia o-ka-ma ama o ahi\\
    1\textsc{sg}.\textsc{s}-go-back.\textsc{f} there emerge 1\textsc{sg}.\textsc{s}-out.\textsc{f} 1\textsc{sg}.\textsc{s}-come-back.\textsc{f} \textsc{sec} 1\textsc{sg}.\textsc{s} here\\
    \glt `{``}I went out again and came out at another place, and then I came back here.'''
\z

\ea\label{ex:02-a191}
Okoma ama ahi.\\
    \gll o-ka-ma ama ahi\\
    1\textsc{sg}.\textsc{s}-come-back.\textsc{f} \textsc{sec} here\\
    \glt `{``}I came here.'''
\z

\ea\label{ex:02-a192}
Okomi winineke haro.\\
    \gll okomi wina-ne-ke haaro\\
    1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) live-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}My mother lives there.'''
\z

\ea\label{ex:02-a193}
Okomine ahi.\\
    \gll o-ka-ma-ne ahi\\
    1\textsc{sg}.\textsc{s}-come-back-\textsc{cont}.\textsc{f} here\\
    \glt `{``}I came back here.'''
\z

\ea\label{ex:02-a194}
Okoma ama ahi.\\
    \gll o-ka-ma ama ahi\\
    1\textsc{sg}.\textsc{s}-come-back.\textsc{f} \textsc{sec} here\\
    \glt `{``}I came here.'''
\z

\ea\label{ex:02-a195}
Yama yofina awineni.\\
    \gll yama yofina awine-ni\\
    place(\textsc{f}) cause\_fear.\textsc{f} seem.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}It's scary.'''
\z

\ea\label{ex:02-a196}
Yama owati tokomi nonofa ra.\\
    \gll yama o-ati to-ka-mi no-nofa ra\\
    place(\textsc{f}) 1\textsc{sg}.\textsc{poss}-voice away-go-back.\textsc{nfin} \textsc{dup}-want \textsc{neg}.\textsc{f}\\
    \glt `{``}I don't think I want to go there.'''\footnote{The word \emph{yama} appears to be a false start.}
\z

\ea\label{ex:02-a197}
Yama fara nima na ya e naoneni.\\
    \gll yama fara ni-ma na ya ee naho-ne-ni\\
    place(\textsc{f}) same.\textsc{f} \textsc{aux}.\textsc{nom}-like \textsc{aux}.\textsc{f} \textsc{adju} 1\textsc{in}.\textsc{s} stay-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}It's just like where we are now.'''
\z

\ea\label{ex:02-a198}
Te te awabana.\\
    \gll tee tee awa-habana\\
    2\textsc{pl} 2\textsc{pl}.\textsc{s} see-\textsc{fut}.\textsc{f}\\
    \glt `{``}You will see it.'''
\z

\ea\label{ex:02-a199}
Te awibe.\\
    \gll tee awa-be\\
    2\textsc{pl}.\textsc{s} see-\textsc{immed}.\textsc{f}\\
    \glt `{``}You will see it.'''
\z

\ea\label{ex:02-a200}
Owati te kasawari.\\
    \gll o-ati tee ka-sawari\\
    1\textsc{sg}.\textsc{poss}-voice 2\textsc{pl}.\textsc{s} \textsc{comit}-frustrate.\textsc{nom}\\
    \glt `{``}You don't believe me.'''
\z

\ea\label{ex:02-a201}
Fara nima na ya e naoneke.\\
    \gll fara ni-ma na ya ee naho-ne-ke\\
    same.\textsc{f} \textsc{aux}.\textsc{nom}-like \textsc{aux}.\textsc{f} \textsc{adju} 1\textsc{in}.\textsc{s} stay-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}It's just like where we are now.'''
\z

\ea\label{ex:02-a202}
Faya yana tonematamonaka.\\
    \gll faya yana to-na-himata-mona-ka\\
    so start \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He was ready to leave.'
\z

\ea\label{ex:02-a203}
Yana tonebona, ma nawahame, amo name, inamatewe komene yana neno, inamatewe komene yana ne komene, ehene ma nawahame, faya inamatewe ihe, Maki ama?\\
    \gll yana to-na-hibona maa na-waha-me amo na-me inamatewe komene yana na-hino inamatewe komene yana ne komene ehene maa na-waha-me faya inamatewe ihe maki ama\\
    start \textsc{ch}-\textsc{aux}-\textsc{int}.\textsc{m} stop \textsc{aux}-change-back.\textsc{m} sleep \textsc{aux}-back.\textsc{m} child(\textsc{m}) pain.\textsc{m} start \textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{m} child(\textsc{m}) pain.\textsc{m} start \textsc{aux}.\textsc{m} pain.\textsc{m} because\_of.\textsc{m} stop \textsc{aux}-change-back.\textsc{m} so child(\textsc{m}) be\_born.\textsc{m} male(\textsc{m}) be\\
    \glt `He was ready to leave, but he turned back and slept there. He turned back because her labor had started. The baby was born. ``Is it a boy?'''
\z

\ea\label{ex:02-a204}
Maki amane.\\
    \gll maki ama-ne\\
    male(\textsc{m}) be-\textsc{bkg}.\textsc{m}\\
    \glt `{``}It's a boy.'''
\z

\ea\label{ex:02-a205}
Maki amakenoho.\\
    \gll maki ama-ke-no-ho\\
    male(\textsc{m}) be-\textsc{decl}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `{``}What do you know, it's a boy.'''
\z

\ea\label{ex:02-a206}
Ene tikabateba amane.\\
    \gll ehene ti-kaba-tee-hiba ama-ne\\
    victim\_of.\textsc{m} 2\textsc{sg}.\textsc{s}-eat-\textsc{hab}-\textsc{fut}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}You will eat his game.'''
\z

\ea\label{ex:02-a207}
Owa tiwarini?\\
    \gll owa ti-awa-ra-ni\\
    1\textsc{sg}.\textsc{o} 2\textsc{sg}.\textsc{s}-see-\textsc{neg}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Didn't you see me?'''
\z

\ea\label{ex:02-a208}
Ene tikabateba amaka.\\
    \gll ehene ti-kaba-tee-hiba ama-ka\\
    victim\_of.\textsc{m} 2\textsc{sg}.\textsc{s}-eat-\textsc{hab}-\textsc{fut}.\textsc{m} \textsc{sec}-\textsc{decl}.\textsc{m}\\
    \glt `{``}You will eat his game.'''
\z

\ea\label{ex:02-a209}
Owa tiwa tike haro bati owa.\\
    \gll owa ti-awa ti-ke haaro bati owa\\
    1\textsc{sg}.\textsc{o} 2\textsc{sg}.\textsc{s}-see 2\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} that\_one.\textsc{f} 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) 1\textsc{sg}\\
    \glt `{``}You have seen me, his father.'''
\z

\ea\label{ex:02-a210}
Awi sa okana wafa me fora.\\
    \gll awi saa o-ka-na wafa me fora\\
    tapir(\textsc{m}) shoot\_with\_arrow 1\textsc{sg}.\textsc{s}-\textsc{comit}-\textsc{aux}.\textsc{list} woolly\_monkey(\textsc{m}) 3\textsc{pl}.\textsc{o} shoot\_with\_blowgun.\textsc{list}\\
    \glt `{``}I have killed a tapir with bow and arrow, and have killed woolly monkeys with a blowgun.'''
\z

\ea\label{ex:02-a211}
Faya kamematamonaka.\\
    \gll faya ka-ma-himata-mona-ka\\
    so come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `So he came back.'
\z

\ea\label{ex:02-a212}
Faya, ma nawame yama yete nebona toka.\\
    \gll faya maa na-waha-me yama yete na-hibona to-ka\\
    so stop \textsc{aux}-change-back.\textsc{m} thing(\textsc{f}) hunt \textsc{aux}-\textsc{int}.\textsc{m} away-go.\textsc{list}\\
    \glt `So he turned back. He went out hunting.'
\z


\ea\label{ex:02-a213}
Yama yete omate, hiba.\\
    \gll yama yete o-mate hiba\\
    thing(\textsc{f}) hunt 1\textsc{sg}.\textsc{s}-short\_time.\textsc{imp}.\textsc{f} wait\\
    \glt `{``}I'm going hunting, wait.'''
\z

\ea\label{ex:02-a214}
Okatao ihaka.\\
    \gll okatao iha-ka\\
    1\textsc{sg}.\textsc{poss}.son(\textsc{m}) be\_born-\textsc{decl}.\textsc{m}\\
    \glt `{``}My son has been born.'''
\z

\ea\label{ex:02-a215}
Okatao okatao okatao ihare ya tikamariyahi mata.\\
    \gll okatao okatao okatao iha-re ya ti-ka-ma-ra-yahi mata\\
    1\textsc{sg}.\textsc{poss}.son(\textsc{m}) 1\textsc{sg}.\textsc{poss}.son(\textsc{m}) 1\textsc{sg}.\textsc{poss}.son(\textsc{m}) be\_born-\textsc{neg}.\textsc{nom}.\textsc{m} \textsc{adju} 2\textsc{sg}.\textsc{s}-go-back-\textsc{neg}-\textsc{dist}.\textsc{imp}.\textsc{f} for\_now\\
    \glt `{``}Since my son hasn't been born yet, you shouldn't go now.'''
\z

\ea\label{ex:02-a216}
Owa noki tinahi mata.\\
    \gll owa noki ti-na-hi mata\\
    1\textsc{sg}.\textsc{o} wait\_for 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{f} for\_now\\
    \glt `{``}Wait for me.'''
\z

\ea\label{ex:02-a217}
Ee, tiwa noki onabana o.\\
    \gll ee tiwa noki o-na-habana o\\
    yes 2\textsc{sg}.\textsc{o} wait\_for 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}\\
    \glt `{``}Yes, I will wait for you.'''
\z

\ea\label{ex:02-a218}
Inamatewe owa komene yana namakiba amane.\\
    \gll inamatewe owa komene yana na-maki-hiba ama-ne\\
    child(\textsc{m}) other.\textsc{m} pain.\textsc{m} start \textsc{aux}-following-\textsc{fut}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}The labor of the other baby will start.'''
\z

\ea\label{ex:02-a219}
Inamatewe owa ihamaki ya okomabana oke, okomi narabi nima onaba owa.\\
    \gll inamatewe owa iha-maki ya o-ka-ma-habana o-ke okomi narabi nima o-na-haba owa\\
    child(\textsc{m}) other.\textsc{m} be\_born-following.\textsc{nom} \textsc{adju} 1\textsc{sg}.\textsc{s}-go-back-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} 1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) ear.\textsc{f} communicate\_to 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}\\
    \glt `{``}When the other baby is born, I will go to speak to my mother.'''
\z

\ea\label{ex:02-a220}
Faya yana yana toneba inamatewe owa ihamaki ya, yana tone, yana tone Inamatewe owa ihamake maki amaka.\\
    \gll faya yana yana to-na-hiba inamatewe owa iha-maki ya yana to-ne yana to-ne inamatewe owa iha-make maki ama-ka\\
    so start start \textsc{ch}-\textsc{aux}-\textsc{fut}.\textsc{m} child(\textsc{m}) other.\textsc{m} be\_born-following.\textsc{nom} \textsc{adju} start \textsc{ch}-\textsc{aux}.\textsc{m} start \textsc{ch}-\textsc{aux}.\textsc{m} child(\textsc{m}) other.\textsc{m} be\_born-following male(\textsc{m}) be-\textsc{decl}.\textsc{m}\\
    \glt `He was going to be ready to leave. When the other baby was born, he was ready to leave. ``The other baby who was born is a boy.'''
\z

\ea\label{ex:02-a221}
Maki me faminekeni.\\
    \gll maki me fama-ne-ke-ni\\
    male(\textsc{m}) 3\textsc{pl}.\textsc{s} be\_two-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}\\
    \glt `{``}What do you know, there are two boys.'''
\z

\newpage
\ea\label{ex:02-a222}
Okomi narabi nima onamamatibe aso.\\
    \gll okomi narabi nima o-na-ma-mata-be aso\\
    1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) ear.\textsc{f} communicate\_to 1\textsc{sg}.\textsc{s}-\textsc{aux}-back-short\_time-\textsc{immed}.\textsc{f} mother\_in\_law(\textsc{f})\\
    \glt `{``}I will go and speak to my mother, Mother-in-law.'''
\z

\ea\label{ex:02-a223}
Aso okomi narabi nima onamamatibe okobi.\\
    \gll aso okomi narabi nima o-na-ma-mata-be okobi\\
    mother\_in\_law(\textsc{f}) 1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) ear.\textsc{f} communicate\_to 1\textsc{sg}.\textsc{s}-\textsc{aux}-back-short\_time-\textsc{immed}.\textsc{f} 1\textsc{sg}.\textsc{poss}.father(\textsc{m})\\
    \glt `{``}Mother-in-law, I will go and speak to my mother, and my father.'''
\z

\ea\label{ex:02-a224}
Faya yana tone, tokomematamonaka.\\
    \gll faya yana to-ne to-ka-ma-himata-mona-ka\\
    so start \textsc{ch}-\textsc{aux}.\textsc{m} away-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He was ready to go, and he went.'
\z

\ea\label{ex:02-a225}
Warematamonaka.\\
    \gll waa-ra-himata-mona-ka\\
    stay-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He didn't stay.'
\z

\ea\label{ex:02-a226}
Okomabana oke ahi.\\
    \gll o-ka-ma-habana o-ke ahi\\
    1\textsc{sg}.\textsc{s}-come-back-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} here\\
    \glt `{``}I will come back here.'''
\z

\ea\label{ex:02-a227}
Okomabana oni.\\
    \gll o-ka-ma-habana o-ni\\
    1\textsc{sg}.\textsc{s}-come-back-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}I will come back.'''
\z

\ea\label{ex:02-a228}
Inamatewe bowara namise ya, okomaba owa awine oke.\\
    \gll inamatewe bowara na-mise ya o-ka-ma-haba owa awine o-ke\\
    child(\textsc{m}) nurse\_well \textsc{aux}-up.\textsc{nom}.\textsc{m} \textsc{adju} 1\textsc{sg}.\textsc{s}-come-back-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s} seem.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}When the baby is nursing well, I will come back.'''
\z

\ea\label{ex:02-a229}
Okomi narabi nima onaharo....\\
    \gll okomi narabi nima o-na-haaro\\
    1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) ear.\textsc{f} communicate\_to 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{pfut}.\textsc{f}\\
    \glt `{``}When I have spoken to my mother....'''\footnote{Siko didn't finish the sentence.}
\z

\ea\label{ex:02-a230}
Faya tokomematamona fahi.\\
    \gll faya to-ka-ma-himata-mona fahi\\
    so away-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `So he went.'
\z

\ea\label{ex:02-a231}
Tokome fawa tosame, Hahi taboro Bahi taboro yofine awane.\\
    \gll to-ka-me fawa to-kosa-me haahi taboro Bahi taboro yofine awa-ne\\
    away-go-back.\textsc{m} disappear \textsc{ch}-middle-back.\textsc{m} this\_one.\textsc{m} place.\textsc{m} Bahi(\textsc{m}) place.\textsc{m} cause\_fear.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `He went, and disappeared into the forest. ``That village of Bahi is scary.'''
\z

\ea\label{ex:02-a232}
Osawirabana oni.\\
    \gll o-sawi-ra-habana o-ni\\
    1\textsc{sg}.\textsc{s}-be\_present-\textsc{neg}-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}I'm not going along.'''
\z

\ea\label{ex:02-a233}
Yama fara toamone ati nareka.\\
    \gll yama fara to-ha-hamone ati na-hare-ka\\
    place(\textsc{f}) same.\textsc{f} \textsc{ch}-be-\textsc{rep}.\textsc{f} say \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `{``}He said it's the same as here.'''
\z

\ea\label{ex:02-a234}
Ati te mitare awane?\\
    \gll ati tee mita-re awa-ne\\
    voice 2\textsc{pl}.\textsc{s} hear-\textsc{neg}.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}Didn't you hear what he said?'''
\z

\ea\label{ex:02-a235}
Yama faramone ya e winineke haro, ati nematamona me ka, ayo ati nematamonaha.\\
    \gll yama fara-mone ya ee wina-ne-ke haaro ati na-himata-mona me ka ayo ati na-himata-mona-ha\\
    place(\textsc{f}) same.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{adju} 1\textsc{in}.\textsc{s} live-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f} that\_one.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} 3\textsc{pl} \textsc{poss} older\_brother(\textsc{m}) say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `{``}He said it's just like the place we live in here,'' their older brother said.'
\z

\ea\label{ex:02-a236}
E tababa eke hinakasima yori onabana, bisa.\\
    \gll ee taba-haba ee-ke hinakasima yori o-na-habana bisa\\
    1\textsc{in}.\textsc{s} be\_together-\textsc{fut}.\textsc{f} 1\textsc{in}.\textsc{s}-\textsc{decl}.\textsc{f} 3\textsc{sg}.\textsc{poss}.younger\_sister(\textsc{f}) have\_sex\_with 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{fut}.\textsc{f} also\\
    \glt `{``}We will go together, so I can have sex with his younger sister, too.'''
\z

\ea\label{ex:02-a237}
Hiba aate oba.\\
    \gll hiba a-ate o-ba\\
    wait \textsc{dup}-ask 1\textsc{sg}.\textsc{s}-\textsc{immed}.\textsc{m}\\
    \glt `{``}Wait, I'll ask him.'''
\z

\ea\label{ex:02-a238}
Kamamata.\\
    \gll ka-ma-mata\\
    come-back-short\_time\\
    \glt `{``}Let him come back.'''
\z

\ea\label{ex:02-a239}
Faya, yana tonebona fawa tosame fawa ne, fawa ne noki hine, me nonoki hie, abariko towitamaki abariko towitamaki abariko towitamaki ne ya, faya tasi namematamona fahi.\\
    \gll faya yana to-na-hibona fawa to-kosa-me fawa ne fawa ne noki hi-ne me no-noki hi-he abariko to-ita-maki abariko to-ita-maki abariko to-ita-maki ne ya faya tasi na-ma-himata-mona fahi\\
    so start \textsc{ch}-\textsc{aux}-\textsc{int}.\textsc{m} disappear \textsc{ch}-middle-back.\textsc{m} disappear \textsc{aux}.\textsc{m} disappear \textsc{aux}.\textsc{m} wait\_for \textsc{oc}-\textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{s} \textsc{dup}-wait\_for \textsc{oc}-\textsc{aux}.\textsc{m} moon(\textsc{m}) away-sit-following.\textsc{list} moon(\textsc{m}) away-sit-following.\textsc{list} moon(\textsc{m}) away-sit-following.\textsc{list} \textsc{aux}.\textsc{nom}.\textsc{m} \textsc{adju} so emerge \textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `He was ready to go, and he disappeared into the forest. He disappeared. He waited for him. They waited for him. After three full moons, he finally appeared again.'
\z

\ea\label{ex:02-a240}
Tasi namematamonaka.\\
    \gll tasi na-ma-himata-mona-ka\\
    emerge \textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He appeared again.'
\z

\ea\label{ex:02-a241}
Tasi name, Bahi kame Bahi kame.\\
    \gll tasi na-me Bahi ka-me Bahi ka-me\\
    emerge \textsc{aux}-back.\textsc{m} Bahi(\textsc{m}) come-back.\textsc{m} Bahi(\textsc{m}) come-back.\textsc{m}\\
    \glt `He appeared again. ``Bahi has come back. Bahi has come back.'''
\z

\ea\label{ex:02-a242}
Faya kame kobo name, Oof, mate, mate yanakeni, haro.\\
    \gll faya ka-me kobo na-me oof matehe matehe yana-ke-ni haaro\\
    so come-back.\textsc{m} arrive \textsc{aux}-back.\textsc{m} hoo children(\textsc{f}) children(\textsc{f}) grow-\textsc{decl}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f} that\_one.\textsc{f}\\
    \glt `He came back, and arrived. ``Hoo, the children have grown.'''
\z

\ea\label{ex:02-a243}
Mate okihaba ama okanani, na ohite.\\
    \gll matehe o-kiha-haba ama o-ka na-hani na o-hi-te\\
    children(\textsc{f}) 1\textsc{sg}.\textsc{s}-have-\textsc{fut}.\textsc{f} \textsc{sec} 1\textsc{sg}.\textsc{s}-\textsc{decl} \textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{f} \textsc{aux} 1\textsc{sg}.\textsc{s}-\textsc{aux}.\textsc{nom}.\textsc{f}-\textsc{rp}.\textsc{n}.\textsc{f}\\
    \glt `{``}Now I finally have children.'''
\z


\ea\label{ex:02-a244}
Okatao me owakamabone, okatao me owakamabone me ka aki ati narake.\\
    \gll okatao me o-ka-ka-ma-habone okatao me o-ka-ka-ma-habone me ka aki ati na-hara-ke\\
    1\textsc{sg}.\textsc{poss}.son(\textsc{m}) 3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-\textsc{comit}-go-back-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{poss}.son(\textsc{m}) 3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-\textsc{comit}-go-back-\textsc{int}.\textsc{f} 3\textsc{pl} \textsc{poss} grandmother(\textsc{f}) say \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}My sons' grandmother said I should take them there.'''
\z

\ea\label{ex:02-a245}
Himata nima te ribeya?\\
    \gll himata ni-ma tee ri-be-ya\\
    what \textsc{aux}.\textsc{nom}-like 2\textsc{pl}.\textsc{s} \textsc{cq}.\textsc{f}-\textsc{immed}.\textsc{f}-now\\
    \glt `{``}What are you going to do?'''
\z

\ea\label{ex:02-a246}
Owa te abiyoni?\\
    \gll owa tee abiyo-ni\\
    1\textsc{sg}.\textsc{o} 2\textsc{pl}.\textsc{s} want\_the\_presence\_of-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Do you want to stay with me?'''
\z

\newpage
\ea\label{ex:02-a247}
E awara awine.\\
    \gll ee awa-ra awine\\
    1\textsc{in}.\textsc{s} see-\textsc{neg}.\textsc{f} seem.\textsc{f}\\
    \glt `{``}I don't know.'''
\z

\ea\label{ex:02-a248}
Bahi yama yofinara awineni?\\
    \gll Bahi yama yofina-ra awine-ni\\
    Bahi(\textsc{m}) place(\textsc{f}) cause\_fear-\textsc{neg}.\textsc{f} seem.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Bahi, isn't it scary?'''
\z

\ea\label{ex:02-a249}
Yama yofinakere.\\
    \gll yama yofina-ke-re\\
    place(\textsc{f}) cause\_fear-\textsc{decl}.\textsc{f}-\textsc{neg}.\textsc{f}\\
    \glt `{``}It's not scary.'''
\z

\ea\label{ex:02-a250}
Fara yama ya e nahoneke haro.\\
    \gll fara yama ya ee naho-ne-ke haaro\\
    same.\textsc{f} place(\textsc{f}) \textsc{adju} 1\textsc{in}.\textsc{s} stay-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}It's the same as where we are now.'''
\z

\ea\label{ex:02-a251}
Ha nima na awineke.\\
    \gll haa ni-ma na awine-ke\\
    \textsc{dem} \textsc{aux}.\textsc{nom}-like \textsc{aux}.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}It's just like here.'''
\z

\ea\label{ex:02-a252}
Ha nima na awineke haro.\\
    \gll haa ni-ma na awine-ke haaro\\
    \textsc{dem} \textsc{aux}.\textsc{nom}-like \textsc{aux}.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}It's just like here.'''
\z

\ea\label{ex:02-a253}
Fara fara yama ya e nahoneni.\\
    \gll fara fara yama ya ee naho-ne-ni\\
    same.\textsc{f} same.\textsc{f} place(\textsc{f}) \textsc{adju} 1\textsc{in}.\textsc{s} stay-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}It's just like where we are now.'''
\z

\ea\label{ex:02-a254}
Bahi bani me tama me awineni?\\
    \gll Bahi bani me tama me awine-ni\\
    Bahi(\textsc{m}) animal(\textsc{m}) 3\textsc{pl}.\textsc{s} be\_many.\textsc{f} 3\textsc{pl}.\textsc{s} seem.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Bahi, is there plenty of game there?'''
\z

\ea\label{ex:02-a255}
Bani me tamaramone ati ne bani fara me toha nineke, fara e wini nima.\\
    \gll bani me tama-ra-hamone ati ne bani fara me to-ha na-ne-ke fara ee wini ni-ma\\
    animal(\textsc{m}) 3\textsc{pl}.\textsc{s} be\_many-\textsc{neg}-\textsc{rep}.\textsc{f} say \textsc{aux}.\textsc{m} animal(\textsc{m}) same.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{ch}-be.\textsc{list} \textsc{aux}-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f} same.\textsc{f} 1\textsc{in}.\textsc{s} live.\textsc{nfin} \textsc{aux}.\textsc{nom}-like\\
    \glt `{``}She is saying there isn't much game, but the game there are the same as where we live.'''
\z

\ea\label{ex:02-a256}
Me yana tona me yana tonabone me, amo nama me amo nama me, Hiba okoso ka, okoso ka bani meba me fora omatibe.\\
    \gll me yana to-na me yana to-na-habone me amo na-ma me amo na-ma me hiba okoso ka okoso ka bani me-ba me fora o-mata-be\\
    3\textsc{pl}.\textsc{s} start \textsc{ch}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} start \textsc{ch}-\textsc{aux}-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{s} sleep \textsc{aux}-back.\textsc{f} 3\textsc{pl}.\textsc{s} sleep \textsc{aux}-back.\textsc{f} 3\textsc{pl}.\textsc{s} wait 1\textsc{sg}.\textsc{poss}.mother\_in\_law(\textsc{f}) \textsc{poss} 1\textsc{sg}.\textsc{poss}.mother\_in\_law(\textsc{f}) \textsc{poss} animal(\textsc{m}) 3\textsc{pl}-\textsc{fut} 3\textsc{pl}.\textsc{o} shoot\_with\_blowgun 1\textsc{sg}.\textsc{s}-short\_time-\textsc{immed}.\textsc{f}\\
    \glt `They were ready to leave. When they were about ready to leave, they slept there some more. ``Wait, I will kill some game for my mother-in-law with my blowgun.'''
\z

\ea\label{ex:02-a257}
Bani me fora oni ya bani me okoso tabasi hinaba, oka koko me tabasi hinaba mati.\\
    \gll bani me fora o-ni ya bani me okoso tabasi hi-na-haba o-ka koko me tabasi hi-na-haba mati\\
    animal(\textsc{m}) 3\textsc{pl}.\textsc{o} shoot\_with\_blowgun 1\textsc{sg}.\textsc{s}-\textsc{aux}.\textsc{nom}.\textsc{f} \textsc{adju} animal(\textsc{m}) 3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{poss}.mother\_in\_law(\textsc{f}) roast\_on\_rack \textsc{oc}-\textsc{aux}-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{poss}-\textsc{poss} father\_in\_law(\textsc{m}) 3\textsc{pl}.\textsc{o} roast\_on\_rack \textsc{oc}-\textsc{aux}-\textsc{fut}.\textsc{f} 3\textsc{pl}.\textsc{o}\\
    \glt `{``}When I kill game with my blowgun, my mother-in-law and father-in-law will roast them on a rack.'''
\z

\ea\label{ex:02-a258}
Okoto tiwakamarawibamakoni tiwati ama?\\
    \gll okoto ti-ka-ka-ma-rawa-ba-makoni ti-ati ama\\
    1\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) 2\textsc{sg}.\textsc{s}-\textsc{comit}-go-back-\textsc{fpl}-\textsc{fut}-for\_this\_reason.\textsc{f} 2\textsc{sg}.\textsc{poss}-voice be\\
    \glt `{``}Are you intending to take my daughters away?'''
\z

\ea\label{ex:02-a259}
Yama yofinara awine?\\
    \gll yama yofina-ra awine\\
    place(\textsc{f}) cause\_fear-\textsc{neg}.\textsc{f} seem.\textsc{f}\\
    \glt `{``}Isn't it scary there?'''
\z

\ea\label{ex:02-a260}
Koko yama yofinakere.\\
    \gll koko yama yofina-ke-re\\
    father\_in\_law(\textsc{m}) place(\textsc{f}) cause\_fear-\textsc{decl}.\textsc{f}-\textsc{neg}.\textsc{f}\\
    \glt `{``}Father-in-law, it's not scary.'''
\z

\ea\label{ex:02-a261}
Fara yama nima yama na ya e naoneke.\\
    \gll fara yama ni-ma yama na ya ee naho-ne-ke\\
    same.\textsc{f} place(\textsc{f}) \textsc{aux}.\textsc{nom}-like place(\textsc{f}) \textsc{aux}.\textsc{f} \textsc{adju} 1\textsc{in}.\textsc{s} stay-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}It's the same as the place where we are.'''
\z


\ea\label{ex:02-a262}
Yama fara tohineni.\\
    \gll yama fara to-ha-ne-ni\\
    place(\textsc{f}) same.\textsc{f} \textsc{ch}-be-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}The places are the same.'''
\z

\ea\label{ex:02-a263}
Fara yama tohineni.\\
    \gll fara yama to-ha-ne-ni\\
    same.\textsc{f} place(\textsc{f}) \textsc{ch}-be-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}The places are the same.'''
\z

\ea\label{ex:02-a264}
Fara yama toaha ya e winineke haro.\\
    \gll fara yama to-haha ya ee wina-ne-ke haaro\\
    same.\textsc{f} place(\textsc{f}) \textsc{ch}-be.\textsc{f} \textsc{adju} 1\textsc{in}.\textsc{s} live-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}It's the same as the place we're living.'''
\z

\ea\label{ex:02-a265}
E yana tona e tokomaro, e tokomaro e kobo tonamaro e, tokoma e tokoma e kobo naba ereke.\\
    \gll ee yana to-na ee to-ka-ma-haaro ee to-ka-ma-haaro ee kobo to-na-ma-haaro ee to-ka-ma ee to-ka-ma ee kobo na-haba ere-ke\\
    1\textsc{in}.\textsc{s} start \textsc{ch}-\textsc{aux}.\textsc{f} 1\textsc{in}.\textsc{s} away-go-back-\textsc{pfut}.\textsc{f} 1\textsc{in}.\textsc{s} away-go-back-\textsc{pfut}.\textsc{f} 1\textsc{in}.\textsc{s} arrive away-\textsc{aux}-back-\textsc{pfut}.\textsc{f} 1\textsc{in}.\textsc{s} away-go-back.\textsc{f} 1\textsc{in}.\textsc{s} away-go-back.\textsc{f} 1\textsc{in}.\textsc{s} arrive \textsc{aux}-\textsc{fut}.\textsc{f} 1\textsc{in}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}When we are ready to leave, and we go, and we arrive somewhere, we will go and go and will arrive.'''
\z

\ea\label{ex:02-a266}
E bosa ni ya yama yaboraba awineke.\\
    \gll ee bosa ni ya yama yabo-ra-haba awine-ke\\
    1\textsc{in}.\textsc{s} get\_up\_early \textsc{aux}.\textsc{nom}.\textsc{f} \textsc{adju} thing(\textsc{f}) be\_far-\textsc{neg}-\textsc{fut}.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}If we get up early, it won't be far.'''
\z

\ea\label{ex:02-a267}
Ai okoma amo onamiro ka oko yobe ware, e amo naro e bosa naro e tokomaro e tokomaro e kobo naba ereke, ati nematamonaka.\\
    \gll ahi o-ka-ma amo o-na-mi-ro ka o-ka yobe waa-re ee amo na-haaro ee bosa na-haaro ee to-ka-ma-haaro ee to-ka-ma-haaro ee kobo na-haba ere-ke ati na-himata-mona-ka\\
    then 1\textsc{sg}.\textsc{s}-go-back.\textsc{f} sleep 1\textsc{sg}.\textsc{s}-\textsc{aux}-back.\textsc{nom}.\textsc{f}-\textsc{rp}.\textsc{e}.\textsc{f} which 1\textsc{sg}.\textsc{poss}-\textsc{poss} house(\textsc{m}) stand-raised\_surface 1\textsc{in}.\textsc{s} sleep \textsc{aux}-\textsc{pfut}.\textsc{f} 1\textsc{in}.\textsc{s} get\_up\_early \textsc{aux}-\textsc{pfut}.\textsc{f} 1\textsc{in}.\textsc{s} away-go-back-\textsc{pfut}.\textsc{f} 1\textsc{in}.\textsc{s} away-go-back-\textsc{pfut}.\textsc{f} 1\textsc{in}.\textsc{s} arrive \textsc{aux}-\textsc{fut}.\textsc{f} 1\textsc{in}.\textsc{s}-\textsc{decl}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `{``}When we have slept in the house I slept in when I went before, when we have gotten up early and gone and gone, we will arrive,'' he said.'
\z

\ea\label{ex:02-a268}
Bahi ati kakasiro raba re.\\
    \gll Bahi ati ka-kasiro raba re\\
    Bahi(\textsc{m}) voice \textsc{dup}-persist a\_bit \textsc{neg}.\textsc{m}\\
    \glt `{``}Bahi really talks.'''
\z

\ea\label{ex:02-a269}
Kokori, kokori ati mite ati fawa ne, Ee koko oko, oko yibote owakamarawaboneke.\\
    \gll kokori kokori ati mite ati fawa ne ee koko o-ka o-ka yibote o-ka-ka-ma-rawa-habone-ke\\
    3\textsc{sg}.\textsc{poss}.father\_in\_law(\textsc{m}) 3\textsc{sg}.\textsc{poss}.father\_in\_law(\textsc{m}) voice hear.\textsc{m} voice stop \textsc{aux}.\textsc{m} yes father\_in\_law(\textsc{m}) 1\textsc{sg}.\textsc{poss}-\textsc{poss} 1\textsc{sg}.\textsc{poss}-\textsc{poss} wife(\textsc{f}) 1\textsc{sg}.\textsc{s}-\textsc{comit}-go-back-\textsc{fpl}-\textsc{int}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He listened to his father-in-law, and his father-in-law stopped speaking. ``Yes, Father-in-law, I'm going to take my wives there.'''
\z

\ea\label{ex:02-a270}
Tiwa tiwa owamabana oke tasa.\\
    \gll tiwa tiwa o-awa-ma-habana o-ke tasa\\
    2\textsc{sg}.\textsc{o} 2\textsc{sg}.\textsc{o} 1\textsc{sg}.\textsc{s}-see-back-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} again\\
    \glt `{``}I will come back and see you again.'''
\z

\ea\label{ex:02-a271}
Okoto tiwakamarawa, otaboro amara.\\
    \gll okoto ti-ka-ka-ma-rawa o-taboro ama-ra\\
    1\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) 2\textsc{sg}.\textsc{s}-\textsc{comit}-go-back-\textsc{fpl}.\textsc{f} 1\textsc{sg}.\textsc{poss}-place be-\textsc{neg}.\textsc{f}\\
    \glt `{``}If you take my daughters to a place that is not my village....'''
\z


\ea\label{ex:02-a272}
Otaborobisa amake.\\
    \gll o-taboro-bisa ama-ke\\
    1\textsc{sg}.\textsc{poss}-place-also be-\textsc{decl}.\textsc{f}\\
    \glt `{``}It's my village.'''
\z

\ea\label{ex:02-a273}
Otaborobisamakoni, oko yibote onakamarawine oke ahi.\\
    \gll o-taboro-bisa-makoni o-ka yibote o-na-ka-ma-rawa-ne o-ke ahi\\
    1\textsc{sg}.\textsc{poss}-place-also-for\_this\_reason.\textsc{f} 1\textsc{sg}.\textsc{poss}-\textsc{poss} wife(\textsc{f}) 1\textsc{sg}.\textsc{s}-\textsc{caus}-come-back-\textsc{fpl}-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} here\\
    \glt `{``}Because it is my village there, I have come here to get my wives.'''
\z

\ea\label{ex:02-a274}
Okobi okobi okomi, me tabori amake.\\
    \gll okobi okobi okomi me tabori ama-ke\\
    1\textsc{sg}.\textsc{poss}.father(\textsc{m}) 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) 1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) 3\textsc{pl}.\textsc{poss} place.\textsc{f} be-\textsc{decl}.\textsc{f}\\
    \glt `{``}It is my father and mother's village.'''
\z

\ea\label{ex:02-a275}
Faya mera towakamematamona fahi.\\
    \gll faya mera to-ka-ka-ma-himata-mona fahi\\
    so 3\textsc{pl}.\textsc{o} away-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `So he took them away.'
\z

\ea\label{ex:02-a276}
Me hiwakama Hine me hiwakami na.\\
    \gll me hi-ka-ka-ma hine me hi-ka-ka-mi na\\
    3\textsc{pl}.\textsc{o} \textsc{oc}-\textsc{comit}-go-back.\textsc{f} only 3\textsc{pl}.\textsc{o} \textsc{oc}-\textsc{comit}-go-back.\textsc{nom}.\textsc{f} \textsc{aux}\\
    \glt `He took them away. ``Let him take them away.'''
\z

\ea\label{ex:02-a277}
Me ka aki me hiwabone mera towakamane.\\
    \gll me ka aki me hi-awa-habone mera to-ka-ka-ma-ne\\
    3\textsc{pl} \textsc{poss} grandmother(\textsc{f}) 3\textsc{pl}.\textsc{o} \textsc{oc}-see-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{o} away-\textsc{comit}-go-back.\textsc{cont}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}He is taking them so their grandmother can see them.'''
\z

\ea\label{ex:02-a278}
Me ahabaraba me awineni.\\
    \gll me ahaba-ra-haba me awine-ni\\
    3\textsc{pl}.\textsc{s} die-\textsc{neg}-\textsc{fut}.\textsc{f} 3\textsc{pl}.\textsc{s} seem.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}They won't die.'''
\z

\ea\label{ex:02-a279}
Ota kami ya ota kama ota kamari ya ota kamarihi, oko yibote yama awarawi ya yama nofarawa, narawaba amake.\\
    \gll otaa ka-mi ya otaa ka-ma otaa ka-ma-ri ya otaa ka-ma-rihi o-ka yibote yama awa-rawi ya yama nofa-rawa na-rawa-haba ama-ke\\
    1\textsc{ex}.\textsc{s} come-back.\textsc{nom}.\textsc{f} \textsc{adju} 1\textsc{ex}.\textsc{s} come-back.\textsc{list} 1\textsc{ex}.\textsc{s} come-back-\textsc{neg}.\textsc{nom}.\textsc{f} \textsc{adju} 1\textsc{ex}.\textsc{s} come-back-\textsc{neg}.\textsc{list} 1\textsc{sg}.\textsc{poss}-\textsc{poss} wife(\textsc{f}) place(\textsc{f}) see-\textsc{fpl}.\textsc{nom}.\textsc{f} \textsc{adju} place(\textsc{f}) like-\textsc{fpl}.\textsc{list} \textsc{aux}-\textsc{fpl}-\textsc{fut}.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I don't know if we will come back soon or not. I don't know if my wives will like the place when they see it.'''
\z

\ea\label{ex:02-a280}
Faya me, faya me, tokoma me tokoma me amo tonama me tokomamina me, wabori aate nebona wabori aate nebona wabori, Okasima titihaba, titihaba tinahani.\\
    \gll faya me faya me to-ka-ma me to-ka-ma me amo to-na-ma me to-ka-ma-mina me wabori a-ate na-hibona wabori a-ate na-hibona wabori okasima ti-iti-haba ti-iti-haba ti-na-hani\\
    so 3\textsc{pl} so 3\textsc{pl}.\textsc{s} away-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} away-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} sleep away-\textsc{aux}-back.\textsc{f} 3\textsc{pl}.\textsc{s} away-go-back-morning.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) \textsc{dup}-ask \textsc{aux}-\textsc{int}.\textsc{m} 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) \textsc{dup}-ask \textsc{aux}-\textsc{int}.\textsc{m} 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) 1\textsc{sg}.\textsc{poss}.younger\_sister(\textsc{f}) 2\textsc{sg}.\textsc{s}-marry-\textsc{fut}.\textsc{f} 2\textsc{sg}.\textsc{s}-marry-\textsc{fut}.\textsc{f} 2\textsc{sg}.\textsc{s}-say-\textsc{ip}.\textsc{n}.\textsc{f}\\
    \glt `So they went. They slept on the way, and went on again in the morning. He wanted to ask his brother-in-law. ``You said you wanted to marry my younger sister.'''\footnote{Non-eyewitness tense is used in the verb \emph{tinahani} not because Bahi wasn't present when the man made his request, but because he didn't know he was going to make this request. This is a mirative use of tense (\cite[205f]{Dixon:2004xr}).}
\z

\ea\label{ex:02-a281}
Okasima manakonerawa ta tinahabone ona, aya sawibisabana ahi.\\
    \gll okasima manakone-rawa taa ti-na-habone o-na aya sawi-bisa-habana ahi\\
    1\textsc{sg}.\textsc{poss}.younger\_sister(\textsc{f}) price-\textsc{fpl} give 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{s}-say.\textsc{f} this be\_present-also-\textsc{fut}.\textsc{f} here\\
    \glt `{``}I want you to pay for my sisters to live here.'''
\z

\ea\label{ex:02-a282}
Aa sawirima nahi.\\
    \gll aa sawi-rima na-hi\\
    no be\_present-\textsc{neg}.\textsc{imp}.\textsc{f} \textsc{aux}-\textsc{imp}.\textsc{f}\\
    \glt `{``}No, I don't want her to live over there.'''
\z

\ea\label{ex:02-a283}
Fara otaboro ya sawihabanake.\\
    \gll fara o-taboro ya sawi-habana-ke\\
    same.\textsc{f} 1\textsc{sg}.\textsc{poss}-place \textsc{adju} be\_present-\textsc{fut}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I want her to live in my village.'''
\z

\newpage
\ea\label{ex:02-a284}
Fara otaboro fara otaboro ya tisawibeya.\\
    \gll fara o-taboro fara o-taboro ya ti-sawi-be-ya\\
    same.\textsc{f} 1\textsc{sg}.\textsc{poss}-place same.\textsc{f} 1\textsc{sg}.\textsc{poss}-place \textsc{adju} 2\textsc{sg}.\textsc{s}-be\_present-\textsc{immed}.\textsc{f}-now\\
    \glt `{``}You live in my village.'''
\z

\ea\label{ex:02-a285}
Yama yofina awineke.\\
    \gll yama yofina awine-ke\\
    place(\textsc{f}) cause\_fear.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}It's scary.'''
\z

\ea\label{ex:02-a286}
Yama yofinakere.\\
    \gll yama yofina-ke-re\\
    place(\textsc{f}) cause\_fear-\textsc{decl}.\textsc{f}-\textsc{neg}.\textsc{f}\\
    \glt `{``}It's not scary.'''
\z

\ea\label{ex:02-a287}
Owati timita fara yama ama haro e yayaka haro.\\
    \gll o-ati ti-mita fara yama ama haaro ee ya-yaka haaro\\
    1\textsc{sg}.\textsc{poss}-voice 2\textsc{sg}.\textsc{s}-hear.\textsc{f} same.\textsc{f} forest(\textsc{f}) be that\_one.\textsc{f} 1\textsc{in}.\textsc{poss} \textsc{dup}-walk that\_one.\textsc{f}\\
    \glt `{``}You heard me, it's the same forest, for walking.'''
\z


\ea\label{ex:02-a288}
Fara hawi tohineke haro.\\
    \gll fara hawi to-ha-ne-ke haaro\\
    same.\textsc{f} trail(\textsc{f}) \textsc{ch}-be-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}It's the same trails.'''
\z

\ea\label{ex:02-a289}
Yama tonokomisamone tina amani.\\
    \gll yama to-noko-misa-hamone ti-na ama-ni\\
    thing(\textsc{f}) away-point-up-\textsc{rep}.\textsc{f} 2\textsc{sg}.\textsc{s}-say.\textsc{f} \textsc{sec}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}You said the trail goes up.'''
\z

\ea\label{ex:02-a290}
Yama tonokomisakere.\\
    \gll yama to-noko-misa-ke-re\\
    thing(\textsc{f}) away-point-up-\textsc{decl}.\textsc{f}-\textsc{neg}.\textsc{f}\\
    \glt `{``}It doesn't go up.'''
\z

\ea\label{ex:02-a291}
Yama yaka ni amosa aware awane.\\
    \gll yama yaka ni amosa awa-re awa-ne\\
    thing(\textsc{f}) walk \textsc{aux}.\textsc{nfin} be\_good.\textsc{f} be\_seen-\textsc{neg}.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}It's good to walk there.'''
\z

\ea\label{ex:02-a292}
Tokoma tokoma tokoma, aya awa wawata na awineke ahi, totokomisama karo.\\
    \gll to-ka-ma to-ka-ma to-ka-ma aya awa wa-wata na awine-ke ahi to-to-ka-misa-ma karo\\
    away-go-back.\textsc{f} away-go-back.\textsc{f} away-go-back.\textsc{f} this tree(\textsc{f}) \textsc{dup}-stand \textsc{aux}.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f} there \textsc{dup}-away-go-up-back which\\
    \glt `{``}The trees are not high, they are like steps.'''
\z

\ea\label{ex:02-a293}
Awa nemeha towatamisineni.\\
    \gll awa nemeha to-wata-misa-ne-ni\\
    tree(\textsc{f}) be\_high.\textsc{f} away-stand-up-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}There are tall trees.'''
\z

\ea\label{ex:02-a294}
Fara toha awineke.\\
    \gll fara to-ha awine-ke\\
    same.\textsc{f} \textsc{ch}-be.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}It's the same.'''
\z


\ea\label{ex:02-a295}
Fara toha awineke.\\
    \gll fara to-ha awine-ke\\
    same.\textsc{f} \textsc{ch}-be.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}It's the same.'''
\z

\ea\label{ex:02-a296}
Wete onisamibe.\\
    \gll wete o-na-risa-ma-be\\
    return 1\textsc{sg}.\textsc{s}-\textsc{aux}-down-back-\textsc{immed}.\textsc{f}\\
    \glt `{``}I will slide back down.'''
\z

\ea\label{ex:02-a297}
Wete nisame.\\
    \gll wete na-risa-me\\
    return \textsc{aux}-down-back.\textsc{m}\\
    \glt `{``}He slides back down.'''
\z

\newpage
\ea\label{ex:02-a298}
Inamatewe tohe tere ne awane.\\
    \gll inamatewe to-he tere ne awa-ne\\
    child(\textsc{m}) \textsc{ch}-be.\textsc{m} crawl \textsc{aux}.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}He crawls because he is a baby.'''
\z

\ea\label{ex:02-a299}
E tere naba watakere.\\
    \gll ee tere na-haba wata-ke-re\\
    1\textsc{in}.\textsc{s} crawl \textsc{aux}-\textsc{fut}.\textsc{f} exist-\textsc{decl}.\textsc{f}-\textsc{neg}.\textsc{f}\\
    \glt `{``}People don't crawl.'''
\z

\ea\label{ex:02-a300}
Faya, me, tokoma me tokomemetemoneke.\\
    \gll faya me to-ka-ma me to-ka-ma-hemete-mone-ke\\
    so 3\textsc{pl}.\textsc{s} away-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} away-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They went.'
\z

\ea\label{ex:02-a301}
Wabori hinasawimake tokomake, tokomake tokomake me tokoma me tokoma me, tokoma me, kobo nemetemoneke bahi fawa nise ya.\\
    \gll wabori hi-na-sawi-make to-ka-make to-ka-make to-ka-make me to-ka-ma me to-ka-ma me to-ka-ma me kobo na-hemete-mone-ke bahi fawa na-rise ya\\
    3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) \textsc{oc}-\textsc{caus}-be\_present-following away-go-following away-go-following away-go-following 3\textsc{pl}.\textsc{s} away-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} away-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} away-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} arrive \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} sun(\textsc{m}) disappear \textsc{aux}-down.\textsc{nom}.\textsc{m} \textsc{adju}\\
    \glt `His brother-in-law made him go along. They went and went, and arrived at sundown.'
\z

\ea\label{ex:02-a302}
Bahi owa, Bahi owa tohawe kamaki, bahi toke ya me kobo na me, me kobo nemetemoneke.\\
    \gll Bahi owa Bahi owa to-ha-wee ka-maki bahi to-ke ya me kobo na me me kobo na-hemete-mone-ke\\
    Bahi(\textsc{m}) brother\_in\_law(\textsc{m}) Bahi(\textsc{m}) brother\_in\_law(\textsc{m}) \textsc{ch}-be-change.\textsc{m} come-following.\textsc{m} sun(\textsc{m}) away-go.\textsc{nom}.\textsc{m} \textsc{adju} 3\textsc{pl}.\textsc{s} arrive \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} arrive \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He had become Bahi's brother-in-law. He came. They arrived at sundown.'
\z

\ea\label{ex:02-a303}
Me kobo na me, amo na me, faya hinakasima haha ne Asima, asima.\\
    \gll me kobo na me amo na me faya hinakasima ha-haa ne asima asima\\
    3\textsc{pl}.\textsc{s} arrive \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} sleep \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} so 3\textsc{sg}.\textsc{poss}.younger\_sister(\textsc{f}) \textsc{dup}-call \textsc{aux}.\textsc{m} younger\_sister(\textsc{f}) younger\_sister(\textsc{f})\\
    \glt `They arrived. They slept. He called his younger sister. ``Sister, Sister.'''
\z

\ea\label{ex:02-a304}
Hoo.\\
    \gll hoo\\
    huh?\\
    \glt `{``}Huh?'''
\z

\ea\label{ex:02-a305}
Asima tikama mata.\\
    \gll asima ti-ka-ma mata\\
    younger\_sister(\textsc{f}) 2\textsc{sg}.\textsc{s}-come-back for\_now\\
    \glt `{``}Sister, come here a minute.'''
\z

\ea\label{ex:02-a306}
Owati timitibe.\\
    \gll o-ati ti-mita-be\\
    1\textsc{sg}.\textsc{poss}-voice 2\textsc{sg}.\textsc{s}-hear-\textsc{immed}.\textsc{f}\\
    \glt `{``}Listen to me.'''
\z

\ea\label{ex:02-a307}
Yama tama ona tiwibeya ahi.\\
    \gll yama tama o-na ti-awa-be-ya ahi\\
    thing(\textsc{f}) hold\_onto 1\textsc{sg}.\textsc{s}-\textsc{aux}.\textsc{f} 2\textsc{sg}.\textsc{s}-see-\textsc{immed}.\textsc{f}-now here\\
    \glt `{``}Look at what I have in my hand.'''
\z

\ea\label{ex:02-a308}
Aya yama okanika tama one haro.\\
    \gll aya yama o-ka-nika tama o-ne haaro\\
    this thing(\textsc{f}) 1\textsc{sg}.\textsc{s}-\textsc{comit}-buy.\textsc{f} hold\_onto 1\textsc{sg}.\textsc{s}-\textsc{cont}.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}I am holding something in my hand that I bought.'''
\z

\ea\label{ex:02-a309}
Ayo himata amariya?\\
    \gll ayo himata ama-ri-ya\\
    older\_brother(\textsc{m}) what be-\textsc{cq}.\textsc{f}-now\\
    \glt `{``}Brother, what is it?'''
\z

\ea\label{ex:02-a310}
Yama tama ona tiwa ha.\\
    \gll yama tama o-na ti-awa haa\\
    thing(\textsc{f}) hold\_onto 1\textsc{sg}.\textsc{s}-\textsc{aux}.\textsc{f} 2\textsc{sg}.\textsc{s}-see \textsc{dem}\\
    \glt `{``}Look at what I am holding in my hand.'''
\z

\ea\label{ex:02-a311}
Aabata, okoyo yokana hiwa tama ne ati ama nanoho.\\
    \gll aabata okoyo yokana hiwa tama ne ati ama na-no-ho\\
    fat 1\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) true 3\textsc{sg}.\textsc{refl}.\textsc{m} hold\_onto \textsc{aux}.\textsc{m} voice be \textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `{``}He is fat. The one my brother is holding looks just like him.'''
\z

\ea\label{ex:02-a312}
Okoyo hiwa tama ne ati ama nanoho.\\
    \gll okoyo hiwa tama ne ati ama na-no-ho\\
    1\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) 3\textsc{sg}.\textsc{refl}.\textsc{m} hold\_onto \textsc{aux}.\textsc{m} voice be \textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `{``}He looks just like my brother.'''
\z

\ea\label{ex:02-a313}
Ayo hibeke bitemete tama tine tiriya?\\
    \gll ayo hibeke bite-mete tama ti-ne ti-ri-ya\\
    older\_brother(\textsc{m}) who.\textsc{m} 3\textsc{sg}.\textsc{poss}.daughter(\textsc{f})-\textsc{fp}.\textsc{n}.\textsc{f} hold\_onto 2\textsc{sg}.\textsc{s}-\textsc{cont}.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{cq}.\textsc{f}-now\\
    \glt `{``}Brother, whose daughters' hands are you holding?'''
\z


\ea\label{ex:02-a314}
Hibeke bitemete tiyabarawine tiriya?\\
    \gll hibeke bite-mete ti-yaba-rawa-ne ti-ri-ya\\
    who.\textsc{m} 3\textsc{sg}.\textsc{poss}.daughter(\textsc{f})-\textsc{fp}.\textsc{n}.\textsc{f} 2\textsc{sg}.\textsc{s}-marry-\textsc{fpl}-\textsc{cont}.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{cq}.\textsc{f}-now\\
    \glt `{``}Whose daughters did you marry?'''
\z

\ea\label{ex:02-a315}
Ino ino owato.\\
    \gll ino ino o-wato\\
    name.\textsc{m} name.\textsc{m} 1\textsc{sg}.\textsc{s}-know\\
    \glt `{``}I don't know his name.'''
\z

\newpage
\ea\label{ex:02-a316}
Ino omiti onofare amaka.\\
    \gll ino o-mita o-nofa-re ama-ka\\
    name.\textsc{m} 1\textsc{sg}.\textsc{s}-hear 1\textsc{sg}.\textsc{s}-recently-\textsc{neg}.\textsc{m} \textsc{sec}-\textsc{decl}.\textsc{m}\\
    \glt `{``}I haven't heard his name.'''
\z

\ea\label{ex:02-a317}
Ino omiti onofare amane ama.\\
    \gll ino o-mita o-nofa-re ama-ne ama\\
    name.\textsc{m} 1\textsc{sg}.\textsc{s}-hear 1\textsc{sg}.\textsc{s}-recently-\textsc{neg}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m} \textsc{sec}\\
    \glt `{``}I haven't heard his name.'''
\z

\ea\label{ex:02-a318}
Hiba, ino omitawahebanaka.\\
    \gll hiba ino o-mita-waha-hibana-ka\\
    wait name.\textsc{m} 1\textsc{sg}.\textsc{s}-hear-change-\textsc{fut}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `{``}Wait, I will hear his name now.'''
\z

\ea\label{ex:02-a319}
Me owamaboneke.\\
    \gll me o-awa-ma-habone-ke\\
    3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-see-back-\textsc{int}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I will go and see them again.'''
\z

\ea\label{ex:02-a320}
Ayo e famaba ereni?\\
    \gll ayo ee fama-haba ere-ni\\
    older\_brother(\textsc{m}) 1\textsc{in}.\textsc{s} be\_two-\textsc{fut}.\textsc{f} 1\textsc{in}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Brother, can we go together?'''
\z

\ea\label{ex:02-a321}
Ee, e famaba ereke.\\
    \gll ee ee fama-haba ere-ke\\
    yes 1\textsc{in}.\textsc{s} be\_two-\textsc{fut}.\textsc{f} 1\textsc{in}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Yes, we will go together.'''
\z

\ea\label{ex:02-a322}
Te famibone itarika hari, sowi keye tiwa nari.\\
    \gll tee fami-bone ita-ri-ka haari sowi keye tiwa na-haari\\
    2\textsc{pl}.\textsc{s} be\_with.\textsc{nom}.\textsc{f}-\textsc{int}.\textsc{f} sit-raised\_surface-\textsc{decl}.\textsc{m} that\_one.\textsc{m} penis have\_an\_erection.\textsc{nfin} hold \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `{``}Your partner is sitting there, holding his hard penis.'''
\z

\ea\label{ex:02-a323}
Asima tika yibotebona amaka.\\
    \gll asima ti-ka yibote-bona ama-ka\\
    younger\_sister(\textsc{f}) 2\textsc{sg}.\textsc{poss}-\textsc{poss} husband(\textsc{m})-\textsc{int}.\textsc{m} be-\textsc{decl}.\textsc{m}\\
    \glt `{``}Sister, he is your future husband.'''
\z

\ea\label{ex:02-a324}
Ai tisawiba ahi?\\
    \gll ahi ti-sawi-ba ahi\\
    there 2\textsc{sg}.\textsc{s}-be\_present.\textsc{nom}-\textsc{fut} there\\
    \glt `{``}Are you going to stay there?'''
\z

\ea\label{ex:02-a325}
E awarini.\\
    \gll ee awa-ra-ni\\
    1\textsc{in}.\textsc{s} see-\textsc{neg}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}I don't know.'''
\z

\ea\label{ex:02-a326}
Osawi osawi ya osawi, osawi okoma fare taboro ya, osawima onahaba.\\
    \gll o-sawi o-sawi ya o-sawi o-sawi o-ka-ma fare taboro ya o-sawi-ma o-na-haba\\
    1\textsc{sg}.\textsc{s}-be\_present.\textsc{nom} 1\textsc{sg}.\textsc{s}-be\_present.\textsc{nom} \textsc{adju} 1\textsc{sg}.\textsc{s}-be\_present.\textsc{list} 1\textsc{sg}.\textsc{s}-be\_present.\textsc{list} 1\textsc{sg}.\textsc{s}-go-back.\textsc{list} same.\textsc{m} place.\textsc{m} \textsc{adju} 1\textsc{sg}.\textsc{s}-be\_present-back.\textsc{list} 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{fut}.\textsc{f}\\
    \glt `{``}I don't want to go to his village and stay there.'''
\z
