 The story of corn.

 ``My son, I'm hungry. I'm going to eat some corn," his father said.

 ``Father, do you want to eat corn? Okay, where is my machete?''

 He went after the corn. He cut down the corn stalks. They carried the corn away. They carried off the corn.

 So that they could eat, he went after their upper lips with a blowgun dart, which had been sharpened.

 ``First one lip is pierced, then the other lip, then their ears and their noses. First this side, then the other side.'' He pierced their lips.\footnote{Nowadays Jarawaras no longer pierce themselves, but Siko and Yowao had pierced noses, and Okomobi, who is Yowao's son, remembers seeing the pierced lips of his maternal grandmother Marikina.}

 ``Eat corn. It brought suffering on you. Eat it.''

 They roasted the corn on coals.

 ``It's not going to be good for me to eat corn [because my mouth hurts].''

 They roasted the corn on coals. They ate the corn. They ate. They had been hungry a long time.

 They ate, eating the corn.

 Then they found manioc. They pulled up the manioc. They ate sweet manioc because of him. The women ate. Their noses hurt.

 The one who became the first snake was an old woman.

 ``Grandma, come here. Eat corn.''

 ``I should eat corn. I can't eat corn. You put in my two teeth, blowgun darts. Put in two blowgun darts, to be my teeth. Put in my two teeth.''

 She didn't want to have her lips pierced. So he put in the snake's two fangs. The snake was ready to leave.

 A boy was standing in front of her. ``Grandma, give me the corn I said you should try eating.''

 She smelled it. \textit{Hss}, \textit{hss}. ``Yum. That thing has a good smell, and it is going to be eaten. It is corn, look at it. It's corn," she said.

 She grabbed the boy's hand. She quickly grabbed a switch. She whipped the boy, \textit{boto}, \textit{boto}.

 ``\textit{Aya}, \textit{aya}, \textit{aya}.''

 She whipped the boy. She didn't eat any corn. She went away and changed into something else.

 She didn't eat corn. She went away. She turned into something else, and stayed at the stream edge. They ate. After he pierced their noses, they ate. After he pierced their upper lips they ate. His son was small.

 ``Come on, I'll pierce your lip. You are going to eat corn.''

 His little son cried out. He pierced his son's lip. Also his wife's. He pierced the lips of all of them, also their noses. He whipped them so they could eat corn. He whipped all of them. He whipped his father, and he whipped his old mother, too. He whipped them, all of them, including the children.

 His switches were stuck under the thatch, all in a row.

 ``Let him whip me. I'm going to eat corn.''

 The pain in his upper lip didn't keep him from eating. The plug hurt him [i.e. the plug in his lower lip hit on the wound in his upper lip]. It was hard to stand.

 There were no crops before, and he found them. These are the crops. He found bitter manioc, sweet manioc, yams, all the crops. He ate \textit{boka} seeds, so he could find the crops. He also ate wild fruits. He ate all kinds of wild fruits.

 Then he set fire to himself [by staying in the garden when they burned it]. Afterwards, he sat in the water, and the \textit{korobo} fish cleaned him off [i.e. they ate the burned flesh]. The \textit{korobo} fish cleaned him off, and he got better.

 He came back, and made a house. That's the one who became the first Paumari, my grandfather used to say, my father reported. Paumaris today have pierced lips, too. They still have holes in their lips, their lips are pierced. He pierced their lips. He changed into a Paumari. He set fire to himself.

 They lived there a long time. The house was full of people. They ate. They had starch to eat with their \textit{mato} fruits. He changed himself so he could get crops. He changed into something else. He was the one who first found the sweet manioc that we eat. He caused us to suffer. Look at our lips. My grandmother said it hurt. ``Pierce your lips [i.e. you try it, and you will see how it hurts]," she said.

 They broke off the ears of corn when its leaves were dry. The pole for holding the ears of corn was laid on top of the rafters. He draped the corn on the pole inside the house.

 ``You eat the bad ears of corn. You toast the kernels from the bad ears in the pan. Eat the bad ears of corn.''

 His grandmothers toasted the kernels from the bad ears of corn in the pan.

 ``She is naughty. I thought she was going to eat corn. I put in her teeth, and she left. The snake left when no one was looking, after I put in her teeth.'' He complained about her.

 She never came back. Her grandson complained about her.

 They lived there. They ate. They grated manioc. They dug up yams when the leaves dried, because he told them to.

 ``When you have dug up the yams, and have dug up the \textit{haso} taro, the \textit{yobo} taro, and the \textit{nakafira} taro, eat other things. You keep the seeds. Put the seeds in a container, to plant them.''

 They dug the crops up. They ate them. They ate. The things they dug up were in a pile, the yams. He dug up yams, that had many tubers.

 He gave names to \textit{kero} yams, \textit{wao} yams, \textit{anabia} yams. ``These are \textit{anabia} yams. These are \textit{wao} yams. These are \textit{kero} yams.'' He gave names to the yams. ``These are \textit{haso} taros. There are the \textit{yobo} taros.''

 The things they carried off were in piles, and he gave them names.

 ``What is that?''

 ``That is a peach palm. It's a peach palm. Those are \textit{biri} plantains there. Those are \textit{biri} plantains, \textit{howe} plantains. Those are \textit{kofiti} bananas. Those are \textit{sawa} bananas. Those are \textit{mawara} bananas. Those are \textit{kofiti} bananas.'' He was giving names to the crops. ``\textit{Kofiti} bananas, \textit{ananika} bananas, that are just this high. Those are \textit{ananika} bananas. Those are \textit{ananika} banana plants, they are not high.''

 He was naming the bananas. ``Where should we stop [picking bananas]?''

 ``You pick them. You eat them. You pick them.''

 They picked the bananas. They carried the bananas off.

 ``They will be eaten, all of them. They have a good smell. Those bananas smell good. They are \textit{kofiti} bananas. Those are \textit{nasa} bananas. They are \textit{nasa} bananas.'' They picked the bananas. They carried the bananas off. They ate the bananas, their bananas.

 They all grated manioc, too, because he told them to.

 ``You grate manioc, it will be eaten with the game meat. You wait for the others, who went hunting.''

 He burned himself up, the one who wouldn't stay there, the one who became the Paumaris. He became the Paumaris, and he didn't stay. He went away. He went away, the one who became the Paumaris. His sons, too.
