\ea\label{ex:03-a1}
Kimi, kimi, kimi me, kimi, Okatao, okatao ofimini.\\
\gll kimi kimi kimi me kimi okatao okatao o-fimi-ni\\
     corn(\textsc{m}) corn(\textsc{m}) corn(\textsc{m}) 3\textsc{pl} corn(\textsc{m}) 1\textsc{sg}.\textsc{poss}.son(\textsc{m}) 1\textsc{sg}.\textsc{poss}.son(\textsc{m}) 1\textsc{sg}.\textsc{s}-be\_hungry-\textsc{bkg}.\textsc{f}\\
\glt `The story of corn. ``My son, I'm hungry.'''
\z

\newpage
\ea\label{ex:03-a2}
Kimi, kimi okobibeya, bati ati nematamonaka, bati.\\
\gll kimi kimi o-kaba-be-ya bati ati na-himata-mona-ka bati\\
     corn(\textsc{m}) corn(\textsc{m}) 1\textsc{sg}.\textsc{s}-eat-\textsc{immed}.\textsc{f}-now 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} 3\textsc{sg}.\textsc{poss}.father(\textsc{m})\\
\glt `{``}I'm going to eat some corn,'' his father said.'
\z

\ea\label{ex:03-a3}
Bati ati ne, Abi kimi kabinima tini?\\
\gll bati ati ne abi kimi kaba-nima ti-ni\\
     3\textsc{sg}.\textsc{poss}.father(\textsc{m}) say \textsc{aux}.\textsc{m} father(\textsc{m}) corn(\textsc{m}) eat-want 2\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
\glt `{``}Father, do you want to eat corn?'''
\z

\ea\label{ex:03-a4}
Ee, oka yimawara?\\
\gll ee o-ka yimawa-ra\\
     yes 1\textsc{sg}.\textsc{poss}-\textsc{poss} knife(\textsc{f})-\textsc{npq}\\
\glt `{``}Okay, where is my machete?'''
\z

\ea\label{ex:03-a5}
Kimi, kimi wati kane kimi ti ne kimi ti ne kimi me iso nemetemone ahi.\\
\gll kimi kimi wati ka-ne kimi tii ne kimi tii ne kimi me iso na-hemete-mone ahi\\
     corn(\textsc{m}) corn(\textsc{m}) go\_after \textsc{comit}-\textsc{aux}.\textsc{m} corn(\textsc{m}) cut\_through \textsc{aux}.\textsc{m} corn(\textsc{m}) cut\_through \textsc{aux}.\textsc{m} corn(\textsc{m}) 3\textsc{pl}.\textsc{s} carry\_off \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
\glt `He went after the corn. He cut down the corn stalks. They carried the corn away.'
\z

\ea\label{ex:03-a6}
Kimi me iso na me tafabone mera, me tafabone me boni wati kanematamona ahi, sare ya, sare siri na ya ahi.\\
\gll kimi me iso na me tafa-habone mera me tafa-habone me boni wati ka-na-himata-mona ahi sarehe ya sarehe siri na ya ahi\\
     corn(\textsc{m}) 3\textsc{pl}.\textsc{s} carry\_off \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} eat-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} eat-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{poss} upper\_lip.\textsc{f} plan\_against \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then blowgun\_dart(\textsc{f}) \textsc{adju} blowgun\_dart(\textsc{f}) be\_scraped \textsc{aux}.\textsc{f} \textsc{adju} then\\
\glt `They carried off the corn. So that they could eat, he went after their upper lips with a blowgun dart, which had been sharpened.'
\z

\newpage
\ea\label{ex:03-a7}
Aya moto, aya moto, me narabi moto, me witi haro haro, ahi.\\
\gll aya moto aya moto me narabi moto me witi haaro haaro ahi\\
     this pierce.\textsc{list} this pierce.\textsc{list} 3\textsc{pl}.\textsc{poss} ear.\textsc{f} pierce.\textsc{list} 3\textsc{pl}.\textsc{poss} nose that\_one.\textsc{f} that\_one.\textsc{f} then\\
\glt `{``}First one lip is pierced, then the other lip, then their ears and their noses.'''
\z

\ea\label{ex:03-a8}
Ha moto, ha ya moto, ahi.\\
\gll haa moto haa ya moto ahi\\
     \textsc{dem} pierce.\textsc{list} \textsc{dem} \textsc{adju} pierce.\textsc{list} then\\
\glt `{``}First this side, then the other side.'''
\z

\ea\label{ex:03-a9}
Faya me boni moto ne, Kimi te kabahi.\\
\gll faya me boni moto ne kimi tee kaba-hi\\
     so 3\textsc{pl}.\textsc{poss} upper\_lip.\textsc{f} pierce \textsc{aux}.\textsc{m} corn(\textsc{m}) 2\textsc{pl}.\textsc{s} eat-\textsc{imp}.\textsc{f}\\
\glt `He pierced their lips. ``Eat corn.'''
\z

\ea\label{ex:03-a10}
Tera yoko amane.\\
\gll tera yoko ama-ne\\
     2\textsc{pl}.\textsc{o} bring\_misfortune\_on.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m}\\
\glt `{``}It brought suffering on you.'''
\z

\ea\label{ex:03-a11}
Te kabaho.\\
\gll tee kaba-ho\\
     2\textsc{pl}.\textsc{s} eat-\textsc{imp}.\textsc{m}\\
\glt `{``}Eat it.'''
\z

\ea\label{ex:03-a12}
Faya me, kimi me yokehemetemone ahi.\\
\gll faya me kimi me yoka-hemete-mone ahi\\
     so 3\textsc{pl} corn(\textsc{m}) 3\textsc{pl}.\textsc{s} roast\_on\_coals-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
\glt `They roasted the corn on coals.'
\z

\ea\label{ex:03-a13}
Me, oko kimi kabi amosareba ama onihi.\\
\gll me o-ka kimi kabi amosa-ra-hiba ama o-ni-hi\\
     3\textsc{pl} 1\textsc{sg}.\textsc{poss}-\textsc{poss} corn(\textsc{m}) be\_eaten.\textsc{nfin} be\_good-\textsc{neg}-\textsc{fut}.\textsc{m} be 1\textsc{sg}.\textsc{poss}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
\glt `{``}It's not going to be good for me to eat corn.'''
\z

\ea\label{ex:03-a14}
Kimi me yoka me kimi me yokemetemoneke.\\
\gll kimi me yoka me kimi me yoka-hemete-mone-ke\\
     corn(\textsc{m}) 3\textsc{pl}.\textsc{s} roast\_on\_coals.\textsc{f} 3\textsc{pl}.\textsc{s} corn(\textsc{m}) 3\textsc{pl}.\textsc{s} roast\_on\_coals-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `They roasted the corn on coals.'
\z

\ea\label{ex:03-a15}
Kimi me kabehemetemone.\\
\gll kimi me kaba-hemete-mone\\
     corn(\textsc{m}) 3\textsc{pl}.\textsc{s} eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
\glt `They ate the corn.'
\z

\ea\label{ex:03-a16}
Me tafehemetemoneke ahi.\\
\gll me tafa-hemete-mone-ke ahi\\
     3\textsc{pl}.\textsc{s} eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
\glt `They ate.'
\z

\ea\label{ex:03-a17}
Me fifimi toa rama ahi.\\
\gll me fi-fimi to-ha rama ahi\\
     3\textsc{pl}.\textsc{s} \textsc{dup}-be\_hungry \textsc{ch}-\textsc{aux}.\textsc{f} unusual then\\
\glt `They had been hungry a long time.'
\z

\ea\label{ex:03-a18}
Me tafa me kimi me kaba, me tafehemetemoneke ahi.\\
\gll me tafa me kimi me kaba me tafa-hemete-mone-ke ahi\\
     3\textsc{pl}.\textsc{s} eat.\textsc{f} 3\textsc{pl}.\textsc{s} corn(\textsc{m}) 3\textsc{pl}.\textsc{s} eat.\textsc{f} 3\textsc{pl}.\textsc{s} eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
\glt `They ate, eating the corn.'
\z

\ea\label{ex:03-a19}
Fowa, fowa me wasiwa fowa me bore nawaha ehene fowa kabe, fowa kabe me kabehemetemoneni.\\
\gll fowa fowa me wasi-waha fowa me bore na-waha ehene fowa\_kabe fowa\_kabe me kaba-hemete-mone-ni\\
     manioc(\textsc{m}) manioc(\textsc{m}) 3\textsc{pl}.\textsc{s} find-change.\textsc{f} manioc(\textsc{m}) 3\textsc{pl}.\textsc{s} pull\_out \textsc{aux}-change.\textsc{f} because\_of.\textsc{m} sweet\_manioc(\textsc{m}) sweet\_manioc(\textsc{m}) 3\textsc{pl}.\textsc{s} eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
\glt `Then they found manioc. They pulled up the manioc. They ate sweet manioc because of him.'
\z

\newpage
\ea\label{ex:03-a20}
Faya me fanawi tafehemetemone, witi faori kawaharo.\\
\gll faya me fanawi tafa-hemete-mone witi fao-ri ka-waha-haaro\\
     so 3\textsc{pl}.\textsc{s} woman eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} nose hurt-\textsc{distr} \textsc{comit}-change-\textsc{ip}.\textsc{e}.\textsc{f}\\
\glt `The women ate. Their noses hurt.'
\z

\ea\label{ex:03-a21}
Makateba, kirabote, kirabote, kirabote, Aki, aki hiya, tikama.\\
\gll maka-tee-ba kirabote kirabote kirabote aki aki hiya ti-ka-ma\\
     snake(\textsc{f})-\textsc{hab}-\textsc{fut} old\_woman(\textsc{f}) old\_woman(\textsc{f}) old\_woman(\textsc{f}) grandmother(\textsc{f}) grandmother(\textsc{f}) come\_on 2\textsc{sg}.\textsc{s}-come-back\\
\glt `The one who became the first snake was an old woman. ``Grandma, come here.'''
\z

\ea\label{ex:03-a22}
Kimi tikabibeya.\\
\gll kimi ti-kaba-be-ya\\
     corn(\textsc{m}) 2\textsc{sg}.\textsc{s}-eat-\textsc{immed}.\textsc{f}-now\\
\glt `{``}Eat corn.'''
\z

\ea\label{ex:03-a23}
Faya kimi okoba ahi.\\
\gll faya kimi o-kaba ahi\\
     so corn(\textsc{m}) 1\textsc{sg}.\textsc{s}-eat then\\
\glt `{``}I should eat corn.'''
\z

\ea\label{ex:03-a24}
Kimi okobene, owiniba, owiniba yoro tinahi ahi, sare ahi.\\
\gll kimi o-kaba-hene o-ini-ba o-ini-ba yoro ti-na-hi ahi sarehe ahi\\
     corn(\textsc{m}) 1\textsc{sg}.\textsc{s}-eat-\textsc{irr}.\textsc{f} 1\textsc{sg}.\textsc{poss}-tooth.\textsc{f}-\textsc{fut} 1\textsc{sg}.\textsc{poss}-tooth.\textsc{f}-\textsc{fut} put\_in\_place 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{f} then blowgun\_dart(\textsc{f}) then\\
\glt `{``}I can't eat corn. You put in my two teeth, blowgun darts.'''
\z

\ea\label{ex:03-a25}
Sare yoro tinahi ahi, owinibana.\\
\gll sarehe yoro ti-na-hi ahi o-ini-bana\\
     blowgun\_dart(\textsc{f}) put\_in\_place 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{f} then 1\textsc{sg}.\textsc{poss}-tooth.\textsc{f}-\textsc{fut}\\
\glt `{``}Put in two blowgun darts, to be my teeth.'''
\z

\ea\label{ex:03-a26}
Owiniba yoro tina.\\
\gll o-ini-ba yoro ti-na\\
     1\textsc{sg}.\textsc{poss}-tooth.\textsc{f}-\textsc{fut} put\_in\_place 2\textsc{sg}.\textsc{s}-\textsc{aux}\\
\glt `{``}Put in my two teeth.'''
\z

\ea\label{ex:03-a27}
Boni, boni kayawa, ini ta yoro hinemetemoneke ahi, ini, makatebana, maka.\\
\gll boni boni ka-yawa ini taa yoro hi-na-hemete-mone-ke ahi ini maka-tee-bana maka\\
     upper\_lip.\textsc{f} upper\_lip.\textsc{f} \textsc{comit}-be\_upset.\textsc{f} tooth.\textsc{f} contrast put\_in\_place \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then tooth.\textsc{f} snake(\textsc{f})-\textsc{hab}-\textsc{fut} snake(\textsc{f})\\
\glt `She didn't want to have her lips pierced. So he put in the snake's two fangs.'
\z

\ea\label{ex:03-a28}
Maka, maka yana tonemetemoneni.\\
\gll maka maka yana to-na-hemete-mone-ni\\
     snake(\textsc{f}) snake(\textsc{f}) start \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
\glt `The snake was ready to leave.'
\z

\ea\label{ex:03-a29}
Inamate wematamona ahi, he nokosi ya.\\
\gll inamate waa-himata-mona ahi hee nokosi ya\\
     child(\textsc{m}) stand-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} there 3\textsc{sg}.\textsc{poss} in\_front\_of \textsc{adju}\\
\glt `A boy was standing in front of her.'
\z

\ea\label{ex:03-a30}
Aki kimi kakaba tinebona one.\\
\gll aki kimi ka-kaba ti-na-hibona o-ne\\
     grandmother(\textsc{f}) corn(\textsc{m}) \textsc{dup}-eat 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{int}.\textsc{m} 1\textsc{sg}.\textsc{s}-say.\textsc{m}\\
\glt `{``}Grandma, you should try eating some corn.'''
\z

\ea\label{ex:03-a31}
Owa kakakawa tinaho.\\
\gll owa ka-ka-kawa ti-na-ho\\
     1\textsc{sg}.\textsc{o} \textsc{dup}-\textsc{comit}-give\_something\_to 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{m}\\
\glt `{``}Give it to me.'''
\z

\ea\label{ex:03-a32}
Mao hisi hinematamona, hss, hss.\\
\gll maho hisi hi-na-himata-mona hss hss\\
     smell.\textsc{m} sniff \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} (sound\_of\_sniffing) (sound\_of\_sniffing)\\
\glt `She smelled it. \textit{Hss}, \textit{hss}.'
\z

\ea\label{ex:03-a33}
Nyai, yama mai amosa kabateraweba tiwahi, kimi.\\
\gll nyai yama mahi amosa kaba-tee-rawa-hiba ti-awa-hi kimi\\
     yum thing(\textsc{f}) smell.\textsc{f} be\_good.\textsc{f} be\_eaten-\textsc{hab}-\textsc{fpl}-\textsc{fut}.\textsc{m} 2\textsc{sg}.\textsc{s}-see-\textsc{imp}.\textsc{f} corn(\textsc{m})\\
\glt `{``}Yum. That thing has a good smell, and it is going to be eaten. It is corn, look at it.'''
\z

\ea\label{ex:03-a34}
Kimi, kimi amane, kimi ati na inamate, inamate mano wara tona, kokosi, kokosi itikabote na inamate kosi nemetemoneke ahi, boto, boto.\\
\gll kimi kimi ama-ne kimi ati na inamate inamate mano wara to-na kokosi kokosi iti-kabote na inamate kosi na-hemete-mone-ke ahi boto boto\\
     corn(\textsc{m}) corn(\textsc{m}) be-\textsc{bkg}.\textsc{m} corn(\textsc{m}) say \textsc{aux}.\textsc{f} child(\textsc{m}) child(\textsc{m}) arm.\textsc{m} grab \textsc{ch}-\textsc{aux}.\textsc{f} switch(\textsc{f}) switch(\textsc{f}) pick\_up-quickly \textsc{aux}.\textsc{f} child(\textsc{m}) flail \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then (sound\_of\_whipping) (sound\_of\_whipping)\\
\glt `{``}It's corn,'' she said. She grabbed the boy's hand. She quickly grabbed a switch. She whipped the boy, \textit{boto}, \textit{boto}.'
\z

\ea\label{ex:03-a35}
Aya, aya, aya, aya, aya inamate kosi na, kimi kabarabana, tokabana, toiyatebana.\\
\gll aya aya aya aya aya inamate kosi na kimi kaba-ra-habana to-ka-habana to-hiya-tee-habana\\
     (expression\_of\_pain) (expression\_of\_pain) (expression\_of\_pain) (expression\_of\_pain) (expression\_of\_pain) child(\textsc{m}) flail \textsc{aux}.\textsc{f} corn(\textsc{m}) eat-\textsc{neg}-\textsc{fut}.\textsc{f} away-go-\textsc{fut}.\textsc{f} \textsc{ch}-be\_different-\textsc{hab}-\textsc{fut}.\textsc{f}\\
\glt `{``}\textit{Aya}, \textit{aya}, \textit{aya}.'' She whipped the boy. She didn't eat any corn. She went away and changed into something else.'
\z

\ea\label{ex:03-a36}
Kimi kabaremetemoneke.\\
\gll kimi kaba-ra-hemete-mone-ke\\
     corn(\textsc{m}) eat-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `She didn't eat corn.'
\z


\ea\label{ex:03-a37}
Tokemetemoneke.\\
\gll to-ka-hemete-mone-ke\\
     away-go-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `She went away.'
\z

\ea\label{ex:03-a38}
Faya tohiyemetemone itineti, fa, fa witi ya.\\
\gll faya to-hiya-hemete-mone ita-ne-ti faha faha witi ya\\
     so \textsc{ch}-be\_different-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} stay-\textsc{cont}.\textsc{f}-\textsc{hab}.\textsc{nom} stream(\textsc{f}) stream(\textsc{f}) edge \textsc{adju}\\
\glt `She turned into something else, and stayed at the stream edge.'
\z

\ea\label{ex:03-a39}
Faya, faya me tafa me me witi moto hina me tafemetemoneke.\\
\gll faya faya me tafa me me witi moto hi-na me tafa-hemete-mone-ke\\
     so so 3\textsc{pl}.\textsc{s} eat.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{poss} nose pierce \textsc{oc}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `They ate. After he pierced their noses, they ate.'
\z

\ea\label{ex:03-a40}
Me tafa me, me boni, me boni moto hina me tafa me, faya biti iisi kabote ne Hiya, tibono moto obe.\\
\gll me tafa me me boni me boni moto hi-na me tafa me faya biti i-isi kabote ne hiya ti-bono moto o-be\\
     3\textsc{pl}.\textsc{s} eat.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{poss} upper\_lip.\textsc{f} 3\textsc{pl}.\textsc{poss} upper\_lip.\textsc{f} pierce \textsc{oc}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} eat.\textsc{f} 3\textsc{pl}.\textsc{s} so 3\textsc{sg}.\textsc{poss}.son(\textsc{m}) \textsc{dup}-be\_small quickly \textsc{aux}.\textsc{m} come\_on 2\textsc{sg}.\textsc{poss}-upper\_lip pierce 1\textsc{sg}.\textsc{s}-\textsc{immed}.\textsc{f}\\
\glt `They ate. After he pierced their upper lips they ate. His son was small. ``Come on, I'll pierce your lip.'''
\z

\ea\label{ex:03-a41}
Kimi tikabibe.\\
\gll kimi ti-kaba-be\\
     corn(\textsc{m}) 2\textsc{sg}.\textsc{s}-eat-\textsc{immed}.\textsc{f}\\
\glt `{``}You are going to eat corn.'''
\z

\ea\label{ex:03-a42}
Biti inamate nafire kiya kanematamonaka.\\
\gll biti inamate nafi-re kiya ka-na-himata-mona-ka\\
     3\textsc{sg}.\textsc{poss}.son(\textsc{m}) child(\textsc{m}) be\_big-\textsc{neg}.\textsc{m} yell \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `His little son cried out.'
\z

\newpage
\ea\label{ex:03-a43}
Biti bono moto ne, fati, fati me nafi, me nafi boni moto nematamona, me witi.\\
\gll biti bono moto ne fati fati me nafi me nafi boni moto na-himata-mona me witi\\
     3\textsc{sg}.\textsc{poss}.son(\textsc{m}) upper\_lip.\textsc{m} pierce \textsc{aux}.\textsc{m} 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) 3\textsc{pl}.\textsc{poss} all 3\textsc{pl}.\textsc{poss} all upper\_lip.\textsc{f} pierce \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} 3\textsc{pl}.\textsc{poss} nose\\
\glt `He pierced his son's lip. Also his wife. He pierced the lips of all of them, also their noses.''
\z

\ea\label{ex:03-a44}
Kimi me, kimi me kababone mera kosi nematamonaka ahi.\\
\gll kimi me kimi me kaba-habone mera kosi na-himata-mona-ka ahi\\
     corn(\textsc{m}) 3\textsc{pl} corn(\textsc{m}) 3\textsc{pl}.\textsc{s} eat-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{o} flail \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `He whipped them so they could eat corn.'
\z

\ea\label{ex:03-a45}
Me nafi kosi nemata.\\
\gll me nafi kosi na-himata\\
     3\textsc{pl}.\textsc{poss} all flail \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}\\
\glt `He whipped all of them.'
\z

\ea\label{ex:03-a46}
Bati kosi, mati, mati bote kosi, mera kosi ne, mera kosi nematamonaka, me nafi, matehe.\\
\gll bati kosi mati mati bote kosi mera kosi ne mera kosi na-himata-mona-ka me nafi matehe\\
     3\textsc{sg}.\textsc{poss}.father(\textsc{m}) flail.\textsc{list} 3\textsc{sg}.\textsc{poss}.mother(\textsc{f}) 3\textsc{sg}.\textsc{poss}.mother(\textsc{f}) old flail.\textsc{list} 3\textsc{pl}.\textsc{o} flail \textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{o} flail \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} 3\textsc{pl}.\textsc{poss} all children(\textsc{f})\\
\glt `He whipped his father, and he whipped his old mother, too. He whipped them. He whipped them, all of them, including the children.'
\z

\ea\label{ex:03-a47}
Hinaka kokosi basasa ni tokomemetemone ahi.\\
\gll hinaka kokosi basa-sa ni to-ka-ma-hemete-mone ahi\\
     3\textsc{sg}.\textsc{poss} switch(\textsc{f}) protrude-\textsc{dup} \textsc{aux}.\textsc{nfin} away-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
\glt `His switches were stuck under the thatch, all in a row.'
\z

\ea\label{ex:03-a48}
Hoo owa kosi neba ama narani.\\
\gll hoo owa kosi na-hiba ama na-rani\\
     whew 1\textsc{sg}.\textsc{o} flail \textsc{aux}-\textsc{fut}.\textsc{m} \textsc{sec} \textsc{aux}.\textsc{m}-?\\
\glt `{``}Let him whip me.'''
\z

\ea\label{ex:03-a49}
Kimi okobaba.\\
\gll kimi o-kaba-ba\\
     corn(\textsc{m}) 1\textsc{sg}.\textsc{s}-eat-\textsc{immed}.\textsc{m}\\
\glt `{``}I'm going to eat corn.'''
\z

\ea\label{ex:03-a50}
Kimi okobibe.\\
\gll kimi o-kaba-be\\
     corn(\textsc{m}) 1\textsc{sg}.\textsc{s}-eat-\textsc{immed}.\textsc{f}\\
\glt `{``}I'm going to eat corn.'''
\z

\ea\label{ex:03-a51}
Bono komene hi kare ahi, bono, bono ewene towatowakosa ne ka bono komebana ahi.\\
\gll bono komene hii ka-re ahi bono bono ewene towa-to-waa-kosa ne ka bono koma-hibana ahi\\
     upper\_lip.\textsc{m} pain.\textsc{m} be\_affected\_by \textsc{comit}-\textsc{neg}.\textsc{m} then upper\_lip.\textsc{m} upper\_lip.\textsc{m} wood.\textsc{m} \textsc{dup}-\textsc{ch}-stand-middle \textsc{aux}.\textsc{nom}.\textsc{m} because upper\_lip.\textsc{m} hurt-\textsc{fut}.\textsc{m} then\\
\glt `The pain in his upper lip didn't keep him from eating. The plug hurt him.'
\z

\ea\label{ex:03-a52}
Nawahemetemone amani.\\
\gll nawaha-hemete-mone ama-ni\\
     take\_a\_long\_time-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec}-\textsc{bkg}.\textsc{f}\\
\glt `It was hard to stand.'
\z

\ea\label{ex:03-a53}
Yama nawahemetemone amani.\\
\gll yama nawaha-hemete-mone ama-ni\\
     thing(\textsc{f}) take\_a\_long\_time-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec}-\textsc{bkg}.\textsc{f}\\
\glt `It was hard to stand.'
\z

\ea\label{ex:03-a54}
Yamata watari ya yamata wasihimatamonaka.\\
\gll yamata wata-ri ya yamata wasi-himata-mona-ka\\
     crops(\textsc{f}) exist-\textsc{neg}.\textsc{nom}.\textsc{f} \textsc{adju} crops(\textsc{f}) find-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `There were no crops, and he found them.'
\z

\ea\label{ex:03-a55}
Yamata amake.\\
\gll yamata ama-ke\\
     crops(\textsc{f}) be-\textsc{decl}.\textsc{f}\\
\glt `There are crops.'
\z

\ea\label{ex:03-a56}
Fowa, fowa kabe, biha, yama nafi wasihimatamonaka.\\
\gll fowa fowa\_kabe biha yama nafi wasi-himata-mona-ka\\
     manioc(\textsc{m}) sweet\_manioc(\textsc{m}) yam(\textsc{f}) thing(\textsc{f}) all find-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `He found bitter manioc, sweet manioc, yams, all the crops.'
\z

\ea\label{ex:03-a57}
Yama wasihiba, yama wasihiba boka noko yome, awa boni, awa boni yome, awa boni nafi yoyome toematamonaha, hiwa sari kanebana.\\
\gll yama wasi-hiba yama wasi-hiba boka noko yome awa boni awa boni yome awa boni nafi yo-yome to-ha-himata-mona-ha hiwa sari ka-na-hibana\\
     thing(\textsc{f}) find-\textsc{fut}.\textsc{m} thing(\textsc{f}) find-\textsc{fut}.\textsc{m} tree.sp(\textsc{m}) seed.\textsc{m} swallow.\textsc{list} tree(\textsc{f}) fruit.\textsc{f} tree(\textsc{f}) fruit.\textsc{f} swallow.\textsc{list} tree(\textsc{f}) fruit.\textsc{f} all \textsc{dup}-swallow \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup} 3\textsc{sg}.\textsc{refl}.\textsc{m} burn \textsc{comit}-\textsc{aux}-\textsc{fut}.\textsc{m}\\
\glt `He ate \textit{boka} seeds, so he could find the crops. He also ate wild fruits. He ate all kinds of wild fruits. Then he set fire to himself.'
\z

\ea\label{ex:03-a58}
Faya hiwa sari kanematamona fahi.\\
\gll faya hiwa sari ka-na-himata-mona fahi\\
     so 3\textsc{sg}.\textsc{refl}.\textsc{m} burn \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
\glt `He set fire to himself.'
\z

\ea\label{ex:03-a59}
Faya kakafowa tohe towake kakafowa toahi me nanamosa toemetemone me ama, korobo mati.\\
\gll faya ka-ka-fowa to-he to-ka-ke ka-ka-fowa to-hahi me na-na-amosa to-ha-hemete-mone me ama korobo mati\\
     so \textsc{dup}-\textsc{comit}-be\_in\_water \textsc{ch}-\textsc{aux}.\textsc{m} away-\textsc{comit}-go.\textsc{m} \textsc{dup}-\textsc{comit}-be\_in\_water \textsc{ch}-\textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{s} \textsc{dup}-\textsc{caus}-be\_good \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{sec} fish.sp(\textsc{m}) 3\textsc{pl}\\
\glt `He sat in the water, and the \textit{korobo} fish cleaned him off.'\footnote{The relative clause \emph{kakafowa toahi} shows that relative clauses in Jarawara are internally headed (\cite[525]{Dixon:2004xr}; \cite[123]{Vogel:2022du}). This relative clause is the O of the main clause of the sentence, and it has no overt noun head.}
\z

\ea\label{ex:03-a60}
Korobo me hinamosahi, korobo me nanamosa hie tafime, kame yobe yayama toematamonaha.\\
\gll korobo me hi-na-amosahi korobo me na-na-amosa hi-he tafi-me ka-me yobe ya-yama to-ha-himata-mona-ha\\
     fish.sp(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{caus}-be\_good.\textsc{m} fish.sp(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{dup}-\textsc{caus}-be\_good \textsc{oc}-\textsc{aux}.\textsc{m} get\_well-back.\textsc{m} come-back.\textsc{m} house(\textsc{m}) \textsc{dup}-make \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
\glt `The \textit{korobo} fish cleaned him off, and he got better. He came back, and made a house.'
\z

\ea\label{ex:03-a61}
Efe, efe, efematamona wakari me tohineke, wakari mati, wakari metebana, wakari metebana, wakaritebana okiti aati toematamona okobi aati toemari amane.\\
\gll efe efe efe-mata-mona wakari me to-ha-ne-ke wakari mati wakari me-tee-bana wakari me-tee-bana wakari-tee-bana okiti a-ati to-ha-himata-mona okobi a-ati to-ha-himari ama-ne\\
     this.\textsc{m} this.\textsc{m} this.\textsc{m}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} Paumari(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{ch}-be-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f} Paumari(\textsc{m}) 3\textsc{pl} Paumari(\textsc{m}) 3\textsc{pl}-\textsc{hab}-\textsc{fut} Paumari(\textsc{m}) 3\textsc{pl}-\textsc{hab}-\textsc{fut} Paumari(\textsc{m})-\textsc{hab}-\textsc{fut} 1\textsc{sg}.\textsc{poss}.grandfather(\textsc{m}) \textsc{dup}-say \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) \textsc{dup}-say \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{e}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m}\\
\glt `That's the one who became the first Paumari, my grandfather used to say, my father reported.''
\z

\ea\label{ex:03-a62}
Wakariteba afa me boni moto te me ama faro.\\
\gll wakari-tee-ba afa me boni moto tee me ama faaro\\
     Paumari(\textsc{m})-\textsc{hab}-\textsc{fut} this.\textsc{f} 3\textsc{pl}.\textsc{poss} upper\_lip.\textsc{f} be\_pierced \textsc{hab} 3\textsc{pl}.\textsc{poss} \textsc{sec} that.\textsc{f}\\
\glt `Paumaris today have pierced lips, too.'
\z

\ea\label{ex:03-a63}
Me boni hoti si tote me amake, me boni moto na mati.\\
\gll me boni hoti sii to-tee me ama-ke me boni moto na mati\\
     3\textsc{pl}.\textsc{poss} upper\_lip.\textsc{f} hole.\textsc{f} stand \textsc{ch}-\textsc{hab} 3\textsc{pl}.\textsc{poss} \textsc{sec}-\textsc{decl}.\textsc{f} 3\textsc{pl}.\textsc{poss} upper\_lip.\textsc{f} be\_pierced \textsc{aux}.\textsc{f} 3\textsc{pl}\\
\glt `They still have holes in their lips, their lips are pierced.'
\z

\ea\label{ex:03-a64}
Me boni moto nemata.\\
\gll me boni moto na-himata\\
     3\textsc{pl}.\textsc{poss} upper\_lip.\textsc{f} pierce \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}\\
\glt `He pierced their lips.'
\z

\ea\label{ex:03-a65}
Tohiyemata.\\
\gll to-hiya-himata\\
     \textsc{ch}-be\_different-\textsc{fp}.\textsc{n}.\textsc{m}\\
\glt `He changed into something else.'
\z

\ea\label{ex:03-a66}
Tohiyematamona amane, wakari.\\
\gll to-hiya-himata-mona ama-ne wakari\\
     \textsc{ch}-be\_different-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m} Paumari(\textsc{m})\\
\glt `He changed into a Paumari.'
\z

\ea\label{ex:03-a67}
Hiwa sari kanematamonaha.\\
\gll hiwa sari ka-na-himata-mona-ha\\
     3\textsc{sg}.\textsc{refl}.\textsc{m} burn \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
\glt `He set fire to himself.'
\z

\ea\label{ex:03-a68}
Faya me wiwinawa toemetemone fahi, yobe nafi me fawa tonia mati.\\
\gll faya me wi-wina-waha to-ha-hemete-mone fahi yobe nafi me fawa to-niha mati\\
     so 3\textsc{pl}.\textsc{s} \textsc{dup}-live-change \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then house(\textsc{m}) all 3\textsc{pl}.\textsc{s} disappear \textsc{ch}-\textsc{caus}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
\glt `They lived there a long time. The house was full of people.'
\z

\ea\label{ex:03-a69}
Me tafatafa toemetemone ahi.\\
\gll me tafa-tafa to-ha-hemete-mone ahi\\
     3\textsc{pl}.\textsc{s} \textsc{dup}-eat \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
\glt `They ate.'
\z

\ea\label{ex:03-a70}
Faya me tatafa toa me mato me kakawi nawaha ahi hoo yama wasihiba hiwa tonahiyemata.\\
\gll faya me ta-tafa to-ha me mato me ka-kawi na-waha ahi hoo yama wasi-hiba hiwa to-na-hiya-himata\\
     so 3\textsc{pl}.\textsc{s} \textsc{dup}-eat \textsc{ch}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} tree.sp(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{dup}-eat\_with\_something \textsc{aux}-change.\textsc{f} then whew thing(\textsc{f}) find-\textsc{fut}.\textsc{m} 3\textsc{sg}.\textsc{refl}.\textsc{m} \textsc{ch}-\textsc{caus}-be\_different-\textsc{fp}.\textsc{n}.\textsc{m}\\
\glt `They ate. They had starch to eat with their \textit{mato} fruits. He changed himself so he could get crops.'
\z

\ea\label{ex:03-a71}
Hiwa tonahiyemata yama wasimata ka yama ota kakaba toineni, fowa kabe hari.\\
\gll hiwa to-na-hiya-himata yama wasi-himata ka yama otaa ka-kaba to-ha-ne-ni fowa\_kabe haari\\
     3\textsc{sg}.\textsc{refl}.\textsc{m} \textsc{ch}-\textsc{caus}-be\_different-\textsc{fp}.\textsc{n}.\textsc{m} thing(\textsc{f}) find-\textsc{fp}.\textsc{n}.\textsc{m} which thing(\textsc{f}) 1\textsc{ex}.\textsc{s} \textsc{dup}-eat \textsc{ch}-\textsc{aux}-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f} sweet\_manioc(\textsc{m}) that\_one.\textsc{m}\\
\glt `He changed into something else. He was the one who first found the sweet manioc that we eat.'
\z

\ea\label{ex:03-a72}
Otara yokohimari amane.\\
\gll otara yoko-himari ama-ne\\
     1\textsc{ex}.\textsc{o} bring\_misfortune\_on-\textsc{fp}.\textsc{e}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m}\\
\glt `He caused us to suffer.'
\z

\ea\label{ex:03-a73}
Otara yokohimari amaka.\\
\gll otara yoko-himari ama-ka\\
     1\textsc{ex}.\textsc{o} bring\_misfortune\_on-\textsc{fp}.\textsc{e}.\textsc{m} \textsc{sec}-\textsc{decl}.\textsc{m}\\
\glt `He caused us to suffer.'
\z

\ea\label{ex:03-a74}
Ota bono tiwa ha, ota bono.\\
\gll otaa bono ti-awa haa otaa bono\\
     1\textsc{ex}.\textsc{poss} upper\_lip 2\textsc{sg}.\textsc{s}-see \textsc{dem} 1\textsc{ex}.\textsc{poss} upper\_lip\\
\glt `Look at our lips.'
\z

\ea\label{ex:03-a75}
Yama komemetemone okaki hinahamaro amake.\\
\gll yama koma-hemete-mone okaki hi-na-hamaro ama-ke\\
     thing(\textsc{f}) hurt-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 1\textsc{sg}.\textsc{poss}.grandmother(\textsc{f}) \textsc{oc}-say-\textsc{fp}.\textsc{e}.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f}\\
\glt `My grandmother said it hurt.'
\z

\ea\label{ex:03-a76}
Tibono moto tina.\\
\gll ti-bono moto ti-na\\
     2\textsc{sg}.\textsc{poss}-upper\_lip pierce 2\textsc{sg}.\textsc{s}-\textsc{aux}\\
\glt `{``}Pierce your lips.'''
\z

\ea\label{ex:03-a77}
Faya me wiwina toha kimi, kimi botorisahi me baka na me, kimi me baka nemetemoneke ahi.\\
\gll faya me wi-wina to-ha kimi kimi boto-risahi me baka na me kimi me baka na-hemete-mone-ke ahi\\
     so 3\textsc{pl}.\textsc{s} \textsc{dup}-live \textsc{ch}-\textsc{aux}.\textsc{f} corn(\textsc{m}) corn(\textsc{m}) dry\_up-down.\textsc{m} 3\textsc{pl}.\textsc{s} break \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} corn(\textsc{m}) 3\textsc{pl}.\textsc{s} break \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
\glt `They broke off the ears of corn when its leaves were dry.'
\z

\ea\label{ex:03-a78}
Kimi me baka na kimi me sako na me, kimi ewene bere niha yobe toro ya kimi bere niha yobe, kimi bere kanihematamonaka ahi.\\
\gll kimi me baka na kimi me sako na me kimi ewene bere niha yobe toro ya kimi bere niha yobe kimi bere ka-niha-himata-mona-ka ahi\\
     corn(\textsc{m}) 3\textsc{pl}.\textsc{s} break \textsc{aux}.\textsc{f} corn(\textsc{m}) 3\textsc{pl}.\textsc{s} tie \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} corn(\textsc{m}) wood.\textsc{m} be\_across \textsc{caus}.\textsc{list} house(\textsc{m}) inside.\textsc{m} \textsc{adju} corn(\textsc{m}) be\_across \textsc{caus}.\textsc{list} house(\textsc{m}) corn(\textsc{m}) be\_across \textsc{comit}-\textsc{caus}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `The pole for holding the ears of corn was laid on top of the rafters. He draped the corn on the pole inside the house.'
\z

\ea\label{ex:03-a79}
Botokori te kabaho.\\
\gll botokori tee kaba-ho\\
     bad\_ear 2\textsc{pl}.\textsc{s} eat-\textsc{imp}.\textsc{m}\\
\glt `{``}You eat the bad ears of corn.'''
\z

\ea\label{ex:03-a80}
Botokori te rawi ha botokori te kaba naho.\\
\gll botokori tee rawi haa botokori tee kaba na-ho\\
     bad\_ear 2\textsc{pl}.\textsc{s} toast.\textsc{nom}.\textsc{f} \textsc{dem} bridge 2\textsc{pl}.\textsc{s} eat.\textsc{list} \textsc{aux}-\textsc{imp}.\textsc{m}\\
\glt `{``}You toast the kernels from the bad ears in the pan. Eat the bad ears of corn.'''
\z

\ea\label{ex:03-a81}
Faya kimi botokori me rawehemetemoneke, hinakakirawaha ahi.\\
\gll faya kimi botokori me rawa-hemete-mone-ke hinakaki-rawa-ha ahi\\
     so corn(\textsc{m}) bad\_ear 3\textsc{pl}.\textsc{s} toast-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} 3\textsc{poss}.\textsc{sg}.grandmother(\textsc{f})-\textsc{fpl}-\textsc{dup} then\\
\glt `His grandmothers toasted the kernels from the bad ears of corn in the pan.'
\z

\ea\label{ex:03-a82}
Kakasi raba raba, kimi kababone ona ini si oniha tokineni.\\
\gll ka-kasi raba ra-haba kimi kaba-habone o-na ini sii o-niha to-ka-ne-ni\\
     \textsc{dup}-be\_naughty a\_bit \textsc{neg}-\textsc{fut}.\textsc{f} corn(\textsc{m}) eat-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{s}-say.\textsc{f} tooth.\textsc{f} stand 1\textsc{sg}.\textsc{s}-\textsc{caus}.\textsc{f} away-go-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
\glt `{``}She is naughty. I thought she was going to eat corn. I put in her teeth, and she left.'''
\z

\ea\label{ex:03-a83}
Ini si oniha tokine ninihi, maka.\\
\gll ini sii o-niha to-ka-ne ni-ni-hi maka\\
     tooth.\textsc{f} stand 1\textsc{sg}.\textsc{s}-\textsc{caus}.\textsc{f} away-go-\textsc{cont}.\textsc{f} \textsc{aux}.\textsc{nom}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup} snake(\textsc{f})\\
\glt `{``}The snake left when no one was looking, after I put in her teeth.'''
\z

\ea\label{ex:03-a84}
Tabiyo hora hina, tabiyo hora hinemetemone.\\
\gll tabiyo hora hi-na tabiyo hora hi-na-hemete-mone\\
     absence complain\_about \textsc{oc}-\textsc{aux}.\textsc{f} absence complain\_about \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
\glt `He complained about her.'
\z

\ea\label{ex:03-a85}
Tabiyo hora hina, kamemetemone itineti.\\
\gll tabiyo hora hi-na ka-ma-hemete-mone ita-ne-ti\\
     absence complain\_about \textsc{oc}-\textsc{aux}.\textsc{f} come-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} stay-\textsc{cont}.\textsc{f}-\textsc{hab}.\textsc{nom}\\
\glt `He complained about her. She never came back.'
\z

\ea\label{ex:03-a86}
Faya kamara tabiyo hinaka noti hora hina, faya me wiwinawa toemetemone fahi.\\
\gll faya ka-ma-ra tabiyo hinaka noti hora hi-na faya me wi-wina-waha to-ha-hemete-mone fahi\\
     so come-back-\textsc{neg}.\textsc{f} absence 3\textsc{sg}.\textsc{poss} grandson(\textsc{m}) complain\_about \textsc{oc}-\textsc{aux}.\textsc{f} so 3\textsc{pl}.\textsc{s} \textsc{dup}-live-change \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
\glt `She didn't come back. Her grandson complained about her. They lived there.'
\z

\ea\label{ex:03-a87}
Me winawa me tafatafa toemetemone.\\
\gll me wina-waha me tafa-tafa to-ha-hemete-mone\\
     3\textsc{pl}.\textsc{s} live-change.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{dup}-eat \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
\glt `They lived there. They ate.'
\z

\ea\label{ex:03-a88}
Fowa me hiti, biha waya nisaha me wi nawaha, ati ehene.\\
\gll fowa me hiti biha waya na-risaha me wii na-waha ati ehene\\
     manioc(\textsc{m}) 3\textsc{pl}.\textsc{s} grate.\textsc{list} yam(\textsc{f}) wilt \textsc{aux}-down.\textsc{f} 3\textsc{pl}.\textsc{s} dig \textsc{aux}-change.\textsc{f} voice because\_of.\textsc{m}\\
\glt `They grated manioc. They dug up yams when the leaves dried, because he told them to.'
\z

\ea\label{ex:03-a89}
Bia te wi na, haso, yobo, nakafira, yama te wi na yama one te kabahi.\\
\gll biha tee wii na haso yobo nakafira yama tee wii na yama one tee kaba-hi\\
     yam(\textsc{f}) 2\textsc{pl}.\textsc{s} dig \textsc{aux}.\textsc{f} taro.sp(\textsc{m}) taro.sp(\textsc{m}) taro.sp(\textsc{f}) thing(\textsc{f}) 2\textsc{pl}.\textsc{s} dig \textsc{aux}.\textsc{f} thing(\textsc{f}) other.\textsc{f} 2\textsc{pl}.\textsc{s} eat-\textsc{imp}.\textsc{f}\\
\glt `{``}When you have dug up the yams, and dug up the \textit{haso} taro, the \textit{yobo} taro, and the \textit{nakafira} taro, eat other things.'''
\z

\ea\label{ex:03-a90}
Yama noki te nahi.\\
\gll yama noki tee na-hi\\
     thing(\textsc{f}) seed.\textsc{f} 2\textsc{pl}.\textsc{s} put\_in-\textsc{imp}.\textsc{f}\\
\glt `{``}You keep the seeds.'''
\z

\ea\label{ex:03-a91}
Noki te, yama wiye ya noki te iharawi ya, noki te koro nabana.\\
\gll noki tee yama wiye ya noki tee iha-rawi ya noki tee koro na-habana\\
     seed.\textsc{f} 2\textsc{pl} thing(\textsc{f}) container.\textsc{f} \textsc{adju} seed.\textsc{f} 2\textsc{pl}.\textsc{s} put-\textsc{fpl}.\textsc{nom}.\textsc{f} \textsc{adju} seed.\textsc{f} 2\textsc{pl}.\textsc{s} plant \textsc{aux}-\textsc{fut}.\textsc{f}\\
\glt `{``}Put the seeds in a container, to plant them.'''
\z

\newpage

\ea\label{ex:03-a92}
Faya yama me wi na yama me kakaba toemetemone fahi.\\
\gll faya yama me wii na yama me ka-kaba to-ha-hemete-mone fahi\\
     so thing(\textsc{f}) 3\textsc{pl}.\textsc{s} dig \textsc{aux}.\textsc{f} thing(\textsc{f}) 3\textsc{pl}.\textsc{s} \textsc{dup}-eat \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
\glt `They dug the crops up. They ate them.'
\z

\ea\label{ex:03-a93}
Yama me kaba me, me tatafa toemetemoneke waha.\\
\gll yama me kaba me me ta-tafa to-ha-hemete-mone-ke waha\\
     thing(\textsc{f}) 3\textsc{pl}.\textsc{s} eat.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} \textsc{dup}-eat \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} now\\
\glt `They ate them. They ate.'
\z

\ea\label{ex:03-a94}
Me tafa me, yama me wi hina ta toremetemoneke ahi, biha habi.\\
\gll me tafa me yama me wii hi-na taa to-re-hemete-mone-ke ahi biha habi\\
     3\textsc{pl}.\textsc{s} eat.\textsc{f} 3\textsc{pl}.\textsc{s} thing(\textsc{f}) 3\textsc{pl}.\textsc{s} dig \textsc{oc}-\textsc{aux}.\textsc{f} be\_piled\_up \textsc{ch}-raised\_surface-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then yam(\textsc{f}) root.\textsc{f}\\
\glt `They ate. The things they dug up were in a pile, the yams.'
\z

\ea\label{ex:03-a95}
Bia habohabo raba terani yama nemata.\\
\gll biha habo-habo raba tee-ra-hani yama na-himata\\
     yam(\textsc{f}) \textsc{dup}-have\_tubers a\_bit \textsc{hab}-\textsc{neg}-\textsc{ip}.\textsc{n}.\textsc{f} work\_on \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}\\
\glt `He dug up yams, that had many tubers.'
\z

\ea\label{ex:03-a96}
Biha, bia kero, bia wao, bia, bia anabia, bia ini yabematamonaka.\\
\gll biha biha\_kero biha\_wao biha biha\_anabia biha ini yaba-himata-mona-ka\\
     yam(\textsc{f}) yam\_variety(\textsc{f}) yam\_variety(\textsc{f}) yam(\textsc{f}) yam\_variety(\textsc{f}) yam(\textsc{f}) name.\textsc{f} say-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `He gave names to \textit{kero} yams, \textit{wao} yams, \textit{anabia} yams.'
\z

\ea\label{ex:03-a97}
Bia, bia anabia amake haro.\\
\gll biha biha\_anabia ama-ke haaro\\
     yam(\textsc{f}) yam\_variety(\textsc{f}) be-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
\glt `{``}They are \textit{anabia} yams.'''
\z

\ea\label{ex:03-a98}
Bia wao ama ha.\\
\gll biha\_wao ama haa\\
     yam\_variety(\textsc{f}) be \textsc{dem}\\
\glt `{``}They are \textit{wao} yams.'''
\z

\ea\label{ex:03-a99}
Biha kero ama haro.\\
\gll biha\_kero ama haaro\\
     yam\_variety(\textsc{f}) be that\_one.\textsc{f}\\
\glt `{``}They are \textit{kero} yams.'''
\z

\ea\label{ex:03-a100}
Bia kero amake haro.\\
\gll biha\_kero ama-ke haaro\\
     yam\_variety(\textsc{f}) be-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
\glt `{``}They are \textit{kero} yams.'''
\z

\ea\label{ex:03-a101}
Biha ini yabe, bia, haso, haso, haso ama hari, haso.\\
\gll biha ini yabe biha haso haso haso ama haari haso\\
     yam(\textsc{f}) name.\textsc{f} say.\textsc{m} yam(\textsc{f}) taro.sp(\textsc{m}) taro.sp(\textsc{m}) taro.sp(\textsc{m}) be that\_one.\textsc{m} taro.sp(\textsc{m})\\
\glt `He gave names to the yams. ``They are \textit{haso} taros.'''
\z

\ea\label{ex:03-a102}
Haso ama ha.\\
\gll haso ama haa\\
     taro.sp(\textsc{m}) be \textsc{dem}\\
\glt `{``}They are \textit{haso} taros.'''
\z

\ea\label{ex:03-a103}
Hahi yobo amaka hari, yobo.\\
\gll haahi yobo ama-ka haari yobo\\
     this\_one.\textsc{m} taro.sp(\textsc{m}) be-\textsc{decl}.\textsc{m} that\_one.\textsc{m} taro.sp(\textsc{m})\\
\glt `{``}There are the \textit{yobo} taros.'''
\z

\ea\label{ex:03-a104}
Yobo ama ha.\\
\gll yobo ama haa\\
     taro.sp(\textsc{m}) be \textsc{dem}\\
\glt `{``}They are \textit{yobo} taros.'''
\z

\ea\label{ex:03-a105}
Yama me iso hina ta tokanare ini yabematamonane.\\
\gll yama me iso hi-na taa to-ka-na-re ini yaba-himata-mona-ne\\
     thing(\textsc{f}) 3\textsc{pl}.\textsc{s} carry\_off \textsc{oc}-\textsc{aux}.\textsc{f} be\_piled\_up \textsc{ch}-\textsc{comit}-\textsc{aux}-raised\_surface name.\textsc{f} say-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
\glt `The things they carried off were in piles, and he gave them names.'
\z

\ea\label{ex:03-a106}
Faya yama ini yabe, Himata amari ha?\\
\gll faya yama ini yabe himata ama-ri haa\\
     so thing(\textsc{f}) name.\textsc{f} say.\textsc{m} what be-\textsc{cq}.\textsc{f} \textsc{dem}\\
\glt `He gave them names. ``What is that?'''
\z

\ea\label{ex:03-a107}
Yawita amane, yawita.\\
\gll yawita ama-ne yawita\\
     peach\_palm(\textsc{m}) be-\textsc{bkg}.\textsc{m} peach\_palm(\textsc{m})\\
\glt `{``}That is a peach palm.'''
\z

\ea\label{ex:03-a108}
Yawita ama hari.\\
\gll yawita ama haari\\
     peach\_palm(\textsc{m}) be that\_one.\textsc{m}\\
\glt `{``}It's a peach palm.'''
\z

\ea\label{ex:03-a109}
Yifari amake haro, yifari biri.\\
\gll yifari ama-ke haaro yifari\_biri\\
     banana(\textsc{f}) be-\textsc{decl}.\textsc{f} that\_one.\textsc{f} plantain\_variety(\textsc{f})\\
\glt `{``}Those are \textit{biri} plantains there.'''
\z

\ea\label{ex:03-a110}
Yifari biri ama haro.\\
\gll yifari\_biri ama haaro\\
     plantain\_variety(\textsc{f}) be that\_one.\textsc{f}\\
\glt `{``}Those are \textit{biri} plantains there.'''
\z

\ea\label{ex:03-a111}
Yifari biri, yifari howe, yifari howe ama haro, yifari howe.\\
\gll yifari\_biri yifari\_howe yifari\_howe ama haaro yifari\_howe\\
     plantain\_variety(\textsc{f}) plantain\_variety(\textsc{f}) plantain\_variety(\textsc{f}) be that\_one.\textsc{f} plantain\_variety(\textsc{f})\\
\glt `{``}Those are \textit{biri} plantains, \textit{howe} plantains.'''
\z

\ea\label{ex:03-a112}
Yifari kofiti, yifari kofiti ama ha.\\
\gll yifari\_kofiti yifari\_kofiti ama haa\\
     banana\_variety(\textsc{f}) banana\_variety(\textsc{f}) be \textsc{dem}\\
\glt `{``}Those are \textit{kofiti} bananas.'''
\z

\ea\label{ex:03-a113}
Yifari sawa ama ha.\\
\gll yifari\_sawa ama haa\\
     banana\_variety(\textsc{f}) be \textsc{dem}\\
\glt `{``}Those are \textit{sawa} bananas.'''
\z

\ea\label{ex:03-a114}
Yifari sawa ama ha.\\
\gll yifari\_sawa ama haa\\
     banana\_variety(\textsc{f}) be \textsc{dem}\\
\glt `{``}Those are \textit{sawa} bananas.'''
\z

\ea\label{ex:03-a115}
Yifari mawara amani.\\
\gll yifari\_mawara ama-ni\\
     banana\_variety(\textsc{f}) be-\textsc{bkg}.\textsc{f}\\
\glt `{``}Those are \textit{mawara} bananas.'''
\z

\ea\label{ex:03-a116}
Yifari mawara ama haro.\\
\gll yifari\_mawara ama haaro\\
     banana\_variety(\textsc{f}) be that\_one.\textsc{f}\\
\glt `{``}Those are \textit{mawara} bananas.'''
\z

\ea\label{ex:03-a117}
Yifari kofiti amani.\\
\gll yifari\_kofiti ama-ni\\
     banana\_variety(\textsc{f}) be-\textsc{bkg}.\textsc{f}\\
\glt `{``}Those are \textit{kofiti} bananas.'''
\z

\ea\label{ex:03-a118}
Yama ini yabematamonaka ahi.\\
\gll yama ini yaba-himata-mona-ka ahi\\
     thing(\textsc{f}) name.\textsc{f} say-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `He was giving names to the crops.'
\z

\ea\label{ex:03-a119}
Yama ini yabe ama.\\
\gll yama ini yabe ama\\
     thing(\textsc{f}) name.\textsc{f} say.\textsc{m} \textsc{sec}\\
\glt `He was giving names to the crops.'
\z

\ea\label{ex:03-a120}
Yifari kofiti, ananika, ananika ama ha, aya wawa naro.\\
\gll yifari\_kofiti ananika ananika ama haa aya wa-waa na-haaro\\
     banana\_variety(\textsc{f}) banana\_variety(\textsc{f}) banana\_variety(\textsc{f}) be \textsc{dem} this \textsc{dup}-stand \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f}\\
\glt `{``}\textit{Kofiti} bananas, \textit{ananika} bananas, that are just this high.'''
\z

\ea\label{ex:03-a121}
Ananika amake haro.\\
\gll ananika ama-ke haaro\\
     banana\_variety(\textsc{f}) be-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
\glt `{``}Those are \textit{ananika} bananas.'''
\z

\ea\label{ex:03-a122}
Ananika ama ha, nemeharaharo.\\
\gll ananika ama haa nemeha-ra-haaro\\
     banana\_variety(\textsc{f}) be \textsc{dem} be\_high-\textsc{neg}-\textsc{ip}.\textsc{e}.\textsc{f}\\
\glt `{``}Those are \textit{ananika} banana plants, they are not high.'''
\z

\ea\label{ex:03-a123}
Faya yifari ini yabe Hibaka yoharibe?\\
\gll faya yifari ini yabe hibaka yowa-ri-be\\
     so banana(\textsc{f}) name.\textsc{f} say.\textsc{m} where reach-\textsc{cq}.\textsc{f}-\textsc{immed}.\textsc{f}\\
\glt `He was naming the bananas. ``Where should we stop?'''
\z

\ea\label{ex:03-a124}
Te kayabahi.\\
\gll tee ka-yaba-hi\\
     2\textsc{pl}.\textsc{s} \textsc{comit}-take\_out-\textsc{imp}.\textsc{f}\\
\glt `{``}You pick them.'''
\z

\ea\label{ex:03-a125}
Te yome nibeya.\\
\gll tee yome na-be-ya\\
     2\textsc{pl}.\textsc{s} swallow \textsc{aux}-\textsc{immed}.\textsc{f}-now\\
\glt `{``}You eat them.'''
\z

\ea\label{ex:03-a126}
Te kayabahi.\\
\gll tee ka-yaba-hi\\
     2\textsc{pl}.\textsc{s} \textsc{comit}-take\_out-\textsc{imp}.\textsc{f}\\
\glt `{``}You pick them.'''
\z

\ea\label{ex:03-a127}
Yifari me kayaba, yifari me iso nawaha, Yome naba amake haro, nafi.\\
\gll yifari me ka-yaba yifari me iso na-waha yome na-haba ama-ke haaro nafi\\
     banana(\textsc{f}) 3\textsc{pl}.\textsc{s} \textsc{comit}-take\_out.\textsc{f} banana(\textsc{f}) 3\textsc{pl}.\textsc{s} carry\_off \textsc{aux}-change.\textsc{f} be\_swallowed \textsc{aux}-\textsc{fut}.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f} that\_one.\textsc{f} all\\
\glt `They picked the bananas. They carried the bananas off. ``They will be eaten, all of them.'''
\z

\ea\label{ex:03-a128}
Mahi amosake haro.\\
\gll mahi amosa-ke haaro\\
     smell.\textsc{f} be\_good-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
\glt `{``}They have a good smell.'''
\z


\newpage

\ea\label{ex:03-a129}
Yifari mai amosa awine faro.\\
\gll yifari mahi amosa awine faaro\\
     banana(\textsc{f}) smell.\textsc{f} be\_good.\textsc{f} seem.\textsc{f} that.\textsc{f}\\
\glt `{``}Those bananas smell good.'''
\z

\ea\label{ex:03-a130}
Yifari kofiti amani.\\
\gll yifari\_kofiti ama-ni\\
     banana\_variety(\textsc{f}) be-\textsc{bkg}.\textsc{f}\\
\glt `{``}They are \textit{kofiti} bananas.'''
\z

\ea\label{ex:03-a131}
Yifari kofiti amake haro.\\
\gll yifari\_kofiti ama-ke haaro\\
     banana\_variety(\textsc{f}) be-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
\glt `{``}They are \textit{kofiti} bananas.'''
\z

\ea\label{ex:03-a132}
Haha nasa ama haro, nasa.\\
\gll haaha nasa ama haaro nasa\\
     this.\textsc{f} banana\_variety(\textsc{f}) be that\_one.\textsc{f} banana\_variety(\textsc{f})\\
\glt `{``}Those are \textit{nasa} bananas.'''
\z

\ea\label{ex:03-a133}
Nasa amake haro.\\
\gll nasa ama-ke haaro\\
     banana\_variety(\textsc{f}) be-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
\glt `{``}They are \textit{nasa} bananas.'''
\z

\ea\label{ex:03-a134}
Faya, me, yifari me kayaba me, yifari me iso na yifari me yoyome toemetemoneke, me ka yifari.\\
\gll faya me yifari me ka-yaba me yifari me iso na yifari me yo-yome to-ha-hemete-mone-ke me ka yifari\\
     so 3\textsc{pl} banana(\textsc{f}) 3\textsc{pl}.\textsc{s} \textsc{comit}-take\_out.\textsc{f} 3\textsc{pl}.\textsc{s} banana(\textsc{f}) 3\textsc{pl}.\textsc{s} carry\_off \textsc{aux}.\textsc{f} banana(\textsc{f}) 3\textsc{pl}.\textsc{s} \textsc{dup}-swallow \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} 3\textsc{pl} \textsc{poss} banana(\textsc{f})\\
\glt `They picked the bananas. They carried the bananas off. They ate the bananas, their bananas.'
\z

\ea\label{ex:03-a135}
Me nafi fowa me hiti nawaha, ati ehene.\\
\gll me nafi fowa me hiti na-waha ati ehene\\
     3\textsc{pl}.\textsc{poss} all manioc(\textsc{m}) 3\textsc{pl}.\textsc{s} grate \textsc{aux}-change.\textsc{list} voice because\_of.\textsc{m}\\
\glt `They all grated manioc, too, because he told them to.'
\z

\ea\label{ex:03-a136}
Fowa te hiti nahi, me tokaha me te noki niba karo, bani me kakawibana.\\
\gll fowa tee hiti na-hi me to-kaha me tee noki ni-ba karo bani me kakawi-bana\\
     manioc(\textsc{m}) 2\textsc{pl}.\textsc{s} grate \textsc{aux}-\textsc{imp}.\textsc{f} 3\textsc{pl}.\textsc{s} away-go.\textsc{f} 3\textsc{pl}.\textsc{o} 2\textsc{pl}.\textsc{s} wait\_for \textsc{aux}.\textsc{nom}.\textsc{f}-\textsc{fut} so\_that animal(\textsc{m}) 3\textsc{pl}.\textsc{poss} starch-\textsc{fut}\\
\glt `{``}You grate manioc, it will be eaten with the game meat. You wait for the others, who went hunting.'''
\z

\ea\label{ex:03-a137}
Sawimareba, sawimareba hiwa sari kanematamona ama, wakari, wakari me toehibana.\\
\gll sawi-ma-ra-hiba sawi-ma-ra-hiba hiwa sari ka-na-himata-mona ama wakari wakari me to-ha-hibana\\
     be\_present-back-\textsc{neg}-\textsc{fut}.\textsc{m} be\_present-back-\textsc{neg}-\textsc{fut}.\textsc{m} 3\textsc{sg}.\textsc{refl}.\textsc{m} burn \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} \textsc{sec} Paumari(\textsc{m}) Paumari(\textsc{m}) 3\textsc{pl}.\textsc{compl} \textsc{ch}-be-\textsc{fut}.\textsc{m}\\
\glt `He burned himself up, the one who wouldn't stay there, the one who became the Paumaris.'
\z

\ea\label{ex:03-a138}
Wakari me toehiba, faya sawimare tokehimatamonaka.\\
\gll wakari me to-ha-hiba faya sawi-ma-re to-ka-himata-mona-ka\\
     Paumari(\textsc{m}) 3\textsc{pl}.\textsc{compl} \textsc{ch}-be-\textsc{fut}.\textsc{m} so be\_present-back-\textsc{neg}.\textsc{m} away-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `He became the Paumaris, and he didn't stay. He went away.'
\z

\ea\label{ex:03-a139}
Tokehimatamona, wakari me tohati, biti mati.\\
\gll to-ka-himata-mona wakari me to-ha-ti biti mati\\
     away-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} Paumari(\textsc{m}) 3\textsc{pl}.\textsc{compl} \textsc{ch}-be-\textsc{hab}.\textsc{nom} 3\textsc{sg}.\textsc{poss}.son(\textsc{m}) 3\textsc{pl}\\
\glt `He went away, the one who became the Paumaris. His sons, too.'
\z
