 The old man was hungry lying in his hammock. The children were crying. The boy was this big. There were two younger brothers and their older brother. There were three of them.

 ``Father, my sons are there. You toast some manioc mash for my sons. There is manioc mash there in the pot. Toast some manioc mash for my sons. Feed them. I'm going to go get my bananas, to be food for my sons.''

 She went with her husband. Her older brother also went out hunting in the forest. ``Younger sister, you come back quickly. I'm going hunting. I'm going to shoot animals with my blowgun.''

 ``Older brother, come back quickly.''

 ``I will not be long in coming back, if I kill an animal with my blowgun,'' he said.

 They went to the garden. They went, and arrived. They got the bananas, also pineapples, ripe bananas. They started back.

 The old man was lying in his hammock. He spoke as he lay in his hammock. He picked up a spear. ``Take your baths quickly. You go on ahead. You take your baths.''

 He got a pot and a knife. His grandsons went.

 ``The two of you go upstream. Go upstream, then turn around and come back downstream here. I will stand here.''

 His spear was against a tree. They came back downstream.

 ``One of you carry the other and come downstream in the water.''

 They came downstream in the water. They came downstream, and he speared both of them with the spear. They died where they were. Their older brother, who was big, jumped and went running away. He went running to their parents.

 Then he cut them up. He cut them in pieces, and put them in the pot. He put them in the pot, and carried them. He brought them, and cooked them. He roasted a leg of one of them stuck on a stick by the fire, for his mother. He roasted his leg on a stick by the fire. The old man toasted manioc mash. He toasted manioc mash, and ate. He ate his grandsons, and then lay in his hammock with a full stomach.

 Then they appeared. He covered the meat for his daughter. When he covered the meat, they appeared. The oldest son had gone and had arrived at the other village. He was a teenager. He was big.  He went to his parents, and arrived. He told the news.

 ``What about my sons?''

 ``Don't say, `What about my sons?' My grandfather speared your two sons. I ran away because of my grandfather, and came here.''

 ``I came here to get food for my sons.''

 Then his daughter cried. She cried. She came crying. She stopped crying. She came, and appeared. She came back crying.

 ``Father, you ate my sons. I went to get food for my sons, and you ate my sons,'' she said, crying.

 ``Don't cry. Don't make trouble for yourself.\footnote{This was a threat to kill his daughter if she didn't stop berating him.} Yours is there in the pot. Eat yours after coming back, without saying anything. Don't cry. Eat it. I ate it. Eat them. Don't cause trouble for yourself.''

 The father of the children didn't say anything. Their father didn't say anything, but his wife cried. His brother-in-law came back, carrying animals. His brother-in-law came back, and threw down the animals. ``Whew.'' He came upon his younger sister as she was crying.

 ``Why are you crying?''

 It was another woman who spoke. ``She is crying because your father ate her sons. Your father ate his grandsons, and he is in the hammock there.''

 The old man's spear was leaning against something beside the fire. He was waiting for them. 	When the old man's son heard what they said, he got up. He got ready. He went to his father as he was in his hammock. He wasn't violent with him.

 ``Father, why did you eat my nephews? They said you ate my nephews. Didn't I bring animals?''

 He was going to pick up the spear, but he grabbed his spear from him. He twisted his spear away from him, but he didn't hit him. He stomped on his spear.

 ``I don't want to kill you. I don't want to kill you with a piece of your own spear,'' he said. He just threatened him.

 His brother-in-law singed the hair off the animals. He singed the hair off the animals, and cut up the animals. They ate the meat. He roasted the meat on a rack.

 ``Isn't your father going to eat it?''

 He took out his shaman's stone. He coughed it up. He had it in his hand. He pressed it against the leg of the animal. He pressed it against the meat.

 ``Here, eat some. Here is some manioc mash. Eat it with some manioc mash. Even though you ate my sons, I'm not mad at you. I am giving you food.''

 So he gave food to his father-in-law. He gave food to his father-in-law, and he ate the meat. When he was finished eating the meat, he drank water. When it was getting dark, he started speaking, sick.

 ``\textit{Aya}, my chest, my chest,'' he said. ``My chest,'' he said. His voice was heard during the night.

 Because his sons had died, he had given his father-in-law something to eat that had a spell on it to make him sick. He gave his father-in-law something to eat that had a spell on it, and his father-in-law was heard during the night.

 Morning came. When morning came, they moved to another place. They moved to another place, and they didn't wait for him. He had chest pain during the night. They didn't wait for him. They moved away. They left. They went. His son didn't take care of him.

 ``You lie here in your hammock. I'm not helping you, because you did a bad thing, eating my nephews. We are leaving.''

 So he left his father behind, and left. He died after they left. He lay dead in his hammock. When he rotted, the vultures ate him. The vultures ate him, the king vultures.

 He came back, his son. He came back to see his father. His son looked at him.

 ``My father has died. Let him die. I don't want to smell him,'' he said.

 He turned around and went back.

 ``Let him die.''

 He went back. He arrived back.

 ``Well?''

 ``You ask, `Well?' ‒ he died. The vultures ate him, the king vultures.''

 So they slept.

 ``He should have stayed calm. He didn't take care of his grandsons. He ate his grandsons. Now they have killed him. He was always wanting more than his share of food,'' one of them said. ``Because he was always wanting more food, he ate his grandsons. Let him die.''

 He died.
