\ea\label{ex:06-a1}
    Kose mati, kose mati inamati, inamati yetene, inamati yetene yetene afo nari, me afo hinahari.\\
    \gll kose\_mati kose\_mati inamati inamati yetene inamati yetene yetene afo na-haari me afo hi-na-haari\\
    name\_of\_monster(\textsc{m}) name\_of\_monster(\textsc{m}) someone(\textsc{m}) someone(\textsc{m}) teenage\_boy someone(\textsc{m}) teenage\_boy teenage\_boy be\_blown\_into \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m} 3\textsc{pl}.\textsc{s} blow\_snuff\_into \textsc{oc}-\textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `The \textit{kose mati} monster. The teenager was trained as a shaman. They trained him as a shaman.'
\z

\ea\label{ex:06-a2}
Me afo hine, me aafo hie me, me sa hikase, wawama toahi, yama yama yete nematamona.\\
\gll me afo hi-ne me a-afo hi-he me me saa hi-ka-kose wa-waa-ma to-hahi yama yama yete na-himata-mona\\
     3\textsc{pl}.\textsc{s} blow\_snuff\_into \textsc{oc}-\textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{s} \textsc{dup}-blow\_snuff\_into \textsc{oc}-\textsc{aux}.\textsc{m} 3\textsc{pl} 3\textsc{pl}.\textsc{s} release \textsc{oc}-\textsc{comit}-middle.\textsc{m} \textsc{dup}-stand-back \textsc{ch}-\textsc{aux}.\textsc{m} thing(\textsc{f}) thing(\textsc{f}) hunt \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
\glt `They trained him as a shaman. They trained him over a period of time, and then they were done with him. He was walking around normally again. He went hunting.'
\z

\ea\label{ex:06-a3}
Abi abi yama yete obana?\\
\gll abi abi yama yete o-bana\\
     father(\textsc{m}) father(\textsc{m}) thing(\textsc{f}) hunt 1\textsc{sg}.\textsc{s}-\textsc{fut}\\
\glt `{``}Father, should I go hunting?'''
\z

\ea\label{ex:06-a4}
Yama yama yete tinahi.\\
\gll yama yama yete ti-na-hi\\
     thing(\textsc{f}) thing(\textsc{f}) hunt 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{f}\\
\glt `{``}Go hunting.'''
\z

\ea\label{ex:06-a5}
Yama yete tinahi.\\
\gll yama yete ti-na-hi\\
     thing(\textsc{f}) hunt 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{f}\\
\glt `{``}Go hunting.'''
\z

\ea\label{ex:06-a6}
Tika.\\
\gll ti-ka\\
     2\textsc{sg}.\textsc{s}-go\\
\glt `{``}Go.'''
\z

\ea\label{ex:06-a7}
Bati hiyose, tokehimatamonaka.\\
\gll bati hi-yose to-ka-himata-mona-ka\\
     3\textsc{sg}.\textsc{poss}.father(\textsc{m}) \textsc{oc}-command away-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `His father ordered him, and he went.'
\z

\ea\label{ex:06-a8}
Me aafo hiahi wawamisama toe, toke, karaboa tiwa kane, karaboa mase hi nematamona, hho hho hohooo ho.\\
\gll me a-afo hi-hahi wa-waa-misa-ma to-he to-ke karaboha tiwa ka-ne karaboha mase hii na-himata-mona hho\_hho\_hohooo\_ho\\
     3\textsc{pl}.\textsc{s} \textsc{dup}-blow\_snuff\_into \textsc{oc}-\textsc{aux}.\textsc{m} \textsc{dup}-stand-up-back \textsc{ch}-\textsc{aux}.\textsc{m} away-go.\textsc{m} blowgun(\textsc{f}) carry \textsc{comit}-\textsc{aux}.\textsc{m} blowgun(\textsc{f}) curassow(\textsc{m}) call \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} (call\_of\_curassow)\\
\glt `They had trained him as a shaman, and he had returned to normal. He went out into the forest, carrying a blowgun on his shoulder. A curassow was calling. ``\textit{Hho} \textit{hho} \textit{hohooo} \textit{ho}.'''
\z

\ea\label{ex:06-a9}
Mase okaraboamatibe.\\
\gll mase o-karaboha-mata-be\\
     curassow(\textsc{m}) 1\textsc{sg}.\textsc{s}-shoot\_with\_blowgun-short\_time-\textsc{immed}.\textsc{f}\\
\glt `{``}I'm going to shoot the curassow with a dart.'''
\z

\ea\label{ex:06-a10}
Mase okaraboamatibe.\\
\gll mase o-karaboha-mata-be\\
     curassow(\textsc{m}) 1\textsc{sg}.\textsc{s}-shoot\_with\_blowgun-short\_time-\textsc{immed}.\textsc{f}\\
\glt `{``}I'm going to shoot the curassow.'''
\z

\ea\label{ex:06-a11}
Eti hawa tini?\\
\gll Eti hawa ti-ha-ni\\
     Edi(\textsc{m}) be\_finished 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{bkg}.\textsc{f}\\
\glt `Edi, are you finished?'
\z

\newpage
\ea\label{ex:06-a12}
Mase karaboebona toke, wasina waha wasina waimetemone amake.\\
\gll mase karaboha-hibona to-ke wasina waaha wasina wai-mete-mone ama-ke\\
     curassow(\textsc{m}) shoot\_with\_blowgun-\textsc{int}.\textsc{m} away-go.\textsc{m} kapok(\textsc{f}) stand.\textsc{f} kapok(\textsc{f}) stand.\textsc{nom}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} be-\textsc{decl}.\textsc{f}\\
\glt `He went on to shoot the curassow. There was a kapok tree.'
\z

\ea\label{ex:06-a13}
    Wasina waha, habi ya, kose mati waiyematamonaka ahi mase, mase hi ni tohahari, tonokomisari ahi, tonokomisari hi nari, Hho hho hohooo o.\\
    \gll wasina waaha habi ya kose\_mati wahiya-himata-mona-ka ahi mase mase hii ni to-ha-haari to-noko-misa-haari ahi to-noko-misa-haari hii na-haari hho\_hho\_hohooo\_o\\
    kapok(\textsc{f}) stand.\textsc{f} root.\textsc{f} \textsc{adju} name\_of\_monster(\textsc{m}) hide-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then curassow(\textsc{m}) curassow(\textsc{m}) call \textsc{aux}.\textsc{nfin} \textsc{ch}-be-\textsc{ip}.\textsc{e}.\textsc{m} away-point-up-\textsc{ip}.\textsc{e}.\textsc{m} then away-point-up-\textsc{ip}.\textsc{e}.\textsc{m} call \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m} (call\_of\_curassow)\\
    \glt `There was a \textit{kose mati} monster hiding in the buttress roots of the kapok tree, making the sound of a curassow calling. It was looking up. ``\textit{Hho} \textit{hho} \textit{hohooo} \textit{o}.'''
\z

\ea\label{ex:06-a14}
Kakehimatamonaka.\\
\gll ka-ke-himata-mona-ka\\
     come-coming-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `He was coming.'
\z

\ea\label{ex:06-a15}
Karaboa karaboa tama ni kake, Mase hi nareka ahi.\\
\gll karaboha karaboha tama na ka-ke mase hii na-hare-ka ahi\\
     blowgun(\textsc{f}) blowgun(\textsc{f}) hold\_onto \textsc{aux}.\textsc{nfin} come-coming curassow(\textsc{m}) call \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m} there\\
\glt `He came with the blowgun in his hand. ``The curassow called there.'''
\z

\ea\label{ex:06-a16}
Towawawitebona kose mati fito nakematamonaka ahi.\\
\gll to-waa-waha-wite-hibona kose\_mati fito na-ke-himata-mona-ka ahi\\
     away-stand-change-out-\textsc{int}.\textsc{m} name\_of\_monster(\textsc{m}) rush \textsc{aux}-coming-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `He moved over to a different position, and the monster attacked him.'
\z

\ea\label{ex:06-a17}
Fito nake he watematamonaka ahi.\\
\gll fito na-ke hee wata-himata-mona-ka ahi\\
     rush \textsc{aux}-coming 3\textsc{sg}.\textsc{o} catch-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `It attacked, and caught him.'
\z

\ea\label{ex:06-a18}
He wate fawa ne bete hikawe he towatamaki, he kakame he kakamematamonaka.\\
\gll hee wate fawa ne bete hi-ka-wee hee to-wata-maki hee ka-ka-me hee ka-ka-ma-himata-mona-ka\\
     3\textsc{sg}.\textsc{o} catch.\textsc{m} disappear \textsc{aux}.\textsc{m} run \textsc{oc}-\textsc{comit}-change.\textsc{m} 3\textsc{sg}.\textsc{o} away-catch-following.\textsc{m} 3\textsc{sg}.\textsc{o} \textsc{comit}-come-back.\textsc{m} 3\textsc{sg}.\textsc{o} \textsc{comit}-come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `It caught him. He got free, and ran away with it. It caught him again. It brought him back.'
\z

\ea\label{ex:06-a19}
He kakame, hinaka hisiri itimetemone amake hisiri.\\
\gll hee ka-ka-me hinaka hisiri iti-hemete-mone ama-ke hisiri\\
     3\textsc{sg}.\textsc{o} \textsc{comit}-come-back.\textsc{m} 3\textsc{sg}.\textsc{poss} basket(\textsc{f}) pick\_up-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f} basket(\textsc{f})\\
\glt `It brought him back. It picked up its basket.'
\z

\ea\label{ex:06-a20}
Hisiri ya hiberise wete hikane, wete hikane hiwa hiwa katomarematamonaka.\\
\gll hisiri ya hi-ibe-rise wete hi-ka-ne wete hi-ka-ne hiwa hiwa katoma-ra-himata-mona-ka\\
     basket(\textsc{f}) \textsc{adju} \textsc{oc}-put\_inside-down.\textsc{m} tie \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} tie \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} 3\textsc{sg}.\textsc{refl}.\textsc{m} 3\textsc{sg}.\textsc{refl}.\textsc{m} damage-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `It put him in the basket, and tied him in. He didn't make a fuss.'
\z

\ea\label{ex:06-a21}
Hiwa katomare wete hikanematamonaka.\\
\gll hiwa katoma-re wete hi-ka-na-himata-mona-ka\\
     3\textsc{sg}.\textsc{refl}.\textsc{m} damage-\textsc{neg}.\textsc{m} tie \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `He didn't make a fuss, and it tied him in.'
\z

\newpage
\ea\label{ex:06-a22}
Wete hikane, Oof, haa, koro okofimataba.\\
\gll wete hi-ka-ne oof haa koro o-ka-fi-mata-ba\\
     tie \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} whew (expression\_of\_exhaustion) throw 1\textsc{sg}.\textsc{s}-\textsc{comit}-water-short\_time-\textsc{immed}.\textsc{m}\\
\glt `It tied him in. ``Whew, I'm going to throw him in the water.'''
\z

\ea\label{ex:06-a23}
Tokomisematamona wasina ya ahi.\\
\gll to-ka-misa-himata-mona wasina ya ahi\\
     away-go-up-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} kapok(\textsc{f}) \textsc{adju} then\\
\glt `It went up into the kapok tree.'
\z

\ea\label{ex:06-a24}
Wasina wa ya tokomise, tokomise Ho-o, ho-o, ho-o, ho-o, ho-o, ho-o weye hikanematamona tati ya.\\
\gll wasina waa ya to-ka-mise to-ka-mise ho\_o ho\_o ho\_o ho\_o ho\_o ho\_o weye hi-ka-na-himata-mona tati ya\\
     kapok(\textsc{f}) stand.\textsc{f} \textsc{adju} away-go-up.\textsc{m} away-go-up.\textsc{m} (expression\_of\_exertion) (expression\_of\_exertion) (expression\_of\_exertion) (expression\_of\_exertion) (expression\_of\_exertion) (expression\_of\_exertion) carry \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} head \textsc{adju}\\
\glt `It went up the kapok tree. It went up. ``Ho-o, ho-o, ho-o.'' It was carrying him attached by its head.'
\z

\ea\label{ex:06-a25}
Weye hikane, kobo hikamise wasina ini batori ya, tokomisawawite fa fa kaya nemetemoneke ahi.\\
\gll weye hi-ka-ne kobo hi-ka-mise wasina ini batori ya to-ka-misa-waha-wite faha faha kaya na-hemete-mone-ke ahi\\
     carry \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} arrive \textsc{oc}-\textsc{comit}-up.\textsc{m} kapok(\textsc{f}) branch.\textsc{f} junction \textsc{adju} away-go-up-change-out water(\textsc{f}) water(\textsc{f}) stand \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} there\\
\glt `It carried him. It arrived up high with him. At the place where there was a branch in the kapok tree, it went around. There was a lake.'
\z

\ea\label{ex:06-a26}
Fa kaya na wasina warebasa, wasina warebasa, tokomisawite bere hikawitematamonaka wasina mani yai tona mani bari ya.\\
\gll faha kaya na wasina waa-re-basa wasina waa-ri-basa to-ka-misa-witi bere hi-ka-wite-himata-mona-ka wasina mani yai to-na mani bari ya\\
     water(\textsc{f}) stand \textsc{aux}.\textsc{f} kapok(\textsc{f}) stand-raised\_surface-edge.\textsc{f} kapok(\textsc{f}) stand-raised\_surface-edge.\textsc{f} away-go-up-out be\_across \textsc{oc}-\textsc{comit}-out-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} kapok(\textsc{f}) branch.\textsc{f} be\_straight \textsc{ch}-\textsc{aux}.\textsc{f} branch.\textsc{f} outer\_surface \textsc{adju}\\
\glt `The kapok tree was on the edge of a lake. It went up the kapok tree, and took him out along a straight limb of the kapok tree, on top of the limb.'
\z

\ea\label{ex:06-a27}
Weye kani hike, weye kani hikematamonaka.\\
\gll weye ka-ni hi-ke weye ka-ni hi-ka-himata-mona-ka\\
     carry \textsc{comit}-\textsc{aux}.\textsc{nfin} \textsc{oc}-go.\textsc{m} carry \textsc{comit}-\textsc{aux}.\textsc{nfin} \textsc{oc}-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `It went carrying him in the basket.'\footnote{The reason the O-construction prefix \textit{hi-} occurs on the intransitive verb \textit{toka} is that there is control, and the \textit{hi-} is transferred from the subordinate clause to the matrix clause (cf. \cite[42f.]{Vogel:2015az}).}
\z

\ea\label{ex:06-a28}
Weye kani hike, Hima, koro okofimataba.\\
\gll weye ka-ni hi-ke hima koro o-ka-fi-mata-ba\\
     carry \textsc{comit}-\textsc{aux}.\textsc{nfin} \textsc{oc}-go.\textsc{m} let's\_go throw 1\textsc{sg}.\textsc{s}-\textsc{comit}-water-short\_time-\textsc{immed}.\textsc{m}\\
\glt `It went carrying him in the basket. ``All right, I will throw him in the water with the basket.'''
\z

\ea\label{ex:06-a29}
Wene hikanematamonaka ahi.\\
\gll wene hi-ka-na-himata-mona-ka ahi\\
     wag \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `It swung him back and forth.'
\z

\newpage
\ea\label{ex:06-a30}
Wene hikane, Koro, fa bai kabokebona, isss sssss towakamematamonaka ahi.\\
\gll wene hi-ka-ne koro faha bai ka-boka-hibona to-ka-ka-ma-himata-mona-ka ahi\\
     wag \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} (sound\_of\_something\_being\_thrown) water(\textsc{f}) middle \textsc{comit}-enter\_water-\textsc{int}.\textsc{m} away-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `It swung him back and forth, and threw him. Instead of landing in the water, he went back.'
\z

\ea\label{ex:06-a31}
Towakame, tihi tokaname, Omm, kabokebona one, hiba.\\
\gll to-ka-ka-me tihi to-ka-na-me omm ka-boka-hibona o-ne hiba\\
     away-\textsc{comit}-go-back.\textsc{m} hit away-\textsc{comit}-\textsc{aux}-back.\textsc{m} humph \textsc{comit}-enter\_water-\textsc{int}.\textsc{m} 1\textsc{sg}.\textsc{s}-say.\textsc{m} wait\\
\glt `He went back, and landed there. ``Humph, I wanted him to go in the water, wait.'''
\z

\ea\label{ex:06-a32}
Kisamake kisamake warewame yama hoti howa toni ya sonehimatamonaka ahi karii beo.\\
\gll ka-risa-make ka-risa-make waa-re-waha-me yama hoti howa to-ni ya sona-himata-mona-ka ahi karii beo\\
     come-down-following come-down-following stand-raised\_surface-change-back.\textsc{m} thing(\textsc{f}) hole.\textsc{f} be\_open \textsc{ch}-\textsc{aux}.\textsc{nom}.\textsc{f} \textsc{adju} fall-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then (sound\_of\_going\_down) (sound\_of\_hitting\_ground)\\
\glt `It went down. It came back on the branch. It went down through an open place in the branches, \textit{karii}, \textit{beo}.'
\z

\ea\label{ex:06-a33}
Faya hiwakitimake, hiwakitimakematamonaka.\\
\gll faya hi-ka-iti-make hi-ka-iti-make-himata-mona-ka\\
     so \textsc{oc}-\textsc{comit}-pick\_up-following \textsc{oc}-\textsc{comit}-take\_out-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `It got him again.'
\z

\ea\label{ex:06-a34}
Hiwakitimake, hiwakitimake, Hee.\\
\gll hi-ka-iti-make hi-ka-iti-make hee\\
     \textsc{oc}-\textsc{comit}-pick\_up-following \textsc{oc}-\textsc{comit}-pick\_up-following hey\\
\glt `It got him again. ``Help.'''
\z

\ea\label{ex:06-a35}
Oo ha rima tinahi.\\
\gll oo haa rima ti-na-hi\\
     oh\_no call \textsc{neg}.\textsc{imp}.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{f}\\
\glt `{``}Hey, don't call.'''
\z

\ea\label{ex:06-a36}
Yama abara awineni, hee.\\
\gll yama ahaba-ra awine-ni hee\\
     thing(\textsc{f}) end-\textsc{neg}.\textsc{f} seem.\textsc{f}-\textsc{bkg}.\textsc{f} hey\\
\glt `{``}There is no one to help me, help.'''
\z

\ea\label{ex:06-a37}
Mitarama ni ha ne ha ni kasiromona ha ni kasiromonamako noko mi hinematamonaka ahi.\\
\gll mita-rama ni haa ne haa ni kasiro-himona haa ni kasiro-mona-mako noko mii hi-na-himata-mona-ka ahi\\
     hear-unusual \textsc{aux}.\textsc{nfin} call \textsc{aux}.\textsc{m} call \textsc{aux}.\textsc{nfin} persist-\textsc{rep}.\textsc{m} call \textsc{aux}.\textsc{nfin} persist-\textsc{rep}.\textsc{m}-for\_this\_reason.\textsc{m} face.\textsc{m} defecate\_on \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `But he kept on calling out. Because he stubbornly kept on calling out, it defecated on his face.'
\z

\ea\label{ex:06-a38}
Noko iyo kowewe kihakiha.\\
\gll noko iyo kowewe kiha-kiha\\
     face.\textsc{m} feces(\textsc{f}) diarrhea(\textsc{f}) \textsc{dup}-have\\
\glt `His face had diarrhea on it.'
\z

\ea\label{ex:06-a39}
Noko iyo kihematamona.\\
\gll noko iyo kiha-himata-mona\\
     face.\textsc{m} feces(\textsc{f}) have-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
\glt `His face had feces on it.'
\z

\newpage
\ea\label{ex:06-a40}
Noko mi hine, noko mi hine hiwakimisamatasematamona.\\
\gll noko mii hi-ne noko mii hi-ne hi-ka-ka-misa-ma-tasa-himata-mona\\
     face.\textsc{m} defecate\_on \textsc{oc}-\textsc{aux}.\textsc{m} face.\textsc{m} defecate\_on \textsc{oc}-\textsc{aux}.\textsc{m} \textsc{oc}-\textsc{comit}-go-up-back-again-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
\glt `It defecated on his face. It took him back up in the basket.'
\z

\ea\label{ex:06-a41}
Hiwakimisamatase, hiwakimisamematamona.\\
\gll hi-ka-ka-misa-ma-tase hi-ka-ka-misa-ma-himata-mona\\
     \textsc{oc}-\textsc{comit}-go-up-back-again.\textsc{m} \textsc{oc}-\textsc{comit}-go-up-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
\glt `It took him back up in the basket.'
\z

\ea\label{ex:06-a42}
Hiwakimisame Ho-o ho-o, ho-o, ho-o ati tokomisamematamona.\\
\gll hi-ka-ka-misa-me ho\_o ho\_o ho\_o ho\_o ati to-ka-misa-ma-himata-mona\\
     \textsc{oc}-\textsc{comit}-go-up-back.\textsc{m} (expression\_of\_exertion) (expression\_of\_exertion) (expression\_of\_exertion) (expression\_of\_exertion) voice away-go-up-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
\glt `It took him back up. ``Ho-o, ho-o, ho-o, ho-o.'' Its voice could be heard as it went back up.'\footnote{The pronunciation \textit{hiwakimisame} is an archaism, younger Jarawaras say [heekimisame]. That is, the younger speakers apply a morphophonological rule that Siko does not apply (\cite[37]{Dixon:2004xr}). See also \textit{hiwakame} in sentence (\ref{ex:06-a99}).}
\z

\ea\label{ex:06-a43}
Ati tokomisame ati tokomisame, ati tokomisame bori tomisame Oof.\\
\gll ati to-ka-misa-me ati to-ka-misa-me ati to-ka-misa-me bori to-misa-me oof\\
     voice away-go-up-back.\textsc{m} voice away-go-up-back.\textsc{m} voice away-go-up-back.\textsc{m} get\_close away-up-back.\textsc{m} hoo\\
\glt `Its voice could be heard as it went back up. It was close. ``Whew.'''
\z

\ea\label{ex:06-a44}
Weye kani forima hine bere hikawitematamonaka.\\
\gll weye ka-ni forima hi-ne bere hi-ka-wite-himata-mona-ka\\
     carry \textsc{comit}-\textsc{aux}.\textsc{nfin} do\_properly \textsc{oc}-\textsc{aux}.\textsc{m} be\_across \textsc{oc}-\textsc{comit}-out-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `It straightened the basket so it would not fall. It went along the limb with him.'
\z

\ea\label{ex:06-a45}
Bere hikawite, bere hikawitematamonaka.\\
\gll bere hi-ka-wite bere hi-ka-wite-himata-mona-ka\\
     be\_across \textsc{oc}-\textsc{comit}-out be\_across \textsc{oc}-\textsc{comit}-out-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `It went along the limb with him.'
\z

\ea\label{ex:06-a46}
Bere hikawite wati hikane tawi hikanematamonaka.\\
\gll bere hi-ka-wite wati hi-ka-ne tawi hi-ka-na-himata-mona-ka\\
     be\_across \textsc{oc}-\textsc{comit}-out plan\_against \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} spin \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `It went along the limb with him. It planned against him. It swung him back and forth.'
\z

\ea\label{ex:06-a47}
Tawi hikane, koro.\\
\gll tawi hi-ka-ne koro\\
     spin \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} (sound\_of\_something\_being\_thrown)\\
\glt `It swung him back and forth, and threw him.'
\z

\ea\label{ex:06-a48}
Faya, towakematamonaka ahi kabokabakani.\\
\gll faya to-ka-ka-himata-mona-ka ahi ka-boka-ba-kani\\
     so away-\textsc{comit}-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then \textsc{comit}-enter\_water-\textsc{fut}-\textsc{cntrfact}\\
\glt `But he went back, even though the monster thought he was going to go in the water.'
\z

\ea\label{ex:06-a49}
Fa fa baki yokana iss sss towakame tihi tokanamematamonaka.\\
\gll faha faha baki yokana to-ka-ka-me tihi to-ka-na-ma-himata-mona-ka\\
     water(\textsc{f}) water(\textsc{f}) inside\_surface.\textsc{f} true away-\textsc{comit}-go-back.\textsc{m} hit away-\textsc{comit}-\textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `He almost touched the surface of the water as he went back. He landed back there.'
\z

\ea\label{ex:06-a50}
Aa, kabokebona one, kabokebona one, kakasi raba reba ama narani, faha ihi toehibona onahari.\\
\gll aa ka-boka-hibona o-ne ka-boka-hibona o-ne ka-kasi raba ra-hiba ama na-rani faha ihi to-ha-hibona o-na-haari\\
     (expression\_of\_displeasure) \textsc{comit}-enter\_water-\textsc{int}.\textsc{m} 1\textsc{sg}.\textsc{s}-say.\textsc{m} \textsc{comit}-enter\_water-\textsc{int}.\textsc{m} 1\textsc{sg}.\textsc{s}-say.\textsc{m} \textsc{dup}-be\_naughty a\_bit \textsc{neg}-\textsc{fut}.\textsc{m} \textsc{sec} \textsc{aux}.\textsc{m}-? water(\textsc{f}) victim\_of.\textsc{f} \textsc{ch}-be-\textsc{int}.\textsc{m} 1\textsc{sg}.\textsc{s}-say-\textsc{ip}.\textsc{e}.\textsc{m}\\
\glt `{``}\textit{Aa}, I wanted him to go in the water. He is naughty. I wanted him to drown.'''
\z

\ea\label{ex:06-a51}
Bere name bere name sonematamonaka ahi.\\
\gll bere na-me bere na-me sona-himata-mona-ka ahi\\
     be\_across \textsc{aux}-back.\textsc{m} be\_across \textsc{aux}-back.\textsc{m} fall-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `It came back on the limb, and went down.'
\z

\ea\label{ex:06-a52}
Sonematamonaka ahi baasa beo.\\
\gll sona-himata-mona-ka ahi baasa beo\\
     fall-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then (sound\_of\_hitting\_leaves) (sound\_of\_hitting\_ground)\\
\glt `It went down, \textit{baasa}, \textit{beo}.'
\z

\ea\label{ex:06-a53}
Faya ha ne Ha nareka.\\
\gll faya haa ne haa na-hare-ka\\
     so call \textsc{aux}.\textsc{m} call \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `He called out. ``He is calling out.'''
\z

\ea\label{ex:06-a54}
Yama abara awineni, hee.\\
\gll yama ahaba-ra awine-ni hee\\
     thing(\textsc{f}) die-\textsc{neg}.\textsc{f} seem.\textsc{f}-\textsc{bkg}.\textsc{f} hey\\
\glt `{``}There is no one to help me, help.'''
\z

\ea\label{ex:06-a55}
Ha rima tinahi.\\
\gll haa rima ti-na-hi\\
     call \textsc{neg}.\textsc{imp}.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{f}\\
\glt `{``}Don't call out.'''
\z

\ea\label{ex:06-a56}
Hiwakitimake, noko mi hinemata.\\
\gll hi-ka-iti-make noko mii hi-na-himata\\
     \textsc{oc}-\textsc{comit}-pick\_up-following face.\textsc{m} defecate\_on \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}\\
\glt `It came and got him. It defecated on his face.'
\z

\ea\label{ex:06-a57}
Ha ni kasiromona noko mi hinematamonaka ahi iyo kowewe ya.\\
\gll haa ni kasiro-himona noko mii hi-na-himata-mona-ka ahi iyo kowewe ya\\
     call \textsc{aux}.\textsc{nfin} persist-\textsc{rep}.\textsc{m} face.\textsc{m} defecate\_on \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then feces(\textsc{f}) diarrhea(\textsc{f}) \textsc{adju}\\
\glt `Because he said he wouldn't stop calling out, it defecated diarrhea on his face.'
\z

\ea\label{ex:06-a58}
Faya noko mi hine, noko mi hine hiwakimisame, Oof, hoo hiba kabokaba nena.\\
\gll faya noko mii hi-ne noko mii hi-ne hi-ka-ka-misa-me oof hoo hiba ka-boka-ba na-hina\\
     so face.\textsc{m} defecate\_on \textsc{oc}-\textsc{aux}.\textsc{m} face.\textsc{m} defecate\_on \textsc{oc}-\textsc{aux}.\textsc{m} \textsc{oc}-\textsc{comit}-go-up-back.\textsc{m} hoo humph wait \textsc{comit}-enter\_water.\textsc{m}-\textsc{fut} \textsc{aux}-\textsc{irr}.\textsc{m}\\
\glt `It defecated on his face. It took him back up. ``Oof, hoo, he should have gone into the water.'''
\z

\ea\label{ex:06-a59}
Hiwa aawa re.\\
\gll hiwa a-awa re\\
     3\textsc{sg}.\textsc{refl}.\textsc{m} \textsc{dup}-see \textsc{neg}.\textsc{m}\\
\glt `{``}Why didn't he?'''
\z

\ea\label{ex:06-a60}
Hiyamata hiri nahi?\\
\gll hiyamata hiri nahi\\
     what do \textsc{aux}.\textsc{m}\\
\glt `{``}What did he do?'''
\z

\ea\label{ex:06-a61}
Hiwakimisamematamona fahi.\\
\gll hi-ka-ka-misa-ma-himata-mona fahi\\
     \textsc{oc}-\textsc{comit}-go-up-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
\glt `It took him back up.'
\z

\ea\label{ex:06-a62}
Hiwakimisame, hiwakimisame, tokomisamematamona.\\
\gll hi-ka-ka-misa-me hi-ka-ka-misa-me to-ka-misa-ma-himata-mona\\
     \textsc{oc}-\textsc{comit}-go-up-back.\textsc{m} \textsc{oc}-\textsc{comit}-go-up-back.\textsc{m} away-go-up-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
\glt `It took him back up. It went back up.'
\z

\ea\label{ex:06-a63}
Tokomisama tokomisame ya, tokomisama tokomisame ya bori tomisama, faya bere hikawitemata fahi.\\
\gll to-ka-misa-ma to-ka-misa-me ya to-ka-misa-ma to-ka-misa-me ya bori to-misa-ma faya bere hi-ka-wite-himata fahi\\
     away-go-up-back.\textsc{list} away-go-up-back.\textsc{nom}.\textsc{m} \textsc{adju} away-go-up-back.\textsc{list} away-go-up-back.\textsc{nom}.\textsc{m} \textsc{adju} get\_close away-up-back.\textsc{list} so be\_across \textsc{oc}-\textsc{comit}-out-\textsc{fp}.\textsc{n}.\textsc{m} then\\
\glt `It went back up. It got close to the top. Then it took him out on the limb.'
\z

\ea\label{ex:06-a64}
Bere hikanamatase, bere hikaname kobo tokaname ya tawi hikanematamonaka ahi.\\
\gll bere hi-ka-na-ma-tase bere hi-ka-na-me kobo to-ka-na-me ya tawi hi-ka-na-himata-mona-ka ahi\\
     be\_across \textsc{oc}-\textsc{comit}-\textsc{aux}-back-again.\textsc{m} be\_across \textsc{oc}-\textsc{comit}-\textsc{aux}-back.\textsc{m} arrive away-\textsc{comit}-\textsc{aux}-back.\textsc{nom}.\textsc{m} \textsc{adju} spin \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `It took him out on the limb again. When it arrived, it swung him back and forth.'
\z

\ea\label{ex:06-a65}
Tawi hikanematamona.\\
\gll tawi hi-ka-na-himata-mona\\
     spin \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
\glt `It swung him back and forth.'
\z

\ea\label{ex:06-a66}
Tawi hikane, koro towake, towake kabokebona towakamatasematamonaka ahi sissss sssi.\\
\gll tawi hi-ka-ne koro to-ka-ke to-ka-ke ka-boka-hibona to-ka-ka-ma-tasa-himata-mona-ka ahi sissss sssi\\
     spin \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} (sound\_of\_something\_being\_thrown) away-\textsc{comit}-go.\textsc{m} away-\textsc{comit}-go.\textsc{m} \textsc{comit}-enter\_water-\textsc{int}.\textsc{m} away-\textsc{comit}-go-back-again-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then (sound\_of\_water) (sound\_of\_water)\\
\glt `It swung him back and forth. \textit{Koro}, he went out. He didn't fall in the water, but went along the surface, \textit{sissss}, \textit{sssi}.'
\z

\ea\label{ex:06-a67}
Fara fara wasina wai ya totokowitamematamonaha.\\
\gll fara fara wasina wai ya to-to-ka-ita-ma-himata-mona-ha\\
     same.\textsc{f} same.\textsc{f} kapok(\textsc{f}) stand.\textsc{nom} \textsc{adju} \textsc{dup}-away-\textsc{comit}-sit-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
\glt `He landed and sat again where the kapok tree was.'
\z

\ea\label{ex:06-a68}
Aa, kakasi raba nahare.\\
\gll aa ka-kasi raba na-hare\\
     (expression\_of\_displeasure) \textsc{dup}-be\_naughty a\_bit \textsc{aux}-\textsc{neg}.\textsc{m}\\
\glt `{``}\textit{Aa}, there he goes again.'''
\z

\ea\label{ex:06-a69}
Kame bere name sonatasematamonaka fahi.\\
\gll ka-me bere na-me sona-tasa-himata-mona-ka fahi\\
     come-back.\textsc{m} be\_across \textsc{aux}-back.\textsc{m} fall-again-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `It came back along the branch again, and came down again.'
\z

\ea\label{ex:06-a70}
Sone, me inamati narabi hawihani, me me inamati inamati narabo hawihino me me famemetemoneke me narabi hawihani mati.\\
\gll sone me\_inamati narabi hawi-hani me me\_inamati inamati narabo hawi-hino me me fama-hemete-mone-ke me narabi hawi-hani mati\\
     fall.\textsc{m} spirits(\textsc{f}) ear.\textsc{f} work-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl} spirits(\textsc{f}) spirit(\textsc{m}) ear.\textsc{m} work-\textsc{ip}.\textsc{n}.\textsc{m} 3\textsc{pl} 3\textsc{pl}.\textsc{s} be\_two-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} 3\textsc{pl}.\textsc{poss} ear.\textsc{f} work-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{poss}\\
\glt `It came down. The spirits heard something. The spirit heard something. There were two of them that heard something.'
\z

\ea\label{ex:06-a71}
Inamati ama narani one ha ne inamati were kanareka owahari.\\
\gll inamati ama na-rani o-ne haa ne inamati were ka-na-hare-ka o-awa-haari\\
     someone(\textsc{m}) be \textsc{aux}.\textsc{m}-? 1\textsc{sg}.\textsc{s}-say.\textsc{m} call \textsc{aux}.\textsc{m} spirit(\textsc{m}) throw \textsc{comit}-\textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m} 1\textsc{sg}.\textsc{s}-see-\textsc{ip}.\textsc{e}.\textsc{m}\\
\glt `{``}The monster threw someone in something who was calling, I saw him.'''
\z

\ea\label{ex:06-a72}
Hari hiwakimisamatase.\\
\gll haari hi-ka-ka-misa-ma-tase\\
     that\_one.\textsc{m} \textsc{oc}-\textsc{comit}-go-up-back-again.\textsc{m}\\
\glt `{``}There, it took him up again.'''
\z

\ea\label{ex:06-a73}
Hibare mata awaraba timata mata.\\
\gll hibare mata awaraba ti-mata mata\\
     wait for\_now leave\_something\_alone 2\textsc{sg}.\textsc{s}-short\_time for\_now\\
\glt `{``}Wait, don't do anything yet.'''
\z

\ea\label{ex:06-a74}
E kakatomamata.\\
\gll ee ka-katoma-mata\\
     1\textsc{in}.\textsc{s} \textsc{comit}-look\_at-short\_time\\
\glt `{``}Let's watch him.'''
\z

\ea\label{ex:06-a75}
Faya bere hikaname, bere hikaname Tekatoma.\\
\gll faya bere hi-ka-na-me bere hi-ka-na-me ti-ka-katoma\\
     so be\_across \textsc{oc}-\textsc{comit}-\textsc{aux}-back.\textsc{m} be\_across \textsc{oc}-\textsc{comit}-\textsc{aux}-back.\textsc{m} 2\textsc{sg}.\textsc{s}-\textsc{comit}-look\_at\\
\glt `It took him out on the limb. ``Look at him.'''
\z

\ea\label{ex:06-a76}
Faya koro hikanamatasematamonaka ahi.\\
\gll faya koro hi-ka-na-ma-tasa-himata-mona-ka ahi\\
     so throw \textsc{oc}-\textsc{comit}-\textsc{aux}-back-again-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `It threw him again.'
\z

\ea\label{ex:06-a77}
Koro hikaname kabokebona kakamatasematamonaka ahi isissss.\\
\gll koro hi-ka-na-me ka-boka-hibona ka-ka-ma-tasa-himata-mona-ka ahi isissss\\
     throw \textsc{oc}-\textsc{comit}-\textsc{aux}-back.\textsc{m} \textsc{comit}-enter\_water-\textsc{int}.\textsc{m} \textsc{dup}-come-back-again-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then (sound\_of\_water)\\
\glt `It threw him. He didn't fall into the water, but went back again, \textit{isissss}.'
\z

\ea\label{ex:06-a78}
Tihi tokanama tihi tokaname ya Owati tikasawariara tini.\\
\gll tihi to-ka-na-ma tihi to-ka-na-me ya o-ati ti-ka-sawari-hara ti-ni\\
     hit away-\textsc{comit}-\textsc{aux}-back.\textsc{list} hit away-\textsc{comit}-\textsc{aux}-back.\textsc{nom}.\textsc{m} \textsc{adju} 1\textsc{sg}.\textsc{poss}-voice 2\textsc{sg}.\textsc{s}-\textsc{comit}-frustrate-\textsc{ip}.\textsc{e}.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
\glt `He landed upright. ``I told you.'''
\z

\ea\label{ex:06-a79}
Inamati, inamati were kane amane, hiba.\\
\gll inamati inamati were ka-ne ama-ne hiba\\
     someone(\textsc{m}) spirit(\textsc{m}) throw \textsc{comit}-\textsc{aux}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m} wait\\
\glt `{``}The monster threw someone in something.'''
\z

\ea\label{ex:06-a80}
Bao nisematamona.\\
\gll bao na-risa-himata-mona\\
     come\_down \textsc{aux}-down-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
\glt `It jumped down.'
\z

\ea\label{ex:06-a81}
Bao nisi me hiwe, fare sosona yokana ya, sosona yokana ya wabise yoro naba me kakiemetemoneni.\\
\gll bao na-risi me hi-awe fare sosona yokana ya sosona yokana ya wabise yoro na-haba me ka-ka-ki-hemete-mone-ni\\
     come\_down \textsc{aux}-down.\textsc{nfin} 3\textsc{pl}.\textsc{s} \textsc{oc}-see.\textsc{m} same.\textsc{m} place\_of\_falling true \textsc{adju} place\_of\_falling true \textsc{adju} spear(\textsc{f}) stand \textsc{aux}-\textsc{fut}.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{comit}-come-coming-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
\glt `They saw it coming down. Two spears would stand right where it came down. The two of them came.'
\z

\ea\label{ex:06-a82}
Me kakiemetemoneke.\\
\gll me ka-ka-ki-hemete-mone-ke\\
     3\textsc{pl}.\textsc{s} \textsc{comit}-come-coming-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `They came.'
\z

\ea\label{ex:06-a83}
Inamati, inamati were kanaka, me, ati na me me, Hiya.\\
\gll inamati inamati were ka-na-ka me ati na me me hiya\\
     spirit(\textsc{m}) spirit(\textsc{m}) throw \textsc{comit}-\textsc{aux}-\textsc{decl}.\textsc{m} 3\textsc{pl}.\textsc{s} say \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl} come\_on\\
\glt `{``}The monster threw someone in something,'' they said. ``Here.'''
\z

\ea\label{ex:06-a84}
Weye hikamisa weye hikamisame ya hiwakimisama, hiwakimisamematamonaka ahi.\\
\gll weye hi-ka-misa weye hi-ka-misa-me ya hi-ka-ka-misa-ma hi-ka-ka-misa-ma-himata-mona-ka ahi\\
     carry \textsc{oc}-\textsc{comit}-up.\textsc{list} carry \textsc{oc}-\textsc{comit}-up-back.\textsc{nom}.\textsc{m} \textsc{adju} \textsc{oc}-\textsc{comit}-go-up-back.\textsc{list} \textsc{oc}-\textsc{comit}-go-up-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `It put him on its back again in the basket. It went up with him.'
\z

\newpage
\ea\label{ex:06-a85}
Hiwakimisame, me kafotoke me, wabise me taba nisikimemetemone ahi tokanokomisaharo ahi.\\
\gll hi-ka-ka-misa-me me ka-foto-ke me wabise me taba na-risa-kima-hemete-mone ahi to-ka-noko-misa-haaro ahi\\
     \textsc{oc}-\textsc{comit}-go-up-back.\textsc{m} 3\textsc{pl}.\textsc{s} \textsc{comit}-appear-coming 3\textsc{pl}.\textsc{s} spear(\textsc{f}) 3\textsc{pl}.\textsc{s} stick\_in \textsc{aux}-down-two-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then away-\textsc{comit}-point-up-\textsc{ip}.\textsc{e}.\textsc{f} then\\
\glt `It took him up. The two of them appeared. They stuck two spears in the ground. They were standing upright.'
\z

\ea\label{ex:06-a86}
Wabise tokanokomisemetemone ama ahi ini famariaro ahi.\\
\gll wabise to-ka-noko-misa-hemete-mone ama ahi ini fama-ri-haaro ahi\\
     spear(\textsc{f}) away-\textsc{comit}-point-up-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec} then point.\textsc{f} be\_two-raised\_surface-\textsc{ip}.\textsc{e}.\textsc{f} then\\
\glt `The spears were pointing up, two together.'
\z

\ea\label{ex:06-a87}
Faya ini famaria me, wabise me taba nisikima me me yoro na me tokome bere hikawitematamonaka.\\
\gll faya ini fama-ria me wabise me taba na-risa-kima me me yoro na me to-ka-me bere hi-ka-wite-himata-mona-ka\\
     so point.\textsc{f} be\_two-raised\_surface.\textsc{f} 3\textsc{pl} spear(\textsc{f}) 3\textsc{pl}.\textsc{s} stick\_in \textsc{aux}-down-two.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} stand \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} away-go-back.\textsc{m} be\_across \textsc{oc}-\textsc{comit}-out-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `The points were together. They stuck the two spears in the ground. They stood there. It went and took him out on the limb.'
\z

\ea\label{ex:06-a88}
Bere hikawite, Kabokaba nena kakasi raba nahare, hiba.\\
\gll bere hi-ka-wite ka-boka-ba na-hina ka-kasi raba na-hare hiba\\
     be\_across \textsc{oc}-\textsc{comit}-out \textsc{comit}-enter\_water.\textsc{m}-\textsc{fut} \textsc{aux}-\textsc{irr}.\textsc{m} \textsc{dup}-be\_naughty a\_bit \textsc{aux}-\textsc{neg}.\textsc{m} wait\\
\glt `It took him out on the limb. ``I want him to go into the water now.'''
\z

\ea\label{ex:06-a89}
Tawi hikanematamonaka ahi.\\
\gll tawi hi-ka-na-himata-mona-ka ahi\\
     spin \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `It swung him back and forth.'
\z

\ea\label{ex:06-a90}
Tawi hikane koro, tefe nabakani, kakamakematamonaka ahi.\\
\gll tawi hi-ka-ne koro tefe na-ba-kani ka-ka-make-himata-mona-ka ahi\\
     spin \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} (sound\_of\_something\_being\_thrown) fall\_in\_water \textsc{aux}.\textsc{m}-\textsc{fut}-\textsc{cntrfact} \textsc{dup}-come-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `It swung him back and forth. \textit{Koro}, it thought he would go into the water, but he came back to the same place.'
\z

\ea\label{ex:06-a91}
Towakame sisssss kobo tokaname, Aa, kakasi raba nahare.\\
\gll to-ka-ka-me sisssss kobo to-ka-na-me aa ka-kasi raba na-hare\\
     away-\textsc{comit}-go-back.\textsc{m} (sound\_of\_water) arrive away-\textsc{comit}-\textsc{aux}-back.\textsc{m} (expression\_of\_displeasure) \textsc{dup}-be\_naughty a\_bit \textsc{aux}-\textsc{neg}.\textsc{m}\\
\glt `He went over the water, \textit{sisss}, and arrived again. ``\textit{Aa}, there he goes again.'''
\z

\ea\label{ex:06-a92}
Kamematamonaka.\\
\gll ka-ma-himata-mona-ka\\
     come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `It came back.'
\z

\ea\label{ex:06-a93}
Bere name bere name, fare sosona ya wabise yoro na rama ahi.\\
\gll bere na-me bere na-me fare sosona ya wabise yoro na rama ahi\\
     be\_across \textsc{aux}-back.\textsc{m} be\_across \textsc{aux}-back.\textsc{m} same.\textsc{m} place\_of\_falling \textsc{adju} spear(\textsc{f}) stand \textsc{aux}.\textsc{f} unusual then\\
\glt `It came back on the limb. The two spears were standing at the place where he always landed.'
\z

\ea\label{ex:06-a94}
Wabise yoro na me, kari, beo nebona wabise tasi hine kiya nematamonaka.\\
\gll wabise yoro na me kari beo na-hibona wabise tasi hi-ne kiya na-himata-mona-ka\\
     spear(\textsc{f}) stand \textsc{aux} 3\textsc{pl} (sound\_of\_dropping) (sound\_of\_hitting\_ground) \textsc{aux}-\textsc{int}.\textsc{m} spear(\textsc{f}) go\_through \textsc{oc}-\textsc{aux}.\textsc{m} yell \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `The two spears were upright. It dropped, and when it was going to hit the ground, the spears went through it, and it cried out.'
\z

\ea\label{ex:06-a95}
Aya aya aya, kiya kanematamonaka.\\
\gll aya aya aya kiya ka-na-himata-mona-ka\\
     (expression\_of\_pain) (expression\_of\_pain) (expression\_of\_pain) yell \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `{``}\textit{Aya}, \textit{aya}, \textit{aya}.'' It cried out.'
\z

\ea\label{ex:06-a96}
Toharisaba ne rama wabise mate ka ya ahi, sona raba namako.\\
\gll to-iha-risa-baa ne rama wabise mate ka ya ahi sona raba na-mako\\
     away-be\_located-down-at\_a\_distance \textsc{aux}.\textsc{m} unusual spear(\textsc{f}) base.\textsc{f} \textsc{loc} \textsc{adju} then fall a\_bit \textsc{aux}-for\_this\_reason.\textsc{m}\\
\glt `It went right down to the hilts of the spears, because it fell hard.'
\z

\ea\label{ex:06-a97}
Faya kiya kane, me fito hikanamaki me hinabowame, me hinabowematamonaka ahi.\\
\gll faya kiya ka-ne me fito hi-ka-na-maki me hi-naabowa-me me hi-naabowa-himata-mona-ka ahi\\
     so yell \textsc{comit}-\textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{s} rush \textsc{oc}-\textsc{comit}-\textsc{aux}-following.\textsc{m} 3\textsc{pl}.\textsc{s} \textsc{oc}-kill-back.\textsc{m} 3\textsc{pl}.\textsc{s} \textsc{oc}-kill-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
\glt `It cried out. They rushed it, and finished it off. They killed it.'
\z

\ea\label{ex:06-a98}
Ahabe ya, abematamonaka.\\
\gll ahabe ya ahaba-himata-mona-ka\\
     die.\textsc{nom}.\textsc{m} \textsc{adju} die-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `It died.'
\z

\ea\label{ex:06-a99}
Abe ya me, me hekitimake me hiwakame, Hee, himata ebe tinahani?\\
\gll ahabe ya me me hi-ka-iti-make me hi-ka-ka-me hee himata ebe ti-na-hani\\
     die.\textsc{nom}.\textsc{m} \textsc{adju} 3\textsc{pl} 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-pick\_up-following 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-come-back.\textsc{m} hey what happen\_to 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{f}\\
\glt `When it died, they went and got him, and brought him back. ``Hey, what happened to you?'''
\z

\ea\label{ex:06-a100}
Tiwa wateino ama?\\
\gll tiwa wata-hino ama\\
     2\textsc{sg}.\textsc{o} catch-\textsc{ip}.\textsc{n}.\textsc{m} \textsc{sec}\\
\glt `{``}Did it capture you?'''
\z

\ea\label{ex:06-a101}
Owa wate amaka ahi.\\
\gll owa wate ama-ka ahi\\
     1\textsc{sg}.\textsc{o} catch.\textsc{m} \textsc{sec}-\textsc{decl}.\textsc{m} then\\
\glt `{``}It captured me.'''
\z

\ea\label{ex:06-a102}
Waino ama ahi.\\
\gll wai-no ama ahi\\
     stand.\textsc{nom}-\textsc{ip}.\textsc{n}.\textsc{m} be there\\
\glt `{``}It was there on the ground.'''
\z

\ea\label{ex:06-a103}
Waino ya owa were kanaka ahi.\\
\gll wai-no ya owa were ka-na-ka ahi\\
     stand.\textsc{nom}-\textsc{ip}.\textsc{n}.\textsc{m} \textsc{adju} 1\textsc{sg}.\textsc{o} throw \textsc{comit}-\textsc{aux}-\textsc{decl}.\textsc{m} then\\
\glt `{``}It threw me from where it was.'''
\z

\ea\label{ex:06-a104}
Waino ama ahi, mase hi ni tohahari.\\
\gll wai-no ama ahi mase hii ni to-ha-haari\\
     stand.\textsc{nom}-\textsc{ip}.\textsc{n}.\textsc{m} be there curassow(\textsc{m}) call \textsc{aux}.\textsc{nfin} \textsc{ch}-be-\textsc{ip}.\textsc{e}.\textsc{m}\\
\glt `{``}It was there making the call of a curassow.'''
\z

\ea\label{ex:06-a105}
Mase hi ni toenoka.\\
\gll mase hii ni to-ha-hino-ka\\
     curassow(\textsc{m}) call \textsc{aux}.\textsc{nfin} \textsc{ch}-be-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `{``}It was making the call of a curassow.'''
\z

\ea\label{ex:06-a106}
Owareno obatasi ka ya owa watareka.\\
\gll o-awa-ra-hino o-batasi ka ya owa wata-hare-ka\\
     1\textsc{sg}.\textsc{s}-see-\textsc{neg}-\textsc{ip}.\textsc{n}.\textsc{m} 1\textsc{sg}.\textsc{poss}-middle\_of\_back \textsc{loc} \textsc{adju} 1\textsc{sg}.\textsc{o} catch-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `{``}I didn't see it, and it captured me from behind.'''
\z

\ea\label{ex:06-a107}
Faya me, me, Homahari ha.\\
\gll faya me me homa-haari haa\\
     so 3\textsc{pl} 3\textsc{pl} lie\_on\_ground-look.\textsc{m} \textsc{dem}\\
\glt `{``}There it is, lying on the ground.'''
\z

\ea\label{ex:06-a108}
Tiwa ha.\\
\gll ti-awa haa\\
     2\textsc{sg}.\textsc{s}-see \textsc{dem}\\
\glt `{``}Look at it.'''
\z

\ea\label{ex:06-a109}
Homakane.\\
\gll homa-kane\\
     lie\_on\_ground-\textsc{mood}.\textsc{m}\\
\glt `{``}Look at it lying there.'''
\z

\ea\label{ex:06-a110}
Kose mati amoamosa nematamonaha.\\
\gll kose\_mati amo-amosa na-himata-mona-ha\\
     name\_of\_monster(\textsc{m}) \textsc{dup}-be\_good \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
\glt `{``}The \textit{kose mati} is huge.'''
\z

\ea\label{ex:06-a111}
Ota kabebanaka.\\
\gll otaa kaba-hibana-ka\\
     1\textsc{ex}.\textsc{o} eat-\textsc{fut}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `{``}We will eat it.'''
\z

\ea\label{ex:06-a112}
Faya he me he me katisemetemoneke fahi.\\
\gll faya hee me hee me ka-tisa-hemete-mone-ke fahi\\
     so 3\textsc{sg}.\textsc{o} 3\textsc{pl} 3\textsc{sg}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{comit}-undo-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
\glt `Then they untied him.'
\z

\ea\label{ex:06-a113}
He me katisa noko me hikatise, me fito hinihe Afi tinahi.\\
\gll hee me ka-tisa noko me hi-ka-tise me fito hi-nihe afi ti-na-hi\\
     3\textsc{sg}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{comit}-undo.\textsc{f} face.\textsc{m} 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-undo.\textsc{m} 3\textsc{pl}.\textsc{s} rush \textsc{oc}-\textsc{caus}.\textsc{m} bathe 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{f}\\
\glt `They untied him. They untied the basket over his face. They let him out. ``Go take a bath.'''
\z

\ea\label{ex:06-a114}
Timo kitake.\\
\gll ti-maho kita-ke\\
     2\textsc{sg}.\textsc{poss}-smell be\_strong-\textsc{decl}.\textsc{f}\\
\glt `{``}You smell bad.'''
\z

\ea\label{ex:06-a115}
Tiwa mi narene.\\
\gll tiwa mii na-hare-ne\\
     2\textsc{sg}.\textsc{o} defecate\_on \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
\glt `{``}It defecated on you.'''
\z

\newpage
\ea\label{ex:06-a116}
Afi nebona tokifiwame afi ne, afi nematamonaka.\\
\gll afi na-hibona to-ka-fi-waha-me afi ne afi na-himata-mona-ka\\
     bathe \textsc{aux}-\textsc{int}.\textsc{m} away-go-water-change-back.\textsc{m} bathe \textsc{aux}.\textsc{m} bathe \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `He went to the water to bathe. He bathed.'
\z

\ea\label{ex:06-a117}
Afi ne, afi ni, afi ni hawa toe ya kimisamematamona.\\
\gll afi ne afi ni afi ni hawa to-he ya ka-misa-ma-himata-mona\\
     bathe \textsc{aux}.\textsc{m} bathe \textsc{aux}.\textsc{nfin} bathe \textsc{aux}.\textsc{nfin} be\_finished \textsc{ch}-\textsc{aux}.\textsc{nom}.\textsc{m} \textsc{adju} come-up-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
\glt `He bathed. When he finished bathing, he came onto the shore.'
\z

\ea\label{ex:06-a118}
Kimisame, kimisamematamonaka.\\
\gll ka-misa-me ka-misa-ma-himata-mona-ka\\
     come-up-back.\textsc{m} come-up-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `He came onto the shore.'
\z

\ea\label{ex:06-a119}
Kimisame ya, Oof owa te kasomake.\\
\gll ka-misa-me ya oof owa tee ka-soma-ke\\
     come-up-back.\textsc{nom}.\textsc{m} \textsc{adju} hoo 1\textsc{sg}.\textsc{o} 2\textsc{pl}.\textsc{s} \textsc{comit}-help-\textsc{decl}.\textsc{f}\\
\glt `He came onto the shore. ``Hoo, you saved me.'''
\z

\ea\label{ex:06-a120}
Tika ha ni tika ha ni ota mita otake.\\
\gll ti-ka haa ni ti-ka haa ni otaa mita otaa-ke\\
     2\textsc{sg}.\textsc{poss}-\textsc{poss} call \textsc{aux}.\textsc{nfin} 2\textsc{sg}.\textsc{poss}-\textsc{poss} call \textsc{aux}.\textsc{nfin} 1\textsc{ex}.\textsc{s} hear.\textsc{f} 1\textsc{ex}.\textsc{s}-\textsc{decl}.\textsc{f}\\
\glt `{``}We heard you calling.'''
\z

\ea\label{ex:06-a121}
Tika ha ni ota mita otani.\\
\gll ti-ka haa ni otaa mita otaa-ni\\
     2\textsc{sg}.\textsc{poss}-\textsc{poss} call \textsc{aux}.\textsc{nfin} 1\textsc{ex}.\textsc{s} hear.\textsc{f} 1\textsc{ex}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
\glt `{``}We heard you calling.'''
\z

\ea\label{ex:06-a122}
Ha tina tika ha ni ota mitimakoni ota kakia, tiwa ota wasikineke.\\
\gll haa ti-na ti-ka haa ni otaa mita-makoni otaa ka-ka-kia tiwa otaa wasi-ki-ne-ke\\
     call 2\textsc{sg}.\textsc{s}-\textsc{aux}.\textsc{f} 2\textsc{sg}.\textsc{poss}-\textsc{poss} call \textsc{aux}.\textsc{nfin} 1\textsc{ex}.\textsc{s} hear-for\_this\_reason.\textsc{f} 1\textsc{ex}.\textsc{s} \textsc{comit}-come-coming.\textsc{f} 2\textsc{sg}.\textsc{o} 1\textsc{ex}.\textsc{s} find-coming-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `{``}When you called, we heard you calling, and for this reason we came and found you.'''
\z

\ea\label{ex:06-a123}
Tiwa ota wasia tiwa ota kasoma.\\
\gll tiwa otaa wasiha tiwa otaa ka-soma\\
     2\textsc{sg}.\textsc{o} 1\textsc{ex}.\textsc{s} find.\textsc{f} 2\textsc{sg}.\textsc{o} 1\textsc{ex}.\textsc{s} \textsc{comit}-help.\textsc{f}\\
\glt `{``}We found you, and saved you.'''
\z

\ea\label{ex:06-a124}
Tiwa kabene amake.\\
\gll tiwa kaba-hene ama-ke\\
     2\textsc{sg}.\textsc{o} eat-\textsc{irr}.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f}\\
\glt `{``}It would have eaten you.'''
\z

\ea\label{ex:06-a125}
Tiwa kabene amani.\\
\gll tiwa kaba-hene ama-ni\\
     2\textsc{sg}.\textsc{o} eat-\textsc{irr}.\textsc{f} \textsc{sec}-\textsc{bkg}.\textsc{f}\\
\glt `{``}It would have eaten you.'''
\z

\ea\label{ex:06-a126}
Tiwa kabehene amake.\\
\gll tiwa kaba-hene ama-ke\\
     2\textsc{sg}.\textsc{o} eat-\textsc{irr}.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f}\\
\glt `{``}It would have eaten you.'''
\z

\ea\label{ex:06-a127}
Tiwa nabowehene amake.\\
\gll tiwa naabowa-hene ama-ke\\
     2\textsc{sg}.\textsc{o} kill-\textsc{irr}.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f}\\
\glt `{``}It would have killed you.'''
\z

\ea\label{ex:06-a128}
Tiwa nabowehene tiwa, tiwa ota kasomineke.\\
\gll tiwa naabowa-hene tiwa tiwa otaa ka-soma-ne-ke\\
     2\textsc{sg}.\textsc{o} kill-\textsc{irr}.\textsc{f} 2\textsc{sg}.\textsc{o} 2\textsc{sg}.\textsc{o} 1\textsc{ex}.\textsc{s} \textsc{comit}-help-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `{``}It would have killed you. We saved you.'''
\z

\ea\label{ex:06-a129}
Hasi tina tikamahi.\\
\gll hasi ti-na ti-ka-ma-hi\\
     escape 2\textsc{sg}.\textsc{s}-\textsc{aux}.\textsc{f} 2\textsc{sg}.\textsc{s}-go-back-\textsc{imp}.\textsc{f}\\
\glt `{``}You have escaped, now go home.'''
\z

\ea\label{ex:06-a130}
Tikamahi.\\
\gll ti-ka-ma-hi\\
     2\textsc{sg}.\textsc{s}-go-back-\textsc{imp}.\textsc{f}\\
\glt `{``}Go home.'''
\z

\ea\label{ex:06-a131}
Ota towakamiba.\\
\gll otaa to-ka-ka-ma-baha\\
     1\textsc{ex}.\textsc{s} away-\textsc{comit}-go-back-before.\textsc{f}\\
\glt `{``}We will take it away.'''
\z

\ea\label{ex:06-a132}
Ota kabibe.\\
\gll otaa kaba-be\\
     1\textsc{ex}.\textsc{s} eat-\textsc{immed}.\textsc{f}\\
\glt `{``}We will eat it.'''
\z

\ea\label{ex:06-a133}
Te kabiya.\\
\gll tee kaba-yahi\\
     2\textsc{pl}.\textsc{s} eat-\textsc{dist}.\textsc{imp}.\textsc{f}\\
\glt `{``}You eat it.'''
\z

\ea\label{ex:06-a134}
Owa owa wataharake.\\
\gll owa owa wata-hara-ke\\
     1\textsc{sg}.\textsc{o} 1\textsc{sg}.\textsc{o} catch-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `{``}It captured me.'''
\z

\ea\label{ex:06-a135}
Awaawa orahani.\\
\gll awa-awa o-ra-hani\\
     \textsc{dup}-see 1\textsc{sg}.\textsc{s}-\textsc{neg}-\textsc{ip}.\textsc{n}.\textsc{f}\\
\glt `{``}I didn't see it.'''
\z

\ea\label{ex:06-a136}
Obatasi ka ya owa wata amake.\\
\gll o-batasi ka ya owa wata ama-ke\\
     1\textsc{sg}.\textsc{poss}-middle\_of\_back \textsc{loc} \textsc{adju} 1\textsc{sg}.\textsc{o} catch.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f}\\
\glt `{``}It captured me from behind.'''
\z

\ea\label{ex:06-a137}
Faya me towakamemetemoneke.\\
\gll faya me to-ka-ka-ma-hemete-mone-ke\\
     so 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `Then the two of them went away.'
\z

\ea\label{ex:06-a138}
Tokome nowati ya me towakama me tiwa himisa me, me tiwa himisa me towakama me hikabemetemone.\\
\gll to-ka-me nowati ya me to-ka-ka-ma me tiwa hi-misa me me tiwa hi-misa me to-ka-ka-ma me hi-kaba-hemete-mone\\
     away-go-back.\textsc{m} after \textsc{adju} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} carry \textsc{oc}-up.\textsc{f} 3\textsc{pl} 3\textsc{pl}.\textsc{s} carry \textsc{oc}-up.\textsc{f} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{oc}-eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
\glt `After he left, they left. They put it on their shoulders, and went, and ate it.'
\z

\ea\label{ex:06-a139}
Hoo inamati inamati were kane ota nabowe ota kakamineke.\\
\gll hoo inamati inamati were ka-ne otaa na-abohe otaa ka-ka-ma-ne-ke\\
     whew someone(\textsc{m}) someone(\textsc{m}) throw \textsc{comit}-\textsc{aux}.\textsc{m} 1\textsc{ex}.\textsc{s} \textsc{caus}-die.\textsc{m} 1\textsc{ex}.\textsc{s} \textsc{comit}-come-back-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `{``}Whew, it threw the man, and we killed it, and we are bringing it.'''
\z

\ea\label{ex:06-a140}
Inamati inamati were kane ota, nabowe ota kakamine, me hina, faya me hikabemetemone fahi.\\
\gll inamati inamati were ka-ne otaa na-abohe otaa ka-ka-ma-ne me hi-na faya me hi-kaba-hemete-mone fahi\\
     someone(\textsc{m}) someone(\textsc{m}) throw \textsc{comit}-\textsc{aux}.\textsc{m} 1\textsc{ex}.\textsc{s} \textsc{caus}-die.\textsc{m} 1\textsc{ex}.\textsc{s} \textsc{comit}-come-back-\textsc{cont}.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{oc}-say.\textsc{f} so 3\textsc{pl}.\textsc{s} \textsc{oc}-eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
\glt `{``}It threw the man, and we killed it, and we are bringing it,'' they said. They ate it.'
\z

\ea\label{ex:06-a141}
Tasi tonamematamonaka.\\
\gll tasi to-na-ma-himata-mona-ka\\
     emerge away-\textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `He arrived back.'
\z

\ea\label{ex:06-a142}
Hoo inamati owa were kanaba owa watani okomine oke.\\
\gll hoo inamati owa were ka-na-haba owa wata-hani o-ka-ma-ne o-ke\\
     whew spirit(\textsc{m}) 1\textsc{sg}.\textsc{o} throw \textsc{comit}-\textsc{aux}-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{o} catch-\textsc{ip}.\textsc{n}.\textsc{f} 1\textsc{sg}.\textsc{s}-come-back-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
\glt `{``}Whew, a spirit captured me to throw me, and I have come back.'''
\z

\ea\label{ex:06-a143}
Inamati, watore tinani tiwa awine ti bati ati nematamonaka.\\
\gll inamati watore ti-na-hani tiwa awine ti bati ati na-himata-mona-ka\\
     spirit(\textsc{m}) forget 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{f} 2\textsc{sg}.\textsc{s} seem.\textsc{f} 2\textsc{sg}.\textsc{s} 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `{``}You were not wary,'' his father said.'\footnote{The word \textit{inamati} is a false start.}
\z

\ea\label{ex:06-a144}
Kose mati toa awine kose mati.\\
\gll kose\_mati to-ha awine kose\_mati\\
     name\_of\_monster(\textsc{m}) \textsc{ch}-be.\textsc{f} seem.\textsc{f} name\_of\_monster(\textsc{m})\\
\glt `{``}It must have been a \textit{kose mati}.'''\footnote{At various points in the story, Siko switches back and forth between masculine and feminine agreement for the monster.}
\z

\ea\label{ex:06-a145}
Inamati wehinoka wasina habi ya, mase hi ni hiri ne tonokomisari ahi.\\
\gll inamati waa-hino-ka wasina habi ya mase hii ni hiri ne to-noko-misa-haari ahi\\
     spirit(\textsc{m}) stand-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{decl}.\textsc{m} kapok(\textsc{f}) root.\textsc{f} \textsc{adju} curassow(\textsc{m}) call \textsc{aux}.\textsc{nfin} make \textsc{aux}.\textsc{m} away-point-up-\textsc{ip}.\textsc{e}.\textsc{m} then\\
\glt `{``}The monster was standing among the buttress roots of a kapok tree, looking up and calling like a curassow.'''
\z

\ea\label{ex:06-a146}
Tonokomiseno ki one obatasi ka owawitebone obatasi owa watareka.\\
\gll to-noko-misa-hino kii o-ne o-batasi ka o-waa-wite-habone o-batasi owa wata-hare-ka\\
     away-point-up-\textsc{ip}.\textsc{n}.\textsc{m} look\_at 1\textsc{sg}.\textsc{s}-\textsc{aux}.\textsc{m} 1\textsc{sg}.\textsc{poss}-middle\_of\_back \textsc{loc} 1\textsc{sg}.\textsc{s}-stand-out-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{poss}-middle\_of\_back 1\textsc{sg}.\textsc{o} catch-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}\\
\glt `{``}It was looking up. I was looking at it, and when I changed my position, it captured me from behind.'''\footnote{The words \textit{obatasi ka} are a false start.}
\z

\ea\label{ex:06-a147}
Kose mati tohe awane.\\
\gll kose\_mati to-he awa-ne\\
     name\_of\_monster(\textsc{m}) \textsc{ch}-be.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m}\\
\glt `{``}It must have been a \textit{kose mati}.'''
\z

\ea\label{ex:06-a148}
Oof totoka tihahi waha ya.\\
\gll oof to-to-ka ti-ha-hi waha ya\\
     hoo \textsc{dup}-away-go 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{f} flooded\_forest(\textsc{f}) \textsc{adju}\\
\glt `{``}You shouldn't go to the flooded forest.'''
\z

\ea\label{ex:06-a149}
Yama watemone amani waha ya.\\
\gll yama wata-hamone ama-ni waha ya\\
     thing(\textsc{f}) be\_located-\textsc{rep}.\textsc{f} \textsc{sec}-\textsc{bkg}.\textsc{f} flooded\_forest(\textsc{f}) \textsc{adju}\\
\glt `{``}The monster is there in the flooded forest.'''
\z

\ea\label{ex:06-a150}
Waha ya watemona me na hihehimari amane tiwa wata nanoho.\\
\gll waha ya wata-himona me na hi-ha-himari ama-ne tiwa wata na-no-ho\\
     flooded\_forest(\textsc{f}) \textsc{adju} be\_located-\textsc{rep}.\textsc{m} 3\textsc{pl}.\textsc{s} say \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{e}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m} 2\textsc{sg}.\textsc{o} catch.\textsc{cont}.\textsc{m} \textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
\glt `{``}People used to say long ago that it is in the flooded forest, and now it captured you.'''\footnote{Although the indirect quote margin means `people used to say' as indicated, the word \textit{na} is not actually a verb meaning `say'. The verb of the clause is \textit{ati na} `say', but there is a morphophonological rule that applies when this verb is used in the iterative construction, and there is a person prefix, in this case \textit{hi-} (\cite[118]{Dixon:2004xr}). I have glossed the auxiliary \textit{na} as `say' in order to make things easier for the reader. When the rule does not apply, i.e. when there is no person prefix, the reduplicated root \textit{aati} appears instead of the auxiliary, as in line (\ref{ex:03-a61}) of \textref{text:03} \textit{The first corn}.}
\z

\ea\label{ex:06-a151}
Owa wate owa were kane me inamati owa kasoma me towakamarake.\\
\gll owa wate owa were ka-ne me\_inamati owa ka-soma me to-ka-ka-ma-hara-ke\\
     1\textsc{sg}.\textsc{o} catch.\textsc{m} 1\textsc{sg}.\textsc{o} throw \textsc{comit}-\textsc{aux}.\textsc{m} spirits(\textsc{f}) 1\textsc{sg}.\textsc{o} \textsc{comit}-help.\textsc{f} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `{``}It captured me, and threw me. Two spirits saved me, and then went away.'''
\z

\ea\label{ex:06-a152}
Owa me kasoma me towakamarake.\\
\gll owa me ka-soma me to-ka-ka-ma-hara-ke\\
     1\textsc{sg}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{comit}-help.\textsc{f} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `{``}They saved me, and then went away.'''
\z

\newpage
\ea\label{ex:06-a153}
Owa owa tiwarene ama tike.\\
\gll owa owa ti-awa-ra-hene ama ti-ke\\
     1\textsc{sg}.\textsc{o} 1\textsc{sg}.\textsc{o} 2\textsc{sg}.\textsc{s}-see-\textsc{neg}-\textsc{irr}.\textsc{f} \textsc{sec} 2\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
\glt `{``}You wouldn't have seen me anymore.'''
\z

\ea\label{ex:06-a154}
Owa wete kanareka isiri ya.\\
\gll owa wete ka-na-hare-ka isiri ya\\
     1\textsc{sg}.\textsc{o} tie \textsc{comit}-\textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m} basket(\textsc{f}) \textsc{adju}\\
\glt `{``}It tied me in a basket.'''
\z

\ea\label{ex:06-a155}
Ha onamone onoko mi hina ama haro.\\
\gll haa o-na-hamone o-noko mii hi-na ama haaro\\
     call 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{rep}.\textsc{f} 1\textsc{sg}.\textsc{poss}-face defecate\_on \textsc{oc}-\textsc{aux}.\textsc{f} \textsc{sec} that\_one.\textsc{f}\\
\glt `{``}It defecated on my face, because it said I was calling out.'''\footnote{The fact that the prefix \textit{hi-} is present in \textit{mi hina} shows that the O is third person. The O is \textit{onoko}, so the \textit{hi-} shows that \textit{onoko} is a third person NP even though it is an inalienably possessed noun with a first person singular possessor. That is, the possessor does not determine the person of the NP. See the discussion in \citet[288f.]{Dixon:2004xr}.}
\z

\ea\label{ex:06-a156}
Iyo wani ama ahi.\\
\gll iyo wani ama ahi\\
     feces(\textsc{f}) be\_attached.\textsc{nom}.\textsc{f} \textsc{sec} then\\
\glt `{``}There were feces on my face.'''
\z

\ea\label{ex:06-a157}
Onoko mi ne owa towakimisamatase, owa koro tokowite, okobokabone owa.\\
\gll o-noko mii ne owa to-ka-ka-misa-ma-tase owa koro to-ka-wite o-ka-boka-habone owa\\
     1\textsc{sg}.\textsc{poss}-face defecate\_on \textsc{aux}.\textsc{m} 1\textsc{sg}.\textsc{o} away-\textsc{comit}-go-up-back-again.\textsc{m} 1\textsc{sg}.\textsc{o} throw away-\textsc{comit}-out 1\textsc{sg}.\textsc{s}-\textsc{comit}-enter\_water-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{s}\\
\glt `{``}It defecated on my face. It took me back up again in the tree, and threw me, so I would go into the water.'''
\z

\ea\label{ex:06-a158}
Fa boti ihi ohabone ati narene.\\
\gll faha boti ihi o-ha-habone ati na-hare-ne\\
     water(\textsc{f}) inside victim\_of.\textsc{f} 1\textsc{sg}.\textsc{s}-be-\textsc{int}.\textsc{f} say \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
\glt `{``}It wanted me to drown.'''
\z

\newpage
\ea\label{ex:06-a159}
Faya okabokabone ati ne, kaboki owatorara oke.\\
\gll faya o-ka-boka-habone ati ne ka-boki o-wato-ra-hara o-ke\\
     so 1\textsc{sg}.\textsc{s}-\textsc{comit}-enter\_water-\textsc{int}.\textsc{f} say \textsc{aux}.\textsc{m} \textsc{comit}-enter\_water.\textsc{nfin} 1\textsc{sg}.\textsc{s}-know-\textsc{neg}-\textsc{ip}.\textsc{e}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
\glt `{``}It wanted me to go in the water, but I couldn't go in the water.'''
\z

\ea\label{ex:06-a160}
Kaboki owatorimakoni hasi one oni.\\
\gll ka-boki o-wato-ra-makoni hasi o-ne o-ni\\
     \textsc{comit}-enter\_water.\textsc{nfin} 1\textsc{sg}.\textsc{s}-know-\textsc{neg}-for\_this\_reason.\textsc{f} escape 1\textsc{sg}.\textsc{s}-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
\glt `{``}Because I couldn't go in the water, I survived.'''
\z

\ea\label{ex:06-a161}
Faya hasi onara fahi.\\
\gll faya hasi o-na-hara fahi\\
     so escape 1\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f} then\\
\glt `{``}I survived.'''
\z

\ea\label{ex:06-a162}
Oko ha ni me mitahani owa me nakiani me, i kiya kanarake okowitabasa owa ni ya.\\
\gll o-ka haa ni me mita-hani owa me na-ka-ki-hani me ihi kiya ka-na-hara-ke o-ka-ita-basa owa ni ya\\
     1\textsc{sg}.\textsc{poss}-\textsc{poss} call \textsc{aux}.\textsc{nfin} 3\textsc{pl}.\textsc{s} hear-\textsc{ip}.\textsc{n}.\textsc{f} 1\textsc{sg}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{caus}-come-coming-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{poss} result\_of.\textsc{f} yell \textsc{comit}-\textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f} \textsc{dup}-\textsc{comit}-sit-edge.\textsc{f} 1\textsc{sg} to \textsc{adju}\\
\glt `{``}They heard me calling, and came after me. Because of them the monster cried out, as I was in the basket on the edge of the lake.'''
\z

\ea\label{ex:06-a163}
Kosi abono me kosi abono memone me ati na me, kosi abono me, owa me kasoma me towakamarake.\\
\gll kosi abono me kosi abono me-mone me ati na me kosi abono me owa me ka-soma me to-ka-ka-ma-hara-ke\\
     palm\_sp(\textsc{m}) spirit 3\textsc{pl} palm\_sp(\textsc{m}) spirit 3\textsc{pl}-\textsc{rep}.\textsc{f} 3\textsc{pl}.\textsc{s} say \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} palm\_sp(\textsc{m}) spirit 3\textsc{pl}.\textsc{s} 1\textsc{sg}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{comit}-help.\textsc{f} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `{``}They said they were spirits of the kosi palm. The two kosi spirits saved me, and then went away.'''
\z

\ea\label{ex:06-a164}
Okoma tasi onamine o.\\
\gll o-ka-ma tasi o-na-ma-ne o\\
     1\textsc{sg}.\textsc{s}-come-back.\textsc{f} emerge 1\textsc{sg}.\textsc{s}-\textsc{aux}-back-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}\\
\glt `{``}I came back, and appeared.'''
\z

\ea\label{ex:06-a165}
Owa te awamarene te amake.\\
\gll owa tee awa-ma-ra-hene tee ama-ke\\
     1\textsc{sg}.\textsc{o} 2\textsc{pl}.\textsc{s} see-back-\textsc{neg}-\textsc{irr}.\textsc{f} 2\textsc{pl}.\textsc{s} \textsc{sec}-\textsc{decl}.\textsc{f}\\
\glt `{``}You wouldn't have seen me anymore.'''
\z

\ea\label{ex:06-a166}
Oabene ama oni fa boti ya.\\
\gll o-ahaba-hene ama o-ni faha boti ya\\
     1\textsc{sg}.\textsc{s}-die-\textsc{irr}.\textsc{f} \textsc{sec} 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f} water(\textsc{f}) inside \textsc{adju}\\
\glt `{``}I would have died in the depths of the water.'''
\z

\ea\label{ex:06-a167}
Faha boti ya ohabene ama oke.\\
\gll faha boti ya o-ahaba-hene ama o-ke\\
     water(\textsc{f}) inside \textsc{adju} 1\textsc{sg}.\textsc{s}-die-\textsc{irr}.\textsc{f} \textsc{sec} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
\glt `{``}I would have died in the water.'''
\z

\ea\label{ex:06-a168}
Haha okomina te mitaharo, okoyo okoyo hikaraboemetemone?\\
\gll haaha o-kamina tee mita-haro okoyo okoyo hi-karaboha-hemete-mone\\
     this.\textsc{f} 1\textsc{sg}.\textsc{s}-tell.\textsc{f} 2\textsc{pl}.\textsc{s} hear-\textsc{rp}.\textsc{e}.\textsc{f} 1\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) 1\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) \textsc{oc}-shoot\_with\_blowgun-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
\glt `Did you already hear the one I told, about what my older brother shot with a blowgun?'
\z

\ea\label{ex:06-a169}
Afa okomina te mitaroke.\\
\gll afa o-kamina tee mita-haro-ke\\
     this.\textsc{f} 1\textsc{sg}.\textsc{s}-tell.\textsc{f} 2\textsc{pl}.\textsc{s} hear-\textsc{rp}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
\glt `I told it, and you heard it.'
\z

\ea\label{ex:06-a170}
    One, oneboneke, kose mati owa.\\
    \gll one one-bone-ke kose\_mati owa\\
     other.\textsc{f} other.\textsc{f}-\textsc{int}.\textsc{f}-\textsc{decl}.\textsc{f} name\_of\_monster(\textsc{m}) other.\textsc{m}\\
    \glt `It will be another story, about another \textit{kose mati}.'
\z