\ea\label{ex:08-a1}
Siraba okominamatibeya inamati hekamemetemonehe.\\
    \gll Siraba o-kamina-mata-be-ya inamati hi-ka-ka-ma-hemete-mone-he\\
    Siraba(\textsc{f}) 1\textsc{sg}.\textsc{s}-tell-short\_time-\textsc{immed}.\textsc{f}-now spirit(\textsc{m}) \textsc{oc}-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `I will tell about Siraba, who a spirit took away.'
\z

\ea\label{ex:08-a2}
Siro me tonaka me towaka me, yama mati forineke faro.\\
    \gll siro me to-na-ka me to-ka-ka me yama mati fori-ne-ke faaro\\
    tree.sp(\textsc{f}) 3\textsc{pl}.\textsc{s} away-\textsc{caus}-go.\textsc{f} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go.\textsc{f} 3\textsc{pl}.\textsc{s} thing(\textsc{f}) line.\textsc{f} lie\_on\_top-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f} that.\textsc{f}\\
    \glt `They went after \textit{siro} fruits. The two of them went. There is a wire there on top.'
\z

\newpage
\ea\label{ex:08-a3}
Siro me tonaka me, siro me tonaka me towaka me, ni ya inamati tokomakematamonaka.\\
    \gll siro me to-na-ka me siro me to-na-ka me to-ka-ka me ni ya inamati to-ka-make-himata-mona-ka\\
    tree.sp(\textsc{f}) 3\textsc{pl}.\textsc{s} away-\textsc{caus}-go.\textsc{f} 3\textsc{pl}.\textsc{s} tree.sp(\textsc{f}) 3\textsc{pl}.\textsc{s} away-\textsc{caus}-go.\textsc{f} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go.\textsc{f} 3\textsc{pl} to \textsc{adju} spirit(\textsc{m}) away-go-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `They went after \textit{siro} fruits. They went. A spirit went to them.'
\z

\largerpage
\ea\label{ex:08-a4}
Inamati tokomake, inamati tokomake, siro me yaba me yoro na me, siro me yaba me yoro na me ni ya tofotomakimatamonane siro me ta toniharewama me yoro na me ni ya.\\
    \gll inamati to-ka-make inamati to-ka-make siro me yaba me yoro na me siro me yaba me yoro na me ni ya to-foto-maki-himata-mona-ne siro me taa to-niha-re-waha-ma me yoro na me ni ya\\
    spirit(\textsc{m}) away-go-following spirit(\textsc{m}) away-go-following tree.sp(\textsc{f}) 3\textsc{pl}.\textsc{s} pick\_up.\textsc{f} 3\textsc{pl}.\textsc{s} stand \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} tree.sp(\textsc{f}) 3\textsc{pl}.\textsc{s} pick\_up.\textsc{f} 3\textsc{pl}.\textsc{s} stand \textsc{aux}.\textsc{f} 3\textsc{pl} to \textsc{adju} away-appear-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m} tree.sp(\textsc{f}) 3\textsc{pl}.\textsc{s} be\_piled\_up \textsc{ch}-\textsc{caus}-raised\_surface-change-back.\textsc{f} 3\textsc{pl}.\textsc{s} stand \textsc{aux}.\textsc{f} 3\textsc{pl} to \textsc{adju}\\
    \glt `The spirit went there. He appeared where they were as they were picking up \textit{siro} fruits, as they were piling the \textit{siro} fruits up.'
\z

\ea\label{ex:08-a5}
Okoyo kamaki ati nemetemoneni.\\
    \gll okoyo ka-maki ati na-hemete-mone-ni\\
    1\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) come-following.\textsc{m} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}My older brother has come,'' she said.'
\z

\ea\label{ex:08-a6}
Ki kabote nemetemoneke.\\
    \gll kii kabote na-hemete-mone-ke\\
    look\_at quickly \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `She looked quickly.'
\z

\ea\label{ex:08-a7}
Fara siro hatira? ati ne.\\
    \gll fara siro hati-ra ati ne\\
    same.\textsc{f} tree.sp(\textsc{f}) ripe-\textsc{npq} say \textsc{aux}.\textsc{m}\\
    \glt `{``}Are there ripe \textit{siro} fruits?'' he said.'
\z

\newpage
\ea\label{ex:08-a8}
Siro hataharani.\\
    \gll siro hata-hara-ni\\
    tree.sp(\textsc{f}) be\_ripe-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}The \textit{siro} fruits are ripe.'''
\z

\ea\label{ex:08-a9}
Siro so tosine faro ati nemetemonehe.\\
    \gll siro soo to-kosa-ne faaro ati na-hemete-mone-he\\
    tree.sp(\textsc{f}) stand \textsc{ch}-middle-\textsc{cont}.\textsc{f} that.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `{``}The \textit{siro} fruits are there in a pile,'' she said.'
\z

\ea\label{ex:08-a10}
Faya Siro, siro siro hati tiwahi ati nemetemoneke.\\
    \gll faya siro siro siro hati ti-awa-hi ati na-hemete-mone-ke\\
    so tree.sp(\textsc{f}) tree.sp(\textsc{f}) tree.sp(\textsc{f}) ripe 2\textsc{sg}.\textsc{s}-see-\textsc{imp}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Look at the \textit{siro} fruits,'' she said.'
\z


\ea\label{ex:08-a11}
Siro hati, itikime siro hawi ne wematamonaha.\\
    \gll siro hati iti-kime siro hawi ne waa-himata-mona-ha\\
    tree.sp(\textsc{f}) ripe pick\_up-two.\textsc{m} tree.sp(\textsc{f}) gnaw \textsc{aux}.\textsc{m} stand-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `He  picked up two of the \textit{siro} fruits. He stood there gnawing on the \textit{siro} fruits.'
\z

\ea\label{ex:08-a12}
We, Ayo ibaka ya tikaribeya? ati nemetemoneke.\\
    \gll wee ayo ibaka ya ti-ka-ri-be-ya ati na-hemete-mone-ke\\
    stand.\textsc{m} older\_brother(\textsc{m}) where \textsc{adju} 2\textsc{sg}.\textsc{s}-go-\textsc{cq}.\textsc{f}-\textsc{immed}.\textsc{f}-now say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He stood there. ``Brother, where are you going?'' she said.'
\z

\ea\label{ex:08-a13}
Okiba ahi Boto Komaha ya.\\
    \gll o-ka-baha ahi Boto\_Komaha ya\\
    1\textsc{sg}.\textsc{s}-go-before.\textsc{f} there Boto\_Komaha(\textsc{f}) \textsc{adju}\\
    \glt `{``}I'm going to Boto Komaha.'''
\z

\newpage
\ea\label{ex:08-a14}
Siro noki koro towiti awa kinarimatamonane.\\
    \gll siro noki koro to-witi awa kina-ri-himata-mona-ne\\
    tree.sp(\textsc{f}) seed.\textsc{f} throw away-out.\textsc{m} tree(\textsc{f}) hit-raised\_surface-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `He threw a \textit{siro} seed and hit a tree.'
\z

\ea\label{ex:08-a15}
Boto Komaha ya okabone oni ati nematamona.\\
    \gll Boto\_Komaha ya o-ka-habone o-ni ati na-himata-mona\\
    Boto\_Komaha(\textsc{f}) \textsc{adju} 1\textsc{sg}.\textsc{s}-go-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `{``}I'm going to Boto Komaha,'' he said.'
\z

\ea\label{ex:08-a16}
Inamati amane ati nemetemoneke one.\\
    \gll inamati ama-ne ati na-hemete-mone-ke one\\
    spirit(\textsc{m}) \textsc{sec}-\textsc{bkg}.\textsc{m} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} other.\textsc{f}\\
    \glt `{``}It's a spirit,'' the other one said.'
\z


\ea\label{ex:08-a17}
Ayo amare ayo tinihine tini ati nemetemonehe.\\
    \gll ayo ama-re ayo ti-niha-ne ti-ni ati na-hemete-mone-he\\
    2\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) be-\textsc{neg}.\textsc{m} 2\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) 2\textsc{sg}.\textsc{s}-\textsc{caus}-\textsc{cont}.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `{``}You are calling him your brother, but he is not your brother,'' she said.'
\z

\ea\label{ex:08-a18}
Me hiwemata awaka.\\
    \gll me hi-awa-himata awa-ka\\
    3\textsc{pl}.\textsc{s} \textsc{oc}-see-\textsc{fp}.\textsc{n}.\textsc{m} seem.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `They recognized him.'
\z

\ea\label{ex:08-a19}
Me hiwemata awane.\\
    \gll me hi-awa-himata awa-ne\\
    3\textsc{pl}.\textsc{s} \textsc{oc}-see-\textsc{fp}.\textsc{n}.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `They recognized him.'
\z

\ea\label{ex:08-a20}
Ayo wawa timatahi.\\
    \gll ayo wa-waa ti-mata-hi\\
    older\_brother(\textsc{m}) \textsc{dup}-stand 2\textsc{sg}.\textsc{s}-short\_time-\textsc{imp}.\textsc{f}\\
    \glt `{``}Brother, you stay here.'''
\z

\ea\label{ex:08-a21}
Siro wiyeba ota tonakamatibeya ati nemetemonehe.\\
    \gll siro wiye-ba otaa to-na-ka-mata-be-ya ati na-hemete-mone-he\\
    tree.sp(\textsc{f}) container.\textsc{f}-\textsc{fut} 1\textsc{ex}.\textsc{s} away-\textsc{caus}-go-short\_time-\textsc{immed}.\textsc{f}-now say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `{``}We are going to go get baskets for the \textit{siro} fruits.'''
\z

\ea\label{ex:08-a22}
Faya me fanawi towaka mera kakatome mera towamakemata awaka.\\
    \gll faya me fanawi to-ka-ka mera ka-katome mera to-awa-make-himata awa-ka\\
    so 3\textsc{pl}.\textsc{s} woman away-\textsc{comit}-go.\textsc{f} 3\textsc{pl}.\textsc{o} \textsc{comit}-look\_at.\textsc{m} 3\textsc{pl}.\textsc{o} away-see-following-\textsc{fp}.\textsc{n}.\textsc{m} seem.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `The women went. He watched them, and he went after them to see them.'
\z

\ea\label{ex:08-a23}
Me ebe, mera kakatomemoneke.\\
    \gll me ebe mera ka-katoma-hamone-ke\\
    3\textsc{pl}.\textsc{poss} \textsc{recip} 3\textsc{pl}.\textsc{o} \textsc{comit}-look\_at-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He watched them.'
\z

\ea\label{ex:08-a24}
Yama me waso me ti kasikima waso me kanasona me, me yoro ra yama me ti kasaremetemoneke.\\
    \gll yama me waso me tii ka-kosa-kima waso me ka-na-sona me me yoro ra yama me tii ka-kosa-ra-hemete-mone-ke\\
    thing(\textsc{f}) 3\textsc{pl} leaf(\textsc{f}) 3\textsc{pl}.\textsc{s} cut\_through \textsc{comit}-middle-two.\textsc{f} leaf(\textsc{f}) 3\textsc{pl}.\textsc{s} \textsc{comit}-\textsc{caus}-fall.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} stay \textsc{neg}.\textsc{f} thing(\textsc{f}) 3\textsc{pl}.\textsc{s} cut\_through \textsc{comit}-middle-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They cut off two palm leaves, and they fell on the ground. They didn't stay there. They didn't keep on cutting off leaves.'
\z

\ea\label{ex:08-a25}
Me kana ni towakamemetemoneke ha ya me tokafotowamabani mati.\\
    \gll me kana ni to-ka-ka-ma-hemete-mone-ke haa ya me to-ka-foto-waha-ma-haba-ni mati\\
    3\textsc{pl}.\textsc{s} run \textsc{aux}.\textsc{nfin} away-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} \textsc{dem} \textsc{adju} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-appear-change-back-\textsc{fut}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `They went running away. They came out on the other side.'
\z

\newpage
\ea\label{ex:08-a26}
Me tokafotowama me kana ni towakamemetemoneni, fare bati itamakino ya isiri kowahari.\\
    \gll me to-ka-foto-waha-ma me kana ni to-ka-ka-ma-hemete-mone-ni fare bati ita-maki-no ya isiri kowa-haari\\
    3\textsc{pl}.\textsc{s} away-\textsc{comit}-appear-change-back.\textsc{f} 3\textsc{pl}.\textsc{s} run \textsc{aux}.\textsc{nfin} away-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} same.\textsc{m} 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) sit-following.\textsc{nom}-\textsc{ip}.\textsc{n}.\textsc{m} \textsc{adju} basket(\textsc{f}) weave-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `They came out, and went running to where her father was sitting, weaving a basket.'
\z

\ea\label{ex:08-a27}
Isiri kowe ite ni ya bite kana ni tofotomemetemoneni.\\
    \gll isiri kowe ite ni ya bite kana ni to-foto-ma-hemete-mone-ni\\
    basket(\textsc{f}) weave.\textsc{m} sit.\textsc{m} to \textsc{adju} 3\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) run \textsc{aux}.\textsc{nfin} away-appear-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `As he sat there weaving the basket, his daughter appeared, running.'
\z

\ea\label{ex:08-a28}
Abi afa ama tini? ati nemetemonehe.\\
    \gll abi afa ama ti-ni ati na-hemete-mone-he\\
    father(\textsc{m}) this.\textsc{f} be 2\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `{``}Father, is it you?'' she said.'
\z

\ea\label{ex:08-a29}
Afa ama okene ati ne Abi inamati otara kiyo kamaki ahi ati nemetemoneni.\\
    \gll afa ama o-ke-ne ati ne abi inamati otara kiyo ka-maki ahi ati na-hemete-mone-ni\\
    this.\textsc{f} be 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}-\textsc{irr}.\textsc{f} say \textsc{aux}.\textsc{m} father(\textsc{m}) spirit(\textsc{m}) 1\textsc{ex}.\textsc{o} chase.\textsc{m} come-following.\textsc{nom} there say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}It is me,'' he said. ``Father, there is a spirit chasing us, and he is coming this way,'' she said.'
\z

\ea\label{ex:08-a30}
Inamati toe awane ati nematamona.\\
    \gll inamati to-he awa-ne ati na-himata-mona\\
    spirit(\textsc{m}) \textsc{ch}-be.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `{``}I guess it is a spirit,'' he said.'
\z

\ea\label{ex:08-a31}
Inamati fotomaki kerewe nematamonaha.\\
    \gll inamati foto-maki kerewe na-himata-mona-ha\\
    spirit(\textsc{m}) appear-following.\textsc{nfin} be\_slow \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `The spirit didn't take long to appear.'
\z

\ea\label{ex:08-a32}
Inamati fotomaki kerewe rematamonaka.\\
    \gll inamati foto-maki kerewe ra-himata-mona-ka\\
    spirit(\textsc{m}) appear-following.\textsc{nfin} be\_slow \textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `The spirit didn't take long to appear.'
\z

\ea\label{ex:08-a33}
Kana ni kamakimatamona ahi, fe bata torebanoho.\\
    \gll kana ni ka-maki-himata-mona ahi fee bata to-re-hiba-no-ho\\
    run \textsc{aux}.\textsc{nfin} come-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then 3\textsc{sg}.\textsc{o} surprise \textsc{ch}-raised\_surface-\textsc{fut}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `He came running, and he surprised him.'
\z

\ea\label{ex:08-a34}
Himata ebe tina tikamakine tiriya? ati nematamona Siraba bati.\\
    \gll himata ebe ti-na ti-ka-maki-ne ti-ri-ya ati na-himata-mona Siraba bati\\
    what purpose 2\textsc{sg}.\textsc{s}-say.\textsc{f} 2\textsc{sg}.\textsc{s}-come-following-\textsc{cont}.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{cq}.\textsc{f}-now say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} Siraba(\textsc{f}) 3\textsc{sg}.\textsc{poss}.father(\textsc{m})\\
    \glt `{``}What are you doing coming here?'' he said, Siraba's father.'
\z

\ea\label{ex:08-a35}
Me okiyorini ati nematamonane.\\
    \gll me o-kiyo-ra-ni ati na-himata-mona-ne\\
    3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-chase-\textsc{neg}-\textsc{bkg}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}I'm not chasing them,'' he said.'
\z

\ea\label{ex:08-a36}
Okoto tikiyoani tikamakine tini.\\
    \gll okoto ti-kiyo-hani ti-ka-maki-ne ti-ni\\
    1\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) 2\textsc{sg}.\textsc{s}-chase-\textsc{ip}.\textsc{n}.\textsc{f} 2\textsc{sg}.\textsc{s}-come-following-\textsc{cont}.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}You have come chasing my daughter.'''
\z

\ea\label{ex:08-a37}
Tiwa owini ati nematamonaha.\\
    \gll tiwa o-awa-ni ati na-himata-mona-ha\\
    2\textsc{sg}.\textsc{o} 1\textsc{sg}.\textsc{s}-see-\textsc{bkg}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `{``}I see you.'''
\z

\ea\label{ex:08-a38}
Tikamiyahi.\\
    \gll ti-ka-ma-yahi\\
    2\textsc{sg}.\textsc{s}-go-back-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `{``}Go away.'''
\z

\ea\label{ex:08-a39}
Tikamiya fara titaboro ya ati nematamona.\\
    \gll ti-ka-ma-yahi fara ti-taboro ya ati na-himata-mona\\
    2\textsc{sg}.\textsc{s}-go-back-\textsc{dist}.\textsc{imp}.\textsc{f} same.\textsc{f} 2\textsc{sg}.\textsc{poss}-place \textsc{adju} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `{``}Go back to your own place,'' he said.'
\z

\ea\label{ex:08-a40}
Faya, wete tonamematamonaka.\\
    \gll faya wete to-na-ma-himata-mona-ka\\
    so return away-\textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `So he went back.'
\z

\ea\label{ex:08-a41}
Tokomake tokomare tokomarematamonaka.\\
    \gll to-ka-make to-ka-ma-re to-ka-ma-ra-himata-mona-ka\\
    away-go-following away-go-back-\textsc{neg}.\textsc{m} away-go-back-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He didn't go back.'
\z


\ea\label{ex:08-a42}
Wete toname, tokome totikosawamakematamonaka.\\
    \gll wete to-na-me to-ka-me to-iti-kosa-waha-make-himata-mona-ka\\
    return away-\textsc{aux}-back.\textsc{m} away-go-back.\textsc{m} away-go\_around-middle-change-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went back, but he came around.'
\z

\ea\label{ex:08-a43}
Tofotowamaki tokomakewematamonane.\\
    \gll to-foto-waha-maki to-ka-maki-waha-himata-mona-ne\\
    away-appear-change-following.\textsc{m} away-go-following-change-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `He came out again and went there.'
\z

\ea\label{ex:08-a44}
Tofotowamaki ka, bai toke ya yama me matehe me me kiyoha, matehe me me kiyohemetemoneke.\\
    \gll to-foto-waha-maki ka bahi to-ke ya yama me matehe me me kiyoha matehe me me kiyo-hemete-mone-ke\\
    away-appear-change-following.\textsc{m} of sun(\textsc{m}) away-go.\textsc{nom}.\textsc{m} \textsc{adju} thing(\textsc{f}) 3\textsc{pl} children(\textsc{f}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} chase.\textsc{f} children(\textsc{f}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} chase-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He came out again. In the late afternoon they chased the children.'
\z

\ea\label{ex:08-a45}
Siraba bati hiwahi hikahabanarehimata awane?\\
    \gll Siraba bati hi-awahi hi-ka-habana-ra-himata awa-ne\\
    Siraba(\textsc{f}) 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) \textsc{oc}-see.\textsc{m} \textsc{oc}-\textsc{comit}-order\_someone\_against-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `When Siraba's father saw him, didn't he send [his familiar spirit] against him?'
\z

\ea\label{ex:08-a46}
Hikabanaremata awaka fare, hiwa kaiyarabise hiwa awe ya.\\
    \gll hi-ka-habana-ra-himata awa-ka fare hiwa ka-hiyara-bise hiwa awe ya\\
    \textsc{oc}-\textsc{comit}-order\_someone\_against-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{m} seem.\textsc{m}-\textsc{decl}.\textsc{m} same.\textsc{m} 3\textsc{sg}.\textsc{refl}.\textsc{m} \textsc{comit}-speak-also.\textsc{m} 3\textsc{sg}.\textsc{refl}.\textsc{m} see.\textsc{nom}.\textsc{m} \textsc{adju}\\
    \glt `Apparently he didn't send [his familiar spirit] against him, even though he was a powerful shaman.'
\z

\ea\label{ex:08-a47}
Faya tokomake, mera noki ne, me haha ni me haha ni saiha me me abe kiyoa mati.\\
    \gll faya to-ka-make mera noki ne me haahaa ni me haahaa ni saiha me me abe kiyoha mati\\
    so away-go-following 3\textsc{pl}.\textsc{o} wait\_for \textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{s} laugh \textsc{aux}.\textsc{nfin} 3\textsc{pl}.\textsc{s} laugh \textsc{aux}.\textsc{nfin} be\_heard.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{poss} \textsc{recip} chase.\textsc{f} 3\textsc{pl}.\textsc{poss}\\
    \glt `He went there and waited for them. Their laughing could be heard as they chased each other.'
\z

\ea\label{ex:08-a48}
Me me kiyoa me, nakiri ya fara me fami me famemetemoneni.\\
    \gll me me kiyoha me nakiri ya fara me fami me fama-hemete-mone-ni\\
    3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} chase.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{sg}.\textsc{poss}.brother's\_wife(\textsc{f}) \textsc{adju} same.\textsc{f} 3\textsc{pl}.\textsc{s} be\_two.\textsc{nfin} 3\textsc{pl}.\textsc{s} be\_two-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `They were chasing them. She was always with her cousin.'
\z

\newpage
\ea\label{ex:08-a49}
Mera kiyoa kana ni tokomake mera moto kawamemetemoneke, yobe wahi bari ya.\\
    \gll mera kiyoha kana ni to-ka-make mera moto ka-waha-ma-hemete-mone-ke yobe waahi bari ya\\
    3\textsc{pl}.\textsc{o} chase.\textsc{f} run \textsc{aux}.\textsc{nfin} away-go-following 3\textsc{pl}.\textsc{o} loop \textsc{comit}-change-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} house(\textsc{m}) stand.\textsc{m} back \textsc{adju}\\
    \glt `She was chasing them. She ran after them. She chased them behind the longhouse.'
\z

\ea\label{ex:08-a50}
Yobe toro ya me winemetemoneni.\\
    \gll yobe toro ya me wina-hemete-mone-ni\\
    house(\textsc{m}) inside.\textsc{m} \textsc{adju} 3\textsc{pl}.\textsc{s} lie\_in\_hammock-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `They were in their hammocks inside the longhouse.'
\z

\ea\label{ex:08-a51}
Mera moto kawama, wara nawebano wematamonane, me ka owamonaha, fara fe fami ka anotimonaha.\\
    \gll mera moto ka-waha-ma wara na-waha-hiba-no waa-himata-mona-ne me ka owa-mona-ha fara fee fami ka anoti-mona-ha\\
    3\textsc{pl}.\textsc{o} loop \textsc{comit}-change-back.\textsc{f} grab \textsc{aux}-change-\textsc{fut}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m} stand-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m} 3\textsc{pl} of man(\textsc{m})-\textsc{rep}.\textsc{m}-\textsc{dup} same.\textsc{f} 3\textsc{sg} be\_with.\textsc{nom}.\textsc{f} which 3\textsc{sg}.\textsc{poss}\_older\_brother(\textsc{m})-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `She chased them around. He was standing there, and grabbed her. She thought he was one of their men, her companion's older brother.'
\z

\ea\label{ex:08-a52}
Fito hikanemata awaka.\\
    \gll fito hi-ka-na-himata awa-ka\\
    rush \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m} seem.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `She ran to him.'
\z

\ea\label{ex:08-a53}
Makoni fe wara tomisa namata, me ka owamonaha.\\
    \gll makoni fee wara to-misa na-mata me ka owa-mona-ha\\
    for\_this\_reason 3\textsc{sg}.\textsc{o} grab away-up.\textsc{list} \textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{fp}.\textsc{n}.\textsc{m} 3\textsc{pl} of man(\textsc{m})-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `So he grabbed her, since she thought he was one of their men.'
\z

\newpage
\ea\label{ex:08-a54}
Faya, Hibake ama tiriya? ati nemetemoneke.\\
    \gll faya hibake ama ti-ri-ya ati na-hemete-mone-ke\\
    so who.\textsc{m} be 2\textsc{sg}.\textsc{s}-\textsc{cq}.\textsc{f}-now say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Who are you?'' she said.'
\z

\ea\label{ex:08-a55}
Ino himitematamona.\\
    \gll ino hi-mita-himata-mona\\
    name.\textsc{m} \textsc{oc}-hear-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `She heard his name.'
\z

\ea\label{ex:08-a56}
Ino owato okara.\\
    \gll ino o-wato o-ka-ra\\
    name.\textsc{m} 1\textsc{sg}.\textsc{s}-know 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{m}-\textsc{neg}.\textsc{m}\\
    \glt `I don't know his name.'
\z

\ea\label{ex:08-a57}
Faya nakiri kana ni kamakiemetemoneni.\\
    \gll faya nakiri kana ni ka-maki-hemete-mone-ni\\
    so 3\textsc{sg}.\textsc{poss}.brother's\_wife(\textsc{f}) run \textsc{aux}.\textsc{nfin} come-following-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `Then her cousin came running.'
\z

\ea\label{ex:08-a58}
Kana ni kamakia, Hibake te fama? ati nemetemonehe.\\
    \gll kana ni ka-makia hibake tee fama ati na-hemete-mone-he\\
    run \textsc{aux}.\textsc{nfin} come-following.\textsc{f} who.\textsc{m} 2\textsc{pl}.\textsc{s} be\_two.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `She came running. ``Who are you with?'' she said.'
\z

\ea\label{ex:08-a59}
Fe okoyo amane? ati nemetemoneke.\\
    \gll fee okoyo ama-ne ati na-hemete-mone-ke\\
    3\textsc{sg} 1\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) be-\textsc{bkg}.\textsc{m} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Is that my brother?'' she said.'
\z

\ea\label{ex:08-a60}
Ayo awane ati nemetemonehe.\\
    \gll ayo awa-ne ati na-hemete-mone-he\\
    2\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) seem.\textsc{m}-\textsc{bkg}.\textsc{m} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `{``}I think it is your brother,'' she said.'
\z

\ea\label{ex:08-a61}
Ayo afa ama tini? ati nemetemoneke.\\
    \gll ayo afa ama ti-ni ati na-hemete-mone-ke\\
    older\_brother(\textsc{m}) this.\textsc{f} be 2\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Brother, is that you?'' she said.'
\z

\ea\label{ex:08-a62}
Fe ama oni ati ne Fe ama oni.\\
    \gll fee ama o-ni ati ne fee ama o-ni\\
    3\textsc{sg} be 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f} say \textsc{aux}.\textsc{m} 3\textsc{sg} be 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}It is me,'' he said.'
\z

\ea\label{ex:08-a63}
Ati yana ni hihiya nemata awaka.\\
    \gll ati yana ni hi-hiya na-himata awa-ka\\
    voice start \textsc{aux}.\textsc{nfin} \textsc{dup}-be\_different \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m} seem.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `I guess he had a different voice.'
\z

\ea\label{ex:08-a64}
Fe ama oni.\\
    \gll fee ama o-ni\\
    3\textsc{sg} be 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}It is me.'''
\z

\ea\label{ex:08-a65}
Okoyo ama tira, okoyo tiwa tinihine tini ati nemetemonehe.\\
    \gll o-ka ayo ama ti-ra okoyo tiwa ti-niha-ne ti-ni ati na-hemete-mone-he\\
    1\textsc{sg}.\textsc{poss}-\textsc{poss} older\_brother(\textsc{m}) be 2\textsc{sg}.\textsc{s}-\textsc{neg}.\textsc{f} 1\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) 2\textsc{sg}.\textsc{o} 2\textsc{sg}.\textsc{s}-\textsc{caus}-\textsc{cont}.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `{``}You aren't my brother, you are just making yourself out to be my brother.'''
\z

\ea\label{ex:08-a66}
Okoyo wine wine wine awane.\\
    \gll okoyo wine wine wine awa-ne\\
    1\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) lie\_in\_hammock.\textsc{m} lie\_in\_hammock.\textsc{m} lie\_in\_hammock.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}My brother is in his hammock.'''
\z

\newpage
\ea\label{ex:08-a67}
Ba, okoyo totowama omatibeya.\\
    \gll hiba okoyo to-to-awa-ma o-mata-be-ya\\
    wait 1\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) \textsc{dup}-away-see-back 1\textsc{sg}.\textsc{s}-short\_time-\textsc{immed}.\textsc{f}-now\\
    \glt `{``}Wait, I will go take a look at my brother.'''
\z

\ea\label{ex:08-a68}
Te yoyoro namatahi ati nemetemoneke.\\
    \gll tee yo-yoro na-mata-hi ati na-hemete-mone-ke\\
    2\textsc{pl}.\textsc{s} \textsc{dup}-stand \textsc{aux}-short\_time-\textsc{imp}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}You two stay here.'''
\z

\ea\label{ex:08-a69}
Faya kana ni tokoma tokiyomamemetemonehe.\\
    \gll faya kana ni to-ka-ma to-ka-yoma-ma-hemete-mone-he\\
    so run \textsc{aux}.\textsc{nfin} away-go-back.\textsc{f} away-go-through-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `She went running home and went inside.'
\z

\ea\label{ex:08-a70}
Faya Siraba fawa kanematamonaka fahi.\\
    \gll faya Siraba fawa ka-na-himata-mona-ka fahi\\
    so Siraba(\textsc{f}) disappear \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `Then he disappeared with Siraba.'
\z

\ea\label{ex:08-a71}
Siraba fawa kane, ni ya kamemetemonehe, mati aate naro.\\
    \gll Siraba fawa ka-ne ni ya ka-ma-hemete-mone-he mati a-ate na-haaro\\
    Siraba(\textsc{f}) disappear \textsc{comit}-\textsc{aux}.\textsc{m} to \textsc{adju} come-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup} 3\textsc{sg}.\textsc{poss}.mother(\textsc{f}) \textsc{dup}-ask \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f}\\
    \glt `As he was disappearing with Siraba, she came home and asked her mother.'
\z

\ea\label{ex:08-a72}
Ami ati nemetemone.\\
    \gll ami ati na-hemete-mone\\
    mother(\textsc{f}) say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `{``}Mother,'' she said.'
\z

\ea\label{ex:08-a73}
Ami okoyo wine awane?\\
    \gll ami okoyo wine awa-ne\\
    mother(\textsc{f}) 1\textsc{sg}.\textsc{poss}.older\_brother(\textsc{m}) lie\_in\_hammock.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}Mother, is my older brother in his hammock?'''
\z

\ea\label{ex:08-a74}
Okatao wine ati nemetemonehe.\\
    \gll okatao wine ati na-hemete-mone-he\\
    1\textsc{sg}.\textsc{poss}.son(\textsc{m}) lie\_in\_hammock.\textsc{m} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `{``}My son is in his hammock,'' she said.'
\z

\ea\label{ex:08-a75}
Inamati Siraba tama ne ware ahi.\\
    \gll inamati Siraba tama ne waa-hare ahi\\
    spirit(\textsc{m}) Siraba(\textsc{f}) hold\_onto \textsc{aux}.\textsc{m} stand-\textsc{ip}.\textsc{e}.\textsc{m} there\\
    \glt `{``}There is a spirit holding onto Siraba there on the plaza.'''
\z

\ea\label{ex:08-a76}
Owamamatabaya ati nemetemone.\\
    \gll o-awa-ma-mata-ba-ya ati na-hemete-mone\\
    1\textsc{sg}.\textsc{s}-see-back-short\_time-\textsc{immed}.\textsc{m}-now say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `{``}I'm going back to see her,'' she said.'
\z

\ea\label{ex:08-a77}
Faya kana ni kama kiyomama, fara me yoro ni ya tokomemetemone.\\
    \gll faya kana ni ka-ma ka-yoma-ma fara me yoro ni ya to-ka-ma-hemete-mone\\
    so run \textsc{aux}.\textsc{nfin} come-back.\textsc{f} come-through-back.\textsc{f} same.\textsc{f} 3\textsc{pl}.\textsc{s} stand \textsc{aux}.\textsc{nom}.\textsc{f} \textsc{adju} away-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `So she came back running. She came out of the longhouse and went to where the two of them had been standing.'
\z

\ea\label{ex:08-a78}
Tokoma yama honaremetemone ama mera kowa nawarebanihi.\\
    \gll to-ka-ma yama hona-ra-hemete-mone ama mera kowa na-wahare-haba-ni-hi\\
    away-go-back.\textsc{f} thing(\textsc{f}) have\_something-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec} 3\textsc{pl}.\textsc{o} whistle\_to \textsc{aux}-all\_over-\textsc{fut}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
    \glt `She went there, but there was no one there. She whistled for them all over.'
\z

\ea\label{ex:08-a79}
Boni ati sai nawaremetemone mera siba naro.\\
    \gll boni ati sai na-wahare-hemete-mone mera siba na-haaro\\
    upper\_lip.\textsc{f} voice be\_heard \textsc{aux}-all\_over-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl}.\textsc{o} look\_for \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f}\\
    \glt `She could be heard whistling all over, as she looked for them.'
\z

\newpage
\ea\label{ex:08-a80}
Mera siba na boni ati sai nawahare, kamawemetemone.\\
    \gll mera siba na boni ati sai na-wahare ka-ma-waha-hemete-mone\\
    3\textsc{pl}.\textsc{o} look\_for \textsc{aux}.\textsc{f} upper\_lip.\textsc{f} voice be\_heard \textsc{aux}-all\_over come-back-change-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `She could be heard whistling all over as she looked for them. She finally came back.'
\z

\ea\label{ex:08-a81}
\textup{Okomobi:} Abi me fami ya me fami me hinihenemete me ama?\\
    \gll abi me fami ya me fami me hi-niha-hene-mete me ama\\
    father(\textsc{m}) 3\textsc{pl}.\textsc{s} be\_two.\textsc{nom}.\textsc{f} \textsc{adju} 3\textsc{pl}.\textsc{s} be\_two.\textsc{nfin} 3\textsc{pl}.\textsc{o} \textsc{oc}-\textsc{caus}-\textsc{irr}.\textsc{f}-\textsc{fp}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{o} \textsc{sec}\\
    \glt `Father, if there were two of them, would he have taken both of them?'
\z

\ea\label{ex:08-a82}
Me fami mera totowakama kabaramata me noki hinemenehe.\\
    \gll me fami mera to-to-ka-ka-ma kaba-ra-mata me noki hi-na-hemenehe\\
    3\textsc{pl}.\textsc{s} be\_two.\textsc{nfin} 3\textsc{pl}.\textsc{o} \textsc{dup}-away-\textsc{comit}-go-back without\_stopping-\textsc{neg}.\textsc{nom}.\textsc{m}-\textsc{fp}.\textsc{n}.\textsc{m} 3\textsc{pl}.\textsc{o} wait\_for \textsc{oc}-\textsc{aux}-\textsc{hypoth}.\textsc{f}\\
    \glt `He could have taken both of them if he had waited.'
\z

\ea\label{ex:08-a83}
Faya tokiyomama, kakomeharo.\\
    \gll faya to-ka-yoma-ma ka-komeha-haaro\\
    so away-go-through-back.\textsc{f} \textsc{comit}-be\_fearsome-\textsc{ip}.\textsc{e}.\textsc{f}\\
    \glt `Then she went inside, because she was afraid.'
\z

\ea\label{ex:08-a84}
Kakomemetemoneke.\\
    \gll ka-komeha-hemete-mone-ke\\
    \textsc{comit}-be\_fearsome-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `She was afraid.'
\z

\ea\label{ex:08-a85}
Siraba, Siraba kamakiremetemone amani.\\
    \gll Siraba Siraba ka-maki-ra-hemete-mone ama-ni\\
    Siraba(\textsc{f}) Siraba(\textsc{f}) come-following-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec}-\textsc{bkg}.\textsc{f}\\
    \glt `Siraba did not come back.'
\z

\newpage
\ea\label{ex:08-a86}
Siraba kawahemetemone amani inamati hikawa hinahabaharo.\\
    \gll Siraba ka-waha-hemete-mone ama-ni inamati hi-ka-waa hi-nahaba-haaro\\
    Siraba(\textsc{f}) \textsc{comit}-dawn-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec}-\textsc{bkg}.\textsc{f} spirit(\textsc{m}) \textsc{oc}-\textsc{comit}-stay \textsc{oc}-at\_night-\textsc{ip}.\textsc{e}.\textsc{f}\\
    \glt `When morning came, Siraba had spent the night with the spirit.'
\z

\ea\label{ex:08-a87}
Tabiyo me hiwemete awineke.\\
    \gll tabiyo me hi-awa-hemete awine-ke\\
    absence 3\textsc{pl}.\textsc{s} \textsc{oc}-feel-\textsc{fp}.\textsc{n}.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They missed her.'
\z

\ea\label{ex:08-a88}
Me hitakora fawa nine nikani mati hinemetemone okomise aati toamaroni, kamakiminaba bosa namabanihi.\\
    \gll me hi-takora fawa na-ne ni-kani mati hi-na-hemete-mone okomise a-ati to-ha-hamaro-ni ka-maki-mina-haba bosa na-ma-haba-ni-hi\\
    3\textsc{pl}.\textsc{s} \textsc{oc}-be\_in\_hammock\_with.\textsc{f} disappear \textsc{aux}-\textsc{cont}.\textsc{f} \textsc{aux}.\textsc{f}-\textsc{cntrfact} 3\textsc{pl}.\textsc{s} \textsc{oc}-say-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 1\textsc{sg}.\textsc{poss}.aunt(\textsc{f}) \textsc{dup}-say \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{e}.\textsc{f}-\textsc{bkg}.\textsc{f} come-following-morning-\textsc{fut}.\textsc{f} get\_up\_early \textsc{aux}-back-\textsc{fut}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
    \glt `Her mother thought that she was gone because she was in someone's hammock with them, and that she would come back early in the morning, my aunt used to say.'
\z

\ea\label{ex:08-a89}
Hinaka yama soki oariemetemoneke.\\
    \gll hinaka yama soki ohari-hemete-mone-ke\\
    3\textsc{sg}.\textsc{poss} thing(\textsc{f}) be\_dark.\textsc{nfin} be\_one-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `She was gone for one night.'
\z

\ea\label{ex:08-a90}
Faya kama kiyomamemetemonehe.\\
    \gll faya ka-ma ka-yoma-ma-hemete-mone-he\\
    so come-back.\textsc{f} come-through-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `She came back, and came inside.'
\z

\ea\label{ex:08-a91}
Me so na me yobiri na mera fiya nama kama, kiyomama, mati ni ya tokoma towinamemetemonehe.\\
    \gll me soo na me yobi-ri na mera fiya na-ma ka-ma ka-yoma-ma mati ni ya to-ka-ma to-wina-ma-hemete-mone-he\\
    3\textsc{pl}.\textsc{s} urinate \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} go\_back\_and\_forth-\textsc{distr} \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{o} pass \textsc{aux}-back.\textsc{f} come-back.\textsc{f} come-through-back.\textsc{f} 3\textsc{sg}.\textsc{poss}.mother(\textsc{f}) to \textsc{adju} away-go-back.\textsc{f} away-lie\_in\_hammock-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `As people were going back and forth and urinating, she came past them. She came back and came inside. She went to her mother, and lay in her hammock.'
\z

\ea\label{ex:08-a92}
Towinama, wete towitia amo nisemetemoneke fahi.\\
    \gll to-wina-ma wete to-witia amo na-risa-hemete-mone-ke fahi\\
    away-lie\_in\_hammock-back.\textsc{f} return away-out.\textsc{f} sleep \textsc{aux}-down-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `She lay in her hammock, and turned and fell asleep.'
\z

\ea\label{ex:08-a93}
Faya yamata me hiwasia mati yamata wasia, mati yamata hiwasia kababone mati mowa hina i hiyarabanihi.\\
    \gll faya yamata me hi-wasiha mati yamata wasiha mati yamata hi-wasiha kaba-habone mati mowa hi-na ihi hiya-ra-haba-ni-hi\\
    so food(\textsc{f}) 3\textsc{pl}.\textsc{s} \textsc{oc}-cook.\textsc{f} 3\textsc{sg}.\textsc{poss}.mother(\textsc{f}) food(\textsc{f}) cook.\textsc{f} 3\textsc{sg}.\textsc{poss}.mother(\textsc{f}) food(\textsc{f}) \textsc{oc}-cook.\textsc{f} eat-\textsc{int}.\textsc{f} 3\textsc{sg}.\textsc{poss}.mother(\textsc{f}) shake \textsc{oc}-\textsc{aux}.\textsc{f} deed.\textsc{f} be\_bad-\textsc{neg}-\textsc{fut}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
    \glt `Her mother shook her so she would eat the food that her mother had prepared. She experienced a bad thing.'
\z

\ea\label{ex:08-a94}
Siraba ati nemetemone.\\
    \gll Siraba ati na-hemete-mone\\
    Siraba(\textsc{f}) say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `{``}Siraba,'' she said.'
\z

\ea\label{ex:08-a95}
Siraba titafihi ati nemetemone.\\
    \gll Siraba ti-tafi-hi ati na-hemete-mone\\
    Siraba(\textsc{f}) 2\textsc{sg}.\textsc{s}-wake\_up-\textsc{imp}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `{``}Siraba, wake up,'' she said.'
\z

\ea\label{ex:08-a96}
Mati hose hikanemetemone kita ya.\\
    \gll mati hose hi-ka-na-hemete-mone kita ya\\
    3\textsc{sg}.\textsc{poss}.mother(\textsc{f}) pull\_out \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} strength(\textsc{f}) \textsc{adju}\\
    \glt `Her mother shook her hard.'
\z

\ea\label{ex:08-a97}
Titafihi.\\
    \gll ti-tafi-hi\\
    2\textsc{sg}.\textsc{s}-wake\_up-\textsc{imp}.\textsc{f}\\
    \glt `{``}Wake up.'''
\z

\ea\label{ex:08-a98}
Tika yibote tiwa me kawa naba me ihi tikabibeya ati nemetemoneni.\\
    \gll ti-ka yibote tiwa me ka-waa nahaba me ihi ti-kaba-be-ya ati na-hemete-mone-ni\\
    2\textsc{sg}.\textsc{poss}-\textsc{poss} husband(\textsc{m}) 2\textsc{sg}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{comit}-stay at\_night.\textsc{f} 3\textsc{pl}.\textsc{poss} victim\_of.\textsc{f} 2\textsc{sg}.\textsc{s}-eat-\textsc{immed}.\textsc{f}-now say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Eat what your boyfriends who were with you all night killed.'''
\z

\ea\label{ex:08-a99}
Faya sa nemetemoneke fahi.\\
    \gll faya saa na-hemete-mone-ke fahi\\
    so vomit \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `Then she vomited.'
\z

\ea\label{ex:08-a100}
Faya ati yana kanemetemoneni.\\
    \gll faya ati yana ka-na-hemete-mone-ni\\
    so voice start \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `Then she spoke, sick.'
\z

\ea\label{ex:08-a101}
Ati yana kana, Ami ami ati nemetemoneke, me toma hinabanihi.\\
    \gll ati yana ka-na ami ami ati na-hemete-mone-ke me toma hi-na-haba-ni-hi\\
    voice start \textsc{comit}-\textsc{aux}.\textsc{f} mother(\textsc{f}) mother(\textsc{f}) say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} 3\textsc{pl}.\textsc{s} suck \textsc{oc}-\textsc{aux}-\textsc{fut}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
    \glt `She spoke. ``Mother, Mother,'' she said. They sucked on her to cure her.'
\z

\ea\label{ex:08-a102}
Me toma hinemetemoneke.\\
    \gll me toma hi-na-hemete-mone-ke\\
    3\textsc{pl}.\textsc{s} suck \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They sucked on her.'
\z

\ea\label{ex:08-a103}
Makibona fe nonofa toare, aawawa hinemetemone ahi.\\
    \gll maki-bona fee no-nofa to-ha-re a-awa-wa hi-na-hemete-mone ahi\\
    3\textsc{sg}.\textsc{poss}.husband(\textsc{m})-\textsc{int}.\textsc{m} 3\textsc{sg}.\textsc{o} \textsc{dup}-like \textsc{ch}-\textsc{aux}-\textsc{neg}.\textsc{m} \textsc{dup}-see-\textsc{dup} \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `The one who was supposed to marry her but didn't like her, felt pity for her.'
\z

\ea\label{ex:08-a104}
Toma hiremete awine?\\
    \gll toma hi-ra-hemete awine\\
    suck \textsc{oc}-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f} seem.\textsc{f}\\
    \glt `Didn't he suck on her?'
\z

\ea\label{ex:08-a105}
Toma hinemete awineke.\\
    \gll toma hi-na-hemete awine-ke\\
    suck \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `I guess he did suck on her.'
\z

\ea\label{ex:08-a106}
Me toma hina, me hikasomemetemonehe.\\
    \gll me toma hi-na me hi-ka-soma-hemete-mone-he\\
    3\textsc{pl}.\textsc{s} suck \textsc{oc}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-help-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `They sucked on her, to cure her.'
\z

\ea\label{ex:08-a107}
Fawa nisemetemoneke.\\
    \gll fawa na-risa-hemete-mone-ke\\
    disappear \textsc{aux}-down-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `She died.'
\z

\ea\label{ex:08-a108}
Fanaki fawa tokanemetemone ati nemetemone, okomise hiyari aati toimaro amani – Abi inamati ehene? – inamati ai hinabaro.\\
    \gll fanaki fawa to-ka-na-hemete-mone ati na-hemete-mone okomise hiyari a-ati to-hi-maro ama-ni abi inamati ehene inamati ahi hi-nahaba-haaro\\
    thigh.\textsc{f} disappear \textsc{ch}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 1\textsc{sg}.\textsc{poss}.aunt(\textsc{f}) speak.\textsc{nfin} \textsc{dup}-say \textsc{ch}-\textsc{aux}.\textsc{nom}.\textsc{f}-\textsc{fp}.\textsc{e}.\textsc{f} be-\textsc{bkg}.\textsc{f} father(\textsc{m}) spirit(\textsc{m}) because\_of.\textsc{m} spirit(\textsc{m}) have\_sex\_with \textsc{oc}-at\_night-\textsc{ip}.\textsc{e}.\textsc{f}\\
    \glt `She said her legs hurt, my aunt used to say – (Okomobi asks:) Father, because of the spirit? – because the spirit had sex with her all night.'
\z

\ea\label{ex:08-a109}
\textup{Okomobi:} Abi ahabimetemonehe?\\
    \gll abi ahabi-mete-mone-he\\
    father(\textsc{m}) end.\textsc{nom}.\textsc{f}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `Father, did she die?'
\z

\newpage
\ea\label{ex:08-a110}
Abemetemoneke.\\
    \gll ahaba-hemete-mone-ke\\
    die-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `She died.'
\z

\ea\label{ex:08-a111}
Abemetemone okomise hiyari aati toamaroni.\\
    \gll ahaba-hemete-mone okomise hiyari a-ati to-ha-hamaro-ni\\
    die-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 1\textsc{sg}.\textsc{poss}.aunt(\textsc{f}) speak.\textsc{nfin} \textsc{dup}-say \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{e}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `She died, my aunt used to say.'
\z

\ea\label{ex:08-a112}
Faya me yawa me kakimetemone amake ahi.\\
    \gll faya me yawa me ka-ki-mete-mone ama-ke ahi\\
    so 3\textsc{pl}.\textsc{s} be\_upset.\textsc{f} 3\textsc{pl}.\textsc{s} come-coming.\textsc{nom}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} be-\textsc{decl}.\textsc{f} then\\
    \glt `They grieved, and they came here.'
\z

\ea\label{ex:08-a113}
Me tokisemete me awineke ahi.\\
    \gll me to-ka-risa-hemete me awine-ke ahi\\
    3\textsc{pl}.\textsc{s} away-go-down-\textsc{fp}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{s} seem.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `I guess they went downstream.'
\z

\ea\label{ex:08-a114}
Me ka awa botemetemone okomise na hiamaro.\\
    \gll me ka awa bote-mete-mone okomise na hi-ha-hamaro\\
    3\textsc{pl} \textsc{poss} garden(\textsc{f}) old-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 1\textsc{sg}.\textsc{poss}.aunt(\textsc{f}) say \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{e}.\textsc{f}\\
    \glt `It's their old garden.'
\z

\ea\label{ex:08-a115}
Okomise yama yoyofi hiamaroke ahi, Fare Wari, ee Sibiri Wa Wari ka, tabora bote na awineke ahi.\\
    \gll okomise yama yo-yofi hi-ha-hamaro-ke ahi Fare\_Wari ee Sibiri\_Waa\_Wari ka tabora bote naa awine-ke ahi\\
    1\textsc{sg}.\textsc{poss}.aunt(\textsc{f}) thing(\textsc{f}) \textsc{dup}-show \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f} there Fare\_Wari(\textsc{f}) (pause) Sibiri\_Waa\_Wari(\textsc{m}) of village(\textsc{f}) old exist.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f} there\\
    \glt `My aunt showed it, Fare Wari, and there is an old village at Sibiri Wa Wari.'
\z

\ea\label{ex:08-a116}
Faya kona watematamona wate awaka fahi Noko Maiyi aati toimaro amake.\\
    \gll faya kona wata-himata-mona wate awa-ka fahi Noko\_Maiyi a-ati to-hi-maro ama-ke\\
    so plant.sp(\textsc{m}) grow-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} grow.\textsc{m} seem.\textsc{m}-\textsc{decl}.\textsc{m} there Noko\_Maiyi(\textsc{f}) \textsc{dup}-say \textsc{ch}-\textsc{aux}.\textsc{nom}.\textsc{f}-\textsc{fp}.\textsc{e}.\textsc{f} be-\textsc{decl}.\textsc{f}\\
    \glt `{``}The \textit{kona} that grew there still is growing there,'' Noko Maiyi used to say.'
\z

\ea\label{ex:08-a117}
\textup{Okomobi:} Abi Siraba e ka fanawirimete ama?\\
    \gll abi Siraba ee ka fanawiri-mete ama\\
    father(\textsc{m}) Siraba(\textsc{f}) 1\textsc{in} \textsc{poss} woman-\textsc{fp}.\textsc{n}.\textsc{f} \textsc{sec}\\
    \glt `Father, was Siraba a Jarawara?'
\z

\ea\label{ex:08-a118}
E ka fanawirimete awineni.\\
    \gll ee ka fanawiri-mete awine-ni\\
    1\textsc{in} \textsc{poss} woman-\textsc{fp}.\textsc{n}.\textsc{f} seem.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `I guess she was a Jarawara.'
\z

\ea\label{ex:08-a119}
Ha me siba remetemone me one e amake, hine me ebe kanemetemone mati.\\
    \gll haa me siba ra-hemete-mone me one ee ama-ke hine me ebe ka-na-hemete-mone mati\\
    \textsc{dem} 3\textsc{pl}.\textsc{s} be\_all\_right \textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl} other.\textsc{f} 1\textsc{in}.\textsc{s} be-\textsc{decl}.\textsc{f} 3.\textsc{refl} 3\textsc{pl}.\textsc{s} divide\_up \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `They split up a long time ago, and we became the Jarawaras.'
\z

\ea\label{ex:08-a120}
Hine me ebe kanemetemone, me yokana me amara me wina me awineke ahi.\\
    \gll hine me ebe ka-na-hemete-mone me yokana me ama-ra me wina me awine-ke ahi\\
    3.\textsc{refl} 3\textsc{pl}.\textsc{s} divide\_up \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl} true 3\textsc{pl}.\textsc{s} be-\textsc{neg}.\textsc{f} 3\textsc{pl}.\textsc{s} live.\textsc{f} 3\textsc{pl}.\textsc{s} seem.\textsc{f}-\textsc{decl}.\textsc{f} there\\
    \glt `They divided up long ago, and the ones who are not Jarawaras live [upstream].'
\z

\ea\label{ex:08-a121}
E ati nima me ati ra me awineke haha mati.\\
    \gll ee ati nima me ati ra me awine-ke haaha mati\\
    1\textsc{in}.\textsc{poss} voice like 3\textsc{pl}.\textsc{poss} voice \textsc{neg}.\textsc{f} 3\textsc{pl}.\textsc{poss} seem.\textsc{f}-\textsc{decl}.\textsc{f} this.\textsc{f} 3\textsc{pl}\\
    \glt `They don't speak our language.'
\z

\ea\label{ex:08-a122}
Me bana me fa fowemetemoneke.\\
    \gll me bana me faha\_fowe-mete-mone-ke\\
    3\textsc{pl}.\textsc{s} move\_away.\textsc{f} 3\textsc{pl}.\textsc{s} high\_water\_season(\textsc{f})-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They moved away. It was the high water season.'
\z

\ea\label{ex:08-a123}
Fa fowe ya me, fa fowe ya me bana me kisakia me, me kisakia me kanawa kanawa ataromatamonane.\\
    \gll faha\_fowe ya me faha\_fowe ya me bana me ka-risa-kia me me ka-risa-kia me kanawa kanawa ataro-mata-mona-ne\\
    high\_water\_season(\textsc{f}) \textsc{adju} 3\textsc{pl} high\_water\_season(\textsc{f}) \textsc{adju} 3\textsc{pl}.\textsc{s} move\_away.\textsc{f} 3\textsc{pl}.\textsc{s} come-down-coming.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} come-down-coming.\textsc{f} 3\textsc{pl}.\textsc{s} tree.sp(\textsc{m}) tree.sp(\textsc{m}) bark.\textsc{m}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `They moved to another place in the high water season. They came downstream. It was a \textit{kanawa} bark canoe.'
\z

\ea\label{ex:08-a124}
Kanawa ataro ya mera iso nematamonaka, fari, e one mati.\\
    \gll kanawa ataro ya mera iso na-himata-mona-ka faari ee one mati\\
    tree.sp(\textsc{m}) bark.\textsc{m} \textsc{adju} 3\textsc{pl}.\textsc{o} carry\_off \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} that\_one.\textsc{m} 1\textsc{in} other.\textsc{f} 3\textsc{pl}\\
    \glt `He carried them in a \textit{kanawa} bark canoe, our people.'
\z

\ea\label{ex:08-a125}
Afa memetemone me e amake.\\
    \gll afa me-mete-mone me ee ama-ke\\
    this.\textsc{f} 3\textsc{pl}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl}.\textsc{compl} 1\textsc{in}.\textsc{s} be-\textsc{decl}.\textsc{f}\\
    \glt `We are the descendants of those people.'
\z


\ea\label{ex:08-a126}
Me wina me awineke.\\
    \gll me wina me awine-ke\\
    3\textsc{pl}.\textsc{s} live.\textsc{f} 3\textsc{pl}.\textsc{s} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They [the others] live [upstream].'
\z

\ea\label{ex:08-a127}
Afa memetemone me e amake one oke, me wete tonamemetemone mati.\\
    \gll afa me-mete-mone me ee ama-ke o-ne o-ke me wete to-na-ma-hemete-mone mati\\
    this.\textsc{f} 3\textsc{pl}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl}.\textsc{compl} 1\textsc{in}.\textsc{s} be-\textsc{decl}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} 3\textsc{pl}.\textsc{s} return away-\textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `The people who went back were Jarawaras.'\footnote{The word \textit{one} is a form of the verb \textit{ati na} `say' even though in this morphological context not even the auxiliary \textit{na} is pronounced.}
\z

\ea\label{ex:08-a128}
Afa memete me amani.\\
    \gll afa me-mete me ama-ni\\
    this.\textsc{f} 3\textsc{pl}-\textsc{fp}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{s} be-\textsc{bkg}.\textsc{f}\\
    \glt `It was them.'
\z

\ea\label{ex:08-a129}
Me wete tonamene me wete tonamemetemone, biti mano ma tokahemona, ati ehene mera ha nemetemoneni, mera iso nari.\\
    \gll me wete to-na-ma-hene me wete to-na-ma-hemete-mone biti mano maa to-ka-ha-himona ati ehene mera haa na-hemete-mone-ni mera iso na-haari\\
    3\textsc{pl}.\textsc{s} return away-\textsc{aux}-back-\textsc{irr}.\textsc{f} 3\textsc{pl}.\textsc{s} return away-\textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{sg}.\textsc{poss}.son(\textsc{m}) arm.\textsc{m} be\_tired \textsc{ch}-\textsc{comit}-\textsc{aux}-\textsc{rep}.\textsc{m} voice because\_of.\textsc{m} 3\textsc{pl}.\textsc{o} call \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} 3\textsc{pl}.\textsc{o} carry\_off \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `They wouldn't have turned back, but they turned back. She called to them because her son, who was carrying them, his arms were tired.'
\z

\ea\label{ex:08-a130}
Me kikisama me mera iiso, me kimisakeha, me nahobasemetemone ahi me ebe noki na mati, fare ohari mera iso namako.\\
    \gll me ki-ka-risa-ma me mera i-iso me ka-misa-keha me naho-basa-hemete-mone ahi me ebe noki na mati fare ohari mera iso na-mako\\
    3\textsc{pl}.\textsc{s} \textsc{dup}-come-down-back 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{o} \textsc{dup}-carry\_off 3\textsc{pl}.\textsc{s} come-up-coming.\textsc{f} 3\textsc{pl}.\textsc{s} sit-edge-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then 3\textsc{pl}.\textsc{poss} \textsc{recip} wait\_for \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{poss} same.\textsc{m} be\_one.\textsc{m} 3\textsc{pl}.\textsc{o} carry\_off \textsc{aux}-for\_this\_reason.\textsc{m}\\
    \glt `They came up the bank. He was carrying them. They came onto the shore and stood on the edge, waiting for the others, because only one man was carrying them.'
\z

\ea\label{ex:08-a131}
Me tama me tonokosa nemetemone ahi me ati hiyawabani mati, e one mati e yokana mati.\\
    \gll me tama me to-naho-kosa na-hemete-mone ahi me ati hiyawa-haba-ni mati ee one mati ee yokana mati\\
    3\textsc{pl}.\textsc{s} be\_many.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{ch}-stand-middle.\textsc{list} \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then 3\textsc{pl}.\textsc{poss} voice have\_an\_idea-\textsc{fut}.\textsc{f}-\textsc{bkg}.\textsc{f} 3\textsc{pl}.\textsc{poss} 1\textsc{in} other.\textsc{f} 3\textsc{pl} 1\textsc{in} true 3\textsc{pl}\\
    \glt `Many people were left standing there. Then they changed their minds, our people, Jarawaras.'
\z

\ea\label{ex:08-a132}
Hee me ati nemetemoneni.\\
    \gll hee me ati na-hemete-mone-ni\\
    hey 3\textsc{pl}.\textsc{s} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Hey,'' they said.'
\z

\ea\label{ex:08-a133}
Ami omano ma tokahara oke.\\
    \gll ami o-mano maa to-ka-ha-hara o-ke\\
    mother(\textsc{f}) 1\textsc{sg}.\textsc{poss}-arm be\_tired \textsc{ch}-\textsc{comit}-\textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f} 1\textsc{sg}.\textsc{poss}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Mother, my arms are tired.'''
\z

\ea\label{ex:08-a134}
Faya te, faya me iso ni amani ati nematamonaha.\\
    \gll faya tee faya me iso ni ama-ni ati na-himata-mona-ha\\
    enough 2\textsc{pl} enough 3\textsc{pl}.\textsc{s} be\_carried\_off \textsc{aux}.\textsc{nfin} be-\textsc{bkg}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `{``}That's enough carrying people,'' he said.'
\z

\ea\label{ex:08-a135}
Okatao mano ma tokahamonaha.\\
    \gll okatao mano maa to-ka-ha-mona-ha\\
    1\textsc{sg}.\textsc{poss}.son(\textsc{m}) arm.\textsc{m} be\_tired \textsc{ch}-\textsc{comit}-\textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `{``}My son's arms are tired.'''
\z

\ea\label{ex:08-a136}
Ota wete tonamibana ati nemetemoneni, ha ni.\\
    \gll otaa wete to-na-mi-bana ati na-hemete-mone-ni haa ni\\
    1\textsc{ex}.\textsc{s} return away-\textsc{aux}-back.\textsc{f}-\textsc{fut} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} call \textsc{aux}.\textsc{nfin}\\
    \glt `{``}We are going back,'' she said, calling.'
\z

\ea\label{ex:08-a137}
Ee ota ota tokomibonehe.\\
    \gll ee otaa otaa to-ka-mi-bone-he\\
    yes 1\textsc{ex} 1\textsc{ex}.\textsc{s} away-go-back.\textsc{nom}.\textsc{f}-\textsc{int}.\textsc{f}-\textsc{dup}\\
    \glt `{``}Okay, we are going back.'''
\z

\ea\label{ex:08-a138}
Ota ota tokomibone.\\
    \gll otaa otaa to-ka-mi-bone\\
    1\textsc{ex} 1\textsc{ex}.\textsc{s} away-go-back.\textsc{nom}.\textsc{f}-\textsc{int}.\textsc{f}\\
    \glt `{``}We are going back.'''
\z

\ea\label{ex:08-a139}
Ota wete tonamibone okatao ati ne ati nemetemoneni.\\
    \gll otaa wete to-na-mi-bone okatao ati ne ati na-hemete-mone-ni\\
    1\textsc{ex}.\textsc{s} return away-\textsc{aux}-back.\textsc{nom}.\textsc{f}-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{poss}.son(\textsc{m}) say \textsc{aux}.\textsc{m} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}My son said we should go back,'' she said.'
\z

\ea\label{ex:08-a140}
Okatao mano ma tokahamona okatao ati ne ati nemetemone.\\
    \gll okatao mano maa to-ka-ha-mona okatao ati ne ati na-hemete-mone\\
    1\textsc{sg}.\textsc{poss}.son(\textsc{m}) arm.\textsc{m} be\_tired \textsc{ch}-\textsc{comit}-\textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{rep}.\textsc{m} 1\textsc{sg}.\textsc{poss}.son(\textsc{m}) say \textsc{aux}.\textsc{m} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `{``}My son said his arms are tired,'' she said.'
\z

\ea\label{ex:08-a141}
Ee te tokomahi, me ati na me ati yana tonemetemoneke fahi.\\
    \gll ee tee to-ka-ma-hi me ati na me ati yana to-na-hemete-mone-ke fahi\\
    yes 2\textsc{pl}.\textsc{s} away-go-back-\textsc{imp}.\textsc{f} 3\textsc{pl}.\textsc{s} say \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{poss} voice start \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `{``}All right, you go,'' they said. They were upset as they were speaking.'
\z


\ea\label{ex:08-a142}
Te ota kakabateba otani me ati nemetemoneni.\\
    \gll tee otaa ka-kaba-tee-haba otaa-ni me ati na-hemete-mone-ni\\
    2\textsc{pl}.\textsc{o} 1\textsc{ex}.\textsc{s} \textsc{comit}-eat-\textsc{hab}-\textsc{fut}.\textsc{f} 1\textsc{ex}.\textsc{s}-\textsc{bkg}.\textsc{f} 3\textsc{pl}.\textsc{s} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}We will eat you in the future,'' they said.'
\z

\newpage
\ea\label{ex:08-a143}
Te ota kakabatebisaba otani me ati nemetemoneni.\\
    \gll tee otaa ka-kaba-tee-bisa-haba otaa-ni me ati na-hemete-mone-ni\\
    2\textsc{pl}.\textsc{o} 1\textsc{ex}.\textsc{s} \textsc{comit}-eat-\textsc{hab}-also-\textsc{fut}.\textsc{f} 1\textsc{ex}.\textsc{s}-\textsc{bkg}.\textsc{f} 3\textsc{pl}.\textsc{s} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}We will eat you, too,'' they said.'
\z

\ea\label{ex:08-a144}
Ee tokatimi ka ya me kisemetemoneke ha ya.\\
    \gll ee to-ka-timi ka ya me ka-risa-hemete-mone-ke haa ya\\
    yes away-go-upstream.\textsc{nom}.\textsc{f} \textsc{loc} \textsc{adju} 3\textsc{pl}.\textsc{s} come-down-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} \textsc{dem} \textsc{adju}\\
    \glt `Yes, they came downstream from up there upstream.'
\z

\ea\label{ex:08-a145}
Kowani ka ama nimete ahi.\\
    \gll kowani ka ama ni-mete ahi\\
    other\_side \textsc{loc} be \textsc{aux}.\textsc{nom}.\textsc{f}-\textsc{fp}.\textsc{n}.\textsc{f} there\\
    \glt `It was on the other side.'
\z

\ea\label{ex:08-a146}
Ha ya me mera iseriwakemata awaka ahi.\\
    \gll haa ya me mera ise-riwaha-ke-himata awa-ka ahi\\
    \textsc{dem} \textsc{adju} 3\textsc{pl} 3\textsc{pl}.\textsc{o} drop\_off-across-coming-\textsc{fp}.\textsc{n}.\textsc{m} seem.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `He brought them across to here.'
\z

\ea\label{ex:08-a147}
Ee me me hekanakosemetemoneke.\\
    \gll ee me me hi-ka-kana-kosa-hemete-mone-ke\\
    yes 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-leave-middle-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `Right, they left them behind.'
\z


\ea\label{ex:08-a148}
Faya me, me ma nisimetemoneke Wara amake haro, okomise hiyari aati toamaroke.\\
    \gll faya me me maa na-risi-mete-mone-ke Wara ama-ke haaro okomise hiyari a-ati to-ha-hamaro-ke\\
    so 3\textsc{pl} 3\textsc{pl}.\textsc{s} stop \textsc{aux}-down.\textsc{nom}.\textsc{f}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} Wara(\textsc{f}) be-\textsc{decl}.\textsc{f} that\_one.\textsc{f} 1\textsc{sg}.\textsc{poss}.aunt(\textsc{f}) speak.\textsc{nfin} \textsc{dup}-say \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `Then they stopped coming downstream at Wara, my aunt used to say.'
\z

\ea\label{ex:08-a149}
Me wiwina toemete me awineke fahi.\\
    \gll me wi-wina to-ha-hemete me awine-ke fahi\\
    3\textsc{pl}.\textsc{s} \textsc{dup}-live \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{s} seem.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `Then they lived there.'
\z

\ea\label{ex:08-a150}
E tamemetemone e, bote me ya e tabakere.\\
    \gll ee tama-hemete-mone ee bote me ya ee taba-ke-re\\
    1\textsc{in}.\textsc{s} be\_many-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 1\textsc{in} old 3\textsc{pl} \textsc{adju} 1\textsc{in}.\textsc{s} be\_together-\textsc{decl}.\textsc{f}-\textsc{neg}.\textsc{f}\\
    \glt `We were numerous, but the old people aren't with us anymore.'
\z

\ea\label{ex:08-a151}
Me te aawa toaramaro te amani.\\
    \gll me tee a-awa to-ha-ra-hamaro tee ama-ni\\
    3\textsc{pl}.\textsc{o} 2\textsc{pl}.\textsc{s} \textsc{dup}-see \textsc{ch}-\textsc{aux}-\textsc{neg}-\textsc{fp}.\textsc{e}.\textsc{f} 2\textsc{pl}.\textsc{s} \textsc{sec}-\textsc{bkg}.\textsc{f}\\
    \glt `You didn't see them.'
\z

\ea\label{ex:08-a152}
Yobe wahi toro ateate na toemari ama me ihi ayaka ini me hiri na mati.\\
    \gll yobe waahi toro ate-ate na to-ha-himari ama me ihi ayaka ini me hiri na mati\\
    house(\textsc{m}) stand.\textsc{m} inside.\textsc{m} \textsc{dup}-make\_noise.\textsc{list} \textsc{aux} \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{e}.\textsc{m} \textsc{sec} 3\textsc{pl}.\textsc{poss} because\_of.\textsc{f} feast(\textsc{f}) name.\textsc{f} 3\textsc{pl}.\textsc{s} say \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `There was a lot of noise in the longhouse because of them, because they were asking for a feast.'\footnote{This sentence is in eyewitness tense, indicating that Yowao was alive when people still lived in longhouses.}
\z
