\noindent I will tell about Siraba, who a spirit took away.

\noindent They went after \textit{siro} fruits. The two of them went. A spirit went to them. The spirit went there. He appeared where they were as they were picking up \textit{siro} fruits, as they were piling the \textit{siro} fruits up.

\noindent ``My older brother has come,'' Siraba's cousin said. Siraba looked quickly.

\noindent ``Are there ripe \textit{siro} fruits?'' he said.

\noindent ``The \textit{siro} fruits are ripe. The \textit{siro} fruits are there in a pile,'' she said. ``Look at the \textit{siro} fruits,'' she said.

\noindent He picked up two of the \textit{siro} fruits. He stood there gnawing on the \textit{siro} fruits. 	
	
\noindent ''Brother, where are you going?'' she said.

\noindent ``I'm going to Boto Komaha.'' He threw a \textit{siro} seed and hit a tree [to show the direction he was heading]. ``I'm going to Boto Komaha,'' he said.

\noindent ``It's a spirit,'' the other one [Siraba] said. ``You are calling him your brother, but he is not your brother,'' she said. They recognized him.

\noindent ``Brother, you stay here. We are going to go get [palm leaves to make] baskets for the \textit{siro} fruits.''

\noindent The women went. He watched them, and he went after them to see them. They cut off two palm leaves, and the leaves fell on the ground. They didn't stay there. They didn't keep on cutting off leaves. They went running away.

\noindent They came out on the other side. They came out, and went running to where Siraba's father was sitting, weaving a basket. As he sat there weaving the basket, his daughter appeared, running.

\noindent ``Father, is it you?'' she said.

\noindent ``It is me,'' he said.

\noindent ``Father, there is a spirit following us, and he is coming this way,'' she said.

\noindent ``I guess it is a spirit,'' he said.

\noindent The spirit didn't take long to appear. He came running, and Siraba's father surprised him.

\noindent ``What are you doing coming here?'' he said, Siraba's father.

\noindent ``I'm not chasing them,'' he said.

\noindent ``You have come chasing my daughter. I see you. Go away. Go back to your own place,'' he said.

\noindent So he went back. He didn't go back. He went back, but he came around. He came out again and went there. He came out again.

\noindent In the late afternoon [in the village] the girls chased the children. When Siraba's father saw him, didn't he send [his familiar spirit] against him? Apparently he didn't send [his familiar spirit] against him, even though he was a powerful shaman.

\noindent The spirit went there and waited for them. Their laughing could be heard as they chased each other. They were chasing them. She was always with her cousin. She was chasing them. She ran after them. She chased them behind the longhouse.

\noindent The others were in their hammocks inside the longhouse. She chased them around. The spirit was standing there, and grabbed her. She thought he was one of their men, her companion's older brother. She ran to him. So he grabbed her, since she thought he was one of their men.

\noindent ``Who are you?'' she said. She heard his name. I don't know his name.

\noindent Then her cousin came running. ``Who are you with?'' she said. ``Is that my brother?'' she said.

\noindent ``I think it is your brother,'' she said.

\noindent ``Brother, is that you?'' she said.

\noindent ``It is me,'' he said. I guess he had a different voice. ``It is me.''

\noindent ``You aren't my brother, you are just making yourself out to be my brother. My brother is in his hammock. Wait, I will go take a look at my brother. You two stay here.''

\noindent She went running home and went inside. Then he disappeared with Siraba. As he was disappearing with Siraba, she came home and asked her mother.

\noindent ``Mother,'' she said. ``Mother, is my older brother in his hammock?''

\noindent ``My son is in his hammock,'' she said.

\noindent ``There is a spirit holding onto Siraba there on the patio. I'm going back to see her,'' she said. 	So she came back running. She came out of the longhouse and went to where the two of them had been standing. She went there, but there was no one there. She whistled for them all over. She could be heard whistling all over, as she looked for them. She finally came back.

\noindent Okomobi: Father, if there were two of them, would he have taken both of them?

\noindent He could have taken both of them if he had waited.

\noindent Then she went inside, because she was afraid. She was afraid. Siraba did not come back. When morning came, Siraba had spent the night with the spirit. They missed her. Her mother thought that she was gone because she was in someone's hammock, and that she would come back early in the morning, my aunt used to say. She was gone for one night.

\newpage
\noindent Siraba came back, and came inside. As people were going back and forth and urinating, she came past them. She came back and came inside. She went to her mother, and lay in her hammock. She lay in her hammock, and turned and fell asleep.

\noindent Her mother shook her so she would eat the food that she had prepared. She experienced a bad thing. ``Siraba,'' she said. ``Siraba, wake up,'' she said. Her mother shook her hard. ``Wake up. Eat what your boyfriend who was with you all night killed.''

\noindent Then she vomited. Then she spoke, sick. She spoke. ``Mother, Mother,'' she said.

\noindent They sucked on her to cure her. The one who was supposed to marry her but didn't like her, felt pity for her. Didn't he suck on her? I guess he did suck on her. They sucked on her, to cure her. She died.

\noindent She said her legs hurt, my aunt used to say.

\noindent Okomobi:  Father, because of the spirit?

\noindent Because the spirit had sex with her all night.

\noindent Okomobi: Father, did she die?

\noindent She died, my aunt used to say. They grieved, and they came here. I guess they went downstream. It's their old garden. My aunt showed it, Fare Wari, and there is an old village at Sibiri Wa Wari. The \textit{kona} that grew there still is growing there, Noko Maiyi used to say.

\noindent Okomobi: Father, was Siraba a Jarawara?

\noindent I guess she was a Jarawara. They split up a long time ago, and we became the Jarawaras. They divided up long ago, and the ones who are not Jarawaras live [upstream]. They don't speak our language.

\noindent They moved away. It was the high water season. They moved to another place in the high water season. They came downstream. It was a \textit{kanawa} bark canoe. He carried them in a \textit{kanawa} bark canoe, our people. We are the descendants of those people. They [the others] live [upstream]. The people who went back were Jarawaras. It was them. They wouldn't have turned back, but they turned back.

\noindent She called to them because her son, who was carrying them in the canoe, his arms were tired. They came up the bank. He was carrying them. They came onto the shore and stood on the edge, waiting for the others, because only one man was carrying them. Many people were left standing there. Then they changed their minds, our people, Jarawaras.

\noindent ``Hey,'' they said.

\noindent ``Mother, my arms are tired. That's enough carrying people,'' he said.

\noindent ``My son's arms are tired. We are going back,'' she said, calling.

\noindent ``Okay, we are going back.''

\noindent ``My son said we should go back,'' she said. ``My son said his arms are tired,'' she said.

\noindent ``All right, you go,'' they said. They were upset as they were speaking. ``We will eat you in the future,'' they said.

\noindent ``We will eat you, too,'' they said.

\noindent Yes, they came downstream from up there upstream. It was on the other side of the Purus River. He brought them across to here. Right, they left them behind. Then they stopped coming downstream at Wara Lake, my aunt used to say. Then they lived there. We were numerous, but the old people aren't with us anymore. You didn't see them. There was a lot of noise in the longhouse because of them, because they were asking for a feast.
