\noindent It was her future sister-in-law. She called her future sister-in-law to go out. It was \textit{bihi}. It was \textit{bihi} palm leaves, to be woven [into fans].

\noindent ``Sister-in-law, let's all go together.''

\noindent ``My mother won't let me.''

\noindent ``Let's go together and play all over. We aren't men. We are all women. When we have gone and picked the \textit{bihi} leaves and brought back the \textit{bihi} leaves, my mother and the others will weave them.''

\noindent ``Your sister-in-law is calling you to go with her. Go with your sister-in-law. If you don't go, she won't call you anymore,'' her mother said.

\noindent Her sister-in-law called her, and they went. They went out into the forest. They picked \textit{bihi} leaves. They picked \textit{bihi} leaves as they came back. They came back picking \textit{bihi} leaves. As they were coming back picking \textit{bihi} leaves, she found some \textit{bihi} fruits and cut the bunch off. She cut off the bunch of \textit{bihi} fruits, and was holding it in her hand. She picked \textit{bihi} leaves and carried them on her shoulder. She came holding the bunch of \textit{bihi} fruits in her hand.

\noindent ``Are you finished?''

\noindent ``Yes, we are finished. Let's go. Let's not separate the \textit{bihi} leaves into strips, let's take the \textit{bihi} leaves home.''

\noindent She wanted to give her \textit{bihi} fruits to her mother. Her sisters-in-law played a joke on her and ate the \textit{bihi} fruits.

\noindent She was upset because they ate them. ``You are eating the \textit{bihi} fruits that are for my mother.''

\noindent They didn't pay any attention to what she was saying. They pulled off the \textit{bihi} fruits. They ate the \textit{bihi} fruits, cracking them in their mouths.

\noindent ``You go ahead and stay here. Go ahead and eat them.''

\noindent ``Go ahead, let her be mad. \textit{Hehei}. Look at her, she's mad.''

\noindent She went back mad. She went back. They were left behind running after each other. Their voices came to her and she stopped to listen. Their voices could be heard. She went on.

\noindent While the girls were playing, the jaguars were also playing. There were two logs across the trail, and the two jaguars were on top of them playing. ``\textit{Ooo}.'' The male was playing with its mate. When she saw it she left and came back. She came running back.

\noindent ``Someone is coming. Someone is coming,'' they said.

\noindent ``Don't say `Someone's coming.' Let's go this way. There are jaguars playing. There are jaguars playing on our trail. Let's go the other way. We will come out on the other side of them.''

\noindent ``She is lying. She's saying that because she saw her boyfriend,'' they said. ``She is coming here saying that because she saw her boyfriend who had come after her.''

\noindent ``That's enough talk. I am telling you to come with me, and you aren't listening. There are jaguars playing there. I am not fooling you. There are jaguars playing on the trail. They are playing on top of two logs on the trail.''

\noindent ``She is coming saying that because she saw her boyfriend,'' they sang. They came singing.

\noindent ``Go ahead back singing.''

\noindent She was behind them as they went back. As they went, there was noise, \textit{kisiririri}, \textit{toko} [sounds of moving leaves, and of jaguars' paws hitting flesh].

\noindent ``\textit{Aya}, \textit{aya},'' \textit{toko}, ``\textit{aya}.'' They were crying out. As they were crying out, the jaguar killed them. The jaguar killed them and they were crying out.

\noindent She came. She was wearing cotton twine adornments, and she undid the cotton and wrapped it around her feet. It was a \textit{baro} tree. She climbed up the \textit{baro}, which was this thick [thick enough to be strong, but thin enough to climb]. She climbed up.

\noindent The jaguars ate her sisters-in-law. She looked down as she heard them eating.

\noindent ``They wouldn't listen to me.''

\noindent She waited until their stomachs were full, then she called out. The sun went down. She was afraid as the sun went down. ``What am I going to do? Where can I go? I'm going to call out,'' she said. ``I think they are full now. I think their stomachs are full.''

\noindent Her older brother was making arrow poison. He ground the \textit{iha} in the mortar. He stirred the poison. He dried the arrow points. He put more poison on the arrow points. The arrow points were dry. The dart points were dry, too.

\noindent The sun was going down. When the sun was going down, he went to urinate. He went, and untied his penis. He stood urinating.

\noindent ``Heei,'' someone said. It was her calling. She was calling, and her older brother heard her calling.

\noindent ``Mother, Mother, someone is calling there, where my sister went with the others,'' he said.

\noindent ``Go look at my daughter.''

\noindent ``Wait.'' He went. He put the arrow points in the sheath. He hung up the darts and went.

\noindent ``I'm going to take a look at them. Let's go together. Come along.''

\noindent The two of them went.

\noindent ``Hee,'' she said.

\noindent He called out. ``Just let her call,'' he said, and he went.

\noindent They went on. ``Someone is calling up there. I think she is standing up there in a tree.''

\noindent As her calling could be heard, they came. The two jaguars were sitting eating. They were eating the bellies. They were eating their bellies, the girls' bellies. They were sitting eating the girls' bellies. They were full. The had eaten their livers, their intestines, the fat of their intestines.

\noindent He came to where they were sitting eating. The jaguar looked. It looked at him as he appeared. It was sitting eating. It got up quickly. He quickly pulled out an arrow. The other one behind him pulled out an arrow, too. They charged them, and they shot them with arrows.

\noindent The jaguar was full. It was staggering. It went a ways and collapsed. The other one shot the other one, and it went a ways and collapsed. They each killed one. They died.

\noindent ``Sister.''

\noindent ``Brother, Brother,'' she said.

\noindent ``What?''

\noindent ``Is it you?''

\noindent ``Yes, it's me.''

\noindent ``Is it really you?''

\noindent ``It's really me. Come down.''

\noindent So his sister came down. She thought her brother was someone else. She looked at her brother, holding on up there. ``Oh, it's my brother.'' She came down.

\noindent The one who was with him was standing a ways away, his brother-in-law. She came down.	``Brother, they wouldn't listen to me, and the jaguars ate them.''

\noindent She cried. She could be heard crying. ``I was coming back. I was holding \textit{bihi} fruits in my hand for my mother, and they took them from me and made me mad, so I came back. As I was coming back, I came upon the jaguars here. I came upon them playing, their log is across the path here. They are lying there by their log.''

\noindent Then she cried. She could be heard crying. ``I went back there. I wanted for us to go away, but they wouldn't listen to me. They came singing here. They came along singing. Now there is no more work for them to do.''

\noindent ``Come on, you go first.''

\noindent The two jaguars were lying there. They left them lying on the ground.

\noindent ``Come on, you two go on first.''

\noindent So they came back. She was crying for her sisters-in-law along the way. She came back crying. They arrived back. She was crying.

\noindent When the mothers heard about their children, they cried, all of them. ``\textit{Aya}, my daughter, my daughter,'' they said, crying.

\noindent ``I wanted to save them. They refused. They almost caused me misfortune. I came upon the jaguars playing,'' she said, crying. She was crying while she talked. ``They said I was fooling them. They said some men were coming after us. I went running. I wanted to take them, but they refused. They passed by me, and the jaguars killed them. I came on behind them, and I climbed a tree. I climbed the tree and heard them eating as I sat there. After the sun was setting I started calling out. I started calling out when the sun was going down.''

\noindent ``Your brother heard you calling when he went to urinate. I sent your brother and he went, and he shot the jaguar with an arrow.''

\noindent ``There were two of them. The two of them shot the two of them with arrows, and they are lying on the ground.''

\noindent Then they cried. They got up early to go after their bones. They got up early because of them. They carried them in their own hammocks. They buried them, all of them. They cried. One of them remembered her daughter, and she cried, missing her daughter. They cried. They stopped crying, and slept. They remembered them again.

\noindent ``The women will not go out in the forest anymore.''

\noindent The girls were true jaguar food.
