\ea\label{ex:10-a1}
Toto abono faneibona fatibone kanike, hinakaso biti, hinakaso ha ne hinakaso, Aso, aso, aso tikoto ota famabone onara oke.\\
    \gll toto abono fana-hibona fati-bone ka-nike hinakaso biti hinakaso haa ne hinakaso aso aso aso tikoto otaa fama-habone o-na-hara o-ke\\
    bamboo.sp(\textsc{f}) spirit be\_married-\textsc{int}.\textsc{m} 3\textsc{sg}.\textsc{poss}.wife(\textsc{f})-\textsc{int}.\textsc{f} \textsc{comit}-ask\_for.\textsc{m} 3\textsc{sg}.\textsc{poss}.father's\_sister(\textsc{f}) 3\textsc{sg}.\textsc{poss}.son(\textsc{m}) 3\textsc{sg}.\textsc{poss}.father's\_sister(\textsc{f}) call \textsc{aux}.\textsc{m} 3\textsc{sg}.\textsc{poss}.father's\_sister(\textsc{f}) father's\_sister(\textsc{f}) father's\_sister(\textsc{f}) father's\_sister(\textsc{f}) 2\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) 1\textsc{ex}.\textsc{s} be\_two-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{s}-say-\textsc{ip}.\textsc{e}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `The spirit of \textit{toto} bamboo wanted to get married, so he asked for a wife. He called his father's sister. ``Aunt, I want to be with your daughter.'''\footnote{The words \textit{hinakaso biti} are a false start.}\footnote{When I first published this story (Vogel 2012), I assumed that \textit{Toto Abono} was a human being. But \textit{toto abono} also means `spirit of \textit{toto} bamboo'. After reading the Portuguese translation of the Jamamadi version of the story in \citet[163f]{Shiratori:2018kn}, in which \textit{hado abono} is `spirit of \textit{hado} bamboo', and only appears as a human being, I wondered whether this might be true for the Jarawara version as well. I asked Bibiri, and he wasn't sure, but he said \textit{toto abono} does seem to be a spirit. I have made the change here, but some doubt remains: if he was hiding his true identity from people, it seems strange to me that he would be known as \textit{Toto Abono}, and addressed as such, as in for example sentence (\ref{ex:10-a31}).}\footnote{The NP \textit{toto abono} consists of the inalienably possessed noun \textit{abono}, which has no marking for gender, and the possessor \textit{toto}, which is a feminine noun. The NP, however, is masculine, as seen in the agreement throughout the text. The reason for this is that all spirits are assumed to be males, unless they are specified with the addition of the noun \textit{fana} as being females, as is the case, for example with \textit{tokowisa abono fana} in \textref{text:12} \textit{Maiko}, line (\ref{ex:12-a216}).}
\z

\ea\label{ex:10-a2}
Otefe wawasi wawata nofara, oko fowa hiti re kawita.\\
    \gll o-tefe wawasi wa-wata nofa-ra o-ka fowa hiti re ka-ita\\
    1\textsc{sg}.\textsc{poss}-food person\_who\_cooks \textsc{dup}-exist recently-\textsc{neg}.\textsc{f} 1\textsc{sg}.\textsc{poss}-\textsc{poss} manioc(\textsc{m}) be\_grated \textsc{neg}.\textsc{m} \textsc{comit}-sit.\textsc{list}\\
    \glt `{``}There is no one to cook my food, and my manioc is sitting in the field without being grated.'''
\z


\ea\label{ex:10-a3}
Faya bati kamematamona fahi.\\
    \gll faya bati ka-ma-himata-mona fahi\\
    so 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `Then her father came home.'
\z

\ea\label{ex:10-a4}
Tikoto bitimi kanike, wa fari.\\
    \gll tikoto bitimi ka-nike waa faari\\
    2\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) 2\textsc{sg}.\textsc{poss}.nephew(\textsc{m}) \textsc{comit}-ask\_for.\textsc{m} stand.\textsc{cont}.\textsc{m} that\_one.\textsc{m}\\
    \glt `{``}Your nephew asked for your daughter. He's standing there.'''
\z

\ea\label{ex:10-a5}
Okoto kanikanoho?\\
    \gll okoto ka-nika-no-ho\\
    1\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) \textsc{comit}-ask\_for.\textsc{nom}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `{``}Did he really ask for my daughter?'''
\z

\ea\label{ex:10-a6}
Okoto hiwakamahi.\\
    \gll okoto hi-ka-ka-ma-hi\\
    1\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) \textsc{oc}-\textsc{comit}-go-back-\textsc{imp}.\textsc{f}\\
    \glt `{``}Let him take my daughter.'''
\z

\ea\label{ex:10-a7}
Hiwakamahi.\\
    \gll hi-ka-ka-ma-hi\\
    \textsc{oc}-\textsc{comit}-go-back-\textsc{imp}.\textsc{f}\\
    \glt `{``}Let him take her.'''
\z

\ea\label{ex:10-a8}
Faya bite ta nematamona fahi, bite kaiwarabanihi.\\
    \gll faya bite taa na-himata-mona fahi bite ka-hiwa-ra-haba-ni-hi\\
    so 3\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) give \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then 3\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) \textsc{comit}-have\_children-\textsc{neg}-\textsc{fut}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
    \glt `So he gave away his daughter, but she didn't have any children.'
\z

\ea\label{ex:10-a9}
Faya toto abono tokomematamonane.\\
    \gll faya toto abono to-ka-ma-himata-mona-ne\\
    so bamboo.sp(\textsc{f}) spirit away-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `So the \textit{toto} spirit went home.'
\z

\ea\label{ex:10-a10}
Tokome fare taboro ya tokome, kobo toname, kobo tonamematamonaka.\\
    \gll to-ka-me fare taboro ya to-ka-me kobo to-na-me kobo to-na-ma-himata-mona-ka\\
    away-go-back.\textsc{m} same.\textsc{m} place.\textsc{m} \textsc{adju} away-go-back.\textsc{m} arrive away-\textsc{aux}-back.\textsc{m} arrive away-\textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went home. He went to his own home. He arrived back home.'
\z

\ea\label{ex:10-a11}
Kobo toname ama.\\
    \gll kobo to-na-me ama\\
    arrive away-\textsc{aux}-back.\textsc{m} \textsc{sec}\\
    \glt `He arrived back home.'
\z

\ea\label{ex:10-a12}
Yama, iyawa fati siremetemone fahi.\\
    \gll yama iyawa fati sira-hemete-mone fahi\\
    thing(\textsc{f}) grated\_mash(\textsc{f}) 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) toast-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `His wife toasted manioc mash.'
\z

\ea\label{ex:10-a13}
Iyawa fati sira, fatibonehe.\\
    \gll iyawa fati sira fati-bone-he\\
    grated\_mash(\textsc{f}) 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) toast.\textsc{f} 3\textsc{sg}.\textsc{poss}.wife(\textsc{f})-\textsc{int}.\textsc{f}-\textsc{dup}\\
    \glt `His wife toasted manioc mash, his bride.'
\z

\ea\label{ex:10-a14}
Faya iyawa siremetemone fahi.\\
    \gll faya iyawa sira-hemete-mone fahi\\
    so grated\_mash(\textsc{f}) toast-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `She toasted manioc mash.'
\z

\ea\label{ex:10-a15}
Faya me tafemetemone fahi, tafamarabanihi.\\
    \gll faya me tafa-hemete-mone fahi tafa-ma-ra-haba-ni-hi\\
    so 3\textsc{pl}.\textsc{s} eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then eat-back-\textsc{neg}-\textsc{fut}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
    \glt `Then they ate, but she didn't eat again.'
\z

\ea\label{ex:10-a16}
Faya me tafa me amo kana me, hitakoraremetemoneke.\\
    \gll faya me tafa me amo ka-na me hi-takora-ra-hemete-mone-ke\\
    so 3\textsc{pl}.\textsc{s} eat.\textsc{f} 3\textsc{pl}.\textsc{s} sleep \textsc{comit}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{oc}-be\_in\_hammock\_with-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They ate. The two of them slept. He didn't sleep in the same hammock with her.'
\z

\ea\label{ex:10-a17}
Hitakorara, hine ya winaha ni ya winabase, ai ni hinofaremetemoneke.\\
    \gll hi-takora-ra hine ya winaha ni ya wina-base ahi ni hi-nofa-ra-hemete-mone-ke\\
    \textsc{oc}-be\_in\_hammock\_with-\textsc{neg}.\textsc{f} only \textsc{adju} lie\_in\_hammock.\textsc{f} to \textsc{adju} lie\_in\_hammock-edge.\textsc{m} have\_sex\_with \textsc{aux}.\textsc{nfin} \textsc{oc}-want-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He didn't sleep in the same hammock with her. He just lay in the hammock beside where she was lying. He didn't want to have sex with her.'\footnote{The verb \textit{winaha} is a relative clause, as seen by the fact that it occurs with \textit{ni ya} (\cite[142f]{Vogel:2022du}). Since it occurs with no NP, it is evidence for Dixon's (\citeyear[525]{Dixon:2004xr}) contention that relative clauses are internally headed.}
\z

\ea\label{ex:10-a18}
Ai ni hinofara, faya toto abono, Oko, oko iyawa abarake.\\
    \gll ahi ni hi-nofa-ra faya toto abono o-ka o-ka iyawa ahaba-hara-ke\\
    have\_sex\_with \textsc{aux}.\textsc{nfin} \textsc{oc}-want-\textsc{neg}.\textsc{f} so bamboo.sp(\textsc{f}) spirit 1\textsc{sg}.\textsc{poss}-\textsc{poss} 1\textsc{sg}.\textsc{poss}-\textsc{poss} grated\_mash(\textsc{f}) end-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He didn't want to have sex with her. ``My manioc mash is all gone.'''
\z

\ea\label{ex:10-a19}
Iyawa abarake.\\
    \gll iyawa ahaba-hara-ke\\
    grated\_mash(\textsc{f}) end-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}The manioc mash is all gone.'''
\z

\ea\label{ex:10-a20}
Fowa bore tikanabone onara oke.\\
    \gll fowa bore ti-ka-na-habone o-na-hara o-ke\\
    manioc(\textsc{m}) pull\_out 2\textsc{sg}.\textsc{s}-\textsc{comit}-\textsc{aux}-\textsc{int}.\textsc{f} 1\textsc{sg}.\textsc{s}-say-\textsc{ip}.\textsc{e}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I want you to pull up some manioc.'''
\z

\ea\label{ex:10-a21}
Fowa bore tikana fowa hiti.\\
    \gll fowa bore ti-ka-na fowa hiti\\
    manioc(\textsc{m}) pull\_out 2\textsc{sg}.\textsc{s}-\textsc{comit}-\textsc{aux}.\textsc{list} manioc(\textsc{m}) grate.\textsc{list}\\
    \glt `{``}When you have pulled up the manioc, grate the manioc, too.'''
\z

\ea\label{ex:10-a22}
Fowa hiti tinisahi.\\
    \gll fowa hiti ti-na-risa-hi\\
    manioc(\textsc{m}) grate 2\textsc{sg}.\textsc{s}-\textsc{aux}-down-\textsc{imp}.\textsc{f}\\
    \glt `{``}Grate the manioc.'''
\z

\ea\label{ex:10-a23}
Fowa hiti tinisa iyawa nanafi nama nabanake.\\
    \gll fowa hiti ti-na-risa iyawa na-nafi na-nama na-habana-ke\\
    manioc(\textsc{m}) grate 2\textsc{sg}.\textsc{s}-\textsc{aux}-down.\textsc{f} grated\_mash(\textsc{f}) \textsc{dup}-be\_big \textsc{aux}-\textsc{super} \textsc{aux}-\textsc{fut}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}When you grate the manioc, there will be a lot of manioc mash.'''
\z

\ea\label{ex:10-a24}
Faya fowa bore na, fowa iso na fowa sowe nemetemoneke.\\
    \gll faya fowa bore na fowa iso na fowa sowe na-hemete-mone-ke\\
    so manioc(\textsc{m}) pull\_out \textsc{aux}.\textsc{f} manioc(\textsc{m}) carry\_off \textsc{aux}.\textsc{f} manioc(\textsc{m}) scrape \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `So she pulled up manioc. She carried the manioc. She peeled the manioc.'
\z

\ea\label{ex:10-a25}
Fowa  sowe na, fowa sowe na fara kawiabanihi.\\
    \gll fowa sowe na fowa sowe na fara kawia-ba-ni-hi\\
    manioc(\textsc{m}) scrape \textsc{aux}.\textsc{f} manioc(\textsc{m}) scrape \textsc{aux}.\textsc{f} same.\textsc{f} accompanying\_food-\textsc{fut}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
    \glt `She peeled the manioc. She peeled the manioc that was to be the starch eaten with herself.'
\z

\newpage
\ea\label{ex:10-a26}
Fara kawiabani ra ai nemetemoneni.\\
    \gll fara kawia-ba-ni ra ahi na-hemete-mone-ni\\
    same.\textsc{f} accompanying\_food-\textsc{fut}-\textsc{ip}.\textsc{n}.\textsc{f} \textsc{o} work\_on \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `She made the starch that was eaten with herself.'
\z

\ea\label{ex:10-a27}
Fowa hiti nemetemoneke.\\
    \gll fowa hiti na-hemete-mone-ke\\
    manioc(\textsc{m}) grate \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `She grated the manioc.'
\z

\ea\label{ex:10-a28}
Fowa hiti na, fowa hiti nemetemoneke.\\
    \gll fowa hiti na fowa hiti na-hemete-mone-ke\\
    manioc(\textsc{m}) grate \textsc{aux}.\textsc{f} manioc(\textsc{m}) grate \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `She grated the manioc.'
\z

\ea\label{ex:10-a29}
Fowa hiti ni ahabe ya, iyawa, fowa nano na, fowa nano nemetemoneke fahi.\\
    \gll fowa hiti ni ahabe ya iyawa fowa nano na fowa nano na-hemete-mone-ke fahi\\
    manioc(\textsc{m}) be\_grated \textsc{aux}.\textsc{nfin} end.\textsc{nom}.\textsc{m} \textsc{adju} grated\_mash(\textsc{f}) manioc(\textsc{m}) squeeze \textsc{aux}.\textsc{f} manioc(\textsc{m}) squeeze \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `When the manioc was finished being grated, she put the manioc through the sieve.'
\z

\ea\label{ex:10-a30}
Fowa nano ni abematamonaka.\\
    \gll fowa nano ni ahaba-himata-mona-ka\\
    manioc(\textsc{m}) be\_squeezed \textsc{aux}.\textsc{nfin} end-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `The manioc was finished being put through the sieve.'
\z

\ea\label{ex:10-a31}
Toto Abono aba, aba me tani tihara tiwa awine tini?\\
    \gll Toto\_Abono aba aba me tani ti-ha-ra tiwa awine ti-ni\\
    Toto\_Abono(\textsc{m}) fish(\textsc{m}) fish(\textsc{m}) 3\textsc{pl}.\textsc{poss} killer 2\textsc{sg}.\textsc{s}-be-\textsc{neg}.\textsc{f} 2\textsc{sg}.\textsc{s} seem.\textsc{f} 2\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Toto Abono, aren't you a fisherman?'''
\z

\ea\label{ex:10-a32}
Aba me sasa kana wati okia okere.\\
    \gll aba me sasaa\_kana wati o-kiha o-ke-re\\
    fish(\textsc{m}) 3\textsc{pl}.\textsc{poss} thing\_for\_spearing arrow(\textsc{m}) 1\textsc{sg}.\textsc{s}-have 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}-\textsc{neg}.\textsc{f}\\
    \glt `{``}I don't have any arrows to spear fish with.'''
\z

\ea\label{ex:10-a33}
Aba me kaya ke.\\
    \gll aba me kaya ke\\
    fish(\textsc{m}) 3\textsc{pl}.\textsc{s} stand \textsc{decl}.\textsc{f}\\
    \glt `{``}The fish are there in the water.'''
\z

\ea\label{ex:10-a34}
Aba me kaya na me wati okiarihi.\\
    \gll aba me kaya na me wati o-kiha-rihi\\
    fish(\textsc{m}) 3\textsc{pl}.\textsc{s} stand \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} arrow(\textsc{m}) 1\textsc{sg}.\textsc{s}-have-\textsc{neg}.\textsc{list}\\
    \glt `{``}The fish are there in the water, but I don't have any arrows.'''
\z

\ea\label{ex:10-a35}
Oye honarimakoni ofimi onofa ama oni.\\
    \gll o-yehe hona-ra-makoni o-fimi o-nofa ama o-ni\\
    1\textsc{sg}.\textsc{poss}-hand have\_something-\textsc{neg}-for\_this\_reason.\textsc{f} 1\textsc{sg}.\textsc{s}-be\_hungry 1\textsc{sg}.\textsc{s}-recently.\textsc{f} \textsc{sec} 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Because I don't have any equipment, I am always hungry.'''
\z

\ea\label{ex:10-a36}
Faya toto abono tokare, kasiro hikane, iyawa hinita kaba, iyawa hinita kabemetemoneni.\\
    \gll faya toto abono to-ka-re kasiro hi-ka-ne iyawa hinita kaba iyawa hinita kaba-hemete-mone-ni\\
    so bamboo.sp(\textsc{f}) spirit away-go-\textsc{neg}.\textsc{m} complain\_about \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} grated\_mash(\textsc{f}) alone eat.\textsc{f} grated\_mash(\textsc{f}) alone eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `So the \textit{toto} spirit didn't go out. She pestered him. She ate manioc meal by itself.'
\z

\ea\label{ex:10-a37}
Iyawa hinita kabi onofarini.\\
    \gll iyawa hinita kabi o-nofa-ra-ni\\
    grated\_mash(\textsc{f}) alone eat.\textsc{nfin} 1\textsc{sg}.\textsc{s}-want-\textsc{neg}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}I don't want to eat manioc meal by itself.'''
\z

\ea\label{ex:10-a38}
Iyawa hinita kabemetemoneke, me amo kanabone me fama mati.\\
    \gll iyawa hinita kaba-hemete-mone-ke me amo ka-na-habone me fama mati\\
    grated\_mash(\textsc{f}) alone eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} 3\textsc{pl}.\textsc{s} sleep \textsc{comit}-\textsc{aux}-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{s} be\_two.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `She ate manioc meal by itself, before the two of them slept.'
\z

\ea\label{ex:10-a39}
Faya amo na, amo nemetemone fahi.\\
    \gll faya amo na amo na-hemete-mone fahi\\
    so sleep \textsc{aux}.\textsc{f} sleep \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `She slept.'
\z

\ea\label{ex:10-a40}
Hato abono ai bai ite ya yifo tafa nematamonaka.\\
    \gll hato abono ahi bahi ite ya yifo tafa na-himata-mona-ka\\
    bamboo.sp(\textsc{f}) spirit there sun(\textsc{m}) sit.\textsc{m} \textsc{adju} firewood(\textsc{f}) split \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `When the sun was this high the \textit{hato} spirit cut firewood.'\footnote{Here and in a couple other places in the story, the spirit is called \textit{hato abono}, the same name used in the Jamamadi version of the story. \textit{Hato} is another species of bamboo.}
\z

\ea\label{ex:10-a41}
Faya yifo tafa ne, yifo tafa nematamonane.\\
    \gll faya yifo tafa ne yifo tafa na-himata-mona-ne\\
    so firewood(\textsc{f}) split \textsc{aux}.\textsc{m} firewood(\textsc{f}) split \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `He cut firewood.'
\z

\ea\label{ex:10-a42}
Yifo ta toniharematamonaka, fati, fatitebone, hiwakama wati kanehinoho.\\
    \gll yifo taa to-niha-re-himata-mona-ka fati fati-tee-bone hi-ka-ka-ma wati ka-na-hino-ho\\
    firewood(\textsc{f}) be\_piled\_up \textsc{ch}-\textsc{caus}-raised\_surface-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) 3\textsc{sg}.\textsc{poss}.wife(\textsc{f})-\textsc{hab}-\textsc{int}.\textsc{f} \textsc{oc}-\textsc{comit}-come-back.\textsc{f} plan\_against \textsc{comit}-\textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `He put the firewood in a pile.  He was planning against his bride, who he had brought.'
\z

\newpage
\ea\label{ex:10-a43}
Wati hikanani, amo na, amo ni tati ba hiremetemoneke, amo naro ahi, yama we kamaki ya.\\
    \gll wati hi-ka-na-hani amo na amo ni tati baa hi-re-hemete-mone-ke amo na-haaro ahi yama wehe ka-maki ya\\
    plan\_against \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{f} sleep \textsc{aux}.\textsc{m} sleep \textsc{aux}.\textsc{nfin} head hit \textsc{oc}-raised\_surface-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} sleep \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f} then thing(\textsc{f}) light.\textsc{f} come-following.\textsc{nom} \textsc{adju}\\
    \glt `He was planning against her. She was sleeping. While she was sleeping, he hit her on the head, at dawn.'
\z

\ea\label{ex:10-a44}
Faya tati ba hiremetemone ahi.\\
    \gll faya tati baa hi-re-hemete-mone ahi\\
    so head hit \textsc{oc}-raised\_surface-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `He hit her on the head.'
\z


\ea\label{ex:10-a45}
Faya ahaba, bosa hikana bobi hinamina hikabaminemetemoneke.\\
    \gll faya ahaba bosa hi-ka-na bobi hi-na-mina hi-kaba-mina-hemete-mone-ke\\
    so die.\textsc{f} get\_up\_early \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{f} slit \textsc{oc}-\textsc{aux}-morning.\textsc{f} \textsc{oc}-eat-morning-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `She died. He got up early with respect to her. He cut her up in the morning. He ate her in the morning.'
\z

\ea\label{ex:10-a46}
Tabasi hina, tabasi hini ya hikaba, fara kakawiabone hiti hinaha ya kakawi hinemetemoneke.\\
    \gll tabasi hi-na tabasi hi-ni ya hi-kaba fara kakawia-bone hiti hi-naha ya ka-kawi hi-na-hemete-mone-ke\\
    roast\_on\_rack \textsc{oc}-\textsc{aux}.\textsc{f} roast\_on\_rack \textsc{oc}-\textsc{aux}.\textsc{nom}.\textsc{f} \textsc{adju} \textsc{oc}-eat.\textsc{f} same.\textsc{f} accompanying\_food-\textsc{int}.\textsc{f} grate \textsc{oc}-\textsc{aux}.\textsc{f} \textsc{adju} \textsc{comit}-eat\_with\_something \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He roasted her. After he roasted her, he ate her. He ate her with the starch that she herself had grated.'
\z

\ea\label{ex:10-a47}
Faya kakawi hina, faya abi ya, faya fati kabi abi ya, yana tone fimihi yana tone, tokomematamonaka.\\
    \gll faya ka-kawi hi-na faya ahabi ya faya fati kabi ahabi ya yana to-ne fimihi yana to-ne to-ka-ma-himata-mona-ka\\
    so \textsc{dup}-eat\_with\_something \textsc{oc}-\textsc{aux}.\textsc{f} so die.\textsc{nom}.\textsc{f} \textsc{adju} so 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) be\_eaten.\textsc{nfin} end.\textsc{nom}.\textsc{f} \textsc{adju} start \textsc{ch}-\textsc{aux}.\textsc{m} be\_hungry.\textsc{m} start \textsc{ch}-\textsc{aux}.\textsc{m} away-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He ate her with starch. When the eating of his wife was finished, he got ready to leave. He was hungry, and he got ready to leave. He went back.'
\z

\ea\label{ex:10-a48}
Tokome, hinaka koma ni ya, hinaka koma ni ya tokome kobo toname, kobo tonamematamonaka.\\
    \gll to-ka-me hinaka koma ni ya hinaka koma ni ya to-ka-me kobo to-na-me kobo to-na-ma-himata-mona-ka\\
    away-go-back.\textsc{m} 3\textsc{sg}.\textsc{poss} son\_in\_law(\textsc{m}) to \textsc{adju} 3\textsc{sg}.\textsc{poss} son\_in\_law(\textsc{m}) to \textsc{adju} away-go-back.\textsc{m} arrive away-\textsc{aux}-back.\textsc{m} arrive away-\textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went back. He went back to his mother-in-law. He arrived.'\footnote{Instead of \textit{hinaka koma}, Siko should have said \textit{hinakaso} `his mother-in-law'.}
\z

\ea\label{ex:10-a49}
Aa, ati maki kame.\\
    \gll aa ati maki ka-me\\
    aha 2\textsc{sg}.\textsc{poss}.older\_sister(\textsc{f}) 3\textsc{sg}.\textsc{poss}.husband(\textsc{m}) come-back.\textsc{m}\\
    \glt `{``}Aha, there comes your older sister's husband.'''\footnote{The feminine equivalent of \textit{kame} is \textit{kamini}, showing the pairing of feminine \textit{-ni} with ``nothing'' rather than the masculine \textit{-ne}. This phenomenon is discussed by \citet[411]{Dixon:2004xr}, who claims, however, that it only applies to polar questions. The masculine \textit{-ne} could be used to form the word \textit{kamane}, but then the word would contain continuous aspect (cf. e.g. \textit{abawahane} in sentence (\ref{ex:10-a231}), and the feminine equivalent would be \textit{kamineni}, which contains the continuous suffix \textit{-ne}, and not \textit{kamini}. See also \citet[36f.]{Vogel:2022du}.
}
\z

\ea\label{ex:10-a50}
Oof, okomine oke.\\
    \gll oof o-ka-ma-ne o-ke\\
    hoo 1\textsc{sg}.\textsc{s}-come-back-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Whew, I have come back.'''
\z

\ea\label{ex:10-a51}
Hoo, oko yibote tera ati na ati ihi okomine oni, fowa te hiti nabone te.\\
    \gll hoo o-ka yibote tera ati na ati ihi o-ka-ma-ne o-ni fowa tee hiti na-habone tee\\
    whew 1\textsc{sg}.\textsc{poss}-\textsc{poss} wife(\textsc{f}) 2\textsc{pl}.\textsc{o} call\_for \textsc{aux}.\textsc{f} voice because\_of.\textsc{f} 1\textsc{sg}.\textsc{s}-come-back-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f} manioc(\textsc{m}) 2\textsc{pl}.\textsc{s} grate \textsc{aux}-\textsc{int}.\textsc{f} 2\textsc{pl}.\textsc{s}\\
    \glt `{``}Whew, my wife called for you. Because of what she said I have come, so you can grate manioc together.'''
\z

\ea\label{ex:10-a52}
Okoto ma toa awineni, fowa hiti naro.\\
    \gll okoto maa to-ha awine-ni fowa hiti na-haaro\\
    1\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) be\_tired \textsc{ch}-\textsc{aux}.\textsc{f} seem.\textsc{f}-\textsc{bkg}.\textsc{f} manioc(\textsc{m}) grate \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f}\\
    \glt `{``}I guess my daughter is tired, because she is grating manioc.'''
\z


\ea\label{ex:10-a53}
Fowa te hiti nabone ati na ati ihi okomine o.\\
    \gll fowa tee hiti na-habone ati na ati ihi o-ka-ma-ne o\\
    manioc(\textsc{m}) 2\textsc{pl}.\textsc{s} grate \textsc{aux}-\textsc{int}.\textsc{f} say \textsc{aux}.\textsc{f} voice because\_of.\textsc{f} 1\textsc{sg}.\textsc{s}-come-back-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}\\
    \glt `{``}She wants you to grate manioc together. Because of what she said I have come.'''
\z

\ea\label{ex:10-a54}
Me tekamiya.\\
    \gll me ti-ka-ka-ma-yahi\\
    3\textsc{pl}.\textsc{o} 2\textsc{sg}.\textsc{s}-\textsc{comit}-go-back-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `{``}You take them.'''
\z

\ea\label{ex:10-a55}
Me famamiyahi.\\
    \gll me fama-ma-yahi\\
    3\textsc{pl}.\textsc{s} be\_two-back-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `{``}Let the two of them be together again.'''
\z

\ea\label{ex:10-a56}
Yana tonematamona fahi.\\
    \gll yana to-na-himata-mona fahi\\
    start \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `He got ready to leave.'
\z

\ea\label{ex:10-a57}
Yana toneibona Aso, aso okoma oke.\\
    \gll yana to-na-hibona aso aso o-ka-ma o-ke\\
    start \textsc{ch}-\textsc{aux}-\textsc{int}.\textsc{m} mother\_in\_law(\textsc{f}) mother\_in\_law(\textsc{f}) 1\textsc{sg}.\textsc{s}-go-back 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `He got ready to leave. ``Mother-in-law, I'm going.'''
\z

\ea\label{ex:10-a58}
Tikoto one ota tofamamibana, hinakati asota nabonehe, fowa me hiti nabone mati.\\
    \gll tikoto one otaa to-fama-mi-bana hinakati asota na-habone-he fowa me hiti na-habone mati\\
    2\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) other.\textsc{f} 1\textsc{ex}.\textsc{s} away-be\_two-back.\textsc{f}-\textsc{fut} 3\textsc{poss}.\textsc{sg}.older\_sister(\textsc{f}) help \textsc{aux}-\textsc{int}.\textsc{f}-\textsc{dup} manioc(\textsc{m}) 3\textsc{pl}.\textsc{s} grate \textsc{aux}-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `{``}Let your other daughter go back with me, to help her older sister, so they will grate manioc together.'''
\z

\ea\label{ex:10-a59}
Ati amosematamonaka.\\
    \gll ati amosa-himata-mona-ka\\
    voice be\_good-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `What he said was acceptable to her.'
\z

\ea\label{ex:10-a60}
Ati maki te tofamamahi.\\
    \gll ati maki tee to-fama-ma-hi\\
    2\textsc{sg}.\textsc{poss}.older\_sister(\textsc{f}) 3\textsc{sg}.\textsc{poss}.husband(\textsc{m}) 2\textsc{pl}.\textsc{s} away-be\_two-back-\textsc{imp}.\textsc{f}\\
    \glt `{``}You go back with your older sister's husband.'''
\z

\ea\label{ex:10-a61}
Ati maki te tofamamahi.\\
    \gll ati maki tee to-fama-ma-hi\\
    2\textsc{sg}.\textsc{poss}.older\_sister(\textsc{f}) 3\textsc{sg}.\textsc{poss}.husband(\textsc{m}) 2\textsc{pl}.\textsc{s} away-be\_two-back-\textsc{imp}.\textsc{f}\\
    \glt `{``}You go back with your older sister's husband.'''
\z

\ea\label{ex:10-a62}
Te towakamahi.\\
    \gll tee to-ka-ka-ma-hi\\
    2\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back-\textsc{imp}.\textsc{f}\\
    \glt `{``}The two of you go.'''
\z

\ea\label{ex:10-a63}
Ati te fafamawa na toharima nahi.\\
    \gll ati tee fa-fama-waha na to-ha-rima na-hi\\
    2\textsc{sg}.\textsc{poss}.older\_sister(\textsc{f}) 2\textsc{pl}.\textsc{s} \textsc{dup}-be\_two-change.\textsc{list} \textsc{aux} \textsc{ch}-\textsc{aux}-intermittent \textsc{aux}-\textsc{imp}.\textsc{f}\\
    \glt `{``}You stay with your older sister.'''
\z

\ea\label{ex:10-a64}
Faya hiwakamemetemone amake.\\
    \gll faya hi-ka-ka-ma-hemete-mone ama-ke\\
    so \textsc{oc}-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f}\\
    \glt `So he took her.'
\z

\ea\label{ex:10-a65}
Hiwakama, hiwakamemetemoneke.\\
    \gll hi-ka-ka-ma hi-ka-ka-ma-hemete-mone-ke\\
    \textsc{oc}-\textsc{comit}-go-back.\textsc{f} \textsc{oc}-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He took her.'
\z

\ea\label{ex:10-a66}
Tokome fawa tosame, fawa tonakosamematamonaka.\\
    \gll to-ka-me fawa to-kosa-me fawa to-na-kosa-ma-himata-mona-ka\\
    away-go-back.\textsc{m} disappear \textsc{ch}-middle-back.\textsc{m} disappear \textsc{ch}-\textsc{aux}-middle-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went back. He disappeared into the forest.'
\z

\ea\label{ex:10-a67}
Okoto fowa hiti namone hinakasima, Toto Abono towakama ne, ma toamone ati ihimonehe.\\
    \gll okoto fowa hiti na-hamone hinakasima Toto\_Abono to-ka-ka-ma ne maa to-ha-hamone ati ihi-mone-he\\
    1\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) manioc(\textsc{m}) grate \textsc{aux}-\textsc{rep}.\textsc{f} 3\textsc{sg}.\textsc{poss}.younger\_sister(\textsc{f}) Toto\_Abono(\textsc{m}) away-\textsc{comit}-go-back.\textsc{list} \textsc{aux}.\textsc{m} be\_tired \textsc{ch}-\textsc{aux}-\textsc{rep}.\textsc{f} voice because\_of.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `{``}My daughter is said to be grating manioc, and now Toto Abono has taken her younger sister, too, because she is said to be tired.'''
\z

\ea\label{ex:10-a68}
Faya me kobo tokanama, fanahi mera yori rari.\\
    \gll faya me kobo to-ka-na-ma fanahi mera yori ra-haari\\
    so 3\textsc{pl}.\textsc{s} arrive away-\textsc{comit}-\textsc{aux}-back.\textsc{f} be\_married.\textsc{m} 3\textsc{pl}.\textsc{o} have\_sex\_with \textsc{neg}-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `The two of them arrived. He was married, but he didn't have sex with them.'
\z

\newpage
\ea\label{ex:10-a69}
Mera yori re, mera iiso tohe mera kakaba toemata.\\
    \gll mera yori re mera i-iso to-he mera ka-kaba to-ha-himata\\
    3\textsc{pl}.\textsc{o} have\_sex\_with \textsc{neg}.\textsc{m} 3\textsc{pl}.\textsc{o} \textsc{dup}-carry\_off \textsc{ch}-\textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{o} \textsc{dup}-eat \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}\\
    \glt `He didn't have sex with them. He was carrying them off. He was eating them.'
\z

\ea\label{ex:10-a70}
Mera kabe, faya hinabowamake, hikabemetemoneke fahi.\\
    \gll mera kabe faya hi-naabowa-make hi-kaba-hemete-mone-ke fahi\\
    3\textsc{pl}.\textsc{o} eat.\textsc{m} so \textsc{oc}-kill-following \textsc{oc}-eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `He ate them. He killed her, also. He ate her.'
\z

\ea\label{ex:10-a71}
Hikaba, kabi abi ya hinaka iyawa ahaba, faya hinaka iyawa, hinaka iyawa aba, Oof, iyawa okiamari.\\
    \gll hi-kaba kabi ahabi ya hinaka iyawa ahaba faya hinaka iyawa hinaka iyawa ahaba oof iyawa o-kiha-ma-ri\\
    \textsc{oc}-eat.\textsc{f} be\_eaten.\textsc{nfin} end.\textsc{nom}.\textsc{f} \textsc{adju} 3\textsc{sg}.\textsc{poss} grated\_mash(\textsc{f}) end.\textsc{f} so 3\textsc{sg}.\textsc{poss} grated\_mash(\textsc{f}) 3\textsc{sg}.\textsc{poss} grated\_mash(\textsc{f}) end.\textsc{f} hoo grated\_mash(\textsc{f}) 1\textsc{sg}.\textsc{s}-have-back-\textsc{neg}.\textsc{nom}.\textsc{f}\\
    \glt `He ate her. After he finished eating her, his manioc mash ran out. ``Whew, I don't have any more manioc mash.'''
\z

\ea\label{ex:10-a72}
Iyawa abake.\\
    \gll iyawa ahaba-ke\\
    grated\_mash(\textsc{f}) end-\textsc{decl}.\textsc{f}\\
    \glt `{``}The manioc mash is all gone.'''
\z

\ea\label{ex:10-a73}
Iyawa hihiti nisaba onakamamate.\\
    \gll iyawa hihiti\_nisa-ba o-na-ka-ma-mate\\
    grated\_mash(\textsc{f}) person\_who\_grates-\textsc{fut} 1\textsc{sg}.\textsc{s}-\textsc{caus}-go-back-short\_time.\textsc{imp}.\textsc{f}\\
    \glt `{``}I'll go get someone to grate manioc mash.'''
\z

\newpage
\ea\label{ex:10-a74}
Faya abemetemone amani.\\
    \gll faya ahaba-hemete-mone ama-ni\\
    so die-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec}-\textsc{bkg}.\textsc{f}\\
    \glt `She died.'
\z

\ea\label{ex:10-a75}
Hikabemetemone amake.\\
    \gll hi-kaba-hemete-mone ama-ke\\
    \textsc{oc}-eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f}\\
    \glt `He ate her.'
\z

\ea\label{ex:10-a76}
Tati ba hiremetemone.\\
    \gll tati baa hi-re-hemete-mone\\
    head hit \textsc{oc}-raised\_surface-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `He hit her on the head.'
\z


\ea\label{ex:10-a77}
Tati ba hire, namiti, namiti niki hikanemetemoneke, amo naro.\\
    \gll tati baa hi-re namiti namiti niki hi-ka-na-hemete-mone-ke amo na-haaro\\
    head hit \textsc{oc}-raised\_surface neck neck press\_on \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} sleep \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f}\\
    \glt `He hit her on the head. He strangled her as she slept.'
\z

\ea\label{ex:10-a78}
Amo ni winaha ra namiti niki kanematamonaka.\\
    \gll amo ni winaha ra namiti niki ka-na-himata-mona-ka\\
    sleep \textsc{aux}.\textsc{nfin} lie\_in\_hammock.\textsc{f} \textsc{o} neck press\_on \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He strangled her as she slept in her hammock.'
\z

\ea\label{ex:10-a79}
Faya namiti niki hikana, hinahabiha, bobi hinamina, bobi hinamina faya tabasi hina sobori hikabemetemoneke ahi.\\
    \gll faya namiti niki hi-ka-na hi-naahabiha bobi hi-na-mina bobi hi-na-mina faya tabasi hi-na sobori hi-kaba-hemete-mone-ke ahi\\
    so neck press\_on \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{f} \textsc{oc}-kill.\textsc{f} slit \textsc{oc}-\textsc{aux}-morning.\textsc{f} slit \textsc{oc}-\textsc{aux}-morning.\textsc{f} so roast\_on\_rack \textsc{oc}-\textsc{aux}.\textsc{f} belly \textsc{oc}-eat-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `He strangled her. He killed her. He cut her up in the morning. Then he roasted her. He ate her belly.'
\z

\ea\label{ex:10-a80}
Sobori hikaba, tafe ya wina nati, Aa, oko bani.\\
    \gll sobori hi-kaba tafe ya wina nahati aa o-ka bani\\
    belly \textsc{oc}-eat.\textsc{f} eat.\textsc{m} \textsc{adju} lie\_in\_hammock daytime.\textsc{m} oh\_yes 1\textsc{sg}.\textsc{poss}-\textsc{poss} animal(\textsc{m})\\
    \glt `He ate her belly. After he ate, he lay in his hammock during the day. ``Oh, my meat.'''
\z

\ea\label{ex:10-a81}
Aa, oko bani tabatabasi okowamate.\\
    \gll aa o-ka bani taba-tabasi o-ka-waha-mate\\
    oh\_yes 1\textsc{sg}.\textsc{poss}-\textsc{poss} animal(\textsc{m}) \textsc{dup}-roast\_on\_rack 1\textsc{sg}.\textsc{s}-\textsc{comit}-change-short\_time.\textsc{imp}.\textsc{f}\\
    \glt `{``}Oh, I need to roast my meat some more.'''
\z

\ea\label{ex:10-a82}
Faya tabasi hina, hikaba hinate, hikabaminatasa, one hikabaminatasi ya aba, faya abemetemoneke.\\
    \gll faya tabasi hi-na hi-kaba hi-nahate hi-kaba-mina-tasa one hi-kaba-mina-tasi ya ahaba faya ahaba-hemete-mone-ke\\
    so roast\_on\_rack \textsc{oc}-\textsc{aux}.\textsc{f} \textsc{oc}-eat \textsc{oc}-daytime \textsc{oc}-eat-morning-again.\textsc{f} other.\textsc{f} \textsc{oc}-eat-morning-again.\textsc{nom}.\textsc{f} \textsc{adju} end.\textsc{f} so end-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `So he roasted her. He ate her all day. He ate her the next day again. He ate another piece the next day again, and then she was all gone.'
\z

\ea\label{ex:10-a83}
Kabi aba, kabi abemetemone fahi, tokomatasebanoho.\\
    \gll kabi ahaba kabi ahaba-hemete-mone fahi to-ka-ma-tasa-hiba-no-ho\\
    be\_eaten.\textsc{nfin} end.\textsc{f} be\_eaten.\textsc{nfin} end-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then away-go-back-again-\textsc{fut}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `The eating of her was done. Then he went back again.'
\z

\ea\label{ex:10-a84}
Me onakamamatibe, me one, oko fowa me hiti naba mati.\\
    \gll me o-na-ka-ma-mata-be me one o-ka fowa me hiti na-haba mati\\
    3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-\textsc{caus}-go-back-short\_time-\textsc{immed}.\textsc{f} 3\textsc{pl} other.\textsc{f} 1\textsc{sg}.\textsc{poss}-\textsc{poss} manioc(\textsc{m}) 3\textsc{pl}.\textsc{s} grate \textsc{aux}-\textsc{fut}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `{``}I'm going to go after them, others of them, for them to grate my manioc.'''
\z

\ea\label{ex:10-a85}
Oof, fowa me hiti nabone mera, kokori biterawa tonakamatasarawe, mera tonakame mera tokanikamatase, Hiya, aa, Toto Abono kamatasari hari.\\
    \gll oof fowa me hiti na-habone mera kokori bite-rawa to-na-ka-ma-tasa-rawe mera to-na-ka-me mera to-ka-nika-ma-tase hiya aa Toto\_Abono ka-ma-tasa-haari haari\\
    hoo manioc(\textsc{m}) 3\textsc{pl}.\textsc{s} grate \textsc{aux}-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{o} 3\textsc{sg}.\textsc{poss}.father\_in\_law(\textsc{m}) 3\textsc{sg}.\textsc{poss}.daughter(\textsc{f})-\textsc{fpl} away-\textsc{caus}-go-back-again-\textsc{fpl}.\textsc{m} 3\textsc{pl}.\textsc{o} away-\textsc{caus}-go-back.\textsc{m} 3\textsc{pl}.\textsc{o} away-\textsc{comit}-ask\_for-back-again.\textsc{m} come\_on aha Toto\_Abono(\textsc{m}) come-back-again-look.\textsc{m} that\_one.\textsc{m}\\
    \glt `He went after his father-in-law's daughters, for them to grate manioc. He went after them. He went and asked for others. ``Look, Toto Abono has come again.'''
\z

\ea\label{ex:10-a86}
Oof, okomine oke.\\
    \gll oof o-ka-ma-ne o-ke\\
    hoo 1\textsc{sg}.\textsc{s}-come-back-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Whew!  I have come back.'''
\z

\ea\label{ex:10-a87}
Me ati kakasiro ra me ati ihi okomatasine oni.\\
    \gll me ati ka-kasiro ra me ati ihi o-ka-ma-tasa-ne o-ni\\
    3\textsc{pl}.\textsc{poss} voice \textsc{dup}-persist \textsc{neg}.\textsc{f} 3\textsc{pl}.\textsc{poss} voice because\_of.\textsc{f} 1\textsc{sg}.\textsc{s}-come-back-again-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}They keep pestering me. Because of what they said I have come again.'''
\z

\ea\label{ex:10-a88}
Fowa e nafi ahi nibeya.\\
    \gll fowa ee nafi ahi na-be-ya\\
    manioc(\textsc{m}) 1\textsc{in}.\textsc{poss} all work\_on \textsc{aux}-\textsc{immed}.\textsc{f}-now\\
    \glt `{``}`Let's all process manioc together.'{''}'
\z

\ea\label{ex:10-a89}
Ota nafi fowa ahi nabanani.\\
    \gll otaa nafi fowa ahi na-habana-ni\\
    1\textsc{ex}.\textsc{poss} all manioc(\textsc{m}) work\_on \textsc{aux}-\textsc{fut}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}`We will all process manioc.'{''}'
\z

\newpage
\ea\label{ex:10-a90}
Ota ka asima tinakamarawahi me ati narake.\\
    \gll otaa ka asima ti-na-ka-ma-rawa-hi me ati na-hara-ke\\
    1\textsc{ex} \textsc{poss} younger\_sister(\textsc{f}) 2\textsc{sg}.\textsc{s}-\textsc{caus}-go-back-\textsc{fpl}-\textsc{imp}.\textsc{f} 3\textsc{pl}.\textsc{s} say \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}`Go get our younger sisters,' they said.'''
\z

\ea\label{ex:10-a91}
Faya me ati na me ati ihi okomine oke.\\
    \gll faya me ati na me ati ihi o-ka-ma-ne o-ke\\
    so 3\textsc{pl}.\textsc{s} say say.\textsc{f} 3\textsc{pl}.\textsc{poss} voice because\_of.\textsc{f} 1\textsc{sg}.\textsc{s}-come-back-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I have come because of what they said.'''
\z

\ea\label{ex:10-a92}
Faya me, me famarawemetemoneke.\\
    \gll faya me me fama-rawa-hemete-mone-ke\\
    so 3\textsc{pl} 3\textsc{pl}.\textsc{s} be\_two-\textsc{fpl}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `There were two of them.'
\z

\ea\label{ex:10-a93}
Famarawa, towakame, me hekamemetemoneke.\\
    \gll fama-rawa to-ka-ka-me me hi-ka-ka-ma-hemete-mone-ke\\
    be\_two-\textsc{fpl}.\textsc{f} away-\textsc{comit}-go-back.\textsc{m} 3\textsc{pl}.\textsc{o} \textsc{oc}-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He took the two of them. He took them.'
\z

\ea\label{ex:10-a94}
Me hekama, Ati ka fowa te sowe na fowa owa te nano niyahi.\\
    \gll me hi-ka-ka-ma ati ka fowa tee sowe na fowa owa tee nano na-yahi\\
    3\textsc{pl}.\textsc{o} \textsc{oc}-\textsc{comit}-go-back.\textsc{f} 2\textsc{sg}.\textsc{poss}.older\_sister(\textsc{f}) \textsc{poss} manioc(\textsc{m}) 2\textsc{pl}.\textsc{s} scrape \textsc{aux}.\textsc{f} manioc(\textsc{m}) other.\textsc{m} 2\textsc{pl}.\textsc{s} squeeze \textsc{aux}-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `He took them. ``You will peel your older sister's manioc. You squeeze the poison juice out of more manioc.'''
\z

\ea\label{ex:10-a95}
Fowa te nano niyahi.\\
    \gll fowa tee nano na-yahi\\
    manioc(\textsc{m}) 2\textsc{pl}.\textsc{s} squeeze \textsc{aux}-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `{``}You squeeze the poison juice out of the manioc.'''
\z

\ea\label{ex:10-a96}
Te ebe kakatomariyahi.\\
    \gll tee ebe ka-katoma-ra-yahi\\
    2\textsc{pl}.\textsc{poss} \textsc{recip} \textsc{comit}-look\_at-\textsc{neg}-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `{``}Don't just look at each other.'''
\z

\ea\label{ex:10-a97}
Okoto ma toarawa awineni.\\
    \gll okoto maa to-ha-rawa awine-ni\\
    1\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) be\_tired \textsc{ch}-\textsc{aux}-\textsc{fpl}.\textsc{f} seem.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}My daughters must be tired.'''
\z

\ea\label{ex:10-a98}
Me ma toarara.\\
    \gll me maa to-ha-ra-hara\\
    3\textsc{pl}.\textsc{s} be\_tired \textsc{ch}-\textsc{aux}-\textsc{neg}-\textsc{ip}.\textsc{e}.\textsc{f}\\
    \glt `{``}They are tired.'''
\z

\ea\label{ex:10-a99}
Me ma toa me ati kasiroarake.\\
    \gll me maa to-ha me ati kasiro-hara-ke\\
    3\textsc{pl}.\textsc{s} be\_tired \textsc{ch}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{poss} voice persist-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}They are tired and they don't stop pestering.'''
\z

\ea\label{ex:10-a100}
Me ma toa me ati kasiroarake.\\
    \gll me maa to-ha me ati kasiro-hara-ke\\
    3\textsc{pl}.\textsc{s} be\_tired \textsc{ch}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{poss} voice persist-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}They are tired and they don't stop pestering.'''
\z

\ea\label{ex:10-a101}
Me ati ihi okomine oke.\\
    \gll me ati ihi o-ka-ma-ne o-ke\\
    3\textsc{pl}.\textsc{poss} voice because\_of.\textsc{f} 1\textsc{sg}.\textsc{s}-come-back-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Because of what they said I have come.'''
\z

\ea\label{ex:10-a102}
Owati amakere.\\
    \gll o-ati ama-ke-re\\
    1\textsc{sg}.\textsc{poss}-voice be-\textsc{decl}.\textsc{f}-\textsc{neg}.\textsc{f}\\
    \glt `{``}It wasn't my idea.'''
\z

\ea\label{ex:10-a103}
Owati amara ya okomine oni.\\
    \gll o-ati ama-ra ya o-ka-ma-ne o-ni\\
    1\textsc{sg}.\textsc{poss}-voice be-\textsc{neg}.\textsc{f} \textsc{adju} 1\textsc{sg}.\textsc{s}-come-back-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}I have come, not because of my idea.'''
\z

\ea\label{ex:10-a104}
Faya kame me kobo namemetemoneke.\\
    \gll faya ka-me me kobo na-ma-hemete-mone-ke\\
    so come-back.\textsc{m} 3\textsc{pl}.\textsc{s} arrive \textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `So he went home. They arrived back.'
\z

\ea\label{ex:10-a105}
Me kobo nama me, bai ite me kobo namemetemoneke ahi.\\
    \gll me kobo na-ma me bahi ite me kobo na-ma-hemete-mone-ke ahi\\
    3\textsc{pl}.\textsc{s} arrive \textsc{aux}-back.\textsc{f} 3\textsc{pl}.\textsc{s} sun(\textsc{m}) sit.\textsc{m} 3\textsc{pl}.\textsc{s} arrive \textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `They arrived back. They arrived back when the sun was this high.'
\z

\ea\label{ex:10-a106}
Oof oko yifoba owamatibe.\\
    \gll oof o-ka yifo-ba o-awa-mata-be\\
    hoo 1\textsc{sg}.\textsc{poss}-\textsc{poss} firewood(\textsc{f})-\textsc{fut} 1\textsc{sg}.\textsc{s}-see-short\_time-\textsc{immed}.\textsc{f}\\
    \glt `{``}Whew! I'm going to get myself some firewood.'''
\z


\ea\label{ex:10-a107}
Yifo okia okereni, mera wati kanenoho, tasa.\\
    \gll yifo o-kiha o-ke-re-ni mera wati ka-na-hino-ho tasa\\
    firewood(\textsc{f}) 1\textsc{sg}.\textsc{s}-have 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}-\textsc{neg}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{o} plan\_against \textsc{comit}-\textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup} again\\
    \glt `{``}I didn't realize I had no firewood.'' He was planning against them, too.'
\z

\ea\label{ex:10-a108}
Hibaka okati toaha?\\
    \gll hibaka okati to-haha\\
    where 1\textsc{sg}.\textsc{poss}.older\_sister(\textsc{f}) \textsc{ch}-be.\textsc{f}\\
    \glt `{``}Where are my older sisters?'''
\z

\ea\label{ex:10-a109}
Me towaka me awineke, fowa me bore naba mati.\\
    \gll me to-ka-ka me awine-ke fowa me bore na-haba mati\\
    3\textsc{pl}.\textsc{s} away-\textsc{comit}-go.\textsc{f} 3\textsc{pl}.\textsc{s} seem.\textsc{f}-\textsc{decl}.\textsc{f} manioc(\textsc{m}) 3\textsc{pl}.\textsc{s} pull\_out \textsc{aux}-\textsc{fut}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `{``}They must have gone out to pull up manioc.'''
\z

\ea\label{ex:10-a110}
Ati, ati, ati ya me ka ati me ha na me, me ka ati haha remetemoneke.\\
    \gll ati ati ati ya me ka ati me haa na me me ka ati ha-haa ra-hemete-mone-ke\\
    older\_sister(\textsc{f}) older\_sister(\textsc{f}) older\_sister(\textsc{f}) \textsc{adju} 3\textsc{pl} \textsc{poss} older\_sister(\textsc{f}) 3\textsc{pl}.\textsc{s} call \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl} \textsc{poss} older\_sister(\textsc{f}) \textsc{dup}-call \textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Sister, sister.'' They called their older sisters, calling, ``Sister.'' Their older sisters didn't call out.'
\z

\ea\label{ex:10-a111}
Yama me siba na me, me moto tokanama, me, hinakati tati tone ra, hinakati tati tone tasi tokasani, mato ya tasi tokasa watehemetemoneni, basioba kawari ya.\\
    \gll yama me siba na me me moto to-ka-na-ma me hinakati tati tone ra hinakati tati tone tasi to-ka-kosa-hani mato ya tasi to-ka-kosa wata-hemete-mone-ni basioba ka-waa-ri ya\\
    thing(\textsc{f}) 3\textsc{pl}.\textsc{s} look\_for \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} loop away-\textsc{comit}-\textsc{aux}-back.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{poss}.\textsc{sg}.older\_sister(\textsc{f}) head bone \textsc{o} 3\textsc{poss}.\textsc{sg}.older\_sister(\textsc{f}) head bone emerge \textsc{ch}-\textsc{comit}-middle-\textsc{ip}.\textsc{n}.\textsc{f} inner\_bark(\textsc{f}) \textsc{adju} emerge \textsc{ch}-\textsc{comit}-middle.\textsc{f} hang-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} palm.sp(\textsc{f}) \textsc{comit}-stand-raised\_surface.\textsc{nom} \textsc{adju}\\
    \glt `They looked around. They looped around. Her sisters' skulls had been strung with a line of inner bark, and were hanging with a line of inner bark through them, on the \textit{basioba} palm wall.'
\z

\ea\label{ex:10-a112}
Faya tati tone me hiwasia Hato Abono okati me kakaba toeino awane.\\
    \gll faya tati tone me hi-wasiha Hato\_Abono okati me ka-kaba to-ha-hino awa-ne\\
    so head bone 3\textsc{pl}.\textsc{s} \textsc{oc}-find.\textsc{f} Hato\_Abono(\textsc{m}) 1\textsc{sg}.\textsc{poss}.older\_sister(\textsc{f}) 3\textsc{pl}.\textsc{o} \textsc{dup}-eat \textsc{ch}-\textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{m} seem.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `They found their skulls. ``Hato Abono has been eating my older sisters.'''
\z

\ea\label{ex:10-a113}
Yifo, yifo ti ne moni saiba nematamona hike ya ahi, teo, teo, teo, teo.\\
    \gll yifo yifo tii ne moni sai-baa na-himata-mona hike ya ahi teo teo teo teo\\
    firewood(\textsc{f}) firewood(\textsc{f}) cut\_through \textsc{aux}.\textsc{m} sound be\_heard-at\_a\_distance \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} far \textsc{adju} then (sound\_of\_hitting) (sound\_of\_hitting) (sound\_of\_hitting) (sound\_of\_hitting)\\
    \glt `The sound of him cutting firewood could be heard in the distance, \textit{teo}, \textit{teo}, \textit{teo}, \textit{teo}.'
\z

\ea\label{ex:10-a114}
Ati e towakamahi.\\
    \gll ati ee to-ka-ka-ma-hi\\
    older\_sister(\textsc{f}) 1\textsc{in}.\textsc{s} away-\textsc{comit}-go-back-\textsc{imp}.\textsc{f}\\
    \glt `{``}Sister, let's go.'''
\z

\ea\label{ex:10-a115}
Faya me yana tokanemetemone fahi, yifo tabo me itikimakabote.\\
    \gll faya me yana to-ka-na-hemete-mone fahi yifo tabo me iti-kima-kabote\\
    so 3\textsc{pl}.\textsc{s} start \textsc{ch}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then hammock(\textsc{m}) bunch.\textsc{m} 3\textsc{pl}.\textsc{s} pick\_up-two-quickly\\
    \glt `So they got ready to go, quickly grabbing their rolled-up hammocks.'
\z

\ea\label{ex:10-a116}
Me kana ni towakamemetemoneke ahi, me ka ati, tati me awa mati, tati tone.\\
    \gll me kana ni to-ka-ka-ma-hemete-mone-ke ahi me ka ati tati me awa mati tati tone\\
    3\textsc{pl}.\textsc{s} run \textsc{aux}.\textsc{nfin} away-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then 3\textsc{pl} \textsc{poss} older\_sister(\textsc{f}) head 3\textsc{pl}.\textsc{s} see.\textsc{f} 3\textsc{pl}.\textsc{s} head bone\\
    \glt `They went running away because they had seen their sisters' skulls.'
\z

\ea\label{ex:10-a117}
Me kana ni towakamemetemoneke.\\
    \gll me kana ni to-ka-ka-ma-hemete-mone-ke\\
    3\textsc{pl}.\textsc{s} run \textsc{aux}.\textsc{nfin} away-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They went running away.'
\z

\ea\label{ex:10-a118}
Me kana ni towakama me, kamematamonaka yifo tiwa ni.\\
    \gll me kana ni to-ka-ka-ma me ka-ma-himata-mona-ka yifo tiwa ni\\
    3\textsc{pl}.\textsc{s} run \textsc{aux}.\textsc{nfin} away-\textsc{comit}-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} firewood(\textsc{f}) carry \textsc{aux}.\textsc{nfin}\\
    \glt `They went running away. He came back carrying the firewood.'
\z

\newpage

\ea\label{ex:10-a119}
Yifo tiwa ni kame, yifo koro nisematamonaka.\\
    \gll yifo tiwa ni ka-me yifo koro na-risa-himata-mona-ka\\
    firewood(\textsc{f}) carry \textsc{aux}.\textsc{nfin} come-back.\textsc{m} firewood(\textsc{f}) throw \textsc{aux}-down-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He came back carrying the firewood. He threw the firewood down.'
\z

\ea\label{ex:10-a120}
Oof, oof, oof, haa, fahara?\\
    \gll oof oof oof haa faha-ra\\
    hoo hoo hoo (expression\_of\_exhaustion) water(\textsc{f})-\textsc{npq}\\
    \glt `{``}Whew, where is some water?'''
\z

\ea\label{ex:10-a121}
Me afi kanabone me afi kana me awine?\\
    \gll me afi ka-na-habone me afi ka-na me awine\\
    3\textsc{pl}.\textsc{s} bathe \textsc{comit}-\textsc{aux}-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{s} bathe \textsc{comit}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} seem.\textsc{f}\\
    \glt `{``}They said they were going to bathe, are they bathing?'''
\z

\ea\label{ex:10-a122}
Me watariya.\\
    \gll me wata-ri-ya\\
    3\textsc{pl}.\textsc{s} exist-\textsc{neg}.\textsc{f}-now\\
    \glt `{``}They aren't here.'''
\z

\ea\label{ex:10-a123}
Me towakamiya.\\
    \gll me to-ka-ka-mi-ya\\
    3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back.\textsc{f}-now\\
    \glt `{``}They went away.'''
\z

\ea\label{ex:10-a124}
Me towakama me nowati ra ha ne mera ha ne Hii, hii, te afi kini?\\
    \gll me to-ka-ka-ma me nowati ra haa ne mera haa ne hii hii tee afi ka-ni\\
    3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} after \textsc{o} call \textsc{aux}.\textsc{m} 3\textsc{pl}.\textsc{o} call \textsc{aux}.\textsc{m} yoohoo yoohoo 2\textsc{pl}.\textsc{s} bathe \textsc{comit}-\textsc{bkg}.\textsc{f}\\
    \glt `They had gone, and he called after them. He called them. ``Yoohoo, are you bathing?'''
\z

\ea\label{ex:10-a125}
Yama fafawa.\\
    \gll yama fa-fawa\\
    thing(\textsc{f}) \textsc{dup}-disappear\\
    \glt `They had disappeared.'
\z

\ea\label{ex:10-a126}
Yama fafawa nemetemoneke ahi.\\
    \gll yama fa-fawa na-hemete-mone-ke ahi\\
    thing(\textsc{f}) \textsc{dup}-disappear \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `They had disappeared.'
\z

\ea\label{ex:10-a127}
Me ha remetemoneke.\\
    \gll me haa ra-hemete-mone-ke\\
    3\textsc{pl}.\textsc{s} call \textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They didn't call.'
\z

\ea\label{ex:10-a128}
Me ha ra me, me ha remetemoneni.\\
    \gll me haa ra me me haa ra-hemete-mone-ni\\
    3\textsc{pl}.\textsc{s} call \textsc{neg}.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} call \textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `They didn't call.'
\z

\ea\label{ex:10-a129}
Kame, yama siba nematamonaka ahi.\\
    \gll ka-me yama siba na-himata-mona-ka ahi\\
    come-back.\textsc{m} thing(\textsc{f}) look\_for \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `He came back. He looked around.'
\z


\ea\label{ex:10-a130}
Hika, me ka yama watakare.\\
    \gll hika me ka yama wata-ka-re\\
    where 3\textsc{pl} \textsc{poss} thing(\textsc{f}) exist-\textsc{decl}-\textsc{neg}.\textsc{f}\\
    \glt `{``}What, their things aren't here.'''
\z

\ea\label{ex:10-a131}
Hika yama me towakamarini?\\
    \gll hika yama me to-ka-ka-ma-ri-ni\\
    where thing(\textsc{f}) 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back-\textsc{\textsc{cq}}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}\\
    \glt `{``}Where did they take the things?'''
\z

\ea\label{ex:10-a132}
Toto abono yana tonematamonaka.\\
    \gll toto abono yana to-na-himata-mona-ka\\
    bamboo.sp(\textsc{f}) spirit start \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `The \textit{toto} spirit was ready to leave.'
\z

\ea\label{ex:10-a133}
Yana tone, yana tone kana ni tokomakematamona ahi.\\
    \gll yana to-ne yana to-ne kana ni to-ka-make-himata-mona ahi\\
    start \textsc{ch}-\textsc{aux}.\textsc{m} start \textsc{ch}-\textsc{aux}.\textsc{m} run \textsc{aux}.\textsc{nfin} away-go-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `He was ready to leave. He went running after them.'
\z

\ea\label{ex:10-a134}
Kana ni tokomake, yama soki, yama soki bata hikane, wete name, kobo name, Hoo, ofimirimakoni me fawa kineni.\\
    \gll kana ni to-ka-make yama soki yama soki bata hi-ka-ne wete na-me kobo na-me hoo o-fimi-ra-makoni me fawa ka-ne-ni\\
    run \textsc{aux}.\textsc{nfin} away-go-following thing(\textsc{f}) be\_dark.\textsc{nfin} thing(\textsc{f}) be\_dark.\textsc{nfin} surprise \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} return \textsc{aux}-back.\textsc{m} arrive \textsc{aux}-back.\textsc{m} humph 1\textsc{sg}.\textsc{s}-be\_hungry-\textsc{neg}-for\_this\_reason.\textsc{f} 3\textsc{pl}.\textsc{s} disappear \textsc{comit}-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `He went running after them. Darkness overtook him. He came back. He arrived back. ``Oh boy, because I wasn't hungry they disappeared.'''
\z

\ea\label{ex:10-a135}
Ofimi ya owa ni me fawa kineke.\\
    \gll o-fimi ya owa ni me fawa ka-ne-ke\\
    1\textsc{sg}.\textsc{s}-be\_hungry.\textsc{nom} \textsc{adju} 1\textsc{sg} to 3\textsc{pl}.\textsc{s} disappear \textsc{comit}-\textsc{cont}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Now I'm hungry, and they aren't here.'''
\z

\ea\label{ex:10-a136}
Ofimi oke haro.\\
    \gll o-fimi o-ke haaro\\
    1\textsc{sg}.\textsc{s}-be\_hungry 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}I'm hungry now.'''
\z

\ea\label{ex:10-a137}
Faya yifo afo kane, ene hiyarebanoho.\\
    \gll faya yifo afo ka-ne ehene hiya-ra-hiba-no-ho\\
    so fire(\textsc{f}) light \textsc{comit}-\textsc{aux}.\textsc{m} deed.\textsc{m} be\_bad-\textsc{neg}-\textsc{fut}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `He lit a fire, and then did something bad.'
\z

\ea\label{ex:10-a138}
Faya fimi amo kane, ati sai nabe, Ofimiaba ama okanani, oko bani me okobabone me ona me towakamani me me okobabone me ona me towakamine ninihi.\\
    \gll faya fimi amo ka-ne ati sai nahabe o-fimi-haba ama o-ka na-hani o-ka bani me o-kaba-habone me o-na me to-ka-ka-ma-hani me me o-kaba-habone me o-na me to-ka-ka-ma-ne ni-ni-hi\\
    so hunger(\textsc{f}) sleep \textsc{comit}-\textsc{aux}.\textsc{m} voice be\_heard at\_night.\textsc{m} 1\textsc{sg}.\textsc{s}-be\_hungry-\textsc{fut}.\textsc{f} \textsc{sec} 1\textsc{sg}.\textsc{s}-\textsc{decl} \textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{f} 1\textsc{sg}.\textsc{poss}-\textsc{poss} animal(\textsc{m}) 3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-eat-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-say.\textsc{f} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-eat-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{o} 1\textsc{sg}.\textsc{s}-say.\textsc{f} 3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back-\textsc{cont}.\textsc{f} \textsc{aux}.\textsc{nom}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
    \glt `He went to sleep hungry. He could be heard in the night. ``Now I'm going to be hungry. When I was going to eat my meat, they went away. I was going to eat them, and they went away.'''
\z

\ea\label{ex:10-a139}
Faya yama wamini ya, yama waminemetemone amake.\\
    \gll faya yama waha-mini ya yama waha-mina-hemete-mone ama-ke\\
    so thing(\textsc{f}) dawn-morning.\textsc{nom}.\textsc{f} \textsc{adju} thing(\textsc{f}) dawn-morning-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f}\\
    \glt `In the morning it dawned.'
\z

\ea\label{ex:10-a140}
Faya Oko sobo okobamakoni, sobo haro.\\
    \gll faya o-ka sobo o-koba-makoni sobo haaro\\
    so 1\textsc{sg}.\textsc{poss}-\textsc{poss} belly(\textsc{f}) 1\textsc{sg}.\textsc{s}-hit-for\_this\_reason.\textsc{f} belly(\textsc{f}) that\_one.\textsc{f}\\
    \glt `{``}I'll eat my belly.'''\footnote{Siko should have said \textit{okobimakoni}.}
\z

\ea\label{ex:10-a141}
Hoho, bayikene, bayikene.\\
    \gll hoho bayi-ke-ne bayi-ke-ne\\
    look be\_thick-\textsc{decl}.\textsc{f}-\textsc{irr}.\textsc{f} be\_thick-\textsc{decl}.\textsc{f}-\textsc{irr}.\textsc{f}\\
    \glt `{``}Look, it's thick.'''
\z

\ea\label{ex:10-a142}
Okobabanake.\\
    \gll o-kaba-habana-ke\\
    1\textsc{sg}.\textsc{s}-eat-\textsc{fut}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I'm going to eat it.'''
\z

\ea\label{ex:10-a143}
Ofimi oke.\\
    \gll o-fimi o-ke\\
    1\textsc{sg}.\textsc{s}-be\_hungry 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I'm hungry.'''
\z

\ea\label{ex:10-a144}
Kero nawamemetemone ahi.\\
    \gll kero na-waha-ma-hemete-mone ahi\\
    be\_curved \textsc{aux}-change-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `It was circular.'
\z

\ea\label{ex:10-a145}
Wisi hisemetemone ahi.\\
    \gll wisi hi-kosa-hemete-mone ahi\\
    cut\_up \textsc{oc}-middle-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `He cut it out.'
\z

\ea\label{ex:10-a146}
Wisi hisa, yoto.\\
    \gll wisi hi-kosa yoto\\
    cut\_up \textsc{oc}-middle.\textsc{list} intestines.\textsc{m}\\
    \glt `He cut the belly out.'\footnote{Probably \textit{yoto} begins a new sentence which Siko did not finish.}
\z

\ea\label{ex:10-a147}
Faya tabasi hinemetemoneke, sobori.\\
    \gll faya tabasi hi-na-hemete-mone-ke sobori\\
    so roast\_on\_rack \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} belly\\
    \glt `Then he roasted the belly.'
\z

\ea\label{ex:10-a148}
Sobori tabasi hina, tabasi hina kaha, ye sosobi hina iyawa sisirihina, iyawa sisirihina nematamonaka.\\
    \gll sobori tabasi hi-na tabasi hi-na kaha yehe so-sobi hina iyawa si-sira-hina iyawa si-sira-hina na-himata-mona-ka\\
    belly roast\_on\_rack \textsc{oc}-\textsc{aux}.\textsc{f} roast\_on\_rack \textsc{oc}-\textsc{aux}.\textsc{f} be\_toasted.\textsc{list} hand \textsc{dup}-lick at\_the\_same\_time.\textsc{list} grated\_mash(\textsc{f}) \textsc{dup}-toast-at\_the\_same\_time.\textsc{list} grated\_mash(\textsc{f}) \textsc{dup}-toast-at\_same\_time \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He roasted the belly. He roasted it. It was ready. He licked his fingers. He toasted manioc mash.'
\z

\ea\label{ex:10-a149}
Iyawa sire iyawa kakabihina, sobori kabematamonaka ahi.\\
    \gll iyawa sire iyawa ka-kaba-hina sobori kaba-himata-mona-ka ahi\\
    grated\_mash(\textsc{f}) toast.\textsc{m} grated\_mash(\textsc{f}) \textsc{comit}-eat-at\_the\_same\_time.\textsc{list} belly eat-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `He toasted manioc mash. He ate some manioc mash. He ate his belly.'
\z

\ea\label{ex:10-a150}
Sobori kabe, sobori kabe, sobori kabematamonaka.\\
    \gll sobori kabe sobo-ri kabe sobori kaba-himata-mona-ka\\
    belly eat.\textsc{m} belly(\textsc{f})-\textsc{pn} eat.\textsc{m} belly eat-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He ate his belly.'
\z

\ea\label{ex:10-a151}
Sobori kabe, sobori kabe ya sobori kabi abe ya, toro tama kanematamona ahi, honofe.\\
    \gll sobori kabe sobori kabe ya sobori kabi ahabe ya toro tama ka-na-himata-mona ahi honofe\\
    belly eat.\textsc{m} belly eat.\textsc{nom}.\textsc{m} \textsc{adju} belly eat.\textsc{nfin} end.\textsc{nom}.\textsc{m} \textsc{adju} guts.\textsc{m} hold\_onto \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then plant.sp(\textsc{m})\\
    \glt `He ate his belly. After he was done eating his belly, he covered the middle of his body with a wild banana leaf.'
\z

\ea\label{ex:10-a152}
Fa hawa tonihawame hinaka fa kawita, fa fawehimatamonaka.\\
    \gll faha hawa to-niha-waha-me hinaka faha ka-ita faha fawa-himata-mona-ka\\
    water(\textsc{f}) be\_finished \textsc{ch}-\textsc{caus}-change-back.\textsc{m} 3\textsc{sg}.\textsc{poss} water(\textsc{f}) \textsc{comit}-sit.\textsc{f} water(\textsc{f}) drink-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He got water ready. His water was in the vessel. He drank water.'
\z

\ea\label{ex:10-a153}
Fa fawematamonaka.\\
    \gll faha fawa-himata-mona-ka\\
    water(\textsc{f}) drink-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He drank water.'
\z

\ea\label{ex:10-a154}
Fa fawe fa nafi fawe, hinaka fa taha tonemetemoneke ahi, ssssss.\\
    \gll faha fawe faha nafi fawe hinaka faha taha to-na-hemete-mone-ke ahi ssssss\\
    water(\textsc{f}) drink.\textsc{m} water(\textsc{f}) all drink.\textsc{m} 3\textsc{sg}.\textsc{poss} water(\textsc{f}) scatter \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then (sound\_of\_water)\\
    \glt `He drank water. He drank all the water. His water came out, \textit{sssssss}.'
\z

\ea\label{ex:10-a155}
See, oko fa ababa amakira.\\
    \gll see o-ka faha ahaba-haba ama-kira\\
    oh\_no 1\textsc{sg}.\textsc{poss}-\textsc{poss} water(\textsc{f}) spill-\textsc{fut}.\textsc{f} \textsc{sec}-\textsc{mood}.\textsc{f}\\
    \glt `{``}Oh no, my water came out.'''\footnote{The suffix \textit{-kira}/\textit{-kara} is an archaic mood suffix (\cite[42f]{Vogel:2022du}). It is used by Siko and Yowao in traditional stories, but younger Jarawaras do not use it.}
\z

\ea\label{ex:10-a156}
Oko fa ababa amakira.\\
    \gll o-ka faha ahaba-haba ama-kira\\
    1\textsc{sg}.\textsc{poss}-\textsc{poss} water(\textsc{f}) spill-\textsc{fut}.\textsc{f} \textsc{sec}-\textsc{mood}.\textsc{f}\\
    \glt `{``}My water came out.'''
\z

\ea\label{ex:10-a157}
Te awa.\\
    \gll tee awa\\
    2\textsc{pl}.\textsc{s} see\\
    \glt `{``}Look!'''
\z

\ea\label{ex:10-a158}
Oko fa ababa amakenihi.\\
    \gll o-ka faha ahaba-haba ama-ke-ni-hi\\
    1\textsc{sg}.\textsc{poss}-\textsc{poss} water(\textsc{f}) spill-\textsc{fut}.\textsc{f} \textsc{sec}-\textsc{decl}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
    \glt `{``}My water came out.'''
\z

\ea\label{ex:10-a159}
Faya toro wete nakose, honofe nawane toro wete nakose, faya winematamonaka, tobokomisari ahi.\\
    \gll faya toro wete na-kose honofe na-wane toro wete na-kose faya wina-himata-mona-ka to-bako-misa-haari ahi\\
    so guts.\textsc{m} tie \textsc{aux}-middle.\textsc{m} plant.sp(\textsc{m}) \textsc{caus}-be\_attached.\textsc{m} guts.\textsc{m} tie \textsc{aux}-middle.\textsc{m} so lie\_in\_hammock-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} away-lie\_belly\_up-up-\textsc{ip}.\textsc{e}.\textsc{m} then\\
    \glt `So he tied inner bark around the middle of his body. He placed wild banana leaves in. He tied the middle of his body. Then he lay in his hammock, on his back.'
\z

\ea\label{ex:10-a160}
Tobokomise winematamona ahi.\\
    \gll to-bako-mise wina-himata-mona ahi\\
    away-lie\_belly\_up-up.\textsc{m} lie\_in\_hammock-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `He lay in his hammock on his back.'
\z

\ea\label{ex:10-a161}
Hoo, bakasi owa oke, fa fawarama ora owa.\\
    \gll hoo bakasi o-awa o-ke faha fawa-rama o-ra owa\\
    humph thirst(\textsc{f}) 1\textsc{sg}.\textsc{s}-feel 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f} water(\textsc{f}) drink-unusual 1\textsc{sg}.\textsc{s}-\textsc{neg}.\textsc{f} 1\textsc{sg}.\textsc{s}\\
    \glt `{``}Hoo, I'm thirsty, because I couldn't drink water.'''
\z

\ea\label{ex:10-a162}
Fa fawarama ora bakasi owine oke.\\
    \gll faha fawa-rama o-ra bakasi o-awa-ne o-ke\\
    water(\textsc{f}) drink-unusual 1\textsc{sg}.\textsc{s}-\textsc{neg}.\textsc{f} thirst(\textsc{f}) 1\textsc{sg}.\textsc{s}-feel-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}-\textsc{decl}.\textsc{f}\\
    \glt `{``}I couldn't drink water and I'm thirsty.'''
\z

\ea\label{ex:10-a163}
Faya fa fawamatase hinaka fa taha tona tasi tona, See, kasiroarake.\\
    \gll faya faha fawa-ma-tase hinaka faha taha to-na tasi to-na see kasiro-hara-ke\\
    so water(\textsc{f}) drink-back-again.\textsc{m} 3\textsc{sg}.\textsc{poss} water(\textsc{f}) scatter \textsc{ch}-\textsc{aux}.\textsc{f} emerge \textsc{ch}-\textsc{aux}.\textsc{f} oh\_no persist-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He drank some more water. His water came pouring out. It came out. ``Oh no, it doesn't stop.'''
\z

\ea\label{ex:10-a164}
Faya towiname, towinamematamonaka.\\
    \gll faya to-wina-me to-wina-ma-himata-mona-ka\\
    so away-lie\_in\_hammock-back.\textsc{m} away-lie\_in\_hammock-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `So he went back and lay in his hammock again.'
\z

\ea\label{ex:10-a165}
Hoo.\\
    \gll hoo\\
    whew\\
    \glt `{``}Whew!'''
\z

\ea\label{ex:10-a166}
Faya toro komene yana ne, toro komene yana nematamona fahi.\\
    \gll faya toro komene yana ne toro komene yana na-himata-mona fahi\\
    so guts.\textsc{m} pain.\textsc{m} start \textsc{aux}.\textsc{m} guts.\textsc{m} pain.\textsc{m} start \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `His gut started to hurt.'
\z

\ea\label{ex:10-a167}
Faya kawine fare wine ya, fare wine ya batematamonaka fahi.\\
    \gll faya ka-wine fare wine ya fare wine ya bata-himata-mona-ka fahi\\
    so \textsc{comit}-lie\_in\_hammock.\textsc{m} same.\textsc{m} lie\_in\_hammock.\textsc{nom}.\textsc{m} \textsc{adju} same.\textsc{m} lie\_in\_hammock.\textsc{nom}.\textsc{m} \textsc{adju} be\_rotten-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `He lay sick in the hammock. He rotted right where he lay.'
\z

\ea\label{ex:10-a168}
Wabori, wabori kamaki, wabori kamakimatamonane.\\
    \gll wabori wabori ka-maki wabori ka-maki-himata-mona-ne\\
    3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) come-following.\textsc{m} 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) come-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `His brother-in-law came.'
\z

\ea\label{ex:10-a169}
Toto Abono wini nofe amake, era kakaba.\\
    \gll Toto\_Abono wina nofe ama-ke era kakaba\\
    Toto\_Abono(\textsc{m}) live recently.\textsc{nom}.\textsc{m} be-\textsc{decl}.\textsc{f} 1\textsc{in}.\textsc{o} eater\\
    \glt `{``}Toto Abono, the people eater, lives here.'''
\z

\ea\label{ex:10-a170}
Era kakaba wini nofe amake ahi.\\
    \gll era kakaba wina nofe ama-ke ahi\\
    1\textsc{in}.\textsc{o} eater live recently.\textsc{nom}.\textsc{m} be-\textsc{decl}.\textsc{f} here\\
    \glt `{``}The people eater lives here.'''
\z

\ea\label{ex:10-a171}
Hike hike ya mao hisi naba hinematamonaka hike ya ahi.\\
    \gll hike hike ya maho hisi na-baa hi-na-himata-mona-ka hike ya ahi\\
    far far \textsc{adju} smell.\textsc{m} smell \textsc{aux}-at\_a\_distance \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} far \textsc{adju} then\\
    \glt `He smelled him from far out.'
\z

\ea\label{ex:10-a172}
Hanf, hanf, hanf, efe amare?\\
    \gll hanf hanf hanf efe ama-re\\
    peeoo peeoo peeoo this.\textsc{m} be-\textsc{neg}.\textsc{m}\\
    \glt `{``}Peeoo, is that him?'''
\z

\ea\label{ex:10-a173}
Batawaha.\\
    \gll bata-waha\\
    be\_rotten-change.\textsc{cont}.\textsc{m}\\
    \glt `{``}He has started to rot.'''
\z

\ea\label{ex:10-a174}
Yoto ene batehino, batematamonaka.\\
    \gll yoto ehene bata-hino bata-himata-mona-ka\\
    intestines.\textsc{m} because\_of.\textsc{m} be\_rotten-\textsc{ip}.\textsc{n}.\textsc{m} be\_rotten-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He was rotting because of his guts.'
\z

\ea\label{ex:10-a175}
Batehino abowi winematamona fahi.\\
    \gll bata-hino abohi wina-himata-mona fahi\\
    be\_rotten-\textsc{ip}.\textsc{n}.\textsc{m} dead\_body lie\_in\_hammock-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `He was rotting. He was dead, lying in the hammock.'
\z

\ea\label{ex:10-a176}
Faya abowi winehino hiwasimaki wabori hora ne, Toto Abono himatabiyo hiwa wisi neno yama yete re nanako raba nofe hiwa wisi nawa nano.\\
    \gll faya abohi wina-hino hi-wasi-maki wabori hora ne Toto\_Abono himatabiyo hiwa wisi na-hino yama yete re na-nako raba nofe hiwa wisi na-waha na-no\\
    so dead\_body lie\_in\_hammock-\textsc{ip}.\textsc{n}.\textsc{m} \textsc{oc}-find-following.\textsc{m} 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) scold \textsc{aux}.\textsc{m} Toto\_Abono(\textsc{m}) why 3\textsc{sg}.\textsc{refl}.\textsc{m} cut\_up \textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{m} thing(\textsc{f}) hunt \textsc{neg}.\textsc{m} \textsc{caus}-be\_famished a\_bit recently.\textsc{m} 3\textsc{sg}.\textsc{refl}.\textsc{m} cut\_up \textsc{aux}-change.\textsc{cont}.\textsc{m} \textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}\\
    \glt `He yelled at his brother-in-law, who he had found lying dead in his hammock. ``Why did Toto Abono cut himself open? He didn't hunt. He was always hungry. Now he has cut himself open.'''
\z

\ea\label{ex:10-a177}
Okasima kabi nofarawenoho.\\
    \gll okasima kaba nofa-rawa-hino-ho\\
    1\textsc{sg}.\textsc{poss}.younger\_sister(\textsc{f}) eat recently-\textsc{fpl}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `{``}He was eating my sisters.'''
\z

\ea\label{ex:10-a178}
Okasima kabi nofaraweno kakahiwa reba okasima niki nofarawahari.\\
    \gll okasima kaba nofa-rawa-hino ka-ka-hiwa ra-hiba okasima niki nofa-rawa-haari\\
    1\textsc{sg}.\textsc{poss}.younger\_sister(\textsc{f}) eat recently-\textsc{fpl}-\textsc{ip}.\textsc{n}.\textsc{m} \textsc{dup}-\textsc{comit}-have\_children \textsc{neg}-\textsc{fut}.\textsc{m} 1\textsc{sg}.\textsc{poss}.younger\_sister(\textsc{f}) press\_on recently-\textsc{fpl}-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `{``}He was eating my sisters. He didn't want to have children. He was strangling my sisters.'''
\z

\ea\label{ex:10-a179}
Hiwa kawinaka.\\
    \gll hiwa ka-wina-ka\\
    only.\textsc{m} \textsc{comit}-lie\_in\_hammock-\textsc{decl}.\textsc{m}\\
    \glt `{``}Let him lie dead in his hammock.'''
\z

\ea\label{ex:10-a180}
Hekanakosematamonaka.\\
    \gll hi-ka-kana-kosa-himata-mona-ka\\
    \textsc{oc}-\textsc{comit}-leave-middle-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He left him there.'
\z

\ea\label{ex:10-a181}
Tokomematamonaka.\\
    \gll to-ka-ma-himata-mona-ka\\
    away-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went home.'
\z

\ea\label{ex:10-a182}
Faya boko, boko yome hirawematamona fahi.\\
    \gll faya boko boko yome hi-rawa-himata-mona fahi\\
    so vulture(\textsc{f}) vulture(\textsc{f}) swallow \textsc{oc}-\textsc{fpl}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `So the vultures ate him.'
\z

\ea\label{ex:10-a183}
Boko yome hirawe, faya toto abono abe, abowi wine hekanakose hikaminame, Toto Abono winaka ahi, hiwa bobi nenoho.\\
    \gll boko yome hi-rawe faya toto abono ahabe abohi wine hi-ka-kana-kose hi-kamina-me Toto\_Abono wina-ka ahi hiwa bobi na-hino-ho\\
    vulture(\textsc{f}) swallow \textsc{oc}-\textsc{fpl}.\textsc{m} so bamboo.sp(\textsc{f}) spirit die.\textsc{m} dead\_body lie\_in\_hammock.\textsc{m} \textsc{oc}-\textsc{comit}-leave-middle.\textsc{m} \textsc{oc}-tell-back.\textsc{m} Toto\_Abono(\textsc{m}) lie\_in\_hammock-\textsc{decl}.\textsc{m} there 3\textsc{sg}.\textsc{refl}.\textsc{m} slit \textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `The vultures ate him. So the toto spirit died. He lay dead in the hammock. He left him there. He told about him back home. ``Toto Abono is lying there in the hammock. He had cut himself up.'''
\z

\ea\label{ex:10-a184}
Toto Abono abareka.\\
    \gll Toto\_Abono ahaba-hare-ka\\
    Toto\_Abono(\textsc{m}) die-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `{``}Toto Abono died.'''
\z

\ea\label{ex:10-a185}
Abawahe.\\
    \gll ahaba-wahe\\
    die-change.\textsc{m}\\
    \glt `{``}He had died.'''
\z

\ea\label{ex:10-a186}
Me ferene me kakamani me amake haro.\\
    \gll me fere-hene me ka-ka-ma-hani me ama-ke haaro\\
    3\textsc{pl}.\textsc{s} wait-\textsc{irr}.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{comit}-come-back-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{sec}-\textsc{decl}.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}They didn't wait, the two of them came home.'''
\z

\ea\label{ex:10-a187}
Me hikabene me amake.\\
    \gll me hi-kaba-hene me ama-ke\\
    3\textsc{pl}.\textsc{o} \textsc{oc}-eat-\textsc{irr}.\textsc{f} 3\textsc{pl}.\textsc{o} \textsc{sec}-\textsc{decl}.\textsc{f}\\
    \glt `{``}He would have eaten them.'''
\z


\ea\label{ex:10-a188}
Toto Abono okati kabi nofaraweno okati tati tone ota wasirawarani.\\
    \gll Toto\_Abono okati kaba nofa-rawa-hino okati tati tone otaa wasi-rawa-hara-ni\\
    Toto\_Abono(\textsc{m}) 1\textsc{sg}.\textsc{poss}.older\_sister(\textsc{f}) eat recently-\textsc{fpl}-\textsc{ip}.\textsc{n}.\textsc{m} 1\textsc{sg}.\textsc{poss}.older\_sister(\textsc{f}) head bone 1\textsc{ex}.\textsc{s} find-\textsc{fpl}-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}Toto Abono had eaten our older sisters, and we found their skulls.'''
\z

\ea\label{ex:10-a189}
Ota nofare ota kakose, yama siba na nanoho.\\
    \gll otaa nofa-re otaa ka-kana-kose yama siba na na-no-ho\\
    1\textsc{ex}.\textsc{s} like-\textsc{neg}.\textsc{m} 1\textsc{ex}.\textsc{s} \textsc{comit}-leave-middle.\textsc{m} thing(\textsc{f}) look\_for \textsc{aux}.\textsc{cont}.\textsc{m} \textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `{``}We didn't like him. We left him behind, looking for us.'''
\z

\newpage
\ea\label{ex:10-a190}
Me abe, me hekanakose me, hiwa kabe abawahematamonaka fahi.\\
    \gll me ahabe me hi-ka-kana-kose me hiwa kabe ahaba-waha-himata-mona-ka fahi\\
    3\textsc{pl} die.\textsc{m} 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-leave-middle.\textsc{m} 3\textsc{pl} 3\textsc{sg}.\textsc{refl}.\textsc{m} eat.\textsc{m} die-change-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `He died. They left him behind. He ate himself. Then he died.'\footnote{\textit{Me} at the beginning is a mistake, also the \textit{me} after \textit{hekanakose}.}
\z

\ea\label{ex:10-a191}
Faya abe, faya abematamonaka ahi.\\
    \gll faya ahabe faya ahaba-himata-mona-ka ahi\\
    so die.\textsc{m} so die-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `He died.'
\z

\ea\label{ex:10-a192}
Abe, boko hikabarawematamonaka, boko, boko sawi.\\
    \gll ahabe boko hi-kaba-rawa-himata-mona-ka boko boko\_sawi\\
    die.\textsc{m} vulture(\textsc{f}) \textsc{oc}-eat-\textsc{fpl}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} vulture(\textsc{f}) king\_vulture(\textsc{m})\\
    \glt `He died. The vultures ate him, the king vultures.'
\z

\ea\label{ex:10-a193}
Boko sawi me hikabematamonane.\\
    \gll boko\_sawi me hi-kaba-himata-mona-ne\\
    king\_vulture(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{oc}-eat-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `The king vultures ate him.'
\z

\ea\label{ex:10-a194}
Hinaka fowa ta kawita, hihiti watarari.\\
    \gll hinaka fowa taa ka-ita hihiti wata-ra-haari\\
    3\textsc{sg}.\textsc{poss} manioc(\textsc{m}) contrast \textsc{comit}-sit.\textsc{list} person\_who\_grates exist-\textsc{neg}-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `But his manioc was still there, with no one to grate it.'
\z

\ea\label{ex:10-a195}
Faya me hiwamarematamona fahi.\\
    \gll faya me hi-awa-ma-ra-himata-mona fahi\\
    so 3\textsc{pl}.\textsc{s} \textsc{oc}-see-back-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `They didn't see him anymore.'
\z

\newpage
\ea\label{ex:10-a196}
Toto Abono te tonakamariyahi.\\
    \gll Toto\_Abono tee to-na-ka-ma-ra-yahi\\
    Toto\_Abono(\textsc{m}) 2\textsc{pl}.\textsc{s} away-\textsc{caus}-go-back-\textsc{neg}-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `{``}Don't go after Toto Abono anymore.'''
\z

\ea\label{ex:10-a197}
Toto Abono bate winarene.\\
    \gll Toto\_Abono bate wina-hare-ne\\
    Toto\_Abono(\textsc{m}) be\_rotten.\textsc{m} lie\_in\_hammock-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}Toto Abono is lying dead in a hammock.'''
\z

\ea\label{ex:10-a198}
Hiwa abaka.\\
    \gll hiwa ahaba-ka\\
    only.\textsc{m} die-\textsc{decl}.\textsc{m}\\
    \glt `{``}Let him die.'''
\z

\ea\label{ex:10-a199}
Sobori bobi nakoseno ahi.\\
    \gll sobori bobi na-kosa-hino ahi\\
    belly slit \textsc{aux}-middle-\textsc{ip}.\textsc{n}.\textsc{m} there\\
    \glt `{``}He cut up his belly.'''
\z

\ea\label{ex:10-a200}
Sobori bobi nakose sobori kabeinoka, eeto raba rinihi.\\
    \gll sobori bobi na-kose sobori kaba-hino-ka e-eheto raba ri-ni-hi\\
    belly slit \textsc{aux}-middle.\textsc{m} belly eat-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{decl}.\textsc{m} \textsc{dup}-be\_good a\_bit \textsc{neg}.\textsc{nom}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
    \glt `{``}He cut up his belly. He ate his belly, it tasted really good.'''
\z

\ea\label{ex:10-a201}
Toro fene itarisareka, fa faweno karo.\\
    \gll toro fehene ita-risa-hare-ka faha fawa-hino karo\\
    inside.\textsc{m} juice.\textsc{m} sit-down-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m} water(\textsc{f}) drink-\textsc{ip}.\textsc{n}.\textsc{m} which\\
    \glt `{``}The liquid from inside him spilled out, the water which he had drunk.'''
\z

\ea\label{ex:10-a202}
Faya kamematamona fahi.\\
    \gll faya ka-ma-himata-mona fahi\\
    so come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `So he came home.'
\z

\newpage
\ea\label{ex:10-a203}
Kame kobo name, Ami, ami tika koma, tika koma hiwa kabehino, abeno winaka ahi.\\
    \gll ka-me kobo na-me ami ami ti-ka koma ti-ka koma hiwa kaba-hino ahaba-hino wina-ka ahi\\
    come-back.\textsc{m} arrive \textsc{aux}-back.\textsc{m} mother(\textsc{f}) mother(\textsc{f}) 2\textsc{sg}.\textsc{poss}-\textsc{poss} son\_in\_law(\textsc{m}) 2\textsc{sg}.\textsc{poss}-\textsc{poss} son\_in\_law(\textsc{m}) 3\textsc{sg}.\textsc{refl}.\textsc{m} eat-\textsc{ip}.\textsc{n}.\textsc{m} die-\textsc{ip}.\textsc{n}.\textsc{m} lie\_in\_hammock-\textsc{decl}.\textsc{m} there\\
    \glt `He came home. He arrived home. ``Mother, your son-in-law ate himself and he died and is lying there in the hammock.'''
\z

\ea\label{ex:10-a204}
Koma oteharekana ene okobarari, okoto kabaterawari.\\
    \gll koma o-tee-hare-ka-na ehene o-kaba-ra-haari okoto kaba-tee-rawa-haari\\
    call\_son\_in\_law 1\textsc{sg}.\textsc{s}-\textsc{hab}-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}-\textsc{irr}.\textsc{m} victim\_of.\textsc{m} 1\textsc{sg}.\textsc{s}-eat-\textsc{neg}-\textsc{ip}.\textsc{e}.\textsc{m} 1\textsc{sg}.\textsc{poss}.daughter(\textsc{f}) eat-\textsc{hab}-\textsc{fpl}-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `{``}He was my son-in-law, but I didn't eat his game; he ate my daughters.'''\footnote{The first two clauses are an example of the ``no effect'' construction discussed in \citet[200f]{Vogel:2022du}. The idea is that a mother-in-law could expect to receive meat when her son-in-law hunted. But the \textit{toto} spirit, instead of hunting, ate her daughters.}
\z

\ea\label{ex:10-a205}
Afa tabiyo hiwani ra hiri nawa nanoho.\\
    \gll afa tabiyo hi-awa-hani ra hiri na-waha na-no-ho\\
    this.\textsc{f} wanting \textsc{oc}-feel-\textsc{ip}.\textsc{n}.\textsc{f} \textsc{o} do \textsc{aux}-change.\textsc{cont}.\textsc{m} \textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `{``}This is what he asked for.'''
\z

\ea\label{ex:10-a206}
Hiwa abakana.\\
    \gll hiwa ahaba-ka-na\\
    only.\textsc{m} die-\textsc{decl}.\textsc{m}-\textsc{irr}.\textsc{m}\\
    \glt `{``}Let him die.'''
\z

\ea\label{ex:10-a207}
Abarare.\\
    \gll ahaba-ra-hare\\
    die-\textsc{neg}-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `{``}He is dead.'''
\z

\ea\label{ex:10-a208}
Watamahare.\\
    \gll wata-ma-hare\\
    exist-back-\textsc{neg}.\textsc{m}\\
    \glt `{``}He is no more.'''
\z

\ea\label{ex:10-a209}
Batareka.\\
    \gll bata-hare-ka\\
    be\_rotten-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `{``}He has rotted.'''
\z

\ea\label{ex:10-a210}
Himatabiyo hiwa katoma namata onarekana.\\
    \gll himatabiyo hiwa katoma na-mata o-na-hare-ka-na\\
    why 3\textsc{sg}.\textsc{refl}.\textsc{m} attack.\textsc{cont}.\textsc{m} \textsc{aux}.\textsc{nom}.\textsc{m}-(?).\textsc{m} 1\textsc{sg}.\textsc{s}-say-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}-\textsc{irr}.\textsc{m}\\
    \glt `{``}Why did he hurt himself?'''
\z

\ea\label{ex:10-a211}
Batareka.\\
    \gll bata-hare-ka\\
    be\_rotten-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `{``}He has rotted.'''
\z

\ea\label{ex:10-a212}
Itihina rareka.\\
    \gll iti-hina ra-hare-ka\\
    pick\_up-can \textsc{neg}-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `{``}He can't be touched.'''
\z

\ea\label{ex:10-a213}
Winareka, bete.\\
    \gll wina-hare-ka bete\\
    lie\_in\_hammock-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m} rotten.\textsc{m}\\
    \glt `{``}He is lying rotten in his hammock.'''
\z

\ea\label{ex:10-a214}
Bete winareka.\\
    \gll bete wina-hare-ka\\
    rotten.\textsc{m} lie\_in\_hammock-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `{``}He is lying rotten in his hammock.'''
\z

\ea\label{ex:10-a215}
Ota towamebanaka, tone.\\
    \gll otaa to-awa-ma-hibana-ka tone\\
    1\textsc{ex}.\textsc{s} away-see-back-\textsc{fut}.\textsc{m}-\textsc{decl}.\textsc{m} bone\\
    \glt `{``}We will go back there and see his bones.'''
\z

\ea\label{ex:10-a216}
Faya wabori me amo ni tere toi ya wabori me hinakame, tati boko sawi ti hiseno tati hekamakino tati itamakematamonaka.\\
    \gll faya wabori me amo ni terei to-hi ya wabori me hi-na-ka-me tati boko\_sawi tii hi-kosa-hino tati hi-ka-ka-maki-hino tati ita-make-himata-mona-ka\\
    so 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) 3\textsc{pl}.\textsc{poss} sleep \textsc{aux}.\textsc{nfin} be\_three \textsc{ch}-\textsc{aux}.\textsc{nom}.\textsc{f} \textsc{adju} 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{caus}-go-back.\textsc{m} head king\_vulture(\textsc{m}) cut\_through \textsc{oc}-middle-\textsc{ip}.\textsc{n}.\textsc{m} head \textsc{oc}-\textsc{comit}-come-following-\textsc{ip}.\textsc{n}.\textsc{m} head sit-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `After three days they went after their brother-in-law. A king vulture had cut off his head and had taken his head away, and it was on the ground.'
\z

\ea\label{ex:10-a217}
Tati itaro ha.\\
    \gll tati ita-haaro haa\\
    head sit-look.\textsc{f} \textsc{dem}\\
    \glt `{``}There is the head on the ground.'''
\z

\ea\label{ex:10-a218}
Boko me hikabematamonaka fahi.\\
    \gll boko me hi-kaba-himata-mona-ka fahi\\
    vulture(\textsc{f}) 3\textsc{pl}.\textsc{s} \textsc{oc}-eat-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `The vultures ate him.'
\z

\ea\label{ex:10-a219}
Faya me hikabe, hikabe abeno ya, boko tone boko kiro natini wani me towasima, faya tafe, tafatafa raba rinihi, hato abono kabaharo.\\
    \gll faya me hi-kabe hi-kabe ahaba-hino ya boko tone boko kiro nahati-hani waa-hani me to-wasi-ma faya tafe tafa-tafa raba ri-ni-hi hato abono kaba-haaro\\
    so 3\textsc{pl}.\textsc{s} \textsc{oc}-eat.\textsc{m} \textsc{oc}-eat.\textsc{m} die-\textsc{ip}.\textsc{n}.\textsc{m} \textsc{adju} vulture(\textsc{f}) bone vulture(\textsc{f}) gnaw daytime-\textsc{ip}.\textsc{n}.\textsc{f} stand-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{s} away-find-back.\textsc{f} so eat.\textsc{m} \textsc{dup}-eat a\_bit \textsc{neg}.\textsc{nom}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup} bamboo.sp(\textsc{f}) spirit eat-\textsc{ip}.\textsc{e}.\textsc{f}\\
    \glt `They ate him. They came upon a vulture on the ground eating him during the day after he died. It ate a lot as it ate the \textit{hato} spirit.'\footnote{The words \textit{ya} and \textit{boko} are mistakes, so that the string should be \textit{abeno tone boko kiro natini}.}
\z

\newpage
\ea\label{ex:10-a220}
Boko, wareharo haro.\\
    \gll boko waa-re-haaro haaro\\
    vulture(\textsc{f}) stand-raised\_surface-look.\textsc{f} that\_one.\textsc{f}\\
    \glt `{``}There is a vulture sitting on a branch.'''
\z

\ea\label{ex:10-a221}
Wareharo ha, haro.\\
    \gll waa-re-haaro haa haaro\\
    stand-raised\_surface-look.\textsc{f} \textsc{dem} that\_one.\textsc{f}\\
    \glt `{``}It's sitting on a branch.'''
\z

\ea\label{ex:10-a222}
Hinaka yifo bete kawinari hari.\\
    \gll hinaka yifo bete ka-wina-haari haari\\
    3\textsc{sg}.\textsc{poss} hammock(\textsc{m}) rotten.\textsc{m} \textsc{comit}-hang-look.\textsc{m} that\_one.\textsc{m}\\
    \glt `{``}His rotted hammock is hanging there.'''
\z

\ea\label{ex:10-a223}
Ataro winari ha.\\
    \gll ataro wina-haari haa\\
    skin.\textsc{m} lie\_in\_hammock-look.\textsc{m} \textsc{dem}\\
    \glt `{``}His skin is in the hammock.'''
\z

\ea\label{ex:10-a224}
Me kakamemetemone fahi.\\
    \gll me ka-ka-ma-hemete-mone fahi\\
    3\textsc{pl}.\textsc{s} \textsc{comit}-come-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `They came home.'
\z

\ea\label{ex:10-a225}
Me kakama me, me kakama me kobo kanami ya, me kobo kanamemetemoneke.\\
    \gll me ka-ka-ma me me ka-ka-ma me kobo ka-na-mi ya me kobo ka-na-ma-hemete-mone-ke\\
    3\textsc{pl}.\textsc{s} \textsc{comit}-come-back.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} \textsc{comit}-come-back.\textsc{f} 3\textsc{pl}.\textsc{s} arrive \textsc{comit}-\textsc{aux}-back.\textsc{nom}.\textsc{f} \textsc{adju} 3\textsc{pl}.\textsc{s} arrive \textsc{comit}-\textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `They came come. They arrived home.'
\z

\ea\label{ex:10-a226}
Me kobo kanama me, Faira?\\
    \gll me kobo ka-na-ma me fahi-ra\\
    3\textsc{pl}.\textsc{s} arrive \textsc{comit}-\textsc{aux}-back.\textsc{f} 3\textsc{pl}.\textsc{s} that-\textsc{npq}\\
    \glt `They arrived home. ``Well?'''
\z

\ea\label{ex:10-a227}
Faira tina winahare.\\
    \gll fahi-ra ti-na wina-hare\\
    that-\textsc{npq} 2\textsc{sg}.\textsc{s}-say.\textsc{f} live-\textsc{neg}.\textsc{m}\\
    \glt `{``}Well, he isn't living there anymore.'''
\z

\ea\label{ex:10-a228}
Watarareka.\\
    \gll wata-ra-hare-ka\\
    exist-\textsc{neg}-\textsc{ip}.\textsc{e}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `{``}He isn't there.'''
\z

\ea\label{ex:10-a229}
Faya tone ta tori ahi.\\
    \gll faya tone taa to-ri ahi\\
    so bone be\_piled\_up \textsc{ch}-raised\_surface.\textsc{nom} there\\
    \glt `{``}His bones are in a pile.'''
\z

\ea\label{ex:10-a230}
Himatabiyo hiwa wisi na namata oneheno hiwa kabehino?\\
    \gll himatabiyo hiwa wisi na na-mata o-na-hino hiwa kaba-hino\\
    why 3\textsc{sg}.\textsc{refl}.\textsc{m} cut\_up \textsc{aux}.\textsc{cont}.\textsc{m} \textsc{aux}.\textsc{nom}.\textsc{m}-(?).\textsc{m} 1\textsc{sg}.\textsc{o}-say-\textsc{ip}.\textsc{n}.\textsc{m} 3\textsc{sg}.\textsc{refl}.\textsc{m} eat-\textsc{ip}.\textsc{n}.\textsc{m}\\
    \glt `{``}Why did he cut into himself and eat himself?'''
\z

\ea\label{ex:10-a231}
Abawahane, bani me, bani me nahorimakoni.\\
    \gll ahaba-waha-ne bani me bani me naho-ra-makoni\\
    die-change.\textsc{cont}.\textsc{m}-\textsc{bkg}.\textsc{m} animal(\textsc{m}) 3\textsc{pl} animal(\textsc{m}) 3\textsc{pl}.\textsc{s} stand-\textsc{neg}-for\_this\_reason.\textsc{f}\\
    \glt `{``}Now he died. There is game.'''
\z


\ea\label{ex:10-a232}
Toto Abono okasima kanahiwarawareba, okasima yabaraweimata.\\
    \gll Toto\_Abono okasima ka-na-hiwa-rawa-ra-hiba okasima yaba-rawa-himata\\
    Toto\_Abono(\textsc{m}) 1\textsc{sg}.\textsc{poss}.younger\_sister(\textsc{f}) \textsc{comit}-\textsc{caus}-have\_children-\textsc{fpl}-\textsc{neg}-\textsc{fut}.\textsc{m} 1\textsc{sg}.\textsc{poss}.younger\_sister(\textsc{f}) marry-\textsc{fpl}-\textsc{fp}.\textsc{n}.\textsc{m}\\
    \glt `{``}Toto Abono didn't want to make my sisters pregnant, but he married my sisters.'''
\z

\ea\label{ex:10-a233}
Hiwa kabi yotoimata.\\
    \gll hiwa kabi yoto-himata\\
    3\textsc{sg}.\textsc{refl}.\textsc{m} eat.\textsc{nfin} follow-\textsc{fp}.\textsc{n}.\textsc{m}\\
    \glt `{``}In the end he ate himself.'''
\z

\ea\label{ex:10-a234}
Abawane.\\
    \gll ahaba-waha-ne\\
    die-change.\textsc{cont}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}Now he died.'''
\z
