The story \textit{Maiko} was told by Yowao in 1992. The corresponding audio file can be downloaded at: \url{https://www.silbrazil.org/sites/brazil/files/yowao-maiko2.mp3}.

Maiko was a powerful shaman, who was able to kill \textit{maka} monsters. The \textit{maka} monster was occuppying Maiko's fishing place, so Maiko turned himself into a kingfisher and stabbed the monster. Maiko caught some fish, and came home.
	
Maiko's two nephews went out fishing, and they offered a large amount of cooked fish to Maiko. They thought they could embarrass him by offering him more than he could eat, but he ate it all. Then he got them back by cooking two batches of taro for them to eat, and they couldn't eat it all.

Maiko went fishing again, this time with his wife. As he was fishing, she went into the forest to defecate, and there she saw a dead deer. She told him about it, and after investigating, he concluded it had been killed by a \textit{maka} monster. They decided to take the deer home to eat, but the monster held onto the deer, trying to keep Maiko from taking it. But they made it home with the deer, along with the fish that Maiko had caught.

Maiko cooked part of the deer and roasted another part, but they only ate the fish. The monster came to the village after the deer, along with its mate. Maiko gave the deer meat to them. When they refused to leave, Maiko called his familiar spirits to help him. But they were overpowered by the monster. So he took a brand from the fire, and put a spell on it, and threw it at the monsters. This killed the female monster, and her mate took her and went away.