\noindent No one saw those people. They didn't see him. We used to hear the stories about him, Maiko. He killed \textit{maka} monsters, he was a shaman. Maiko killed monsters, because he had a shaman's stone.
 
\noindent He went into the forest to fish. He went into the forest, and he came upon a \textit{maka} monster standing in the shallow water. The monster was standing in the water hitting a reed on the surface of the water.
 
\noindent ``Huh, it is hitting a reed on my water that I was going to fish in. I'm going to spear it.''
 
\noindent He decided what to do. He took snuff. Then he went to where the monster was standing, hitting on the water with a reed. His water went, \textit{koho}.
 
\noindent Then Maiko came. ``\textit{Kere}, \textit{kere},'' he said, imitating the call of a kingfisher.
 
\noindent The monster was standing in the water. The monster had its back turned to him. He aimed at the monster.
 
\noindent \textit{Tiihiiii}, went the sound of the monster being hit [by the kingfisher's beak].
 
\noindent \textit{Toooooo} went the sound of the water. It was the sound of the monster. The sound of the monster's fruits could be heard.\footnote{That is, the monster made a sound like the sound of many fruits falling in the water.}
 
\noindent He came out far away. ``\textit{Kere}, \textit{kere}.'' He went away calling.
 
\noindent He went around and sat down again. ``I speared the one who was hitting a reed on the water where I was going to fish. I will fish.''
 
\noindent Then he fished. He speared fish. He had speared the monster.
 
\noindent He weaved a basket to carry the fish in. He came back to the village.
 
\noindent ``My uncle has come back, carrying fish.''
 
\noindent ``I speared a monster, Boyori [his wife]. It was hitting the water with a reed, and I speared it, and it died,'' he said.
 
\noindent He killed the monster. He told his nephews about the monster back in the village. ``You can fish in the place where it was. I speared it. It died.''
 
\noindent They speared fish. ``My uncle eats a lot. I'm going to offer a feast to my uncle. He will eat fish.'' He planned to offer a feast to his uncle.
 
\noindent He offered a feast to his uncle. They gave him a feast. He ate all their food. One person normally doesn't have such a big stomach.
 
\noindent They cooked two pots of fish. They called him. ``Uncle, uncle.''
 
\noindent ``Yes?'' he said.
 
\noindent ``Uncle, fish for you to eat. There is fish for you to eat on a leaf there.''
 
\noindent ``Right, let it lie there. I will take a look at it.''
 
\noindent They served the fish. He ate the fish all by himself. He didn't have any companion. He ate the fish. He drank the broth, too.
 
\noindent ``There are the two empty pots. I'm going home.'' There was a pile of bones from his fish on the leaf. He went home and lay in his hammock.
 
\noindent ``You made a feast for me by myself. Now I will do the same for you. You will be just like me. You attacked me when I was alone.'' He went home and lay in his hammock.
 
\noindent The two of them went out into the forest again. After they left, he went and got \textit{nakarifa} taro in the garden. He brought back the taro, and cooked it. After he cooked the taro, he put it in a basket. He cooked some more.
 
\noindent They speared fish, and came back. They were angry with their wives. ``We are with women who don't prepare food, we are hungry. They don't prepare food for us,'' they said.
 
\noindent ``Wait a little. Your uncle is planning something for you,'' she said.
 
\noindent ``There are two pots of food for you to eat there. Don't criticize your wives,'' he said. ``You gave me a feast when I was alone. Now I have done the same to you. Eat it all.''
 
\noindent So they ate the taro, and they were full. Their stomachs didn't stretch, the ones who had given a feast to him.
 
\noindent ``Uncle, we are full. There is taro left over.''
 
\noindent ``You can eat more later. Leave the two pots of food there for now.''
 
\noindent They left the two pots of taro, and they ate more later, just as he had said. They ate some taro, and left some.
 
\noindent They went fishing. Maiko went into the forest with his wife. ``Let's go together. I'm going to fish. You watch me.''
 
\noindent So the two of them went. Maiko was standing, fishing. He was standing there, and his wife spoke.
 
\noindent ``Maiko, Maiko.''
 
\noindent ``What?'' he said.
 
\noindent ``Maiko, I need to defecate,'' she said.
 
\noindent ``Go and defecate. I am continuing to fish.''
 
\noindent He speared fish. His wife went to defecate. She came upon the monster. She was standing with the kill of the monster without knowing it. The monster flew away. There was a dead deer on the ground, and she looked at it. She defecated and then came back.
 
\noindent ``Maiko, Maiko, there is a dead animal there on the ground,'' she said.
 
\noindent ``What animal is it?''
 
\noindent ``It's a deer. It's a \textit{bato mawa} deer,'' she said. ``There is a \textit{bato mawa} deer on the ground there.''
 
\noindent ``Wait, I will take a look at it.''
 
\noindent He speared a fish, and hit the fish on the head.
 
\noindent ``I'm going. You put the fish in the basket. Let's go look at the animal.''
 
\noindent They were ready to go look at the animal. They were ready, and they went.
 
\noindent ``The animal is there on the ground,'' she said.
 
\noindent ``Look, the animal is there on the ground.''
 
\noindent He didn't want to just look at the animal. He looked at the kill of the monster.
 
\noindent ``We don't know how the animal died. It will be eaten. I will tie it up.''
 
\noindent He pulled down some vines, and tied up the deer.
 
\noindent ``Maiko, I think it was killed by the thing that flew away,'' she said.\footnote{The monster had taken the form of a bird.}
 
\noindent ``It was killed by it. It was killed by the monster. The deer was killed by it. Go on ahead. I'm going to carry it.''
 
\noindent He hit his feet on the ground in order to lift the deer onto his back, \textit{bao}, \textit{bao} \textit{bao}.\footnote{Maiko was in a sitting position, to stand up with the load, but the monster was holding onto the deer.}
 
\noindent ``It is holding me back,'' he said.
 
\noindent The monster let go. Maiko carried the deer, and he finally came to where his wife was.
 
\noindent ``You carry the fish. Let's go.''
 
\noindent They were ready, and they came back to his village. They cooked the fish back in the village. As they cooked the fish, he devoted himself to preparing the deer. In order to burn the hair off of the deer, he went after firewood, and cut the firewood, and brought the firewood back to the village. He split the firewood back in the village. He lit the firewood, and burned the hair off the deer. Then he cut it up. He made a roasting rack for it, and he roasted it.
 
\noindent They only ate the fish. Maiko's wife ate fish.
 
\noindent He took snuff, and slept. When he was going to sleep, a child began to cry. There was the sound of a child crying, coming toward the village. ``\textit{One}, \textit{one}, \textit{one},'' the child said.
 
\noindent ``Oh no, he didn't put my grandson to sleep early, now he is putting my grandson to sleep late at night,'' he said.
 
\noindent Then his wife spoke. ``Maiko, it is the voice of the monster, whose kill we brought back,'' she said.
 
\noindent ``Yes, it is its voice,'' he said.
 
\noindent It came.
 
\noindent ``I will take a look at it,'' he said.
 
\noindent He took a look at it. He opened the door. The monster had come and was standing on the plaza. It was standing there. Maiko had brought back the monster's kill, and the cooked ribs of the deer were there in the pot.
 
\noindent The monster came crying to the door. It went into Maiko's house.
 
\noindent Maiko didn't sleep. ``I want to sleep. No one ate it. It is there on the rack, I roasted it. The ribs are in the pot there.''
 
\noindent ``I will give it food. I want it to go away.''
 
\noindent He gave food to the monster, cooked deer. He gave it food. The sound of its throat could be heard [as it swallowed].
 
\noindent ``Maiko, there are two monsters.''
 
\noindent ``I guess there are two of them.''
 
\noindent He put a spell on the cooked meat. He gave broth for it to drink, too.
 
\noindent ``OK, there is roasted meat on the rack there, too. I didn't eat the meat.''
 
\noindent Then he gave it the roasted meat. He gave the roasted meat to it, and it ate the roasted deer meat. People talk about it, the monster that kills us. He gave the roasted meat to it, and it ate the roasted meat. It gave food to its mate.
 
\noindent It hadn't eaten. Then it ate. It stayed with him.
 
\noindent ``You two go away. Go away. I am going to sleep. Go away.''
 
\noindent He told it to go away, but it didn't go away. It stayed with him.
 
\noindent Maiko had eaten. He wanted to sleep, and he kept looking at it.
 
\noindent ``Oh no, it wants to stay with me. I want to sleep. It is staying with me.''
 
\noindent ``I'm going to take you there. I will take care of it.''
 
\noindent Maiko took his wife away. He took his wife to the familiar spirits.
 
\noindent ``Don't give any food to my wife.\footnote{If she ate the food of the familiar spirits, she would die.} My wife has already eaten. I'm going to go and take care of the monster that stayed with me.''
 
\noindent He called them to go, but he went instead of them.
 
\noindent ``I'm going to go and take care of the monster that stayed with me.''
 
\noindent He went back, and his wife stayed behind. He came back with the familiar spirits. He came up out of the ground, and he took a look. It was still there.
 
\noindent ``There the monster is, I am telling you. I wanted to sleep, but it stayed with me.''
 
\noindent He took a brand out of the fire, and put a spell on the brand. He threw the brand at it and hit it.
 
\noindent ``You go on ahead back. Don't get in my way.''
 
\noindent After they were gone, he hit it with the brand. \textit{Koro}, \textit{tiihii}, went the sound of it.
 
\noindent The monster set the house on fire. It went away hurt, because Maiko had hit it with the brand. Maiko hit it.
 
\noindent ``It was hit. I wanted to sleep, but it stayed with me. It brought misfortune on itself.''
 
\noindent His familiar spirits came down to the ground. \textit{Baowa}, \textit{barorere}.
 
\noindent ``Oh, over here. Help me.'' His voice could be heard as he called his familiar spirits.
 
\noindent The familiar spirits were all dead as they came down to earth, because of the monster.
 
\noindent ``Female \textit{tokowisa}\footnote{\textit{Tokowisa} is a tree in the forest.} spirit, help me.''
 
\noindent The female \textit{tokowisa} spirit filled her mouth with \textit{kawaki} grass. She came down to earth dead.
 
\noindent ``I called her, but she is not going to help me.''
 
\noindent The female \textit{tokowisa} spirit came down dead. He called others of his familiar spirits. He called his familiar spirits. Another female \textit{tokowisa} spirit came down to earth, and spit \textit{kawaki} grass juice out.
 
\noindent ``Help me, Grandma.''
 
\noindent She spit it out. ``You messed with the monster,'' she said. The female \textit{tokowisa} spirit spit out \textit{kawaki} grass juice.
 
\noindent ``Come on, I'm going to sleep. I'm going to go get my wife.''
 
\noindent He went and picked up his wife. He went after his wife. He brought his wife back.
 
\noindent ``Tie up the hammocks now that you are back. Let's sleep.''
 
\noindent There were not many people living in Maiko's village. His relatives lived somewhere else. 	So he slept finally. He slept, and woke up.
 
\noindent ``I threw something at the monster and hit it. Look, there are the signs of it. Its mate was there on the ground. Its mate fell on the ground. It took its mate away.''
 
\noindent It went away hurt. He looked at the signs of it. He blew on the place. He blew to take away its smell.
 
\noindent Then he got his bow and arrows. ``You stay here. I'm going there to take a look at it. I'm going to spear a fish.''
 
\noindent He went there. He speared fish, and brought fish back. He came back carrying fish.
 
\noindent He killed the monsters. Then he moved away. He lived there a long time, at a place where there were no monsters, after he killed the monsters. Where did he go? He must have gone somewhere else. His descendents live there. No one saw them. They live everywhere.
 
\noindent His old village was there. The foundation poles of his house used to be there. I have said the name of Maiko. The foundation poles of Maiko's house used to be there at Toto Boti.
