\ea\label{ex:13-a1}
Yima, yima mera wati kane, yima, yima mera wati kane, awani awa hoti ka me kaomate mera teo kane awani me awani me me teo kana awani me me warara natiemetemone, awani tosi mati.\\
    \gll yima yima mera wati ka-ne yima yima mera wati ka-ne awani awa hoti ka me ka-homa-tee mera teo ka-ne awani me awani me me teo ka-na awani me me wara-ra nahati-hemete-mone awani tosi mati\\
    Yima(\textsc{m}) Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} plan\_against \textsc{comit}-\textsc{aux}.\textsc{m} Yima(\textsc{m}) Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} plan\_against \textsc{comit}-\textsc{aux}.\textsc{m} wasp(\textsc{m}) tree(\textsc{f}) hole.\textsc{f} \textsc{loc} 3\textsc{pl}.\textsc{s} \textsc{comit}-lie\_on\_ground-\textsc{hab} 3\textsc{pl}.\textsc{o} hit \textsc{comit}-\textsc{aux}.\textsc{m} wasp(\textsc{m}) 3\textsc{pl} wasp(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} hit \textsc{comit}-\textsc{aux}.\textsc{f} wasp(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} grab-\textsc{dup} daytime-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} wasp(\textsc{m}) wasp.sp(\textsc{m}) 3\textsc{pl}\\
    \glt `He was planning to go after the Yimas. He hit with a stick on the buttress roots of a tree where there were wasps in a hole. They were hitting on the tree where the wasps were, and they were grabbing the wasps with their hands, tosi wasps.'
\z

\ea\label{ex:13-a2}
Tosi me me teo kana tosi me me warara natiha me, awani me me tonafiyoha me, awani me me tonafiyoha me, awani haha awani me bara me me tohawawitetasa awani me me mowa tasa me, awani me me nabitasemetemoneni, yima me me mowa nabone mati.\\
    \gll tosi me me teo ka-na tosi me me wara-ra nahatiha me awani me me to-na-fiyoha me awani me me to-na-fiyoha me awani haaha awani me bara me me to-iha-waha-wite-tasa awani me me mowa tasa me awani me me naahabiha-tasa-hemete-mone-ni yima me me mowa na-habone mati\\
    wasp.sp(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} hit \textsc{comit}-\textsc{aux}.\textsc{f} wasp.sp(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} grab-\textsc{dup} daytime.\textsc{f} 3\textsc{pl}.\textsc{s} wasp(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{ch}-\textsc{caus}-be\_all\_gone.\textsc{f} 3\textsc{pl}.\textsc{s} wasp(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{ch}-\textsc{caus}-be\_all\_gone.\textsc{f} 3\textsc{pl}.\textsc{s} wasp(\textsc{m}) this.\textsc{f} wasp(\textsc{m}) 3\textsc{pl} other.\textsc{f} 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} away-take\_in\_turn-change-out-again.\textsc{f} wasp(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} fight again.\textsc{f} 3\textsc{pl}.\textsc{s} wasp(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} kill-again-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} fight \textsc{aux}-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `They hit on the tree where the tosi wasps were. They were grabbing the tosi wasps with their hands. They ran out of wasps. Then they went looking for other wasps, and they fought against those wasps, too, and they killed them. They did this so that they could fight against the Yimas.'
\z

\ea\label{ex:13-a3}
Yima me me mowa nabone awani me me nahabikimaha me, Yima me yima me ababana.\\
    \gll yima me me mowa na-habone awani me me naahabiha-kimaha me yima me yima me ahaba-habana\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} fight \textsc{aux}-\textsc{int}.\textsc{f} wasp(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} kill-two.\textsc{f} 3\textsc{pl}.\textsc{s} Yima(\textsc{m}) 3\textsc{pl} Yima(\textsc{m}) 3\textsc{pl}.\textsc{s} die-\textsc{fut}.\textsc{f}\\
    \glt `They killed wasps on two days, so they could fight against the Yimas. ``The Yimas will die.'''
\z

\ea\label{ex:13-a4}
Me e me e towamakemata.\\
    \gll me ee me ee to-awa-make-mata\\
    3\textsc{pl} 1\textsc{in} 3\textsc{pl}.\textsc{o} 1\textsc{in}.\textsc{s} away-see-following-short\_time\\
    \glt `{``}Let's go see them.'''
\z

\ea\label{ex:13-a5}
Me ye hone, me ye honemetemoneke fa mati.\\
    \gll me yehe hone me yehe hona-hemete-mone-ke faa mati\\
    3\textsc{pl}.\textsc{poss} hand weapon.\textsc{f} 3\textsc{pl}.\textsc{poss} hand have\_something-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} that 3\textsc{pl}\\
    \glt `They had weapons.'
\z

\ea\label{ex:13-a6}
Ya hona me bokori nabisemetemoneni me ye hone.\\
    \gll ya\_hona me bokori na-bisa-hemete-mone-ni me yehe hone\\
    weapon(\textsc{f}) 3\textsc{pl}.\textsc{s} hold \textsc{aux}-also-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} 3\textsc{pl}.\textsc{poss} hand weapon.\textsc{f}\\
    \glt `They had weapons, their weapons.'
\z

\ea\label{ex:13-a7}
Yima me me tonakamakia me, yima me me tonakamakia me, me me hikawemetemoneni, yobe wahi toro ya.\\
    \gll yima me me to-na-ka-makia me yima me me to-na-ka-makia me me me hi-ka-waa-hemete-mone-ni yobe waahi toro ya\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} away-\textsc{caus}-go-following.\textsc{f} 3\textsc{pl}.\textsc{s} Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} away-\textsc{caus}-go-following.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-stand-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} house(\textsc{m}) stand.\textsc{m} inside.\textsc{m} \textsc{adju}\\
    \glt `They went after the Yimas. They went after the Yimas, and waited for them as they were inside the longhouse.'
\z

\ea\label{ex:13-a8}
Yima me me hikawa me, yima watamino kiya kanematamona.\\
    \gll yima me me hi-ka-waa me yima watami-hino kiya ka-na-himata-mona\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-stand.\textsc{f} 3\textsc{pl}.\textsc{o} Yima(\textsc{m}) dream-\textsc{ip}.\textsc{n}.\textsc{m} yell \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `They waited for them. A Yima cried out when he was dreaming.'
\z

\ea\label{ex:13-a9}
Kiya kane, yima watami kiya kane, owa tafihino, fati mowa ne fati kakiyomame me kiyomamemetemoneke.\\
    \gll kiya ka-ne yima watami kiya ka-ne owa tafi-hino fati mowa ne fati ka-ka-yoma-me me ka-yoma-ma-hemete-mone-ke\\
    yell \textsc{comit}-\textsc{aux}.\textsc{m} Yima(\textsc{m}) dream.\textsc{m} yell \textsc{comit}-\textsc{aux}.\textsc{m} other.\textsc{m} buoy\_up-\textsc{ip}.\textsc{n}.\textsc{m} 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) shake \textsc{aux}.\textsc{m} 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) \textsc{comit}-come-through-back.\textsc{m} 3\textsc{pl}.\textsc{s} come-through-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He cried out. The Yima cried out as he was dreaming. Another man woke up. He shook his wife. He brought her outside. They came out.'
\z

\newpage
\ea\label{ex:13-a10}
Me kiyomama me, yama sokimetemoneni.\\
    \gll me ka-yoma-ma me yama soki-mete-mone-ni\\
    3\textsc{pl}.\textsc{s} come-through-back.\textsc{f} 3\textsc{pl}.\textsc{s} thing(\textsc{f}) be\_dark.\textsc{nom}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `They came out. It was dark.'
\z

\ea\label{ex:13-a11}
Homakabote nematamonane yima me me bata hikana mati, yima fana so hinebanoho.\\
    \gll homa-kabote na-himata-mona-ne yima me me bata hi-ka-na mati yima fana soo hi-na-hiba-no-ho\\
    lie\_on\_ground-quickly \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m} Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} surprise \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{o} Yima(\textsc{m}) female(\textsc{f}) urinate\_on \textsc{oc}-\textsc{aux}-\textsc{fut}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `He quickly lay down on the ground, because the Yimas caught them by surprise. The Yima woman urinated on him.'
\z

\ea\label{ex:13-a12}
Yima fana so hine, Ee e ati nene karimematamona yima fana so hinari.\\
    \gll yima fana soo hi-ne ee ee ati na-hene karima-himata-mona yima fana soo hi-na-haari\\
    Yima(\textsc{m}) female(\textsc{f}) urinate\_on \textsc{oc}-\textsc{aux}.\textsc{m} (expression\_of\_surprise) 1\textsc{in}.\textsc{s} say \textsc{aux}-\textsc{irr}.\textsc{f} suffer\_from-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} Yima(\textsc{m}) female(\textsc{f}) urinate\_on \textsc{oc}-\textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `The Yima woman urinated on him. He was suffering, because he felt like speaking, as the Yima woman was urinating on him.'
\z

\ea\label{ex:13-a13}
Ma kanikimemetemone ahi, he so naro.\\
    \gll maa na-kanikima-hemete-mone ahi hee soo na-haaro\\
    stop \textsc{aux}-scattered-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then 3\textsc{sg}.\textsc{o} urinate\_on \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{f}\\
    \glt `She squatted, urinating on him.'
\z

\ea\label{ex:13-a14}
Faya me tokiyomamemetemone.\\
    \gll faya me to-ka-yoma-ma-hemete-mone\\
    so 3\textsc{pl}.\textsc{s} away-go-through-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `Then they went back inside.'
\z

\newpage
\ea\label{ex:13-a15}
Me tokiyomama me, yima me me, yima me me amo tonihamemetemoneni, yama we kamaki ya.\\
    \gll me to-ka-yoma-ma me yima me me yima me me amo to-niha-ma-hemete-mone-ni yama wehe ka-maki ya\\
    3\textsc{pl}.\textsc{s} away-go-through-back.\textsc{f} 3\textsc{pl}.\textsc{s} Yima(\textsc{m}) 3\textsc{pl} 3\textsc{pl} Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} sleep \textsc{ch}-\textsc{caus}-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} thing(\textsc{f}) light.\textsc{f} come-following.\textsc{nom} \textsc{adju}\\
    \glt `They went back inside. They made the Yimas go back to sleep before dawn.'
\z


\ea\label{ex:13-a16}
Tokiyomamake yima me ye hone iso ne, yima me ye hone nawatawamematamona.\\
    \gll to-ka-yoma-make yima me yehe hone iso ne yima me yehe hone na-wata-waha-ma-himata-mona\\
    away-go-through-following Yima(\textsc{m}) 3\textsc{pl}.\textsc{poss} hand weapon.\textsc{f} carry\_off \textsc{aux}.\textsc{m} Yima(\textsc{m}) 3\textsc{pl}.\textsc{poss} hand weapon.\textsc{f} \textsc{caus}-be\_located-change-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `He went inside. He carried away the weapons of the Yimas. He piled up the weapons of the Yimas.'
\z

\ea\label{ex:13-a17}
Yima me ye hone nawatawame, me one haha ne he one me, E tokomibeya me ati nemetemoneni, yima me me wati kana mati.\\
    \gll yima me yehe hone na-wata-waha-me me one ha-haa ne hee one me ee to-ka-ma-be-ya me ati na-hemete-mone-ni yima me me wati ka-na mati\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{poss} hand weapon.\textsc{f} \textsc{caus}-be\_located-change-back.\textsc{m} 3\textsc{pl} other.\textsc{f} \textsc{dup}-call \textsc{aux}.\textsc{m} 3\textsc{sg} other.\textsc{f} 3\textsc{pl} 1\textsc{in}.\textsc{s} away-go-back-\textsc{immed}.\textsc{f}-now 3\textsc{pl}.\textsc{s} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} plan\_against \textsc{comit}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `He piled up the weapons of the Yimas. He called to the others. ``Let's go home,'' they said. They were going to attack the Yimas.'\footnote{The string \textit{he one me} is a false start.}
\z


\newpage
\ea\label{ex:13-a18}
Yima me me wati kana yima me me mowa nabone me... yima me me mowa nabone, ya hona iso ne yima me ye hone nawatawame.\\
    \gll yima me me wati ka-na yima me me mowa na-habone me yima me me mowa na-habone ya\_hona iso ne yima me yehe hone na-wata-waha-me\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} plan\_against \textsc{comit}-\textsc{aux}.\textsc{f} Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} fight \textsc{aux}-\textsc{int}.\textsc{f} 3\textsc{pl} Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} fight \textsc{aux}-\textsc{int}.\textsc{f} weapon(\textsc{f}) carry\_off \textsc{aux}.\textsc{m} Yima(\textsc{m}) 3\textsc{pl}.\textsc{poss} hand weapon.\textsc{f} \textsc{caus}-be\_located-change-back.\textsc{m}\\
    \glt `So that they could fight against the Yimas, he carried away the weapons, and piled them up at another place.'
\z

\ea\label{ex:13-a19}
Iti.\\
    \gll iti\\
    grandfather(\textsc{m})\\
    \glt `Grandpa.'
\z

\ea\label{ex:13-a20}
Haa?\\
    \gll haa\\
    huh?\\
    \glt `What?'
\z

\ea\label{ex:13-a21}
Yima me me mowa nawemetemoneheni.\\
    \gll yima me me mowa na-waha-hemete-mone-he-ni\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} fight \textsc{aux}-change-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}-\textsc{bkg}.\textsc{f}\\
    \glt `Then they fought against the Yimas.'
\z

\ea\label{ex:13-a22}
Yima fati takoreno, itawematamonaka.\\
    \gll yima fati takora-hino ita-waha-himata-mona-ka\\
    Yima(\textsc{m}) 3\textsc{sg}.\textsc{poss}.wife(\textsc{f}) be\_in\_hammock\_with-\textsc{ip}.\textsc{n}.\textsc{m} pierce-change-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He ran a Yima through with a spear as he was in the hammock with his wife.'
\z

\ea\label{ex:13-a23}
Yima tiwa mowa obone.\\
    \gll yima tiwa mowa o-bone\\
    Yima(\textsc{m}) 2\textsc{sg}.\textsc{o} fight 1\textsc{sg}.\textsc{s}-\textsc{int}.\textsc{f}\\
    \glt `{``}Yima, I am going to fight you.'''
\z

\ea\label{ex:13-a24}
Tiwa mowa obonehe.\\
    \gll tiwa mowa o-bone-he\\
    2\textsc{sg}.\textsc{o} fight 1\textsc{sg}.\textsc{s}-\textsc{int}.\textsc{f}-\textsc{dup}\\
    \glt `{``}I am going to fight you.'''
\z

\ea\label{ex:13-a25}
Yima tafimatamona amane ahi.\\
    \gll yima tafi-himata-mona ama-ne ahi\\
    Yima(\textsc{m}) wake\_up-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m} then\\
    \glt `The Yima woke up.'
\z


\ea\label{ex:13-a26}
Yima tiwa mowa obonehe.\\
    \gll yima tiwa mowa o-bone-he\\
    Yima(\textsc{m}) 2\textsc{sg}.\textsc{o} fight 1\textsc{sg}.\textsc{s}-\textsc{int}.\textsc{f}-\textsc{dup}\\
    \glt `{``}Yima, I'm going to fight you.'''
\z

\ea\label{ex:13-a27}
Tiihii, yima kiya kanematamona amane.\\
    \gll tiihii yima kiya ka-na-himata-mona ama-ne\\
    (sound\_of\_hitting) Yima(\textsc{m}) yell \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m}\\
    \glt `\textit{Tiihii}, the Yima cried out.'
\z

\ea\label{ex:13-a28}
Me ye kawemetemone ahi.\\
    \gll me yee ka-waha-hemete-mone ahi\\
    3\textsc{pl}.\textsc{s} yell \textsc{comit}-change-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `They yelled.'
\z

\ea\label{ex:13-a29}
Me hiihii kana nemetemone.\\
    \gll me hiihii ka-na na-hemete-mone\\
    3\textsc{pl}.\textsc{s} yell\_hiihii \textsc{comit}-\textsc{aux} \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `They yelled, ``\textit{Hiihii}.'''
\z

\ea\label{ex:13-a30}
Yima mera katome moni saihimatamona ahi, yima me mowa nari.\\
    \gll yima mera katome moni sai-himata-mona ahi yima me mowa na-haari\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} attack.\textsc{m} sound be\_heard-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} fight \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `The sound of him wounding the Yimas could be heard, as he fought against the Yimas.'
\z

\newpage
\ea\label{ex:13-a31}
Yima me mowa ne moni saimatamona yima mera nabiari.\\
    \gll yima me mowa ne moni sai-himata-mona yima mera naahabiha-haari\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} fight \textsc{aux}.\textsc{m} sound be\_heard-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} kill-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `The sound of him fighting with Yimas could be heard, as he was killing the Yimas.'
\z


\ea\label{ex:13-a32}
Yima mera nabi me yima me me nabiemetemoneke.\\
    \gll yima mera naahabi me yima me me naahabiha-hemete-mone-ke\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} kill.\textsc{m} 3\textsc{pl} Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} kill-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He was killing the Yimas. They were killing the Yimas.'
\z

\ea\label{ex:13-a33}
Yima me me nabia me, me me me me were hinemetemone ahi.\\
    \gll yima me me naahabiha me me me me me were hi-na-hemete-mone ahi\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} kill.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl} 3\textsc{pl} 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} throw \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `They were killing the Yimas. They were throwing them in a pile.'
\z

\ea\label{ex:13-a34}
Yima fana, yima fana, yima fana me hinahabiha me, yima me me nabia me me me ati yana tonemetemonehe.\\
    \gll yima fana yima fana yima fana me hi-naahabiha me yima me me naahabiha me me me ati yana to-na-hemete-mone-he\\
    Yima(\textsc{m}) female(\textsc{f}) Yima(\textsc{m}) female(\textsc{f}) Yima(\textsc{m}) female(\textsc{f}) 3\textsc{pl}.\textsc{s} \textsc{oc}-kill.\textsc{f} 3\textsc{pl}.\textsc{o} Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} kill.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl} 3\textsc{pl}.\textsc{poss} voice start \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `They killed a Yima woman. They killed Yimas. They spoke.'\footnote{Since the meaning was a single Yima woman, the verb should have been \textit{nabia} as in the second clause, and in this case the \textit{me} after the verb would refer to the subject.}
\z

\ea\label{ex:13-a35}
E tokoma.\\
    \gll ee to-ka-ma\\
    1\textsc{in}.\textsc{s} away-go-back\\
    \glt `{``}Let's go home.'''
\z

\ea\label{ex:13-a36}
E tokomahi.\\
    \gll ee to-ka-ma-hi\\
    1\textsc{in}.\textsc{s} away-go-back-\textsc{imp}.\textsc{f}\\
    \glt `{``}Let's go.'''
\z

\ea\label{ex:13-a37}
Yima me one watara me awine.\\
    \gll yima me one wata-ra me awine\\
    Yima(\textsc{m}) 3\textsc{pl} other.\textsc{f} exist-\textsc{neg}.\textsc{f} 3\textsc{pl}.\textsc{s} seem.\textsc{f}\\
    \glt `{``}There aren't any more Yimas.'''
\z

\ea\label{ex:13-a38}
Yama e siba nene ere.\\
    \gll yama ee siba na-hene ere\\
    thing(\textsc{f}) 1\textsc{in}.\textsc{s} look\_for \textsc{aux}-\textsc{irr}.\textsc{f} 1\textsc{in}.\textsc{s}\\
    \glt `{``}Let's not look anymore.'''
\z

\ea\label{ex:13-a39}
Me one winarini?\\
    \gll me one wina-ra-ni\\
    3\textsc{pl} other.\textsc{f} live-\textsc{neg}-\textsc{ip}.\textsc{n}.\textsc{f}\\
    \glt `{``}Aren't there any others living here?'''
\z

\ea\label{ex:13-a40}
Me one winara me awineke.\\
    \gll me one wina-ra me awine-ke\\
    3\textsc{pl} other.\textsc{f} live-\textsc{neg}.\textsc{f} 3\textsc{pl}.\textsc{s} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Apparently there aren't any others living here.'''
\z

\ea\label{ex:13-a41}
E tokomahi.\\
    \gll ee to-ka-ma-hi\\
    1\textsc{in}.\textsc{s} away-go-back-\textsc{imp}.\textsc{f}\\
    \glt `{``}Let's go home.'''
\z

\ea\label{ex:13-a42}
Ee.\\
    \gll ee\\
    yes\\
    \glt `{``}Yes.'''
\z

\ea\label{ex:13-a43}
Faya me kamemetemone fahi.\\
    \gll faya me ka-ma-hemete-mone fahi\\
    so 3\textsc{pl}.\textsc{s} come-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `So they came back.'
\z

\ea\label{ex:13-a44}
Me kama me, me kamehemetemoneni.\\
    \gll me ka-ma me me ka-ma-hemete-mone-ni\\
    3\textsc{pl}.\textsc{s} come-back.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} come-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `They came back.'
\z

\ea\label{ex:13-a45}
Me kama me hiyaramemetemonehe.\\
    \gll me ka-ma me hiyara-ma-hemete-mone-he\\
    3\textsc{pl}.\textsc{s} come-back.\textsc{f} 3\textsc{pl}.\textsc{s} speak-back-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `They came back. They talked back home.'
\z

\ea\label{ex:13-a46}
Yima otara me waka nine, me mowa ni ini hiri ona mati.\\
    \gll yima otara me waka na-ne me mowa ni ini hiri o-na mati\\
    Yima(\textsc{m}) 1\textsc{ex}.\textsc{o} 3\textsc{pl}.\textsc{s} kill \textsc{aux}-\textsc{cont}.\textsc{f} 3\textsc{pl}.\textsc{o} fight \textsc{aux}.\textsc{nfin} name.\textsc{f} say 1\textsc{sg}.\textsc{s}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{poss}\\
    \glt `{``}The Yimas coudln't kill us, because I called the others to fight against them.'''
\z

\ea\label{ex:13-a47}
Me abarake me nafi.\\
    \gll me ahaba-hara-ke me nafi\\
    3\textsc{pl}.\textsc{s} die-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{decl}.\textsc{f} 3\textsc{pl}.\textsc{poss} all\\
    \glt `{``}They all died.'''
\z

\ea\label{ex:13-a48}
Me ota nabiarani, Saba ati nematamona, hiyaramahari.\\
    \gll me otaa naahabiha-hara-ni Saba ati na-himata-mona hiyara-ma-haari\\
    3\textsc{pl}.\textsc{o} 1\textsc{ex}.\textsc{s} kill-\textsc{ip}.\textsc{e}.\textsc{f}-\textsc{bkg}.\textsc{f} Saba(\textsc{m}) say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} speak-back-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `{``}We killed them,'' Saba said back at the village.'
\z

\ea\label{ex:13-a49}
Saba hiyara kaminame me hiyara me ati saihemetemone me ama.\\
    \gll Saba hiyara kamina-me me hiyara me ati sai-hemete-mone me ama\\
    Saba(\textsc{m}) story(\textsc{f}) tell-back.\textsc{m} 3\textsc{pl}.\textsc{s} speak.\textsc{f} 3\textsc{pl}.\textsc{poss} voice be\_heard-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl}.\textsc{poss} \textsc{sec}\\
    \glt `Saba told the news back home. They spoke, and their voices could be heard.'
\z

\newpage
\ea\label{ex:13-a50}
Me hiyara me me ati saia me, me banawemetemoneni.\\
    \gll me hiyara me me ati saiha me me bana-waha-hemete-mone-ni\\
    3\textsc{pl}.\textsc{s} speak.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{poss} voice be\_heard.\textsc{f} 3\textsc{pl}.\textsc{poss} 3\textsc{pl}.\textsc{s} move\_away-change-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `They spoke, and their voices could be heard. Then they moved away to another place.'
\z


\ea\label{ex:13-a51}
Me towinawawitiemetemoneke me ihi itiha mati.\\
    \gll me to-wina-waha-witi-hemete-mone-ke me ihi itiha mati\\
    3\textsc{pl}.\textsc{s} away-live-change-out-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} 3\textsc{pl}.\textsc{poss} victim\_of.\textsc{f} exist.\textsc{f} 3\textsc{pl}.\textsc{poss}\\
    \glt `They went and lived somewhere else, because they had killed people.'
\z

\ea\label{ex:13-a52}
Yima me me nonoki toemetemoneni.\\
    \gll yima me me no-noki to-ha-hemete-mone-ni\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{dup}-wait\_for \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `They waited for the Yimas a long time.'
\z

\ea\label{ex:13-a53}
Me, me mari nabone me, me mari nabone me ebe ha na me me ha hina me, Ee e mari naba ereni.\\
    \gll me me mari na-habone me me mari na-habone me ebe haa na me me haa hi-na me ee ee mari na-haba ere-ni\\
    3\textsc{pl} 3\textsc{pl}.\textsc{s} feast \textsc{aux}-\textsc{int}.\textsc{f} 3\textsc{pl} 3\textsc{pl}.\textsc{s} feast \textsc{aux}-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{poss} \textsc{recip} call \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} call \textsc{oc}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{o} yes 1\textsc{in}.\textsc{s} feast \textsc{aux}-\textsc{fut}.\textsc{f} 1\textsc{in}.\textsc{s}-\textsc{bkg}.\textsc{f}\\
    \glt `They wanted to have a feast, so they invited each other. They invited the others. ``Yes, we will have a feast.'''
\z

\ea\label{ex:13-a54}
Me me ha hina me yana tonemetemone.\\
    \gll me me haa hi-na me yana to-na-hemete-mone\\
    3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} call \textsc{oc}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} start \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `They invited them, and they were ready.'
\z

\ea\label{ex:13-a55}
Te tai toama.\\
    \gll tee tai to-ha-ma\\
    2\textsc{pl}.\textsc{s} be\_ahead \textsc{ch}-\textsc{aux}-back\\
    \glt `{``}You go on ahead.'''
\z

\ea\label{ex:13-a56}
Ota tokomakeminaba ota.\\
    \gll otaa to-ka-make-mina-haba otaa\\
    1\textsc{ex}.\textsc{s} away-go-following-morning-\textsc{fut}.\textsc{f} 1\textsc{ex}.\textsc{s}\\
    \glt `{``}We will follow tomorrow.'''
\z



\ea\label{ex:13-a57}
Faya, me yana tonamina, Abi tiwitibana?\\
    \gll faya me yana to-na-mina abi ti-ita-bana\\
    so 3\textsc{pl}.\textsc{s} start \textsc{ch}-\textsc{aux}-morning.\textsc{f} father(\textsc{m}) 2\textsc{sg}.\textsc{s}-stay.\textsc{f}-\textsc{fut}\\
    \glt `The next day they were ready to leave. ``Father, are you going to stay here?'''
\z

\ea\label{ex:13-a58}
Okomakeba owa awine o.\\
    \gll o-ka-make-haba owa awine o\\
    1\textsc{sg}.\textsc{s}-go-following-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s} seem.\textsc{f} 1\textsc{sg}.\textsc{s}\\
    \glt `{``}I will come later.'''
\z

\ea\label{ex:13-a59}
Oyotoabana o te nowati.\\
    \gll o-yoto-habana o tee nowati\\
    1\textsc{sg}.\textsc{s}-follow-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s} 2\textsc{pl}.\textsc{poss} behind\\
    \glt `{``}I will come after you.'''
\z

\ea\label{ex:13-a60}
Kamakikabote tinahi.\\
    \gll ka-maki-kabote ti-na-hi\\
    come-following-quickly 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{f}\\
    \glt `{``}Come quickly.'''
\z

\ea\label{ex:13-a61}
E totabamakibeya.\\
    \gll ee to-taba-maki-be-ya\\
    1\textsc{in}.\textsc{s} away-be\_together-following-\textsc{immed}.\textsc{f}-now\\
    \glt `{``}Let's go there together.'''
\z

\ea\label{ex:13-a62}
Okomakeba owa awine o.\\
    \gll o-ka-make-haba owa awine o\\
    1\textsc{sg}.\textsc{s}-go-following-\textsc{fut}.\textsc{f} 1\textsc{sg}.\textsc{s} seem.\textsc{f} 1\textsc{sg}.\textsc{s}\\
    \glt `{``}I will go.'''
\z

\ea\label{ex:13-a63}
Yifo tikatisahi.\\
    \gll yifo ti-ka-tisa-hi\\
    hammock(\textsc{m}) 2\textsc{sg}.\textsc{s}-\textsc{comit}-undo-\textsc{imp}.\textsc{f}\\
    \glt `{``}Take down the hammock.'''
\z

\ea\label{ex:13-a64}
Era me hima ni.\\
    \gll era me hima ni\\
    1\textsc{in}.\textsc{o} 3\textsc{pl}.\textsc{s} call\_to\_go \textsc{bkg}.\textsc{f}\\
    \glt `{``}They called us to go.'''
\z

\ea\label{ex:13-a65}
Ota tai toamakiba.\\
    \gll otaa tai to-ha-maki-baha\\
    1\textsc{ex}.\textsc{s} be\_ahead \textsc{ch}-\textsc{aux}-following-before\\
    \glt `{``}We are going on ahead.'''
\z

\ea\label{ex:13-a66}
Ee te tai toamake.\\
    \gll ee tee tai to-ha-make\\
    yes 2\textsc{pl}.\textsc{s} be\_ahead \textsc{ch}-\textsc{aux}-following\\
    \glt `{``}Right, you go on ahead.'''
\z

\ea\label{ex:13-a67}
Me tai toamakemetemone me mari nabone mati.\\
    \gll me tai to-ha-make-hemete-mone me mari na-habone mati\\
    3\textsc{pl}.\textsc{s} be\_ahead \textsc{ch}-\textsc{aux}-following-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl}.\textsc{s} feast \textsc{aux}-\textsc{int}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `They went on ahead, to participate in the feast.'
\z

\ea\label{ex:13-a68}
Me ka abi kamaki, me ka abi kamaki, me ka abi kamakimatamonaka.\\
    \gll me ka abi ka-maki me ka abi ka-maki me ka abi ka-maki-himata-mona-ka\\
    3\textsc{pl} \textsc{poss} father(\textsc{m}) come-following.\textsc{m} 3\textsc{pl} \textsc{poss} father(\textsc{m}) come-following.\textsc{m} 3\textsc{pl} \textsc{poss} father(\textsc{m}) come-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `Their father came afterwards.'
\z

\ea\label{ex:13-a69}
Me ka abi yotohimatamonane.\\
    \gll me ka abi yoto-himata-mona-ne\\
    3\textsc{pl} \textsc{poss} father(\textsc{m}) follow-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `Their father came behind.'
\z


\newpage
\ea\label{ex:13-a70}
Me tai toamake hibati me kobo tonamakemetemoneke faro, me ka abi tabiyo me awabani mati.\\
    \gll me tai to-ha-make hibati me kobo to-na-make-hemete-mone-ke faaro me ka abi tabiyo me awa-haba-ni mati\\
    3\textsc{pl}.\textsc{s} be\_ahead \textsc{ch}-\textsc{aux}-following right\_away 3\textsc{pl}.\textsc{s} arrive away-\textsc{aux}-following-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} that.\textsc{f} 3\textsc{pl} \textsc{poss} father(\textsc{m}) absence 3\textsc{pl}.\textsc{s} feel-\textsc{fut}.\textsc{f}-\textsc{ip}.\textsc{n}.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `They went ahead. Soon they arrived. Then they missed their father.'
\z

\ea\label{ex:13-a71}
Bai yabomako kama namata bati tabiyo.\\
    \gll bahi yabo-mako ka-ma na-mata bati tabiyo\\
    sun(\textsc{m}) be\_far-for\_this\_reason.\textsc{m} come-back.\textsc{cont}.\textsc{m} \textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{fp}.\textsc{n}.\textsc{m} 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) absence\\
    \glt `It was late in the day, so he came back looking for his father.'
\z

\ea\label{ex:13-a72}
Okobi kamakibona kakamaki ra.\\
    \gll okobi ka-maki-hibona ka-ka-maki ra\\
    1\textsc{sg}.\textsc{poss}.father(\textsc{m}) come-following-\textsc{int}.\textsc{m} \textsc{dup}-come-following \textsc{neg}.\textsc{cont}.\textsc{m}\\
    \glt `{``}My father was going to come, but he didn't come.'''
\z

\ea\label{ex:13-a73}
Abi kamakibona ati nano?\\
    \gll abi ka-maki-hibona ati na-no\\
    2\textsc{sg}.\textsc{poss}.father(\textsc{m}) come-following-\textsc{int}.\textsc{m} say \textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}\\
    \glt `{``}Did your father said he would come?'''
\z

\ea\label{ex:13-a74}
Ee, okobi kamakibona kamakira.\\
    \gll ee okobi ka-maki-hibona ka-maki-ra\\
    yes 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) come-following-\textsc{int}.\textsc{m} come-following-\textsc{neg}.\textsc{cont}.\textsc{m}\\
    \glt `{``}Yes, my father was going to come, but he hasn't come.'''
\z

\ea\label{ex:13-a75}
Totowama tinaho.\\
    \gll to-to-awa-ma ti-na-ho\\
    \textsc{dup}-away-see-back 2\textsc{sg}.\textsc{s}-\textsc{aux}-\textsc{imp}.\textsc{m}\\
    \glt `{``}Go back to see him.'''
\z

\ea\label{ex:13-a76}
Okobi totowama omatibe.\\
    \gll okobi to-to-awa-ma o-mata-be\\
    1\textsc{sg}.\textsc{poss}.father(\textsc{m}) \textsc{dup}-away-see-back 1\textsc{sg}.\textsc{s}-short\_time-\textsc{immed}.\textsc{f}\\
    \glt `{``}I will go back to see my father.'''
\z

\ea\label{ex:13-a77}
Fa moni nafike hiba.\\
    \gll faha moni nafi-ke hiba\\
    water(\textsc{f}) sound be\_big-\textsc{decl}.\textsc{f} wait\\
    \glt `The sound of the rain is loud, wait.'
\z

\ea\label{ex:13-a78}
Kamematamonaka, bati nakamari.\\
    \gll ka-ma-himata-mona-ka bati na-ka-ma-haari\\
    come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) \textsc{caus}-come-back-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `He came back after his father.'
\z

\ea\label{ex:13-a79}
Bati nakame kamematamonaka.\\
    \gll bati na-ka-me ka-ma-himata-mona-ka\\
    3\textsc{sg}.\textsc{poss}.father(\textsc{m}) \textsc{caus}-come-back.\textsc{m} come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He came back after his father.'
\z

\ea\label{ex:13-a80}
Me anoti kakosematamonane.\\
    \gll me anoti ka-kana-kosa-himata-mona-ne\\
    3\textsc{pl} 3\textsc{sg}.\textsc{poss}\_older\_brother(\textsc{m}) \textsc{comit}-leave-middle-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `He left his older brother behind.'
\z

\ea\label{ex:13-a81}
Kamematamona, bati yima me wete hinenoho.\\
    \gll ka-ma-himata-mona bati yima me wete hi-na-hino-ho\\
    come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) Yima(\textsc{m}) 3\textsc{pl}.\textsc{s} tie \textsc{oc}-\textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `He came back. The Yimas had tied his father up.'
\z

\ea\label{ex:13-a82}
Bati yima me wete hineno, bati ene wasimematamonaka.\\
    \gll bati yima me wete hi-na-hino bati ehene wasi-ma-himata-mona-ka\\
    3\textsc{sg}.\textsc{poss}.father(\textsc{m}) Yima(\textsc{m}) 3\textsc{pl}.\textsc{s} tie \textsc{oc}-\textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{m} 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) result\_of.\textsc{m} find-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `The Yimas had tied up his father, and he came upon the signs of violence.'
\z

\ea\label{ex:13-a83}
Bati ene wasime Okobi me inamati watani awineke.\\
    \gll bati ehene wasi-me okobi me\_inamati wata-hani awine-ke\\
    3\textsc{sg}.\textsc{poss}.father(\textsc{m}) result\_of.\textsc{m} find-back.\textsc{m} 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) people(\textsc{f}) catch-\textsc{ip}.\textsc{n}.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He came upon the signs of violence. ``Someone has caught my father.'''
\z


\ea\label{ex:13-a84}
Okomi okobi ene nari ha.\\
    \gll okomi okobi ehene naa-haari haa\\
    1\textsc{sg}.\textsc{poss}.mother(\textsc{f}) 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) result\_of.\textsc{m} exist-look.\textsc{m} \textsc{dem}\\
    \glt `{``}There are signs of my father.'''
\z

\ea\label{ex:13-a85}
E namako kamakira nano.\\
    \gll ee na-mako ka-maki-ra na-no\\
    be\_similar \textsc{aux}-for\_this\_reason.\textsc{m} come-following-\textsc{neg}.\textsc{cont}.\textsc{m} \textsc{aux}.\textsc{nom}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}\\
    \glt `{``}So that's why he didn't come.'''
\z

\ea\label{ex:13-a86}
Okobi me inamati, wete nani i ama ha.\\
    \gll o-ka abi me\_inamati wete na-hani ihi ama haa\\
    1\textsc{sg}.\textsc{poss}-\textsc{poss} father(\textsc{m}) people(\textsc{f}) tie \textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{f} result\_of.\textsc{f} be \textsc{dem}\\
    \glt `{``}There are signs of someone having tied up my father.'''
\z

\ea\label{ex:13-a87}
Me namoni okomamatibe.\\
    \gll me namoni o-ka-ma-mata-be\\
    3\textsc{pl}.\textsc{poss} report 1\textsc{sg}.\textsc{s}-go-back-short\_time-\textsc{immed}.\textsc{f}\\
    \glt `{``}I'm going back to tell the others about them.'''
\z

\ea\label{ex:13-a88}
Kana ni tokomematamonaka.\\
    \gll kana ni to-ka-ma-himata-mona-ka\\
    run \textsc{aux}.\textsc{nfin} away-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went running back.'
\z

\ea\label{ex:13-a89}
Kana ni tokome, Ayo okobi e namako okobi kamakirenoho.\\
    \gll kana ni to-ka-me ayo okobi ee na-mako okobi ka-maki-ra-hino-ho\\
    run \textsc{aux}.\textsc{nfin} away-go-back.\textsc{m} older\_brother(\textsc{m}) 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) be\_similar \textsc{aux}-for\_this\_reason.\textsc{m} 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) come-following-\textsc{neg}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `He went running back. ``Brother, this is why my father didn't come.'''
\z

\ea\label{ex:13-a90}
Okobi me inamati hekameinoho.\\
    \gll okobi me\_inamati hi-ka-ka-ma-hino-ho\\
    1\textsc{sg}.\textsc{poss}.father(\textsc{m}) people(\textsc{f}) \textsc{oc}-\textsc{comit}-go-back-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `{``}Someone took my father away.'''
\z

\ea\label{ex:13-a91}
Okobi e towamamatahi ati nematamonane.\\
    \gll okobi ee to-awa-ma-mata-hi ati na-himata-mona-ne\\
    1\textsc{sg}.\textsc{poss}.father(\textsc{m}) 1\textsc{in}.\textsc{s} away-see-back-short\_time-\textsc{imp}.\textsc{f} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `{``}Let's go back and look at my father,'' he said.'
\z

\ea\label{ex:13-a92}
Ee okobi me inamati towakama me awineke.\\
    \gll ee okobi me\_inamati to-ka-ka-ma me awine-ke\\
    yes 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) spirits(\textsc{f}) away-\textsc{comit}-go-back.\textsc{f} 3\textsc{pl}.\textsc{s} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `{``}Right, someone has taken away my father.'''
\z

\ea\label{ex:13-a93}
Me one hima nematamona fara marina me itiha mati.\\
    \gll me one hima na-himata-mona fara marina me itiha mati\\
    3\textsc{pl} other.\textsc{f} call\_to\_go \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} same.\textsc{f} feast(\textsc{f}) 3\textsc{pl}.\textsc{s} begin.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `He called the others to go, who were sponsoring the feast.'
\z

\ea\label{ex:13-a94}
Me me mari namataremetemonehe.\\
    \gll me me mari na-mata-ra-hemete-mone-he\\
    3\textsc{pl} 3\textsc{pl}.\textsc{s} feast \textsc{aux}-short\_time-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `They stopped the feast.'
\z

\ea\label{ex:13-a95}
Me kama me, me kama me ihi me wasima me, yama me siba nemetemone.\\
    \gll me ka-ma me me ka-ma me ihi me wasi-ma me yama me siba na-hemete-mone\\
    3\textsc{pl}.\textsc{s} come-back.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{pl}.\textsc{s} come-back.\textsc{f} 3\textsc{pl}.\textsc{poss} result\_of.\textsc{f} 3\textsc{pl}.\textsc{s} find-back.\textsc{f} 3\textsc{pl}.\textsc{poss} thing(\textsc{f}) 3\textsc{pl}.\textsc{s} look\_for \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `They came back. They came upon the signs of them. They looked around.'
\z

\ea\label{ex:13-a96}
Yama me siba na me, Okobi me inamati hekameno awaka.\\
    \gll yama me siba na me okobi me\_inamati hi-ka-ka-ma-hino awa-ka\\
    thing(\textsc{f}) 3\textsc{pl}.\textsc{s} look\_for \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) people(\textsc{f}) \textsc{oc}-\textsc{comit}-go-back-\textsc{ip}.\textsc{n}.\textsc{m} seem.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `They looked around. ``Someone took away my father.'''
\z

\ea\label{ex:13-a97}
Me hawi tokomaro ha.\\
    \gll me hawi to-ka-ma-haaro haa\\
    3\textsc{pl}.\textsc{poss} trail.\textsc{f} away-go-back-look.\textsc{f} \textsc{dem}\\
    \glt `{``}There is their trail going that way.'''
\z

\ea\label{ex:13-a98}
Me e towamakemata.\\
    \gll me ee to-awa-make-mata\\
    3\textsc{pl}.\textsc{o} 1\textsc{in}.\textsc{s} away-see-following-short\_time\\
    \glt `{``}Let's go look at them.'''
\z

\ea\label{ex:13-a99}
Tokomakematamonaka.\\
    \gll to-ka-make-himata-mona-ka\\
    away-go-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He went.'
\z

\ea\label{ex:13-a100}
Tokomake me tokomakemetemone.\\
    \gll to-ka-make me to-ka-make-hemete-mone\\
    away-go-following 3\textsc{pl}.\textsc{s} away-go-following-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}\\
    \glt `He went. They went.'
\z

\ea\label{ex:13-a101}
Wati hinama me ye hone toehemetemone ama ahi.\\
    \gll wati hinama me yehe hone to-ha-hemete-mone ama ahi\\
    arrow(\textsc{m}) all 3\textsc{pl}.\textsc{poss} hand weapon.\textsc{f} \textsc{ch}-be-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec} then\\
    \glt `Their weapons were all arrows.'
\z

\ea\label{ex:13-a102}
Tokomakematamona ahi.\\
    \gll to-ka-make-himata-mona ahi\\
    away-go-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `He went there.'
\z

\ea\label{ex:13-a103}
Tokomake, me ihi me hikiyoemetemoneni, me hawi.\\
    \gll to-ka-make me ihi me hi-kiyo-hemete-mone-ni me hawi\\
    away-go-following 3\textsc{pl}.\textsc{poss} result\_of.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{oc}-chase-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} 3\textsc{pl}.\textsc{poss} trail.\textsc{f}\\
    \glt `He went. They followed their trail.'
\z

\ea\label{ex:13-a104}
Me hawi me hikiyoa me me me hinakamakiemetemoneni.\\
    \gll me hawi me hi-kiyoha me me me hi-na-ka-maki-hemete-mone-ni\\
    3\textsc{pl}.\textsc{poss} trail.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{oc}-follow.\textsc{f} 3\textsc{pl}.\textsc{poss} 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{caus}-go-following-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `They followed their trail. They went after them.'
\z

\ea\label{ex:13-a105}
Me me hinakamakia me, me ati me himitamakiemetemoneni me, me hiyara mati.\\
    \gll me me hi-na-ka-makia me me ati me hi-mita-maki-hemete-mone-ni me me hiyara mati\\
    3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{caus}-go-following.\textsc{f} 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{poss} voice 3\textsc{pl}.\textsc{s} \textsc{oc}-hear-following-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} 3\textsc{pl} 3\textsc{pl}.\textsc{s} speak.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `They went after them. They heard their voices as they were talking.'
\z

\ea\label{ex:13-a106}
Faya me ma tonamakemetemoneke fahi.\\
    \gll faya me maa to-na-make-hemete-mone-ke fahi\\
    so 3\textsc{pl}.\textsc{s} stop away-\textsc{aux}-following-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} then\\
    \glt `Then they stopped.'
\z

\ea\label{ex:13-a107}
Faya yima kamematamona amaka me nao me tonomakia me nokosi ya.\\
    \gll faya yima ka-ma-himata-mona ama-ka me naho me to-naho-makia me nokosi ya\\
    so Yima(\textsc{m}) come-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} \textsc{sec}-\textsc{decl}.\textsc{m} 3\textsc{pl}.\textsc{s} stand 3\textsc{pl}.\textsc{s} away-stand-following.\textsc{f} 3\textsc{pl}.\textsc{poss} in\_front\_of \textsc{adju}\\
    \glt `A Yima man came right facing where they were standing.'\footnote{The verb \textit{me nao} is a false start.}
\z

\ea\label{ex:13-a108}
Kifewame, kifewamematamonane.\\
    \gll ka-fe-waha-me ka-fe-waha-ma-himata-mona-ne\\
    come-water-change-back.\textsc{m} come-water-change-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `He came to the water.'
\z

\ea\label{ex:13-a109}
Kifewame, kanatafo katise tokifiwawiti, iyawa itimatamonaha.\\
    \gll ka-fe-waha-me kanatafo ka-tise to-ka-fi-waha-witi iyawa iti-himata-mona-ha\\
    come-water-change-back.\textsc{m} breechcloth(\textsc{f}) \textsc{comit}-undo.\textsc{m} away-go-water-change-out.\textsc{m} grated\_mash(\textsc{f}) take\_out-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `He came to the water. He took off his breechcloth. He went to the water, and took out the manioc mash.'
\z

\ea\label{ex:13-a110}
Iyawa iti, iyawa tiwa ni kimisame iyawa towakematamona ahi.\\
    \gll iyawa iti iyawa tiwa ni ka-misa-me iyawa to-ka-ka-himata-mona ahi\\
    grated\_mash(\textsc{f}) take\_out.\textsc{m} grated\_mash(\textsc{f}) carry \textsc{aux}.\textsc{nfin} come-up-back.\textsc{m} grated\_mash(\textsc{f}) away-\textsc{comit}-go-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `He took out the manioc mash. He came to the bank carrying the manioc mash. He took the manioc mash away.'
\z

\ea\label{ex:13-a111}
Iyawa tonawawitematamonaka.\\
    \gll iyawa to-na-waa-wite-himata-mona-ka\\
    grated\_mash(\textsc{f}) away-\textsc{caus}-stand-out-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m}\\
    \glt `He set the basket of manioc mash down.'
\z

\largerpage
\ea\label{ex:13-a112}
Iyawa tonawawite, iyawa hokowabone ra tonawawitematamonane, hehe me hekame kakawibonaha.\\
    \gll iyawa to-na-waa-wite iyawa hoko-waha-habone ra to-na-waa-wite-himata-mona-ne hehe me hi-ka-ka-me kakawi-bona-ha\\
    grated\_mash(\textsc{f}) away-\textsc{caus}-stand-out grated\_mash(\textsc{f}) be\_dry-change-\textsc{int}.\textsc{f} \textsc{o} away-\textsc{caus}-stand-out-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m} this\_one.\textsc{m} 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-go-back.\textsc{m} starch-\textsc{int}.\textsc{m}-\textsc{dup}\\
    \glt `He set the manioc mash down. He set the manioc mash down so that the mash would dry out, so it would be eaten with the one that they had brought.'
\z

\ea\label{ex:13-a113}
Faya, faya afi ne, afi ne hawa toe tokome nowati ya me tokomakemetemoneke.\\
    \gll faya faya afi ne afi ne hawa to-he to-ka-me nowati ya me to-ka-make-hemete-mone-ke\\
    so so bathe \textsc{aux}.\textsc{m} bathe \textsc{aux}.\textsc{m} be\_finished \textsc{ch}-\textsc{aux}.\textsc{m} away-go-back.\textsc{m} behind \textsc{adju} 3\textsc{pl}.\textsc{s} away-go-following-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `He bathed. He finished bathing. He went, and they went behind him.'
\z

\ea\label{ex:13-a114}
Tokomakematamonaha.\\
    \gll to-ka-make-himata-mona-ha\\
    away-go-following-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `He followed.'
\z

\ea\label{ex:13-a115}
Tokomake kori kasematamona.\\
    \gll to-ka-make kori ka-kosa-himata-mona\\
    away-go-following be\_afraid \textsc{comit}-middle-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `He followed. He wasn't afraid.'
\z

\ea\label{ex:13-a116}
Tokomake, me one tokobasawawitemetemone ai yobe bari ya.\\
    \gll to-ka-make me one to-ka-basa-waha-wite-hemete-mone ahi yobe bari ya\\
    away-go-following 3\textsc{pl} other.\textsc{f} away-go-edge-change-out-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} there house(\textsc{m}) outer\_surface \textsc{adju}\\
    \glt `He followed. The others surrounded the longhouse.'
\z

\ea\label{ex:13-a117}
Me towakiyomamakiemetemoneni.\\
    \gll me to-ka-ka-yoma-maki-hemete-mone-ni\\
    3\textsc{pl}.\textsc{s} away-\textsc{comit}-go-through-following-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `Two of them went in.'
\z

\ea\label{ex:13-a118}
Inamate ohi nematamona amane nisorinoho.\\
    \gll inamate ohi na-himata-mona ama-ne nisori-no-ho\\
    child(\textsc{m}) cry \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} \textsc{sec}-\textsc{bkg}.\textsc{m} 3\textsc{sg}.\textsc{poss}.younger\_brother(\textsc{m})-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup}\\
    \glt `A child cried. It was his younger brother.'
\z

\ea\label{ex:13-a119}
Okobi e raya? ati nematamona.\\
    \gll okobi ee ra-ya ati na-himata-mona\\
    1\textsc{sg}.\textsc{poss}.father(\textsc{m}) (question) \textsc{cq}.\textsc{m}-now say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}\\
    \glt `{``}Where is my father?'' he said.'
\z



\ea\label{ex:13-a120}
Okobi okobi te kakameno onakamakine o.\\
    \gll okobi okobi tee ka-ka-ma-hino o-na-ka-maki-ne o\\
    1\textsc{sg}.\textsc{poss}.father(\textsc{m}) 1\textsc{sg}.\textsc{poss}.father(\textsc{m}) 2\textsc{pl}.\textsc{s} \textsc{comit}-come-back-\textsc{ip}.\textsc{n}.\textsc{m} 1\textsc{sg}.\textsc{s}-\textsc{caus}-come-following-\textsc{cont}.\textsc{f} 1\textsc{sg}.\textsc{s}\\
    \glt `{``}I have come to get my father, who you brought here.'''
\z

\ea\label{ex:13-a121}
Abi homa ha.\\
    \gll abi homa haa\\
    2\textsc{sg}.\textsc{poss}.father(\textsc{m}) lie\_on\_ground.\textsc{cont}.\textsc{m} \textsc{dem}\\
    \glt `{``}Your father is lying there on the ground.'''
\z

\ea\label{ex:13-a122}
Abi abi nahore.\\
    \gll abi abi naho-re\\
    2\textsc{sg}.\textsc{poss}.father(\textsc{m}) 2\textsc{sg}.\textsc{poss}.father(\textsc{m}) be\_touched-\textsc{neg}.\textsc{m}\\
    \glt `{``}Your father has not been touched.'''
\z

\ea\label{ex:13-a123}
Abi homa hari me ati nemetemone yima mati.\\
    \gll abi homa haari me ati na-hemete-mone yima mati\\
    2\textsc{sg}.\textsc{poss}.father(\textsc{m}) lie\_on\_ground.\textsc{cont}.\textsc{m} that\_one.\textsc{m} 3\textsc{pl}.\textsc{s} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} Yima(\textsc{m}) 3\textsc{pl}\\
    \glt `{``}Your father is lying there,'' they said, the Yimas.'
\z

\ea\label{ex:13-a124}
Okobi te tisa.\\
    \gll okobi tee tisa\\
    1\textsc{sg}.\textsc{poss}.father(\textsc{m}) 2\textsc{pl}.\textsc{s} undo\\
    \glt `{``}Unbind my father.'''
\z

\ea\label{ex:13-a125}
Okobi owakamibeya.\\
    \gll okobi o-ka-ka-ma-be-ya\\
    1\textsc{sg}.\textsc{poss}.father(\textsc{m}) 1\textsc{sg}.\textsc{s}-\textsc{comit}-go-back-\textsc{immed}.\textsc{f}-now\\
    \glt `{``}I'm going to take my father home.'''
\z

\ea\label{ex:13-a126}
Faya me, yima me hitisematamonane.\\
    \gll faya me yima me hi-tisa-himata-mona-ne\\
    so 3\textsc{pl} Yima(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{oc}-undo-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `So the Yimas unbound him.'
\z

\ea\label{ex:13-a127}
Me hitise, inamatewe me tisa nemetemone ahi.\\
    \gll me hi-tise inamatewe me tisa na-hemete-mone ahi\\
    3\textsc{pl}.\textsc{s} \textsc{oc}-undo.\textsc{m} child(\textsc{m}) 3\textsc{pl}.\textsc{s} undo.\textsc{list} \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `They unbound him. They also unbound the child.'
\z

\ea\label{ex:13-a128}
Mati me hitisa hinemetemoneni me wete hinanihi.\\
    \gll mati me hi-tisa hi-na-hemete-mone-ni me wete hi-na-hani-hi\\
    3\textsc{sg}.\textsc{poss}.mother(\textsc{f}) 3\textsc{pl}.\textsc{s} \textsc{oc}-undo.\textsc{list} \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} 3\textsc{pl}.\textsc{s} tie \textsc{oc}-\textsc{aux}-\textsc{ip}.\textsc{n}.\textsc{f}-\textsc{dup}\\
    \glt `They also unbound his mother, who they had bound.'
\z

\ea\label{ex:13-a129}
Mera hihiyara kawematamonaka ahi.\\
    \gll mera hi-hiyara ka-waha-himata-mona-ka ahi\\
    3\textsc{pl}.\textsc{o} \textsc{oc}-speak\_to \textsc{comit}-change-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{decl}.\textsc{m} then\\
    \glt `He continued to speak to them.'
\z

\ea\label{ex:13-a130}
Yima ati me hikamitematamona me naoa mati.\\
    \gll yima ati me hi-ka-mita-himata-mona me nahoha mati\\
    Yima(\textsc{m}) voice 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-hear-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} 3\textsc{pl}.\textsc{s} stand.\textsc{f} 3\textsc{pl}.\textsc{s}\\
    \glt `The Yimas listened to him as they stood there.'
\z

\ea\label{ex:13-a131}
Me naoa ati me hikamite ati yana tonematamonane.\\
    \gll me nahoha ati me hi-ka-mite ati yana to-na-himata-mona-ne\\
    3\textsc{pl}.\textsc{s} stand.\textsc{f} voice 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-hear.\textsc{m} voice start \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `They stood listening to him. He spoke.'
\z

\ea\label{ex:13-a132}
Okobi te towakama.\\
    \gll okobi tee to-ka-ka-ma\\
    1\textsc{sg}.\textsc{poss}.father(\textsc{m}) 2\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back\\
    \glt `{``}Take my father away.'''
\z

\ea\label{ex:13-a133}
Okobi te towakama.\\
    \gll okobi tee to-ka-ka-ma\\
    1\textsc{sg}.\textsc{poss}.father(\textsc{m}) 2\textsc{pl}.\textsc{s} away-\textsc{comit}-go-back\\
    \glt `{``}Take my father away.'''
\z

\ea\label{ex:13-a134}
Me ati omiti forimamatibeya.\\
    \gll me ati o-miti forima-mata-be-ya\\
    3\textsc{pl}.\textsc{poss} voice 1\textsc{sg}.\textsc{s}-hear.\textsc{nfin} do\_properly-short\_time-\textsc{immed}.\textsc{f}-now\\
    \glt `{``}I'm going to finish listening to them.'''
\z

\ea\label{ex:13-a135}
Me ati mitehibona, bati iseme bati me hekamematamonane.\\
    \gll me ati mita-hibona bati ise-me bati me hi-ka-ka-ma-himata-mona-ne\\
    3\textsc{pl}.\textsc{poss} voice hear-\textsc{int}.\textsc{m} 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) drop\_off-back.\textsc{m} 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{bkg}.\textsc{m}\\
    \glt `He wanted to listen to them, and he handed off his father. They took his father away.'
\z

\ea\label{ex:13-a136}
Bati me hekame, bati me hekamematamonaha.\\
    \gll bati me hi-ka-ka-me bati me hi-ka-ka-ma-himata-mona-ha\\
    3\textsc{sg}.\textsc{poss}.father(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-go-back.\textsc{m} 3\textsc{sg}.\textsc{poss}.father(\textsc{m}) 3\textsc{pl}.\textsc{s} \textsc{oc}-\textsc{comit}-go-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `They took his father away.'
\z

\ea\label{ex:13-a137}
Faya tokiyomamatasematamonaha.\\
    \gll faya to-ka-yoma-ma-tasa-himata-mona-ha\\
    so away-go-through-back-again-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m}-\textsc{dup}\\
    \glt `Then he went back inside.'
\z

\ea\label{ex:13-a138}
Wabori towakose, tokiyomamematamona ahi.\\
    \gll wabori to-waa-kose to-ka-yoma-ma-himata-mona ahi\\
    3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) \textsc{ch}-stand-middle.\textsc{m} away-go-through-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `His brother-in-law stayed outside, while he went inside.'
\z

\ea\label{ex:13-a139}
Te winibana?\\
    \gll tee wini-bana\\
    2\textsc{pl}.\textsc{s} live.\textsc{f}-\textsc{fut}\\
    \glt `{``}Are you going to stay here?'''
\z

\ea\label{ex:13-a140}
Ota mari nabone ota ama.\\
    \gll otaa mari na-habone otaa ama\\
    1\textsc{ex}.\textsc{s} feast \textsc{aux}-\textsc{int}.\textsc{f} 1\textsc{ex}.\textsc{s} \textsc{sec}\\
    \glt `{``}We are going to have a feast.'''
\z

\ea\label{ex:13-a141}
Ee ee yima me ati nemetemone me amani ahi.\\
    \gll ee ee yima me ati na-hemete-mone me ama-ni ahi\\
    yes yes Yima(\textsc{m}) 3\textsc{pl}.\textsc{s} say \textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{sec}-\textsc{bkg}.\textsc{f} then\\
    \glt `{``}Right,'' the Yimas said.'
\z

\ea\label{ex:13-a142}
Ota ota winineni.\\
    \gll otaa otaa wina-ne-ni\\
    1\textsc{ex} 1\textsc{ex}.\textsc{s} live-\textsc{cont}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `{``}We are living here.'''
\z

\ea\label{ex:13-a143}
Te winiyahi.\\
    \gll tee wina-yahi\\
    2\textsc{pl}.\textsc{s} live-\textsc{dist}.\textsc{imp}.\textsc{f}\\
    \glt `{``}Live here.'''
\z

\ea\label{ex:13-a144}
Faya wati soba nakematamona fahi.\\
    \gll faya wati soba na-ke-himata-mona fahi\\
    so arrow(\textsc{m}) pull\_out \textsc{aux}-coming-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `Then he took an arrow out of the sheath.'
\z

\ea\label{ex:13-a145}
Wati soba nake, wati yori tomisematamona neme ka ya.\\
    \gll wati soba na-ke wati yori to-misa-himata-mona neme ka ya\\
    arrow(\textsc{m}) pull\_out \textsc{aux}-coming arrow(\textsc{m}) insert away-up-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} up(\textsc{f}) \textsc{loc} \textsc{adju}\\
    \glt `He pulled out an arrow, and inserted the arrow pointing up.'
\z

\ea\label{ex:13-a146}
Hoka namematamona ahi.\\
    \gll hoka na-ma-himata-mona ahi\\
    yank \textsc{aux}-back-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `He pulled back on the arrow.'
\z

\newpage
\ea\label{ex:13-a147}
Te ka yobe bako sa okani toe o, yima bako sa kanebanoho wati weo nisahari.\\
    \gll tee ka yobe bako saa o-ka-na to-ha o yima bako saa ka-na-hiba-no-ho wati weho na-risa-haari\\
    2\textsc{pl} \textsc{poss} house(\textsc{m}) inner\_surface.\textsc{m} shoot\_with\_arrow 1\textsc{sg}.\textsc{s}-\textsc{comit}-\textsc{aux} \textsc{ch}-\textsc{aux} 1\textsc{sg}.\textsc{s} Yima(\textsc{m}) chest.\textsc{m} shoot\_with\_arrow \textsc{comit}-\textsc{aux}-\textsc{fut}.\textsc{m}-\textsc{ip}.\textsc{n}.\textsc{m}-\textsc{dup} arrow(\textsc{m}) drop\_down \textsc{aux}-down-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `{``}I want to shoot an arrow at the ceiling of your longhouse.'' But he lowered the arrow and shot a Yima in the chest.'
\z

\ea\label{ex:13-a148}
Ayata sa hikane rama kiya kanematamona ahi.\\
    \gll ayata saa hi-ka-ne rama kiya ka-na-himata-mona ahi\\
    near(\textsc{f}) shoot\_with\_arrow \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{m} unusual yell \textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then\\
    \glt `He shot him from very close by, and he cried out.'
\z

\ea\label{ex:13-a149}
Me ye kawemetemone ahi.\\
    \gll me yee ka-waha-hemete-mone ahi\\
    3\textsc{pl}.\textsc{s} yell \textsc{comit}-change-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `They yelled.'
\z

\ea\label{ex:13-a150}
Yima fana, yima fana kana ni yana torawemetemone amani.\\
    \gll yima fana yima fana kana ni yana to-rawa-hemete-mone ama-ni\\
    Yima(\textsc{m}) female(\textsc{f}) Yima(\textsc{m}) female(\textsc{f}) run \textsc{aux}.\textsc{nfin} start \textsc{ch}-\textsc{fpl}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} \textsc{sec}-\textsc{bkg}.\textsc{f}\\
    \glt `The Yima women ran away.'
\z

\ea\label{ex:13-a151}
Yima fana tokiyomarawemetemone ahi.\\
    \gll yima fana to-ka-yoma-rawa-hemete-mone ahi\\
    Yima(\textsc{m}) female(\textsc{f}) away-go-through-\textsc{fpl}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `The Yima women went out.'
\z

\newpage
\ea\label{ex:13-a152}
Me, mera tisa ne ati saihimatamona me me mowa na mati wabori famahari.\\
    \gll me mera tisa ne ati sai-himata-mona me me mowa na mati wabori fama-haari\\
    3\textsc{pl} 3\textsc{pl}.\textsc{o} shoot\_with\_arrow \textsc{aux}.\textsc{m} voice be\_heard-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} fight \textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{s} 3\textsc{sg}.\textsc{poss}.brother\_in\_law(\textsc{m}) be\_with-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `He was yelling as he was shooting them with arrows, as he and his brother-in-law were attacking them.'
\z



\ea\label{ex:13-a153}
Yima me fito tonemetemone me ama ahi.\\
    \gll yima me fito to-na-hemete-mone me ama ahi\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{s} rush \textsc{ch}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} 3\textsc{pl}.\textsc{s} \textsc{sec} then\\
    \glt `The Yimas were rushing out.'
\z

\ea\label{ex:13-a154}
Me me tisa hinemetemone ahi yima fanarawaha.\\
    \gll me me tisa hi-na-hemete-mone ahi yima fana-rawa-ha\\
    3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} shoot\_with\_arrow \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then Yima(\textsc{m}) female(\textsc{f})-\textsc{fpl}-\textsc{dup}\\
    \glt `The others were shooting them with arrows, also the Yima women.'
\z

\ea\label{ex:13-a155}
Me me tisa hinemetemoneni.\\
    \gll me me tisa hi-na-hemete-mone-ni\\
    3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} shoot\_with\_arrow \textsc{oc}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f}\\
    \glt `They were shooting them.'
\z

\ea\label{ex:13-a156}
Me me tisa hina me, ye tonakosematamona ahi, yima mera mowa nari.\\
    \gll me me tisa hi-na me yee to-na-kosa-himata-mona ahi yima mera mowa na-haari\\
    3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} shoot\_with\_arrow \textsc{oc}-\textsc{aux}.\textsc{f} 3\textsc{pl}.\textsc{o} yell \textsc{ch}-\textsc{aux}-middle-\textsc{fp}.\textsc{n}.\textsc{m}-\textsc{rep}.\textsc{m} then Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} fight \textsc{aux}-\textsc{ip}.\textsc{e}.\textsc{m}\\
    \glt `They were shooting them. He yelled out as he attacked them.'
\z

\ea\label{ex:13-a157}
Yima me mowa ne yima, yima me me kiyoemetemone ahi.\\
    \gll yima me mowa ne yima yima me me kiyo-hemete-mone ahi\\
    Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} fight \textsc{aux}.\textsc{m} Yima(\textsc{m}) Yima(\textsc{m}) 3\textsc{pl}.\textsc{o} 3\textsc{pl}.\textsc{s} chase-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} then\\
    \glt `He was attacking the Yimas. They were chasing the Yimas.'
\z

\ea\label{ex:13-a158}
Yima fana naba koro, kana ni tokifiwawitia bokemetemoneke faro.\\
    \gll yima fana naba\_koro kana ni to-ka-fi-waha-witia boka-hemete-mone-ke faaro\\
    Yima(\textsc{m}) female(\textsc{f}) pregnant run \textsc{aux}.\textsc{nfin} away-go-water-change-out.\textsc{f} enter\_water-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} that.\textsc{f}\\
    \glt `A pregnant Yima woman went running to the water, and went in the water.'
\z

\ea\label{ex:13-a159}
One me sa hikanemetemoneni tame baki.\\
    \gll one me saa hi-ka-na-hemete-mone-ni tame baki\\
    one.\textsc{f} 3\textsc{pl}.\textsc{s} shoot\_with\_arrow \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{bkg}.\textsc{f} foot.\textsc{f} inside\_surface.\textsc{f}\\
    \glt `They shot one of the women in the sole of her foot.'
\z

\ea\label{ex:13-a160}
Tame baki me sa hikanemetemonehe.\\
    \gll tame baki me saa hi-ka-na-hemete-mone-he\\
    foot.\textsc{f} inside\_surface.\textsc{f} 3\textsc{pl}.\textsc{s} shoot\_with\_arrow \textsc{oc}-\textsc{comit}-\textsc{aux}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{dup}\\
    \glt `They shot her in the sole of her foot.'
\z

\ea\label{ex:13-a161}
Tame baki me sa hikana, abemetemoneke faro.\\
    \gll tame baki me saa hi-ka-na ahaba-hemete-mone-ke faaro\\
    foot.\textsc{f} inside\_surface.\textsc{f} 3\textsc{pl}.\textsc{s} shoot\_with\_arrow \textsc{oc}-\textsc{comit}-\textsc{aux}.\textsc{f} die-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f} that.\textsc{f}\\
    \glt `They shot her in the sole of her foot, and she died.'
\z

\ea\label{ex:13-a162}
One ta tafiremetemoneke.\\
    \gll one taa tafi-ra-hemete-mone-ke\\
    other.\textsc{f} contrast buoy\_up-\textsc{neg}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `The other woman didn't come to the surface.'
\z

\ea\label{ex:13-a163}
Fe bokimetemone ka fa itamaro ita awineke.\\
    \gll fehe boki-mete-mone ka faha ita-hamaro ita awine-ke\\
    water.\textsc{f} enter\_water.\textsc{nom}.\textsc{f}-\textsc{fp}.\textsc{n}.\textsc{f}-\textsc{rep}.\textsc{f} which water(\textsc{f}) be\_located-\textsc{fp}.\textsc{e}.\textsc{f} be\_located.\textsc{f} seem.\textsc{f}-\textsc{decl}.\textsc{f}\\
    \glt `Long ago I saw the lake where she went into the water. It must still be there.'
\z
