\noindent That's what they were singing about, \textit{kona} poison roots. It was \textit{kona}, a big pile. His name has gone away quickly from my memory. Don't you know, Boroko? The one I used to talk about, who sang about \textit{kona}. Oh, Aifa, yes, Aifa. They dug up \textit{kona} roots. They started into the forest to the place where they would beat the \textit{kona}. My father was small. He was very small. Aifa was thinking about the \textit{kona}. They were thinking about the \textit{kona}. ``You beat \textit{kona}. You go on ahead. You beat \textit{kona},'' Aifa said. ``You go on ahead. You put the poles for beating \textit{kona} on the ground. I'll come later. I'm going to take snuff now.'' He took snuff. He went into the forest to defecate. He went and defecated. He came back.

\noindent ``We are going to beat \textit{kona} for a while,'' he said [to his wife, in the village]. He came to the others. He had painted his face red with annatto. He came.

\noindent ``Well? Have you begun the \textit{kona} song?'' he said.

\noindent ``We haven't begun the song yet.''

\noindent ``They are taking a long time. Sit in two rows.'' They sat in two rows on the poles, facing each other. One group sat on one side, and the other group sat on the other side, and he started the song.

\noindent He sat down on one of the poles. He grabbed a baton, to beat the \textit{kona} with. The baton went, \textit{teioo}, as he beat the \textit{kona}. He started the E Onira song, the \textit{kona} song. ``My younger sister doesn't know how to take out fish gall bladder. My younger sister doesn't know how to take out fish gall bladder,'' he said, singing the \textit{kona} song. They sang the E Onira song. The sound of them beating the \textit{kona} went, \textit{baro}, \textit{baro}. They were in lines beating the \textit{kona}, my father and the others used to say. The E Onira song could be heard. They kept bending their heads down as they beat the \textit{kona}.

\noindent He came to start the E Onira song, and all of a sudden he started the E Onira song. ``They never sang this song a long time ago,'' he said. ``Our adolescent boys are beating to the song that was never sung a long time ago. They will learn it.''

\noindent \textit{Baro}, \textit{baro}. It was the \textit{kona} song. They beat the \textit{kona}.

\noindent As they beat the \textit{kona}, he sang the E Onira song, the \textit{kona} song. He sang about it, the \textit{bahama} catfish. ``When the \textit{bahama} fish hear the enemy Indians [i.e. the Jarawaras], they `talk' to warn each other,'' he sang. They beat the \textit{kona} really fast, by the E Onira song. They beat the \textit{kona} by the E Onira song. ``The \textit{bahama} fish has a white moustache,'' he sang. He composed the E Onira song, my father said. This is what he was like: he had painted his face red, and his knees went up over his head as he sat, and he was between two others. They kept moving their heads up and down as they beat the \textit{kona}. The \textit{bahama} song didn't exist before. That \textit{kona} song didn't exist before.

\noindent He started another right after that one. ``A piece of a \textit{tawi} fish goes around and around in the water as it hits the enemies' fish weir,'' he sang. [The \textit{tawi} fish was cut in half by piranhas.] ``E Onira, E Onira,'' he sang. They walked around to the song, wearing toucan feather ornaments.

\noindent They sang, singing the \textit{kona} song. ``The \textit{wasa} fish blows to keep the fever of the enemy Indians away,'' he said, singing the \textit{kona} song, which they had composed and taught. I am telling what I heard my father say long ago.

\noindent He started another right after that one. ``The \textit{base} fish turns its anus up when it hits the enemy Indians' fish weir,'' he sang.

\noindent They took turns starting songs. They took turns starting other songs about animals and fish. ``The \textit{wisi} tamarin goes, `\textit{Sisibarara}'. The \textit{wisi} tamarin goes, `\textit{Sisibarara},' on the fork in the \textit{awa sawa} tree,'' he sang. That's what he was like.

\noindent He started the \textit{kiso} capuchin song right away. ``The pregnant \textit{kiso} capuchin says, `\textit{Hao},''' he sang. ``The pregnant \textit{kiso} capuchin says, `\textit{Hao}.' '' They sang it all very fast and loudly, and it was good, my father used to say.

\noindent ``The \textit{kabobo} lizard, the \textit{yifoma} lizard; it is pregnant, and just hits the water repeatedly [instead of running on the surface]. E Onira, E Onira,'' they sang.

\noindent There were songs about both fish and animals. ``The \textit{bobo} owl hoots on the tree branch,'' he sang.

\noindent When one was over, he started another. He started it. ``I swam to get the fat of the \textit{aba} fish,'' he sang. ``I swam to get the fat of the \textit{aba} fish.''

\noindent When the beating of the \textit{kona} was done, they left to go on the trail singing the E Onira, decorated. They went back. The women got them all wet with mud. They got mud all over them with leaves. They wet them with mud and leaves. The women's fire went out because there were so many of them. [The men came back to the village and went over the fire, and the mud that was dripping off them put out the fire. This was symbolic of the \textit{aba} that comes to the surface because of the \textit{kona} poison.] The fire went out. The women had burned them with brands [which they touched to their skin], and now the fire went out. [The men are upset at being burned by the women, and they put the fire out, so they couldn't burn them anymore.] They had burned them. They were burned because of the women. They were burned, because they burned them. They got mud on them so they could eat, so they could eat fish. They got mud on them and they were burned, they burned them, they were hurt. They put out their fire, so they could shoot fish the next day, and kill many fish.

\noindent They got up early and put the \textit{kona} in the water. They killed lots of fish in a trap. They brought the fish that had been caught by the fish weir, my father used to say. They carried lots of fish in the fish trap. [Several men carried one trap full of fish.]

\noindent Back at the village, they decided to start preparing the fish, and they wrapped the fish in leaves. They roasted the fish on racks. ``Aren't they going to light the feast?'' They danced after they ate. They were still dancing at dawn. They slept during the day. They ate in the morning and then slept, when it was day.
