% \newcommand*{\orcid}[1]{}


\makeatletter
\let\thetitle\@title
\let\theauthor\@author
\makeatother

\newcommand{\togglepaper}[1][0]{
   \bibliography{../localbibliography}
   \papernote{\scriptsize\normalfont
     \theauthor.
     \titleTemp.
     To appear in:
     Adam J.R. Tallman, Sandra Auderset \& Hiroto Uchihara (eds.).
     Constituency and convergence in the Americas.
     Berlin: Language Science Press. [preliminary page numbering]
   }
   \pagenumbering{roman}
   \setcounter{chapter}{#1}
   \addtocounter{chapter}{-1}
}

\newbool{bookcompile}
\booltrue{bookcompile}
\newcommand{\bookorchapter}[2]{\ifbool{bookcompile}{#1}{#2}}

\newfontfamily\cherokeefont{FreeSerif}


% leipzig custom glosses
\newleipzig{III}{III}{set III bound person form}
\newleipzig{II}{II}{set II bound person form}
\newleipzig{I}{I}{set I bound person form}

\newleipzig{abil}{abil}{abilitative}
\newleipzig{abst}{abs.st}{absolute state}
\newleipzig{act}{act}{active}
\newleipzig{add}{add}{additive}
\newleipzig{advr}{advr}{adversative}
\newleipzig{aff}{aff}{affective}
\newleipzig{al}{al}{animals and liquids}
\newleipzig{ali}{ali}{alienable}
\newleipzig{amb}{amb}{ambulative}
\newleipzig{an}{an}{animate}
\newleipzig{and/}{and/}{and, switch-reference marker}
\newleipzig{andt}{andt}{andative}
\newleipzig{anim}{anim}{animals}
\newleipzig{anml}{anml}{animal}
\newleipzig{anph}{anph}{anaphoric marker}
\newleipzig{ant}{ant}{anterior}
\newleipzig{antic}{antic}{anticausative}
\newleipzig{ap}{ap}{applicative/adposition}
\newleipzig{appos}{appos}{apposition}
\newleipzig{asp}{asp}{aspect suffix}
\newleipzig{asr}{asr}{assertive}
\newleipzig{assist}{assist}{assistive}
\newleipzig{assm}{assm}{assumed}
\newleipzig{aug}{aug}{augmentative}
\newleipzig{ana}{ana}{anaphoric demonstrative ('aforementioned')}

\newleipzig{bas}{bas}{basic}
\newleipzig{barg}{b}{set B patientive}

\newleipzig{cisl}{cisl}{cislocative}
\newleipzig{co}{co}{coordinator}
\newleipzig{cmpl}{cmpl}{completive}
\newleipzig{cntg}{cntg}{contingent mood}
\newleipzig{cntr}{cntr}{contrastive}
\newleipzig{cntrfct}{cntrfct}{counterfactual}
\newleipzig{coll}{coll}{collective}
\newleipzig{conc}{conc}{concessive}
\newleipzig{conj}{conj}{conjunction}
\newleipzig{conseq}{conseq}{consequential mood}
\newleipzig{contin}{contin}{continuative}
\newleipzig{contmp}{contmp}{contemporative mood}
\newleipzig{cp}{cp}{completive}

\newleipzig{dcsd}{dcsd}{deceased}
\newleipzig{dep}{dep}{dependent}
\newleipzig{deprec}{deprec}{depreciative}
\newleipzig{des}{des}{desiderative}
\newleipzig{detOne}{det$_{1}$}{determiner 1: not visible} %ka
\newleipzig{detTwo}{det$_{2}$}{determiner 2: going, far}%so
\newleipzig{detThree}{det$_{3}$}{determiner 3: coming, close}%na
\newleipzig{detFour}{det$_{4}$}{determiner 4: static, vertical}%ɾa
\newleipzig{detFive}{det\_{5}}{determiner 5: static, horizontal}%dʒi
\newleipzig{detSix}{det\_{6}}{determiner type 6: static, sitting}
\newleipzig{diff}{diff}{different referent, switch-reference}
\newleipzig{dim}{dim}{diminutive}
\newleipzig{dir}{dir}{directional}
\newleipzig{disap}{disap}{disapprove}
\newleipzig{dm}{dm}{discourse marker}
\newleipzig{down}{down}{downwards}
\newleipzig{dt}{dt}{delimiter}
\newleipzig{ds}{ds}{different-subject conjunction}
\newleipzig{dst}{dst}{distant past}
\newleipzig{dub}{dub}{dubitative}
\newleipzig{dynm}{dynm}{dynamic}

\newleipzig{e}{e}{e}
\newleipzig{emot}{emot}{emotive}
\newleipzig{emph}{emph}{emphasis}
\newleipzig{epen}{epen}{epenthetic}
\newleipzig{epis}{epis}{epistemic}
\newleipzig{evid}{evid}{evidential}
\newleipzig{exist}{exist}{existential}

\newleipzig{far}{far}{far past}
\newleipzig{fact}{fact}{factitive}
\newleipzig{fill}{fill}{filler}
\newleipzig{fin}{fin}{finite}
\newleipzig{final}{final}{final}
\newleipzig{fl}{fl}{flexible}
\newleipzig{flr}{flr}{filler}
\newleipzig{for}{for}{formal}
\newleipzig{freq}{freq}{frequentive}
\newleipzig{frust}{frust}{frustrative}

\newleipzig{gerund}{gerund}{gerundive}

\newleipzig{hab}{hab}{habitual}
\newleipzig{hon}{hon}{honorific}
\newleipzig{hum}{hum}{human}
\newleipzig{hsy}{hsy}{hearsay}

\newleipzig{inan}{inan}{inanimate}
\newleipzig{incept}{incept}{inceptive}
\newleipzig{inch}{inch}{inchoative}
\newleipzig{incomp}{incomp}{incompletive}
\newleipzig{incmpl}{incmpl}{incompletive}
\newleipzig{infr}{infr}{inferred}
\newleipzig{infl}{infl}{inflectional suffix}
\newleipzig{intrc}{intrc}{interactional}
\newleipzig{int}{int}{int}
\newleipzig{inter}{inter}{interrogative}
\newleipzig{intens}{intens}{intensifier}
\newleipzig{inv}{inv}{inverse}
\newleipzig{impf}{impf}{imperfective}
\newleipzig{impers}{impers}{impersonal}
\newleipzig{iter}{iter}{iterative}
\newleipzig{itg}{itg}{intangible}
\newleipzig{itrd}{itrd}{derived intransitive}
\newleipzig{itr}{itr}{intransitivizer}


\newleipzig{jen}{jen}{pluractional/downwards}

\newleipzig{lg}{lg}{long}
\newleipzig{limit}{limit}{limitative}
\newleipzig{linkOne}{link$_{1}$}{linker type 1}% nakoʔ
\newleipzig{linkTwo}{link$_{2}$}{linker type 2}%kan
\newleipzig{lnk}{lnk}{linker}
\newleipzig{locOne}{loc$_{1}$}{locative type 1}%lek
\newleipzig{locTwo}{loc$_{2}$}{locative type 2}%ge
\newleipzig{locThree}{loc$_{3}$}{locative type 3}%git
\newleipzig{locFour}{loc$_{4}$}{locative type 4}%gi
\newleipzig{lv}{lv}{light verb}

\newleipzig{med}{med}{medial}
\newleipzig{mid}{mid}{middle}
\newleipzig{mir}{mir}{mirative}
\newleipzig{mod}{mod}{modality}
\newleipzig{mode}{mode}{mode suffix}
\newleipzig{mods}{mod}{modality suffix}
\newleipzig{mom}{mom}{momentary}
\newleipzig{mot}{mot}{motion}

\newleipzig{nfin}{nfin}{non-finite}
\newleipzig{nfut}{nfut}{non-future}
\newleipzig{nhon}{nhon}{non-honorific}
\newleipzig{nonsg}{nonsg}{non-singular}
\newleipzig{nonvis}{nonvis}{nonvisual}
\newleipzig{nvneg}{nvneg}{non-verbal negation}
\newleipzig{nspec}{nspec}{non-specific (3rd) person}

\newleipzig{oblig}{oblig}{obligative}
\newleipzig{opt}{opt}{optative}
\newleipzig{or}{or}{orientation}
\newleipzig{ov}{ov}{oval things}

\newleipzig{pauc}{pauc}{paucal}
\newleipzig{part}{part}{partitive}
\newleipzig{pft}{pft}{perfect}
\newleipzig{plact}{plact}{pluractional}
\newleipzig{pp}{pp}{participle}
\newleipzig{pot}{pot}{potential}
\newleipzig{pos}{pos}{positional}
\newleipzig{post}{post}{posterior}
\newleipzig{prep}{prep}{preposition}
\newleipzig{prior}{prior}{prior}
\newleipzig{pr}{pr}{pronoun}
\newleipzig{pron}{pron}{independent pronoun}
\newleipzig{pronom}{pronom}{pronominal element}
\newleipzig{prosp}{prosp}{prospective}
\newleipzig{prx}{prx}{proximative aspect}
\newleipzig{punct}{punct}{punctual}

\newleipzig{qr}{qr}{quantifier}

\newleipzig{real}{real}{realis}
\newleipzig{regret}{regret}{regretative}
\newleipzig{rela}{rel}{relationalizing prefix}
\newleipzig{relc}{rel}{relative case}
\newleipzig{relev}{relev}{relevance} 
\newleipzig{rep}{rep}{reportative}
\newleipzig{rem}{rem}{remote}
\newleipzig{resp}{resp}{respect}
\newleipzig{rest}{rest}{restorative}

\newleipzig{same}{same}{same referent, switch-reference}
\newleipzig{sap}{sap}{speech act participant}
\newleipzig{sh}{sh}{super high}
\newleipzig{sim}{sim}{simultaneous}
\newleipzig{spat}{spat}{spatial}
\newleipzig{sploan}{$_{sp}$}{Spanish loanword}
\newleipzig{ss}{ss}{same-subject conjunction}
\newleipzig{stem}{stem}{stem}
\newleipzig{stat}{stat}{stative}
\newleipzig{sub}{sub}{subordinator}
\newleipzig{synt}{synt}{syntactic suffix}

\newleipzig{tag}{tag}{tag}
\newleipzig{tel}{tel}{telic}
\newleipzig{thing}{thing}{things, abstract concepts}
\newleipzig{tprl}{tprl}{temporal}
\newleipzig{tra}{tr}{transitivity prefix}
\newleipzig{trnsl}{trnsl}{translocative}
\newleipzig{trprt}{trprt}{transitive participle mood}
\newleipzig{trvz}{trvz}{transitivizer}
\newleipzig{tplz}{tplz}{topicalizer}

\newleipzig{up}{up}{upwards}

\newleipzig{v}{v}{verb}
\newleipzig{vblz}{vblz}{verbalizer}
\newleipzig{ven}{ven}{venitive}
\newleipzig{vis}{vis}{visual}
\newleipzig{vm}{vm}{valence modifier}

\newleipzig{when}{when}{when, switch-reference marker}
\newleipzig{wood}{wood}{wooden things}

\newleipzig{x}{x}{unspecified person}


\newleipzig{assimil}{assimil.}{assimilation}
\newleipzig{ciscat}{ciscat.}{ciscategoriality}
\newleipzig{ccompl}{compl.}{complement}
\newleipzig{coal}{coal.}{coalescence}
\newleipzig{cons}{cons.}{consonant}
\newleipzig{dds}{DS}{different subject}
\newleipzig{exp}{exp.}{exponent}
\newleipzig{extend}{extend.}{extended}
\newleipzig{max}{max.}{maximum}
\newleipzig{min}{min.}{minimum}
\newleipzig{nointerrupt}{no-interrupt.}{non-interruptability}
\newleipzig{nopermut}{no-permut.}{non-permutability}
\newleipzig{occurr}{occurr.}{occurrence}
\newleipzig{rec}{rec.}{recursion-based}
\newleipzig{rrep}{rep.}{repetition}
\newleipzig{sel}{sel.}{selection}
\newleipzig{sss}{SS}{same subject}
\newleipzig{vclust}{V-clust.}{vowel cluster}


% define color for xcolor
    \definecolor{filled}{gray}{0.6}

% counter
\newcounter{verbslots}[chapter]
\def\positions{\refstepcounter{verbslots}(\theverbslots)}
% define column type based on counter for use in tables
\newcolumntype{S}{>{\positions}r} 

% for nominal planar structure
\newcounter{nounslots}[chapter]
\def\positionsnoun{\refstepcounter{nounslots}(\thenounslots)}
% define column type based on counter for use in tables
\newcolumntype{T}{>{\positionsnoun}r} 

\newcounter{advslots}[chapter]
\def\positionsadv{\refstepcounter{advslots}(\theadvslots)}
% define column type based on counter for use in tables
\newcolumntype{U}{>{\positionsadv}r}

% for diagnostics - comment out if not needed
\newcounter{diagnostics}
\def\rownumberd{}


\newcommand{\lab}{ʷ}
\newcommand{\ssn}{ⁿ}
\newcommand{\1}{¹}
\newcommand{\2}{²}
\newcommand{\3}{³}
\newcommand{\4}{⁴}
\newcommand{\ff}{⁽⁴⁾}
\renewcommand{\st}{ˈ}
\newcommand{\tlg}{ː}
\newcommand{\xy}{ɕ}


\newcommand{\gm}[1]{\textsc{#1}}
\newcommand{\separator}{\begin{center}*\ \ *\ \ *\end{center}}
\newcommand{\phofont}[1]{#1}
\newcommand{\latin}[1]{#1}

\makeatletter
\newcommand\cdotfill{%
    \leavevmode\cleaders\hb@xt@.44em{\hss$\cdot$\hss}\hfill\kern\z@
}
\makeatother
\newcommand{\gc}{\cellcolor{lightgray}}
\newcommand{\dc}{\cellcolor{gray}}

\newcommand{\citegentv}[1]{\citeauthor{#1}'s (\citeyear*{#1} [this volume])}
