\ea \label{PM.1}
\textit{ʕerzû xizmetît kerû heta min jinîyenim pase meşhûr bîyen şewê şê ʕeladînî, şewê peɫê merekuř peɫê merekuř mê şarezûr.} \\ 
\gll ʕerz-û xizmet-î=t ker-û heta min jinîye=n=im pase meşhûr bîye=n şew(e)-ê şê ʕeladîn-î şew(e)-ê peɫ-ê merekuř peɫ-ê merekuř m-ê şarezûr \\ 
 petition\textsc{.m-ez.gen} service\textsc{.m-sg.obl=2sg:R} do\textsc{.prs.ind-1sg:A} even \textsc{1sg} hear\textsc{.pst.ptcp.m=cop.3sg.m:O=1sg:A} such famous be\textsc{.pst.ptcp.m=cop.3sg.m:S} night\textsc{-f.sg.obl} sheikh\textsc{.m} \textsc{pn-m.sg.obl} night\textsc{-f.sg.obl} \textsc{clf-indf} grasshopper \textsc{clf-indf} grasshopper \textsc{ind-}come\textsc{.prs.3sg:S} \textsc{pn} \\ 
\glt `Let me tell you .\footnote{Lit. May a petition be for your service.} I have even heard (this). It is well known that a swarm of grasshoppers came to Sharazour one night.'
\z 
 
\ea \label{PM.2}
\textit{çêrhur mê.} \\ 
\gll çêr=hur m-ê \\ 
 under\textsc{=post} \textsc{ind-}come\textsc{.prs.3sg:S} \\ 
\glt `They came from the south.'
\z 
 
\ea \label{PM.3}
\textit{her yagêwe mê. terefû xwaywe mê.} \\ 
\gll her yagê-(e)we m-ê teref-û xwa-î=we m-ê \\ 
 just place\textsc{.f-indf} \textsc{ind-}come\textsc{.prs.3sg:S} side\textsc{-ez.gen} God\textsc{.m-sg.obl=post} \textsc{ind-}come\textsc{.prs.3sg:S} \\ 
\glt `They came from every direction. They came from God.'
\z 
 
\ea \label{PM.4}
\textit{mê şarezûr.} \\ 
\gll m-ê şarezûr \\ 
 \textsc{ind-}come\textsc{.prs.3sg:S} \textsc{pn} \\ 
\glt `They came to Sharazour.'
\z 
 
\ea \label{PM.5}
\textit{fermandew merekuřekey, řeʕîsekeşa maço, ``ême bê a şexsî metawmê, eçê menmêwe zerer deymê çî mentêqey.''} \\ 
\gll fermande-û merekuř-ekey řeʕîs-eke=şa m-aç-o ême bê a şexs-î me-taw-mê e=çê m\'en-mê=we zerer d\'e-ymê ç=î mentêqe-î \\
 leader\textsc{.m-ez.gen} grasshopper\textsc{-def.m.sg.obl} leader\textsc{.m-def.m.sg.dir=3pl:PSR} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{1pl} without \textsc{dem.dist} saint\textsc{-m.sg.obl} \textsc{neg.ind-}can\textsc{.prs-1pl:A} in=here stay\textsc{.prs.sbjv-1pl:S=compl} harm\textsc{.m} give\textsc{.prs.sbjv-1pl:A} in=\textsc{dem.prox} region\textsc{.m-sg.obl} \\ 
\glt `The head of the grasshoppers said, ``Without [permission from] this person, we cannot stay here and do harm to this region.'''
\z 
 
\ea \label{PM.6}
\textit{nimayinde kîyano xizmetû şê ʕeladînî, sahêbû hîne gewrew merekuřekey.} \\ 
\gll nimayinde kîyan-o xizmet-û şê ʕeladîn-î sahêb-û hîne gewre-û {} merekuř-ekey \\ 
 representative\textsc{.m} send\textsc{.prs.ind-3sg:A} service\textsc{.m-ez.gen} sheikh\textsc{.m} \textsc{pn-m.sg.obl} owner\textsc{.m-ez.gen} \textsc{fill} head\textsc{.m-ez.gen} {} grasshopper\textsc{-def.m.sg.obl} \\ 
\glt `He, i.e., the head of the grasshoppers, sent a representative to Sheikh Aladin.'
\z 
 
\ea \label{PM.7}
\textit{nimayindey kîyano bîyarê.} \\ 
\gll nimayinde-î kîyan-o bîyarê \\ 
 representative\textsc{.m-sg.obl} send\textsc{.prs.ind-3sg:A} \textsc{pn} \\ 
\glt `He sent a messenger to Biyare,'
\z 
 
\newpage
\ea \label{PM.8}
\textit{fermawo, ``luwe xizmetû ađî!} \\ 
\gll fermaw-o l\'u-e xizmet-û ađî \\ 
 say\textsc{.prs.ind-3sg:A} go\textsc{.prs.imp-2sg:S} service\textsc{.m-ez.gen} \textsc{3sg.obl.m} \\ 
\glt `[and] said, ``Go into his presence!'
\z 
 
\ea \label{PM.9}
\textit{îcaze, bizane ađ îcaze miđo ême eçê nehar kermê yam ne.''} \\ 
\gll îcaze bi-zan-e ađ îcaze mi-đ(e)-o ême e=çê nehar k\'er-mê yam ne \\
 permission \textsc{imp-}know\textsc{.prs-2sg:A} \textsc{3sg.m.dir} permission \textsc{ind-}give\textsc{.prs-3sg:A} \textsc{1pl} in=here lunch\textsc{.m} do\textsc{.prs.sbjv-1pl:A} or no \\ 
\glt `See if he lets us stay here for lunch or not.'''
\z 
 
\ea \label{PM.10}
\textit{nimayindeke mê xizmetû şêxî.} \\ 
\gll nimayinde-(e)ke m-ê xizmet-û şêx-î \\ 
 representative\textsc{.m-def.m.sg.dir} \textsc{ind-}come\textsc{.prs.3sg:S} service\textsc{.m-ez.gen} sheikh\textsc{.m-sg.obl} \\ 
\glt `The messenger came to the Sheikh.'
\z 
 
\ea \label{PM.11}
\textit{maço, ``qurban ême tayfênmê amεnmêre.} \\ 
\gll m-aç-o qurban ême tayf(e)-ê=nmê amε=nmê=re \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A} sir\textsc{.voc} \textsc{1pl} tirbe\textsc{.m-indf=cop.1pl:S} come\textsc{.pst.ptcp.pl=cop.1pl:S=povb} \\ 
\glt `[and] said, ``Sir, we are a family [who] have come here.'
\z 
 
\ea \label{PM.12}
\textit{înaymê fiɫane yagê.} \\ 
\gll îna-ymê fiɫan-e yagê \\ 
 \textsc{deic-1pl:S} such\_and\_such\textsc{-ez.cmpd} place\textsc{.f} \\ 
\glt `We are at such-and-such a place.'
\z 
 
\ea \label{PM.13}
\textit{kîyasenaşa xizmetû to.} \\ 
\gll kîyase=na=şa xizmet-û to \\ 
 send\textsc{.pst.ptcp.m=cop.1sg:O=3pl:A} service\textsc{.m-ez.gen} \textsc{2sg} \\ 
\glt `I have been sent to you.'
\z 
 
\newpage
\ea \label{PM.14}
\textit{eger to îcaze miđey ême eçê neharê kermê îcaze be desû ton.} \\ 
\gll eger to îcaze mi-đe-î ême e=çê nehar-ê k\'er-mê îcaze be des-û to=n \\
 if \textsc{2sg} permission \textsc{ind-}give\textsc{.prs-2sg:A} \textsc{1pl} in=here lunch\textsc{.m-indf} do\textsc{.prs.sbjv-1pl:A} permission to hand\textsc{.m-ez.gen} \textsc{2sg=cop.3sg.m:S} \\ 
\glt `It is in your hands whether to let us stay here for lunch.'
\z 
 
\ea \label{PM.15}
\textit{amεnmê xizmetû to îcazema biđey.''} \\ 
\gll amε=nmê xizmet-û to îcaze=ma bi-đ\'e-î \\
 come\textsc{.pst.ptcp.pl=cop.1pl:S} service\textsc{.m-ez.gen} \textsc{2sg} permission\textsc{=1pl:R} \textsc{sbjv}-give\textsc{.prs-2sg:A} \\ 
\glt `I\footnote{Lit. we} have come to your service so that you might permit us.'''
\z 
 
\ea \label{PM.16}
\textit{şêxî ʕeladînî fermawo, ``lodêra şarezûr!} \\ 
\gll şêx-î ʕeladîn-î fermaw-o l\'o-dê=ra şarezûr \\
 sheikh\textsc{-ez} \textsc{pn-m.sg.obl} say\textsc{.prs.ind-3sg:A} go\textsc{.prs.imp-2pl:S=povb} \textsc{pn} \\ 
\glt `Sheikh Aladin said, ``Go down to Sharazour!'
\z 
 
\ea \label{PM.17}
\textit{ana ême wêma mehelêma hen tûl.''} \\ 
\gll an(e)=a ême wê=ma mehel(e)-ê=ma hen-∅ tûl \\ 
 \textsc{dem.dist.m.3sg.dir=deic} \textsc{1pl} \textsc{reflx=1pl:PSR} neighbourhood\textsc{.m-indf=1pl:NC} \textsc{exist-3sg.m:S} long \\ 
\glt `There, I, myself, own a big piece of farmland.''\footnote{Lit. We, ourselves, have a big piece of farmland.}'
\z 
 
\ea \label{PM.18}
\textit{muɫkêşa bîyebo çane. temamû muɫkekey kuliş ce zeraʕetne bîyebo amabonew.} \\ 
\gll muɫk-ê=şa bîye=b-o ç=a=ne temam-û muɫk-ekey kul=iş ce zeraʕet=ne bîye=b-o ama=b-one=û \\ 
 property\textsc{.m-indf=3pl:NC} be\textsc{.pst.ptcp.m}=be\textsc{.prs-3sg:S} in=\textsc{dem.dist=post} all\textsc{-ez.gen} property\textsc{.m-def.m.sg.obl} all\textsc{=3sg:PSR} from agriculture\textsc{.m=post} be\textsc{.pst.ptcp.m}=be\textsc{.prs-3sg:S} come\textsc{.pst.ptcp.m}=be\textsc{.prs-3sg:S}=and \\ 
\glt `-- It appears that he\footnote{they} had a big property. The property might have been filled with crops, and they might have been ripe.'
\z 
 
\ea \label{PM.19}
\textit{menzûriş zeraʕetî bîyebê yawabo penê.} \\ 
\gll menzûr=iş zeraʕet-î bîye=b-ê yawa=b-o penê \\ 
 intention\textsc{.m=3sg:PSR} agriculture\textsc{.m-sg.obl} be\textsc{.pst.ptcp.m}=be\textsc{.prs-aug.3sg:S} arrive\textsc{.pst.ptcp.m}=be\textsc{.prs-3sg:S} to \\ 
\glt `He meant the crops [which] were ready to be harvested.'
\z 
 
\ea \label{PM.20}
\textit{fermawebê, fermawo ``lodê eça muɫkû ême wêma, herçêma hen çane a firmande wêş zano heqû ême kama.} \\ 
\gll fermawe=b-ê fermaw-o l\'o-dê e=ça muɫk-û ême wê=ma herçê=ma hen-∅ ç=a=ne a firmande wê=ş zan-o heq-û ême kam=a \\
 say\textsc{.pst.ptcp.m}=be\textsc{.prs-aug.3sg:O} say\textsc{.prs.ind-3sg:A} go\textsc{.prs.imp-2pl:S} in=there property\textsc{.m-ez.gen} \textsc{1pl} \textsc{reflx=1pl:PSR} whatever\textsc{=1pl:NC} \textsc{exist-3sg.m:S} in=\textsc{dem.dist=post} \textsc{dem.dist} leader\textsc{.m} \textsc{reflx=3sg:PSR} know\textsc{.prs.ind-3sg:A} right\textsc{.m-ez.gen} \textsc{1pl} which\textsc{=cop.3sg.m:S} \\ 
\glt `He said, ``Go! In that property of mine, whatever we possess there, your boss knows which portion [of land] is mine.'
\z 
 
\ea \label{PM.21}
\textit{êşew nehar kerdê, pey neharû wêtaw. seʕbê xwahafîzta!} \\ 
\gll êşew nehar k\'er-dê pey nehar-û wê=ta=û seʕb(e)-ê xwahafîz=ta \\
 tonight lunch\textsc{.m} do\textsc{.prs.imp-2pl:A} for lunch\textsc{.m-ez.gen} \textsc{reflx=2pl:PSR}=and morning\textsc{-f.sg.obl} goodbye\textsc{.m=2pl:PSR} \\ 
\glt `[Take it] as a camp tonight [and] eat from it as your lunch. [Then,] tomorrow, bid farewell!'
\z 
 
\ea \label{PM.22}
\textit{nemendêwe eçê!''} \\ 
\gll ne-men-dê=we e=çê \\ 
 \textsc{proh-}remain\textsc{.prs-2pl:S=compl} in=here \\
\glt `Don't stay here!'''
\z 
 
\largerpage[2]
\ea \label{PM.23}
\textit{şewê merekuř nîşo milo mehelekew şêxî ʕeladînî.} \\ 
\gll şew(e)-ê merekuř nîş-o mi-l-o mehele-(e)ke-û şêx-î ʕeladîn-î \\ 
 night\textsc{-f.sg.obl} grasshopper\textsc{.m} sit\textsc{.prs.ind-3sg:S} \textsc{ind-}go\textsc{.prs-3sg:S} neighbourhood\textsc{.m-def.m.sg.dir-ez.gen} sheikh\textsc{-ez} \textsc{pn-m.sg.obl} \\ 
\glt `The grasshoppers stayed\footnote{Lit. sat} [there] during the night. They went to Sheikh Aladin's land.'
\z\newpage
 
\ea \label{PM.24}
\textit{herçê bo seʕbê hurmêzawe miđya tenha geɫεwe nemenen.} \\ 
\gll herçê b-o seʕb(e)-ê hur-m-êz-a=we mi-đy(e)-a tenha geɫεwe ne-mene=n \\ 
 whatever be\textsc{.prs.ind-3sg:S} morning\textsc{-f.sg.obl} \textsc{pvb-ind-}rise\textsc{.prs-3pl:S=compl} \textsc{ind-}look\textsc{.prs-3pl:S} not\_even leaf\textsc{.m.indf} \textsc{neg-}remain\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `whatever there was, the next day they (i.e., people from Sheikh Aladin's entourage) got up [and] noticed that nothing was left.'
\z 
 
\ea \label{PM.25}
\textit{çoɫ bo luwa! çoɫ.} \\ 
\gll çoɫ b-o luwa çoɫ \\ 
 deserted be\textsc{.prs.ind-3sg:S} go\textsc{.pst.3sg:S} deserted \\ 
\glt `[The field] was bare.'
\z 
 
\ea \label{PM.26}
\textit{ħîçiş nemenen merekuř namêş nemenyêne.} \\ 
\gll ħîç=iş ne-mene=n merekuř namê=ş ne-menyê=ne \\ 
 nothing\textsc{=3sg:NC} \textsc{neg-}remain\textsc{.pst.ptcp.m=cop.3sg.m:S} grasshopper name\textsc{.f=3sg:NC} \textsc{neg-}remain\textsc{.pst.ptcp.f=cop.3sg.f:S} \\ 
\glt `There was nothing left in it. The grasshoppers had disappeared.\footnote{Lit. The grasshopper, its name had not remained.}'
\z 
 
\ea \label{PM.27}
\textit{her şewêw meheleke fire bîyenû kem wardenşû luwan.} \\ 
\gll her şew(e)-ê=û mehele-(e)ke fire bîye=n=û kem warde=n=ş=û luwa=n \\ 
 just night\textsc{-f.sg.obl}=and neighbourhood\textsc{.m-def.m.sg.dir} a\_lot be\textsc{.pst.ptcp.m=cop.3sg.m:S}=and little eat\textsc{.pst.ptcp.m=cop.3sg.m:O=3sg:A}=and go\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `Whatever had been there on that piece of farmland, whether large or small, they (i.e., the grasshoppers) had eaten it during the night [and] gone away.'
\z 
 
\ea \label{PM.28}
\textit{``tate řow bawke řo!'' verzîyer mê pey xizmetû şa ʕeladînî.} \\ 
\gll tate řo=û bawke řo verzîyer m-ê pey xizmet-û şa ʕeladîn-î \\ 
 father\textsc{.m} \textsc{onom}=and fahter\textsc{.voc} \textsc{onom} farm\_labourer\textsc{.m} \textsc{ind-}come\textsc{.prs.3sg:S} for service\textsc{.m-ez.gen} king\textsc{.m} \textsc{pn-m.sg.obl} \\ 
\glt `[Sheikh Aladin's farm-labourer shouted], ``Oh, dear God!'' The farm labourer went to Sheikh Aladin.'
\z 
 
\ea \label{PM.29}
\textit{maço `qurban, îne çêşit kerdû?} \\ 
\gll m-aç-o qurban \textsc{îne} çêş=it kerd{-\O}=û \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A} sir\textsc{.voc} \textsc{prst} what\textsc{=2sg:A} do\textsc{.pst-3sg.m:O}=and \\  
\glt `They said, ``Sir, what have you done?'
\z 
 
\ea \label{PM.30}
\textit{îne bê îcazew to îne nekiryan.} \\ 
\gll îne bê îcaze-û to îne ne-kir-ya=n \\ 
 \textsc{dem.prox.m.3sg.dir} without permission\textsc{-ez.gen} \textsc{2sg} \textsc{dem.prox.m.3sg.dir} \textsc{neg-}do\textsc{-pst.pass=cop.3sg.m:S} \\ 
\glt `This couldn't have been done without your permission.'
\z 
 
\ea \label{PM.31}
\textit{tenha hinû ton, geɫεwew kesî zereriş neginan ce şarezûrne} \\ 
\gll tenha hin-û to=n geɫεwe-û kes-î zerer=iş ne-gina=n ce şarezûr=ne \\ 
 only \textsc{ez.pron-ez.gen} \textsc{2sg=cop.3sg.m:S} leaf\textsc{.m.indf-ez.gen} person\textsc{.m-sg.obl} harm\textsc{.m=3sg:R} \textsc{neg-}fall\textsc{.pst.ptcp.m=cop.3sg.m:S} in \textsc{pn=post} \\ 
\glt `It's all yours, no one else's leaves have been harmed in Sharazour.'
\z 
 
\ea \label{PM.32}
\textit{îne çî e pêse?} \\ 
\gll îne çî e=pêse \\ 
 \textsc{prst} why \textsc{emph}=like \\ 
\glt `Why [is] this situation like this?'
\z 
 
\ea \label{PM.33}
\textit{fermawo, nařihet nebî!} \\ 
\gll fermaw-o nařihet ne-b-î \\ 
 say\textsc{.prs.ind-3sg:A} sad \textsc{neg.sbjv-}be\textsc{.prs-2sg:S} \\ 
\glt `He said, ``May you not be worried!''
\z 
 
\ea \label{PM.34}
\textit{nařihet nebî ane meʕmûre bê terefû xwaywe amabê.} \\ 
\gll nařihet ne-b-î ane meʕmûre b-ê teref-û xwa-î=we ama=b-ê \\ 
 sad \textsc{neg.sbjv-}be\textsc{.prs-2sg:S} \textsc{dem.dist.m.3sg.dir} agent\textsc{.f}  be\textsc{.prs-aug.3sg:S} side\textsc{-ez.gen} God\textsc{.m-sg.obl=post} come\textsc{.pst.ptcp.m}=be\textsc{.prs-aug.3sg:S} \\ 
\glt `Don't worry! It was an agent [who] was sent  by God.'
\z 
 
\ea \label{PM.35}
\textit{eger yerê řowê şarezûrne menêwe geɫayê sewz şarezûrne nazê.} \\ 
\gll eger yerê řo-ê şarezûr=ne men-ê=we geɫa-ê sewz şarezûr=ne n(e)-az-ê \\ 
 if three day\textsc{.m-pl.dir} \textsc{pn=post} remain\textsc{.prs-aug.3pl:S=compl} leaf\textsc{.m-indf} green \textsc{pn=post} \textsc{neg-}let\textsc{.prs-aug.3sg:A} \\ 
\glt `If the grasshoppers had stayed in Sharazour for three days, they wouldn't have left behind a single green leaf in Sharazour.'
\z 
 
\ea \label{PM.36}
\textit{min pewkî îcazem dan hinekew ême wero, bilo zerer ce kesî negino.} \\ 
\gll min pewkî îcaze=m da=n hin-eke-û ême w\'er-o bi-l-o zerer ce kes-î ne-gin-o \\
 \textsc{1sg} that\_is\_why permission\textsc{=1sg:A} give\textsc{.pst.ptcp.m=cop.3sg.m:O} \textsc{ez.pron-def.m.sg.dir-ez.gen} \textsc{1pl} eat\textsc{.prs.sbjv-3sg:A} \textsc{sbjv-}go\textsc{.prs-3sg:S} harm\textsc{.m} from person\textsc{.m-sg.obl} \textsc{neg.sbjv-}fall\textsc{.prs-3sg:S} \\ 
\glt `I let them [eat our crop] so that they would eat our crop and go away, [but] not do harm to others.'
\z 
 
\ea \label{PM.37}
\textit{ême girdû şarezûrî zano êsaɫ miɫkekema nemenenû zeraʕetma nîyaw; bencanîma nîyenew; xîyarma nîyaw; genmîma nîyenew; îne nîyaw; ûne nîyaw; ûne nîya.} \\ 
\gll ême gird-û şarezûr-î zan-o êsaɫ miɫk-eke=ma ne-mene=n=û zeraʕet=ma nîy(e)=a=û bencanî=ma nîye=ne=û xîyar=ma nîy(e)=a=û genmî=ma nîye=ne=û îne nîy(e)=a=û ûne nîy(e)=a=û ûne nîy(e)=a \\ 
 \textsc{1pl} all\textsc{-ez.gen} \textsc{pn-m.sg.obl} know\textsc{.prs.ind-3sg:A} this\_year property\textsc{.m-def.m.sg.dir=1pl:PSR} \textsc{neg-}remain\textsc{.pst.ptcp.m=cop.3sg.m:S}=and agriculture\textsc{.m=1pl:NC} \textsc{neg.exist=cop.3sg.m:S}=and tomato\textsc{.f=1pl:NC} \textsc{neg.exist=cop.3sg.f:S}=and cucmber\textsc{.m=1pl:NC} \textsc{neg.exist=cop.3sg.m:S}=and wheat\textsc{.f=1pl:NC} \textsc{neg.exist=cop.3sg.f:S}=and \textsc{dem.prox.m.3sg.dir} \textsc{neg.exist=cop.3sg.m:S}=and \textsc{dem.dist} \textsc{neg.exist=cop.3sg.m:S}=and \textsc{dem.dist} \textsc{neg.exist=cop.3sg.m:S} \\ 
\glt `Everyone in Sharazour knows that we have not cultivated much land this year; we don't have much agriculture; we don't have cucumbers; we don't have tomatoes; we don't have wheat; This [crop] is not available; that [crop] is not available; that [crop] is not available.'
\z 
 
\ea \label{PM.38}
\textit{girđû şarezûrî êsaɫ meraʕetû ême kero ce parî zîyater pey toyçû pey êmeyç.} \\ 
\gll girđ-û şarezûr-î êsaɫ meraʕet-û ême ker-o ce par-î zîya-ter pey to=îç=û pey ême=îç \\ 
 all\textsc{-ez.gen} \textsc{pn-m.sg.obl} this\_year care\textsc{.m-ez.gen} \textsc{1pl} do\textsc{.prs.ind-3sg:A} from last\_year\textsc{-m.sg.obl} increased\textsc{-cmpr} for \textsc{2sg=add}=and for \textsc{1pl=add} \\ 
\glt `This year, all [the people in] Sharazour care for us -- both you and I\footnote{Lit. we} --, more than the previous year.'
\z 
 
\ea \label{PM.39}
\textit{hîçkamma mihtacê mebîmê pîney.} \\ 
\gll hîçkam=ma mihtac-ê me-b-îmê p=îney \\ 
 no\_one\textsc{=1pl:PSR} needy\textsc{-pl} \textsc{neg.ind-}be\textsc{.prs-1pl:S} to=\textsc{\textsc{dem.prox}.obl.m.3sg} \\ 
\glt `None of us would be in need of these [products].'
\z 
 
\ea \label{PM.40}
\textit{bilowe xemit nebo!} \\ 
\gll bi-lo=we xem=it ne-b-o \\ 
 \textsc{sbjv-}go\textsc{.prs.2sg:S=compl} sorrow\textsc{.m=2sg:PSR} \textsc{neg.sbjv-}be\textsc{.prs-3sg:S} \\ 
\glt `Go back; don't worry!'''
\z 
 
\ea \label{PM.41}
\textit{wextêwe şarezûr mênêw wesîle yawo penîyû.} \\ 
\gll wext-êwe şarezûr m-ê-nê=û wesîle yaw-o penî=û \\ 
 time\textsc{.m-indf} \textsc{pn} \textsc{ind-}come\textsc{.prs-3pl:S}=and product\textsc{.m} arrive\textsc{.prs.ind-3sg:S} at=and \\ 
\glt `When they came back from Sharazour, the crops were ripe for harvesting.'
\z 
 
\ea \label{PM.42}
\textit{şêxî ʕeladîn hemîşey bîyarêwe girđçê zîya kero.} \\ 
\gll şêx-î ʕeladîn hemîşe-î bîyarê=we girđçê zîya ker-o \\ 
 sheikh\textsc{.m-sg.obl} \textsc{pn} always\textsc{-m.sg.obl} \textsc{pn=post} everything increased do\textsc{.prs.ind-3sg:A} \\ 
\glt `Sheikh Aladin was used to having more agricultural products in Biyare.'
\z 
 
\ea \label{PM.43}
\textit{girđçê bencanî bo xîyarî bo bermê ba pey verzîyerîyû pey îđîyû pey ewîyû pey ewîyû pey ewî.} \\ 
\gll girđçê bencanî b-o xîyarî b-o ber-mê ba pey verzîyer-î=û pey îđî=û pey ewî=û pey ewî=û pey ewî \\ 
 everything tomato\textsc{.f} be\textsc{.prs.ind-3sg:S} cucumber be\textsc{.prs.ind-3sg:S} take\textsc{.prs.ind-1pl:A} \textsc{hort} for farm\_labourer\textsc{.m-sg.obl}=and for \textsc{3sg.obl.m}=and for \textsc{3sg.obl.m}=and for \textsc{3sg.obl.m}=and for \textsc{3sg.obl.m} \\ 
\glt `Everything [was there]; There were cucumbers; there were tomatoes. [People said], ``Let's take [these crops] to the farm-labourer; to this [person]; to that [person]; to that [person] and so on.'
\z 
 
\ea \label{PM.44}
\textit{werzîyereke maço, weɫahî şêxî ʕeladîne řasiş fermawan.} \\ 
\gll werzîyer-eke m-aç-o weɫahî şêx-î ʕeladîn=e řas=iş fermawa=n \\ 
 farm\_labourer\textsc{.m-def.m.sg.dir} \textsc{ind-}say\textsc{.prs-3sg:A} indeed sheikh\textsc{.m-sg.obl} \textsc{pn=dem} truth\textsc{=3sg:A} say\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `The farm-labourer said; ``Indeed, Sheikh Aladin was right.'
\z 
 
\ea \label{PM.45}
\textit{min îse î nîştenare îsrahetîç kerû.} \\ 
\gll min îse î nîşte=na=re îsrahet=îç ker-û \\ 
 \textsc{1sg} now \textsc{dem.prox} sit\textsc{.pst.ptcp.m=cop.1sg:S=povb} rest\textsc{.m=add} do\textsc{.prs.ind-1sg:A} \\ 
\glt `Now I am seated, and I am resting.'
\z 
 
\ea \label{PM.46}
\textit{çowe ew mê peym maro.} \\ 
\gll ç=o=we ew m-ê pey=m m-ar-o \\ 
 in=\textsc{dem.dist=post} \textsc{dem.dist} \textsc{ind-}come\textsc{.prs.3sg:S} for\textsc{=1sg:R} \textsc{ind-}bring\textsc{.prs-3sg:A} \\ 
\glt `From that direction, he  comes [to me] and brings [food] for me.'
\z 
 
\ea \label{PM.47}
\textit{êđ çowe mê peym maro.} \\ 
\gll êđ ç=o=we m-ê pey=m m-ar-o \\ 
 \textsc{3sg.prox.dir} in=\textsc{dem.dist=post} \textsc{ind-}come\textsc{.prs.3sg:S} for\textsc{=1sg:R} \textsc{ind-}bring\textsc{.prs-3sg:A} \\ 
\glt `From this direction, he  comes [to me] and brings [food] for me.'
\z 
 
\ea \label{PM.48}
\textit{ew çowe mê peym maro.} \\ 
\gll ew ç=o=we m-ê pey=m m-ar-o \\ 
 \textsc{dem.dist} in=\textsc{dem.dist=post} \textsc{ind-}come\textsc{.prs.3sg:S} for\textsc{=1sg:R} \textsc{ind-}bring\textsc{.prs-3sg:A} \\ 
\glt `From that direction, he  comes [to me] and brings [food] for me.'
\z 
 
\ea \label{PM.49}
\textit{fermawo dey êtir min çî nařihet bîya!} \\ 
\gll fermaw-o dey êtir min çî nařihet bî-a \\ 
 say\textsc{.prs.ind-3sg:A} \textsc{disc.ptcl} \textsc{disc.ptcl} \textsc{1sg} why sad become\textsc{.pst-1sg:S} \\ 
\glt `So, why did I get worried at all?'
\z 
 
\ea \label{PM.50}
\textit{çî adiz bîya?} \\ 
\gll çî adiz bî-a \\ 
 why upset become\textsc{.pst-1sg:S} \\ 
\glt `Why did I get upset?'
\z 
 
\ea \label{PM.51}
\textit{îne heɫbet xway meylişa îne} \\ 
\gll îne heɫbet xwa-î meyl=ş=a îne \\ 
 \textsc{dem.prox.m.3sg.dir} obviously God\textsc{.m-sg.obl} willingness\textsc{=3sg:NC=cop.3sg.m:S} \textsc{dem.prox.m.3sg.dir} \\ 
\glt `This is indeed God's will.'
\z 
 
\ea \label{PM.52}
\textit{îne heɫbet xway meylşa.} \\ 
\gll îne heɫbet xwa-î meyl=ş=a \\ 
 \textsc{dem.prox.m.3sg.dir} obviously God\textsc{.m-sg.obl} willingness\textsc{=3sg:NC=cop.3sg.m:S} \\ 
\glt `This is indeed God's will.'''
\z 
 
\ea \label{PM.53}
\textit{haɫîta?} \\ 
\gll haɫî=t=a \\ 
 understood\textsc{=2sg:NC=cop.3sg.m:S} \\ 
\glt `You see?'
\z
