
\ea \label{ZB.1}
\textit{řowê ce řowa yuwe mê yoyç mê xizmetû şê ʕeladînî, her a şê ʕeladîne.} \\ 
\gll řo-ê ce řo-a yuwe m-ê yo=îç m-ê xizmet-û şê ʕeladîn-î her a şê ʕeladîn=e \\ 
 day\textsc{.m-indf} from day\textsc{.m-pl.obl} one\textsc{.f} \textsc{ind-}come\textsc{.prs.3sg:S} one\textsc{.m=add} \textsc{ind-}come\textsc{.prs.3sg:S} service\textsc{.m-ez.gen} sheikh\textsc{.m} \textsc{pn-m.sg.obl} \textsc{emph} \textsc{dem.dist} sheikh\textsc{.m} \textsc{pn=dem} \\ 
\glt `Once,\footnote{Lit. One day of days} a woman and a man\footnote{Lit. One (f.) came, one (m.) came.} came to the service of Sheikh Aladin, Sheikh Aladin himself.'
\z 
 
\ea \label{ZB.2}
\textit{wişkesaɫî bo pêse îsey.} \\ 
\gll wişkesaɫî b-o pêse îse-î \\ 
 drought\textsc{.m} be\textsc{.prs.ind-3sg:S} like now\textsc{-m.sg.obl} \\ 
\glt `[This region] was in drought, like [it is] now.'
\z 
 
\ea \label{ZB.3}
\textit{heywandar bo.} \\ 
\gll heywandar\textsc{.m} b-o \\ 
 herdsman be\textsc{.prs.ind-3sg:S} \\ 
\glt `He was a herdsman.'
\z 
 
\ea \label{ZB.4}
\textit{mênê qeredax.} \\ 
\gll m-ê-nê qeredax \\ 
 \textsc{ind-}come\textsc{.prs-3pl:S} \textsc{pn} \\ 
\glt `They (the man and his wife) came to Qaradagh.'
\z 
 
\ea \label{ZB.5}
\textit{eça xizone eşkewtêwe. heywandare heywanekeyş bero memliketê çoɫû.} \\ 
\gll e=ça xiz-one eşkewt(e)-êwe heywandar-e heywan-ekey=ş ber-o memliket-ê çoɫ=û \\ 
in=there creep\textsc{.prs.ind-3sg:S} cave\textsc{.f-indf} herdsman\textsc{-def} animal\textsc{.m-def.m.sg.obl=3sg:O} take\textsc{.prs.ind-3sg:A} country\textsc{.m-indf} deserted\textsc{.m}=and\\ 
\glt `There, they crept into a cave. The herdsman took his animals to a deserted region.'
\z 
 
\ea \label{ZB.6}
\textit{a wextîyekey ça nîşorew ta hamin bo.} \\ 
\gll a wext-î-ekey ça nîş-o=re=o ta hamin b-o \\ 
\textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} there sit\textsc{.prs.ind-3sg:S=povb=compl} until summer\textsc{.m} be\textsc{.prs.ind-3sg:S} \\ 
\glt `Then, they settled there until the summer.'
\z 
 
\ea \label{ZB.7}
\textit{ginmê dewrû haminî weharî temamîyo.} \\ 
\gll gin-mê dewr-û hamin-î wehar-î temamîy(e)-o \\ 
 fall\textsc{.prs.ind-1pl:S} around\textsc{-ez.gen} summer\textsc{.m-sg.obl} spring\textsc{.m-sg.obl} finish\textsc{.prs-3sg:S} \\ 
\glt `It was\footnote{Lit. We fell.} near the end of summer, [or] spring.'
\z 
 
\ea \label{ZB.8}
\textit{mê şewê ce şewa jenêweş bo. wêşû kinaçê ezemêç laşo bo.} \\ 
\gll m-ê şew(e)-ê ce şew(e)-a jen(î)-êwe=ş b-o wê=ş=û kinaçê ezem-(e)-ê=ç la=ş=o b-o \\ 
 \textsc{ind-}come\textsc{.prs.3sg:S} night\textsc{.f-indf} from night\textsc{.f-pl.obl} woman\textsc{.f-indf=3sg:NC} be\textsc{.prs.ind-3sg:S} \textsc{reflx=3sg:PSR}=and girl\textsc{.f} adolescent\textsc{-f-indf=add} with\textsc{=3sg:R=post} be\textsc{.prs.ind-3sg:S} \\ 
\glt `There came a night when -- he had a wife. She and an adolescent daughter [of his] were with him.'
\z 
 
\ea \label{ZB.9}
\textit{kuřêş ba. kuřekêş germîyanne nimênêş çene.} \\ 
\gll kuř-ê=ş b-a kuř-ekê=ş germîyan=ne nim(e)-ê-nê=ş çene \\ 
 son\textsc{.m-pl.dir=3sg:NC} be\textsc{.prs.ind-3pl:S} son\textsc{.m-def.pl.dir=3sg:PSR} \textsc{pn=post} \textsc{neg.ind-}come\textsc{.prs-3pl:S=3sg:R} with \\ 
\glt `He had sons. His sons did not accompany him [; they stayed in] Garmiyan.'
\z 
 
\ea \label{ZB.10}
\textit{eça nîşore. şewê waranê waro.} \\ 
\gll e=ça nîş-o=re şew(e)-ê waran-ê war-o \\ 
 in=there sit\textsc{.prs.ind-3sg:S=povb} night\textsc{.f-indf} rain\textsc{-indf} rain\textsc{.prs.ind-3sg:S} \\ 
\glt `They settled there (i.e. in that region). One night, it was raining.'
\z 
 
\ea \label{ZB.11}
\textit{be ʕemrû xwa hewrê gino řû we asmanî.} \\ 
\gll be ʕemr(e)-û xwa hewr-ê gin-o řû=we asman-î \\ 
 to order\textsc{.f-ez.gen} God\textsc{.m} cloud\textsc{-indf} fall\textsc{.prs.ind-3sg:S} up=to sky\textsc{.m-sg.obl} \\ 
\glt `By the order of God, a cloud covered the sky.\footnote{Lit. a cloud fell on the sky.}'
\z 
 
\ea \label{ZB.12}
\textit{tecrebeş bo. ʕemriş dewrû ṣewû wîs saɫane bo wîs saɫê sî saɫê pa dewrne.} \\ 
\gll tecrebe=ş b-o ʕemr=iş dewr-û ṣew=û wîs saɫ(e)-a=ne b-o wîs saɫ(e)-ê sî saɫ(e)-ê p=a dewr=ne \\ 
 experience\textsc{=3sg:NC} be\textsc{.prs.ind-3sg:S} age\textsc{.m=3sg:PSR} around\textsc{-ez.gen} hundred=and twenty year\textsc{.f-pl.obl=post} be\textsc{.prs.ind-3sg:S} twenty year\textsc{.f-pl.dir} thirty year\textsc{.f-pl.dir} at=\textsc{dem.dist} around\textsc{=post} \\ 
\glt `He was experienced. He was around 120 years old, about 120 or 130.'
\z 
 
\ea \label{ZB.13}
\textit{maço, ``jenekê!''} \\ 
\gll m-aç-o jen(î)-ekê \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A} woman\textsc{.f-def.f.sg} \\ 
\glt `He said, ``Wife!'''
\z 
 
\ea \label{ZB.14}
\textit{maço, ``beɫê!''} \\ 
\gll m-aç-o beɫê \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A} yes \\ 
\glt `She said, ``Yes!'''
\z 
 
\ea \label{ZB.15}
\textit{maço, ``êşew waro; tekbîrma mekiryo.''} \\ 
\gll m-aç-o êşew war-o tekbîr=ma {} me-kiry(e)-o \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A} tonight rain\textsc{.prs.ind-3sg:S} solution\textsc{.m=1pl:PSR} {} \textsc{neg.ind-}do\textsc{.prs.pass-3sg:S} \\ 
\glt `He said, ``It's going to rain tonight; we can't handle (it)\footnote{Lit. Our solution won't be done.}'''
\z 
 
\ea \label{ZB.16}
\textit{ewêç maço, ``waranû çêşî?''} \\ 
\gll ewê=ç m-aç-o waran-û çêş-î \\ 
 \textsc{3sg.f.dir=add} \textsc{ind-}say\textsc{.prs-3sg:A} rain\textsc{-ez.gen} what\textsc{-m.sg.obl} \\ 
\glt `She said, ``What rain?!'''
\z 
 
\ea \label{ZB.17}
\textit{maço, ``be xwa waro. ana hewrê sîyaw pise çetrê kewt asman.''} \\ 
\gll m-aç-o be xwa war-o an(e)=a hewr-ê sîyaw pise çetr(e)-ê kewt-∅ asman \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A} to God\textsc{.m} rain\textsc{.prs.ind-3sg:S} \textsc{dem.dist.m.3sg.dir=emph} cloud\textsc{.m-indf} black\textsc{.m} like umbrella\textsc{-f.sg.obl} fall\textsc{.pst-3sg.m:S} sky\textsc{.m.dir} \\ 
\glt `He said, ``[I swear] to God, it will rain. There, a black cloud has covered the sky like an umbrella!'''
\z 
 
\ea \label{ZB.18}
\textit{şewê firê bijyawe.} \\ 
\gll şew(e)-ê fir(e)-ê bijy(e)-a=we \\ 
 night\textsc{-f.sg.obl} a\_lot\textsc{-indf} toss\textsc{.prs-3pl:S=compl} \\ 
\glt `During the night, they tossed and turned [trying to get comfortable].'
\z 
 
\ea \label{ZB.19}
\textit{ca a wextî çawre bo. çawrekêne fire ginawe kem.} \\ 
\gll ca a wext-î çawre b-o çawre-(e)kê=ne fire gin-a=we kem \\ 
 afterwards \textsc{dem.dist} time\textsc{.m-sg.obl} tent\textsc{.f} be\textsc{.prs.ind-3sg:S} tent\textsc{.f-def.f.sg=post} a\_lot fall\textsc{.prs.ind-3pl:S=compl} little \\ 
\glt `Back then, there were tents. They tossed and turned [in bed] in the tent for a while.'
\z 
 
\ea \label{ZB.20}
\textit{wextê weyowe, lawaw aman çawre peře bîye awî.} \\ 
\gll wext-ê wey(e)-o=we lawaw ama=n çawre peř-e bî-e awî \\ 
 time\textsc{.m-indf} wake\textsc{.prs.ind-3sg:S=compl} flood\textsc{.m} come\textsc{.pst.ptcp.m=cop.3sg.m:S} tent\textsc{.f} full\textsc{-f} be\textsc{.pst-3sg.f:S} water\textsc{.f} \\ 
\glt `When they finally woke up, the flood had come [and] the tent was filled with water.'
\z 
 
\ea \label{ZB.21}
\textit{heywane awê berde.} \\ 
\gll heywane aw(î)-ê berd-e \\ 
 animal\textsc{.f.sg.dir} water\textsc{.f-sg.obl} take\textsc{.pst-3sg.f:O} \\ 
\glt `The flood\footnote{Lit. water} took away the animals.'
\z 
 
\newpage
\ea \label{ZB.22}
\textit{epare beʕzêş piřnaşre berd wisiş řoxane.} \\ 
\gll e=pare beʕzê=ş piřna=ş=re berd-∅ wis-∅=iş řoxane\textsc{.m} \\ 
 in=down\_there some\textsc{=3sg:PSR} break\_off\textsc{.pst=3sg:A=povb} take\textsc{.pst-3sg.m:O} throw\textsc{.pst-3sg.m:O=3sg:A} river \\ 
\glt `Down there, [the flood] took some [animals and] cast [them] into the river.'
\z 
 
\ea \label{ZB.23}
\textit{beʕzêçiş meno milwa xizana meřekê.} \\ 
\gll beʕzê=ç=iş men-o mi-lw-a xiz-anê meře-(e)kê \\ 
 some\textsc{=add=3sg:PSR} remain\textsc{.prs.ind-3sg:S} \textsc{ind-}go\textsc{.prs-3pl:S} creep\textsc{.prs.ind-3pl:S} cave\textsc{.f-def.f.sg} \\ 
\glt `Some of [the animals] which had survived went [and] squeezed themselves into the cave.'
\z 
 
\ea \label{ZB.24}
\textit{şewê be ʕemrew xway jenêç nîşore duwê zaroɫê wîno, her a şewê.} \\ 
\gll şew(e)-ê be ʕemre-û xwa-î jen(î)-ê=ç nîş-o=re duwê zaroɫ(e)-ê wîn-o her a şew(e)-ê \\ 
 night\textsc{-f.sg.obl} by order\textsc{.f-ez.gen} God\textsc{.m-sg.obl} woman\textsc{-f.sg.obl=add} sit\textsc{.prs.ind-3sg:S=povb} two child\textsc{-pl.dir} see\textsc{.prs.ind-3sg:A} \textsc{emph} \textsc{dem.dist} night\textsc{-f.sg.obl} \\ 
\glt `During that very night, by the order of God, the wife gave birth to two babies.\footnote{Lit. She sat down [and] saw two babies.}'
\z 
 
\ea \label{ZB.25}
\textit{wizoşare dilû awê.} \\ 
\gll wiz-o=şa=re dil-û aw(î)-ê \\ 
 throw\textsc{.prs.ind-3sg:A=3pl:O=povb} inside\textsc{-ez.gen} water\textsc{.f-sg.obl} \\ 
\glt `She delivered them in the water [which had flooded in the tent].'
\z 
 
\ea \label{ZB.26}
\textit{ça menawe ta seʕbe řoj bowe berzepey miđrarê.} \\ 
\gll ça men-a=we ta seʕbe řoj b-o=we berzepey miđr-a=rê \\ 
 there remain\textsc{.prs.ind-3pl:S=compl} until morning\textsc{.f} daylight\textsc{.m} become\textsc{.prs.ind-3sg:S=compl} standing stop\textsc{.prs-3pl:S=povb} \\ 
\glt `They remained there [in the tent] standing up\footnote{Lit. stopped standing by} until the next day.\footnote{Lit. Until early morning became daylight.}'
\z 
 
\ea \label{ZB.27}
\textit{seʕbê mila wextî miđya penc bizê menênêw.} \\ 
\gll seʕb(e)-ê mi-l-a wext-î mi-đy(e)-a penc biz(e)-ê menê=nê=û \\ 
 morning\textsc{-f.sg.obl} \textsc{ind-}go\textsc{.prs-3pl:S} time\textsc{.m-sg.obl} \textsc{ind-}look\textsc{.prs-3pl:S} five goat\textsc{.f-pl.dir} remain\textsc{.pst.ptcp.pl=cop.3pl:S}=and \\ 
\glt `The next day, when they looked around, they saw that five goats had survived [the flood].'
\z 
 
\ea \label{ZB.28}
\textit{awî luwεnêrew serû ane hardîşa bîyênêw.} \\ 
\gll awî luwε=nê=re=û ser-û ane hardî=şa bîyê=nê=û \\ 
 water\textsc{.f} go\textsc{.pst.ptcp.pl=cop.3pl:S=povb}=and on\textsc{-ez.gen} \textsc{dem.dist.m.3sg.dir} flour\textsc{.f.pl.dir=3pl:NC} be\textsc{.ptcp.pl=cop.3pl:S}=and \\ 
\glt `The water had risen to the level of the flour [sacks] that they had.'
\z 
 
\ea \label{ZB.29}
\textit{kuçê qenû çayû a wextîyekey, ne nan hen, ne çay hen, ne çîw hen, ħîç.} \\ 
\gll kuç-ê qen=û çay=û a wext-î-ekey ne nan hen-∅ ne çay hen-∅ ne çîw hen-∅ ħîç \\ 
 little\textsc{-indf} sugar\_cube=and tea\textsc{.m}=and \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} neither bread\textsc{.m} \textsc{exist-3sg.m:S} nor tea\textsc{.m} \textsc{exist-3sg.m:S} nor thing\textsc{.m} \textsc{exist-3sg.m:S} nothing \\ 
\glt `[There was] some sugar and tea, well then, there was neither bread, nor water, nor anything else, [there was] nothing [left].'
\z 
 
\ea \label{ZB.30}
\textit{awê berdenû luwan.} \\ 
\gll aw(î)-ê berde=n=û luwa=n \\ 
 water\textsc{.f-sg.obl} take\textsc{.pst.ptcp.m=cop.3sg.m:O}=and go\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `The water had taken [everything] away, [and] it was gone.'
\z 
 
\ea \label{ZB.31}
\textit{ħeywanê miđa wer. girđ wizaş weɫê.} \\ 
\gll ħeywan-ê mi-đ(e)-a wer girđ wiz-a=ş weɫê \\ 
 animal\textsc{.m-pl.dir} \textsc{ind-}give\textsc{.prs-3pl:A} out all throw\textsc{.prs.ind-3pl:A=3sg:O} front \\ 
\glt `They herded the animals out. They drove all [of them] forth.'
\z 
 
\ea \label{ZB.32}
\textit{a wextîyekey zaroɫê duwê bawşîşane.} \\ 
\gll a wext-î-ekey zaroɫ(e)-ê duwê bawşî=şa=ne \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} child\textsc{-pl.dir} two embrace\textsc{.f=3pl:PSR=post} \\ 
\glt `And they held the babies in their arms.'
\z 
 
\ea \label{ZB.33}
\textit{yejenekê biřê pîyake maço, ``melama pey.''} \\ 
\gll yejen(î)-ekê biřê pîya-(e)ke m-aç-o me-l-a=ma pey \\ 
 woman\textsc{.f-def.f.sg} a\_while man\textsc{.m-def.m.sg.dir} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{neg.ind-}go\textsc{.prs-3pl:S=1pl:R} for \\ 
\glt `The woman (or) the man said several times, ``It's not possible for us [to take the babies]. \footnote{Lit. they don't go for us.}'''
\z 
 
\ea \label{ZB.34}
\textit{tate xway îrayeşa.} \\ 
\gll tate xwa-î îraye=ş=a \\ 
 father\textsc{.m} God\textsc{.m-sg.obl} volition\textsc{=3sg:NC=cop.3sg.m:S} \\ 
\glt ``Man, this is God's will.''
\z 
 
\ea \label{ZB.35}
\textit{mekiryoyç î kuřî xinîkno.} \\ 
\gll me-kiry(e)-o=îç î kuř-î xin\stackunder[-10pt]{\^{i}}{\'{}}kn-o \\
 \textsc{neg.ind-}do\textsc{.prs.pass-3sg:S=add} \textsc{dem.prox} boy\textsc{.m-sg.obl} drown\textsc{.prs.sbjv-3sg:A} \\ 
\glt `He should not drown the [baby] boys.'\footnote{Lit. It can't be done that he drowns the baby boys.}
\z 
 
\ea \label{ZB.36}
\textit{bero mizyoş qiřoɫû derê, hînêwe meřû darê.} \\ 
\gll ber-o mi-zy(e)-o=ş qiřoɫ-û der(e)-ê hîn-êwe meř(e)-û dar-ê \\ 
 take\textsc{.prs.ind-3sg:A} \textsc{ind-}leave\textsc{.prs-3sg:A=3sg:O} hollow\textsc{.m-ez.gen} valley\textsc{.m-indf} \textsc{fill-indf} cave\textsc{.f-ez.gen} tree\textsc{.m-indf} \\ 
\glt `He took him (i.e., one of the baby boys), [and] put him in a hole in a valley, a thingummy, in the hollow of a tree.'
\z 
 
\newpage
\ea \label{ZB.37}
\textit{îne min înem jinîyen law, fermawiştû şêxîn, şê ʕeladînîwe.} \\ 
\gll îne min îne=m jinîye=n la-û fermawişt-û şêx-î=n şê ʕeladîn-î=we \\ 
 \textsc{dem.prox.m.3sg.dir} \textsc{1sg} \textsc{dem.prox.m.3sg.dir=1sg:A} hear\textsc{.pst.ptcp.m=cop.3sg.m:O} at\_the\_place\_of\textsc{-ez.gen} speech\textsc{.m-ez.gen} sheikh\textsc{.m-sg.obl=cop.3sg.m:S} sheikh \textsc{pn-m.sg.obl=post} \\ 
\glt `I have heard this tale from -- it is the Sheikh's words -- Sheikh Aladin.'
\z 
 
\ea \label{ZB.38}
\textit{a wextîyekey minyoş qiřoɫû beřû darêwew.} \\ 
\gll a wext-î-ekey mi-ny(e)-o=ş qiřoɫ-û beřû dar-êwe=û \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} \textsc{ind-}put\textsc{.prs-3sg:A=3sg:O} hollow\textsc{.m-ez.gen} oak\textsc{.m} tree\textsc{.m-indf}=and \\ 
\glt `Anyhow, he put him into the hollow of an oak tree.'
\z 
 
\ea \label{ZB.39}
\textit{kûç kera milawe pey germîyanî.} \\ 
\gll kûç ker-a mi-l-a=we pey germîyan-î \\ 
 migration\textsc{.m} do\textsc{.prs.ind-3pl:A} \textsc{ind-}go\textsc{.prs-3pl:S=compl} for \textsc{pn-m.sg.obl} \\ 
\glt `They (i.e., the family) migrated [and] went to Garmiyan.'
\z 
 
\ea \label{ZB.40}
\textit{mila ew kuřekey yoyşa bera. aneşa zilterû ʕalter bo ađî bera.} \\ 
\gll mi-l-a ew kuř-ekey yo-î=şa ber-a ane=şa zil-ter=û ʕal-ter b-o ađî ber-a \\ 
 \textsc{ind-}go\textsc{.prs-3pl:S} \textsc{dem.dist} son\textsc{.m-def.m.sg.obl} one\textsc{.m-sg.obl=3pl:PSR} take\textsc{.prs.ind-3pl:A} \textsc{dem.dist.m.3sg.dir=3pl:PSR} big\textsc{-cmpr}=and good\textsc{-cmpr} be\textsc{.prs.ind-3sg:S} \textsc{3sg.obl.m} take\textsc{.prs.ind-3pl:A} \\ 
\glt `They went away [and took] that son. They took one of them (i.e., of the boys), the one who was bigger and healthier; they took him.'
\z 
 
\ea \label{ZB.41}
\textit{ewî minya qiřoɫû darêwe.} \\ 
\gll ewî mi-ny(e)-a qiřoɫ-û dar-êwe \\ 
 \textsc{3sg.obl.m} \textsc{ind-}put\textsc{.prs-3pl:A} hollow\textsc{.m-ez.gen} tree\textsc{.m-indf} \\ 
\glt `They left him (i.e., the other one) in the hole in the tree.'
\z 
 
\ea \label{ZB.42}
\textit{çî mede xwawinû ʕalemîyanî pase îrayeşa bizêw menêne cîyay sayqekewe.} \\ 
\gll ç=î mede xwawin-û ʕalemîyan-î pase îraye=ş=a biz(e)-êw menê=ne cîyay sayqe-(e)ke=we \\ 
 at=\textsc{dem.prox} time owner\textsc{-ez.gen} worlds\textsc{.m-sg.obl} like volition\textsc{=3sg:NC=cop.3sg.m:S} goat\textsc{.f-indf} remain\textsc{.pst.ptcp.f=cop.3sg.f:S} behind lightning\textsc{.m-def.m.sg.dir=post} \\ 
\glt `The will of the lord of all worlds was such that a goat was [accidentally] left behind [alive] from the storm [that caused the flood].'
\z 
 
\ea \label{ZB.43}
\textit{xway pase îrayeşa ađê mila řeđê baw.} \\ 
\gll xwa-î pase îraye=ş=a ađê mi-l-a řeđ-ê b-a=û \\ 
 God\textsc{.m-sg.obl} such volition\textsc{=3sg:NC=cop.3sg.m:S} \textsc{3pl.dir} \textsc{ind-}go\textsc{.prs-3pl:S} crossing\textsc{-pl.dir} be\textsc{.prs-3pl:S}=and \\ 
\glt `God's will was such that they went away, [and] crossed [that region].'
\z 
 
\ea \label{ZB.44}
\textit{bizê her werew mê.} \\ 
\gll biz(e)-ê her wer=ew m-ê \\ 
 goat\textsc{.f-pl.dir} \textsc{emph} front\textsc{.m=post} \textsc{ind-}come\textsc{.prs.3sg:S} \\ 
\glt `and the goat kept going\footnote{Lit. coming forward} [to the tree]'
\z 
 
\ea \label{ZB.45}
\textit{pey minyo serû nerekeyû çêrhur zaroɫeke şot wero.} \\ 
\gll pey mi-ny(e)-o ser-û nere-(e)key=û çêr=hur zaroɫe-(e)ke şot wer-o \\ 
 foot \textsc{ind-}put\textsc{.prs-3sg:A} on\textsc{-ez.gen} oak\_tree\textsc{.m-def.m.sg.obl}{=and} under\textsc{=post} child\textsc{-def.m.sg.dir} milk\textsc{.m} eat\textsc{.prs.ind-3sg:A} \\ 
\glt `[and] putting its feet on the trunk of the oak tree, and the baby drank [its] milk from below.'
\z 
 
\ea \label{ZB.46}
\textit{saɫê ter mênêwe, maço, ``bilmê bizanmê tekbîrî êtir řawêjma çêş bo'' ađê, wişkesaɫîn.} \\ 
\gll saɫ(e)-ê ter m-ê-nê=we m-aç-o bi-l-mê bi-zan-mê tekbîr-î êtir řawêj=ma çêş b-o ađê {} wişkesaɫî=n \\
 year\textsc{.f-indf} other \textsc{ind-}come\textsc{.prs-3pl:S=compl} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{sbjv-}go\textsc{.prs-1pl:S} \textsc{sbjv-}know\textsc{.prs-1pl:A} solution\textsc{-m.sg.obl} \textsc{disc.ptcl} consultation\textsc{=1pl:PSR} what be\textsc{.prs.sbjv-3sg:S} \textsc{3pl.dir} {} drought\textsc{=cop.3sg.m:S} \\ 
\glt `The following year, they went back. He (the father) said, ``We shall go [and] see what is ahead of us with this period of drought.'''
\z 
 
\ea \label{ZB.47}
\textit{mênê yagekê.} \\ 
\gll m-ê-nê yag(ê)-ekê \\ 
 \textsc{ind-}come\textsc{.prs-3pl:S} place\textsc{.f-def.f.sg} \\ 
\glt `They went to the same place [as last year].'
\z 
 
\ea \label{ZB.48}
\textit{wextê miđyo kabra qiřoɫû nerekeyne sere ber mêw milo řû zemînîre.} \\ 
\gll wext-ê mi-đy(e)-o kabra qiřoɫ-û nere-(e)key=ne sere ber m-ê=û mi-l-o řû zemîn-î=re \\ 
 time\textsc{.m-indf} \textsc{ind-}look\textsc{.prs-3sg:S} fellow hollow\textsc{.m-ez.gen} oak\_tree\textsc{.m-def.m.sg.obl=post} head\textsc{.m} out \textsc{ind-}come\textsc{.prs.3sg:S=}and \textsc{ind-}go\textsc{.prs-3sg:S} face land\textsc{.m-sg.obl=post} \\ 
\glt `When he looked [at the tree], he saw someone put his head out of the hollow of the oak tree and pulled it (his head) back again.'
\z 
 
\ea \label{ZB.49}
\textit{milo wextê miđyo zaroɫeke bîyen peřû qiřoɫekey.} \\ 
\gll mi-l-o wext-ê mi-đy(e)-o zaroɫe-(e)ke bîye=n peř-û qiřoɫ-ekey \\ 
 \textsc{ind-}go\textsc{.prs-3sg:S} time\textsc{.m-indf} \textsc{ind-}look\textsc{.prs-3sg:S} child\textsc{-def.m.sg.dir} be\textsc{.pst.ptcp.m=cop.3sg.m:S} full\textsc{-ez.gen} hollow\textsc{.m-def.m.sg.obl} \\ 
\glt `The father went [closer and] noticed that the baby boy [had grown up so much that] he had filled the hollow of the tree.'
\z 
 
\ea \label{ZB.50}
\textit{a wextîyekey îđîçşa ardenû.} \\ 
\gll a wext-î-ekey îđ=îç=şa arde=n=û \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} \textsc{3sg.m.dir=add=3pl:A} bring\textsc{.pst.ptcp.m=cop.3sg.m:O}=and \\ 
\glt `They (the family) brought him back.'
\z 
 
\ea \label{ZB.51}
\textit{marowe kêşoş şiş kîlowê.} \\ 
\gll m-ar-o=we kêş-o=ş şiş kîlo-ê \\ 
 \textsc{ind-}bring\textsc{.prs-3sg:A=compl} weigh\textsc{.prs.ind-3sg:A=3sg:O} six kilo\textsc{.m-pl.dir} \\ 
\glt `The father weighed him, [and] he weighed six kilos [more than the other son].'
\z 
 
\ea \label{ZB.52}
\textit{hinû dilû qiřoɫû nerekey şiş kîlowê çîđî zîya bo.} \\ 
\gll hin-û dil-û qiřoɫ-û nere-(e)key şiş kîlo-ê ç=îđî zîya b-o \\ 
 \textsc{ez.pron-ez.gen} inside\textsc{-ez.gen} hollow\textsc{.m-ez.gen} oak\_tree\textsc{.m-def.m.sg.obl} six kilo\textsc{.m-pl.dir} from=\textsc{3sg.obl.m} more be\textsc{.prs.ind-3sg:S} \\ 
\glt `The one raised in the hollow of the tree weighed six kilos more than him (the one they took with them).'
\z 
 
\ea \label{ZB.53}
\textit{karû xway eçanen.} \\ 
\gll kar-û xwa-î e=çane=n \\ 
 task\textsc{.m-ez.gen} God\textsc{.m-sg.obl} \textsc{emph}=like\_that\textsc{=cop.3sg.m:S} \\ 
\glt `God's work is like this.'
\z 
 
\ea \label{ZB.54}
\textit{î bizêçe powe awêzane bo mê eçê êđ çeknoşû milo la yagekêş.} \\ 
\gll î biz(e)-ê=ç=e p=o=we awêzan-e b-o m-ê e=çê êđ çekn-o=ş=û mi-l-o la yag(ê)-ekê=ş \\ 
 \textsc{dem.prox} goat\textsc{.f-sg.obl=add=dem} at=\textsc{dem.dist=post} hanging\textsc{-f} be\textsc{.prs.ind-3sg:S} \textsc{ind-}come\textsc{.prs.3sg:S} in=here \textsc{3sg.prox.dir} suckle\textsc{.prs-3sg:A=3sg:O}=and \textsc{ind-}go\textsc{.prs-3sg:S} to place\textsc{.f-def.f.sg=3sg:PSR} \\ 
\glt `The goat would stay in a hanging position, he (the child) would feed from its udder, and it would go back to its place.'
\z 
 
\ea \label{ZB.55}
\textit{haɫîta?} \\ 
\gll haɫî=t=a \\ 
 understood\textsc{=2sg:NC=cop.3sg.m:S} \\ 
\glt `You see?'
\z 
 
\ea \label{ZB.56}
\textit{wextê kuřekey maro milo řûweşo waçê pêse neçîrî piřa, piřa.} \\ 
\gll wext-ê kuř-ekey m-ar-o mi-l-o řûwe=ş=o waç-ê pêse neçîr-î piřa-∅ piřa-∅ \\ 
 time\textsc{.m-indf} son\textsc{.m-def.m.sg.obl} \textsc{ind-}bring\textsc{.prs-3sg:A} \textsc{ind-}go\textsc{.prs-3sg:S} face\textsc{=3sg:PSR=post} say\textsc{.prs-aug.3sg:A} like prey\textsc{.m-sg.obl} fly\textsc{.pst-3sg.m:S} fly\textsc{.pst-3sg.m:S} \\ 
\glt `When he approached the son [in the hollow of the tree], it was said that he jumped like a prey; he jumped.'
\z 
 
\ea \label{ZB.57}
\textit{heɫbete bero gêroş. kêştebêş şiş kîlowê ađ çewî zîyater bîyebê.} \\ 
\gll heɫbete ber-o gêr-o=ş kêşte=b-ê=ş şiş kîlo-ê ađ ç=ewî zîya-ter bîye=b-ê \\ 
 of\_course take\textsc{.prs.ind-3sg:A} take\textsc{.prs.ind-3sg:A=3sg:O} weigh\textsc{.pst.ptcp.m}=be\textsc{.prs-aug.sg:O=3sg:A} six kilo\textsc{.m-pl.dir} \textsc{3sg.m.dir} from=\textsc{3sg.obl.m} increased\textsc{-cmpr} be\textsc{.pst.ptcp.m}=be\textsc{.prs-aug.3sg:S} \\ 
\glt `Indeed, he (the father) took him; he grabbed him. When he weighed him, he weighed six kilos more than him (the other son).'
\z 
 
\ea \label{ZB.58}
\textit{îne dastanû gewregerîyû a meşeyxane.} \\ 
\gll îne dastan(e)-û gewregerî-û a meşeyxa=ne \\ 
 \textsc{dem.prox.m.3sg.dir} story\textsc{.f-ez.gen} greatness\textsc{-ez.gen} \textsc{dem.dist} sheikh\textsc{.pl.obl=cop.3sg.f:S} \\ 
\glt `This is the story of the greatness of the sheikhs.'
\z 
 
\ea \label{ZB.59}
\textit{herçêşa kerdebo herçêşa watebo la xwaywe qisêşa meqboɫe bîyêne.} \\ 
\gll herçê=şa kerde=b-o herçê=şa wate=b-o la xwa-î=we qisê=şa meqboɫ-e bîyê=ne \\ 
 whatever\textsc{=3pl:A} do\textsc{.pst.ptcp.m}=be\textsc{.prs-3sg:O} whatever\textsc{=3pl:A} say\textsc{.pst.ptcp.m}=be\textsc{.prs-3sg:O} at\_the\_place\_of God\textsc{.m-sg.obl=post} talk\textsc{.f.sg=3pl:PSR} acceptable\textsc{-f} be\textsc{.pst.ptcp.f=cop.3sg.f:S} \\ 
\glt `Whatever they did, whatever they said, their remarks were accepted by God.'
\z 
 
\ea \label{ZB.60}
\textit{welê îse be qurʕan ne şêx menen, ne meɫa menen, ne kesê hen be weş dađiş berî la.} \\ 
\gll welê îse be qurʕan ne şêx mene=n ne meɫa mene=n ne kes-ê hen-∅ be weş dađ=iş b\'er-î la \\
 but now to \textsc{pn} neither sheikh\textsc{.m} remain\textsc{.pst.ptcp.m=cop.3sg.m:S} nor mullah\textsc{.m} remain\textsc{.pst.ptcp.m=cop.3sg.m:S} nor person\textsc{.m-indf} \textsc{exist-3sg.m:S} by good cry\textsc{.m=3sg:R} take\textsc{.prs.sbjv-2sg:A} to \\ 
\glt `However, nowadays, I swear on the Quran, neither a [true] sheikh nor a [true] mullah remains, nor is there a person with whom one can take counsel in peace.'
\z 
 
\ea \label{ZB.61}
\textit{haɫîta?} \\ 
\gll haɫî=t=a \\ 
 understood\textsc{=2sg:NC=cop.3sg.m:S} \\ 
\glt `You see?'
\z 
