\ea \label{JP.1}
\textit{namê heqetîneş sey misefan.} \\ 
\gll namê-(î) heqetîne=ş sey misefa=n \\ 
 name\textsc{.f-ez.attr} real\textsc{=3sg:PSR} \textsc{pn} \textsc{pn=cop.3sg.m:S} \\ 
\glt `His real name was Say Mustafa.'
\z 
 
\ea \label{JP.2}
\textit{namê wêş pîr şelîyare sey misefan.} \\ 
\gll namê wê=ş pîr şelîyar-e sey misefa=n \\ 
 name\textsc{.f} \textsc{reflx=3sg:PSR} \textsc{pn} \textsc{pn-def} \textsc{pn} \textsc{pn=cop.3sg.m:S} \\ 
\glt `Pir Shaliyar's name was [initially] Say Mustafa.'
\z 
 
\ea \label{JP.3}
\textit{maçaş pîr şelîyar, yanê luwan pêwaw xwaywe.} \\ 
\gll m-aç-a=ş pîr şelîyar yanê luwa=n pêwa-û xwa=we \\ 
 \textsc{ind-}say\textsc{.prs-3pl:A=3sg:O} \textsc{pn} \textsc{pn} that\_means go\textsc{.pst.ptcp.m=cop.3sg.m:S} welcoming\textsc{.m-ez.gen} God\textsc{.m=post} \\ 
\glt `He is referred to as Pir Shaliyar, meaning that he has gone [and] embraced God.'
\z 
 
\ea \label{JP.4}
\textit{pewkîma leqemşa miđa maçaş pîr şelîyar.} \\ 
\gll pewkî=ma leqem=şa mi-đ(e)-a m-aç-a=ş pîr şelîyar \\ 
 that\_is\_why\textsc{=1pl:PSR} title\textsc{.m=3pl:R} \textsc{ind-}give\textsc{.prs-3pl:A} \textsc{ind-}say\textsc{.prs-3pl:A=3sg:O} \textsc{pn} \textsc{pn} \\ 
\glt `That is why he's been given this title [and] is called Pir Shaliyar.'
\z 
 
\ea \label{JP.5}
\textit{tateş namêş pîr bayêɫa.} \\ 
\gll tate=ş namê=ş pîr bayêɫ=a \\ 
 father\textsc{.m=3sg:PSR} name\textsc{.f=3sg:PSR} \textsc{pn} \textsc{pn=cop.3sg.m:S} \\ 
\glt `His father's name\footnote{Lit. his father, his name} was Pir Bayel.'
\z 
 
\ea \label{JP.6}
\textit{anen îse îna law hanew degε.} \\ 
\gll ane=n îse îna-∅ la-û hane-û degε \\ 
 \textsc{dem.dist.m.3sg.dir=cop.3sg.m:S} now \textsc{deic-3sg.m:S} next\_to\textsc{-ez.gen} water\_spring\textsc{.m-ez.gen} village\textsc{.f.sg.obl} \\ 
\glt `He is the one [whose shrine] is next to the spring of the village.'
\z 

\ea \label{JP.7}
\textit{ane tateşa.} \\ 
\gll ane tate=ş=a \\ 
 \textsc{dem.dist.m.3sg.dir} father\textsc{.m=3sg:PSR=cop.3sg.m:S} \\ 
\glt `That is his father.'
\z 
 
\ea \label{JP.8}
\textit{tateş fewt kero. pîr bayêɫ fewt kero.} \\ 
\gll tate=ş fewt ker-o pîr bayêɫ fewt ker-o \\ 
 father\textsc{.m=3sg:PSR} death\textsc{.m} do\textsc{.prs.ind-3sg:A} \textsc{pn} \textsc{pn} death\textsc{.m} do\textsc{.prs.ind-3sg:A} \\ 
\glt `His father passed away. Pir Bayel passed away.'
\z 
 
\ea \label{JP.9}
\textit{hetîm gino.} \\ 
\gll hetîm gin-o \\ 
 orphan fall\textsc{.prs.ind-3sg:S} \\ 
\glt `He was left an orphan.'
\z 
 
\ea \label{JP.10}
\textit{pîr şelîyar hetîm bo.} \\ 
\gll pîr şelîyar hetîm b-o \\ 
 \textsc{pn} \textsc{pn} orphan be\textsc{.prs.ind-3sg:S} \\ 
\glt `Pir Shaliyar was an orphan.'
\z 
 
\ea \label{JP.11}
\textit{milo law mamoyş, mamo yosoy.} \\ 
\gll mi-l-o la-û mamo-î=ş mamo yoso-î \\ 
 \textsc{ind-}go\textsc{.prs-3sg:S} to\textsc{-ez.gen} paternal\_uncle\textsc{-obl.m=3sg:PSR} paternal\_uncle \textsc{pn-m.sg.obl} \\ 
\glt `He went to his uncle, Uncle Yoso.'
\z 
 
\ea \label{JP.12}
\textit{mamo yoso, a pîr heme yoso mamoşa.} \\ 
\gll mamo yoso a pîr heme yoso mamo=ş=a \\ 
 paternal\_uncle \textsc{pn} \textsc{dem.dist} \textsc{pn} \textsc{pn} \textsc{pn} paternal\_uncle\textsc{=3sg:PSR=cop.3sg.m:S} \\ 
\glt `Uncle Yoso, [otherwise called] Pir Hama Yoso, was his uncle.'
\z 
 
\ea \label{JP.13}
\textit{êtir milo law ađî.} \\ 
\gll êtir mi-l-o la-û ađî \\ 
 \textsc{disc.ptcl} \textsc{ind-}go\textsc{.prs-3sg:S} to\textsc{-ez.gen} \textsc{3sg.obl.m} \\ 
\glt `Anyhow, he went to him.'
\z 
 
\ea \label{JP.14}
\textit{zaroɫe bo hetîm bo milo la ađî.} \\ 
\gll zaroɫe b-o hetîm b-o mi-l-o la ađî \\ 
 child be\textsc{.prs.ind-3sg:S} orphan be\textsc{.prs.ind-3sg:S} \textsc{ind-}go\textsc{.prs-3sg:S} to \textsc{3sg.obl.m} \\ 
\glt `He was a child. He was an orphan. He went to him (his uncle).'
\z 
 
\ea \label{JP.15}
\textit{miqdarê yawo pene.} \\ 
\gll miqdarê yaw-o pene \\ 
 a\_little arrive\textsc{.prs.ind-3sg:S} to \\ 
\glt `He grew up a bit.'
\z 
 
\ea \label{JP.16}
\textit{tawo şuwanew bizeɫa kero. tawo şuwanew heywanî kero.} \\ 
\gll taw-o şuwane-û bizeɫa k\'er-o taw-o şuwane-û heywan-î {} k\'er-o \\
 can\textsc{.prs.ind-3sg:A} shepherd\textsc{.m-ez.gen} goat\textsc{.dim.pl.obl} do\textsc{.prs.sbjv-3sg:A} can\textsc{.prs.ind-3sg:A} shepherd\textsc{.m-ez.gen} animal\textsc{.m-sg.obl} {} do\textsc{.prs.sbjv-3sg:A} \\ 
\glt `He was able to tend goats. He was able to tend livestock.'
\z 
 
\ea \label{JP.17}
\textit{şuwanew heywanî bo.} \\ 
\gll şuwane-û heywan-î b-o \\ 
 shepherd\textsc{.m-ez.gen} animal\textsc{.m-sg.obl} be\textsc{.prs.ind-3sg:S} \\ 
\glt `He was a shepherd.'
\z 
 
\ea \label{JP.18}
\textit{kîyanaş maça, ``heywanekey bere''} \\ 
\gll kîyan-a=ş m-aç-a heywan-ekey b\'er-e \\
 send\textsc{.prs.ind-3pl:A=3sg:O} \textsc{ind-}say\textsc{.prs-3pl:A} animal\textsc{.m-def.m.sg.obl} take\textsc{.prs.imp-2sg:A} \\ 
\glt `They (his uncle's family) sent him [shepherding]. They said, ``Take the animals ....'''
\z 
 
\ea \label{JP.19}
\textit{ana ew dîmû qeɫaw kitay.} \\ 
\gll an(e)=a ew dîm-û qeɫa-û kita-î \\ 
 \textsc{dem.dist.m.3sg.dir=deic} \textsc{dem.dist} side\textsc{.m-ez.gen} castle\textsc{.f-ez.gen} \textsc{pn-m.sg.obl} \\ 
\glt `-- Look, it is [located] on the other side of Kita castle.'
\z 
 
\ea \label{JP.20}
\textit{yagê hen maçaş nehaɫû pîr şelîyara} \\ 
\gll yag(ê)-ê hen-∅ m-aç-a=ş nehaɫ-û pîr şelîyar-a \\ 
 place\textsc{.f-indf} \textsc{exist-3sg.m:S} \textsc{ind-}say\textsc{.prs-3pl:A=3sg:O} valley\textsc{-ez.gen} \textsc{pn} \textsc{pn-pl.obl} \\ 
\glt `There is a place called the valley of Pir Shaliyar.'
\z 
 
\ea \label{JP.21}
\textit{``heywanekey bere pey ewa.''} \\ 
\gll heywan-ekey b\'er-e pey ewa \\
 animal\textsc{.m-def.m.sg.obl} take\textsc{.prs.imp-2sg:A} to there \\ 
\glt ``Take the animals there [the valley of Pir Shaliyar].''
\z 
 
\ea \label{JP.22}
\textit{heremeşû pîr heme yoso kîyanaş çenû biza pey ewagey} \\ 
\gll hereme=ş=û pîr heme yoso kîyan-a=ş çenû biz(e)-a pey ewage-î \\ 
 wife\textsc{.f=3sg:PSR}=and \textsc{pn} \textsc{pn} \textsc{pn} send\textsc{.prs.ind-3pl:A=3sg:O} with\textsc{.ez.gen} goat\textsc{.f-pl.obl} to there\textsc{-m.sg.obl} \\ 
\glt `Pir Hama Yoso and his wife sent him there [as a shepherd] together with the goats.'
\z 
 
\ea \label{JP.23}
\textit{milo mêwe.} \\ 
\gll mi-l-o m-ê=we \\ 
 \textsc{ind-}go\textsc{.prs-3sg:S} \textsc{ind-}come\textsc{.prs.3sg:S=compl} \\ 
\glt `He went [shepherding and] came back [home].'
\z 
 
\ea \label{JP.24}
\textit{milo mêwew maça, ``hizmêç bare.''} \\ 
\gll mi-l-o m-ê=we=û m-aç-a hizmê=ç b-ar-e \\ 
 \textsc{ind-}go\textsc{.prs-3sg:S} \textsc{ind-}come\textsc{.prs.3sg:S=compl}=and \textsc{ind-}say\textsc{.prs-3pl:A} firewood\textsc{.f.pl.dir=add} \textsc{imp-}bring\textsc{.prs-2sg:A} \\ 
\glt `He went [shepherding and] came back [home]. They said, ``Bring firewood as well.'''
\z 
 
\ea \label{JP.25}
\textit{derdeyçiş hêzmê maro.} \\ 
\gll derdey=ç=iş hêzm(î)-ê m-ar-o \\ 
 load\textsc{.m=add=3sg:R} firewood\textsc{-f.sg.obl} \textsc{ind-}bring\textsc{.prs-3sg:A} \\ 
\glt `He brought [home] a load of firewood on his shoulders.'
\z 
 
\ea \label{JP.26}
\textit{îmtihaniş kera.} \\ 
\gll îmtihan=iş ker-a \\ 
 test\textsc{.m=3sg:O} do\textsc{.prs.ind-3pl:A} \\ 
\glt `They were testing him.'
\z 
 
\ea \label{JP.27}
\textit{maça, ``dey to metawî --sey misefaw-- kuçêş yewê berî nehaɫû pîr şelîyara şanîşa pijgînîşa ba bizya.''} \\ 
\gll m-aç-a dey to me-taw-î sey misefa=û kuç-ê=ş yew(e)-ê b\'er-î nehaɫ-û pîr şelîyar-a ş\'an-î=şa p\'ijgîn-î=şa ba bi-zy(e)-a \\
 \textsc{ind-}say\textsc{.prs-3pl:A} \textsc{disc.ptcl} \textsc{2sg} \textsc{neg.ind-}can\textsc{.prs-2sg:A} \textsc{pn} \textsc{pn}=and little\textsc{-indf=3sg:PSR} barley\textsc{-f.sg.obl} take\textsc{.prs.sbjv-2sg:A} valley\textsc{-ez.gen} \textsc{pn} \textsc{pn-pl.obl} throw\textsc{.prs.sbjv-2sg:A=3pl:O} scatter\textsc{.prs.sbjv-2sg:A=3pl:O} \textsc{hort} \textsc{sbjv-}grow\textsc{.prs-3pl:S} \\ 
\glt `They said, ``You are not able, Say Mustafa, to take some barley seeds to the valley of Pir Shaliyar, [and] scatter them [on the field], [and] plant them, so that they grow.'''
\z 
 
\ea \label{JP.28}
\textit{êtir xo ađ meselen hetîm bo. law mamoyşa bo. řûweş mebo waço, ``ya metawû ya tawû.''} \\ 
\gll êtir xo ađ meselen hetîm b-o la-û mamo-î=şa b-o řûwe=ş me-b-o w\'aç-o ya me-taw-û ya taw-û \\
 \textsc{disc.ptcl} \textsc{disc.ptcl} \textsc{3sg.m.dir} for\_example orphan be\textsc{.prs.ind-3sg:S} at\_the\_place\_of\textsc{-ez.gen} paternal\_uncle\textsc{-obl.m=3pl:PSR} be\textsc{.prs.ind-3sg:S} face\textsc{=3sg:NC} \textsc{neg.ind-}be\textsc{.prs-3sg:S} say\textsc{.prs.sbjv-3sg:A} either \textsc{neg.ind-}can\textsc{.prs-1sg:A} or can\textsc{.prs.ind-1sg:A} \\ 
\glt `Well, he was an orphan. He was at his uncle's. He was ashamed\footnote{Lit. He didn't have a face.} to say, ``I cannot'', or ``I can.'''
\z 
 
\ea \label{JP.29}
\textit{werû mecbûrî yewayç bero.} \\ 
\gll wer-û mecbûrî-(î) yew(e)-a=îç {} ber-o \\ 
 out\_of\textsc{-ez.gen} obligation\textsc{-m.sg.obl} barley\textsc{.f-pl.obl=add} {} take\textsc{.prs.ind-3sg:A} \\ 
\glt `Out of obligation, he took the barley seeds, too.'
\z 
 
\ea \label{JP.30}
\textit{zengene wişɫalê bero.} \\ 
\gll zengen(e)-e wişɫal(e)-ê ber-o \\ 
 hoe\textsc{.m-ez.cmpd} small\textsc{-indf} take\textsc{.prs.ind-3sg:A} \\ 
\glt `He took a small hoe, too.'
\z 
 
\ea \label{JP.31}
\textit{milo ew dîmû qeɫaw kita.} \\ 
\gll mi-l-o ew dîm-û qeɫa-û kita \\ 
 \textsc{ind-}go\textsc{.prs-3sg:S} \textsc{dem.dist} side\textsc{.m-ez.gen} castle\textsc{.f-ez.gen} \textsc{pn} \\ 
\glt `He went to the other side of Kita castle.'
\z 
 
\ea \label{JP.32}
\textit{yewaşê tomekey şano zemînne yeweka.} \\ 
\gll yewaşê tom-ekey şan-o zemîn=ne yewe-(e)ka \\ 
 then seed\textsc{.m-def.m.sg.obl} scatter\textsc{.prs.ind-3sg:A} land\textsc{.m=post} barley\textsc{.f-def.pl.obl} \\ 
\glt `Then, he scattered the seeds on the land, the barley seeds.'
\z 
 
\ea \label{JP.33}
\textit{wêş lûlejen bo.} \\ 
\gll wê=ş lûlejen b-o \\ 
 \textsc{reflx=3sg:PSR} flute\_player be\textsc{.prs.ind-3sg:S} \\ 
\glt `He was a flute player.'
\z 
 
\ea \label{JP.34}
\textit{zengenekey minyowe payînû zemînekey.} \\ 
\gll zengene-(e)key mi-ny(e)-o=we payîn-û zemîn-ekey \\ 
 hoe\textsc{.m-def.m.sg.obl} \textsc{ind-}put\textsc{.prs-3sg:A=compl} bottom\textsc{-ez.gen} land\textsc{.m-def.m.sg.obl} \\ 
\glt `He put the hoe at the south end of the land.'
\z 
 
\ea \label{JP.35}
\textit{zengeneke zengenare kero.} \\ 
\gll zengene-(e)ke zengenare ker-o \\ 
 hoe\textsc{.m-def.m.sg.dir} ploughing do\textsc{.prs.ind-3sg:A} \\ 
\glt `The hoe [started to] plough.'
\z 
 
\ea \label{JP.36}
\textit{yeweka kaɫo.} \\ 
\gll yewe-(e)ka kaɫ-o \\ 
 barley\textsc{.f-def.pl.obl} plant\textsc{.prs.ind-3sg:A} \\ 
\glt `He planted the barley seeds.'
\z 
 
\ea \label{JP.37}
\textit{bizekêçiş be lûlekê hurpiřa cîya leweřyayşa.} \\ 
\gll bize-(e)kê=ç=iş be lûle-(e)kê hur-piř-a cîya leweřyay=şa \\ 
 goat\textsc{.f-def.pl.dir=add=3sg:PSR} with flute\textsc{.f-def.f.sg} \textsc{pvb-}jump\textsc{.prs.ind-3pl:S} instead\_of graze\textsc{.inf=3pl:PSR} \\ 
\glt `With the sound of the flute, the goats danced instead of grazing.'
\z 
 
\ea \label{JP.38}
\textit{meselen aro kewçêş kîyana pene.} \\ 
\gll meselen aro kewç(e)-ê=ş kîyan-a pene \\ 
 for\_example today six\_kilos\textsc{-indf=3sg:R} send\textsc{.prs.ind-3pl:A} to \\ 
\glt `For instance, he (the uncle) would send him half a kilo [of barley to sow].'
\z 
 
\ea \label{JP.39}
\textit{milo mêwe maça, ``şanat tomeke?''} \\ 
\gll mi-l-o m-ê=we m-aç-a şana=t tom-eke \\ 
 \textsc{ind-}go\textsc{.prs-3sg:S} \textsc{ind-}come\textsc{.prs.3sg:S=compl} \textsc{ind-}say\textsc{.prs-3pl:A} scatter\textsc{.pst.3sg:O=2sg:A} seed\textsc{-def.m.sg.dir}\\ 
\glt `He would go [to the field] and come back. They would say, ``Did you plant the seeds?'''
\z 
 
\ea \label{JP.40}
\textit{maço, ``erê.''} \\ 
\gll m-aç-o erê \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A} yes \\ 
\glt `He would say, ``Yes.'''
\z 
 
\ea \label{JP.41}
\textit{seba kewçê teriş pene miđa.} \\ 
\gll seba kewç(e)-ê ter=iş pene mi-đ(e)-a \\ 
 tomorrow\textsc{.m} six\_kilos\textsc{-indf} another\textsc{=3sg:R} to \textsc{ind-}give\textsc{.prs-3pl:A} \\ 
\glt `The next day, he (the uncle) would give him another six kilos [of barley seed].'
\z 
 
\ea \label{JP.42}
\textit{meʕmûrê kîyano.} \\ 
\gll meʕmûr-ê kîyan-o \\ 
 officer\textsc{.m-indf} send\textsc{.prs.ind-3sg:A} \\ 
\glt `He (the uncle) sent a servant [to spy on Pir Shaliyar].'
\z 
 
\ea \label{JP.43}
\textit{maço bêdengê mamo yoso, ``da luwe şûnû sey misefayre bizane řas maço!} \\ 
\gll m-aç-o bêdengê mamo yoso da l\'u-e şûn-û sey misefa-î=re bi-z\'an-e řas m-aç-o \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A} secretly paternal\_uncle \textsc{pn} \textsc{hort} go\textsc{.prs.imp-2sg:S} after\textsc{-ez.gen} \textsc{pn} \textsc{pn-m.sg.obl=post} \textsc{imp-}know\textsc{.prs-2sg:A} truth \textsc{ind-}say\textsc{.prs-3sg:A} \\ 
\glt `Uncle Yoso secretly said [to the servant], ``Go after Say Mustafa, see if he is telling the truth!'
\z 
 
\ea \label{JP.44}
\textit{î yewêşe şanεnê pî zemînne?} \\ 
\gll î yew(e)-ê=ş=e şanε=nê p=î zemîn=ne \\ 
 \textsc{dem.prox} barley\textsc{.f-pl.dir=3sg:A=dem} scatter\textsc{.pst.ptcp.pl=cop.3pl:O} at=\textsc{dem.prox} land\textsc{.m=post} \\ 
\glt `Has he planted the seeds in the field?'
\z 
 
\ea \label{JP.45}
\textit{ya î yewêşe zaxêre kerdênê?} \\ 
\gll ya î yew(e)-ê=ş=e zaxêre kerdê=nê \\ 
 or \textsc{dem.prox} barley\textsc{.f-pl.dir=3sg:A=dem} storage do\textsc{.pst.ptcp.pl=cop.3pl:O} \\ 
\glt `Or has he [instead] stored the seeds [for himself]?'
\z 
 
\ea \label{JP.46}
\textit{ađ çenî tawo kewçê yewê şano zemînnew wêrega bêwe?} \\ 
\gll ađ çenî taw-o kewç-ê yew(e)-ê ş\'an-o zemîn=ne=û wêrega b-ê=we \\
 \textsc{3sg.m.dir} how can\textsc{.prs.ind-3sg:A} six\_kilos\textsc{-indf} barley\textsc{-f.sg.obl} scatter\textsc{.prs.sbjv-3sg:A} land\textsc{.m=post}=and evening \textsc{sbjv-}come\textsc{.prs.3sg:S=compl} \\ 
\glt `How can he plant six kilos [of barley] in the field [and] return in the evening,'
\z 
 
\ea \label{JP.47}
\textit{derdey hêzmê koɫêşa,} \\ 
\gll derdey hêzm(î)-ê koɫê=ş=a \\ 
 load\textsc{.m} firewood\textsc{-f.sg.obl} shoulder\textsc{=3sg:PSR=cop.3sg.m:S} \\ 
\glt `[while simultaneously] a load of firewood is on his shoulders,'
\z 
 
\ea \label{JP.48}
\textit{şuwanew bizayç bo?} \\ 
\gll şuwane-û biz(e)-a=îç b-o \\
 shepherd\textsc{.m-ez.gen} goat\textsc{.f-pl.obl=add} be\textsc{.prs.sbjv-3sg:S} \\ 
\glt `And he is shepherding the goats?'''
\z 
 
\ea \label{JP.49}
\textit{îneş netawan kero.''} \\ 
\gll îne=ş ne-tawa=n {} k\'er-o \\
 \textsc{dem.prox.m.3sg.dir=3sg:A} \textsc{neg-}can\textsc{.pst.ptcp.m=cop.3sg.m:O} {} do\textsc{.prs.sbjv-3sg:A} \\ 
\glt `He cannot do these [tasks].'''
\z 
 
\ea \label{JP.50}
\textit{yo kîyano. bê dengê milo weraweriş.} \\ 
\gll yo kîyan-o bê-deng-ê mi-l-o werawer=iş \\ 
 one\textsc{.m} send\textsc{.prs.ind-3sg:A} without-noise\textsc{.m-indf} \textsc{ind-}go\textsc{.prs-3sg:S} opposite\textsc{=3sg:PSR} \\ 
\glt `He sent one [servant] secretly. He (the servant) went to him (Pir Shaliyar).'
\z 
 
\ea \label{JP.51}
\textit{milo weraweriş miđyo epase min wat tomekeş şana zemînekeynew.} \\ 
\gll mi-l-o werawer=iş mi-đy(e)-o e=pase min wat tom-eke=ş şana {} zemîn-ekey=ne=û \\ 
 \textsc{ind-}go\textsc{.prs-3sg:S} opposite\textsc{=3sg:PSR} \textsc{ind-}look\textsc{.prs-3sg:S} \textsc{emph}=like \textsc{1sg} say\textsc{.pst} seed\textsc{.m-def.m.sg.dir=3sg:A} scatter\textsc{.pst.3sg:O} {} land\textsc{.m-def.m.sg.obl=post}=and \\ 
\glt `He went to him [and] saw that --as I said-- he (Pir Shaliyar) scattered the seeds in the field.'
\z 
 
\ea \label{JP.52}
\textit{zengenekeş wistrew pane zengene zemînekey wêş.} \\ 
\gll zengene-(e)ke=ş wist{-\O}=re=we p=ane zengene zemîn-ekey wê=ş \\ 
 hoe\textsc{.m-def.m.sg.dir=3sg:A} throw\textsc{.pst-3sg:O=povb=compl} at=\textsc{dem.dist.m.3sg.dir} hoe\textsc{.m} land\textsc{.m-def.m.sg.obl} \textsc{reflx=3sg:PSR} \\ 
\glt `He put the hoe there, in his field.'
\z 
 
\ea \label{JP.53}
\textit{zengenareş kerdû.} \\ 
\gll zengenare=ş kerd{-\O}=û \\ 
 ploughing\textsc{=3sg:A} do\textsc{.pst-3sg:O}=and \\ 
\glt `The hoe ploughed the field.'
\z 
 
\ea \label{JP.54}
\textit{bizekê ... wêş lûlê jeno.} \\ 
\gll bize-(e)kê wê=ş lûl(e)-ê jen-o \\ 
 goat\textsc{.f-def.pl.dir} \textsc{reflx=3sg:PSR} flute\textsc{-f.sg.obl} play\textsc{.prs.ind-3sg:A} \\ 
\glt `The goats ... he (Pir Shaliyar) played the flute.'
\z 
 
\ea \label{JP.55}
\textit{bizekê cîya leweřyayşa hurpiřa.} \\ 
\gll bize-(e)kê cîya leweřyay=şa hur-piř-a \\ 
 goat\textsc{.f-def.pl.dir} instead\_of graze\textsc{.inf=3pl:PSR} \textsc{pvb-}jump\textsc{.prs.ind-3pl:S} \\ 
\glt `The goats were dancing instead of grazing.'
\z 
 
\ea \label{JP.56}
\textit{bizekê semay kera hurpiřa.} \\ 
\gll bize-(e)kê sema-î ker-a hur-piř-a \\ 
 goat\textsc{.f-def.pl.dir} dance\textsc{.m-sg.obl} do\textsc{.prs.ind-3pl:A} \textsc{pvb-}jump\textsc{.prs.ind-3pl:S} \\ 
\glt `The goats were dancing.'
\z 
 
\ea \label{JP.57}
\textit{pîyake mêwe ʕerzû heme yosoy kero maço, ``weɫahî qurban min tûşû ceryanê eçîney amanê.} \\ 
\gll pîya-(e)ke m-ê=we ʕerz-û heme yoso-î ker-o m-aç-o weɫahî qurban min tûş-û ceryan-ê e=çîne-î ama-(a)nê \\ 
 man\textsc{.m-def.m.sg.dir} \textsc{ind-}come\textsc{.prs.3sg:S=compl} petition\textsc{.m-ez.gen} \textsc{pn} \textsc{pn-m.sg.obl} do\textsc{.prs.ind-3sg:A} \textsc{ind-}say\textsc{.prs-3sg:A} indeed sir\textsc{.voc} \textsc{1sg} encounter\textsc{-ez.gen} story\textsc{.m-indf} \textsc{emph}=such\textsc{-m.sg.obl} come\textsc{.pst-1sg:S} \\ 
\glt `The man returned [and said] to Hama Yoso, ``Indeed, Sir! I have witnessed an event like this.'
\z 
 
\ea \label{JP.58}
\textit{min tûşû ceryanê eçîney amanê.''} \\ 
\gll min tûş-û ceryan-ê e=çîne-î ama-(a)nê \\ 
 \textsc{1sg} encounter\textsc{-ez.gen} story\textsc{.m-indf} \textsc{emph}=such\textsc{-m.sg.obl} come\textsc{.pst-1sg:S} \\ 
\glt `I have witnessed an event like this.'''
\z 
 
\ea \label{JP.59}
\textit{bawiřiş pene kero.} \\ 
\gll bawiř=iş pene ker-o \\ 
 belief\textsc{=3sg:R} in do\textsc{.prs.ind-3sg:A} \\ 
\glt `He (Hama Yoso) believed him.'
\z 
 
\ea \label{JP.60}
\textit{maço, ``erê řas maçî.''} \\ 
\gll m-aç-o erê řas m-aç-î \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A} yes truth \textsc{ind-}say\textsc{.prs-2sg:A} \\ 
\glt `He said, ``Yes, you are telling the truth.'''
\z 
 
\ea \label{JP.61}
\textit{mê, mêwe wêrega.} \\ 
\gll m-ê m-ê=we wêrega \\ 
 \textsc{ind-}come\textsc{.prs.3sg:S} \textsc{ind-}come\textsc{.prs.3sg:S=compl} evening \\ 
\glt `He returned in the evening.'
\z 
 
\ea \label{JP.62}
\textit{wêrega pîr şelîyar mêwe. sey misefa wextê mêwe, ʕerzû jenekêş kero maço, ``ba êşew îmti̩hanêş kermê law wêmawe newsinmêş.} \\ 
\gll wêrega pîr şelîyar m-ê=we sey misefa wext-ê m-ê=we ʕerz-û jen(î)-ekê=ş ker-o m-aç-o ba êşew îmti̩han-ê=ş k\'er-mê la-û wê=ma=we ne-wsin-mê=ş \\
 evening \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs.3sg:S=compl} \textsc{pn} \textsc{pn} time\textsc{.m-indf} \textsc{ind-}come\textsc{.prs.3sg:S=compl} petition\textsc{.m-ez.gen} woman\textsc{.f-def.f.sg=3sg:R} do\textsc{.prs.ind-3sg:A} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{hort} tonight test\textsc{.m-indf=3sg:O} do\textsc{.prs.sbjv-1pl:A} next\_to\textsc{-ez.gen} \textsc{reflx=1pl:PSR=post} \textsc{neg.sbjv-}put\_to\_sleep\textsc{.prs-1pl:A=3sg:O} \\ 
\glt `Pir Shaliyar returned in the evening. When Say Mustafa returned, he (Hama Yoso) said to his wife, ``Let's test him tonight; let's not put him to sleep next to us.'
\z 
 
\newpage
\ea \label{JP.63}
\textit{yagê cîyekare bûsinmêş.} \\ 
\gll yag(ê)-ê cîyekar-e b-ûsin-mê=ş \\ 
 place\textsc{.f-indf} separate\textsc{-f} \textsc{sbjv-}put\_to\_sleep\textsc{.prs-1pl:A=3sg:O} \\ 
\glt `We will put him to sleep in a separate place.'
\z 
 
\ea \label{JP.64}
\textit{îmtîhanê teriş kermê bizanmê řasa diron.''} \\ 
\gll îmtîhan-ê ter=iş k\'er-mê bi-zan-mê řas=a diro=n \\
 test\textsc{.m-indf} another\textsc{=3sg:O} do\textsc{.prs.sbjv-1pl:A} \textsc{sbjv-}know\textsc{.prs-1pl:A} true\textsc{=cop.3sg.m:S} lie\textsc{=cop.3sg.m:S} \\ 
\glt `We will test him again, see if it (his supernatural power) is right [or] wrong.'''
\z 
 
\ea \label{JP.65}
\textit{şewê yagêş cîya kerawe. yagê ter mûsnaş.} \\ 
\gll şew(e)-ê yagê=ş cîya ker-a=we yag(ê)-ê ter m-ûsn-a=ş \\ 
 night\textsc{-f.sg.obl} place\textsc{.f=3sg:PSR} separate do\textsc{.prs.ind-3pl:A=compl} place\textsc{.f-indf} other \textsc{ind-}put\_to\_sleep\textsc{.prs-3pl:A=3sg:O} \\ 
\glt `At night, they kept his bed apart [from theirs]. They put him to bed in another place.'
\z 
 
\ea \label{JP.66}
\textit{fire mûsaw kem mûsaw.} \\ 
\gll fire m-ûs-a=û kem m-ûs-a=û \\ 
 a\_lot \textsc{ind-}sleep\textsc{.prs-3pl:S}=and a\_little \textsc{ind-}sleep\textsc{.prs-3pl:S}=and \\ 
\glt `They slept a lot; they slept a little.'
\z 
 
\ea \label{JP.67}
\textit{wurmêzawe bila seriş bila zana ceryan çêşa.} \\ 
\gll wur-m-êz-a=we bi-l-a ser=iş bi-l-a z\'an-a ceryan çêş=a \\
 \textsc{pvb-ind-rise.prs-3pl:S=compl} \textsc{sbjv-}go\textsc{.prs-3pl:S} on\textsc{=3sg:R} \textsc{sbjv-}go\textsc{.prs-3pl:S} know\textsc{.prs.sbjv-3pl:A} story\textsc{.m} what\textsc{=cop.3sg.m:S} \\ 
\glt `They (the uncle and his wife) woke up to go to his bed and went to see what the story was (i.e. what was happening).'
\z 
 
\ea \label{JP.68}
\textit{wextê miđya îna ane ce zemînî hewawe wuten.} \\ 
\gll wext-ê mi-đy(e)-a îna-∅ ane ce zemîn-î hewa=we wute=n \\ 
 time\textsc{.m-indf} \textsc{ind-}look\textsc{.prs-3pl:S} \textsc{deic-3sg.m:S} \textsc{dem.dist.m.3sg.dir} from earth\textsc{.m-sg.obl} air\textsc{=post} sleep\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
 \newpage
\glt `They saw that he (Pir Shaliyar) had levitated and slept aloft in the air at this height.\footnote{Lit. that [much]}\footnote{The narrator indicates the height of the bed with his hand}'
\z 
 
\ea \label{JP.69}
\textit{ane ce zemînî pêse waçî xway ketê pey kîyasen.} \\ 
\gll ane ce zemîn-î pêse w\'aç-î xwa-î ket-ê pey kîyase=n \\
 \textsc{dem.dist.m.3sg.dir} from earth\textsc{.m-sg.obl} as\_if say\textsc{.prs.sbjv-2sg:A} God\textsc{.m-sg.obl} bed\textsc{-indf} to send\textsc{.pst.ptcp.m=cop.3sg.m:R} \\ 
\glt `[He had levitated] this much from the ground, as if God had sent him a bed.'
\z 
 
\ea \label{JP.70}
\textit{ane îna hewawe.} \\ 
\gll ane îna-∅ hewa=we \\ 
 \textsc{dem.dist.m.3sg.dir} \textsc{deic-3sg.m:S} air\textsc{=post} \\ 
\glt `He (his body) was aloft in the air [at this height].'
\z 
 
\ea \label{JP.71}
\textit{pîr heme yoso wêş milo temaşeş kero; miđyo ane îna hewawe. zano îne êtir karî xwayîn.} \\ 
\gll pîr heme yoso wê=ş mi-l-o temaşe=ş ker-o mi-đy(e)-o ane îna-∅ hewa=we zan-o îne êtir kar-î xwayî=n \\ 
 \textsc{pn} \textsc{pn} \textsc{pn} \textsc{reflx=3sg:PSR} \textsc{ind-}go\textsc{.prs-3sg:S} looking\textsc{=3sg:O} do\textsc{.prs.ind-3sg:A} \textsc{ind-}look\textsc{.prs-3sg:S} \textsc{dem.dist.m.3sg.dir} \textsc{deic-3sg.m:S} air\textsc{=post} know\textsc{.prs.ind-3sg:A} \textsc{dem.prox.m.3sg.dir} \textsc{disc.ptcl} task\textsc{.m-ez.attr} divine\textsc{=cop.3sg.m:S} \\ 
\glt `Pir Hama Yoso went [and] looked at him; he noticed that he (Pir Shaliyar) was aloft in the air [at a certain height]. [And] he came to realise that that was God's work.'
\z 
 
\ea \label{JP.72}
\textit{ʕerzû heremêş kero, fermawo be heremêş, ``weɫahî êtir ême metawmê, sey misefa metawmê hîç karêş pene kermê îla kîyanmêş bilo weno, bilo kîyanmêş weno.''} \\ 
\gll ʕerz-û herem(e)-ê=ş ker-o fermaw-o be herem(e)-ê=ş weɫahî êtir ême me-taw-mê sey misefa me-taw-mê hîç kar-ê=ş pene k\'er-mê îla kîy\'an-mê=ş bi-l-o w\'en-o bi-l-o kîy\'an-mê=ş w\'en-o \\
 petition\textsc{.m-ez.gen} wife\textsc{-f.sg.obl=3sg:PSR} do\textsc{.prs.ind-3sg:A} say\textsc{.prs.ind-3sg:A} to wife\textsc{-f.sg.obl=3sg:PSR} indeed \textsc{disc.ptcl} \textsc{1pl} \textsc{neg.ind-}can\textsc{.prs-1pl:A} \textsc{pn} \textsc{pn} \textsc{neg.ind-}can\textsc{.prs-1pl:A} nothing task\textsc{.m-indf=3sg:R} to do\textsc{.prs.sbjv-1pl:A} except send\textsc{.prs.sbjv-1pl:A=3sg:O} \textsc{sbjv-}go\textsc{.prs-3sg:S} read\textsc{.prs.sbjv-3sg:A} \textsc{sbjv-}go\textsc{.prs-3sg:S} send\textsc{.prs.sbjv-1pl:A=3sg:O} read\textsc{.prs.sbjv-3sg:A} \\ 
\glt `He (Pir Hama Yoso) said to his wife, ``Indeed, we can no longer be helpful to Say Mustafa,\footnote{Lit. we can do nothing to him.} except by sending him to study; let us send him to go [and] study.'''
\z 
 
\ea \label{JP.73}
\textit{haɫîyeş kero.} \\ 
\gll haɫî-e=ş ker-o \\ 
 understood\textsc{-f=3sg:O} do\textsc{.prs.ind-3sg:A} \\ 
\glt `He explained [it] to her.'
\z 
 
\ea \label{JP.74}
\textit{maço, ``min înem dî be sey misefaywe.} \\ 
\gll m-aç-o min îne=m dî-∅ be sey misefa-î=we \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{1sg} \textsc{dem.prox.m.3sg.dir=1sg:A} see\textsc{.pst-3sg.m:O} about \textsc{pn} \textsc{pn-m.sg.obl=post} \\ 
\glt `He said, ``I witnessed this concerning Say Mustafa.'
\z 
 
\ea \label{JP.75}
\textit{êtir mebo ême karêş pene kermê.} \\ 
\gll êtir me-b-o ême kar-ê=ş pene k\'er-mê \\
 no\_longer \textsc{neg.ind-}be\textsc{.prs-3sg:S} \textsc{1pl} task\textsc{.m-indf=3sg:R} to do\textsc{.prs.sbjv-1pl:A} \\ 
\glt `We can no longer do anything for him.'
\z 
 
\ea \label{JP.76}
\textit{çemim pene kotû. řasaw.} \\ 
\gll çem=im pene kot=û řas=a=û \\ 
 eye\textsc{=1sg:NC} to fall\textsc{.pst.3sg.m:S/R}=and true\textsc{=cop.3sg.m:S}=and \\ 
\glt `I caught sight of him. It's true [that he has supernatural powers].'
\z 
 
\newpage
\ea \label{JP.77}
\textit{û pîyayçe řasiş wat ûneş gêɫnawew.} \\ 
\gll û pîya=îç=e řas=iş wat ûne=ş gêɫna=we=û \\ 
 \textsc{dem.dist} man\textsc{.m=add=dem} truth\textsc{=3sg:A} say\textsc{.pst} \textsc{dem.dist=3sg:A} narrate\textsc{.pst.3sg:O=post}=and \\ 
\glt `And the man (the servant) told the truth when he said that [those words].'
\z 
 
\ea \label{JP.78}
\textit{meger kîyanmêş bilo weno be feqîgerî.''} \\ 
\gll meger kîy\'an-mê=ş bi-l-o w\'en-o be feqîgerî \\
 if\_only send\textsc{.prs.sbjv-1pl:A=3sg:O} \textsc{sbjv-}go\textsc{.prs-3sg:S} read\textsc{.prs.sbjv-3sg:A} about Islamic\_jurisprudence \\ 
\glt `If only we could send him to study Islamic jurisprudence.'''
\z 
 
\ea \label{JP.79}
\textit{kîyanaş weno. miqdarê weno.} \\ 
\gll kîyan-a=ş wen-o miqdarê wen-o \\ 
 send\textsc{.prs.ind-3pl:A=3sg:O} read\textsc{.prs.ind-3sg:A} a\_little read\textsc{.prs.ind-3sg:A} \\ 
\glt `He (the uncle) sent him (Pir Shaliyar) [and] he studied. He studied a little.'
\z 
 
\ea \label{JP.80}
\textit{ce her yagê weno, ça dimayê wextê miđya ênneş wena qabilyetiş peya kerd kîyanaş pey beẍay pey xizmetû, kîyanaş pey beẍay wenay.} \\ 
\gll ce her yag(ê)-ê wen-o ç=a dima-ê wext-ê mi-đy(e)-a ênne=ş wena qabilyet=iş peya kerd kîyan-a=ş pey beẍa-î pey xizmet-û kîyan-a=ş pey beẍa-î wenay \\ 
 from \textsc{emph} place\textsc{.f-indf} read\textsc{.prs.ind-3sg:A} in=\textsc{dem.dist} afterwards\textsc{-obl.f} time\textsc{.m-indf} \textsc{ind-}look\textsc{.prs-3pl:S} so\_much\textsc{=3sg:A} read\textsc{.pst} capability\textsc{.m=3sg:A} visible do\textsc{.pst} send\textsc{.prs.ind-3pl:A=3sg:O} to \textsc{pn-m.sg.obl} to service\textsc{.m-ez.gen} send\textsc{.prs.ind-3pl:A=3sg:O} to \textsc{pn-m.sg.obl} read\textsc{.inf} \\ 
\glt `Wherever he studied, after a while, they realised that he studied so well that he deserved to be sent to Baghdad into the presence of His Highness [Ghaws]. They sent him to Baghdad so that he could read.'
\z 
 
\ea \label{JP.81}
\textit{milo miqdarê weno.} \\ 
\gll mi-l-o miqdarê wen-o \\ 
 \textsc{ind-}go\textsc{.prs-3sg:S} a\_little read\textsc{.prs.ind-3sg:A} \\ 
\glt `He went [to Baghdad and] studied for a while.'
\z 
 
\ea \label{JP.82}
\textit{şûnû miqdarêre hezretû xosî fermawo duwê yerê îmtihanêş kero} \\ 
\gll şûn-û miqdarê=re hezret-û xos-î fermaw-o duwê yerê îmtihan-ê=ş ker-o \\ 
 after\textsc{-ez.gen} a\_little\textsc{=post} his\_highness\textsc{.m-ez.gen} \textsc{pn-m.sg.obl} say\textsc{.prs.ind-3sg:A} two three test\textsc{.m-pl.dir=3sg:O} do\textsc{.prs.ind-3sg:A} \\ 
\glt `After a while, His Highness Ghaws said -- He tested him two, three times.'
\z 
 
\ea \label{JP.83}
\textit{îmtihanêş kero, fermawo, \textsuperscript{A} exî mistefa recilun kamilun. \textsuperscript{A} yanê be ʕerebî wêşa paseş fermawan hezretû xosî.} \\ 
\gll îmtihan-ê=ş ker-o fermaw-o exî mistefa recilun kamilun yanê be ʕerebî wê=şa pase=ş fermawa=n hezret-û xos-î \\ 
 test\textsc{.m-pl.dir=3sg:O} do\textsc{.prs.ind-3sg:A} say\textsc{.prs.ind-3sg:A} my\_brother \textsc{pn} man complete that\_means in Arabic \textsc{reflx=3pl:PSR} such\textsc{=3sg:A} say\textsc{.pst.ptcp.m=cop.3sg.m:O} his\_highness\textsc{.m-ez.gen} \textsc{pn-m.sg.obl} \\ 
\glt `He (his Highness Ghaws) tested him [and] said, ``My brother Mustafa is a well-rounded man', which His Highness Ghaws said in this manner in his\footnote{Lit. their} Arabic.'
\z 
 
\ea \label{JP.84}
\textit{misefa yawan pene tawo ce şerîʕetîç îstîfade kero ce terêqetîç.} \\ 
\gll misefa yawa=n pene taw-o ce şerîʕet-î=ç îstîfade k\'er-o ce terêqet-î=ç \\
 \textsc{pn} arrive\textsc{.pst.ptcp.m=cop.3sg.m:S} to can\textsc{.prs.ind-3sg:A} from sharia\_law\textsc{-obl.m=add} use do\textsc{.prs.sbjv-3sg:A} from denomination\textsc{-obl.m=add} \\ 
\glt `Mustafa has become [knowledgeable] enough that he can both practise Sharia law and lead an order (of dervishes).'
\z 
 
 \newpage
\ea \label{JP.85}
\textit{yawan pene.} \\ 
\gll yawa=n pene \\ 
 arrive\textsc{.pst.ptcp.m=cop.3sg.m:S} to \\ 
\glt `He has mastered [Islamic jurisprudence].'
\z 
 
\ea \label{JP.86}
\textit{\textsuperscript{A} exî mistefa recilun kamilun. \textsuperscript{A}} \\ 
\gll exî mistefa recilun kamilun \\ 
 my\_brother \textsc{pn} man complete \\ 
\glt `My brother Mustafa is a well-rounded man.'
\z 
 
\ea \label{JP.87}
\textit{mişyo bilowe pey menteqew wêşa xizmet be merđimî kero.} \\ 
\gll mişyo bi-l-o=we pey menteqe-û wê=şa xizmet be merđim-î k\'er-o \\
 \textsc{aux} \textsc{sbjv-}go\textsc{.prs-3sg:S=compl} to region\textsc{-ez.gen} \textsc{reflx=3pl:PSR} service\textsc{.m} by people\textsc{.m-sg.obl} do\textsc{.prs.sbjv-3sg:A} \\ 
\glt `He should go back to his region [and] and be of service to people,'
\z 
 
\ea \label{JP.88}
\textit{çûn ce şerêʕetîçnew ce terêqetîçne yawan pene.} \\ 
\gll çûn ce şerêʕet-î=ç=ne=û ce terêqet-î=ç=ne yawa=n pene \\ 
 because from sharia\_law\textsc{-obl.m=add=post}=and from denomination\textsc{-obl.m=add=post} arrive\textsc{.pst.ptcp.m=cop.3sg.m:S} to \\ 
\glt `because he has both mastered Sharia law and [is capable of leading an] order (of dervishes).'
\z 
 
\ea \label{JP.89}
\textit{desûriş pene miđo.} \\ 
\gll desûr=iş pene mi-đ(e)-o \\ 
 task/order\textsc{.m=3sg:R} to \textsc{ind-}give\textsc{.prs-3sg:A} \\ 
\glt `He (Ghaws) gave him (Pir Shaliyar) an order.'
\z 
 
\ea \label{JP.90}
\textit{îcazeş pene miđo.} \\ 
\gll îcaze=ş pene mi-đ(e)-o \\ 
 permission\textsc{=3sg:R} to \textsc{ind-}give\textsc{.prs-3sg:A} \\ 
\glt `He gave him permission.'
\z 
 
\ea \label{JP.91}
\textit{fermawo hezretû xos, ``dey to mişyo teşrîf berî bilîwe, pey menteqew wêtaw.} \\ 
\gll fermaw-o hezret-û xos dey to mişyo teşrîf b\'er-î bi-l-î=we pey menteqe-û wê=ta=û \\
 say\textsc{.prs.ind-3sg:A} his\_highness\textsc{.m-ez.gen} \textsc{pn} \textsc{disc.ptcl} \textsc{2sg} \textsc{aux} honouring take\textsc{.prs.sbjv-2sg:A} \textsc{sbjv-}go\textsc{.prs-2sg:S=compl} to region\textsc{-ez.gen} \textsc{reflx=2pl:PSR}=and \\ 
\glt `His Highness Ghaws said, ``You should leave [here and] return to your region.'
\z 
 
\ea \label{JP.92}
\textit{bilî çay xizmetû mirîđû mensûb weş kerîyû.} \\ 
\gll bi-l-î ça-î xizmet-û mirîđ=û mensûb weş k\'er-î=û \\
 \textsc{sbjv-}go\textsc{.prs-2sg:S} there\textsc{-m.sg.obl} service\textsc{.m-ez.gen} adherent=and appointees well do\textsc{.prs.sbjv-2sg:A}=and \\ 
\glt `You should go there and be of good service to adherents and appointees.'
\z 
 
\ea \label{JP.93}
\textit{bilîwe pey a menteqew wêtaw, şerʕî ders waçîyû} \\ 
\gll bi-l-î=we pey a menteqe-û wê=ta=û şerʕî ders w\'aç-î=û \\
 \textsc{sbjv-}go\textsc{.prs-2sg:S=compl} to \textsc{dem.dist} region\textsc{-ez.gen} \textsc{reflx=2pl:PSR}=and sharia lesson\textsc{.m} say\textsc{.prs.sbjv-2sg:A}=and \\ 
\glt `Go to your region [and] teach Sharia law.'
\z 
 
\ea \label{JP.94}
\textit{hem şerêʕetû hem terêqet to amadenî tekmêɫnî.} \\ 
\gll hem şerêʕet=û hem terêqet to amade=nî tekmêɫ=nî \\ 
 both sharia\_law=and and denomination \textsc{2sg} ready\textsc{=cop.2sg:S} complete\textsc{=cop.2sg:S} \\ 
\glt `You are ready both [to practise] Sharia law and [to lead] an order (of dervishes); you are complete.'
\z 
 
\ea \label{JP.95}
\textit{luwewe!''} \\ 
\gll l\'u-e=we \\
 go\textsc{.prs.imp-2sg:S=compl} \\ 
\glt `Go back!'''
\z 
 
\ea \label{JP.96}
\textit{kîyanoşewe. mêwe.} \\ 
\gll kîyan-o=ş=ewe m-ê=we \\ 
 send\textsc{.prs.ind-3sg:A=3sg:O=compl} \textsc{ind-}come\textsc{.prs.3sg:S=compl} \\ 
\glt `He sent him back. He returned [to Hewraman].'
\z 
 
 \newpage
\ea \label{JP.97}
\textit{řane mênê pane ya guzerêwene mesîrêwene çenû hecîcîyê gina pêwere.} \\ 
\gll řa=ne m-ê-nê p=ane ya guzer-êwe=ne mesîr-êwe=ne çenû hecîc-î-ê gin-a pêwere \\ 
 road\textsc{.f=post} \textsc{ind-}come\textsc{.prs-3pl:S} at=\textsc{dem.dist.m.3sg.dir} or path\textsc{-indf=post} path\textsc{-indf=post} with\textsc{.ez.gen} \textsc{pn-adjzr-indf} fall\textsc{.prs.ind-3pl:S} together \\ 
\glt `On the road, on a path, he ran into a person from Hajij.'
\z 
 
\ea \label{JP.98}
\textit{êtir herçî mênê a zemane wesîlew maşînû çîwî mebo.} \\ 
\gll êtir herçî m-ê-nê a zeman=e wesîle=û maşîn=û çîw-î me-b-o \\ 
 \textsc{disc.ptcl} what\_ever \textsc{ind-}come\textsc{.prs-3pl:S} \textsc{dem.dist} time\textsc{.m=dem} tool\textsc{.m}=and car\textsc{.m}=and thing\textsc{-m.sg.obl} \textsc{neg.ind-}be\textsc{.prs-3sg:S} \\ 
\glt `Well, in that period, there were no means of transportation, cars, etc.'
\z 
 
\ea \label{JP.99}
\textit{wesîlew suwarî mebo be peyê mênê řare.} \\ 
\gll wesîle=û suwarî me-b-o be peyê m-ê-nê řa=re \\ 
 tool\textsc{.m}=and driving \textsc{neg.ind-}be\textsc{.prs-3sg:S} by foot\textsc{.obl} \textsc{ind-}come\textsc{.prs-3pl:S} road\textsc{.f=post} \\ 
\glt `There were no means of transportation. They (people) would set off on the road on foot.'
\z 
 
\ea \label{JP.100}
\textit{her mênê mênê. mênê nizîkû hecîcî ginawe.} \\ 
\gll her m-ê-nê m-ê-nê m-ê-nê nizîk-û hecîc-î gin-a=we \\ 
 \textsc{emph} \textsc{ind-}come\textsc{.prs-3pl:S} \textsc{ind-}come\textsc{.prs-3pl:S} \textsc{ind-}come\textsc{.prs-3pl:S} close\textsc{-ez.gen} \textsc{pn-m.sg.obl} fall\textsc{.prs.ind-3pl:S=post} \\ 
\glt `They kept walking. They arrived [at a place] close to Hajij.'
\z 
 
\ea \label{JP.101}
\textit{nizîkû hecîcî ginawe. miqdarê zeřşa pene bo.} \\ 
\gll nizîk-û hecîc-î gin-a=we miqdarê zeř=şa pene b-o \\ 
 close\textsc{-ez.gen} \textsc{pn-m.sg.obl} fall\textsc{.prs.ind-3pl:S=compl} a\_little money\textsc{.m=3pl:NC} with be\textsc{.prs.ind-3sg:S} \\ 
\glt `They\footnote{Lit. he} arrived [at a place] close to Hajij. They had some money with them\footnote{Lit. him}.'
\z 
 
\ea \label{JP.102}
\textit{hecîcîyeke maço, ``dey î min îna nizîkû hecîcî kotenawe.} \\ 
\gll hecîc-î-eke m-aç-o dey î min îna-∅ nizîk-û hecîc-î kote=na=we \\ 
 \textsc{pn-adjzr-def.m.sg.dir} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{disc.ptcl} \textsc{dem.prox} \textsc{1sg} \textsc{deic-3sg.m:S} close\textsc{-ez.gen} \textsc{pn-m.sg.obl} fall\textsc{.pst.ptcp.m=cop.1sg:S=compl} \\ 
\glt `The man from Hajij said, ``I have arrived [at a place] close to Hajij.'''
\z 
 
\ea \label{JP.103}
\textit{to luwe zeřekey bere çenû wêt pey xercît lodê.} \\ 
\gll to l\'u-e zeř-ekey b\'er-e çenû wê=t pey xercî=t l\'o-dê \\
 \textsc{2sg} go\textsc{.prs.imp-2sg:S} money\textsc{.m-def.m.sg.obl} take\textsc{.prs.imp-2sg:A} with\textsc{.ez.gen} \textsc{reflx=2sg:PSR} for provision\textsc{=2sg:PSR} go\textsc{.prs.imp-2pl:S} \\ 
\glt `You go, take the money with you for your expenses.'
\z 
 
\ea \label{JP.104}
\textit{zeřekey bere. min, qey mekero.''} \\ 
\gll zeř-ekey b\'er-e min qey me-ker-o \\
 money\textsc{.m-def.m.sg.obl} take\textsc{.prs.imp-2sg:A} \textsc{1sg} limit \textsc{neg.ind-}do\textsc{.prs-3sg:A} \\ 
\glt `Take the money. I will be fine.'''
\z 
 
\ea \label{JP.105}
\textit{ađîç fermawo, ``ne qey mekero ba law towe bo. min xercîm gerek nîya. milûwe.''} \\ 
\gll ađ=îç fermaw-o ne qey me-ker-o ba la-û to=we b-o min xercî=m gerek nîy(e)=a mi-l-û=we \\
 \textsc{3sg.m.dir=add} say\textsc{.prs.ind-3sg:A} no limit \textsc{neg.ind-}do\textsc{.prs-3sg:A} \textsc{hort} with\textsc{-ez.gen} \textsc{2sg=post} be\textsc{.prs.sbjv-3sg:S} \textsc{1sg} provision\textsc{=1sg:NC} necessary \textsc{neg.exist=cop.3sg.m:S} \textsc{ind-}go\textsc{.prs-1sg:S=compl} \\ 
\glt `He (Pir Shaliyar) said, ``No, it is all right! Let it be with you. I do not need provisions. I will go back.'''
\z 
 
\ea \label{JP.106}
\textit{fermawo, ``na na meger berîş.''} \\ 
\gll fermaw-o na na meger b\'er-î=ş \\
 say\textsc{.prs.ind-3sg:A} no no must take\textsc{.prs.sbjv-2sg:A=3sg:O} \\ 
\glt `He (the man from Hajij) said, ``No, you must take it.'''
\z 
 
\ea \label{JP.107}
\textit{be zor miđoş milû sey misefayre zerey.} \\ 
\gll be zor mi-đ(e)-o=ş mil-û sey misefa-î=re zer-e-y \\ 
 by force \textsc{ind-}give\textsc{.prs-3sg:A=3sg:O} on\textsc{-ez.gen} \textsc{pn} \textsc{pn-m.sg.obl=post} money\textsc{.m-def-m.sg.obl} \\ 
\glt `He (the man from Hajij) did not take no for an answer\footnote{Lit. with force} [and] gave him (Pir Shaliyar) the money.'
\z 
 
\ea \label{JP.108}
\textit{miqdarê miđoş milîşre.} \\ 
\gll miqdarê mi-đ(e)-o=ş mil-î=ş=re \\ 
 a\_little \textsc{ind-}give\textsc{.prs-3sg:A=3sg:O} shoulder\textsc{.m-sg.obl=3sg:R=post} \\ 
\glt `He (the man from Hajij) forced him to accept the money.'
\z 
 
\ea \label{JP.109}
\textit{miqdarê mê pêsewe be keramat zeřey kîyanowe pey hecîcîyekey.} \\ 
\gll miqdarê m-ê pês=ewe be keramat zeř-e-î kîyan-o=we pey hecîc-î-ekey \\ 
 a\_little \textsc{ind-}come\textsc{.prs.3sg:S} front\textsc{=post} by supernatural\_power\textsc{.m} money\textsc{.m-def-m.sg.obl} send\textsc{.prs.ind-3sg:A=compl} to \textsc{pn-adjzr-def.m.sg.obl} \\ 
\glt `He (the man from Hajij) went on; he passed by. He (Pir Shaliyar) had sent the money back to him using his supernatural powers.'
\z 
 
\ea \label{JP.110}
\textit{be keramat kîyanoşo zeřeş pey kîyanowew.} \\ 
\gll be keramat kîyan-o=ş=o zeř-e=ş pey kîyan-o=we=û \\ 
 by supernatural\_power\textsc{.m} send\textsc{.prs.ind-3sg:A=3sg:O=compl} money\textsc{.m-def=3sg:R} to send\textsc{.prs.ind-3sg:A=compl}=and \\ 
\glt `Using his supernatural powers, he had sent it back; he had sent back the money.'
\z 
 
\ea \label{JP.111}
\textit{zeřeş gerek mebo.} \\ 
\gll zeř-e=ş gerek me-b-o \\ 
 money\textsc{.m-def=3sg:NC} necessary \textsc{neg.ind-}be\textsc{.prs-3sg:S} \\ 
\glt `He did not need the money.'
\z 
 
 \newpage
\ea \label{JP.112}
\textit{hecîcîyeke çay dima walyore miđyo îna zeřeş penen.} \\ 
\gll hecîc-î-eke ça-î dima waly(e)-o=re mi-đy(e)-o îna-∅ zeř-e=ş pene=n \\ 
 \textsc{pn-adjzr-def.m.sg.dir} there\textsc{-m.sg.obl} afterwards be\_astonished\textsc{.prs-3sg:S=povb} \textsc{ind-}look\textsc{.prs-3sg:S} \textsc{prst} money\textsc{.m-def=3sg:NC} with\textsc{=cop.3sg.m:S} \\ 
\glt `The man from Hajij noticed later. He realised, there it was, he had money with him.'
\z 
 
\ea \label{JP.113}
\textit{miqdarêş ađî be sey misefay ardeniş.} \\ 
\gll miqdarê=ş ađî be sey misefa-î arde=n=iş \\ 
 a\_little\textsc{=3sg:PSR} \textsc{3sg.obl.m} by \textsc{pn} \textsc{pn-m.sg.obl} bring\textsc{.pst.ptcp.m=cop.3sg.m:O=3sg:A} \\ 
\glt `The money which he had made Say Mustafa take.\footnote{Lit. He brought it by Say Mustafa.}'
\z 
 
\ea \label{JP.114}
\textit{miđyo zeřey îna baxeɫişne.} \\ 
\gll mi-đy(e)-o zeř-e-î îna-∅ baxeɫ=iş=ne \\ 
 \textsc{ind-}look\textsc{.prs-3sg:S} money\textsc{.m-def-m.sg.obl} \textsc{deic-3sg.m:S} embrace\textsc{=3sg:PSR=post} \\ 
\glt `He noticed that there was money in his hands.'
\z 
 
\ea \label{JP.115}
\textit{ađî kîyasen dimawe ama.} \\ 
\gll ađî kîyase=n dima=we ama \\ 
 \textsc{3sg.obl.m} send\textsc{.pst.ptcp.m=cop.3sg.m:O} back\textsc{=post} come\textsc{.pst.3sg:S} \\ 
\glt `He (Pir Shaliyar) had sent it back.'
\z 
 
\ea \label{JP.116}
\textit{mêwe êtir. a zemane pîr şelîyar mêwe menteqe. des kero, carê ane minya maçaş meřû asinî} \\ 
\gll m-ê=we êtir a zeman=e pîr şelîyar m-ê=we menteqe des ker-o carê ane mi-ny(e)-a m-aç-a=ş meř(e)-û asin-î \\ 
 \textsc{ind-}come\textsc{.prs.3sg:S=compl} \textsc{disc.ptcl} \textsc{dem.dist} time\textsc{.m=dem} \textsc{pn} \textsc{pn} \textsc{ind-}come\textsc{.prs.3sg:S=compl} region hand\textsc{.m} do\textsc{.prs.ind-3sg:A} once \textsc{dem.dist.m.3sg.dir} \textsc{ind-}put\textsc{.prs-3pl:A} \textsc{ind-}say\textsc{.prs-3pl:A=3sg:O} cave\textsc{.f-ez.gen} iron\textsc{-m.sg.obl} \\ 
\glt `He (Pir Shaliyar) returned [to Hewraman]. Then, Pir Shaliyar returned to the region. He started to build [a thing] called Maru Asini ``Iron cave.'''
\z 
 
\ea \label{JP.117}
\textit{êtir hezretû xosî kîyaseno.} \\ 
\gll êtir hezret-û xos-î kîyase=n=o \\ 
 \textsc{disc.ptcl} his\_highness\textsc{.m-ez.gen} \textsc{pn-m.sg.obl} send\textsc{.pst.ptcp.m=cop.3sg.m:O=compl} \\ 
\glt `Anyhow, His Highness Ghaws had sent him.'
\z 
 
\ea \label{JP.118}
\textit{luwanre meřû asinî çilêş kêştêne.} \\ 
\gll luwa=n=re meř(e)-û asin-î çilê=ş kêştê=ne \\ 
 go\textsc{.pst.ptcp.m=cop.3sg.m:S=povb} cave\textsc{.f-ez.gen} iron\textsc{-m.sg.obl} forty\_days\_fasting\textsc{.f=3sg:A} pull\textsc{.pst.ptcp.f=cop.3sg.f:O} \\ 
\glt `He went to the Maru Asini and fasted for forty days.'
\z 
 
\ea \label{JP.119}
\textit{çane weɫêne luwanre wa çilêş kêştêne.} \\ 
\gll ç=a=ne weɫêne luwa=n=re wa çilê=ş kêştê=ne \\ 
 from=\textsc{dem.dist=post} in\_the\_beginning go\textsc{.pst.ptcp.m=cop.3sg.m:S=povb} there forty\_days\_fasting\textsc{.f.sg=3sg:A} pull\textsc{.pst.ptcp.f=cop.3sg.f:O} \\ 
\glt `From long ago, he would go there and fast for forty days.'
\z 
 
\ea \label{JP.120}
\textit{çilexanêş bîyenû.} \\ 
\gll çilexan(e)-ê=ş bîye=n=û\\ 
 \textsc{pn-indf=3sg:NC} be\textsc{.pst.ptcp.m=cop.3sg.m:S}=and \\ 
\glt `It (the cave) was a place for forty days of fasting.'
\z 
 
\ea \label{JP.121}
\textit{eça meřû asnîne çilêş kêştênew.} \\ 
\gll e=ça meř(e)-û asin-î=ne çilê=ş kêştê=ne=û \\ 
 in=there cave\textsc{.f-ez.gen} iron\textsc{-obl.m=post} forty\_days\_fasting\textsc{.f=3sg:A} pull\textsc{.pst.ptcp.f=cop.3sg.f:O}=and \\ 
\glt `He would fast rounds of forty days of fasting in that Iron Cave.'
\z 
 
\ea \label{JP.122}
\textit{luwan pasarû leme çilêş kêştênew.} \\ 
\gll luwa=n pasar-û leme çilê=ş kêştê=ne=û \\ 
 go\textsc{.pst.ptcp.m=cop.3sg.m:S} \textsc{pn-ez.gen} \textsc{pn} forty\_days\_fasting\textsc{.f=3sg:A} pull\textsc{.pst.ptcp.f=cop.3sg.f:O}=and \\ 
\glt `He would go to [a place called] \textit{Pasaru Lama} ``guardians of the belly'' and would fast for forty days.'
\z 
 
\ea \label{JP.123}
\textit{a çilexane çilexane lemê.} \\ 
\gll a çilexane çilexane lemê \\ 
 \textsc{dem.dist} \textsc{pn} \textsc{pn} belly\textsc{f.sg.obl} \\ 
\glt `[he would fast] in that place of fasting [called] \textit{Chilaxana Lama}.'
\z 
 
\ea \label{JP.124}
\textit{a wextî xeberê yawno hezretû xosî.} \\ 
\gll a wext-î xeber-ê yawn-o hezret-û xos-î \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl} news\textsc{.m-indf} cause\_to\_arrive\textsc{.prs.ind-3sg:S} his\_highness\textsc{.m-ez.gen} \textsc{pn-m.sg.obl} \\ 
\glt `Then, the news reached His Highness Ghaws.'
\z 
 
\ea \label{JP.125}
\textit{maça kabrayê pêse îse a wextîç qisekerê her bîyênê.} \\ 
\gll m-aç-a kabra-ê pêse îse a wext-î=ç qiseker-ê her bîyê=nê \\ 
 \textsc{ind-}say\textsc{.prs-3pl:A} man\textsc{-indf} as\_if now \textsc{dem.dist} time\textsc{.m-sg.obl=add} gossiper\textsc{-pl.dir} \textsc{emph} be\textsc{.ptcp.pl=cop.3pl:S} \\ 
\glt `They said -- Like nowadays, back then, there were gossips too.'
\z 
 
\ea \label{JP.126}
\textit{maça, ``qurban to desûrêt da be sey misefay bilowe şerêʕetû terêqet ders waço.} \\ 
\gll m-aç-a qurban to desûr-ê=t da-{\O} be sey misefa-î bi-l-o=we şerêʕet=û terêqet ders w\'aç-o \\
 \textsc{ind-}say\textsc{.prs-3pl:A} sir\textsc{.voc} \textsc{2sg} order\textsc{.m-indf=2sg:A} give\textsc{.pst-3sg.m:O} to \textsc{pn} \textsc{pn-m.sg.obl} \textsc{sbjv-}go\textsc{.prs-3sg:S=compl} sharia\_law=and denomination lesson\textsc{.m} say\textsc{.prs.sbjv-3sg:A} \\ 
\glt `They (the gossips) said, ``Sir, you ordered Say Mustafa to go back [to Hewraman and] teach Sharia law and [the ways of] orders (of dervishes).'
\z 
 
\ea \label{JP.127}
\textit{îse maça şêt bîyen.} \\ 
\gll îse m-aç-a şêt bîye=n \\ 
 now \textsc{ind-}say\textsc{.prs-3pl:A} crazy be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `Now, rumour has it that he has gone crazy.''
\z 
 
\ea \label{JP.128}
\textit{îse maça şêt bîyen luwan daniş keş.''} \\ 
\gll îse m-aç-a şêt bîye=n luwa=n da=n=iş keş \\ 
 now \textsc{ind-}say\textsc{.prs-3pl:A} crazy be\textsc{.pst.ptcp.m=cop.3sg.m:S} go\textsc{.pst.ptcp.m=cop.3sg.m:S} give\textsc{.pst=cop.3sg.m:S=3sg:A} mountain\textsc{.m} \\ 
\glt `Now rumour has it that he has gone crazy [and] gone and taken refuge in a mountain.'''
\z 
 
\ea \label{JP.129}
\textit{hezretû xosî be mexfî pîyay kîyano menteqew hewramanatî.} \\ 
\gll hezret-û xos-î be mexfî pîya-î kîyan-o menteqe-û hewraman-at-î \\ 
 his\_highness\textsc{.m-ez.gen} \textsc{pn-m.sg.obl} in secret man\textsc{.m-sg.obl} send\textsc{.prs.ind-3sg:A} region\textsc{-ez.gen} \textsc{pn-pl-obl.m} \\ 
\glt `His Highness Ghaws secretly sent a man to the Hewraman region.'
\z 
 
\ea \label{JP.130}
\textit{xeberiş perso.} \\ 
\gll xeber=iş pers-o \\ 
 news\textsc{.m=3sg:PSR} ask\textsc{.prs.ind-3sg:A} \\ 
\glt `He (the man) asked for his (Pir Shaliyar's) news.'
\z 
 
\ea \label{JP.131}
\textit{miđya îna meřû asinî.} \\ 
\gll mi-đy(e)-a îna-∅ meř(e)-û asin-î \\ 
 \textsc{ind-}look\textsc{.prs-3pl:S} \textsc{deic-3sg.m:S} cave\textsc{.f-ez.gen} iron\textsc{-m.sg.obl} \\ 
\glt `They realised that he was in the Iron Cave.'
\z 
 
\ea \label{JP.132}
\textit{pîyakew hezretû xosî milo meřû asinî tûşiş mê.} \\ 
\gll pîya-(e)ke-û hezret-û xos-î mi-l-o meř(e)-û asin-î tûş=iş m-ê \\ 
 man\textsc{.m-def.m.sg.dir-ez.gen} his\_highness\textsc{.m-ez.gen} \textsc{pn-m.sg.obl} \textsc{ind-}go\textsc{.prs-3sg:S} cave\textsc{.f-ez.gen} iron\textsc{-m.sg.obl} encounter\textsc{=3sg:PSR} \textsc{ind-}come\textsc{.prs.3sg:S} \\ 
\glt `His Highness Ghaws's man went [and] ran into him (Pir Shaliyar) in the Iron cave.'
\z 
 
\ea \label{JP.133}
\textit{meřû asinî tûşiş mê.} \\ 
\gll meř(e)-û asin-î tûş=iş m-ê \\ 
 cave\textsc{.f-ez.gen} iron\textsc{-m.sg.obl} encounter\textsc{=3sg:PSR} \textsc{ind-}come\textsc{.prs.3sg:S} \\ 
\glt `He ran into him (Pir Shaliyar) in the Iron Cave.'
\z 
 
\ea \label{JP.134}
\textit{xo ađ mezano hezretû xosî kîyaseniş.} \\ 
\gll xo ađ me-zan-o hezret-û xos-î kîyase=n=iş \\ 
 \textsc{disc.ptcl} \textsc{3sg.m.dir} \textsc{neg.ind-}know\textsc{.prs-3sg:A} his\_highness\textsc{.m-ez.gen} \textsc{pn-m.sg.obl} send\textsc{.pst.ptcp.m=cop.3sg.m:O=3sg:A} \\ 
\glt `Well, he (Pir Shaliyar) did not realise that His Highness Ghaws had sent him (the man).'
\z 
 
\ea \label{JP.135}
\textit{suwaleş çene kero, maço, ``çî amanîwe pey înîşa?''} \\ 
\gll suwale=ş çene ker-o m-aç-o çî ama=nî=we pey înîşa \\ 
 question\textsc{.f=3sg:R} with do\textsc{.prs.ind-3sg:A} \textsc{ind-}say\textsc{.prs-3sg:A} why come\textsc{.pst.ptcp.m=cop.2sg:S=compl} to \textsc{dem.prox.3pl.obl} \\ 
\glt `He asked him [and] said, ``Why have you come here?\footnote{Lit. to [here] and so on}'''
\z 
 
\ea \label{JP.136}
\textit{fermawo, ``min her noʕê amana amanaw înanê çêgew.''} \\ 
\gll fermaw-o min her noʕ-ê ama=na ama=na=û îna-(a)nê çêge=û \\ 
 say\textsc{.prs.ind-3sg:A} \textsc{1sg} \textsc{emph} way\textsc{-indf} come\textsc{.pst.ptcp.m=cop.1sg:S} come\textsc{.pst.ptcp.m=cop.1sg:S}=and \textsc{deic-1sg:S} here=and \\ 
\glt `He said, ``Well, I have come here anyway. I am here.'''
\z 
 
\ea \label{JP.137}
\textit{a wextîyekey her noʕê maçoş lawew.} \\ 
\gll a wext-î-ekey her noʕ-ê m-aç-o=ş la=we=û \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} \textsc{emph} way\textsc{-indf} \textsc{ind-}say\textsc{.prs-3sg:A=3sg:R} to\textsc{=post}=and \\ 
\glt `Then, he (the man) told him (Pir Shaliyar) in some way.'
\z 
 
\ea \label{JP.138}
\textit{maçoş lawe maço, ``min pey îney amana amanaw.''} \\ 
\gll m-aç-o=ş la=we m-aç-o min pey îney ama=na ama=na=û \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A=3sg:R} to\textsc{=post} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{1sg} for \textsc{dem.prox.m.3sg.obl} come\textsc{.pst.ptcp.m=cop.1sg:S} come\textsc{.pst.ptcp.m=cop.1sg:S}=and \\ 
\glt `He said to him, ``I have come for this [reason].'''
\z 
 
\ea \label{JP.139}
\textit{miqdarê çilê kêşo. çil çilê kêşo.} \\ 
\gll miqdarê çilê kêş-o çil çilê kêş-o \\ 
 a\_little forty\_days\_fasting\textsc{.f.sg} pull\textsc{.prs.ind-3sg:A} forty forty\_days\_fasting\textsc{.f.pl.dir} pull\textsc{.prs.ind-3sg:A} \\ 
\glt `He (Pir Shaliyar) did some 40 days of fasting. He fasted for forty rounds of forty days of fasting.'
\z 
 
\ea \label{JP.140}
\textit{çil danê çilê kêşo.} \\ 
\gll çil dan(e)-ê çilê kêş-o \\ 
 forty \textsc{clf-pl.dir} forty\_days\_fasting\textsc{.f.sg} pull\textsc{.prs.ind-3sg:A} \\ 
\glt `He fasted for forty rounds of forty days of fasting.'
\z 
 
\ea \label{JP.141}
\textit{mê mêwe menteqew mêwe dilû hewramanîyû. yaneşa yaneşa eça bo milo wa.} \\ 
\gll m-ê m-ê=we menteqe=û m-ê=we dil-û hewraman-î=û yane=şa yane=şa e=ça b-o mi-l-o wa \\ 
 \textsc{ind-}come\textsc{.prs.3sg:S} \textsc{ind-}come\textsc{.prs.3sg:S=compl} region=and \textsc{ind-}come\textsc{.prs.3sg:S=compl} inside\textsc{-ez.gen} \textsc{pn-m.sg.obl}{=and} house\textsc{.m=3pl:PSR} house\textsc{.m=3pl:PSR} in=there be\textsc{.prs.ind-3sg:S} \textsc{ind-}go\textsc{.prs-3sg:S} there \\ 
\glt `He (Pir Shaliyar) returned to the region, returned to Hewraman. His\footnote{Lit. their} house was there.'
\z 
 
\ea \label{JP.142}
\textit{yanew tateyş ça bîyebo.} \\ 
\gll yane-û tate-î=ş ça bîye=b-o \\ 
 house\textsc{.m-ez.gen} father\textsc{.m-sg.obl=3sg:PSR} there be\textsc{.pst.ptcp.m}=be\textsc{.prs-3sg:S} \\ 
\glt `His father's house was apparently there.'
\z 
 
\ea \label{JP.143}
\textit{milowe yanew êtir pane namêwe peya kero.} \\ 
\gll mi-l-o=we yane=û êtir p=ane namê-(e)we peya ker-o \\ 
 \textsc{ind-}go\textsc{.prs-3sg:S=compl} house\textsc{.m}=and \textsc{disc.ptcl} at=\textsc{dem.dist.m.3sg.dir} name\textsc{.f-indf} visible do\textsc{.prs.ind-3sg:A} \\ 
\glt `He went there (to Hewraman) and got a bit famous.\footnote{Lit. He found a name.}'
\z 
 
\ea \label{JP.144}
\textit{welî emaneketê xas xeɫk bawiřiş pene mekero.} \\ 
\gll welî emaneketê xas xeɫk bawiř=iş pene me-ker-o \\ 
 but but well people\textsc{.m} belief\textsc{=3sg:R} in \textsc{neg.ind-}do\textsc{.prs-3sg:A} \\ 
\glt `But people still did not believe in him that much [as a leader].'
\z 
 
\ea \label{JP.145}
\textit{miqdarê bawiřiş pene kera miqdarê mekera.} \\ 
\gll miqdarê bawiř=iş pene ker-a miqdarê me-ker-a \\ 
 a\_little belief\textsc{=3sg:R} in do\textsc{.prs.ind-3pl:A} a\_little \textsc{neg.ind-}do\textsc{.prs-3pl:A} \\ 
\glt `Some believed in him, some did not.'
\z 
 
\ea \label{JP.146}
\textit{padşaw mîsrî kinaçêş bo felece bo.} \\ 
\gll padşa-û mîsr-î kinaç(ê)-ê=ş b-o felec-e b-o \\ 
 king\textsc{.m-ez.gen} \textsc{pn-m.sg.obl} daughter\textsc{.f-indf=3sg:NC} be\textsc{.prs.ind-3sg:S} disabled\textsc{-f} be\textsc{.prs.ind-3sg:S} \\ 
\glt `The king of Egypt had a daughter [who] was disabled.'
\z 
 
\ea \label{JP.147}
\textit{ħîç dukturê metawo weşeş kerowe.} \\ 
\gll ħîç duktur-ê me-taw-o weş-e=ş k\'er-o=we \\
 no physician\textsc{.m-indf} \textsc{neg.ind-}can\textsc{.prs-3sg:A} well\textsc{-f=3sg:O} do\textsc{.prs.sbjv-3sg:A=compl} \\ 
\glt `No doctor could cure her.'
\z 
 
\ea \label{JP.148}
\textit{hîç dukturê, heçî dukturê, heçî kîşwerê beroş.} \\ 
\gll hîç duktur-ê heçî duktur-ê heçî kîşwer-ê ber-o=ş \\ 
 no physician\textsc{.m-indf} every physician\textsc{.m-pl.dir} every country\textsc{-pl.dir} take\textsc{.prs.ind-3sg:A=3sg:O} \\ 
\glt `No doctor, no doctor [could cure her no matter] how many countries he took her to.'
\z 
 
\ea \label{JP.149}
\textit{kinaçekê felece bo. felecîyekeyş ce diɫişne bo.} \\ 
\gll kinaç(ê)-ekê felec-e b-o felecî-ekey=ş ce diɫ=iş=ne b-o \\ 
 girl\textsc{.f-def.f.sg} disabled\textsc{-f} be\textsc{.prs.ind-3sg:S} disability\textsc{-def.m.sg.obl=3sg:PSR} from heart\textsc{.m=3sg:PSR=post} be\textsc{.prs.ind-3sg:S} \\ 
\glt `The girl was disabled. Her disability came from her tongue\footnote{Lit. heart}'
\z 
 
\ea \label{JP.150}
\textit{qisê metawo kero. nitqiş bînîyo.} \\ 
\gll qis(ê)-ê me-taw-o k\'er-o nitq=iş bînîy(e)-o \\
 talk\textsc{.f-pl.dir} \textsc{neg.ind-}can\textsc{.prs-3sg:A} do\textsc{.prs.sbjv-3sg:A} speech\textsc{=3sg:PSR} close\textsc{.prs.pass-3sg:S} \\ 
\glt `She was not able to speak. She had lost the ability to speak.\footnote{Lit. Her speech was closed.}'
\z 
 
\ea \label{JP.151}
\textit{nitqiş bînîyo.} \\ 
\gll nitq=iş bînîy(e)-o \\ 
 speech\textsc{=3sg:PSR} close\textsc{.prs.pass-3sg:S} \\ 
\glt `She had lost the ability to speak.'
\z 
 
\ea \label{JP.152}
\textit{hîç dukturê metawo weşeş kerowe.} \\ 
\gll hîç duktur-ê me-taw-o weş-e=ş k\'er-o=we \\
 nothing physician\textsc{.m-indf} \textsc{neg.ind-}can\textsc{.prs-3sg:A} well\textsc{-f=3sg:O} do\textsc{.prs.sbjv-3sg:A=compl} \\ 
\glt `No doctor could cure her.'
\z 
 
\ea \label{JP.153}
\textit{şewê wêş wermê wîno.} \\ 
\gll şew(e)-ê wê=ş werm-ê wîn-o \\ 
 night\textsc{-f.sg.obl} \textsc{reflx=3sg:PSR} sleep\textsc{-indf} see\textsc{.prs.ind-3sg:A} \\ 
\glt `One night, he (the king) had a dream.'
\z 
 
\ea \label{JP.154}
\textit{wermekeşne maçaş pene menteqew hewramanatî pîyawê hen tawo î kinaçê to weşe kerowe.} \\ 
\gll werm-eke=ş=ne m-aç-a=ş pene menteqe-û hewraman-at-î pîya-ê hen-∅ taw-o î kinaçê to weş-e k\'er-o=we \\
 sleep\textsc{-def.m.sg.dir=3sg:PSR=post} \textsc{ind-}say\textsc{.prs-3pl:A=3sg:R} to region\textsc{-ez.gen} \textsc{pn-pl-obl.m} man\textsc{.m-indf} \textsc{exist-3sg.m:S} can\textsc{.prs.ind-3sg:A} \textsc{dem.prox} girl\textsc{.f} \textsc{2sg} well\textsc{-f} do\textsc{.prs.sbjv-3sg:A=compl} \\
\glt `In the dream, it was said to him\footnote{Lit. They told him.} that there was a man in the Hewraman region who could cure his daughter.'
\z 
 
\ea \label{JP.155}
\textit{wermekeş haɫîş kera puxte.} \\ 
\gll werm-eke=ş haɫî=ş ker-a {} puxte \\ 
 sleep\textsc{-def.m.sg.dir=3sg:O} understood\textsc{=3sg:O} do\textsc{.prs.ind-3pl:A} {} properly \\
 \glt `The dream explained it to him clearly.\footnote{Lit. It made him understand well.}'
\z 
 
\ea \label{JP.156}
\textit{maça, ``î kinaçê to tawo weşe kerowe.''} \\ 
\gll m-aç-a î kinaçê to taw-o weş-e k\'er-o=we \\
 \textsc{ind-}say\textsc{.prs-3pl:A} \textsc{dem.prox} girl\textsc{.f} \textsc{2sg} can\textsc{.prs.ind-3sg:A} well\textsc{-f} do\textsc{.prs.sbjv-3sg:A=compl} \\ 
\glt `It was said [to him in the dream], ``He can cure this daughter of yours.'''
\z 
 
\ea \label{JP.157}
\textit{padşayç maço be lalo kinaçekê, be şalyare sîyawîş.} \\ 
\gll padşa=îç m-aç-o be lalo kinaç(ê)-ekê be şalyar-e sîyaw-î=ş \\ 
 king\textsc{.m=add} \textsc{ind-}say\textsc{.prs-3sg:A} to maternal\_uncle\textsc{.m} girl\textsc{.f-def.f.sg} to \textsc{pn-ez.cmpd} black\textsc{-obl.m=3sg:PSR} \\ 
\glt `The king said to [his] daughter's uncle, to Shaliyar Syaw.'
\z 
 
 \newpage
\ea \label{JP.158}
\textit{laloşa, lalo heremû pîr şelyarîn.} \\ 
\gll lalo=ş=a lalo herem(e)-û pîr şelyar-î=n \\ 
 maternal\_uncle\textsc{.m=3sg:PSR=cop.3sg.m:S} maternal\_uncle\textsc{.m} wife\textsc{.f-ez.gen} \textsc{pn} \textsc{pn-m.sg.obl=cop.3sg.m:S} \\ 
\glt `-- He was her uncle. He was the uncle of Pir Shaliyar's wife.'
\z 
 
\ea \label{JP.159}
\textit{`to bo luwe! î kinaçê bere.} \\ 
\gll to b-o l\'u-e î kinaçê b\'er-e \\
 \textsc{2sg} \textsc{imp-}come\textsc{.prs.2sg:S} go\textsc{.prs.imp-2sg:S} \textsc{dem.prox} girl\textsc{.f} take\textsc{.prs.imp-2sg:A} \\ 
\glt `Come [here and] go! Take the girl.'
\z 
 
\ea \label{JP.160}
\textit{min wêm padşana.} \\ 
\gll min wê=m padşa=na \\ 
 \textsc{1sg} \textsc{reflx=1sg:PSR} king\textsc{.m=cop.1sg:S} \\ 
\glt `I myself am the king.'
\z 
 
\ea \label{JP.161}
\textit{metawû serû textîwe wezûrew bilû ginû gelû kinaçêw be peyû luway řarew dega be dega.} \\ 
\gll me-taw-û ser-û text-î=we w\'ez-û=re=û bi-l-û g\'in-û gel-û kinaçê=û be pey=û luway řa=re=û dega be dega \\
 \textsc{neg.ind-}can\textsc{.prs-1sg:A} on\textsc{-ez.gen} throne\textsc{-obl.m=post} climb\textsc{.prs.sbjv-1sg:S=povb}=and \textsc{sbjv-}go\textsc{.prs-1sg:S} fall\textsc{.prs.sbjv-1sg:S} with\textsc{-ez.gen} daughter\textsc{.f}=and by foot=and go\textsc{.inf} road\textsc{.f=post}=and village\textsc{.f} by village\textsc{.f} \\ 
\glt `I cannot leave the throne [and] go [and] accompany the girl, [while] walking on foot on the roads [that go] village by village.'
\z 
 
\ea \label{JP.162}
\textit{dêhat be dêhat î kinaçê bere.} \\ 
\gll dêhat be dêhat î kinaçê b\'er-e \\
 village\textsc{.m} by village\textsc{.m} \textsc{dem.prox} girl\textsc{.f} take\textsc{.prs.imp-2sg:A} \\ 
\glt `Take the girl village by village.'
\z 
 
 \newpage
\ea \label{JP.163}
\textit{eger to î kinaçête berdew, î pîya her kes hen,} \\ 
\gll eger to î kinaçê=t=e berd-e=û î pîya her kes hen-∅ \\ 
 if \textsc{2sg} \textsc{dem.prox} girl\textsc{.f.sg=2sg:A=dem} take\textsc{.pst-3sg.f:O}{=and} \textsc{dem.prox} man\textsc{.m} \textsc{emph} person\textsc{.m} \textsc{exist-3sg.m:S} \\ 
\glt `If you take the girl, [no matter] who this man (i.e., the doctor) is,'
\z 
 
\ea \label{JP.164}
\textit{tawaş î duktire î kinaçêşe weşe kerdew,} \\ 
\gll tawa=ş î duktir=e î kinaçê=ş=e weş-e kerd-e=û \\ 
 can\textsc{.pst=3sg:A} \textsc{dem.prox} physician\textsc{=dem} \textsc{dem.prox} girl\textsc{.f.sg=3sg:A=dem} well\textsc{-f} do\textsc{.pst.3sg.f:O}{=and} \\ 
\glt `If the doctor can cure the girl,'
\z 
 
\ea \label{JP.165}
\textit{to wekêɫû min be be mezebû wêşa î kinaçeşa pey nîkah kerew. narîşo!'} \\ 
\gll to wekêɫ-û min b-e be mezeb-û wê=şa î kinaç(ê)=e=şa pey nîkah k\'er-e=û n(e)-ar-î=ş=o \\
 \textsc{2sg} advocate\textsc{.m-ez.gen} \textsc{1sg} be\textsc{.prs.imp-2sg:S} by religion\textsc{-ez.gen} \textsc{reflx=3pl:PSR} \textsc{dem.prox} daughter\textsc{.f=dem=3pl:R} to marriage\textsc{.m} do\textsc{.prs.imp-2sg:A}{=and} \textsc{neg.sbjv-}bring\textsc{.prs-2sg:A=3sg:O=compl} \\ 
\glt `Be my advocate [and] marry the girl to him according to [the customs of] their religion. May you not bring her back!'
\z 
 
\ea \label{JP.166}
\textit{lalo gino gelû kinaçê. maroş. dey dega be degaw kêşwer be kêşwerû maroş silêne.} \\ 
\gll lalo gin-o gel-û kinaçê m-ar-o=ş dey dega be dega=û kêşwer be kêşwer=û m-ar-o=ş silêne \\ 
 maternal\_uncle\textsc{.m} fall\textsc{.prs.ind-3sg:S} with\textsc{-ez.gen} girl\textsc{.f} \textsc{ind-}bring\textsc{.prs-3sg:A=3sg:O} \textsc{disc.ptcl} village\textsc{.f} by village\textsc{.f}=and country by country=and \textsc{ind-}bring\textsc{.prs-3sg:A=3sg:O} \textsc{pn} \\ 
\glt `The uncle set off on the road with the girl. He took her. He took her village by village and country by country until [they arrived] at Silên.'
\z 
 
\ea \label{JP.167}
\textit{silênne řeđê ba. tewenê ça beynne bê beynû silênê biɫberîne bê.} \\ 
\gll silên=ne řeđ-ê b-a tewen(î)-ê ç=a beyn=ne b-ê beyn-û silênê biɫber-î=ne b-ê \\ 
 \textsc{pn=post} crossing\textsc{-pl.dir} be\textsc{.prs.ind-3pl:S} stone\textsc{.f-indf} at=\textsc{dem.dist} between\textsc{=post} be\textsc{.prs-aug.3sg:S} between\textsc{-ez.gen} \textsc{pn.f.sg.obl} \textsc{pn-m.sg.obl=post} be\textsc{.prs-aug.3sg:S} \\ 
\glt `They passed Silên. There was a stone in the road between Silên and Biɫber.'
\z 
 
\ea \label{JP.168}
\textit{leqemiş waçênê tewenû sireway.} \\ 
\gll leqem=iş waç-ên-ê tewen(î)-û sireway \\ 
 title\textsc{.m=3sg:PSR} say\textsc{.prs-aug-3pl:A} stone\textsc{.f-ez.gen} heal\textsc{.inf} \\ 
\glt `The stone is called ``the healing stone.'''
\z 
 
\ea \label{JP.169}
\textit{mênê çenû kinaçekê serû tewenekêwe nîşare.} \\ 
\gll m-ê-nê çenû kinaç(ê)-ekê ser-û tewen(î)-ekê=we nîş-a=re \\ 
 \textsc{ind-}come\textsc{.prs-3pl:S} with\textsc{.ez.gen} girl\textsc{-def.f.sg} on\textsc{-ez.gen} stone\textsc{.f-def.f.sg=post} sit\textsc{.prs.ind-3pl:S=povb} \\ 
\glt `He (the uncle) went with the girl [and] sat on the stone.'
\z 
 
\ea \label{JP.170}
\textit{îse cawe luwan serişne tewenekê.} \\ 
\gll îse cawe luwa=n ser=iş=ne tewen(î)-ekê \\ 
 now road go\textsc{.pst.ptcp.m=cop.3sg.m:S} on\textsc{=3sg:R=post} stone\textsc{.f-def.f.sg} \\ 
\glt `Now, a road has been built on the stone.\footnote{Lit. A road has gone on the stone.}'
\z 
 
\ea \label{JP.171}
\textit{tewenê pane bê. nîşare seyawe.} \\ 
\gll tewen(î)-ê pan-e b-ê nîş-a=re sey-a=we \\ 
 stone\textsc{.f-indf} wide\textsc{-f} be\textsc{.prs-aug.3sg:S} sit\textsc{.prs.ind-3pl:S=povb} rest\textsc{.prs.ind-3pl:S=compl} \\ 
\glt `It was a wide stone. There was a stone there. They sat [on the stone] and rested.'
\z 
 
\ea \label{JP.172}
\textit{kinaçekê maço, ``aaay!''} \\ 
\gll kinaç(ê)-ekê m-aç-o aaay \\ 
 girl\textsc{.f.sg-def.f.sg} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{intj} \\ 
\glt `The girl said, ``Oooh!'''
\z 
 
\ea \label{JP.173}
\textit{qisê kero maço, ``aaay!''} \\ 
\gll qis(ê)-ê ker-o m-aç-o aaay \\ 
 talk\textsc{.f-pl.dir} do\textsc{.prs.ind-3sg:A} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{intj} \\ 
\glt `The girl spoke [and] said, ``Oooh!'''
\z 
 
\ea \label{JP.174}
\textit{lalowekeş maço, ``erê weɫa sirewa.''} \\ 
\gll lalo-eke=ş m-aç-o erê weɫa sirewa \\ 
 maternal\_uncle\textsc{.m-def.m.sg.dir=3sg:PSR} \textsc{ind-}say\textsc{.prs-3sg:A} yes indeed be\_cured\textsc{.pst.3sg.f:S} \\ 
\glt `Her uncle said, ``Indeed, she is cured.'''
\z 
 
\ea \label{JP.175}
\textit{tumez mekere î ay wextê ço cimayre yawa gerûberû silênêw nawêw e pagene nizîkê bawe pîr şelîyar eçê xeber zano.} \\ 
\gll tumez me-ker-e î ay wext-ê ç=o cim-a=re yaw-a gerûber-û silênê=û nawê=û e=pagene nizîk-ê b-a=we pîr şelîyar e=çê xeber zan-o \\ 
 surprisingly \textsc{proh-}do\textsc{.prs-2sg.imp:A} \textsc{dem.prox} \textsc{intj} time\textsc{.m-indf} in=\textsc{dem.dist} move\textsc{.prs.ind-3pl:S=povb} arrive\textsc{.prs.ind-3pl:S} around\textsc{-ez.gen} \textsc{pn.f.sg.obl}=and \textsc{pn.obl}=and in=there close\textsc{-pl} be\textsc{.prs.ind-3pl:S=compl} \textsc{pn} \textsc{pn} in=here news\textsc{.m.sg.dir} know\textsc{.prs.ind-3sg:A} \\ 
\glt `But, as it turned out,\footnote{Lit. You wouldn't say/see!} he (Pir Shaliyar) became aware of [the sound] ``Oh'' [that the girl said] while they were walking and were getting close to somewhere between Silên and Naw.'
\z 
 
\ea \label{JP.176}
\textit{tewecowêşa kero. wextê miđyo duwê danê dêwê yoşa îna peştîşne mê řare yoşa werişne.} \\ 
\gll teweco-ê=şa ker-o wext-ê mi-đy(e)-o duwê dan(e)-ê dêw-ê yo=şa îna-∅ peştî=ş=ne m-ê řa=re yo=şa wer=iş=ne \\ 
 caution\textsc{-indf=3pl:R} do\textsc{.prs.ind-3sg:A} time\textsc{.m-indf} \textsc{ind-}look\textsc{.prs-3sg:S} two \textsc{clf-pl.dir} ogre\textsc{.m-pl.dir} one\textsc{.m=3pl:PSR} \textsc{deic-3sg.m:S} back\textsc{.f=3sg:PSR=post} \textsc{ind-}come\textsc{.prs.3sg:S} road\textsc{.f=post} one\textsc{.m=3pl:PSR} front\textsc{.m=3sg:PSR=post} \\ 
\glt `He noticed [their presence]. He realised that two ogres [were with them], one behind her (the girl), one in front of her.'
\z 
 
\ea \label{JP.177}
\textit{î dêwênê î kinaçêşa bestêne.} \\ 
\gll î dêw-ê=nê î kinaçê=şa bestê=ne \\ 
 \textsc{dem.prox} ogre\textsc{.m-pl.dir=cop.3pl:S} \textsc{dem.prox} girl\textsc{.f.sg=3pl:A} tie\textsc{.pst.ptcp.f=cop.3sg.f:O} \\ 
\glt `It was the ogres who had muted the girl.'
\z 
 
\ea \label{JP.178}
\textit{nitqşa girtêne î duwe dêwene.} \\ 
\gll nitq=şa girtê=ne î duwe dêw=e=ne \\ 
 speech\textsc{=3pl:A} grab\textsc{.pst.ptcp.f=cop.3sg.f:O} \textsc{dem.prox} two ogre\textsc{.m=dem=cop.3sg.f:S} \\ 
\glt `They had muted her speech, these two ogres.'
\z 
 
\ea \label{JP.179}
\textit{eça tewecoşa wene kero.} \\ 
\gll e=ça teweco=şa wene ker-o \\ 
 in=there caution\textsc{=3pl:R} at do\textsc{.prs.ind-3sg:A} \\ 
\glt `He eyed them (the ogres).'
\z 
 
\ea \label{JP.180}
\textit{yoşa heyadar bo gêɫowe.} \\ 
\gll yo=şa heyadar b-o {} gêɫ-o=we \\ 
 one\textsc{.m=3pl:PSR} ashamed be\textsc{.prs.ind-3sg:S} {} wander\textsc{.prs.ind-3sg:S=compl} \\ 
\glt `One [of the ogres] was ashamed; he went back.'
\z 
 
\ea \label{JP.181}
\textit{kinaçekê maço, ``aaay! aaay!''} \\ 
\gll kinaç(ê)-ekê m-aç-o aaay aaay \\ 
 girl\textsc{.f-def.f.sg} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{intj} \textsc{intj} \\ 
\glt `[Then] the girl said, ``Oooh! Oooh!'''
\z 
 
\ea \label{JP.182}
\textit{pîr şelîyar tewecowekey kero.} \\ 
\gll pîr şelîyar teweco-ekey ker-o \\ 
 \textsc{pn} \textsc{pn} caution\textsc{-def.m.sg.obl} do\textsc{.prs.ind-3sg:A} \\ 
\glt `Pir Shaliyar paid attention to her.'
\z 
 
\ea \label{JP.183}
\textit{kinaçekê maço, ``aaay!''} \\ 
\gll kinaç(ê)-ekê m-aç-o aaay \\ 
 girl\textsc{.f-def.f.sg} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{intj} \\ 
\glt `The girl said, ``Ooooh!'''
\z 
 
\ea \label{JP.184}
\textit{laloş maço, ``weɫa erê weɫa sirewa.''} \\ 
\gll lalo=ş m-aç-o weɫa erê weɫa sirewa \\ 
 maternal\_uncle\textsc{.m=3sg:PSR} \textsc{ind-}say\textsc{.prs-3sg:A} indeed yes indeed be\_cured\textsc{.pst.3sg.f:S} \\ 
\glt `Her uncle said, ``Indeed, she is cured.'''
\z 
 
\ea \label{JP.185}
\textit{êti namêşa nîya tewenekê `tewenû sireway'.} \\ 
\gll êti namê=şa nîya tewen(î)-ekê tewen(î)-û sireway \\ 
 \textsc{disc.ptcl} name\textsc{.f.sg=3pl:A} put\textsc{.pst} stone\textsc{.f-def.f.sg} stone\textsc{.f-ez.gen} heal\textsc{.inf} \\ 
\glt `That is why the stone is called Tawanu Sraway, `the healing stone.''
\z 
 
\ea \label{JP.186}
\textit{mênê, mênê pêse maçmê biɫbeřû hurweza milegaw.} \\ 
\gll m-ê-nê m-ê-nê pêse m-aç-mê biɫbeř=û hur-wez-a milega=û \\ 
 \textsc{ind-}come\textsc{.prs-3pl:S} \textsc{ind-}come\textsc{.prs-3pl:S} as\_if \textsc{ind-}say\textsc{.prs-1pl:A} \textsc{pn}=and \textsc{pvb-}climb\textsc{.prs-3pl:S} top\_of\_mountain=and \\ 
\glt `They (the girl and her uncle) came [in the direction of Hewraman]; they climbed a peak of the mountain.'
\z 
 
\ea \label{JP.187}
\textit{milega a wextî mizya werawerû dûtifɫey.} \\ 
\gll milega a wext-î mi-zy(e)-a werawer-û dûtifɫe-y \\ 
 top\_of\_mountain \textsc{dem.dist} time\textsc{.m-sg.obl} \textsc{ind-}go\_out\textsc{.prs-3pl:S} opposite\textsc{-ez.gen} \textsc{pn-m.sg.obl} \\ 
\glt `Then they came out in front of  Dutifla.'
\z 
 
\ea \label{JP.188}
\textit{pîr şelîyar ce yanekew wêşane.} \\ 
\gll pîr şelîyar ce yane-(e)ke-û wê=ş=a=ne \\ 
 \textsc{pn} \textsc{pn} from house\textsc{.m-def.m.sg.dir-ez.gen} \textsc{reflx=3sg:PSR=post} \\ 
\glt `Pir Shaliyar [was] in his house.'
\z 
 
\ea \label{JP.189}
\textit{hezretû sey misefay ça miđyo. xo kes ter mewînoş.} \\ 
\gll hezret-û sey misefa-î ça mi-đy(e)-o xo kes ter me-wîn-o=ş \\ 
 his\_highness\textsc{.m-ez.gen} \textsc{pn} \textsc{pn-m.sg.obl} there \textsc{ind-}look\textsc{.prs-3sg:S} \textsc{disc.ptcl} person\textsc{.m} other \textsc{neg.ind-}see\textsc{.prs-3sg:A=3sg:O} \\ 
\glt `His Highness Say Mustafa watched [them] from above. No one else could see her.'
\z 
 
\ea \label{JP.190}
\textit{ađîşa mewêno.} \\ 
\gll ađîşa me-wên-o \\ 
 \textsc{3pl.obl} \textsc{neg.ind-}see\textsc{.prs-3sg:A} \\ 
\glt `He did not see them.'
\z 
 
\ea \label{JP.191}
\textit{miđyo dûtifɫene duwê neferê amε.} \\ 
\gll mi-đy(e)-o dûtifɫe=ne duwê nefer-ê amε \\ 
 \textsc{ind-}look\textsc{.prs-3sg:S} \textsc{pn=post} two person\textsc{.m-pl.dir} come\textsc{.pst.3pl:S} \\ 
\glt `He saw that two people were coming from Dutifla.'
\z 
 
\ea \label{JP.192}
\textit{dêwê keřetê milo peştîşane keřetê mê werşane.} \\ 
\gll dêw-ê keřet-ê mi-l-o peştî=şa=ne keřet-ê m-ê wer=şa=ne \\ 
 ogre\textsc{.m-indf} time\textsc{.m-indf} \textsc{ind-}go\textsc{.prs-3sg:S} back\textsc{.f=3pl:PSR=post} time\textsc{.m-indf} \textsc{ind-}come\textsc{.prs.3sg:S} front\textsc{.m=3pl:PSR=post} \\ 
\glt `An ogre would sometimes [lit. once] follow them, [and] sometimes [lit. once] go ahead of them.'
\z 
 
\ea \label{JP.193}
\textit{keřetê milo peştîşane keřetê mê werşane.} \\ 
\gll keřet-ê mi-l-o peştî=şa=ne keřet-ê m-ê wer=şa=ne \\ 
 time\textsc{.m-indf} \textsc{ind-}go\textsc{.prs-3sg:S} back\textsc{.f=3pl:PSR=post} time\textsc{.m-indf} \textsc{ind-}come\textsc{.prs.3sg:S} front\textsc{.m=3pl:PSR=post} \\ 
\glt `[An ogre] would sometimes follow them, [and] sometimes go ahead of them.'
\z 
 
\ea \label{JP.194}
\textit{duwê yerê keřetê xuřoş wene. bêheya bo megêɫowe.} \\ 
\gll duwê yerê keřet-ê xuř-o=ş wene bê-heya b-o me-gêɫ-o=we \\ 
 two three time\textsc{.m-pl.dir} shout\textsc{.prs.ind-3sg:A=3sg:R} \textsc{povb} without-decency be\textsc{.prs.ind-3sg:S} \textsc{neg.ind-}turn\textsc{-3sg:S=compl} \\ 
\glt `He (Pir Shaliyar) shouted at him (the ogre) two or three times. [But] he (the ogre) was shameless; he did not return.'
\z 
 
\ea \label{JP.195}
\textit{mê. eha taylê hen?} \\ 
\gll m-ê eha tayl(e)-ê hen-∅ \\ 
 \textsc{ind-}come\textsc{.prs.3sg:S} \textsc{intj} tree\_name\textsc{-indf} \textsc{exist-3sg.m:S} \\ 
\glt `He (the ogre) kept walking. Have you noticed there is a tree?'
\z 
 
\ea \label{JP.196}
\textit{tenûrû maçaş pene ``tenûrû dêwa."''} \\ 
\gll tenûr-û m-aç-a=ş pene tenûr-û dêw-a \\ 
 oven\textsc{.m-ez.gen} \textsc{ind-}say\textsc{.prs-3pl:A=3sg:R} to oven\textsc{.m-ez.gen} ogre\textsc{.m-pl.obl} \\ 
\glt `They call [that place next to the tree] \textit{Tanuru Dewa} ``the ogres' oven.'''
\z 
 
\ea \label{JP.197}
\textit{mê henû taylekey.} \\ 
\gll m-ê hen-û tayle-(e)key \\ 
 \textsc{ind-}come\textsc{.prs.3sg:S} side\textsc{-ez.gen} tree\_name\textsc{-def.m.sg.obl} \\ 
\glt `He (the ogre) arrived next to the tree.'
\z 
 
\ea \label{JP.198}
\textit{fermawo, ``weɫa çûn ana bêheyanî kerût penû ʕaɫemî.} \\ 
\gll fermaw-o weɫa çûn an(e)=a bêheya=nî ker-û=t pen-û ʕaɫem-î \\ 
 say\textsc{.prs.ind-3sg:A} indeed as \textsc{dem.dist.m.3sg.dir=deic} shameless\textsc{=cop.2sg:S} do\textsc{.prs.ind-1sg:A=2sg:O} advice\textsc{.m-ez.gen} world\textsc{-m.sg.obl} \\ 
\glt `He (Pir Shaliyar) said, ``Indeed, since you are so shameless, I (hereby) expose your impropriety to the world. \footnote{Lit. I [hereby] make you the news of the world.}'''
\z 
 
\ea \label{JP.199}
\textit{çûn bêheyanî aşkirat kerû.} \\ 
\gll çûn bêheya=nî aşkira=t ker-û \\ 
 because shameless\textsc{=cop.2sg:S} visible\textsc{=2sg:O} do\textsc{.prs.ind-1sg:A} \\ 
\glt `Since you are [so] shameless, I will expose you.'
\z 
 
\ea \label{JP.200}
\textit{tewecowêş wene kero. pêse gaw til bore e pa tenûrû dêware til bowew ginorewe. anen maçaş pene tenûrû dêwa.} \\ 
\gll teweco-ê=ş wene ker-o pêse gaw til b-o=re e=pa tenûr-û dêw-a=re til b-o=we=û gin-o=re=we. ane=n m-aç-a=ş pene tenûr-û dêw-a \\
 caution\textsc{-indf=3sg:R} at do\textsc{.prs.ind-3sg:A} as\_if bull\textsc{.m} rolling be\textsc{.prs.sbjv-3sg:S=compl} in=that oven\textsc{.m-ez.gen} ogre\textsc{.m-pl.obl=post} rolling be\textsc{.prs.ind-3sg:S=povb}=and fall\textsc{.prs.ind-3sg:S=povb=compl} \textsc{dem.dist.m.3sg.dir=cop.3sg.m:S} \textsc{ind-}say\textsc{.prs-3pl:A=3sg:R} to oven\textsc{.m-ez.gen} ogre\textsc{.m-pl.obl} \\ 
\glt `He (Pir Shaliyar) paid attention to him. He (the ogre) rolled down as if a cow had rolled down. It is there. It is called Tanuru Dewa.'
\z 
 
\ea \label{JP.201}
\textit{tenûrû dêwa êtir pêse gaw til bowe eça werawerû girdû mexeleqeɫaw hewramanîwe dêweke til bowe ginore soɫawe.} \\ 
\gll tenûr-û dêw-a êtir pêse gaw til b-o=we e=ça werawer-û gird-û mexeleqeɫa-û hewraman-î=we dêw-eke til b-o=we gin-o=re soɫawe \\
 oven\textsc{.m-ez.gen} ogre\textsc{.m-pl.obl} \textsc{disc.ptcl} as\_if bull\textsc{.m} rolling be\textsc{.prs.sbjv-3sg:S=compl} in=there opposite\textsc{-ez.gen} all\textsc{-ez.gen} people\textsc{.m-ez.gen} \textsc{pn-m.sg.obl=post} ogre\textsc{.m-def.m.sg.dir} rolling be\textsc{.prs.ind-3sg:S=compl} fall\textsc{.prs.ind-3sg:S=povb} \textsc{pn} \\ 
\glt `Then it was as if a cow had rolled down; the [ogre] rolled down and fell [into a place called] Solawa before all the people's eyes.'
\z 
 
\ea \label{JP.202}
\textit{dêwe çane topo miro.} \\ 
\gll dêw-e ç=a=ne top-o mir-o \\ 
 ogre\textsc{-def} in=\textsc{dem.dist=post} die\textsc{.prs.ind-3sg:S} die\textsc{.prs.ind-3sg:S} \\ 
\glt `The ogre died there.'
\z 
 
\ea \label{JP.203}
\textit{kabra lalow şabare xatûnî mê dilû degεw.} \\ 
\gll kabra lalo-û şabare xatûn-î m-ê dil-û degε=û \\ 
 man maternal\_uncle\textsc{.m-ez.gen} \textsc{pn} lady\textsc{-m.sg.obl} \textsc{ind-}come\textsc{.prs.3sg:S} inside\textsc{-ez.gen} village\textsc{.f.sg.obl}{=and} \\ 
\glt `The fellow, the uncle of Shahbara Khatun, went into the village.'
\z 
 
\ea \label{JP.204}
\textit{perso, ``ceryan çikonû? î yanew kabray kênû îne kênû?''} \\ 
\gll pers-o ceryan çiko=n=û î yane-û kabra-î kê=n=û îne kê=n=û \\ 
 ask\textsc{.prs.ind-3sg:A} story\textsc{.m} where\textsc{=cop.3sg.m:S}=and \textsc{dem.prox} house\textsc{.m-ez.gen} man\textsc{-m.sg.obl} who\textsc{=cop.3sg.m:S}=and \textsc{dem.prox.m.3sg.dir} who\textsc{=cop.3sg.m:S}=and \\ 
\glt `[He came to the village.] He asked [people], ``What is the story? Where is the house of the fellow? Who is he?'''
\z 
 
\ea \label{JP.205}
\textit{muxteserû kelamî, yaney yozawew. mila beraş xizmetû pîrîyû.} \\ 
\gll muxteser-û kelam-î yane-î yoz-a=we=û mi-l-a ber-a=ş xizmet-û pîr-î=û \\ 
 summary\textsc{.m-ez.gen} speech\textsc{-m.sg.obl} house\textsc{.m-sg.obl} find\textsc{.prs.ind-3pl:A=compl}{=and} \textsc{ind-}go\textsc{.prs-3pl:S} take\textsc{.prs.ind-3pl:A=3sg:O} service\textsc{.m-ez.gen} \textsc{pn-m.sg.obl}=and \\ 
\glt `To cut a long story short,\footnote{Lit. gist of speech} they found the house. They (the people) went [and] took him to Pir [Shaliyar].'
.\z 
 
\ea \label{JP.206}
\textit{a wextî lalo kinaçê maço, min, taze padşay kerdena wekêɫ.} \\ 
\gll a wext-î lalo kinaçê m-aç-o min taze padşa-î kerde=na wekêɫ \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl} maternal\_uncle\textsc{.m} girl\textsc{.f} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{1sg} anyway king\textsc{.m-sg.obl} do\textsc{.pst.ptcp.m=cop.1sg:O} advocate\textsc{.m} \\ 
\glt `Then, the girl's uncle said, ``As for me, well, the king has put me in charge.'\footnote{Lit. The king has made me advocate.}
\z 
 
\ea \label{JP.207}
\textit{mişyo î kinaçê nîkah kerû pey tow be mezebû wêtaw.} \\ 
\gll mişyo î kinaçê nîkah k\'er-û pey to=û be mezeb-û wê=ta=û \\
 \textsc{aux} \textsc{dem.prox} girl\textsc{.f} marriage\textsc{.m} do\textsc{.prs.sbjv-1sg:A} to \textsc{2sg}=and in religion\textsc{-ez.gen} \textsc{reflx=2pl:PSR}=and \\ 
\glt `I shall marry this girl to you [according to the customs] of your religion.'
\z 
 
\ea \label{JP.208}
\textit{xo min meberûşo.} \\ 
\gll xo min me-ber-û=ş=o \\ 
 \textsc{disc.ptcl} \textsc{1sg} \textsc{neg.ind-}take\textsc{.prs-1sg:A=3sg:O=compl} \\ 
\glt `I will not take her back.'
\z 
 
\ea \label{JP.209}
\textit{pađşay desûriş dan be minû.''} \\ 
\gll pađşa-î desûr=iş da=n be min=û \\ 
 king\textsc{.m-sg.obl} order\textsc{.m=3sg:A} give\textsc{.pst.ptcp.m=cop.3sg.m:O} to \textsc{1sg}=and \\ 
\glt `The king has ordered me [to do this].'''
\z 
 
\ea \label{JP.210}
\textit{pîr şelîyarîç fermawo, ``dey min kabrayêna aha xeplazeřetînî werûnew. îne padşazayênew derdû min meweronew.} \\ 
\gll pîr şelîyar=îç fermaw-o dey min kabra-ê=na aha xeplazeřetînî wer-ûne=û îne padşazayê=ne=û derd(e)-û min me-wer-one=û \\ 
 \textsc{pn} \textsc{pn=add} say\textsc{.prs.ind-3sg:A} \textsc{disc.ptcl} \textsc{1sg} man\textsc{-indf=cop.1sg:S} \textsc{intj} \textsc{pn} eat\textsc{.prs.ind-1sg:A}=and \textsc{dem.prox.m.3sg.dir} princess\textsc{.f=cop.3sg.f:S}=and illness\textsc{.f-ez.gen} \textsc{1sg} \textsc{neg.ind-}eat\textsc{.prs-3sg:A}=and \\ 
\glt `Pir Shaliyar said, ``Well, I am a fellow -- look —- who eats xaplazaratine (a bread made of corn flour)''. She is a princess. She is not suitable for me.'
\z 
 
\ea \label{JP.211}
\textit{mebarekû sahêbîş bo!} \\ 
\gll mebarek-û sahêb-î=ş b-o \\
 blessing\textsc{.m-ez.gen} owner\textsc{.m-sg.obl=3sg:PSR} be\textsc{.prs.sbjv-3sg:S} \\ 
\glt `May she be happy with her owner (i.e., her father)!'
\z 
 
\ea \label{JP.212}
\textit{bereşo. qey mekeronew!} \\ 
\gll b\'er-e=ş=o qey me-ker-one=û \\
 take\textsc{.prs.imp-2sg:A=3sg:O=compl} limit \textsc{neg.ind-}do\textsc{.prs-3sg:A}=and \\ 
\glt `Take her back. It is all right!'
\z 
 
\ea \label{JP.213}
\textit{înî be min mewero kinaçê padşay.} \\ 
\gll înî be min me-wer-o kinaçê {} padşa-î \\ 
 \textsc{dem.prox.f.3sg.dir} with \textsc{1sg} \textsc{neg.ind-}eat\textsc{.prs-3sg:A} girl\textsc{.f.dir} {} king\textsc{.m-sg.obl} \\ 
\glt `She will not eat with me [under the same roof], the king's daughter.'
\z 
 
\ea \label{JP.214}
\textit{min ane nanma xeplezeřatînen.''} \\ 
\gll min ane nan=m=a xeplezeřatîne=n \\ 
 \textsc{1sg} \textsc{dem.dist.m.3sg.dir} meal\textsc{.m=1sg:PSR=cop.3sg.m:S} \textsc{pn=cop.3sg.m:S} \\ 
\glt `Look, my meal is xaplazaratine.'''
\z 
 
\ea \label{JP.215}
\textit{ane zeratiş şanî miđo, zerate sûrî.} \\ 
\gll ane zerat=iş şanî mi-đ(e)-o zerat-e sûrî \\ 
 \textsc{dem.dist.m.3sg.dir} corn\textsc{=3sg:R} show \textsc{ind-}give\textsc{.prs-3sg:A} corn\textsc{-ez.cmpd} Syrian \\ 
\glt `He (Pir Shaliyar) showed him some corn, [some] red corn.'
\z 
 
\ea \label{JP.216}
\textit{ane şanî mido.} \\ 
\gll ane şanî mi-d(e)-o \\ 
 \textsc{dem.dist.m.3sg.dir} show \textsc{ind-}give\textsc{.prs-3sg:A} \\ 
\glt `He showed him some [corn].'
\z 
 
\ea \label{JP.217}
\textit{fermawo, ``min îne nanmarêw.} \\ 
\gll fermaw-o min îne nan=m=a=rê=û \\ 
 say\textsc{.prs.ind-3sg:A} \textsc{1sg} \textsc{dem.prox.m.3sg.dir} meal\textsc{.m=1sg:PSR=cop.3sg.m:S=povb}=and \\ 
\glt `He said, ``This is my meal.'
\z 
 
\ea \label{JP.218}
\textit{eçîne nanê eçîne werû.''} \\ 
\gll e=çîne nan-ê e=çîne wer-û \\ 
 \textsc{emph}=such meal\textsc{.m-indf} \textsc{emph}=such eat\textsc{.prs.ind-1sg:A} \\ 
\glt `I eat such food.'''
\z 
 
\ea \label{JP.219}
\textit{xo a wextî wêçiş yaneş her mebo.} \\ 
\gll xo a wext-î wê=ç=iş yane=ş her me-b-o \\ 
 \textsc{disc.ptcl} \textsc{dem.dist} time\textsc{.m-sg.obl} \textsc{reflx=add=3sg:PSR} house\textsc{.m=3sg:NC} \textsc{emph} \textsc{neg.ind-}be\textsc{.prs-3sg:S} \\ 
\glt `He had no house at that time.'
\z 
 
\ea \label{JP.220}
\textit{a wextî maça, ``dey qurban!''} \\ 
\gll a wext-î m-aç-a dey qurban \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl} \textsc{ind-}say\textsc{.prs-3pl:A} \textsc{disc.ptcl} sir\textsc{.voc} \\ 
\glt `Then, they (people) said, Oh, Sir!'''
\z 
 
\ea \label{JP.221}
\textit{ma[ça] ... xeɫk dewreş miđo.} \\ 
\gll m-aç-a xeɫk dewre=ş mi-đ(e)-o \\ 
 \textsc{ind-}say\textsc{.prs-3pl:A} people\textsc{.m} rotation\textsc{=3sg:O} \textsc{ind-}give\textsc{.prs-3sg:A} \\ 
\glt `People [came and] surrounded him.'
\z 
 
\ea \label{JP.222}
\textit{maça, ``îne çêşa?} \\ 
\gll m-aç-a îne çêş=a \\ 
 \textsc{ind-}say\textsc{.prs-3pl:A} \textsc{dem.prox.m.3sg.dir} what\textsc{=cop.3sg.m:S} \\ 
\glt `They (the people) said, ``What is this?'
\z 
 
\ea \label{JP.223}
\textit{înê î qisê çêşene? çêş maçdê? maça çêş?''} \\ 
\gll înê î qisê çêşe=ne çêş m-aç-dê m-aç-a çêş \\ 
 \textsc{dem.prox.f.3sg.dir} \textsc{dem.prox} talk\textsc{.f.sg} what\textsc{.f=cop.3sg.f:S} what \textsc{ind-}say\textsc{.prs-2pl:A} \textsc{ind-}say\textsc{.prs-3pl:A} what \\ 
\glt `What is this talk? What are you saying? What are they saying?'''
\z 
 
\ea \label{JP.224}
\textit{dewriş miđa maça...} \\ 
\gll dewr=iş mi-đ(e)-a m-aç-a \\ 
 round\textsc{=3sg:O} \textsc{ind-}give\textsc{.prs-3pl:A} \textsc{ind-}say\textsc{.prs-3pl:A} \\ 
\glt `They surrounded him [and] said...'
\z 
 
\ea \label{JP.225}
\textit{her laloweke maço, ``mişyo î kinaçê min meberûşo. mareş biřûnew''} \\ 
\gll her lalo-eke m-aç-o mişyo î kinaçê min me-ber-û=ş=o mare=ş biř-ûne=û \\ 
 \textsc{emph} maternal\_uncle\textsc{.m-def.m.sg.dir} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{aux} \textsc{dem.prox} girl\textsc{.f} \textsc{1sg} \textsc{neg.ind-}take\textsc{.prs-1sg:A=3sg:O=compl} marraige\textsc{=3sg:O} cut\textsc{.prs.ind-1sg:A}=and \\ 
\glt `The uncle kept saying, ``I shall not take this girl back. I shall marry her [to Pir Shaliyar], and so on.'''
\z 
 
\ea \label{JP.226}
\textit{ađîç maço, ``min metawû î kinaçê wey kerû. îne nanma.''} \\ 
\gll ađ=îç m-aç-o min me-taw-û î kinaçê wey k\'er-û îne nan=m=a \\
 \textsc{3sg.m.dir=add} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{1sg} \textsc{neg.ind-}can\textsc{.prs-1sg:A} \textsc{dem.prox} girl\textsc{.f} raising do\textsc{.prs.sbjv-1sg:A} \textsc{dem.prox.m.3sg.dir} meal\textsc{.m=1sg:PSR=cop.3sg.m:S} \\ 
\glt `People … he (Pir Shaliyar) said, ``I cannot take care of this girl. This is my food.'''
\z 
 
\ea \label{JP.227}
\textit{xeɫkîç maço, ``dey kake gîyan! qey mekero. to îse hîçit nîya, ême, yoma ane hiɫoşet miđo.} \\ 
\gll xeɫk=îç m-aç-o dey kake gîyan qey me-ker-o to îse hîç=it nîy(e)=a ême yo=ma ane hiɫoşe=t mi-đ(e)-o \\ 
 people\textsc{.m=add} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{disc.ptcl} brother\textsc{.voc} dear\textsc{.voc} limit \textsc{neg.ind-}do\textsc{.prs-3sg:A} \textsc{2sg} now\_that nothing\textsc{=2sg:NC} \textsc{neg.exist=cop.3sg.m:S} \textsc{1pl} one\textsc{.m=1pl:PSR} \textsc{dem.dist.m.3sg.dir} cracked\_wheat\textsc{.f.sg.dir=2sg:R} \textsc{ind-}give\textsc{.prs-3sg:A} \\ 
\glt `The people said, ``Oh, dear brother! Take it easy. Now that you have nothing, one of us will give you cracked wheat.'
\z 
 
\ea \label{JP.228}
\textit{yoma ane tişit miđo.} \\ 
\gll yo=ma ane tiş=it mi-đ(e)-o \\ 
 one\textsc{.m=1pl:PSR} \textsc{dem.dist.m.3sg.dir} molasses\textsc{=2sg:R} \textsc{ind-}give\textsc{.prs-3sg:A} \\ 
\glt `One of us will give [pomegranate] molasses.'
\z 
 
\ea \label{JP.229}
\textit{yoyçma kuçê pîyazît miđo.''} \\ 
\gll yo=îç=ma kuç-ê pîyaz-î=t mi-đ(e)-o \\ 
 one\textsc{.m=add=1pl:PSR} little\textsc{-indf} onion\textsc{.m-sg.obl=2sg:R} \textsc{ind-}give\textsc{.prs-3sg:A} \\ 
\glt `One of us will give you some onions.'''
\z 
 
\ea \label{JP.230}
\textit{yoyç ma[ço]... yoyç maço, ``kake gîyan! minîç şimşû yerû heywanêt midew pene.} \\ 
\gll yo=îç m-aç-o yo=îç m-aç-o kake gîyan min=îç şimş(î)=û yeher-û heywan-ê=t mi-de-w pene \\ 
 one\textsc{.m=add} \textsc{ind-}say\textsc{.prs-3sg:A} one\textsc{.m=add} \textsc{ind-}say\textsc{.prs-3sg:A} brother\textsc{.voc} dear\textsc{.voc} \textsc{1sg=add} spleen=and liver\textsc{.m-ez.gen} animal\textsc{.m-indf=2sg:R} \textsc{ind-}give\textsc{.prs-1sg:A} to \\ 
\glt `One [said], One person said, ``Dear brother! I will give you the spleen and liver of an animal.'
\z 
 
\ea \label{JP.231}
\textit{çaştekê kermê.} \\ 
\gll çaşt(î)-ekê k\'er-mê \\
 meal\textsc{.f-def.f.sg} do\textsc{.prs.sbjv-1pl:A} \\ 
\glt `Let us make the meal.'
\z 
 
\ea \label{JP.232}
\textit{çaştekê kermê kermêş be zemawine peyt.} \\ 
\gll çaşt(î)-ekê k\'er-mê k\'er-mê=ş be zemawine pey=t \\
 meal\textsc{.f-def.f.sg} do\textsc{.prs.sbjv-1pl:A} do\textsc{.prs.sbjv-1pl:A=3sg:O} into wedding for\textsc{=2sg:R} \\ 
\glt `Let's cook the meal and hold a wedding for you.'
\z 
 
\ea \label{JP.233}
\textit{nîkahiş kermê peyt.} \\ 
\gll nîkah=iş k\'er-mê pey=t \\
 marriage\textsc{.m=3sg:O} do\textsc{.prs.sbjv-1pl:A} for\textsc{=2sg:R} \\ 
\glt `Let us marry her to you.'
\z 
 
\ea \label{JP.234}
\textit{qey mekero. nimaymê peket gino.} \\ 
\gll qey me-ker-o nim(e)-az-mê peke=t g\'in-o \\
 limit \textsc{neg.ind-}do\textsc{.prs-3sg:A} \textsc{neg.ind-}let\textsc{.prs-1pl:A} strength\textsc{.f=2sg:PSR} fall\textsc{.prs.sbjv-3sg:S} \\ 
\glt `It is all right. We will not let you worry about it.'
\z 
 
\ea \label{JP.235}
\textit{nimaymê peket gino.''} \\ 
\gll nim(e)-az-mê peke=t g\'in-o \\
 \textsc{neg.ind-}let\textsc{.prs-1pl:A} strength\textsc{.f=2sg:PSR} fall\textsc{.prs.sbjv-3sg:S} \\ 
\glt `We will not let you worry about it.'''
\z 
 
\ea \label{JP.236}
\textit{``min hiɫoşet miđew.''} \\ 
\gll min hiɫoşe=t mi-đe-û \\ 
 \textsc{1sg} cracked\_wheat\textsc{.f.sg.dir=2sg:R} \textsc{ind-}give\textsc{.prs-1sg:A} \\ 
\glt `[One of them said], ``I will give you cracked wheat.'''
\z 
 
\ea \label{JP.237}
\textit{yoyç maço, ``minîç tişit miđew.''} \\ 
\gll yo=îç m-aç-o min=îç tiş=it mi-đe-û \\ 
 one\textsc{.m=add} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{1sg=add} molasses\textsc{=2sg:R} \textsc{ind-}give\textsc{.prs-1sg:A} \\ 
\glt `One said, ``I will give you [pomegranate] molasses.'''
\z 
 
 \newpage
\ea \label{JP.238}
\textit{yoyç maço, ``minîç şimşû yeherit miđew.''} \\ 
\gll yo=îç m-aç-o min=îç şimş(î)=û yeher=it mi-đe-û \\ 
 one\textsc{.m=add} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{1sg=add} spleen=and liver\textsc{.m=2sg:R} \textsc{ind-}give\textsc{.prs-1sg:A} \\ 
\glt `One said, ``I will give you spleen and liver.'''
\z 
 
\ea \label{JP.239}
\textit{yoyç maço, ``minîç mekî marû.''} \\ 
\gll yo=îç m-aç-o min=îç mekî m-ar-û \\ 
 one\textsc{.m=add} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{1sg=add} salt\textsc{.f} \textsc{ind-}bring\textsc{.prs-1sg:A} \\ 
\glt `One said, ``I will give you salt.'''
\z 
 
\ea \label{JP.240}
\textit{yoyç maço, ``minîç pîyazî marû.''} \\ 
\gll yo=îç m-aç-o min=îç pîyaz-î m-ar-û \\ 
 one\textsc{.m=add} \textsc{ind-}say\textsc{.prs-3sg:A} \textsc{1sg=add} onion\textsc{.m-sg.obl} \textsc{ind-}bring\textsc{.prs-1sg:A} \\ 
\glt `One said, ``I will give you onions.'''
\z 
 
\ea \label{JP.241}
\textit{çaştekê hiɫoşînekê kermê.} \\ 
\gll çaşt(î)-ekê hiɫoşîne-(e)kê k\'er-mê \\
 meal\textsc{.f-def.f.sg} \textsc{pn-def.f.sg} do\textsc{.prs.sbjv-1pl:A} \\ 
\glt `Let us make the meal called Hiloshina.'
\z 
 
\ea \label{JP.242}
\textit{qey mekero, madam desûr e pêsen padşa řazîye nîya î jenî bilowe î kinaçê bilowe, qey mekero!''} \\ 
\gll qey me-ker-o madam desûr e=pêse=n padşa řazîye nîy(e)=a î jenî bi-l-o=we î kinaçê bi-l-o=we qey me-ker-o \\ 
 limit \textsc{neg.ind-}do\textsc{.prs-3sg:A} as\_long\_as task\textsc{.m} \textsc{emph}=such\textsc{=cop.3sg.m:S} king\textsc{.m} satisfied \textsc{neg.exist=cop.3sg.m:S} \textsc{dem.prox} woman\textsc{.f} \textsc{sbjv-}go\textsc{.prs-3sg:S=compl} \textsc{dem.prox} girl\textsc{.f.sg} \textsc{sbjv-}go\textsc{.prs-3sg:S=compl} limit \textsc{neg.ind-}do\textsc{.prs-3sg:A} \\ 
\glt `There is no problem. Now that things have developed this way, [and] the king is not happy for the girl to return, it's all right! [She can stay here].'''
\z 
 
\ea \label{JP.243}
\textit{pîr şelîyar fermawo, ``tehaa weɫahî îne şime desûrta daw watta çûnke ađê be diɫta wat řasta wat heta xway dinyaş weşe kerdêne her menowe kemîç mekero her zîya kero.} \\ 
\gll pîr şelîyar fermaw-o tehaa weɫahî îne şime desûr=ta da=û wat=ta çûnke ađê be diɫ=ta wat řas=ta wat, heta xwa-î dinya=ş weş-e kerdê=ne, her men-o=we kem=îç me-ker-o; her zîya ker-o \\ 
 \textsc{pn} \textsc{pn} say\textsc{.prs.ind-3sg:A} \textsc{intj} indeed \textsc{dem.prox.m.3sg.dir} \textsc{2pl} order\textsc{.m=2pl:A} give\textsc{.pst}=and say\textsc{.pst=2pl:A} since \textsc{3pl.dir} by heart\textsc{.m=2pl:A} say\textsc{.pst} truth=\textsc{2pl:A} say\textsc{.pst} as\_long\_as God\textsc{.m-sg.obl} world\textsc{.f.sg.dir=3sg:A} well\textsc{-f} do\textsc{.pst.ptcp.f=cop.3sg.f:O} \textsc{emph} remain\textsc{.prs.ind-3sg:S=compl} small\textsc{=add} \textsc{neg.ind-}do\textsc{.prs-3sg:A} \textsc{emph} increased do\textsc{.prs.ind-3sg:A} \\ 
\glt `Pir Shaliyar said, ``Indeed, what you just ordered and said [was right] since you said it with good faith, so long as the world continues to exist, [this ceremony] will continue to exist \footnote{Lit. remains}; it will not be diminished; it will grow.'
\z 
 
\ea \label{JP.244}
\textit{î bezme şime wattaw î çaştîyû îne î desûre şime îcrata kerd, heta xway dinyaş weşe kerdêne, şime řasta wat; be diɫta wat.} \\ 
\gll î bezm=e şime wat=ta=û î çaştî=û îne î desûr=e şime îcra=ta kerd heta xwa-î dinya=ş weş-e kerdê=ne şime řas=ta wat be diɫ=ta wat \\ 
 \textsc{dem.prox} jollification\textsc{.m.sg.dir=dem} \textsc{2pl} say\textsc{.pst=2pl:A}=and \textsc{dem.prox} meal\textsc{.f}=and \textsc{dem.prox.m.3sg.dir} \textsc{dem.prox} order\textsc{.m=dem} \textsc{2pl} performance=\textsc{2pl:A} do\textsc{.pst} as\_long\_as God\textsc{.m-sg.obl} world\textsc{.f.sg.dir=3sg:A} well\textsc{-f} do\textsc{.pst.ptcp.f=cop.3sg.f:O} \textsc{2pl} truth=\textsc{2pl:A} say\textsc{.pst} by heart\textsc{.m=2pl:A} say\textsc{.pst} \\ 
\glt `[After] the things you said, and the meals which you prepared, so long as the world continues to exist, you are right [in so doing]; you did these [things] in good faith.'
\z 
 
\ea \label{JP.245}
\textit{î merasême her menowe, hîçkeyç kem mekero. her zîyay kero. îlay zîyayne bo zîyayiş pene kero.''} \\ 
\gll î merasêm=e her men-o=we hîçke=îç kem me-ker-o her zîyay ker-o îlay zîyay=ne b-o zîyay=iş pene ker-o \\ 
 \textsc{dem.prox} ceremony\textsc{.m=dem} \textsc{emph} remain\textsc{.prs.ind-3sg:S=compl} never\textsc{=add} little \textsc{neg.ind-}do\textsc{.prs-3sg:A} \textsc{emph} increase do\textsc{.prs.ind-3sg:A} for\_sure increased\textsc{=post} be\textsc{.prs.ind-3sg:S} increased\textsc{=3sg:R} to do\textsc{.prs.ind-3sg:A} \\ 
\glt `This ceremony will be repeated; it will never diminish. It will continue to grow. It will grow [into bigger ceremonies].'''
\z 
 
\ea \label{JP.246}
\textit{desûr miđa.} \\ 
\gll desûr mi-đ(e)-a \\ 
 order\textsc{.m} \textsc{ind-}give\textsc{.prs-3pl:A} \\ 
\glt `They (the people) ordered [that the ceremony should take place].'
\z 
 
\ea \label{JP.247}
\textit{yo hiɫoşê marow.} \\ 
\gll yo hiɫoş(e)-ê m-ar-o=û \\ 
 one\textsc{.m} cracked\_wheat\textsc{-f.sg.obl} \textsc{ind-}bring\textsc{.prs-3sg:A}=and \\ 
\glt `One fetched cracked wheat.'
\z 
 
\ea \label{JP.248}
\textit{yo pîyaz marow.} \\ 
\gll yo pîyaz m-ar-o=û \\ 
 one\textsc{.m} onion\textsc{.m} \textsc{ind-}bring\textsc{.prs-3sg:A}=and \\ 
\glt `One fetched onions.'
\z 
 
\ea \label{JP.249}
\textit{e wextîyekey yo tişe henar marow.} \\ 
\gll e=wext-î-ekey yo tiş-e henar m-ar-o=û \\ 
 \textsc{dem}=time\textsc{.m-sg.obl-def.m.sg.obl} one\textsc{.m} molasses\textsc{-ez.cmpd} pomegranate\textsc{.m} \textsc{ind-}bring\textsc{.prs-3sg:A}=and \\ 
\glt `One fetched pomegranate molasses.'
\z 
 
\ea \label{JP.250}
\textit{kîyanaş kabrayê serûpîrî qesab bo.} \\ 
\gll kîyan-a=ş kabra-ê serûpîrî qesab b-o \\ 
 send\textsc{.prs.ind-3pl:A=3sg:O} man\textsc{-indf} \textsc{pn} butcher\textsc{.m} be\textsc{.prs.ind-3sg:S} \\ 
\glt `They sent for a fellow from Saru Piri, who was a butcher. [They fetched him].'
\z 
 
\ea \label{JP.251}
\textit{kîyanaş serûpîrîwe qesabê bo maçaş tayfew padşay.} \\ 
\gll kîyan-a=ş serûpîrî=we qesab-ê b-o m-aç-a=ş tayfe-û padşa-î \\ 
 send\textsc{.prs.ind-3pl:A=3sg:O} \textsc{pn=post} butcher\textsc{.m-indf} be\textsc{.prs.ind-3sg:S} \textsc{ind-}say\textsc{.prs-3pl:A=3sg:O} tirbe\textsc{.m-ez.gen} king\textsc{.m-sg.obl} \\ 
\glt `They sent [one] to Saru Piri; there was a butcher whose [tribe] is called the king's tribe.'
\z 
 
\ea \label{JP.252}
\textit{tayfew padşayaş waçênê.} \\ 
\gll tayfe-û padşa-ya=ş waç-ên-ê \\ 
 tirbe\textsc{.m-ez.gen} king\textsc{.m-pl.obl=3sg:O} say\textsc{.prs-aug-3pl:A} \\ 
\glt `They (his tribe) were called the kings' tribe.'
\z 
 
\ea \label{JP.253}
\textit{maraş. serê heywan sere biřo. çaştekê kera.} \\ 
\gll m-ar-a=ş ser-ê heywan sere biř-o çaşt(î)-ekê ker-a \\ 
 \textsc{ind-}bring\textsc{.prs-3pl:A=3sg:O} \textsc{clf-indf} animal\textsc{.m} head\textsc{.m} cut\textsc{.prs.ind-3sg:A} meal\textsc{.f-def.f.sg} do\textsc{.prs.ind-3pl:A} \\ 
\glt `They fetched him. He butchered one animal. They cooked the meal.'
\z 
 
\ea \label{JP.254}
\textit{çaştekê încam miđa.} \\ 
\gll çaşt(î)-ekê încam mi-đ(e)-a \\ 
 meal\textsc{.f-def.f.sg} accomplishment \textsc{ind-}give\textsc{.prs-3pl:A} \\ 
\glt `They made the meal.'
\z 
 
\ea \label{JP.255}
\textit{çaştekê încam miđanêw.} \\ 
\gll çaşt(î)-ekê încam mi-đ(e)-anê=û \\ 
 meal\textsc{.f-def.f.sg} accomplishment \textsc{ind-}give\textsc{.prs-3pl:A}=and \\ 
\glt `They made the meal.'
\z 
 
 \newpage
\ea \label{JP.256}
\textit{merasêmekew pîr şelîyarî bergozar keraw.} \\ 
\gll merasêm-eke-û pîr şelîyar-î bergozar ker-a=û \\ 
 ceremony\textsc{.m-def.m.sg.dir-ez.gen} \textsc{pn} \textsc{pn-m.sg.obl} holding do\textsc{.prs.ind-3pl:A}=and \\ 
\glt `They held Pir Shaliyar's ceremony.'
\z 
 
\ea \label{JP.257}
\textit{jenî be mezebû hewramanî mare biřa pey pîr şelîyarîyû.} \\ 
\gll jenî be mezeb-û hewraman-î mare biř-a pey pîr şelîyar-î=û \\ 
 woman\textsc{.f} by religion\textsc{-ez.gen} \textsc{pn-m.sg.obl} marraige cut\textsc{.prs.ind-3pl:A} to \textsc{pn} \textsc{pn-m.sg.obl}=and \\ 
\glt `They married the woman to Pir Shaliyar] according to] the religious customs in Hewraman.'
\z 
 
\ea \label{JP.258}
\textit{a wextîyekey lalo pîr şelîyar hîn lalo heremû pîr şelîyarî milowe xeber berowe pey padşay.} \\ 
\gll a wext-î-ekey lalo pîr şelîyar hîn lalo herem(e)-û pîr şelîyar-î mi-l-o=we xeber ber-o=we pey padşa-î \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} maternal\_uncle\textsc{.m} \textsc{pn} \textsc{pn} \textsc{fill} maternal\_uncle\textsc{.m} wife\textsc{.f-ez.gen} \textsc{pn} \textsc{pn-m.sg.obl} \textsc{ind-}go\textsc{.prs-3sg:S=compl} news\textsc{.m.sg.dir} take\textsc{.prs.ind-3sg:A=compl} to king\textsc{.m-sg.obl} \\ 
\glt `Then, Pir Shaliyar's uncle, eh, Pir Shaliyar's wife's uncle, went back [to Egypt] and passed on the news to the king.'
\z 
 
\ea \label{JP.259}
\textit{xeber berowe pey padşay maço, ``haɫû cereyanê eçînen.''} \\ 
\gll xeber ber-o=we pey padşa-î m-aç-o haɫ-û cereyan-ê e=çîne=n \\ 
 news\textsc{.m.sg.dir} take\textsc{.prs.ind-3sg:A=compl} to king\textsc{.m-sg.obl} \textsc{ind-}say\textsc{.prs-3sg:A} state\textsc{.m-ez.gen} story\textsc{.m-indf} \textsc{emph}=such\textsc{=cop.3sg.m:S} \\ 
\glt `He passed on the news to the king [and] said, ``The story is like this.'''
\z 
 
 \newpage
\ea \label{JP.260}
\textit{padşa fermawo, ``dey mubarekeş bo.} \\ 
\gll padşa fermaw-o dey mubarek-e=ş b-o \\
 king\textsc{.m} say\textsc{.prs.ind-3sg:A} \textsc{disc.ptcl} blessed\textsc{-f=3sg:R} be\textsc{.prs.sbjv-3sg:S} \\ 
\glt `The king said, ``May he be blessed with happiness.'\footnote{Lit. May she be blessed to him.}
\z 
 
\ea \label{JP.261}
\textit{mađam weşeş kerdêbowe mubarekeş bonew.} \\ 
\gll mađam weş-e=ş kerdê=b-o=we mubarek-e=ş b-one=û \\
 as\_long\_as well\textsc{-f=3sg:A} do\textsc{.pst.ptcp.f}=be\textsc{.prs-3sg:O=compl} blessed\textsc{-f=3sg:R} be\textsc{.prs.sbjv-3sg:S}=and \\ 
\glt `Now that it appears that he has healed her [thoroughly], may he be blessed with happiness.'
\z 
 
\ea \label{JP.262}
\textit{luwe! to luwewe hem wesaylîç bere pey î pîyayû î kinaçêw} \\ 
\gll l\'u-e to lu-e=we hem wesayl=îç b\'er-e pey î pîya-î=û î kinaçê=û \\
 go\textsc{.prs.imp-2sg:S} \textsc{2sg} go\textsc{.prs.imp-2sg:S=compl} \textsc{emph} equipment\textsc{.m=add} take\textsc{.prs.imp-2sg:A} to \textsc{dem.prox} man\textsc{.m-sg.obl}=and \textsc{dem.prox} daughter\textsc{.sg.dir}=and \\ 
\glt `Go! Go back! Take some household equipment for this man and [my] daughter.'
\z 
 
\ea \label{JP.263}
\textit{êtir to luwe meylû wêta.''} \\ 
\gll êtir to l\'u-e meyl-û wê=t=a \\ 
 \textsc{disc.ptcl} \textsc{2sg} go\textsc{.prs.imp-2sg:S} willingness\textsc{-ez.gen} \textsc{reflx=2sg:PSR=cop.3sg.m:S} \\ 
\glt `Then it's up to you [to stay in Hewraman]."
\z 
 
\ea \label{JP.264}
\textit{lalo mêwe wesayil marowe pey kinaçê pêse îsey maçmê teleş pey bera.} \\ 
\gll lalo m-ê=we wesayil m-ar-o=we pey kinaçê pêse îse-î m-aç-mê tele=ş pey ber-a \\ 
 maternal\_uncle\textsc{.m} \textsc{ind-}come\textsc{.prs.3sg:S=compl} equipment\textsc{.m} \textsc{ind-}bring\textsc{.prs-3sg:A=compl} to girl\textsc{.f} like now\textsc{-m.sg.obl} \textsc{ind-}say\textsc{.prs-1pl:A} wedding\_gift\textsc{=3sg:R} to take\textsc{.prs.ind-3pl:A} \\ 
\glt `The uncle went back. He brought household equipment for the girl, just like nowadays when we say, ``They (the bride's family) take gifts for her (the bride).'''
\z 
 
\ea \label{JP.265}
\textit{teleş pey maro. maro mê milowe.} \\ 
\gll tele=ş pey m-ar-o m-ar-o m-ê mi-l-o=we \\ 
 wedding\_gift\textsc{=3sg:R} to \textsc{ind-}bring\textsc{.prs-3sg:A} \textsc{ind-}bring\textsc{.prs-3sg:A} \textsc{ind-}come\textsc{.prs.3sg:S} \textsc{ind-}go\textsc{.prs-3sg:S=compl} \\ 
\glt `He (the uncle) took gifts to her. He took [the gifts] and went back to Hewraman.'
\z 
 
\ea \label{JP.266}
\textit{êti melowe ađîç melowe.} \\ 
\gll êti me-l-o=we ađ=îç me-l-o=we \\ 
 \textsc{disc.ptcl} \textsc{neg.ind-}go\textsc{.prs-3sg:S=compl} \textsc{3sg.m.dir=add} \textsc{neg.ind-}go\textsc{.prs-3sg:S=compl} \\ 
\glt `After all, he didn't go back [to Egypt]. He did not go back either.'
\z 
 
\ea \label{JP.267}
\textit{anen îse şelîyare sîyawa.} \\ 
\gll ane=n îse şelîyar-e sîyaw=a \\ 
 \textsc{dem.dist.m.3sg.dir=cop.3sg.m:S} now \textsc{pn-ez.cmpd} black\textsc{=cop.3sg.m:S} \\ 
\glt `[His shrine] is over there now; [he is called] Shaliyar Siyaw.'
\z 
 
\ea \label{JP.268}
\textit{\textsuperscript{A} weselam ʕeleykum we řehmetuɫahê we berekatun} \textsuperscript{A}\\ 
\gll weselam ʕeleykum we řehmetuɫahê we berekatun \\ 
 peace upon\_you and God's\_mercy and blessing \\ 
\glt `May peace and God's mercy and blessings be upon you.'
\z
