\ea \label{JM.1}
\textit{heşta saɫê ʕemirma.} \\ 
\gll heşta saɫ(e)-ê ʕemr=m=a \\ 
 eighty year\textsc{.f-pl.dir} age\textsc{.m=1sg:PSR=cop.3sg.m:S} \\ 
\glt `I am eighty years old.'
\z 
 
\ea \label{JM.2}
\textit{tatem duwê jenîş ardênê.} \\ 
\gll tate=m duwê jenî=ş ardê=nê \\ 
 father\textsc{.m=1sg:PSR} two woman\textsc{.f.sg=3sg:A} take\textsc{.pst.ptcp.pl=cop.3pl:O} \\ 
\glt `My father married two women.'
\z 
 
\ea \label{JM.3}
\textit{min kuřû duwêna.} \\ 
\gll min kuř-û duw(ê)-ê=na \\ 
 \textsc{1sg} son\textsc{.m-ez.gen} two\textsc{-ord=cop.1sg:S} \\ 
\glt `I am the son of the second [wife].'
\z 
 
\ea \label{JM.4}
\textit{şûnû aneyre, şûnû řay eweɫ ce karîgerîwe serem berşîyen.} \\ 
\gll şûn-û aney=re şûn-û řa-î eweɫ ce karîgerî=we sere=m {} ber-şîye=n \\ 
 after\textsc{-ez.gen} \textsc{dem.dist.m.3sg.obl=post} after\textsc{-ez.gen} road\textsc{.f-ez.attr} first from labourer\_job\textsc{=post} head\textsc{.m=1sg:PSR} {} out-go\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `Well, I first learned\footnote{Lit. My head went out.}  to work as a labourer.'
\z 
 
\ea \label{JM.5}
\textit{ce herbeneywe serem berşîyen.} \\ 
\gll ce herbene-î=we sere=m ber-şîye=n \\ 
 from donkey\_keeper\textsc{.m-sg.obl=post} head\textsc{.m=1sg:PSR} out-go\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `I learned to work as a donkey keeper.'
\z 
 
\ea \label{JM.6}
\textit{şûnîre amana desim kerđen karêz biřyey.} \\ 
\gll şûnî=re ama=na des=im kerđe=n karêz biřyey \\ 
 afterwards\textsc{=post} come\textsc{.pst.ptcp.m=cop.1sg:S} hand\textsc{.m=1sg:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O} subterranean\_canal\textsc{.m} cut\textsc{.inf} \\ 
\glt `Then, I started\footnote{Lit. hand do} to dig subterranean canals.'
\z 
 
\ea \label{JM.7}
\textit{ce karêz biřyeywe pî heşta saɫe. îse ħewt saɫê} \\ 
\gll ce karêz biřyey=we p=î heşta saɫ(e)=e îse ħewt saɫ(e)-ê \\ 
 from subterranean\_canal\textsc{.m} cut\textsc{.inf=post} at=\textsc{dem.prox} eighty year\textsc{.f=dem} now seven year\textsc{.f-pl.dir} \\ 
\glt `From [the time I was] digging canals until now that I am eighty years old [I have been working].'
\z 
 
\ea \label{JM.8}
\textit{îse ħewt saɫêç amanawe serû îney dûkandarna.} \\ 
\gll îse ħewt saɫ(e)-ê=ç ama=na=we ser-û îney dûkandar=na \\ 
 now seven year\textsc{.f-pl.dir=add} come\textsc{.pst.ptcp.m=cop.1sg:S=compl} on\textsc{-ez.gen} \textsc{dem.prox.m.3sg.obl} shopkeeper\textsc{.m=cop.1sg:S} \\ 
\glt `Now, I have been a shopkeeper for seven years.'
\z 
 
\ea \label{JM.9}
\textit{qaçim meřyanû kewtena yanew.} \\ 
\gll qaç=im meřya=n=û kewte=na yane=û \\ 
 leg\textsc{.m=1sg:PSR} break\textsc{.pst.ptcp.m=cop.3sg.m:S}=and fall\textsc{.pst.ptcp.m=cop.1sg:S} house\textsc{.m}=and \\ 
\glt `My leg is broken. I am confined to home, and so on.\footnote{Lit. I have fallen at home.}'
\z 
 
\ea \label{JM.10}
\textit{a wextîyekey pîr bîyena.} \\ 
\gll a wext-î-ekey pîr bîye=na \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} old be\textsc{.pst.ptcp.m=cop.1sg:S} \\ 
\glt `I have aged.'
\z 
 
\ea \label{JM.11}
\textit{çemêm korê bîyênêw.} \\ 
\gll çem-ê=m kor-ê bîyê=nê=û \\ 
 eye\textsc{-pl.dir=1sg:PSR} blind\textsc{-pl.dir} be\textsc{.ptcp.pl=cop.3pl:S}=and \\ 
\glt `My eyesight has deteriorated.\footnote{My eyes have gotten blind.}'
\z 
 
\ea \label{JM.12}
\textit{îne dastanû mina.} \\ 
\gll îne dastan(e)-û min=a \\ 
 \textsc{dem.prox.m.3sg.dir} story\textsc{.f-ez.gen} \textsc{1sg=cop.3sg.m:S} \\ 
\glt `This is my story.'
\z 
 
\ea \label{JM.13}
\textit{eger tûlû ʕemrîm temamû dinyεm wişkinan pey jîwayû dinyε.} \\ 
\gll eger tûl-û ʕemr-î=m temam-û dinyε=m wişkina=n pey jîway-û dinyε \\ 
 if lenght\textsc{.m-ez.gen} age\textsc{.m-sg.obl=1sg:PSR} all\textsc{-ez.gen} world\textsc{.f.sg.obl=1sg:A} scour\textsc{.pst.ptcp.m=cop.3sg.m:O} for live\textsc{.inf-ez.gen} world\textsc{.f.sg.obl} \\ 
\glt `Throughout my life, I have searched all around the world [for ways] to make ends meet.'
\z 
 
\ea \label{JM.14}
\textit{dizîm nekerđenû hîzîm nekerdenû. girđkar nebîyena.} \\ 
\gll dizî=m ne-kerđe=n=û hîzî=m ne-kerde=n=û girđkar {} ne-bîye=na \\ 
 theft\textsc{.m.sg=1sg:A} \textsc{neg-}do\textsc{.pst.ptcp.m=cop.3sg.m:O}=and adultery\textsc{.m.sg=1sg:A} \textsc{neg-}do\textsc{.pst.ptcp.m=cop.3sg.m:O}=and know\_all {} \textsc{neg-}be\textsc{.pst.ptcp.m=cop.1sg:S} \\ 
\glt `I have not committed theft or adultery. I was not a know-all.'
\z 
 
\ea \label{JM.15}
\textit{a wextîyekey eçî newʕe řû be werû wêmre řû be xwa řisqêm peya kerđen.} \\ 
\gll a wext-î-ekey eçî newʕ=e řû be wer-û wê=m=re řû be xwa řisq-ê=m peya kerđe=n \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} like\_this way\textsc{=dem} face by front\textsc{.m-ez.gen} \textsc{reflx=1sg:PSR=post} face by God\textsc{.m} living\textsc{-indf=1sg:A} visible do\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `I have taken refuge in God and found a living.'
\z 
 
\ea \label{JM.16}
\textit{daraw îse elʕan haɫêhazirî daraw penc kuřanaw yerê kinaça, heştû wêmû jenê îne de neferê.} \\ 
\gll dara-û îse elʕan haɫêhazir-î dara-û penc kuř-a=na=û yerê kinaç(ê)-a heşt=û wê=m=û jen(î)-ê îne de nefer-ê \\ 
 owner\textsc{.m-ez.gen} now now at\_the\_moment\textsc{-m.sg.obl} owner\textsc{.m-ez.gen} five son\textsc{.m-pl.obl=cop.1sg:S}=and three daughter\textsc{.f-pl.obl} eight=and \textsc{reflx=1sg:PSR}=and woman\textsc{.f-indf} \textsc{dem.prox.m.3sg.dir} ten person\textsc{.m-pl.dir} \\ 
\glt `I have, at the moment, five sons and three daughters, [which makes] eight, [plus] myself and a wife, [making us] ten people.'
\z 
 
\ea \label{JM.17}
\textit{ta îse yawana be řehmû xwaywe.} \\ 
\gll ta îse yawa=na be řehm-û xwa-î=we \\ 
 until now arrive\textsc{.pst.ptcp.m=cop.1sg:S} by mercy\textsc{.m-ez.gen} God\textsc{.m-sg.obl=post} \\ 
\glt `I have made\footnote{Lit. arrived} it so far, thanks to God's blessing.'
\z 
 
\ea \label{JM.18}
\textit{ta wê yawana.} \\ 
\gll ta wê yawa=na \\ 
 until there arrive\textsc{.pst.ptcp.m=cop.1sg:S} \\ 
\glt `I have made it so far.'
\z 
 
\ea \label{JM.19}
\textit{î menteqew ême kemûkuřîn.} \\ 
\gll î menteqe-û ême kemûkořî=n \\ 
 \textsc{dem.prox} region\textsc{-ez.gen} \textsc{1pl} poverty\textsc{.m=cop.3sg.m:S} \\ 
\glt `Our region is in poverty.'
\z 
 
\ea \label{JM.20}
\textit{min zemanê zemaniş bîyen jenîm ardêne girđû karêw nebîyey bîyen.} \\ 
\gll min zeman-ê zeman=iş bîye=n jenî=m ardê=ne girđ-û kar-ê-û nebîyey bîye=n \\ 
 \textsc{1sg} time\textsc{.m-indf} time\textsc{.m=3sg:PSR} be\textsc{.pst.ptcp.m=cop.3sg.m:S} woman\textsc{.f.sg.dir=1sg:A} bring\textsc{.pst.ptcp.f=cop.3sg.f:O} all\textsc{-ez.gen} task\textsc{.m-pl.dir-ez.gen} poverty\textsc{.m} be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `As for me, once I got married, there was poverty all around.'
\z 
 
\ea \label{JM.21}
\textit{î mentêqew ême nebîyeyş hen nebîyey.} \\ 
\gll î mentêqe-û ême nebîyey=ş hen-∅ nebîyey \\ 
 \textsc{dem.prox} region\textsc{.m-ez.gen} \textsc{1pl} poverty\textsc{.m=3sg:NC} \textsc{exist-3sg.m:S} poverty\textsc{.m} \\ 
\glt `Our region is in poverty, [and] shortage.'
\z 
 
\ea \label{JM.22}
\textit{a wextî mişyê bilmê ẍaricne çîw barmêw çîw bermê.} \\ 
\gll a wext-î mişyê bi-l-mê ẍaric=ne çîw b-ar-mê=û çîw b\'er-mê \\
 \textsc{dem.dist} time\textsc{.m-sg.obl} \textsc{aux.pst} \textsc{sbjv-}go\textsc{.prs-1pl:S} abroad\textsc{=post} thing\textsc{.m} \textsc{sbjv-}bring\textsc{.prs-1pl:A}=and thing\textsc{.m} take\textsc{.prs.sbjv-1pl:A} \\ 
\glt `We had to go out [of Hewraman] to fetch and take stuff [to sell].'
\z 
 
\ea \label{JM.23}
\textit{luwεnmê ferz taran karma kerđen şiş mangê.} \\ 
\gll luwε=nmê ferz taran kar=ma kerđe=n şiş mang(e)-ê \\ 
 go\textsc{.pst.ptcp.pl=cop.1pl:S} guess\textsc{.m} \textsc{pn} work\textsc{.m=1pl:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O} six month\textsc{.f-pl.dir} \\ 
\glt `We would go, say, to Tehran and work for six months.'
\z 
 
\ea \label{JM.24}
\textit{luwεnmê kirman karma kerđen şiş mangê.} \\ 
\gll luwε=nmê kirman kar=ma kerđe=n şiş mang(e)-ê \\ 
 go\textsc{.pst.ptcp.pl=cop.1pl:S} \textsc{pn} work\textsc{.m=1pl:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O} six month\textsc{.f-pl.dir} \\ 
\glt `We would go to Kerman and work for six months.'
\z 
 
\ea \label{JM.25}
\textit{cezîrew kîşîne karma kerđen şiş mangê.} \\ 
\gll cezîre-û kîş-î=ne kar=ma kerđe=n şiş mang(e)-ê \\ 
 island\textsc{-ez.gen} \textsc{pn-m.sg.obl=post} work\textsc{.m=1pl:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O} six month\textsc{.f-pl.dir} \\ 
\glt `[We would go] and work on the Island of Kish for six months.'
\z 
 
\ea \label{JM.26}
\textit{a wextîyekey ħewt no, yanze zaroɫêm bîyênê.} \\ 
\gll a wext-î-ekey ħewt no yanze zaroɫ(e)-ê=m bîyê=nê \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} seven nine eleven child\textsc{-pl.dir=1sg:NC} be\textsc{.ptcp.pl=cop.3pl:S} \\ 
\glt `I had seven or nine or eleven children.'
\z 
 
\ea \label{JM.27}
\textit{î yaŋze zaroɫê eçê asênêm cîya milû î jenîme milû nebîyeywe.} \\ 
\gll î yaŋze zaroɫ(e)-ê e=çê asê=nê=m cîya mil-û î jenî=m=e mil-û nebîyey=we \\ 
 \textsc{dem.prox} eleven child\textsc{-pl.dir} in=here leave\textsc{.pst.ptcp.pl=cop.3pl:O=1sg:A} separate on\textsc{-ez.gen} \textsc{dem.prox} woman\textsc{.f=1sg:PSR=dem} on\textsc{-ez.gen} poverty\textsc{.m=post} \\ 
\glt `Because of destitution, [I had to] leave these eleven children to my wife.'
\z 
 
\ea \label{JM.28}
\textit{min eçagene karîger bîyena. şiş mangê xeberêm nezanan.} \\ 
\gll min e=çagene karîger bîye=na şiş mang(e)-ê xeber-ê=m ne-zana=n \\ 
 \textsc{1sg} in=there labourer\textsc{.m} be\textsc{.pst.ptcp.m=cop.1sg:S} six month\textsc{.f-pl.dir} news\textsc{.m-indf=1sg:A} \textsc{neg-}know\textsc{.pst.ptcp.m=cop.3sg.m:PSR} \\ 
\glt `I was a worker there. I was unaware of them\footnote{Lit. I didn't know their news.} (my children) for six months.'
\z 
 
\ea \label{JM.29}
\textit{bizanû kewtî kewtênê warđîşa warden dizîş dizîyen.} \\ 
\gll bi-zan-û kewt-î kewtê=nê warđ-î=şa warde=n diz(î)-î=ş dizîye=n \\ 
 \textsc{sbjv-}know\textsc{.prs-1sg:A} fall\textsc{.pst-nmlz} fall\textsc{.pst.ptcp.pl=cop.3pl:S} eat\textsc{.pst-nmlz=3pl:A} eat\textsc{.pst.ptcp.m=cop.3sg.m:O} steal\textsc{.pst-nmlz=3sg:A} steal\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `I [was not around to] witness [if] they (the children) had fallen, [if] they had eaten, or stolen [something].'
\z 
 
\ea \label{JM.30}
\textit{a wextîyekey řûtê bîyênê qûtê bîyênê bîyenşa nebîyenşa. xeberêm nezanênê.} \\ 
\gll a wext-î-ekey řût-ê bîyê=nê qût-ê bîyê=nê bîye=n=şa ne-bîye=n=şa xeber-ê=m ne-zanê=nê \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} naked\textsc{-pl} be\textsc{.ptcp.pl=cop.3pl:S} bare\textsc{-pl} be\textsc{.ptcp.pl=cop.3pl:S} be\textsc{.pst.ptcp.m=cop.3sg.m:S=3pl:NC} \textsc{neg-}be\textsc{.pst.ptcp.m=cop.3sg.m:S=3pl:NC} news\textsc{.m-pl.dir=1sg:A} \textsc{neg-}know\textsc{.pst.ptcp.pl=cop.3pl:PSR} \\ 
\glt `[I was unaware if] they were without clothes; if they were poor, if they had [enough]; if they didn't have [enough]. I didn't know their news.'
\z 
 
\ea \label{JM.31}
\textit{dilê bedbextîne bîyênmê.} \\ 
\gll dilê bedbextî=ne bîyê=nmê \\ 
 inside hardship\textsc{.m=post} be\textsc{.ptcp.pl=cop.1pl:S} \\ 
\glt `We experienced hardship.'
\z 
 
\ea \label{JM.32}
\textit{a wextîyekey eçage hîç xeberêma nebîyenû. telefûnîçû jimarew telefûnî nebêw.} \\ 
\gll a wext-î-ekey e=çage hîç xeber-ê=ma ne-bîye=n=û telefûn=îç=û jimare-û telefûn-î ne-b-ê=û \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} in=there nothing news\textsc{.m-indf=1pl:NC} \textsc{neg-}be\textsc{.pst.ptcp.m=cop.3sg.m:S}=and telephone\textsc{.m=add}=and number\textsc{.m-ez.gen} telephone\textsc{.m-sg.obl} \textsc{neg-}be\textsc{.prs-aug.3sg:S}=and \\ 
\glt `I\footnote{Lit. we}  had no news [of my family] from there. There was no phone or phone number at the time.'
\z 
 
\ea \label{JM.33}
\textit{şesû şiş saʕetê ême be maşîn ca luwεnmê binderʕebasû.} \\ 
\gll şes-û şiş saʕet-ê ême be maşîn ca luwε=nmê binderʕebas=û \\ 
 sixty\textsc{-ez.gen} six hour\textsc{.m-pl.dir} \textsc{1pl} by car\textsc{.m} afterwards go\textsc{.pst.ptcp.pl=cop.1pl:S} \textsc{pn}=and \\ 
\glt `It took us sixty-three hours to reach Bandar Abbas by car.'
\z 
 
\ea \label{JM.34}
\textit{luwεnmê çarekû kerdenma ew des.} \\ 
\gll luwε=nmê çarek=û kerde=n=ma ew des \\ 
 go\textsc{.pst.ptcp.pl=cop.1pl:S} \textsc{pn}=and do\textsc{.pst.ptcp.m=cop.3sg.m:O=1pl:A} \textsc{dem.dist} hand\textsc{.m} \\ 
\glt `We went to Charak and went to the other side.'
\z 
 
\ea \label{JM.35}
\textit{şiş mangê wêma ça karma kerđen.} \\ 
\gll şiş mang(e)-ê wê=ma ça kar=ma kerđe=n \\ 
 six month\textsc{.f-pl.dir} \textsc{reflx=1pl:PSR} there work\textsc{.m=1pl:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `I worked there for [periods of] six months.'
\z 
 
\newpage
\ea \label{JM.36}
\textit{î şiş mange xeberû hîçîma nezanan.} \\ 
\gll î şiş mang(e)=e xeber-û hîç-î=ma ne-zana=n \\ 
 \textsc{dem.prox} six month\textsc{.f=dem} news\textsc{.m-ez.gen} nothing\textsc{-obl.m=1pl:A} \textsc{neg-}know\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `I didn't have any updates [from my family] during those periods of six months.'
\z 
 
\ea \label{JM.37}
\textit{ħaɫîta?} \\ 
\gll ħaɫî=t=a \\ 
 understood\textsc{=2sg:NC=cop.3sg.m:S} \\ 
\glt `You see?'
\z 
 
\ea \label{JM.38}
\textit{extîyarma miştû a farsane bîyen cezîrew kîşî.} \\ 
\gll extîyar=ma mişt-û a fars-a=ne bîye=n cezîre-û kîş-î \\ 
 control\textsc{.m=1pl:PSR} fist\textsc{.m-ez.gen} \textsc{dem.dist} Persian\textsc{-pl.obl=post} be\textsc{.pst.ptcp.m=cop.3sg.m:S} island\textsc{-ez.gen} \textsc{pn-m.sg.obl} \\ 
\glt `We were under the rule of Persians at Kish island.'
\z 
 
\ea \label{JM.39}
\textit{imrîkayî amεnê eça karma pey kerdênê.} \\ 
\gll imrîka-î amε=nê e=ça kar=ma pey kerdê=nê \\ 
 \textsc{pn-adjzr} come\textsc{.pst.ptcp.pl=cop.3pl:S} in=there work\textsc{.m=1pl:A} to do\textsc{.pst.ptcp.pl=cop.3pl:R} \\ 
\glt `Americans came [there]. We had them work [for us].'
\z 
 
\ea \label{JM.40}
\textit{ême çagene saxtmanê saznênmê.} \\ 
\gll ême çagene saxtman-ê sazn-ên-mê \\ 
 \textsc{1pl} there building\textsc{.m-pl.dir} build\textsc{.prs-aug-1pl:A} \\ 
\glt `We built buildings there.'
\z 
 
\ea \label{JM.41}
\textit{kuweyt duwê saʕetê beynşa çenû kuweytî.} \\ 
\gll kuweyt duwê saʕet-ê beyn=ş=a çenû kuweyt-î \\ 
 \textsc{pn} two hour\textsc{.m-pl.dir} between\textsc{=3sg:R=cop.3sg:S} with\textsc{.ez.gen} \textsc{pn-m.sg.obl} \\ 
\glt `Kuwait was two hours away from there.'
\z 
 
\ea \label{JM.42}
\textit{kuweyt, ađîç wêş kêşwerêwa kuweyt.} \\ 
\gll kuweyt ađ=îç wê=ş kêşwer-êw=a kuweyt \\ 
 \textsc{pn} \textsc{3sg.m.dir=add} \textsc{reflx=3sg:PSR} country\textsc{-indf=cop.3sg.m:S} \textsc{pn} \\ 
\glt `Kuwait is a country on its own.'
\z 
 
\ea \label{JM.43}
\textit{a zemane kuweyt ce girđû kêşwereka dewɫetmenter bê.} \\ 
\gll a zeman=e kuweyt ce girđ-û kêşwer-eka dewɫetmen-ter b-ê \\ 
 \textsc{dem.dist} time\textsc{.m=dem} \textsc{pn} from all\textsc{-ez.gen} country\textsc{-def.pl.obl} rich\textsc{-cmpr} be\textsc{.prs-aug.3sg:S} \\ 
\glt `Back then, Kuwait was the richest of all countries.'
\z 
 
\ea \label{JM.44}
\textit{ce taranî kuweyt dewɫetmenter bê.} \\ 
\gll ce taran-î kuweyt dewɫetmen-ter b-ê \\ 
 from \textsc{pn-m.sg.obl} \textsc{pn} rich\textsc{-cmpr} be\textsc{.prs-aug.3sg:S} \\ 
\glt `Kuwait was more affluent than Tehran.'
\z 
 
\ea \label{JM.45}
\textit{ême cezîre çenû a cezîrey çeneş bênmê duwê saʕetê beyniş bê.} \\ 
\gll ême cezîre çenû a cezîre-î çene=ş b-ên-mê duwê saʕet-ê beyn=iş b-ê \\ 
 \textsc{1pl} island with\textsc{.ez.gen} \textsc{dem.dist} island\textsc{-m.sg.obl} in\textsc{=3sg:PSR} be\textsc{.prs-aug-1pl:S} two hour\textsc{.m-pl.dir} distance\textsc{.m=3sg:PSR} be\textsc{.prs-aug.3sg:S} \\ 
\glt `It (Kuwait) was two hours away from the island we were on.'
\z 
 
\ea \label{JM.46}
\textit{řowê sî penc timenê kar kerênmê.} \\ 
\gll řo-ê sî penc timen-ê kar ker-ên-mê \\ 
 day\textsc{.m-indf} thirty five \textsc{pn-pl.dir} task\textsc{.m} do\textsc{.prs-aug-1pl:A} \\ 
\glt `We used to work for a daily salary of thirty-five tomans.'
\z 
 
\ea \label{JM.47}
\textit{penc řowê çane min řemezanim gêrten.} \\ 
\gll penc řo-ê ç=a=ne min řemezan=im gêrte=n \\ 
 five day\textsc{.m-pl.dir} in=\textsc{dem.dist=post} \textsc{1sg} \textsc{pn=1sg:A} grab\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `[while there] I fasted five days during Ramadan.'
\z 
 
\ea \label{JM.48}
\textit{haɫîta?} \\ 
\gll haɫî=t=a \\ 
 understood\textsc{=2sg:NC=cop.3sg.m:S} \\ 
\glt `You see?'
\z 
 
\ea \label{JM.49}
\textit{řowê sî penc timenê şesû yerê saɫê çêweɫî kero êtir meyanû saɫû çinê bê.} \\ 
\gll řo-ê sî penc timen-ê şes=û yerê saɫ(e)-ê çêweɫî ker-o êtir me-zan-û saɫ(e)-û çinê b-ê \\ 
 day\textsc{.m-indf} thirty five \textsc{pn-pl.dir} sixty=and three year\textsc{.f-pl.dir} ago do\textsc{.prs.ind-3sg:A} \textsc{disc.ptcl} \textsc{neg.ind-}know\textsc{.prs-1sg:A} year\textsc{.f-ez.gen} how\_many\textsc{.obl} be\textsc{.prs-aug.3sg:S} \\ 
\glt `A daily salary of thirty-five tomans -- it was sixty-three years ago. I don't know which year it was.'
\z 
 
\ea \label{JM.50}
\textit{ême çane karma kerđenû.} \\ 
\gll ême ç=a=ne kar=ma kerđe=n=û \\ 
 \textsc{1pl} in=\textsc{dem.dist=post} work\textsc{.m=1pl:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O}=and \\ 
\glt `We worked there.'
\z 
 
\ea \label{JM.51}
\textit{wextê amεnmêwe î mentêqe êtir çê musteqerê bîyênmê.} \\ 
\gll wext-ê amε=nmê=we î mentêqe êtir çê musteqer-ê bîyê=nmê \\ 
 time\textsc{.m-indf} come\textsc{.pst.ptcp.pl=cop.1pl:S=compl} \textsc{dem.prox} region\textsc{.m} \textsc{disc.ptcl} here settled\textsc{-pl} be\textsc{.ptcp.pl=cop.1pl:S} \\ 
\glt `When I returned to this region, I settled here.'
\z 
 
\ea \label{JM.52}
\textit{luwεnmê benenû amεnmêwew.} \\ 
\gll luwε=nmê benen=û amε=nmê=we=û \\ 
 go\textsc{.pst.ptcp.pl=cop.1pl:S} \textsc{pn}=and come\textsc{.pst.ptcp.pl=cop.1pl:S=compl}=and \\ 
\glt `We would go to Banan (the summer habitat) and return [Home].'
\z 
 
\ea \label{JM.53}
\textit{ʕerzû biraw wêm kerû. amεnmêwe î mentêqe.} \\ 
\gll ʕerz-û bira-û wê=m k\'er-û amε=nmê=we î mentêqe \\
 petition\textsc{.m-ez.gen} brother\textsc{.m-ez.gen} \textsc{reflx=1sg:PSR} do\textsc{.prs.sbjv-1sg:A} come\textsc{.pst.ptcp.pl=cop.1pl:S=compl} \textsc{dem.prox} region\textsc{.m} \\ 
\glt `Let me tell you. I returned to this region.'
\z 
 
\ea \label{JM.54}
\textit{êtir îsey amεnmê ta wê yawεnmê be kemûkuřîyû hewramanî.} \\ 
\gll êtir îse-î amε=nmê ta wê yawε=nmê be kemûkořî-û hewraman-î \\ 
 \textsc{disc.ptcl} now\textsc{-m.sg.obl} come\textsc{.pst.ptcp.pl=cop.1pl:S} until there arrive\textsc{.pst.ptcp.pl=cop.1pl:S} by poverty\textsc{.m-ez.gen} \textsc{pn-m.sg.obl} \\ 
\glt `I have made it so far, [despite] poverty and destitution in Hewraman.'
\z 
 
\ea \label{JM.55}
\textit{êtir îse weʕzû hewramanî ta menteqê terê kuçê xastera.} \\ 
\gll êtir îse weʕz(e)-û hewraman-î ta menteq(e)-ê ter-ê kuç-ê xas-ter=a \\ 
 \textsc{disc.ptcl} now situation\textsc{.f-ez.gen} \textsc{pn-m.sg.obl} than region\textsc{-pl.dir} other\textsc{-pl.dir} little\textsc{-indf} good\textsc{-cmpr=cop.3sg.m:S} \\ 
\glt `The situation in Hewraman is a bit better than in other regions.'
\z 
 
\ea \label{JM.56}
\textit{çêweɫ nebîyey bê êtir îseyç xû nîya.} \\ 
\gll çêweɫ nebîyey b-ê êtir îse=îç xû nîy(e)=a \\ 
 previously poverty\textsc{.m} be\textsc{.prs-aug.3sg:S} \textsc{disc.ptcl} now\textsc{=add} \textsc{disc.ptcl} \textsc{neg.exist=cop.3sg.m:S} \\ 
\glt `In the past, there was poverty. But even nowadays, there is poverty.'
\z 
 
\ea \label{JM.57}
\textit{a wextîyekey ême îne zinđigîş pene kermê be nebîyey.} \\ 
\gll a wext-î-ekey ême îne zinđigî=ş pene ker-mê be nebîyey \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} \textsc{1pl} \textsc{dem.prox.m.3sg.dir} life\textsc{.m=3sg:R} with do\textsc{.prs.ind-1pl:A} in poverty\textsc{.m} \\ 
\glt `Well, we live in poverty.'
\z 
 
\ea \label{JM.58}
\textit{be kemûkořî nebîyey zinđigîş pene kermê.} \\ 
\gll be kemûkořî nebîyey zinđigî=ş pene ker-mê \\ 
 in poverty\textsc{.m} poverty\textsc{.m} life\textsc{.m=3sg:R} with do\textsc{.prs.ind-1pl:A} \\ 
\glt `We live in poverty and destitution.'
\z 
 
\ea \label{JM.59}
\textit{enane ême waryatêma nîya.} \\ 
\gll ena=ne ême waryat-ê=ma nîy(e)=a \\ 
 otherwise\textsc{=post} \textsc{1pl} income\textsc{.m-indf=1pl:NC} \textsc{neg.exist=cop.3sg.m:S} \\ 
\glt `because we have no income.'
\z 
 
 \newpage
\ea \label{JM.60}
\textit{warîyatma nîya kasbîma nîya karma nîya.} \\ 
\gll warîyat=ma nîy(e)=a kasbî=ma nîy(e)=a kar=ma nîy(e)=a \\ 
 income\textsc{.m=1pl:NC} \textsc{neg.exist=cop.3sg.m:S} business\textsc{.m=1pl:NC} \textsc{neg.exist=cop.3sg.m:S} work\textsc{.m=1pl:NC} \textsc{neg.exist=cop.3sg.m:S} \\ 
\glt `We have no income, no business, no jobs.'
\z 
 
\ea \label{JM.61}
\textit{herkes tawo bilo koɫê baro koɫê milwênê koɫê duwê milwênê.} \\ 
\gll herkes taw-o bi-l-o koɫ-ê b-ar-o koɫ-ê milwên-ê koɫ-ê duwê milwên-ê \\ 
 anyone can\textsc{.prs.ind-3sg:A} \textsc{sbjv-}go\textsc{.prs-3sg:S} load\textsc{.m-indf} \textsc{sbjv-}bring\textsc{.prs-3sg:A} load\textsc{.m-indf} million\textsc{.m-indf} load\textsc{.m-indf} two million\textsc{.m-pl.dir} \\ 
\glt `Anyone who can go [to the mountains] as a porter can earn one or two million [tomans] per load.'
\z 
 
\ea \label{JM.62}
\textit{herkesîç metawo pêse min ke nelabû êtir ane hîçê derameđêş nîya.} \\ 
\gll herkes=îç me-taw-o pêse min ke ne-la=b-û êtir ane hîç-ê derameđ-ê=ş nîy(e)=a \\ 
 anyone\textsc{=add} \textsc{neg.ind-}can\textsc{.prs-3sg:A} like \textsc{1sg} \textsc{rel} \textsc{neg-}go\textsc{.pst.ptcp.m}=be\textsc{.prs-1sg:S} \textsc{disc.ptcl} \textsc{dem.dist.m.3sg.dir} nothing\textsc{-indf} income\textsc{-indf=3sg:NC} \textsc{neg.exist=cop.3sg.m:S} \\ 
\glt `Anyone who is not able [to work as a porter], like me, who has probably not been a porter, well, he has no income.'
\z 
 
\ea \label{JM.63}
\textit{epase êtir.} \\ 
\gll e=pase êtir \\ 
 \textsc{emph}=like\_this \textsc{disc.ptcl} \\ 
\glt `Well, [the situation is] like this.'
\z 
