\ea \label{JE.1}
\textit{dey çêgeyç qeyîm ʕaɫifşa kennenû.} \\ 
\gll dey çêge=îç qeyîm ʕaɫif=şa kenne=n=û \\ 
 \textsc{disc.ptcl} here\textsc{=add} old\_time grass\textsc{.m.sg.dir=3pl:A} mow\textsc{.pst.ptcp.m=cop.3sg.m:O}=and \\ 
\glt `In the past, they (people) mowed grass.'
\z 
 
\ea \label{JE.2}
\textit{ħeywanşa wey kerdenû.} \\ 
\gll ħeywan=şa wey kerde=n=û \\ 
 animal\textsc{.m.sg.dir=3pl:A} raising do\textsc{.pst.ptcp.m=cop.3sg.m:O}=and \\ 
\glt `They raised animals.'
\z 
 
\ea \label{JE.3}
\textit{hewarêşa wişkinεnê.} \\ 
\gll hewar-ê=şa wişkinε=nê \\ 
 summer\_habitat\textsc{.m-pl.dir=3pl:A} scour\textsc{.pst.ptcp.pl=cop.3pl:O} \\ 
\glt `They scoured the summer habitats [searching for food etc.].'
\z 
 
\ea \label{JE.4}
\textit{î dega toş vînî çoɫe bîyêne.} \\ 
\gll î dega to=ş vîn-î çoɫ-e bîyê=ne \\ 
 \textsc{dem.prox} village{\textsc{.f}} \textsc{2sg=3sg:O} see\textsc{.prs.ind-2sg:A} deserted\textsc{-f} be\textsc{.pst.ptcp.f=cop.3sg.f:S} \\ 
\glt `This village, which you see, was deserted.'
\z 
 
\ea \label{JE.5}
\textit{ħîç nebîyen.} \\ 
\gll ħîç ne-bîye=n \\ 
 nothing \textsc{neg-}be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `There was nothing [here].'
\z 
 
\ea \label{JE.6}
\textit{awîş nebîyêne.} \\ 
\gll awî=ş ne-bîyê=ne \\ 
 water\textsc{.f.sg.dir=3sg:NC} \textsc{neg-}be\textsc{.pst.ptcp.f=cop.3sg.f:S} \\ 
\glt `It (the village) had no water.'
\z 
 
\ea \label{JE.7}
\textit{hîç awû çîwiş nebîyen.} \\ 
\gll hîç aw(î)=û çîw=iş ne-bîye=n \\ 
 nothing water\textsc{.f}=and thing\textsc{.m=3sg:NC} \textsc{neg-}be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `It had no water or anything.'
\z 
 
\ea \label{JE.8}
\textit{luwεnê hewar.} \\ 
\gll luwε=nê hewar \\ 
 go\textsc{.pst.ptcp.pl=cop.3pl:S} summer\_habitat\textsc{.m} \\ 
\glt `They (people) would go to the summer habitat.'
\z 
 
\ea \label{JE.9}
\textit{îse luwεnê ta penca çil şewê zim[san], çil şewê payîz şîyen ca koçşa kerdeno.} \\ 
\gll îse luwε=nê ta penca çil şew(e)-ê zimsan çil şew(e)-ê payîz şîye=n ca koç=şa kerde=n=o \\ 
 now go\textsc{.pst.ptcp.pl=cop.3pl:S} until fifty forty night\textsc{.f-pl.dir} winter\textsc{.m} forty night\textsc{.f-pl.dir} autumn go\textsc{.pst.ptcp.m=cop.3sg.m:S} afterwards migration\textsc{.m.sg.dir=3pl:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O=compl} \\ 
\glt `They used to go [to the summer habitat] at this time of year. They would migrate back [to the village] around 40 or 50 days after autumn [began].'
\z 
 
\ea \label{JE.10}
\textit{her hewarne bîyênê.} \\ 
\gll her hewar=ne bîyê=nê \\ 
 still summer\_habitat\textsc{.m=post} be\textsc{.ptcp.pl=cop.3pl:S} \\ 
\glt `They were still at the summer habitat [until  40 or 50 days after autumn began].'
\z 
 
\ea \label{JE.11}
\textit{êtir pagene girđkes ħeywandarîş kerdenû.} \\ 
\gll êtir pagene girđkes ħeywandarî=ş kerde=n=û \\ 
 \textsc{disc.ptcl} there everyone animal\_husbandry\textsc{=3sg:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O}=and \\ 
\glt `There, everyone tended animals.'
\z 
 
\ea \label{JE.12}
\textit{girđkes ʕaɫfiş kennenû.} \\ 
\gll girđkes ʕaɫif=iş kenne=n=û \\ 
 everyone grass\textsc{.m.sg.dir=3sg:A} mow\textsc{.pst.ptcp.m=cop.3sg.m:O}=and \\ 
\glt `Everyone mowed grass.'
\z 
 
\ea \label{JE.13}
\textit{ʕaɫfiş pêtenû ta koçşa kerden ta ʕaɫifşa ardeno.} \\ 
\gll ʕaɫif=iş pête=n=û ta koç=şa kerde=n ta ʕaɫif=şa arde=n=o \\ 
 grass\textsc{.m.sg.dir=3sg:A} fold\textsc{.pst.ptcp.m=cop.3sg.m:O}=and until migration\textsc{.m.sg.dir=3pl:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O} until grass\textsc{.m.sg.dir=3pl:A} bring\textsc{.pst.ptcp.m=cop.3sg.m:O=compl} \\ 
\glt `[and] made grass into stacks until they returned [to the village]; until they brought back grass'
\z 
 
\ea \label{JE.14}
\textit{êtir zimsan aman.} \\ 
\gll êtir zimsan ama=n \\ 
 \textsc{disc.ptcl} winter\textsc{.m} come\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `Then the winter came.'
\z 
 
\ea \label{JE.15}
\textit{zimsanîç aman ta luwεnê hane teqanû.} \\ 
\gll zimsan=îç ama=n ta luwε=nê hane teqa=n=û \\ 
 winter\textsc{.m=add} come\textsc{.pst.ptcp.m=cop.3sg.m:S} until go\textsc{.pst.ptcp.pl=cop.3pl:S} water\_spring\textsc{.m} gush\textsc{.pst.ptcp.m=cop.3sg.m:S}=and \\ 
\glt `The winter came. They would go. The water springs would gush with water.'
\z 
 
\ea \label{JE.16}
\textit{luwεnê awîşa ardêne. be kune awîşa ardêne.} \\ 
\gll luwε=nê awî=şa ardê=ne be kune awî=şa ardê=ne \\ 
 go\textsc{.pst.ptcp.pl=cop.3pl:S} water\textsc{.f.sg.dir=3pl:A} bring\textsc{.pst.ptcp.f=cop.3sg.f:O} by clay\_pot\textsc{.m} water\textsc{.f.sg.dir=3pl:A} bring\textsc{.pst.ptcp.f=cop.3sg.f:O} \\ 
\glt `They used to go and bring water. They used to fetch water using clay pots.'
\z 
 
\ea \label{JE.17}
\textit{be kune awîşa ardênew. nîyεneşarew.} \\ 
\gll be kune awî=şa ardê=ne=û nîyε=ne=şa=re=û \\ 
 by clay\_pot\textsc{.m} water\textsc{.f.sg.dir=3pl:A} bring\textsc{.pst.ptcp.f=cop.3sg.f:O}=and put\textsc{.pst.ptcp.f=cop.3sg.f:O=3pl:A=povb}=and \\ 
\glt `They fetched water in clay pots. They unloaded [the water].'
\z 
 

\ea \label{JE.18}
\textit{desû dîmşa gêrten nimaşa kerdêne.} \\ 
\gll des=û dîm=şa gêrte=n nima=şa kerdê=ne \\ 
 hand\textsc{.m}=and face\textsc{.m=3pl:A} grab\textsc{.pst.ptcp.m=cop.3sg.m:O} prayer\textsc{.f.sg.dir=3pl:A} do\textsc{.pst.ptcp.f=cop.3sg.f:O} \\ 
\glt `They performed ablutions [and] prayed.'
\z 
 
\ea \label{JE.19}
\textit{nanşa peten.} \\ 
\gll nan=şa pete=n \\ 
 bread\textsc{.m.sg.dir=3pl:A} bake\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `They baked bread.'
\z 
 
\ea \label{JE.20}
\textit{çayşa kerden.} \\ 
\gll çay=şa kerde=n \\ 
 tea\textsc{.m.sg.dir=3pl:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `They made tea.'
\z 
 
\ea \label{JE.21}
\textit{awê her kunene her netemamyεne.} \\ 
\gll aw(î)-ê her kune=ne her ne-temamyε=ne \\ 
 water\textsc{.f-sg.obl} \textsc{emph} clay\_pot\textsc{.m=post} \textsc{emph} \textsc{neg-}finish\textsc{.pst.ptcp.f=cop.3sg.f:S} \\ 
\glt `The water [was] in the clay pots. It would not run out [quickly].'
\z 
 
\ea \label{JE.22}
\textit{her firê bîyêne êtir epase.} \\ 
\gll her fir(e)-ê bîyê=ne êtir e=pase \\ 
 \textsc{emph} plentiful\textsc{-f} be\textsc{.pst.ptcp.f=cop.3sg.f:S} \textsc{disc.ptcl} \textsc{emph}=like\_this \\ 
\glt `It (the water) was plentiful. Well, [it was] like this.'
\z 
 
\ea \label{JE.23}
\textit{çêgeyç awî her nebîyêne bînder.} \\ 
\gll çêge=îç awî her ne-bîyê=ne bînder \\ 
 here\textsc{=add} water\textsc{.f.sg.dir} \textsc{emph} \textsc{neg-}be\textsc{.pst.ptcp.f=cop.3sg.f:S} at\_all \\ 
\glt `There was no water here.'
\z 
 
\ea \label{JE.24}
\textit{luwεnê dûrew awîşa ardênê.} \\ 
\gll luwε=nê dûr=ew awî=şa ardê=nê \\ 
 go\textsc{.pst.ptcp.pl=cop.3pl:S} far\textsc{=post} water\textsc{.pl.dir=3pl:A} bring\textsc{.pst.ptcp.pl=cop.3pl:O} \\ 
\glt `They (people) would go far and fetch water.'
\z 
 
\ea \label{JE.25}
\textit{yanew êtir ne yane bîyenû.} \\ 
\gll yane=û êtir ne yane bîye=n=û\\ 
 house\textsc{.m}=and \textsc{disc.ptcl} neither house\textsc{.m.sg.dir}  be\textsc{.pst.ptcp.m=cop.3sg.m:S}=and \\ 
\glt `As for houses, there was no [proper] house [to live in].'
\z 
 
\ea \label{JE.26}
\textit{denîwle girđ yo dilne yane xakîne bîyen.} \\ 
\gll denîwle girđ yo dilne yan(e)-e xakî=ne bîye=n \\ 
 small all one\textsc{.m} inside house\textsc{.m-ez.cmpd} earthen\textsc{=post} be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `[In] a small [house], inside [it], all [the members] lived\footnote{Lit. were} in an earthen house.'
\z 
 
\ea \label{JE.27}
\textit{de neferê serû camoɫê çaştîşa nîyεnere.} \\ 
\gll de nefer-ê ser-û camoɫ(e)-ê çaştî=şa nîyε=ne=re \\ 
 ten person\textsc{.m-pl.dir} on\textsc{-ez.gen} small\_pot\textsc{-indf} meal\textsc{.f.sg.dir=3pl:A} put\textsc{.pst.ptcp.f=cop.3sg.f:O=povb} \\ 
\glt `They would cook meals in a single small pot for ten persons.'
\z 
 
\ea \label{JE.28}
\textit{şûnû de neferekeyre her çemçêşa bîyen.} \\ 
\gll şûn-û de nefer-ekey=re her çemç(e)-ê=şa bîye=n \\ 
 after\textsc{-ez.gen} ten person\textsc{.m-def.m.sg.obl=post} only spoon\textsc{.m-indf=3pl:NC} be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `Then, there was only one spoon for the ten people.'
\z 
 
\newpage
\ea \label{JE.29}
\textit{çemçekene girđ nobenobey çaştîşa wardêne.} \\ 
\gll çemçe-(e)ke=ne girđ nobenobey çaştî=şa wardê=ne \\ 
 spoon\textsc{.m-def.m.sg.dir=post} all one\_by\_one meal\textsc{.f.sg.dir=3pl:A} eat\textsc{.pst.ptcp.f=cop.3sg.f:O} \\ 
\glt `They would take turns to eat their meal using the spoon.'
\z 
 

\ea \label{JE.30}
\textit{êtir yo xeleş bîyenû.} \\ 
\gll êtir yo xele=ş bîye=n=û\\ 
 \textsc{disc.ptcl} one\textsc{.m} grain\textsc{.m.sg.dir=3sg:NC} be\textsc{.pst.ptcp.m=cop.3sg.m:S}=and \\ 
\glt `One [person] had grains [as an agricultural product].'
\z 
 
\ea \label{JE.31}
\textit{yo bancanîş bîyêne.} \\ 
\gll yo bancanî=ş bîyê=ne \\ 
 one\textsc{.m} tomato\textsc{.f.sg.dir=3sg:NC} be\textsc{.pst.ptcp.f=cop.3sg.f:S} \\ 
\glt `One [person] had tomatoes.'
\z 
 
\ea \label{JE.32}
\textit{yo sêfêş bîyênew.} \\ 
\gll yo sêfê=ş bîyê=ne=û \\ 
 one\textsc{.m} potato\textsc{.f.sg=3sg:NC} be\textsc{.pst.ptcp.f=cop.3sg.f:S}=and \\ 
\glt `One had potatoes.'
\z 
 
\ea \label{JE.33}
\textit{luwεnê eweɫû wextî luwεnê hêzmîşa biřyênê.} \\ 
\gll luwε=nê eweɫ-û wext-î luwε=nê hêzmî=şa biřyê=nê \\ 
 go\textsc{.pst.ptcp.pl=cop.3pl:S} first\textsc{-ez.gen} time\textsc{.m-sg.obl} go\textsc{.pst.ptcp.pl=cop.3pl:S} firewood\textsc{.pl.dir=3pl:A} cut\textsc{.pst.ptcp.pl=cop.3pl:O} \\ 
\glt `Early in the morning, they would go and chop wood.'
\z 
 
\ea \label{JE.34}
\textit{be derze hêzmîşa ardênê.} \\ 
\gll be derze hêzmî=şa ardê=nê \\ 
 by shoulder\textsc{.m} firewood\textsc{.pl.dir=3pl:A} bring\textsc{.pst.ptcp.pl=cop.3pl:O} \\ 
\glt `They fetched firewood on their shoulders.'
\z 
 
\newpage·
\ea \label{JE.35}
\textit{be hesere hêzmîşa ardênê pey zimsanî.} \\ 
\gll be hesere hêzmî=şa ardê=nê pey zimsan-î \\ 
 by mule\textsc{.f} firewood\textsc{.pl.dir=3pl:A} bring\textsc{.pst.ptcp.pl=cop.3pl:O} for winter\textsc{.m-sg.obl} \\ 
\glt `They fetched firewood for the winter on mules.'
\z 
 
\ea \label{JE.36}
\textit{yaneyçşa pêř kerdenn hêzmî.} \\ 
\gll yane=yç=şa pêř kerde=n hêzmî \\ 
 house\textsc{.m.sg.dir=add=3pl:A} full do\textsc{.pst.ptcp.m=cop.3sg.m:O} firewood\textsc{.f} \\ 
\glt `They filled their houses with firewood.'
\z 
 
\ea \label{JE.37}
\textit{hêzmî êtir zimsan aman hurêsênê yerê awîr, awîre kuçêşa nîyεnêre bî kipê.} \\ 
\gll hêzmî êtir zimsan ama=n hur-êsê=nê yerê awîr awîr-e kuç-ê=şa nîyε=nê=re bî-∅ kip-ê \\ 
 firewood\textsc{.f} \textsc{disc.ptcl} winter\textsc{.m} come\textsc{.pst.ptcp.m=cop.3sg.m:S} \textsc{pvb-}rise\textsc{.pst.ptcp.pl=cop.3pl:S} three fire\textsc{.m} fire\textsc{.m-ez.cmpd} little\textsc{-pl.dir=3pl:A} put\textsc{.pst.ptcp.pl=cop.3pl:O=povb} become\textsc{.pst-3sg.m:S} sealed\textsc{-pl} \\ 
\glt `Firewood …  Then when it was winter they would get up and start three small fires [and the fire would] die down.'
\z 
 
\ea \label{JE.38}
\textit{nîyanşare. êtir çayşa kerden be nanekeyû.} \\ 
\gll nîya=n=şa=re êtir çay=şa kerde=n be nan-ekey=û \\ 
 put\textsc{.pst.ptcp.m=cop.3sg.m:O=3pl:A=povb} \textsc{disc.ptcl} tea\textsc{.m.sg.dir=3pl:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O} with bread\textsc{.m-def.m.sg.obl}=and \\ 
\glt `They started [the fire].  Then, they made tea [to be served] with the food.'
\z 
 
\ea \label{JE.39}
\textit{çaştekêşa kerdêne be awîrgakêw.} \\ 
\gll çaşt(î)-ekê=şa kerdê=ne be awîrga-(e)kê=û \\ 
 meal\textsc{.f-def.f.sg=3pl:A} do\textsc{.pst.ptcp.f=cop.3sg.f:O} on hearth\textsc{.f-def.f.sg}=and \\ 
\glt `They made the food on the hearth.'
\z 

\newpage
\ea \label{JE.40}
\textit{her sikeɫ bîyen. berqû çîw nebîyen.} \\ 
\gll her sikeɫ bîye=n berq=û çîw ne-bîye=n \\ 
 just ember\textsc{.m.sg.dir} be\textsc{.pst.ptcp.m=cop.3sg.m:S} electricity\textsc{.m.dir}=and thing\textsc{.m.dir} \textsc{neg-}be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `There were only embers [to light the house]. There was no electricity or anything.'
\z 
 
\ea \label{JE.41}
\textit{deseçirawê bîyen. deseçiraweke çeqnanşa keɫekere cîyaw řoşnayîşa.} \\ 
\gll deseçiraw-ê bîye=n deseçiraw-eke çeqna=n=şa keɫeke=re cîyaw řoşnayî=şa \\ 
 hand\_lump\textsc{-indf} be\textsc{.pst.ptcp.m=cop.3sg.m:S} hand\_lump\textsc{-def.m.sg.dir} install\textsc{.pst.ptcp.m=cop.3sg.m:O=3pl:A} wall\textsc{=post} in\_place\_of light\textsc{.m=3pl:PSR} \\ 
\glt `There were hand-held oil lamps. They would install the hand-held oil lamps on the wall instead of lights.'
\z 
 
\ea \label{JE.42}
\textit{çirawe desîşa, çirawekeşa weş kerden bila pey yanew pîr şelîyarî pey çîwî baranşo.} \\ 
\gll çiraw-e des-î=şa çiraw-eke=şa weş-∅ kerde=n bi-l-a pey yane-û pîr şelîyar-î pey çîw-î b-ar-an=ş=o \\ 
 lamp\textsc{.m-ez.cmpd} hand\textsc{-adjzr=3pl:PSR} lamp\textsc{.m-def.m.sg.dir=3pl:A} good\textsc{-m} do\textsc{.pst.ptcp.m=cop.3sg.m:O} \textsc{sbjv-}go\textsc{.prs-3pl:S} to house\textsc{.m-ez.gen} \textsc{pn} \textsc{pn-m.sg.obl} to thing\textsc{.m-sg.obl} \textsc{sbjv-}bring\textsc{.prs-3pl:A=3sg:O=compl} \\ 
\glt `They lit the hand-held oil lamps to go to Pir Shaliyar's shrine\footnote{Lit. house} [and] to fetch things.'
\z 
 
\ea \label{JE.43}
\textit{eđaw tatew min her beđbext bîyen.} \\ 
\gll eđa=û tate-û min her beđbext bîye=n \\ 
 mother\textsc{.f}=and father\textsc{.m-ez.gen} \textsc{1sg} \textsc{emph} poor be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `My parents\footnote{Lit. my mother and father} were poor.'
\z 
 
\ea \label{JE.44}
\textit{denîwlê yane bîyen ħewt kiɫfetêş bîyênê.} \\ 
\gll denîwlê yane bîye=n ħewt kiɫfet-ê=ş bîyê=nê \\ 
 small house\textsc{.m.sg.dir} be\textsc{.pst.ptcp.m=cop.3sg.m:S} seven child\textsc{-pl.dir=3sg:NC} be\textsc{.ptcp.pl=cop.3pl:S} \\ 
\glt `There was a small house [for us to live in]. We were seven children.'
\z 
 
\ea \label{JE.45}
\textit{camoɫêw çaştîş nîyεnere.} \\ 
\gll camoɫê-û çaştî=ş {} nîyε=ne=re \\ 
 small\_pot\textsc{-ez.gen} meal\textsc{.f.sg.dir=3sg:A} {} put\textsc{.pst.ptcp.f=cop.3sg.f:O=povb} \\ 
\glt `She (my mother) would make a small pot of food.'
\z 
 
\ea \label{JE.46}
\textit{penc çemçêşa nîyanre.} \\ 
\gll penc çemç(e)-ê=şa nîya=n=re \\ 
 five spoon\textsc{.m-pl.dir=3pl:A} put\textsc{.pst.ptcp.m=cop.3sg.m:O=povb} \\ 
\glt `They (my family) had set five spoons [on the tablecloth].'
\z 
 
\ea \label{JE.47}
\textit{nobenobey çemçêşa bîyen.} \\ 
\gll nobenobey çemç(e)-ê=şa bîye=n \\ 
 one\_by\_one spoon\textsc{.m-indf=3pl:NC} be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `They would [eat with] the spoons one after another.'
\z 
 
\ea \label{JE.48}
\textit{nanşa wardenû çayşa warden.} \\ 
\gll nan=şa warde=n=û çay=şa warde=n \\ 
 bread\textsc{.m.dir=3pl:A} eat\textsc{.pst.ptcp.m=cop.3sg.m:O}=and tea\textsc{.m.dir=3pl:A} drink\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `They would eat food and drink tea.'
\z 
 
\ea \label{JE.49}
\textit{watenşa, ``xwa gîyan hemdû sena pey to.} \\ 
\gll wate=n=şa xwa gîyan hemd=û sena pey to \\ 
 say\textsc{.pst.ptcp.m=cop.3sg.m:O=3pl:A} God\textsc{.m} dear\textsc{.voc} praise=and admiration for \textsc{2sg} \\ 
\glt `[Then] they would say, ``Dear God, may our praises be for you!'
\z 
 
\newpage
\ea \label{JE.50}
\textit{î dowînete dεnmêw. î şelemînête dεnmê.} \\ 
\gll î dowîne=t=e dε=nmê=û î şelemînê=t=e dε=nmê \\ 
 \textsc{dem.prox} \textsc{pn=2sg:A=dem} give\textsc{.pst.ptcp.pl=cop.1pl:R}=and \textsc{dem.prox} \textsc{pn.f.sg.obl=2sg:A=dem} give\textsc{.pst.ptcp.pl=cop.1pl:R} \\ 
\glt `You gave us this dwina.\footnote{Dish made of cracked wheat and diluted yoghurt.}'
\z 
 
\ea \label{JE.51}
\textit{î şelemînête dεnmê.} \\ 
\gll î şelemînê=t=e dε=nmê \\ 
 \textsc{dem.prox} \textsc{pn.f.sg.obl=2sg:A=dem} give\textsc{.pst.ptcp.pl=cop.1pl:R} \\ 
\glt `you gave us this shalamina.\footnote{Dish made of cracked wheat and turnip.}'
\z 
 
\ea \label{JE.52}
\textit{î xeplête dεnmê.} \\ 
\gll î xeplê=t=e dε=nmê \\ 
 \textsc{dem.prox} \textsc{corn\_bread.f.sg.obl=2sg:A=dem} give\textsc{.pst.ptcp.pl=cop.1pl:R} \\ 
\glt `You gave us this xapla (a bread made of corn flour).'
\z 
 
\ea \label{JE.53}
\textit{înêma wardêne.} \\ 
\gll înê=ma wardê=ne \\ 
 \textsc{dem.prox.f.3sg.dir=1pl:A} eat\textsc{.pst.ptcp.f=cop.3sg.f:O} \\ 
\glt `We have eaten this [meal].'
\z 
 
\ea \label{JE.54}
\textit{xwa hemdû sena pey to!''} \\ 
\gll xwa hemd=û sena pey to \\ 
 God\textsc{.m} praise=and admiration for \textsc{2sg} \\ 
\glt `God, may our praises be for you!'''
\z 
 
\ea \label{JE.55}
\textit{pase zawlêşa wey kerdê.} \\ 
\gll pase zawlê=şa wey kerd-ê \\ 
 like\_this child\textsc{.pl.dir=3pl:A} raising do\textsc{.pst-3pl:O} \\ 
\glt `They raised children in this way.'
\z 
 
\ea \label{JE.56}
\textit{çaweɫî sextî nebîyen.} \\ 
\gll çaweɫ-î sextî ne-bîye=n \\ 
 in\_the\_past\textsc{-m.sg.obl} difficulty\textsc{.m.dir} \textsc{neg-}be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `In the past there was no hardship.'
\z 
 
\ea \label{JE.57}
\textit{çaweɫî hîç nebîyen.} \\ 
\gll çaweɫ-î hîç ne-bîye=n \\ 
 in\_the\_past\textsc{-m.sg.obl} nothing \textsc{neg-}be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `In the past there was nothing.'
\z 
 
\ea \label{JE.58}
\textit{temaʕe nebîyêne.} \\ 
\gll temaʕe ne-bîyê=ne \\ 
 greed\textsc{.f} \textsc{neg-}be\textsc{.pst.ptcp.f=cop.3sg.f:S} \\ 
\glt `There was no greed.'
\z 
 
\ea \label{JE.59}
\textit{seʕbê ta wêrega min duwê nanê bê.} \\ 
\gll seʕb(e)-ê ta wêrega min duwê nan-ê {} b-ê \\ 
 morning\textsc{-f.sg.obl} until evening \textsc{1sg} two meal\textsc{.m-pl.dir} {} be\textsc{.prs-aug.3sg:S} \\ 
\glt `I, there were a few\footnote{Lit. two} [loads of] bread [for me to cook] from morning to evening.'
\z 
 
\ea \label{JE.60}
\textit{nanim peten.} \\ 
\gll nan=im pete=n \\ 
 bread\textsc{.m.sg.dir=1sg:A} bake\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `I would bake bread.'
\z 
 
\ea \label{JE.61}
\textit{awîm ardêne.} \\ 
\gll awî=m ardê=ne \\ 
 water\textsc{.f.sg.dir=1sg:A} bring\textsc{.pst.ptcp.f=cop.3sg.f:O} \\ 
\glt `I brought water.'
\z 
 
\ea \label{JE.62}
\textit{luwεna ħeywanim bîyen.} \\ 
\gll luwε=na ħeywan=im bîye=n \\ 
 go\textsc{.pst.ptcp.f=cop.1sg:S} animal\textsc{.m.sg.dir=1sg:NC} be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `I went … I had animals.'
\z 
 
\ea \label{JE.63}
\textit{ħeywandarîm kerden.} \\ 
\gll ħeywandarî=m kerde=n \\ 
 animal\_husbandry\textsc{.m.sg=1sg:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `I tended animals.'
\z 
 
\ea \label{JE.64}
\textit{herkam girewεba bînîyenim koɫîmre.} \\ 
\gll herkam girewε=b-a bînîye=n=im koɫî=m=re \\ 
 whoever cry\textsc{.pst.ptcp.pl}=be\textsc{.prs-3pl:S} tie\textsc{.pst.ptcp.m=cop.3sg.m:O=1sg:A} shoulder\textsc{.f=1sg:PSR=post} \\ 
\glt `Each [of my children] who might have cried, I would put on my shoulders.'
\z 
 
\ea \label{JE.65}
\textit{bêşkêm, bêşkêne bîyênê,} \\ 
\gll bêşk(e)-ê=m bêşk(e)-ê=ne bîyê=nê \\ 
 cot\textsc{.f-indf=1sg:PSR} cot\textsc{.f-indf=post} be\textsc{.ptcp.pl=cop.3pl:S} \\ 
\glt `They were in a cot.'
\z 
 
\ea \label{JE.66}
\textit{hurim gêrtênê.} \\ 
\gll hur=im gêrtê=nê \\ 
 \textsc{pvb=1sg:A} take\textsc{.pst.ptcp.pl=cop.3pl:O} \\ 
\glt `I lifted them.'
\z 
 
\ea \label{JE.67}
\textit{awû nanim dεnê nîyεnêm nam bêşkekê.} \\ 
\gll aw(î)=û nan=im dε=nê nîyε=nê=m nam bêşke-(e)kê \\ 
 water\textsc{.f}=and bread\textsc{.m=1sg:A} give\textsc{.pst.ptcp.pl=cop.3pl:R} put\textsc{.pst.ptcp.pl=cop.3pl:O=1sg:A} inside cot\textsc{.f-def.f.sg} \\ 
\glt `I gave them food [and] put them back in the cot.'
\z 
 
\ea \label{JE.68}
\textit{girđ yo zilê bîyênê pey wêşa.} \\ 
\gll girđ yo zil-ê bîyê=nê pey wê=şa \\ 
 all one\textsc{.m} big\textsc{-pl} be\textsc{.ptcp.pl=cop.3pl:S} for \textsc{reflx=3pl:PSR} \\ 
\glt `They all grew up.'
\z 
 
\ea \label{JE.69}
\textit{yo şuwane bîyenû.} \\ 
\gll yo şuwane bîye=n=û\\ 
 one\textsc{.m} shepherd\textsc{.m.sg.dir} become\textsc{.pst.ptcp.m=cop.3sg.m:S}=and \\ 
\glt `One of them has become a shepherd.'
\z 
 
\ea \label{JE.70}
\textit{yo ħeywanben bîyen.} \\ 
\gll yo ħeywanben bîye=n \\ 
 one\textsc{.m.dir} herdsman become\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `One has become a herdsman.'
\z 
 
\ea \label{JE.71}
\textit{pane weşɫe weşɫeyî yanekoɫêşa kerden.} \\ 
\gll p=ane weşɫe weşɫeyî yanekoɫê=şa kerde=n \\ 
 at=\textsc{dem.dist.m.3sg.dir} well happy\_manner moving\_house\textsc{=3pl:A} do\textsc{.pst.ptcp.m=cop.3sg.m:O} \\ 
\glt `They left the house in a happy manner.'
\z 
 
\ea \label{JE.72}
\textit{pêse îsey nebîyen girđçê ʕal bo weʕze hîn bo.} \\ 
\gll pêse îse-î ne-bîye=n girđçê ʕal b-o weʕze hîn b-o \\
  like now\textsc{-m.sg.obl} \textsc{neg-}be\textsc{.pst.ptcp.m=cop.3sg.m:S} everything good be\textsc{.prs.sbjv-3sg:S} situation\textsc{.f} \textsc{fill} be\textsc{.prs.sbjv-3sg:S} \\ 
\glt `It wasn't like nowadays, when everything is fine, and the situation is thingummy.'
\z 
 
\ea \label{JE.73}
\textit{a wextî to mîsal watεt, a fiɫane kesem gerekene.''} \\ 
\gll a wext-î to mîsal wat-ε=t a fiɫan-e kese=m gerek-e=ne \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl} \textsc{2sg} for\_example say\textsc{.pst-cond.aug=2sg:A} \textsc{dem.dist} such\_and\_such\textsc{-ez.cmpd} person\textsc{.f=1sg:NC} necessary\textsc{-f=cop.3sg.f:S} \\ 
\glt `In earlier times, let us say\footnote{Lit. for instance} you said, ``I want that certain person (i.e., girl).''' 
\z 
 
\ea \label{JE.74}
\textit{luwênî law î kuřî waçênî, ``da luwe waçe eđêş bizane î kinaçêşe miđom pene ya ne!''} \\ 
\gll lu-ên-î la-û î kuř-î waç-ên-î da l\'u-e w\'aç-e eđê=ş bi-z\'an-e î kinaçê=ş=e mi-đ(e)-o=m pene ya ne \\
 go\textsc{.prs-aug-2sg:S} to\textsc{-ez.gen} \textsc{dem.prox} boy\textsc{.m-sg.obl} say\textsc{.prs-aug-2sg:A} \textsc{hort} go\textsc{.prs.imp-2sg:S} say\textsc{.prs.imp-2sg:A} mother\textsc{.obl.f=3sg:PSR} \textsc{imp-}know\textsc{.prs-2sg:A} \textsc{dem.prox} daughter\textsc{.f=3sg:PSR=dem} \textsc{ind-}give\textsc{.prs-3sg:A=1sg:R} to or no \\ 
\glt `You would go to this boy [seated next to the narrator] and would say, ``Go [and] tell her mother [about me], see\footnote{Lit. know} if she gives me her daughter or not!'''
\z 
 
\ea \label{JE.75}
\textit{kuřekeyç ê maçî, ``kinaçekêt miđey pî kuřî?''} \\ 
\gll kuř-ekey=ç ê m-aç-î kinaç(ê)-ekê=t mi-đe-î p=î kuř-î \\ 
 boy\textsc{.m-def.m.sg.obl=add} come\textsc{.prs.pst}c\textsc{.3sg} \textsc{ind-}say\textsc{.prs-2sg:A} daughter\textsc{.f-def.f.sg=2sg:PSR} \textsc{ind-}give\textsc{.prs-2sg:A} to=\textsc{dem.prox} boy\textsc{.m-sg.obl} \\ 
\glt `The boy comes. [Addressing her grandson] You would say [to her mother], ``Will you give your daughter to this boy [in marriage]?'''
\z 

\ea \label{JE.76}
\textit{waçê, ``weɫa ba kinaçekê donmê.''} \\ 
\gll waç-ê weɫa ba kinaç(ê)-ekê d\'on-mê \\
 say\textsc{.prs-aug.3sg:A} indeed \textsc{hort} girl\textsc{.f-def.f.sg} talk\_to\textsc{.prs.sbjv-1pl:A} \\ 
\glt `She (the mother) would say, ``Let us talk to the girl.'''
\z 
 
\ea \label{JE.77}
\textit{eger kinaçekêşa donε watεş, ``erê, kerû''} \\ 
\gll eger kinaç(ê)-ekê=şa donε wat-ε=ş erê ker-û \\ 
 if girl\textsc{.f-def.f.sg=3pl:A} talk\_to\textsc{.pst.cond.aug.3sg:R} say\textsc{.pst-cond.aug=3sg:A} yes do\textsc{.prs.ind-1sg:A} \\ 
\glt `If they had talked to the girl [and] she had said, ``Yes, I will [marry him],'''
\z 
 
\ea \label{JE.78}
\textit{waçê, ``dey luwe duwê yerê pîyε bênê.''} \\ 
\gll waç-ê dey l\'u-e duwê yerê pîyε b-ê-nê \\ 
 say\textsc{.prs-aug.3sg:A} \textsc{disc.ptcl} go\textsc{.prs.imp-2sg:S} two three man\textsc{.pl.dir} \textsc{sbjv-}come\textsc{.prs-3pl:S} \\ 
\glt `she (the mother) would say, ``Go! Let two, three men come.'
\z 
 
\ea \label{JE.79}
\textit{duwê pîyε hîcbîş kera.} \\ 
\gll duwê pîyε hîcbî=ş ker-a \\ 
 two man\textsc{.pl.dir} betrothal\textsc{=3sg:O} do\textsc{.prs.ind-3pl:A} \\ 
\glt `Two men acted as intermediaries.'
\z 
 
\ea \label{JE.80}
\textit{ênê hîcbîş kerê. waçê, ``î kinaçê miđey pî kuřî?''} \\ 
\gll ênê hîcbî=ş ker-ê waç-ê î kinaçê mi-đe-î p=î kuř-î \\ 
 come\textsc{.prs.aug.3pl:S} betrothal\textsc{=3sg:O} do\textsc{-prs-aug.3sg:A} say\textsc{.prs-aug.3sg:A} \textsc{dem.prox} girl\textsc{.f} \textsc{ind-}give\textsc{.prs-2sg:A} to=\textsc{dem.prox} boy\textsc{.m-sg.obl} \\ 
\glt `They would come [and] act as intermediaries to ask for the hand of the girl. They would say, ``Will you give this girl to this boy?'''
\z 
 
\ea \label{JE.81}
\textit{waçê, ``erê miđewş.''} \\ 
\gll waç-ê erê mi-đe-û=ş \\ 
 say\textsc{.prs-aug.3sg:A} yes \textsc{ind-}give\textsc{.prs-1sg:A=3sg:O} \\ 
\glt `They would say, ``Yes, I will give her [to him].'''
\z 
 
\ea \label{JE.82}
\textit{waçê, ``dey luwe yanew wêt nîşere ta duwê mangê ter yerê mangê ter.} \\ 
\gll waç-ê dey l\'u-e yane-û wê=t nîş-\'e=re ta duwê mang(e)-ê ter yerê mang(e)-ê ter \\
 say\textsc{.prs-aug.3sg:A} \textsc{disc.ptcl} go\textsc{.prs.imp-2sg:S} house\textsc{.m-ez.gen} \textsc{reflx=2sg:PSR} sit\textsc{.prs.imp-2sg:S=povb} until two month\textsc{.f-pl.dir} another three month\textsc{.f-pl.dir} another \\ 
\glt `They (the girl's family) would say [to the boy], ``Go back home [and] wait\footnote{Lit. sit} for another two or three months,'
\z 
 
\ea \label{JE.83}
\textit{ta qewmekama wiş kermê. qewmeka wiş kermê bizanmê î qewmîma řezaşane ya řezaşa nîyene.''} \\ 
\gll ta qewm-eka=ma wiş k\'er-mê qewm-eka wiş k\'er-mê bi-zan-mê î qewm-î=ma řeza=şa=ne ya řeza=şa nîye=ne \\
 until relative\textsc{.m-def.pl.obl=1pl:PSR} memory\textsc{.m} do\textsc{.prs.sbjv-1pl:A} relative\textsc{.m-def.pl.obl} memory\textsc{.m} do\textsc{.prs.sbjv-1pl:A} \textsc{sbjv-}know\textsc{.prs-1pl:A} \textsc{dem.prox} relative\textsc{.m-sg.obl=1pl:PSR} satisfaction\textsc{.f=3pl:NC=cop.3sg.f:S} or satisfaction\textsc{.f=3pl:NC} \textsc{neg.exist=cop.3sg.f:S} \\ 
\glt `[until] we inform our relatives. We will inform our relatives to see if they agree [to the  marriage] or not.'''
\z 
 
\ea \label{JE.84}
\textit{êtir ênêwe eger řezaşa bîyε, kinaçê dênê.} \\ 
\gll êtir ênê=we eger řeza=şa bî-ε kinaçê d(e)-ên-ê \\ 
 \textsc{disc.ptcl} come\textsc{.prs.pst}c\textsc{.3pl=compl} if satisfaction\textsc{.f=3pl:NC} be.\textsc{pst-cond.aug.3sg:S} girl\textsc{.f} give\textsc{.prs-aug-3pl:A} \\ 
\glt `They (the relatives) would come [to the girl's family] and if they agreed to it, they would give the girl in marriage.'
\z 
 
\ea \label{JE.85}
\textit{eger řezaşa nebîyε neđênê.} \\ 
\gll eger řeza=şa ne-bî-ε ne-đ(e)-ên-ê \\ 
 if satisfaction\textsc{.f=3pl:NC} \textsc{neg-}be\textsc{.pst-cond.aug.3sg:S} \textsc{neg-}give\textsc{.prs-aug-3pl:A} \\ 
\glt `If they didn't agree, they wouldn't give her.'
\z 
