\ea \label{DG.1}
\textit{şê ʕeladînû şa ʕosman.} \\ 
\gll şê ʕeladîn=û şê ʕosman \\ 
 sheikh\textsc{.m} \textsc{pn}=and sheikh\textsc{.m} \textsc{pn.m} \\ 
\glt `Sheikh Aladin and Sheikh Osman.'
\z 
 
\ea \label{DG.2}
\textit{şê ʕeladîn tatew şê ʕosmanî bîyen.} \\ 
\gll şê ʕeladîn tate-û şê ʕosman-î bîye=n \\ 
 sheikh\textsc{.m} \textsc{pn} father\textsc{.m-ez.gen} sheikh\textsc{.m} \textsc{pn.m-sg.obl} be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `Sheikh Aladin was Sheikh Osman's father.'
\z 
 
\ea \label{DG.3}
\textit{a wextîyekey yerê kuřêş bîyênê.} \\ 
\gll a wext-î-ekey yerê kuř-ê=ş bîyê=nê \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} three son\textsc{.m-pl.dir=3sg:NC} be\textsc{.ptcp.pl=cop.3pl:S} \\ 
\glt `[Sheikh Aladin] had three sons.'
\z 
 
\ea \label{DG.4}
\textit{şûnû îney yerê kuřêş bîyênê. êtir î memlekete ta milo kursan, ta milo beẍa, î şa ʕeladîne a wextîyekey xwa payebiɫindiş keronew xwa a wextîyekey ʕefweş keronew ême çêrû himayeyş bîyênmê} \\ 
\gll şûn-û îney yerê kuř-ê=ş bîyê=nê êtir î memleket=e ta mi-l-o kursan ta mi-l-o beẍa, î şa ʕeladîn=e a wext-î-ekey xwa payebiɫind=iş k\'er-one=û xwa a wext-î-ekey ʕefwe=ş k\'er-one=û ême çêr-û himaye-î=ş bîyê=nmê \\
 after\textsc{-ez.gen} \textsc{\textsc{dem.prox}.obl.m.3sg} three son\textsc{.m-pl.dir=3sg:NC} be\textsc{.ptcp.pl=cop.3pl:S} \textsc{disc.ptcl} \textsc{dem.prox} country\textsc{.m=dem} until \textsc{ind-}go\textsc{.prs-3sg:S} \textsc{pn.m} until \textsc{ind-}go\textsc{.prs-3sg:S} \textsc{pn.m} \textsc{dem.prox} king\textsc{.m} \textsc{pn=dem} \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} God\textsc{.m} grand\textsc{=3sg:O} do\textsc{.prs.sbjv-3sg:A}=and God\textsc{.m} \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} pardon\textsc{=3sg:O} do\textsc{.prs.sbjv-3sg:A}=and \textsc{1pl} under\textsc{-ez.gen} support\textsc{.m-sg.obl=3sg:PSR} be\textsc{.ptcp.pl=cop.1pl:S} \\ 
\glt `Well, he had three sons. May God grant him glory; may God forgive him; We were under their rule, that is, Sheikh Osman, [well rather] Sheikh Aladin, in this region and far beyond until Sanandaj and, eh, Baghdad.'
\z 
 
\ea \label{DG.5}
\textit{eger neweşîyêma bîyebo, duktur nebîyenû.} \\ 
\gll eger neweşî-ê=ma bîye=b-o duktur ne-bîye=n=û \\ 
 if illness\textsc{.m-indf=1pl:NC} be\textsc{.pst.ptcp.m}=be\textsc{.prs-3sg:S} physician\textsc{.m} \textsc{neg-}be\textsc{.pst.ptcp.m=cop.3sg.m:S}{=and} \\ 
\glt `If we happened to be ill, there was no doctor.'
\z 
 
\ea \label{DG.6}
\textit{a wextîyekey luwεnmê e wage duʕaw xeyrîş kerdênew.} \\ 
\gll a wext-î-ekey luwε=nmê e=wage duʕa-û xeyr-î=ş kerdê=ne=û \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} go\textsc{.pst.ptcp.pl=cop.1pl:S} in=there prayer\textsc{.f.sg.dir-ez.gen} goodness\textsc{.m-sg.obl=3sg:A} do\textsc{.pst.ptcp.f=cop.3sg.f:O}=and \\ 
\glt `We would go there [i.e. to them and] they would bless us.'
\z 
 
\ea \label{DG.7}
\textit{xway derdema deʕfeş kerdêne.} \\ 
\gll xwa-î derde=ma deʕfe=ş kerdê=ne \\ 
 God\textsc{.m-sg.obl} illness\textsc{.f=1pl:PSR} exclusion\textsc{.f=3sg:A} do\textsc{.pst.ptcp.f=cop.3sg.f:O} \\ 
\glt `[And] God would protect us from illness.'
\z 
 
\ea \label{DG.8}
\textit{zemanê ce zemaniş bîyen ađê fers kere ce şêxî ẍeyasedînîwe ta aro ađê şêxû î memleketî bîyênê.} \\ 
\gll zeman-ê ce zeman=iş bîye=n ađê fers k\'er-e ce şêx-î ɫyaʕdîn-î=we ta aro ađê şêx-û î memleket-î bîyê=nê \\
 time\textsc{.m-indf} from time\textsc{.m=3sg:PSR} be\textsc{.pst.ptcp.m=cop.3sg.m:S} \textsc{3pl.dir} guess\textsc{.m} do\textsc{.prs.imp-2sg:A} from sheikh\textsc{.m-ez.attr} \textsc{pn.m-sg.obl=post} until today \textsc{3pl.dir} sheikh\textsc{.m-ez.gen} \textsc{dem.prox} country\textsc{.m-sg.obl} be\textsc{.ptcp.pl=cop.3pl:S} \\ 
\glt `For a long time, let's say from the period of Sheikh Ghayasadin until today, they have been the Sheikhs in our region.'
\z 
 
\ea \label{DG.9}
\textit{egerketekey serawê fire gewre fire gewre fire gewre bîyen.} \\ 
\gll egerkete-(e)key seraw-ê fire gewre fire gewre fire gewre bîye=n \\ 
 if\textsc{-def.m.sg.obl} large\_area\textsc{-indf} very big very big very big be\textsc{.pst.ptcp.m=cop.3sg.m:S} \\ 
\glt `Imagine there has been a large area of land [under their control],'
\z 
 
\ea \label{DG.10}
\textit{ʕerzit kerû ta milo kursanû ta milo ta wa, tirkmen sara ađê î memliketşa gêɫnan be dîndarî.} \\ 
\gll ʕerz=it k\'er-û ta mi-l-o kursan=û ta mi-l-o ta wa tirkmen sara ađê î memliket=şa gêɫna=n be dîndarî \\
 petition\textsc{.m=2sg:R} do\textsc{.prs.sbjv-1sg:A} until \textsc{ind-}go\textsc{.prs-3sg:S} \textsc{pn.m}{=and} until \textsc{ind-}go\textsc{.prs-3sg:S} until there \textsc{pn} \textsc{pn} \textsc{3pl.dir} \textsc{dem.prox} country\textsc{.m=3pl:A} lead\textsc{.pst.ptcp.m=cop.3sg.m:O} by piety\textsc{.m} \\ 
\glt `[then] let me tell you\footnote{Lit. May a petition be for your service.} that the Sheikhs have ruled this region with piousness as far as Sanandaj and Turkmen Sahra.'
\z 
 
\ea \label{DG.11}
\textit{řowê ce řowa yo mê xizmetiş, xizmetû şê ʕeladînî.} \\ 
\gll řo-ê ce řo-a yo m-ê xizmet=iş xizmet-û şê ʕeladîn-î \\ 
 day\textsc{.m-indf} from day\textsc{.m-pl.obl} one\textsc{.m} \textsc{ind-}come\textsc{.prs.3sg:S} service\textsc{.m=3sg:PSR} service\textsc{.m-ez.gen} sheikh\textsc{.m} \textsc{pn-m.sg.obl} \\ 
\glt `Once\footnote{Lit. one day of days}, a man\footnote{Lit. one}  came into his presence, to Sheikh Aladin.'
\z 
 
\ea \label{DG.12}
\textit{biraş tûşû derdê mê.} \\ 
\gll bira=ş tûş-û derd(e)-ê m-ê \\ 
 brother\textsc{.m=3sg:PSR} encounter\textsc{-ez.gen} illness\textsc{.f-indf} \textsc{ind-}come\textsc{.prs.3sg:S} \\ 
\glt `His brother was suffering from an illness.'
\z 
 
\ea \label{DG.13}
\textit{derdekêş, a wextîyekey, derdû gulî bo, dûrû goşdara.} \\ 
\gll derde-(e)kê=ş a wext-î-ekey derd(e)-û gulî b-o dûr-û goşdar-a \\ 
 illness\textsc{.f-def.f.sg=3sg:PSR} \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} illness\textsc{.f-ez.gen} leprosy be\textsc{.prs.ind-3sg:S} far\textsc{-ez.gen} listener\textsc{.m-pl.obl} \\ 
\glt `May it be far from the audience, his illness was leprosy.'
\z 
 
\ea \label{DG.14}
\textit{derdû gulî bo.} \\ 
\gll derd(e)-û gulî b-o \\ 
 illness\textsc{.f-ez.gen} leprosy be\textsc{.prs.ind-3sg:S} \\ 
\glt `His illness was leprosy.'
\z 
 
\ea \label{DG.15}
\textit{a wextîyekey, birakeyş ber kera.} \\ 
\gll a wext-î-ekey bira-(e)key=ş ber ker-a \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} brother\textsc{.m-def.m.sg.obl=3sg:PSR} out do\textsc{.prs.ind-3pl:A} \\ 
\glt `His brother was expelled [from the village].'
\z 
 
\ea \label{DG.16}
\textit{kabrayç werû řehmû wêşû milo paɫ midowe birakeyşo.} \\ 
\gll kabra=îç wer-û řehm-û wê=ş=û mi-l-o paɫ mi-d(e)-o=we bira-(e)key=ş=o \\ 
 man\textsc{=add} out\_of\textsc{-ez.gen} mercy\textsc{.m-ez.gen} \textsc{reflx=3sg:PSR}=and \textsc{ind-}go\textsc{.prs-3sg:S} side/flank\textsc{.m} \textsc{ind-}give\textsc{.prs-3sg:A=compl} brother\textsc{.m-def.m.sg.obl=3sg:PSR=post} \\ 
\glt `The fellow, out of mercy, went and supported his brother.'
\z 
 
\ea \label{DG.17}
\textit{beroş mila milo eşkewtêwe.} \\ 
\gll ber-o=ş mi-l-a mi-l-o eşkewt(e)-êwe \\ 
 take\textsc{.prs.ind-3sg:A=3sg:O} \textsc{ind-}go\textsc{.prs-3pl:S} \textsc{ind-}go\textsc{.prs-3sg:S} cave\textsc{.f-indf} \\ 
\glt `He took him to… He went to a cave.'
\z 
 
\ea \label{DG.18}
\textit{dere řo milo epare nan wazo.} \\ 
\gll dere řo mi-l-o e=pare nan waz-o \\ 
 valley\textsc{.m} day\textsc{.m} \textsc{ind-}go\textsc{.prs-3sg:S} in=there bread\textsc{.m} request\textsc{.prs.ind-3sg:A} \\ 
\glt `During the day, he would go and collect the food people would leave.'
\z 
 
\ea \label{DG.19}
\textit{şewêne beroşû be ... milowe eşkewtekê la birakeyş.} \\ 
\gll şew(e)-ê=ne ber-o=ş=û be mi-l-o=we eşkewte-(e)kê la bira-(e)key=ş \\ 
 night\textsc{-f.sg.obl=post} take\textsc{.prs.ind-3sg:A=3sg:O}=and to \textsc{ind-}go\textsc{.prs-3sg:S=compl} cave\textsc{.f-def.f.sg} to brother\textsc{.m-def.m.sg.obl=3sg:PSR} \\ 
\glt `Late at night, he would take it (the food) and would go to the cave to his brother.'
\z 
 
\ea \label{DG.20}
\textit{haɫît bo?} \\ 
\gll haɫî=t b-o \\ 
 understood\textsc{=2sg:NC} be\textsc{.prs.ind-3sg:S} \\ 
\glt `You see?'
\z 
 
\ea \label{DG.21}
\textit{a wextîyekey ta yerê saɫê eça keşo ađ guzeran kero la birakeyşo.} \\ 
\gll a wext-î-ekey ta yerê saɫ(e)-ê e=ça keş=o ađ guzeran ker-o la bira-(e)key=ş=o \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} until three year\textsc{.f-pl.dir} in=there mountain\textsc{.m=post} \textsc{3sg.m.dir} subsistence\textsc{.m} do\textsc{.prs.ind-3sg:A} with brother\textsc{.m-def.m.sg.obl=3sg:PSR=post} \\ 
\glt `For three years, he stayed with his brothers in those mountains.'
\z 
 
\ea \label{DG.22}
\textit{birakeyş, derde, laşeş derde kero. awî şano. derdû gulî bo.} \\ 
\gll bira-(e)key=ş derde laşe=ş derde ker-o awî şan-o derd(e)-û gulî b-o \\ 
 brother\textsc{.m-def.m.sg.obl=3sg:PSR} illness\textsc{.f} body\textsc{.m=3sg:PSR} illness\textsc{.f} do\textsc{.prs.ind-3sg:A} water\textsc{.f.sg.dir} ooze\textsc{.prs.ind-3sg:A} illness\textsc{.f-ez.gen} leprosy be\textsc{.prs.ind-3sg:S} \\ 
\glt `His brother had pain in his body. It would ooze liquid\footnote{Lit. water}. [The illness] was leprosy.'
\z 
 
\ea \label{DG.23}
\textit{a wextîyekey, nimaza bilo aweyanî.} \\ 
\gll a wext-î-ekey nim(e)-az-a bi-l-o aweyanî \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} \textsc{neg.ind-}let\textsc{.prs-3pl:A} \textsc{sbjv-}go\textsc{.prs-3sg:S} habitat\textsc{\textsc{.m}} \\ 
\glt `People wouldn't let him go to the village.'
\z 
 
\ea \label{DG.24}
\textit{řowê ce řowa maça, ``pîyayê hen namêş şê ʕeladînarêw.} \\ 
\gll řo-ê ce řo-a m-aç-a pîya-ê hen-∅ namê=ş şê ʕeladîn=a=rê=û \\ 
 day\textsc{.m-indf} from day\textsc{.m-pl.obl} \textsc{ind-}say\textsc{.prs-3pl:A} man\textsc{.m-indf} \textsc{exist-3sg.m:S} name\textsc{.f=3sg:PSR} sheikh\textsc{.m} \textsc{pn=cop.3sg.m:S=povb}=and \\ 
\glt `Once, people said, ``There is a man called Sheikh Aladin.'
\z 
 
\ea \label{DG.25}
\textit{îna yaneş bîyarêw.} \\ 
\gll îna-∅ yane=ş bîyarê=û \\ 
 \textsc{deic-3sg.m:S} house\textsc{=3sg:PSR} \textsc{pn}=and \\ 
\glt `He lives in Biyare.'\footnote{Lit. His house is in Biyare.}
\z 
 
\ea \label{DG.26}
\textit{to mişo î birayte şewêne bere.} \\ 
\gll to mişo î bira-î=t=e şew(e)-ê=ne b\'er-e \\
 \textsc{2sg} \textsc{aux} \textsc{dem.prox} brother\textsc{.m-sg.obl=2sg:PSR=dem} night\textsc{-f.sg.obl=post} take\textsc{.prs.imp-2sg:A} \\ 
\glt `You should take this brother of yours [there] at night.'
\z 
 
\ea \label{DG.27}
\textit{luwe wêt wize milû paɫayeka şê ʕeladînîre.} \\ 
\gll l\'u-e wê=t w\'iz-e mil-û paɫa-eka şê ʕeladîn-î=re \\
 go\textsc{.prs.imp-2sg:S} \textsc{reflx=2sg:PSR/O} throw\textsc{.prs.imp-2sg:A} on\textsc{-ez.gen} shoe\textsc{-def.pl.obl} sheikh\textsc{.m} \textsc{pn-m.sg.obl=post} \\ 
\glt `Go, throw yourself at the feet\footnote{Lit. shoes} of Sheikh Aladin.'
\z 
 
\ea \label{DG.28}
\textit{ga hen laɫêwe xway xwa derdeş deʕfe kero.''} \\ 
\gll gahen laɫy(e)-ê=we xwa-î xwa derde=ş deʕfe k\'er-o \\
 maybe beseech\textsc{-aug.3sg:S=compl} God\textsc{.m-sg.obl} God\textsc{.m} illness\textsc{.f=3sg:PSR} exclusion\textsc{.f} do\textsc{.prs.sbjv-3sg:A} \\ 
\glt `and beg God [to cure him]; may God cure him.'''
\z 
 
\ea \label{DG.29}
\textit{şewê, de, de řowê bo, ya paŋze řowê bo ya wîs řowê bo, ce qeredaxo maroş ta heɫebce.} \\ 
\gll şew(e)-ê de de řo-ê b-o ya pangze řo-ê b-o ya wîs řo-ê b-o ce qeredax=o m-ar-o=ş ta heɫebce \\
 night\textsc{.f-indf} ten ten day\textsc{.m-pl.dir} be\textsc{.prs.sbjv-3sg:S} or fifteeen day\textsc{.m-pl.dir} be\textsc{.prs.sbjv-3sg:S} or twenty day\textsc{.m-pl.dir} be\textsc{.prs.sbjv-3sg:S} from \textsc{pn=post} \textsc{ind-}bring\textsc{.prs-3sg:A=3sg:O} until \textsc{pn} \\ 
\glt `Whether it took one night, ten days, fifteen days, or twenty days, he took his brother from Qaradax to Halabja.'
\z 
 
\ea \label{DG.30}
\textit{mênê degaw çinarîş maça.} \\ 
\gll m-ê-nê dega-û çinar-î=ş m-aç-a \\ 
 \textsc{ind-}come\textsc{.prs-3pl:S} village\textsc{.f-ez.gen} \textsc{pn-m.sg.obl=3sg:R} \textsc{ind-}say\textsc{.prs-3pl:A} \\ 
\glt `They came to a village called Chinar.'
\z 
 
\ea \label{DG.31}
\textit{mênê eça birakeyş bero teɫanêwene beynû çinarane.} \\ 
\gll m-ê-nê e=ça bira-(e)key=ş ber-o teɫan-êwe=ne beyn-û çinar-a=ne \\ 
 \textsc{ind-}come\textsc{.prs-3pl:S} in=there brother\textsc{.m-def.m.sg.obl=3sg:PSR} take\textsc{.prs.ind-3sg:A} terrace\_farm\textsc{-indf=post} between\textsc{-ez.gen} \textsc{pn-pl.obl=post} \\ 
\glt `They went to a terrace farm in between the two sides of Chinar village.'
\z 
 
\ea \label{DG.32}
\textit{minyoşrew wêş mê xaneqaw bîyarê.} \\ 
\gll mi-ny(e)-o=ş=re=û wê=ş m-ê xaneqa-û bîyarê \\ 
 \textsc{ind-}put\textsc{.prs-3sg:A=3sg:O=povb}=and \textsc{reflx=3sg:PSR} \textsc{ind-}come\textsc{.prs.3sg:S} monastery\textsc{.m-ez.gen} \textsc{pn} \\ 
\glt `He left his brother there, and went to the monastery.'
\z 
 
\ea \label{DG.33}
\textit{xaneqaw bîyarêne, a wextîyekey dađ kero.} \\ 
\gll xaneqa-û bîyarê=ne a wext-î-ekey dađ ker-o \\ 
 monastery\textsc{.m-ez.gen} \textsc{pn=post} \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} cry\textsc{.m} do\textsc{.prs.ind-3sg:A} \\ 
\glt `He gave a cry of despair at the monastery of Biyare.'
\z 
 
\ea \label{DG.34}
\textit{maço, ``seweqet bo, min birayêwim henû îne waʕzemnew îne karmaw îne waʕzemne.} \\ 
\gll m-aç-o seweqe=t b-o min bira-êw=im hen-∅=û îne waʕze=m=ne=û îne kar=m=a=û îne waʕze=m=ne \\ 
 \textsc{ind-}say\textsc{.prs-3sg:A} sacrifice\textsc{=2sg:PSR} be\textsc{.prs.ind-3sg:S} \textsc{1sg} brother\textsc{.m-indf=1sg:NC} \textsc{exist-3sg.m:S}=and \textsc{dem.prox.m.3sg.dir} situation\textsc{.f=1sg:NC=\textsc{cop\textsc{.3sg.f}}}=and \textsc{dem.prox.m.3sg.dir} task\textsc{.m=1sg:PSR=cop.3sg.m:S}{=and} \textsc{dem.prox.m.3sg.dir} situation\textsc{.f=1sg:PSR=\textsc{cop\textsc{.3sg.f}}} \\ 
\glt `He said [to Sheikh Aladin], ``May I be your sacrifice! I have a brother; my situation is like this; my task is like this; my situation is like this.'
\z 
 
\ea \label{DG.35}
\textit{nimaza bilmê aweyanî.} \\ 
\gll nim(e)-az-a bi-l-mê aweyanî \\ 
 \textsc{neg.ind-}let\textsc{.prs-3pl:A} \textsc{sbjv-}go\textsc{.prs-1pl:S} habitat\textsc{.m} \\ 
\glt `We are not allowed to go to the village.'
\z 
 
\ea \label{DG.36}
\textit{be ... çaweɫ ħeywandarê bîyênmêw.} \\ 
\gll be çaweɫ ħeywandar-ê bîyê=nmê=û \\ 
 by in\_the\_past herdsman\textsc{-pl.dir} be\textsc{.ptcp.pl=cop.1pl:S}=and \\ 
\glt `In the past, we were herdsmen.'
\z 
 
\ea \label{DG.37}
\textit{baxdarê bîyênmêw.} \\ 
\gll baxdar-ê bîyê=nmê=û \\ 
 gardener\textsc{.m-pl.dir} be\textsc{.ptcp.pl=cop.1pl:S}=and \\ 
\glt `We were gardeners.'
\z 
 
\ea \label{DG.38}
\textit{a wextîyekey, êtir, îse, min yerê saɫê î birame tûşû î derdê amanû.} \\ 
\gll a wext-î-ekey êtir îse min yerê saɫ(e)-ê î bira=m=e tûş-û î derd(e)-ê ama=n=û \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} \textsc{disc.ptcl} now \textsc{1sg} three year\textsc{.f-pl.dir} \textsc{dem.prox} brother\textsc{.m=1sg:PSR=dem} encounter\textsc{-ez.gen} \textsc{dem.prox} illness\textsc{-f.sg.obl} come\textsc{.pst.ptcp.m=cop.3sg.m:S}=and \\ 
\glt `Now, as for me, it's been three years that my brother is suffering from this illness.'
\z 
 

\newpage
\ea \label{DG.39}
\textit{nimaza bilû aweyanî.} \\ 
\gll nim(e)-az-a bi-l-û aweyanî \\ 
 \textsc{neg.ind-}let\textsc{.prs-3pl:A} \textsc{sbjv-}go\textsc{.prs-1sg:S} habitat\textsc{.f} \\ 
\glt `I am not allowed to go to the village.'
\z 
 
\ea \label{DG.40}
\textit{şewêne, řone milû nan peya kerûnew.} \\ 
\gll şew(e)-ê=ne řo=ne mi-l-û nan peya ker-ûne=û \\ 
 night\textsc{-f.sg.obl=post} day\textsc{.m=post} \textsc{ind-}go\textsc{.prs-1sg:S} bread\textsc{.m} visible do\textsc{.prs.ind-1sg:A}=and \\ 
\glt `I go [and] find bread during the day.'
\z 
 
\ea \label{DG.41}
\textit{şewêne milûwe laş.} \\ 
\gll şew(e)-ê=ne mi-l-û=we la=ş \\ 
 night\textsc{-f.sg.obl=post} \textsc{ind-}go\textsc{.prs-1sg:S=compl} to\textsc{=3sg:R} \\ 
\glt `I go back to him at night.'
\z 
 
\ea \label{DG.42}
\textit{nizîkû aweyanîyê nimaza ginmê.} \\ 
\gll nizîk-û aweyanî-ê nim(e)-az-a g\'in-mê \\
 close\textsc{-ez.gen} habitat\textsc{\textsc{.m}-indf} \textsc{neg.ind-}let\textsc{.prs-3pl:A} fall\textsc{.prs.sbjv-1pl:S} \\ 
\glt `They don't let us anywhere near a village.'
\z 
 
\ea \label{DG.43}
\textit{î birame derdeş, derdêş îna po derdû gulîş maça dađê!} \\ 
\gll î bira=m=e derde=ş derd(e)-ê=ş îna-∅ p=o derd(e)-û gulî=ş m-aç-a dađ-ê \\ 
 \textsc{dem.prox} brother\textsc{.m=1sg:PSR=dem} illness\textsc{.f=3sg:R} illness\textsc{.f-indf=3sg:R} \textsc{deic-3sg.m:S} at\textsc{=post} illness\textsc{.f-ez.gen} leprosy\textsc{=3sg:R} \textsc{ind-}say\textsc{.prs-3pl:A} succour\textsc{.m-pl.dir} \\ 
\glt `This brother of mine is ill; it is called leprosy; Please help!'
\z 
 
\ea \label{DG.44}
\textit{îđîç fermawo, luwe ça nîşere.} \\ 
\gll îđ=îç fermaw-o l\'u-e ça n\stackunder[-10pt]{\^{i}}{\'{}}ş-e=re \\
 \textsc{3sg.prox.dir.m=add} say\textsc{.prs.ind-3sg:A} go\textsc{.prs.imp-2sg:S} there sit\textsc{.prs.imp-2sg:S=povb} \\ 
\glt `He (Sheikh Aladin) said, ``Go [and] sit there.'
\z 
 

\newpage
\ea \label{DG.45}
\textit{luwe xaneqa ta xeberit dewwe pene.''} \\ 
\gll l\'u-e xaneqa ta xeber=it d\'e-û=we pene \\
 go\textsc{.prs.imp-2sg:S} monastery\textsc{.m} until news\textsc{.m=2sg:R} give\textsc{.prs.sbjv-1sg:A=compl} to \\ 
\glt `Go to the monastery until I let you know.'
\z 
 
\ea \label{DG.46}
\textit{şêx mutewec bo, wextê miđyo derdû î pîyay la xwayû xwawinû ʕaɫemîyanî mutewec bo, řanimonîş kero.} \\ 
\gll şêx mutewec b-o wext-ê mi-đy(e)-o derd(e)-û î pîya-î la xwa-î=û xwawin-û ʕaɫemîyan-î mutewec b-o řanimonî=ş ker-o \\ 
 sheikh\textsc{.m} aware be\textsc{.prs.ind-3sg:S} time\textsc{.m-indf} \textsc{ind-}look\textsc{.prs-3sg:S} illness\textsc{.f-ez.gen} \textsc{dem.prox} man\textsc{.m-sg.obl} at\_the\_place\_of God\textsc{.m-sg.obl}=and owner\textsc{-ez.gen} worlds\textsc{.m-sg.obl} aware be\textsc{.prs.ind-3sg:S} guidance\textsc{.m=3sg:O} do\textsc{.prs.ind-3sg:A} \\ 
\glt `The Sheikh understood. When he looked at the man's illness, the lord of all universes realised [what] his illness [was and] guided him (i.e. the Sheikh).'
\z 
 
\ea \label{DG.47}
\textit{winû kîseɫê sawî be...} \\ 
\gll win(î)-û kîseɫ(î)-ê s\'aw-î be \\
 blood\textsc{.f-ez.gen} tortoise\textsc{-f.sg.obl} rub\textsc{.prs.sbjv-2sg:A} to \\ 
\glt `[The Sheikh said to the man], ``You may rub tortoise's blood on...'
\z 
 
\ea \label{DG.48}
\textit{şime be kîseɫê çêş maçdê?} \\ 
\gll şime be kîseɫ(î)-ê çêş m-aç-dê \\ 
 \textsc{2pl} by tortoise\textsc{-f.sg.obl} what \textsc{ind-}say\textsc{.prs-2pl:A} \\ 
\glt `What do you say for \textit{kîseɫ} `tortoise'''?'
\z 
 
\ea \label{DG.49}
\textit{winû kîseɫê zamdare kere.} \\ 
\gll win(î)-û kîseɫ(î)-ê zamdar-e k\'er-e \\
 blood\textsc{.f-ez.gen} tortoise\textsc{-f.sg.obl} wounded\textsc{-f} do\textsc{.prs.imp-2sg:A} \\ 
\glt `Wound a tortoise,'
\z 
 
\newpage
\ea \label{DG.50}
\textit{winekêş sawe pa qelbû pîyayre.} \\ 
\gll win(î)-ekê=ş s\'aw-e p=a qelb-û pîya-î=re \\
 blood\textsc{.f-def.f.sg=3sg:PSR} rub\textsc{.prs.imp-2sg:A} by\textsc{=dem.dist} heart\textsc{.m-ez.gen} man\textsc{.m-sg.obl=post} \\ 
\glt `[then] rub its blood on the heart of the [sick] man'
\z 
 
\ea \label{DG.51}
\textit{î pose ane gino, pos maro nawe.} \\ 
\gll î pos=e ane gin-o pos m-ar-o nawe \\ 
 \textsc{dem.prox} skin\textsc{.m=dem} \textsc{dem.dist.m.3sg.dir} fall\textsc{.prs.ind-3sg:S} skin\textsc{.m} \textsc{ind-}bring\textsc{.prs-3sg:A} inside \\ 
\glt `his skin will fall off, and [a new] skin will replace it.'
\z 
 
\ea \label{DG.52}
\textit{bowe pise wêş.} \\ 
\gll b-o=we pise wê=ş \\ 
 become\textsc{.prs.ind-3sg:S=compl} like \textsc{reflx=3sg:PSR} \\ 
\glt `He will become like he was.'\footnote{Lit. He will become himself.}
\z 
 
\ea \label{DG.53}
\textit{şa ʕeladîn îney îney maço be ... fermawo be pîya, hînî, biraw pîyakey.} \\ 
\gll şa ʕeladîn îney îney m-aç-o be fermaw-o be pîya hîn-î bira-û pîya-(e)key \\ 
 king\textsc{.m} \textsc{pn} \textsc{\textsc{dem.prox}.obl.m.3sg} \textsc{\textsc{dem.prox}.obl.m.3sg} \textsc{ind-}say\textsc{.prs-3sg:A} to say\textsc{.prs.ind-3sg:A} to man\textsc{.m} \textsc{fill-obl.m} brother\textsc{.m-ez.gen} man\textsc{.m-def.m.sg.obl} \\ 
\glt `Sheikh Aladin said this to the man, to the brother of the [sick] man.'
\z 
 
\ea \label{DG.54}
\textit{biraş înarê degaw çinarî, metawo bilo aweyanî.} \\ 
\gll bira=ş îna-∅=rê dega-û çinar-î me-taw-o bi-l-o aweyanî \\ 
 brother\textsc{.m=3sg:PSR} \textsc{deic-3sg.m:S=povb} village\textsc{.f-ez.gen} \textsc{pn-m.sg.obl} \textsc{neg.ind-}can\textsc{.prs-3sg:A} \textsc{sbjv-}go\textsc{.prs-3sg:S} habitat\textsc{\textsc{.m}} \\ 
\glt `His brother was in Chinar village. He wasn't allowed to go to the village.'
\z 
 
\newpage
\ea \label{DG.55}
\textit{a wextîyekey îne wîyero. qerar bo} \\ 
\gll a wext-î-ekey îne wîyer-o qerar b-o \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} \textsc{dem.prox.m.3sg.dir} pass\textsc{.prs-3sg:S} arrangement be\textsc{.prs.ind-3sg:S} \\ 
\glt `Then, this was settled. The deal was ...'
\z 
 
\ea \label{DG.56}
\textit{fermawo, ``hoşit neşowe bilewe îne derdû î biraw tone.} \\ 
\gll fermaw-o hoş=it ne-ş-o=we bi-l-e=we îne derd(e)-û î bira-û to=ne \\ 
 say\textsc{.prs.ind-3sg:A} memory\textsc{.m=2sg:PSR} \textsc{neg.sbjv-}go\textsc{.prs-3sg:S=compl} \textsc{imp-}go\textsc{.prs-2sg:S=compl} \textsc{dem.prox.m.3sg.dir} illness\textsc{.f-ez.gen} \textsc{dem.prox} brother\textsc{.m-ez.gen} \textsc{2sg=cop.3sg.f:S} \\ 
\glt `He (Sheikh Aladin) said, ``Don't forget! Go back [to your brother]. This is the medicine for your brother.'''
\z 
 
\ea \label{DG.57}
\textit{luwe xaneqa. nan bere, bideş penew.} \\ 
\gll l\'u-e xaneqa nan b\'er-e bi-d(e)-e=ş pene=û \\
 go\textsc{.prs.imp-2sg:S} monastery\textsc{.m} bread\textsc{.m} take\textsc{.prs.imp-2sg:A} \textsc{imp-}give\textsc{.prs-2sg:A=3sg:R} to=and \\ 
\glt `Go to the monastery. Take food [and] give it to him.'
\z 
 
\ea \label{DG.58}
\textit{a wextîyekey, gêɫdê lodê. pena bere be xway.} \\ 
\gll a wext-î-ekey gêɫ-dê l\'o-dê pena b\'er-e be xwa-î \\
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} wander\textsc{.prs.imp-2pl:S} go\textsc{.prs.imp-2pl:S} refuge take\textsc{.prs.imp-2sg:A} with God\textsc{.m-sg.obl} \\ 
\glt `Then, go back [to your place]. Take refuge with God.'
\z 
 
\ea \label{DG.59}
\textit{gahes pî qisê mine xwa řehm kero, weş bo.''} \\ 
\gll gahes p=î qisê min=e xwa řehm ker-o weş b-o \\ 
 maybe by\textsc{=dem}.prox talk\textsc{.f.sg} \textsc{1sg=dem} God\textsc{.m} mercy\textsc{.m} do\textsc{.prs.ind-3sg:A} well be\textsc{.prs.ind-3sg:S} \\ 
\glt `Through my intervention may God have mercy on him so that he gets well.'''
\z 
 
\ea \label{DG.60}
\textit{şûnû îneyre aha la xwawe inne duʕaş meqboɫe bo.} \\ 
\gll şûn-û îney=re aha la xwa=we inne duʕa=ş meqboɫ-e b-o \\ 
 after\textsc{-ez.gen} \textsc{\textsc{dem.prox}.obl.m.3sg=post} \textsc{intj} at\_the\_place\_of God\textsc{.m=post} so\_much prayer\textsc{.f=3sg:PSR} acceptable\textsc{-f} be\textsc{.prs.ind-3sg:S} \\ 
\glt `Well then, his prayers were accepted by God.'
\z 
 
\ea \label{DG.61}
\textit{kîseɫêwe ça kîseɫa milo xizmetû şê ʕeladînî.} \\ 
\gll kîseɫ(î)-êwe ç=a kîseɫ(î)-a mi-l-o xizmet-û şê ʕeladîn-î \\ 
 tortoise\textsc{.f-indf} from\textsc{=dem.dist} tortoise\textsc{.f-pl.obl} \textsc{ind-}go\textsc{.prs-3sg:S} service\textsc{.m-ez.gen} sheikh\textsc{.m} \textsc{pn-m.sg.obl} \\ 
\glt `A tortoise\footnote{Lit. A tortoise of those tortoises.} came into the presence of Sheikh Aladin.'
\z 
 
\ea \label{DG.62}
\textit{teqe teq pilekanî ba xaneqa hurwezo, duwê pilekanî hurwezo, temenεwe gêro pey şexî.} \\ 
\gll teqeteq pilekanî b-a xaneqa hur-wez-o duwê pilekanî hur-wez-o temenεwe gêr-o pey şex-î \\ 
 rattling stair\textsc{.f} be\textsc{.prs.ind-3pl:S} monastery\textsc{.m} \textsc{pvb-}climb\textsc{.prs.ind-3sg:S} two stair\textsc{.pl.dir} \textsc{pvb-}climb\textsc{.prs.ind-3sg:S} supplication\textsc{.m.indf} take\textsc{.prs.ind-3sg:A} to sheikh\textsc{-m.sg.obl} \\ 
\glt `He went up the stairs at the monastery, making a noise [as he went]. He went up two stairs. He pleaded with the Sheikh.'
\z 
 
\ea \label{DG.63}
\textit{bîyenû meʕlûma.} \\ 
\gll bîye=n=û meʕlûm=a \\ 
 be\textsc{.pst.ptcp.m=cop.3sg.m:S}=and obvious\textsc{=cop.3sg.m:S} \\ 
\glt `This [event] has occurred, and it is obvious.'
\z 
 
\ea \label{DG.64}
\textit{a wextîyekey `beɫê,' `min ... bilo a memleketû wêşa, kiseɫîş çenê. ême zamdarê nekero. ême keçê nekero.} \\ 
\gll a wext-î-ekey beɫê min bi-l-o a memleket-û wê=şa kiseɫî=ş çenê ême zamdar-ê ne-ker-o ême keç-ê ne-ker-o \\ 
 \textsc{dem.dist} time\textsc{.m-sg.obl-def.m.sg.obl} yes \textsc{1sg} \textsc{sbjv-}go\textsc{.prs-3sg:S} \textsc{dem.dist} country\textsc{.m-ez.gen} \textsc{reflx=3pl:PSR} tortoise\textsc{.f=3sg:R} in \textsc{1pl} wounded\textsc{-pl} \textsc{neg.sbjv-}do\textsc{.prs-3sg:A} \textsc{1pl} crippled\textsc{-pl} \textsc{neg.sbjv-}do\textsc{.prs-3sg:A} \\ 
\glt `[The Sheikh said], ``Yes!''[The tortoise said] I [say] he [should] go to his region; [there are] tortoises there. He should not injure us. He should not make us crippled.'
\z 
 
\ea \label{DG.65}
\textit{beroşo memliketû wêşa, kîseɫê zamdar kero.} \\ 
\gll b\'er-o=ş=o memliket-û wê=şa kîseɫ(î)-ê zamdar k\'er-o \\
 take\textsc{.prs.sbjv-3sg:A=3sg:O=compl} country\textsc{.m-ez.gen} \textsc{reflx=3pl:PSR} tortoise\textsc{.f-indf} wounded\textsc{.m} do\textsc{.prs.sbjv-3sg:A} \\ 
\glt `He should take him to his own region [and] injure a tortoise from there.'
\z 
 
\ea \label{DG.66}
\textit{ê.. dewayekeş sawe pore da ême keçê nebîmê.'} \\ 
\gll ê.. dewayî-eke=ş s\'aw-e pore da ême keç-ê ne-b-îmê \\
 eh medication\textsc{-def.m.sg.dir=3sg:PSR} rub\textsc{.prs.imp-2sg:A} at \textsc{hort} \textsc{1pl} crrippled\textsc{-pl} \textsc{neg.sbjv-}be\textsc{.prs-1pl:S} \\ 
\glt `He should rub the medicine on him; accordingly, may we not be crippled.'
\z 
 
\ea \label{DG.67}
\textit{şê ʕeladînîç fermawo, ``birakem wextê înê zana min gunakê may mê milimre.} \\ 
\gll şê ʕeladîn-î=ç fermaw-o bira-(e)ke=m wext-ê înê zan-a min guna-(e)kê m-ay m-ê mil=im=re \\ 
 sheikh\textsc{.m} \textsc{pn-m.sg.obl=add} say\textsc{.prs.ind-3sg:A} brother\textsc{.m-def.m.sg.dir=1sg:PSR} time\textsc{.m-indf} \textsc{dem.prox.3pl} know\textsc{.prs.ind-3pl:A} \textsc{1sg} sin\textsc{.f-def.f.sg} \textsc{ind-}come\textsc{.prs.nmlz} \textsc{ind-}come\textsc{.prs.3sg:S} shoulder\textsc{.m=1sg:PSR=post} \\ 
\glt `Sheikh Aladin said, ``Brother! as they (i.e. the tortoises) know [about the medicine], I won't shoulder the burden of injuring them.\footnote{Lit. The sin will come to me.}'
\z 
 
\newpage
\ea \label{DG.68}
\textit{bereşo memliketû wêta, welê serma deydê.''} \\ 
\gll b\'er-e=ş=o memliket-û wê=ta welê ser=ma d\'e-ydê \\
 take\textsc{.prs.imp-2sg:A=3sg:O=compl} country\textsc{.m-ez.gen} \textsc{reflx=2pl:PSR} but head\textsc{.m=1pl:R} give\textsc{.prs.imp-2pl:A} \\ 
\glt `Take him to your region, but pay us a visit.'''
\z 
 
\ea \label{DG.69}
\textit{milowe. mangêş pene meşo mêwe.} \\ 
\gll mi-l-o=we mang(e)-ê=ş pene me-ş-o m-ê=we \\ 
 \textsc{ind-}go\textsc{.prs-3sg:S=compl} month\textsc{.f-indf=3sg:R} to \textsc{neg.ind-}go\textsc{.prs-3sg:S} \textsc{ind-}come\textsc{.prs.3sg:S=compl} \\ 
\glt `He went back. A month had barely passed [and] he came back.'
\z 
 
\ea \label{DG.70}
\textit{kuře wextê miđyo ana ça şazayêwe.} \\ 
\gll kuř-e wext-ê mi-đy(e)-o an(e)=a ça şazay(e)-êwe \\ 
 boy\textsc{.m-def.dir} time\textsc{.m-indf} \textsc{ind-}look\textsc{.prs-3sg:S} \textsc{dem.dist.m.3sg.dir=deic} there prince\textsc{.m-indf} \\ 
\glt `The [sick] boy has become like a prince.'
\z 
 
\ea \label{DG.71}
\textit{weş bîyeno.} \\ 
\gll weş bîye=n=o \\ 
 well be\textsc{.pst.ptcp.m=cop.3sg.m:S=compl} \\ 
\glt `He had been healed.'
\z 
 
\ea \label{DG.72}
\textit{haɫîta?} \\ 
\gll haɫî=t=a \\ 
 understood\textsc{=2sg:NC=cop.3sg.m:S} \\ 
\glt `You see?'
\z 
 
\ea \label{DG.73}
\textit{înê dastanû î şêxaw ême bîyen gewregerîşa.} \\ 
\gll înê dastan(e)-û î şêx-a-û ême bîye=n gewregerî=şa \\ 
 \textsc{dem.prox.f.3sg.dir} story\textsc{.f-ez.gen} \textsc{dem.prox} sheikh\textsc{.m-pl.obl-ez.gen} \textsc{1pl} be\textsc{.pst.ptcp.m=cop.3sg.m:S} greatness\textsc{=3pl:PSR} \\ 
\glt `This is the story of our Sheikhs and their greatness.'
\z 
