\addchap{\lsAcknowledgementTitle} 
 Numerous people and institutions have made this book possible, and I would like to express my gratitude to all of them.
 
To begin with, my work has received generous funding from the German Research Council (DFG), project \#{}453412235.
 
 
It was Raija Kramer who first got me interested in the topic of phasal polarity.  At the English department at FSU Jena I encountered a welcoming and supportive working environment. Many an insight came from discussions at our research colloquium, and from long conversations with Volker Gast. I am indebted to the anonymous peer reviewers for Language Science Press, as well as to Caroline de Becker, Eva Schultze-Berndt, Itamar Francez, Jens G. Fischer, Martin Haspelmath, Rozenne Guérois, Maj-Britt Hansen, Stefan Savić, Thera Crane, Tom Bossuyt, Volker Gast, and Yael Greenberg for feedback on earlier drafts of different parts of this book. I also wish to thank the community proofreaders and Kathy Wolf for fixing my non-native English, as well as Sebastian Nordhoff and Felix Kopecky for help with typesetting.

Last, but far from least, this book would have been impossible without the support of the countless people who endured my endless, surely bizarre questions, shared unpublished data with me, helped with transliteration and tricky glosses, or provided references that were hard to come by. These are (in alphabetical order of their given names): 

\begin{quote}
Alena Witzlack-Makarevich, Andrey Nefedov, Antoine Guillaume, Bruno Olsson, Christian Döhler, Cindy Boyer, Dan Ke, Drungdrung a Tjakivusung, Elia Weber, Eline Visser, Elizabeth Zeitoun, Eva Schultze-Berndt, Jim Boyer, Felicity Meakins, Gerd Jandraschek, Gertrud Schneider-Blum, Gilles Polian, Guillaume Jacques, Harald Hammerström, Heinrich Werner, Helma Pasch, Irina Nikolaeva, Itamar Francez, Jack Martin, James Essegbey, Jens G. Fischer, John Elliot, John McLaughlin, Jurica Polančec, Katharina Haude, Klaudia Dombrowsky-Hahn, Kristian Roncero, Lalo a tahesh kaybabaw, Maj-Britt Hansen, Marian Klamer, Marijana Kresić Vukosav, Maud Devos, Michel Launey, Miriam van Staden, Onelisa Slater, Peter Barclay, Petrus Mabena, Philomena Dol, Phumelele Lovisa, Ponsiano Kanijo, Rachid Ridouane, Rasmus Bernander, Sérgio Meira, Stefan Dienst, Stefan Georg, Stefan Savić, Tapani Salminen, Thera Crane, Wei-chen Huang, Yael Greenberg, Zhuang Chen.
\end{quote}

Needless to say, all mistakes remain my own.
