\addchap{Acknowledgements}

This book grew out of a postdoctoral research project I undertook at the University of Bamberg in the years 2012–2018. I would like to thank my colleagues in English Linguistics there, who were always available for discussions of different aspects of my work and who gave me feedback and food for thought concerning the host of theoretical and methodo\-logical challenges that I faced. I am particularly grateful to Manfred Krug, Lukas Sönning, Julia Schlüter, Fabian Vetter and Gabriele Knappe. Manfred's role as a mentor and friend has been very special throughout this project and during the years leading up to it~– thank you! Discussions and joint activities with Lukas in the areas of (Bayesian) regression modelling and data visualisation had a profoundly positive influence on my work, and his generous practical advice was invaluable. In Bamberg and beyond, Geoffrey Haig, Edgar Schneider and Martin Hilpert kindly agreed to join Manfred on the board of reviewers for my postdoctoral degree~– thank you for your advice and support! I am also grateful to everybody who took an interest in discussing my work at various conferences across Europe. The University of Bamberg funded many of those trips and provided a research environment that was fantastic in every conceivable way~– it was a joy to work there. At Language Science Press, I am grateful to the editors of the \textit{Language Variation} series for accepting my monograph, the community proofreaders who spotted so much that I had overlooked, and Felix Kopecky, who made the entire production process very smooth and even instructive. Many thanks are also due to Lena Senger at Leipzig University, who did a marvellous job preparing the final version of the manuscript. Finally, and once again, I would like to thank Steffi and Arved, who supported me when things were not running smoothly, celebrated with me when they were, and kept reminding me of one important fact: Although it may be interesting to look at concessive constructions in varieties of English, there is so much more to life than this. I dedicate this book to you.\bigskip\\
\noindent Berlin, October 2023
