\title{(Dia)lects in the 21st century}
\subtitle{Selected papers from Methods in Dialectology XVII}
\BackBody{This book offers an in-depth exploration of contemporary issues and methodologies in the fields of dialectology and sociolinguistics. Readers will find a diverse collection of studies that examine how language varies and changes across different regions, communities, and social contexts. The book covers a wide range of languages, including German, English, Yiddish, Russian, and Japanese, providing a global perspective on linguistic diversity.

Key themes include the use of modern data sources, such as social media, to study language patterns and the impact of digital communication on regional dialects. The book also addresses the dynamics of language contact in expatriate communities, revealing how speakers adapt and merge linguistic features from different dialects.

Several chapters focus on the evolution of dialectological research, offering critiques and new approaches to studying regional language variations. Readers will also encounter innovative methods, such as cognitive geography, which uses mental representations of space to understand dialect variation, and tone distance measures, which are crucial for studying tonal languages.

Additionally, the book presents case studies on how non-experts perceive and categorize dialects, providing insights into the public's understanding of linguistic diversity. It also tackles challenges in selecting dialect speakers for research, especially in urban environments, where traditional criteria may no longer apply.

Overall, this book is a valuable resource for linguists, researchers, and anyone interested in the complex and ever-changing landscape of human language. It highlights the importance of adapting research methods to keep pace with the evolving nature of language and offers fresh perspectives on how we study and understand dialects and language variation.}

\author{Susanne Wagner and Ulrike Stange-Hundsdörfer} 


\renewcommand{\lsISBNdigital}{978-3-96110-502-1}
\renewcommand{\lsISBNhardcover}{978-3-98554-131-7}
\BookDOI{10.5281/zenodo.14925847}
\typesetter{Hannah Schleupner}
\proofreader{Amir Ghorbanpour,
Andreas Hölzl,
Andrew Kato,
Brett Reynolds,
Christopher Straughn,
Elliott Pearl,
Erk Deniz Vargez,
Georgios Vardakis,
Hella Olbertz,
Katja Politt,
Killian Kiuttu,
Lea Schäfer,
Mary Ann Walter,
Jean Nitzke,
Rainer Schulze,
Raquel Benítez Burraco,
Sarah Warchhold}
% \lsCoverTitleSizes{51.5pt}{17pt}% Font setting for the title page


\renewcommand{\lsSeries}{lv}
\renewcommand{\lsSeriesNumber}{10}
\renewcommand{\lsID}{447}
