\chapter{Appendices}
\section{APPENDIX A – Film Corpus Text Elements}\label{appendix_a}

\newcommand{\filmbox}[5]{%
\parbox{6mm}{#1\vphantom{fj}}~%
\parbox[t]{70mm}{\raggedright #2\vphantom{fj}}~%
\parbox{17mm}{#3\vphantom{fj}}~%
\parbox{17mm}{#4\vphantom{fj}}~%
\parbox{17mm}{#5\vphantom{fj}}\\\medskip%
}
 
\filmbox{Nr.}{Original Title}{Release}{Ranking}{Analysed}{} 
\noindent 
\filmbox{1}{The Lord of the Rings: The Fellowship of the Ring}{2001}{92.25}{y}
\filmbox{2}{The Lord of the Ring: The Return of the King}{2005}{89.5}{y}
\filmbox{3}{Avatar}{2009}{86}{y}
\filmbox{4}{Finding Nemo}{2003}{82.75}{y}
\filmbox{5}{The Lord of the Rings: The Two Towers}{2002}{77.75}{y}
\filmbox{6}{Harry Potter and the Sorcerer's Stone}{2001}{71}{y}
\filmbox{7}{Harry Potter and the Goblet of Fire}{2005}{69.75}{y}
\filmbox{8}{The Incredibles}{2004}{65}{y}
\filmbox{9}{Star Wars: Episode III - Revenge of the Sith}{2005}{62}{y}
\filmbox{10}{The Dark Knight}{2008}{59.75}{n}
\filmbox{11}{Pirates of the Caribbean: The Curse of the Black Pearl}{2003}{59.25}{y}
\filmbox{12}{Harry Potter and the Prisoner of Azkaban}{2004}{57.5}{y}
\filmbox{13}{Harry Potter and the Half-Blood Prince}{2009}{55.5}{y}
\filmbox{14}{Ratatouille}{2007}{52.75}{y}
\filmbox{15}{Harry Potter and the Order of the Phoenix}{2007}{52.5}{y}
\filmbox{16}{Up}{2009}{50.25}{y}
\filmbox{17}{Pirates of the Caribbean: Dead Man's Chest}{2006}{49.5}{y}
\filmbox{18}{Star Wars: Episode II - Attack of the Clones}{2002}{48.5}{y}
\filmbox{19}{Harry Potter and the Chamber of Secrets}{2002}{48}{y}
\filmbox{20}{Star Trek}{2009}{47.5}{y}
\filmbox{21}{Spider-Man}{2002}{47}{y}
\filmbox{22}{The Chronicles of Narnia: The Lion, the Witch and the Wardrobe}{2005}{42.25}{y}
\filmbox{23}{Pirates of the Caribbean: At World's End}{2007}{41.75}{y}
\filmbox{24}{Shrek}{2001}{41.5}{y}
\filmbox{25}{Shrek 2}{2004}{41.25}{n}
\filmbox{26}{WALL·E}{2008}{41}{y}
\filmbox{27}{Spider-Man 2}{2004}{40.75}{n}
\filmbox{28}{Casino Royale}{2008}{39.25}{y}
\filmbox{29}{The Departed}{2006}{38.25}{y}
\filmbox{30}{How the Grinch Stole Christmas}{2000}{38}{n}
\filmbox{31}{Gladiator}{2000}{37.5}{y}
\filmbox{32}{Troy}{2004}{37}{n}
\filmbox{33}{No Country for Old Men}{2007}{36.5}{n}
\filmbox{34}{Monsters, Inc.}{2001}{35.25}{y}
\filmbox{35}{Iron Man}{2008}{34}{y}
\filmbox{36}{The Matrix: Reloaded}{2003}{33.25}{y}
\filmbox{37}{Shrek the Third}{2007}{32.75}{y}
\filmbox{38}{300}{2006}{30.75}{y}
\filmbox{39}{Eternal Sunshine of the Spotless Mind}{2004}{30.5}{n}
\filmbox{40}{Ice Age: The Meltdown}{2006}{30}{n}
\filmbox{41}{Ice Age: Dawn of the Dinosaurs}{2009}{30}{y}
\filmbox{42}{The Bourne Ultimatum}{2007}{29.75}{n}
\filmbox{43}{Cast Away}{2000}{29.75}{n}
\filmbox{44}{The Twilight Saga: New Moon}{2009}{29.5}{y}
\filmbox{45}{The Da Vinci Code}{2006}{29.25}{n}
\filmbox{46}{There Will Be Blood}{2007}{29}{n}
\filmbox{47}{Twilight}{2008}{28.5}{y}
\filmbox{48}{Batman Begins}{2005}{28}{y}
\filmbox{49}{Madagascar}{2005}{27.75}{n}
\filmbox{50}{X-Men: The Last Stand}{2006}{27.5}{y}
\filmbox{51}{Elf}{2003}{26.25}{n}
\filmbox{52}{Pride \& Prejudice}{2005}{26.25}{y}
\filmbox{53}{Spider-Man 3}{2007}{26.25}{n}
\filmbox{54}{The Hangover}{2009}{26}{y}
\filmbox{55}{Transformers}{2007}{25.75}{y}
\filmbox{56}{The Simpsons Movie}{2007}{25.25}{n}
\filmbox{57}{Love Actually}{2003}{25}{n}
\filmbox{58}{What Women Want}{2000}{25}{n}
\filmbox{59}{Ice Age}{2002}{24.75}{y}
\filmbox{60}{Mission: Impossible II}{2000}{24.75}{n}
\filmbox{61}{Hancock}{2008}{24.25}{n}
\filmbox{62}{Sideways}{2004}{24}{n}
\filmbox{63}{Transformers: Revenge of the Fallen}{2009}{23.75}{y}
\filmbox{64}{The Fast and the Furious}{2001}{23}{y}
\filmbox{65}{Mean Girls}{2004}{22.75}{n}
\filmbox{66}{The Queen}{2006}{22.75}{n}
\filmbox{67}{Pearl Harbor}{2001}{22.75}{n}
\filmbox{68}{Men in Black II}{2002}{22.5}{n}
\filmbox{69}{Anchorman: The Legend of Ron Burgundy}{2004}{22.25}{n}
\filmbox{70}{Madagascar: Escape 2 Africa}{2008}{22.25}{n}
\filmbox{71}{Inglorious Basterds}{2009}{22}{n}
\filmbox{72}{Almost Famous}{2000}{22}{n}
\filmbox{73}{Gosford Park}{2001}{21.75}{y}
\filmbox{74}{X-Men Origins: Wolverine}{2009}{21.5}{y}
\filmbox{75}{United 93}{2006}{21.5}{n}
\filmbox{76}{Indiana Jones and the Kingdom of the Crystal Skull}{2008}{21.5}{n}
\filmbox{77}{Lost in Translation}{2003}{21.25}{y}
\filmbox{78}{Bruce Almighty}{2003}{21.25}{n}
\filmbox{79}{Meet the Fockers}{2004}{21.25}{n}
\filmbox{80}{Quantum of Solace}{2008}{21}{n}
\filmbox{81}{Borat}{2006}{20.75}{n}
\filmbox{82}{The Notebook}{2004}{20.5}{y}
\filmbox{83}{Kung Fu Panda}{2008}{20.5}{y}
\filmbox{84}{The Mummy Returns}{2001}{20.25}{n}
\filmbox{85}{Donnie Darko}{2001}{20}{y}
\filmbox{86}{Letters from Iwo Jima}{2006}{20}{n}
\filmbox{87}{Fast \& Furious}{2009}{19.5}{n}
\filmbox{88}{Capote}{2005}{19.5}{y}
\filmbox{89}{Chicken Run}{2000}{19.25}{n}
\filmbox{90}{Requiem for a Dream}{2000}{19}{n}
\filmbox{91}{District 9}{2009}{19}{y}
\filmbox{92}{The Prestige}{2006}{18.75}{y}
\filmbox{93}{Wallace \& Gromit: The Curse of the Were-Rabbit}{2005}{18.75}{n}
\filmbox{94}{Brokeback Mountain}{2005}{18.5}{n}
\filmbox{95}{King Kong}{2005}{18.5}{n}
\filmbox{96}{Ocean's Eleven}{2001}{18.5}{n}
\filmbox{97}{Night at the Museum}{2006}{18.5}{n}
\filmbox{98}{Slumdog Millionaire}{2008}{18.25}{n}
\filmbox{99}{American Pie 2}{2001}{18.25}{n}
\filmbox{100}{Million Dollar Baby}{2004}{18}{n}  

\section{{APPENDIX B – List of Referenced Films}}\label{appendix_b}
\sloppy

\medskip\noindent\textit{300.} 2006. Directed by Zack Snyder. USA: Warner Bros., Legendary Entertainment, Virtual Studios; Warner Bros.
\footnote{Films will be listed with main director, main production companies, and main distributor. Further information can be found on http://www.imdb.com [2016-10-20]. Country codes are based on ISO 3166.}



\medskip\noindent\textit{Annie Hall.} 1977. Directed by Woody Allen. USA: Rollins-Joffe Productions; MGM Home Entertainment.



\medskip\noindent\textit{Apocalypse Now.} 1979. Directed by Francis Ford Coppola. USA: Zoetrope Studios; Paramount Home Entertainment.



\medskip\noindent\textit{Austin Powers in Goldmember.} 2002. Directed by Jay Roach. USA: New Line Cinema, Gratitude International, Team Todd; Warner Home Video.



\medskip\noindent\textit{Avatar.} 2009. Directed by James Cameron. USA/UK: Twentieth Century Fox Film Corporation, Dune Entertainment, Ingenious Film Partners; Twentieth Century Fox Home Entertainment.



\medskip\noindent\textit{Bad Company.} 2002. Directed by Joel Schumacher. USA/CZE: Touchstone Pictures, Jerry Bruckheimer Films, Stillking Films; Buena Vista Home Entertainment.



\medskip\noindent\textit{Batman Begins.} 2005. Directed by Christopher Nolan. USA/UK: Warner Bros., Syncopy, DC Comics; Warner Bros.



\medskip\noindent\textit{Being Human} [Television Series]. 2008-2013. Created by Toby Whithouse. UK: Touchpaper Television; British Broadcasting Corporation.



\medskip\noindent\textit{Brother Bear.} 2003. Directed by Aaron Blaise. USA: Walt Disney Feature Animation, Walt Disney Pictures; Buena Vista International.



\medskip\noindent\textit{Carry On Only} [Short]. 2013. Directed by Christopher Loope. USA: no company credits.



\medskip\noindent\textit{Casablanca.} 1942. Directed by Michael Curtiz. USA: Warner Bros.; Warner Bros.



\medskip\noindent\textit{Die Hard.} 1988. Directed by John McTiernan. USA: Twentieth Century Fox Film Corporation, Gordon Company, Silver Pictures; Twentieth Century Fox Home Entertainment.



\medskip\noindent\textit{District 9.} 2009. Directed by Neill Blomkamp. ZA/USA/NZ/CAN: TriStar Pictures, Block/Hanson, WingNut Films; Sony Pictures Releasing.



\medskip\noindent\textit{Dvevnoy Dozor} [Day Watch]. 2006. Directed by Timur Bekmambetov. RUS: Bazelevs Production, Challen One Russia, TABBAK; Twentieth Century Fox Home Entertainment.



\medskip\noindent\textit{Eagle Eye.} 2008. Directed by D.~J. Caruso. USA: DreamWorks SKG, Goldcrest Pictures, KMP Film Invest; Paramount Home Entertainment.



\medskip\noindent\textit{Eternal Sunshine of the Spotless Mind.} 2004. Directed by Michel Gondry. USA: Focus Features, Anonymous Content, This Is That Productions; Constantin Film.



\medskip\noindent\textit{Ex Machina}. 2015. Directed by Alex Garland. UK: Universal Pictures International, Film4, DNA Films; Universal Pictures International.



\medskip\noindent\textit{Fast Five}. 2011. Directed by Justin Lin. USA: Original Film, One Race Films, Dentsu; Universal Pictures.



\medskip\noindent\textit{Fast \& Furious 6}. 2013. Directed by Justin Lin. USA: Universal Pictures, Relativity Media, Original Film; Universal Pictures International.



\medskip\noindent\textit{Fast \& Furious 7}. 2015. Directed by James Wan. USA/JP: Universal Pictures, MRC, China Film Co.; Universal Pictures International.



\medskip\noindent\textit{Finding Nemo}. 2003. Directed by Andrew Stanton. USA: Walt Disney Pictures, Pixar Animation Studios, Disney Enterprises; Buena Vista International.
 
\medskip\noindent\textit{Game of Thrones} [Television Series]. 2011-. Created by David Benioff and D.~B. Weiss. USA/UK: Home Box Office, Television 360, Grok! Studio; HBO Home Entertainment.
 


\medskip\noindent\textit{Gone Girl}. 2014. Directed by David Fincher. USA: Twentieth Century Fox Film Corporation, Regency Enterprises, TSG Entertainment; Twentieth Century Fox Home Entertainment.



\medskip\noindent\textit{Grand Budapest Hotel}. 2014. Directed by Wes Anderson. USA/DE/UK: Fox Searchlight Pictures, Indian Paintbrush, Studio Babelsberg; Twentieth Century Fox Home Entertainment.



\medskip\noindent\textit{Harry Potter} [Franchise]. 2001-2011. UK/USA: Warner Bros.; Warner Bros.



\medskip\noindent\textit{Heroes} [Television Series]. 2006-2010. Created by Tim Kring. USA: Tailwind Productions; Universal Pictures.



\medskip\noindent\textit{Heroes Reborn} [Television Series]. 2015-2016. Created by Tim Kring. USA: Tailwind Productions, Imperative Entertainment, Universal Television; Universal Pictures.



\medskip\noindent\textit{Hot Fuzz.} 2007. Directed by Edgar Wright. UK/FR/USA: Universal Pictures, StudioCanal, Working Title Films; Universal Pictures.



\medskip\noindent\textit{House of Cards} [Television Series]. 2013-. Created by Beau Willimon. USA: Media Rights Capital, Panic Pictures (II), Trigger Street Productions; Sony Pictures Home Entertainment.



\medskip\noindent\textit{Ice Age.} 2002. Directed by Chris Wedge and Carlos Saldanha. USA: Twentieth Century Fox Film Corporation, Blue Sky Studios, Twentieth Century Fox Animation; Twentieth Century Fox Home Entertainment.



\medskip\noindent\textit{In Bruges.} 2008. Directed by Martin McDonagh. UK/USA: Blueprint Pictures, Focus Features; Scion Films; Universum Film.



\medskip\noindent\textit{Iron Man.} 2008. Directed by Jon Favreau. USA: Paramount Pictures, Marvel Enterprises, Marvel Studios; Concorde Home Entertainment.



\medskip\noindent\textit{Jaws}. 1975. Directed by Steven Spielberg. USA: Zanuck/Brown Productions, Universal Pictures; Universal Pictures.



\medskip\noindent\textit{John Wick}. 2014. Directed by Chad Stahelski and David Lynch. USA: 87Eleven; Lionsgate Entertainment.



\medskip\noindent\textit{Joining the Dots} [Short]. 2012. Directed by Pablo Romero-Fresco. UK: no company credits.



\medskip\noindent\textit{La Antena}. 2007. Directed by Esteban Sapir. AR: LadobleA; Capelight Pictures.



\medskip\noindent\textit{Lilo \& Stitch}. 2002. Directed by Dean DeBlois and Chris Sanders. USA: Walt Disney Pictures, Walt Disney Feature Animation; Buena Vista International.



\medskip\noindent\textit{Limitless}. 2011. Directed by Neil Burger. USA: Relativity Media, Virgin Produced, Rogue; Twentieth Century Fox Home Entertainment.



\medskip\noindent\textit{M}. 1951. Directed by Joseph Losey. USA: Superior Pictures; Darker Images Video.



\medskip\noindent\textit{Man on Fire.} 2004. Directed by Tony Scott. USA/UK: Warner Bros. International; Twentieth Century Fox.



\medskip\noindent\textit{Montreal Stories.} CAN 1991. Various directors. CAN: Atlantis Films, Cinémaginaire, National Film Board of Canada.



\medskip\noindent\textit{Monty Python and the Holy Grail.} 1975. Directed by Terry Gilliam and Terry Jones. UK: Michael White Productions, National Film Trustee Company, Python (Monty) Pictures; Sony Pictures Home Entertainment.



\medskip\noindent\textit{Nochnoy Dozor} [Night Watch]. 2004. Directed by Timur Bekmambetov. RUS: Channel One; Twentieth Century Fox.



\medskip\noindent\textit{Non-Stop}. 2014. Directed by Jaume Collet-Serra. UK/FR/USA/CAN: StudioCanal, Anton Capital Entertainment, Silver Pictures; StudioCanal.



\medskip\noindent\textit{Notes on Blindness}. 2016. Directed by Peter Middleton and James Spinney. UK: Arte France, Creative England, Impact Partners; ARTE.



\medskip\noindent\textit{Pirates of the Carribean: The Curse of the Black Pearl.} 2003. Directed by Gore Verbinski. USA: Walt Disney Pictures, Jerry Bruckheimer Films; Buena Vista International.



\medskip\noindent\textit{Pirates of the Carribean: Dead Man’s Chest.} 2006. Directed by Gore Verbinski. USA: Walt Disney Pictures, Jerry Bruckheimer Films, Second Mate Productions; Buena Vista International.



\medskip\noindent\textit{Pirates of the Carribean: At World’s End.} 2007. Directed by Gore Verbinski. USA: Walt Disney Pictures, Jerry Bruckheimer Films, Second Mate Productions; Buena Vista International.



\medskip\noindent\textit{Pride \& Prejudice.} 2005. Directed by Joe Wright. FR/UK/USA: Focus Features, Universal Pictures, StudioCanal; Universal Studios Home Entertainment.



\medskip\noindent\textit{Psycho.} 1960. Directed by Alfred Hitchcock. USA: Shamley Productions; Universal Pictures.



\medskip\noindent\textit{Ronin.} 1998. Directed by John Frankenheimer. UK/FR/USA: FGM Entertainment, United Artists Corporation, United Artists; MGM Home Entertainment.



\medskip\noindent\textit{Se7en.} 1995. Directed by David Fincher. USA: Cecchi Gori Pictures, Juno Pix, New Line Cinema; Warner Home Video.



\medskip\noindent\textit{Seven Pounds.} 2008. Directed by Gabriele Muccino. USA: Columbia Pictures, Relativity Media, Overbrook Entertainment; Sony Pictures Home Entertainment.



\medskip\noindent\textit{Sherlock} [Television Series]. 2010-. Created by Mark Gatiss and Steven Moffat. UK/USA: Hartswood Films, BBC Wales, Masterpiece Theatre; British Broadcasting Corporation.



\medskip\noindent\textit{Sherlock Holmes – A Game of Shadows}. 2011. Directed by Guy Ritchie. USA: Warner Bros., Village Roadshow Pictures, Silver Pictures; Warner Home Video.



\medskip\noindent\textit{Sin City.} 2005. Directed by Frank Miller and Robert Rodiguez. USA: Dimension Films, Troublemaker Studios; Buena Vista Home Entertainment.



\medskip\noindent\textit{Slumdog Millionaire}. 2008. Directed by Danny Boyle and Loveleen Tandan. UK/FR/USA: Celador Films Ltd.; Twentieth Century Fox.



\medskip\noindent\textit{Star Trek.} 2009. Directed by J.~J. Abrams. USA: Paramount Pictures, Spyglass Entertainment, Bad Robot; Paramount Home Entertainment.



\medskip\noindent\textit{Star Trek Into Darkness}. 2013. Directed by J.J. Abrams. USA: Paramount Pictures, Spyglass Entertainment, Bad Robot; Paramount Pictures.



\medskip\noindent\textit{Star Wars} [Franchise]. 1977-. USA: various.



\medskip\noindent\textit{Star Wars: Episode III – Revenge of the Sith.} 2005. Directed by George Lucas. USA: Lucasfilm; Twentieth Century Fox Home Entertainment.



\medskip\noindent\textit{Stranger Than Fiction.} 2006. Directed by Marc Forster. USA: Columbia Pictures, Mandate Pictures, Three Strange Angels; Sony Pictures Home Entertainment.



\medskip\noindent\textit{The Big Bang Theory} [Television Series]. 2007-. Created by Chuck Lorre and Bill Prady. USA: Chuck Lorre Productions, Warner Bros. Television; Warner Home Video.



\medskip\noindent\textit{The Dancer.} 2000. Directed by Frédéric Garson. FR: EuropaCorp, TF1 Films Production; Tobis StudioCanal.



\medskip\noindent\textit{The Departed}. 2006. Directed by Martin Scorsese. USA/HK: Warner Bros., Plan B Entertainment, Initial Entertainment Group; Warner Home Video.



\medskip\noindent\textit{The Godfather: Part II.} 1974. Directed by Fracis Ford Coppola. USA: Paramount Pictures, The Coppola Company; Paramount Home Entertainment.



\medskip\noindent\textit{The Incredibles}. 2004. Directed by Brad Bird. USA: Walt Disney Pictures, Pixar Animation Studios; Buena Vista Home Entertainment.



\medskip\noindent\textit{The Lord of the Rings} [Franchise]. 2001-2003. Directed by Peter Jackson. NZ/USA: New Line Cinema, WingNut Films, The Saul Zaentz Company; Warner Home Video.



\medskip\noindent\textit{The Lord of the Rings: The Fellowship of the Ring.} 2001. Directed by Peter Jackson. NZ/USA: New Line Cinema, WingNut Films, The Saul Zaentz Company; Warner Home Video.



\medskip\noindent\textit{The Lord of the Rings – The Two Towers.} 2002. Directed by Peter Jackson. NZ/USA: New Line Cinema, Wingnut Films, The Saul Zaentz Company; Warner Home Video.



\medskip\noindent\textit{The Matrix} [Franchise]. 1999-2003. Directed by Lana and Lilly Wachowski. USA: Warner Bros., Roadshow Pictures; Warner Bros.



\medskip\noindent\textit{The Rundown.} 2003. Directed by Peter Berg. USA: Columbia Pictures Corporation, Universal Pictures, WWE Studios; Columbia TriStar Home Entertainment.



\medskip\noindent\textit{The Twilight Saga: New Moon.} 2009. Directed by Chris Weitz. USA: Temple Hill Entertainment, Maverick Films, Imprint Entertainment; Concorde Home Entertainment.



\medskip\noindent\textit{Train of Life.} 1998. Directed by Radu Mihaileanu. FR/BE/NL/ISR/ROU: Belfilms, Canal+, Centre National de la Cinématographie; Sunfilm Entertainment.



\medskip\noindent\textit{Trainspotting.} 1996. Directed by Danny Boyle. UK: Channel Four Films; Figment Films, The Noel Gay Motion Picture Company; Prokino Filmverleih.  

\medskip\noindent\textit{Vivre Sa Vie.} 1962. Directed by Jean-Luc Godard. FR: Les Films de la Pléiade, Pathé Consortium Cinéma; Panthéon Distribution.
