\author{Constantijn Kaland}  
\title{Word stress in prosodic theory}
\dedication{\textit{voor mijn ouders}}
% \subtitle{Add subtitle here if it exists}
\renewcommand{\lsISBNdigital}{978-3-96110-512-0}
\renewcommand{\lsISBNhardcover}{978-3-98554-141-6}
\BackBody{This book is composed of four studies that all investigate different aspects of word stress in Papuan Malay, an Austronesian language spoken in eastern Indonesia. These aspects, in order of presentation, include acoustic realisation, auditory perception, lexical analyses and word disambiguation. The introduction provides the theoretical background against which the studies are undertaken. All studies are empirical in nature; they either report acoustic analyses, production or perception experiments, or corpus-based analyses. Taken together, the results of all studies pose a challenge to maintaining a stressless analysis of Papuan Malay. At the same time, the type of word stress that emerges from the reported results is unlike its common theoretical conception and therefore requires more work to be integrated in prosodic theory. Given the controversy on word stress in Indonesian languages, the results are always discussed and carefully interpreted in a cross-linguistic context. In this way, the current thesis extends and deepens our knowledge and understanding of word stress in prosodic theory.}
\BookDOI{10.5281/zenodo.15437940}
\typesetter{Constantijn Kaland}
\proofreader{Bruno Behling,
Jeroen van de Weijer,
Ludger Paschen,
Mary Ann Walter,
Raquel Benítez Burraco
}
\renewcommand{\lsID}{472}
% \lsCoverTitleSizes{51.5pt}{17pt}
\renewcommand{\lsSeries}{silp}
\renewcommand{\lsSeriesNumber}{15}
