\addchap{\lsAcknowledgementTitle} 

This work is the fruit of many years research in \is{Linguistics!Missionary linguistics}Australian missionary linguistics. Early versions were presented at meetings of the Society for the History of Linguistics in the Pacific, with the encouragement of \ia{McGregor, William B.}Bill McGregor. For the present work, I thank \ia{Wafer, Jim}Jim Wafer who suggested I write an introduction to a proposed modern edition of \ia{Threlkeld, L.E.}Threlkeld’s corpus of scripture translations. He generously read and corrected my original draft, and provided the transcriptions into modern orthography for the \il{Hunter River Lake Macquarie language (HRLM)}Hunter River Lake Macquarie language (HRLM).

Further research has been supported by a grant from the UK Arts and Humanities Council (AH/X001881/1) and the German Deutsche Forschungsgemeinschaft (DFG) as part of the Global Bible project.

The \citeauthor{fraser_gospel_1891} (\citeyear{fraser_gospel_1891}) edition of \ia{Threlkeld, L.E.}Threlkeld’s Gospel of St Luke, is reproduced with permission from Cambridge University Library from the archive of the British and Foreign Bible Society.

GMS 83 is reproduced with permission from Auckland Council Libraries Heritage Collections.

For copyediting and formatting the digital files, I am more than grateful to Dustin Saynisch, whose work went above and beyond anything I might have expected. His expertise contributed significantly to the quality of this volume.

I would also like to thank Keith Giles, Photograph Collections Librarian of Auckland Libraries, who kindly posted me high resolution images of GMS 83, which enabled the text to be read using Transkribus handwriting recognition software.

\bigskip

\noindent Hilary M. Carey\\
University of Bristol\\
August 2024
