\addchap{\lsAcknowledgementTitle}
\largerpage[2.75]

The editors would like to acknowledge the support of the following:

\begin{itemize}\itemsep=.5\itemsep
\item German Research Foundation (DFG grant WA 2432/4-1)
\item Johannes-Gutenberg University Mainz (Stufe I inneruniversitäre Förderung)
\item Publishers: de Gruyter Mouton, Edinburgh University Press, John Benjamins Publishing Company, Universitätsverlag Winter (Heidelberg)
\end{itemize}

We are also tremendously grateful to our full-time student helpers, without whom the conference wouldn't have been half as well-organised as it was. Major thanks goes to \textbf{Natascha Kessler} and \textbf{Johannes Schindler}, who spent too many hours in unairconditioned offices during the hottest time of the year with their heads pretty much in their laptops to figure out how to accommodate all scheduling requests while still coming up with a half-way logical programme (you should have seen the Excel sheet!).

A big thank-you is also due to our other student helpers, \textbf{Victoria Fox} and \textbf{Lena Jubelius}, who made bag-packing a science and worked overtime willingly because they liked the conference atmosphere so much. 

\textbf{Christiane Dinger} probably dreamt of money for a couple of months while she was figuring out the finances with our administration – thank you, Christiane! Susanne would also like to thank Ulrike, who took over most of the editorial work while Susanne had to handle some unforeseen commitments (including becoming vice dean) – thank you, Ulrike! Heartfelt thanks also go to \textbf{Emelie Lambertz} and \textbf{Daniel Kastenholz} for their invaluable assistance with the finer details of the editorial work on this volume, which significantly accelerated the publication process.

Last but not least, the biggest thank-you goes out to our plenary speakers, presenters and participants – obviously, without you, Methods XVII wouldn't have been possible. While we had to wait two years for the Covid pandemic to pass (the conference was originally scheduled for 2020), it didn't take away any of the ``Methods magic". We are looking forward to many more successful conferences in the series.

\bigskip
\noindent
Mainz, \monthname{} \the\year{} \hfill Susanne Wagner \& Ulrike Stange-Hundsdörfer
