% Appendix B
\chapter{The Rat, the Swamp Hen, and the Squid (Philippe Gohoupe)}
\label{text:squid2}
This story is known all over New Caledonia, and originally from Tiga in the Loyalty Islands. The underlying information of this story, that squids hate rats, and that rats make excellent bait to catch squids, is known beyond the New Caledonian archipelago.\largerpage[2]

\ea%<Hc2-1>

%%\langinfo{}{}{} koin hupwan maa hmasut xahut ka cahma i khîî ahan moo kon maa
\gll Koin hut pwa-n maa a=hma-sut xahut ka cahma i=khîî a=han moo ko-n maa\\ 
end go.down on-\gl{nspec} reef 3\gl{sg}=arrive-go.down below \gl{cnj} \gl{top} \gl{def}.\gl{sg}=swamp.hen \gl{3}\gl{sg}=go stay on-\gl{nspec} reef\\ 
\glt \qu{When the receding tide was low, the Swamp Hen went to stay on the reef [to fish]}

\z
\ea%<Hc2-2>

%%\langinfo{}{}{} ka cahma i siimbwi kon moo cai wâng
\gll ka cahma i=siibwi koon moo ca i=wang\\ 
\gl{cnj} \gl{top} \gl{def}.\gl{sg}=rat \gl{prog} stay in \gl{def}.\gl{sg}=boat\\ 
\glt \qu{and the Rat stayed in the boat.}

\z
\ea%<Hc2-3>

%\langinfo{}{}{} ca i wângalu, a lu vwa wângalu
\gll Ca i=wanga-lu a= lu=vwa wanga-lu
\\ in \gl{def}.\gl{sg}=boat-\gl{3}\gl{du}.\gl{poss} \gl{rel}= \gl{3}\gl{du}=do boat-\gl{3}\gl{du}.\gl{poss}
\\ \glt \qu{In their boat that they made for themselves...}

\z
\ea%<Hc2-4>

%\langinfo{}{}{} ko i ujep. ma moo ka i siibwi a xhuti
\gll Ko i=ujep ma a=moo ka=i=siibwi a xhuti\\ 
\gl{obl} \gl{def}.\gl{sg}=sugar.cane \gl{subr} 3\gl{sg}=stay \gl{sbj}=\gl{def}.\gl{sg}=rat \gl{3}\gl{sg}=scrunch.(sugarcane)\\ 
\glt \qu{...from sugar cane. When the Rat stayed like this, it ate it.}
\z

\ea%<Hc2-5>
%\langinfo{}{}{} cama hamwame ka khîî ma a hmacamwame, paa cika mwa wâng
\gll Cama han-mwa-me ka=khîî, ma=a=hma-ca-mwa-me paa cika mwa wang\\ 
when go-\gl{rep}-\gl{dir.cp} \gl{sbj}=swamp.hen when=\gl{3}\gl{sg}=arrive-go.up-\gl{rep}-\gl{dir.cp} \gl{prf} \gl{neg}.\gl{exist} \gl{rep} boat\\ 
\glt \qu{When Swamp Hen came back, when she came back, the boat was already gone.}

\z
\ea%<Hc2-6>

%\langinfo{}{}{} uwang sin ujep wang-lu ma siibwi pa xhuti meka na siibwi
\gll I=wang sin ujep wang lu ma siibwi paa xhuti meekan a siibwi\\ \gl{def}.\gl{sg}=boat arm sugar.cane boat \gl{3}\gl{du} \gl{subr} rat \gl{pfv} scrunch.(sugarcane) all \gl{sbj} rat\\ \glt \qu{The boat [of] sugar cane branches, their boat, but Rat had already eaten everything.}

\z
\ea%<Hc2-7>

%\langinfo{}{}{} amwa fajimwa ka vi hapi
\gll A=mwa fwajimwake ka vi hapi
\\ 3\gl{sg}=\gl{deict} ask and say \gl{comp}
\\ \glt \qu{She asked him and he said that...}

\z
\ea%<Hc2-8>

%\langinfo{}{}{} a moo a hmanan a paa xhuti
\gll A=moo a=hmana-n a=paa xhuti\\ 
\gl{3}\gl{sg}=stay \gl{3}\gl{sg}=hungry-3\gl{sg} \gl{3}\gl{sg}=\gl{prf} scrunch.(sugarcane)\\ 
\glt \qu{...he stayed, he was hungry, and he ate...}

\z
\ea%<Hc2-9>

%\langinfo{}{}{} xhuti i wâng
\gll xhuti i=wang\\ 
scrunch.(sugarcane) \gl{def}.\gl{sg}=boat\\ 
\glt \qu{...ate the boat.}

\z
\ea%<Hc2-10>

%\langinfo{}{}{} go a vi nyakon a khîî
\gll Go a=vi nyakoo-n a=khîî\\ 
\gl{disc} \gl{3}\gl{sg}=say for-3\gl{sg}.\gl{poss} \gl{sbj}=swamp.hen\\
 \glt \qu{Then Swamp Hen told him:}

\z
\ea%<Hc2-11>

%\langinfo{}{}{} bwa evaang go go cika ca=hun-ta-mwa ko yo vwa li=vuun sung
\gll Bwa e-vaang go go cika eca=hun-ta-mwa ko yo vwa li=vwun s-ung\\ \gl{ipfv} \gl{mid}-unknown \gl{2}\gl{sg} \gl{2}\gl{sg} \gl{neg}.\gl{exist} \gl{indf}.\gl{sg}=\gl{nmlz}-move.up-\gl{rep} because \gl{1}\gl{sg} do \gl{def}.\gl{pl}=hair arm-1\gl{sg}.\gl{poss}\\ \glt \qu{``Well I don't know about you, you don't have any means to go back up, whereas I have my wings."}

\z
\ea%<Hc2-12>

%\langinfo{}{}{} go tha ja tamwame
\gll Go tha a=ja ta-mwa-me\\ then \gl{ass} 3\gl{sg}=\gl{prf} move.up-\gl{rep}-\gl{dir.cp}\\ \glt \qu{And she went back up to the land.}

\z
\ea%<Hc2-13>

%\langinfo{}{}{} ya tha ja thêên thên tamwame
\gll Ya tha ja thêên thêên ta-mwa-me\\ 3\gl{sg} \gl{ass} \gl{prf} run run move.up-\gl{rep}-\gl{dir.cp}\\ \glt \qu{She, she flew up fast to go home.}

\z
\ea%<Hc2-14>

%\langinfo{}{}{} cama i siibwi a balan tabo kon se
\gll Cama i=siibwi a=balan tabo koon se\\ if \gl{def}.\gl{sg}=rat \gl{3}\gl{sg}=finally sit \gl{prog} cry\\ \glt \qu{But the Rat, he just sat there crying.}

\z
\ea%<Hc2-15>

%\langinfo{}{}{} ko i vai a foovwat kon tabo xahut pwan bwanma
\gll Ko i=vai a= fovwat koon tabo xahut pwan bwan-maa\\ on \gl{def}.\gl{sg}=stone \gl{rel}= white \gl{prog} sit below on top-reef\\ \glt \qu{on the white [coral] rock, he was sitting down there on the reef.}

\z
\ea%<Hc2-16>

%\langinfo{}{}{} koin a kon a bwa tena ka i ibwen
\gll Koin a=kon a=bwa tena-a ka i=ibwen\\ stop \gl{3}\gl{sg}=\gl{prog} \gl{3}\gl{sg}=\gl{ipfv} hear-3\gl{sg}.\gl{obj} \gl{sbj} \gl{def}.\gl{sg}=squid\\ \glt \qu{Then he was - the Squid heard him.}

\z
\ea%<Hc2-17>

%\langinfo{}{}{} go a hame a hmacame koon a mwa fajimwa go kon se ko hmwaeke
\gll Go a=hame a=hma-ca-me koo-n a=mwa fwajimwa ``go=kon se ko hmwaeke?"\\ 2\gl{sg} 3\gl{sg}=come 3\gl{sg}=arrive-go.up-\gl{dir.cp} because-\gl{psm} \gl{3}\gl{sg}=\gl{deict} ask 2\gl{sg}=\gl{prog} cry because how\\ \glt \qu{Then he came he came there to him, he was like ``why do you cry?"}

\z
\ea%<Hc2-18>

%\langinfo{}{}{} kavi hapi a se-a moo a=hmanong
\gll Ka a=vi hapi a=se-a moo a=hman-ong\\ and 3\gl{sg}=say \gl{comp} 3\gl{sg}=one-\gl{3}\gl{sg} stay \gl{3}\gl{sg}=hungry-\gl{1}\gl{sg}.\gl{poss}\\ \glt \qu{And he said that he was alone and ``I was hungry".}

\z
\ea%<Hc2-19>

%\langinfo{}{}{} a xhuti wâng aju
\gll A=xhuti i=wanga-ju\\ \gl{3}\gl{sg}=scrunch.(sugarcane) \gl{def}.\gl{sg}=boat.\gl{poss}-3\gl{du}.\gl{poss}\\ \glt \qu{He ate our boat.}

\z
\ea%<Hc2-20>

%\langinfo{}{}{} paa ja pa cika mwa ca=wang-abu m=abu taa-mwa kabu 
\gll Paa ja paa cika mwa ca=wang-abu m=abu ta-mwa ka=bu\\ \gl{pfv} \gl{prf} \gl{pfv} \gl{neg}.\gl{exist} \gl{rep} \gl{indf}.\gl{sg}=boat-\gl{1}\gl{du}.\gl{excl} \gl{subr}=\gl{1}\gl{du}.\gl{excl} move.up-\gl{rep} \gl{sbj}=1\gl{du}.\gl{excl}\\ \glt \qu{And now there is no more, no more boat of ours with which we could go back up.}
\z

\ea%<Hc2-21>
%\langinfo{}{}{} go ya avi hapi go tha vaang go yo theja thêên tamwa ko vwa li=vuun sung
\gll  Go ya a=vi hapi go tha vaang go yo thêên tamwa ko vwa li=vwun s-ung\\  \gl{disc} \gl{3}\gl{sg} \gl{3}\gl{sg}=say \gl{comp} \gl{2}\gl{sg} \gl{ass} inconnu \gl{2}\gl{sg} \gl{1}\gl{sg} run go.back.up because \gl{exist} \gl{def}.\gl{pl}=hair arm-1\gl{sg}.\gl{poss}\\ \glt \qu{And she said “I don't know about you, I'm leaving because I have wings."}
\z

\ea%<Hc2-22>
%\langinfo{}{}{} ko camwa siibwi ko tabo kon se se se a tena ka ibwen a hame
\gll Ko cahma siibwi koon tabo koon se se se a=tena ka ibwen a=ha-me\\ \gl{cnj} \gl{top} rat \gl{prog} sit \gl{prog} cry cry cry \gl{3}\gl{sg}=hear \gl{sbj} squid \gl{3}\gl{sg}=go-\gl{dir.cp}\\ \glt \qu{And Rat, he sits, he cries, cries, cries, Squid hears and comes.}

\z
\ea%<Hc2-23>

%\langinfo{}{}{} a fajimwa ha go siibwi go se se ko hmwaeke
\gll A=fwajimwake ha go siibwi go=se se ko hmwaeke\\ \gl{3}\gl{sg}=ask \gl{expl} \gl{2}\gl{sg} rat \gl{2}\gl{sg}=cry cry \gl{obl} how\\ \glt \qu{He asks ``Hey you, Rat, why do you cry, cry?"}
\z

\ea%<Hc2-24>
%\langinfo{}{}{} e=se ko e=xhuti i wangabe, abe hupwe ma khîî kon ya tha paa tamwa
\gll E=se ko e=xhuti i=wang-abe abe=hup-e ma khîî kon ya tha paa ta-mwa\\ \gl{1}\gl{sg}=cry because 1\gl{sg}=scrunch.(sugarcane) \gl{def}.\gl{sg}=boat-\gl{1}\gl{pl}.\gl{excl}.\gl{poss}\gl{1}\gl{pl}.\gl{excl}=go.down-\gl{dir.cp} \gl{subr} swamp.hen \gl{cnj} \gl{3}\gl{sg} \gl{ass} \gl{prf} go.up-\gl{rep}\\ \glt \qu{``I cry because I ate our boat, we came down with Swamp Hen, and she already went back up."}

\z
\ea%<Hc2-25>
%\langinfo{}{}{} go ka go hame ko e=bo fetamwako
\gll Go ka go=ha-me ko e=bo fe-ta-mwa-ko\\ \gl{2}\gl{sg} \gl{sbj} \gl{2}\gl{sg}=go-\gl{dir.cp} because \gl{1}\gl{sg}=\gl{irr} take-move.up \gl{rep}-2\gl{sg}\\ \glt \qu{``You, come, because I will take you back up."}

\z
\ea%<Hc2-26>

%\langinfo{}{}{} hê a ja han tabo pwan bwan kona ja fetamwamea
\gll Hê a=ja han tabo pwa-n bwa-n kona a=ja fe-ta-mwa-me-a\\ yes \gl{3}\gl{sg}=\gl{prf} go sit on-\gl{nspec} head-3\gl{sg}.\gl{poss} \gl{cnj} 3\gl{sg}=\gl{prf} take-move.up-\gl{rep}-\gl{dir.cp}-\gl{3}\gl{sg}.\gl{obj}\\ \glt \qu{Yes, he goes to sit on his head and he takes him back up.}

\z
\ea%<Hc2-27>

%\langinfo{}{}{} lu hmacamwame can hmeewan
\gll Lu=hma-ca-mwa-me ca-n hmeewan\\ \gl{3}\gl{du}=arrive-go.up-\gl{rep}-\gl{dir.cp} on-\gl{nspec} sand\\ \glt \qu{The two arrive up at the beach,}

\z
\ea%<Hc2-28>

%\langinfo{}{}{} ma xhasat nyadamwa siibwi phâêû
\gll Ma xhasaat nya-da-mwa siibwi phâêû\\ \gl{subr} jump towards-up-\gl{deict} rat dry.land\\ \glt \qu{so that Rat [can] jump up back onto dry land.}

\z
\ea%<Hc2-29>

%\langinfo{}{}{} go cahma i ibwen a ja hupwa a konja hupwale can we
\gll Go cahma i=ibwen a=ja hut a=koon ja hup-wa-le ca-n we\\ \gl{2}\gl{sg} \gl{top} \gl{def}.\gl{sg}=squid \gl{3}\gl{sg}=\gl{prf} move.down \gl{3}\gl{sg}=\gl{prog} \gl{prf} go.down-\gl{rep}-\gl{dir.cf} in-\gl{nspec} water\\ \glt \qu{And the Squid, he goes back down, he is going back away into the water.}
\z
\pagebreak
\ea%<Hc2-30>

%\langinfo{}{}{} a cuut mwa ka siibwi a emwadia a emwadia a mwadike mwa i ape xhwata can bwan
\gll A=cuut mwa ka siibwi mwa a=e-mwadi-a a=e-mwadi-a a=mwadi-ke i=ape-xhwata ca-n bwa-n\\ \gl{3}\gl{sg}=stand \gl{deict} \gl{sbj} rat \gl{deict} 3\gl{sg}=\gl{refl}-laugh-3\gl{sg}.\gl{obj} 3\gl{sg}=\gl{refl}-laugh-3\gl{sg}.\gl{obj} 3\gl{sg}=-laugh-\gl{tr} \gl{def}.\gl{sg}=\gl{loc}.\gl{nmlz}-bald in-\gl{nspec} head-3\gl{sg}.\gl{poss}\\ \glt \qu{Rat stands now and laughs to himself, laughs about the bald spot on his head.}

\z
\ea%<Hc2-31>

%\langinfo{}{}{} ibwan ibwen a xhwata
\gll I=bwa-n ibwen a= xhwata\\ 
\gl{def}.\gl{sg}=head-\gl{nspec} squid \gl{rel}= bald\\ 
\glt \qu{The bald squid head.}

\z
\ea%<Hc2-32>

%\langinfo{}{}{} aemwadia kona tena ka i ibwen
\gll A=e-mwadi-a kona a-tena-a ka i=ibwen\\ 3\gl{sg}=\gl{refl}-laugh-3\gl{sg}.\gl{obj} then hear \gl{sbj} \gl{def}.\gl{sg}=squid\\ \glt \qu{He laughed to himself, then he was heard by the Squid.}

\z
\ea%<Hc2-33>

%\langinfo{}{}{} ha go go mwadike i da?
\gll Ha go go=mwadi-ke i=da\\ 
\gl{expl} \gl{2}\gl{sg} \gl{2}\gl{sg}=laugh-\gl{tr} \gl{def}.\gl{sg}=what\\ 
\glt \qu{``Hey you! What are you laughing about?"}

\z
\ea%<Hc2-34>

%\langinfo{}{}{} e=mwadike xhwatakam
\gll E=mwadi-ke xhwata ka-m\\ 
\gl{1}\gl{sg}=laugh-\gl{tr} bald \gl{clf}.\gl{poss}-\gl{2}\gl{sg}.\gl{poss}\\ 
\glt \qu{``I laugh of your baldness."}
\z
\ea%<Hc2-35>

%\langinfo{}{}{} go go thuang-o ta-me na yo efetame-ko can! 
\gll Go go=thuang-o ta-me na yo e=fe-ta-me-ko can, \\ \gl{2}\gl{sg} \gl{2}\gl{sg}=joke-\gl{1}\gl{sg}.\gl{obj} move.up-\gl{dir.cp} \gl{dem} \gl{1}\gl{sg} 1\gl{sg}=take-move.up-\gl{dir.cp}-2\gl{sg}.\gl{obj} in\\ \glt \qu{``You, you laugh of me walking up [to dry land], it is me who brought you up here!"}

\z

\ea

%\langinfo{}{}{} a thake nyadame i thaangan kona thaakea
\gll A=thake nya-da-me i=thaangan koo-n a=thake-a\\  \gl{3}\gl{sg}=throw send-go.up-\gl{dir.cp} \gl{def}.\gl{sg}=tentacle on-3\gl{sg} \gl{3}\gl{sg}=throw-\gl{3}\gl{sg}.\gl{obj}\\ \glt \qu{Then he threw his tentacle and then hit him.}
\z
\pagebreak

\ea%<Hc2-36>
%\langinfo{}{}{} thakea thakea hut can we go a yabilo-a
\gll Thake-a thake-a hut ca-n we go a=yabilo-a\\ throw-\gl{3}\gl{sg}.\gl{obj} throw-\gl{3}\gl{sg}.\gl{obj} move.down in water then \gl{3}\gl{sg}=kill-3\gl{sg}.\gl{obj}\\ 
\glt \qu{Threw him, threw him down into the water, then he killed him.}
\z
\ea%<Hc2-37>

%\langinfo{}{}{} go ethuago e=nyauko can we ma bup nyanaam 
\gll Go=e-thuang-o e=nya-ut-ko ca-n we ma bup hnyanaa-m\\ \gl{2}\gl{sg}=\gl{refl}-joke-\gl{1}\gl{sg}.\gl{obj} \gl{1}\gl{sg}=put-go.down-2\gl{sg}.\gl{obl} in-\gl{nspec} water \gl{subr} thunder.strike breath-\gl{2}\gl{sg}.\gl{poss}\\ \glt \qu{``You laugh about me, I'll drag you down into the water to burst your breath."}

\z
\ea%<Hc2-38>

%\langinfo{}{}{} go aja vwa mwa mekai siibwi
\gll Go a=ja vwa ma me ka=i=siibwi\\ 
then \gl{3}\gl{sg}=\gl{prf} do \gl{subr} die \gl{sbj}=\gl{def}.\gl{sg}=rat\\ 
\glt \qu{And he made the Rat die.}

\z
\ea%<Hc2-39>

%\langinfo{}{}{} cahma ibwen thaja hupwa
\gll Cahma ibwen tha ja hup-wa\\ \gl{top} squid \gl{ass} \gl{prf} go.down-\gl{rep}\\ \glt \qu{And the Squid, it went back down.}

\z
\ea%<Hc2-40>

%\langinfo{}{}{} thaja koin mwa siibwi thapa mea thapaja seada
\gll Tha ja koin mwa siibwi tha pa me-a tha=a=pa ja sea-da\\ \gl{ass} \gl{prf} stop \gl{rep} rat \gl{ass} already die-\gl{3}\gl{sg}.\gl{obj} \gl{ass}=3\gl{sg}=\gl{pfv} \gl{prf} look-go.up\\ \glt \qu{This is the end now, Rat is dead, he looks upwards now.}

\z


