\documentclass{article}
\title{The 1917 Tipije War - English translation}
\usepackage{glossingtool}
\usepackage{todonotes}
%\label{Appendixme} % Change X to a consecutive letter; for referencing this appendix elsewhere, use \ref{AppendixX}
\begin{document}
\maketitle


Question (asked by Nigai, ~50):


In \ref{ex:koin_while}, “while" would work better if it were \textit{koon} \todo{check}

\pex \label{ques}
\a \label{ex:koin_while}
\begingl
\gla a=kon vi hapi na gasu <enregistrer> koin gase=bo pala
// \glb 3\gl{sg}=\gl{koon} say that \gl{na} 1.\gl{du}.\gl{incl} record finish 1\gl{pl}.\gl{incl}=\gl{bo} say// \glft  \qu{He is saying that we will record while we will be speaking}
//\endgl
\a
\begingl
\gla hêê da cama nyima-m ma go <enregistrer> ce-ma go=bo pala ma bo fainake koin cama nyima-m ma go xaahni ma i=evwakan nyakoom.
// \glb yes what if/\gl{irr} will-2.\gl{sg}.\gl{poss} \gl{subr} 2.\gl{sg} record if/\gl{irr} 2.\gl{sg}=\gl{bo} speak \gl{subr} \gl{bo} show finish if will-2.\gl{sg}.\gl{poss} \gl{subr} 2.\gl{sg} keep with \gl{def}.\gl{sg}=\gl{nmlz}.\gl{ins}-do-\gl{nmlz} for-2.\gl{sg}.\gl{poss}
// \glft \qu{Yes, um, if you want to record when you will speak in order to show after(?) if you want to preserve it with the tool for yourself}
// \endgl

\a
\begingl
\gla ma xaani nyakoo... ya a=vi canbwen
// \glb \gl{subr} keep for 3.\gl{sg} 3.\gl{sg}=say yesterday
// \glft \qu{To keep it for... he said it yesterday}
// \endgl

\a
\begingl
\gla evwakan nyako li=mwan somun gase bo bwaa pala koon cama go ca aman thapoke fwajimwake
// \glb tool for \gl{def}.\gl{pl}=house lesson 1.\gl{pl}.\gl{incl} \gl{bo} \gl{bwa} speak \gl{obl}-3.\gl{sg} if 2.\gl{sg}/well in? something begin ask
// \glft \qu{Tool for the schools, we will be speaking about it if, well/you something begin asking}
// \endgl

\a
\begingl
\gla enyima-n ma caihnan i=hun-moo-kan cala le=fe ka li=popwaale ce veandan vwa vaa. hê ko i=va. 
// \glb ?-will-3.\gl{sg}.\gl{poss} \gl{subr} know \gl{def}.\gl{sg}-\gl{nmlz}-stay-\gl{nmlz} when.\gl{real} 3.\gl{pl}=fetch \gl{agt} \gl{def}.\gl{pl}=European in time \gl{exist} war yes about \gl{def}.\gl{sg}=war
// \glft \qu{He wants to know the way of being [what happened] when the Europeans took [conquered?], at the time of the war. Yes, about the war}
// \endgl

\xe

Story (Dego ~80):

\pex
\a
\begingl
\gla kona li=pupwaale tha le=pa fe cahni ko-nya seen hnya hnute
// \glb because? \gl{def}.\gl{pl}=Europeans \gl{ass} 3.\gl{pl}=\gl{pa} take here until border over downstream
// \glft \qu{Thing is? the Europeans had taken from here to the border over there}
// \endgl

\a
\begingl
\gla le=hmasa-me ka li=pupwaale kon cahni li=apuli li=xhaomû moo sen xahnut can-la-la li=xhwaapwê
// \glb 3.\gl{pl}=arrive-\gl{me} \gl{agt} \gl{def}.\gl{pl}-Europeans ? here \gl{def}.\gl{pl}=person \gl{def}.\gl{pl}=elder stay border downstream in?-\gl{prox}-\gl{prox} \gl{def}.\gl{pl}=colonnary.pine
// \glft \qu{The Europeans arrived and here the people, the elders stayed at the border down there at the \textit{araucaria} pines} 
// \endgl

\a
\begingl
\gla kone le=hma wati-le nya cahni hma xaa-le thapoke le cahni na tha i=seen nyaande
// \glb when? 3.\gl{pl}=arrive chase-3.\gl{pl} until? here arrive hit-3.\gl{pl} begin 2.\gl{pl} here \gl{na} \gl{ass} \gl{def}.\gl{sg}=order ?
// \glft \qu{When they arrived chasing them until here, arrived striking them beginning here until the border over there}// 
\endgl

\a
\begingl
\gla li=apuli li=xhaohmu kona li=pupwaale kona tha i=seen hnya hnut-te
// \glb \gl{def}.\gl{pl}=person \gl{def}.\gl{pl}=elder ? \gl{def}.\gl{pl}=European = \gl{ass} \gl{def}.\gl{sg}=border towards? downstream-\gl{te}
// \glft \qu{The people, the elders ? the Europeans ? the border down there}
// \endgl

\a
\begingl
\gla koi le=hma wati-le ma xaa-le hnuuda-me cahni
// \glb \gl{obl} 3.\gl{pl} arrive chase-3.\gl{pl} \gl{subr} strike-3.\gl{pl} upstream-\gl{me} here
// \glft \qu{And they arrived chasing them to strike them upstream of here}
// \endgl

\a
\begingl
\gla koin le=bwa hân hân cop xada koin le hup-wa xahan xahut le seejanit koin le=bwa-da ha thê
// \glb finish 3.\gl{pl}=\gl{bwa} go go go.over up.there finish 3.\gl{pl} go.down-\gl{mwa} over.there down.there 3.\gl{pl} T. finish 3.\gl{pl}=\gl{bwa}-? go? rock
// \glft \qu{They walked [from there (French version)], walked, went up over there, then they came down again over there, they were in Tiendanit. Then they ? rock} 
// \endgl

\a
\begingl
\gla bwa xada thathee mwa maahma ca i=dingan xahan thêêdo
// \glb \gl{bwa} up.there kill \gl{mwa}? chief in \gl{def}.\gl{sg}=creek over.there T.
// \glft \qu{They killed the chief over there in the creek at Tendo.}
// \endgl

\a
\begingl
\gla go le=fwade-a nyamwa-la koi ya mwa si-n i=mwa-n hmat ko i=vaa a=hnyâda-me ka thêamâ
// \glb then 3.\gl{pl}=seek.\gl{anim} ?-\gl{prox} because 3\gl{sg} container hand-3.\gl{sg}\gl{poss} \gl{def}.\gl{sg}=container-\gl{poss} customary.money for \gl{sg}.\gl{def}=war 3-\gl{sg}=send-\gl{me} \gl{agt} chief
// \glft  \qu{Puis ils le cherchaient en ce moment-là. Ce'était lui, il avait la monnaie kanak dans sa main, envoyée par le grand chef}
// \endgl

\a
\begingl
\gla 
i=se mwâ-n solda a=môô xahnut pwanvai ka i=se a=môô xahut dewanwe
// \glb \gl{def}.\gl{sg}=other container-\gl{poss} soldier 3.\gl{sg}=stay downstream.there P. \gl{cntr} \gl{def}.\gl{sg}=other 3.\gl{sg}=stay down.there T.
// \glft \qu{Barracks was down at Pwanvai and another in Tiouandé}
// \endgl

\a
\begingl
\gla 
cipa gase=va caihnan wa nyala hun-moo-kan vaa
// \glb \gl{neg} 1.\gl{pl}.\gl{incl}=strongly know from.there ? \gl{nmlz}-stay-\gl{nmlz} war?
// \glft \qu{We do not know well ? nature of fact} \\ \qu{We do not know well the cause of this fact (direct translation of the French version)}
// \endgl

\a
\begingl
\gla kona a=nyadaa-me ka bwaaxat tha i=hmat-aen 
// \glb because 3.\gl{sg}=send-\gl{me} \gl{agt} B. \gl{ass} \gl{def}.\gl{sg}=customary.money-\gl{dem}
// \glft \qu{This money was sent by Bwaxat}
// \endgl

\a
\begingl
\gla go le=ja thathe-a mwâ go le=nya siwa-mwâ 
// \glb then 3.\gl{pl}=\gl{ja} kill-3.\gl{pl} \gl{mwa} then 3.\gl{pl}=? return-\gl{mwa}
// \glft \qu{And they killed him, then went back}
// \endgl

\a
\begingl
\gla 
cama hmamaca-me ka i=mwâ-n hmat ka lu=tua lu=mae a lu ma
// \glb when.\gl{irr} arrive-\gl{me} \gl{agt} \gl{def}.\gl{sg}=container-\gl{poss} customary.money \gl{cntr} 3.\gl{du}=take.out 3.\gl{du}=? \gl{rel}? 3.\gl{du} \gl{subr}?
// \glft \qu{When the war money arrived, they unwrapped it ?}
// \endgl

\a
\begingl
\gla a=vi maman Henri ma maama-le-mwa lu=tua a=fe i=dipi ka i=see koin a=fe sanan ka i=see 
// \glb 3.\gl{sg}=say mother H. \gl{com} mother-3.\gl{pl}\gl{poss}-\gl{mwa} 3.\gl{du}=take.out 3.\gl{sg}=take \gl{def}.\gl{sg}=cover \gl{agt} \gl{def}.\gl{sg}=other finish 3.\gl{sg}=take \gl{def}.\gl{sg}=content \gl{agt} \gl{def}.\gl{sg}=other
// \glft  \qu{That is, Henri's mother and the mother of the others, the two unwrapped it and one took the cover, while the other took the content}
// \endgl

\a
\begingl
\gla 
ma le=han ma ja han thathe-a-mwa kon le=ja siwa-mwa 
// \glb \gl{subr} 3.\gl{pl}=go \gl{subr} \gl{ja} go kill-3.\gl{sg}-\gl{mwa} ? 3.\gl{pl}-\gl{ja} return-\gl{mwa}
// \glft \qu{When they went and he had already been killed, they went back (direct translation from French version)}
// \endgl

\a
\begingl
\gla ma le=ja hmaca-mwa Noumea le=ja fee-le-mwa jevwan-mwa acan ka li=xhaomu mwa le=môô cahni 
// \glb \gl{irr} 3.\gl{pl}=\gl{ja} arrive-\gl{mwa} N. 3.\gl{pl}=\gl{ja} take-3.\gl{pl}-\gl{mwa} several among \gl{agt} \gl{def}.\gl{pl}=elder \gl{mwa} 3.\gl{pl}=stay here
// \glft \qu{When they arrived in Noumea, they took them [prisoners], several of which were people from here}\\ \qu{When they arrived in Noumea, the prisoners, among the elders who stayed here (direct translation from the French version}
// \endgl

\a
\begingl
\gla acan ka i=hao-nea tha u vwa mwa li=been
// \glb among \gl{agt} \gl{def}.\gl{sg}=grandfather-3.\gl{sg}.\gl{poss} \gl{ass} ? \gl{exist} \gl{mwa} \gl{def}.\gl{pl}=brothers
// \glft \qu{Among them was his[man present in the audience] grand-father, and others also}
// \endgl

\a
\begingl
\gla na cahni tha xhwan see-a a thathe-a cahni a thathe-a na i=bee i=papa-n ena xahnuut pwanbaut mwa
// \glb \gl{na} here \gl{ass} a.bit be.one-3.\gl{sg} \gl{rel} kill-3.\gl{sg} here \gl{rel} kill-3.\gl{sg} \gl{na} \gl{def}.\gl{sg}=brother \gl{def}.\gl{sg}=father-\gl{poss} \gl{dist} downstream P. \gl{mwa}
// \glft \qu{Here, there was only a single who was killed here, the one who was killed was the brother of the father of those downstream at Pwanbaut.}
// \endgl

\a
\begingl
\gla kavi i=bee-n i=papa-nle yatan bwaungan
// \glb but \gl{def}.\gl{sg}=brother-\gl{poss} \gl{def}.\gl{sg}=father-3.\gl{pl}\gl{poss} name B.
// \glft \qu{But the brother of their father was called Bwaungan.}
// \endgl


\a
\begingl
\gla thathe-a nya xahan waneut ma le=siwa-mwa ka le=ja fwadae-mwa ka li=pupwaale a=ja ha thaoot ka bwaaxat-ae vwa koo-n ka bwaaxat li=apuli-nea
// \glb kill-3.\gl{sg} \gl{prox} over.there O.-K. \gl{irr}? 3.\gl{pl}-return-\gl{mwa} \gl{cntr} 3.\gl{pl}=\gl{ja} seek.\gl{inan}-\gl{mwa} \gl{agt} \gl{def}.\gl{pl}=European 3.\gl{sg}=\gl{ja} ? come.out \gl{agt} B.-\gl{dem} do \gl{obl}-3.\gl{sg} \gl{agt} B. \gl{def}.\gl{pl}=man-3.\gl{sg}.\gl{poss}
// \glft \qu{They killed him way over there in Wen Kuc as? they returned and/because the Europeans were/had been? investigating, and it had emerged that this Bwaaxat, it was done by Bwaaxat, his men}
// \endgl

\xe


Death of Bwaxat

\pex
\a
\begingl
\gla fe-ta ma le=up-e thai-le ma le=ta mwa-n suki aman
// \glb take-go.up \gl{subr} 3.\gl{pl}=go.down-\gl{me}? attach-3.\gl{pl} \gl{subr} 3.\gl{pl}=go.up house-\gl{poss} pay something
// \glft \qu{They came down north to gather them and put them in prison}
// \endgl

\a
\begingl
\gla hê fwandai ca wuukin hapi na wukin Bwaxat xhu-pwa-mee i=wang le=bwa vii hapi na ya ya thapoke i=vaa
// \glb yes seek.\gl{inan} some reason that \gl{na} reason B. go.down-\gl{mwa}-\gl{me} \gl{def}.\gl{sg}=boat 3.\gl{pl}=\gl{bwa} say that \gl{na} 3.\gl{sg} 3.\gl{sg} begin \gl{def}.\gl{sg}=war
// \glft \qu{Yes, looking for some reason, that the reason was Bwaxat. The boat came up north. They were saying that it was him, he started the war.}
// \endgl

\a
\begingl
\gla a=ja hu-pwa-mee umwang hu-pwa-me fe-a ce wandan thangavi nyôô-n apuli
// \glb 3.\gl{sg}=\gl{ja} go.down-\gl{mwa}-\gl{me} boat go.down-\gl{mwa}-\gl{me} take-3.\gl{sg} in time chop neck-\gl{poss} person
// \glft \qu{When the boat went down north, went down to catch him, at this time they chopped off people's necks}
// \endgl

\a
\begingl
\gla ce wadan a=hma-cuut ka i=umwang ka le=fe-ta-a koma le=fe-a ka i=wang a=han-me hma-cuut ya a=ta wainyô nyece i=juu mwâ
// \glb in time 3.\gl{sg}=arrive-stand \gl{agt} \gl{def}.\gl{sg}=ship \gl{cntr} 3.\gl{pl}=take-go.up-3.\gl{sg} because? 3.\gl{pl}=take-3.\gl{sg} \gl{cntr} \gl{def}.\gl{pl}=boat 3.\gl{sg}-go-\gl{me} arrive-stand 3.\gl{sg} 3.\gl{sg}-go.up tie.neck in(building) \gl{def}.\gl{sg}=real house
// \glft \qu{At the moment when they boat arrived and they went up to take him [in the mountains], take him, and the boat had come, he had hanged himself in his house.}
// \endgl

\a
\begingl
\gla ma le=ta ma le=xale-a ko ma le=fe-a ma le=ta ma le=fe-a ya a=pa ja me-a ma le=hmasa-mwa xada ma le=ja fwandai ta-mwa
// \glb \gl{com} 3.\gl{pl}=go.up \gl{com}? 3.\gl{pl}=see-3.\gl{sg} because \gl{subr} 3.\gl{pl}=take-3.\gl{sg} \gl{com} 3.\gl{pl}=go.up \gl{subr} 3.\gl{pl}=take-3.\gl{sg} 3.\gl{sg} 3.\gl{sg}=\gl{pa} \gl{ja} die-3.\gl{sg} xx 3.\gl{pl}=arrive-\gl{mwa} up.there xx 3.\gl{pl}=\gl{ja} seek.\gl{anim} go.up-\gl{mwa}
// \glft  \qu{When they went up to see him because they wanted to take him, went up to take him, he had already died when they arrived up there to look for him.}
// \endgl

\a
\begingl
\gla 
go a=ja ta-mwa ka i=thamo i=mama-n maama ma
// \glb then 3.\gl{sg}=\gl{ja} go.up-\gl{mwa} \gl{agt} \gl{def}.\gl{sg}=woman \gl{def}.\gl{sg}=mother-\gl{poss} chief\footnote{lit. \qu{elder brother}} \gl{subr}
// \glft \qu{Then came up the woman, the mother of the chief, to...}
// \endgl

\a
\begingl
\gla 
a=fainake i=mwâ-n hmat a=nya ka ena na ehni xaaci civije
// \glb 3.\gl{sg}=show \gl{def}.\gl{sg}=container-\gl{poss} customary.money 3.\gl{sg}=give \gl{agt} \gl{dist} \gl{na} \gl{prox} strike T.
// \glft \qu{...show the war money which that one [war chief Bwaaxat] had sent the other [her son] to strike Tipije.}
// \endgl

\a
\begingl
\gla 
le=ja vatipwe-mwa kon ya tha pa ja me-a go tha le=ja thuat-mwâ moo ka li=xhaomu ja tha le=ja ha-mwa-me
// \glb 3.\gl{pl}=ja drop-\gl{mwa} because 3.\gl{sg} \gl{ass} \gl{pa} \gl{ja} die-3.\gl{sg} then \gl{ass} 3.\gl{pl}=\gl{ja} emerge-\gl{mwa} stay \gl{agt} \gl{def}.\gl{pl}=elder \gl{ja} \gl{ass} 3.\gl{pl}=\gl{ja} go-\gl{mwa}-\gl{me}
// \glft  \qu{They were released because he [Bwaaxat] had already died and the ancestors came out [of prison] and they went back [home].}
// \endgl

\a
\begingl
\gla 
go cama li=xayu le=thai-le hnya-da-le wailu
// \glb and if\footnote{concerning, “quant aux"} \gl{def}.\gl{pl}=boy 3.\gl{pl}=catch?-3.\gl{pl} send-up?-3.\gl{pl} H.
// \glft \qu{And concerning the men, they tied/gathered them and sent them to Houaïlou}
// \endgl

\a
\begingl
\gla 
can goakan-ae le=hma-cu-pwa-me ka li=thamo ma le=hup-waa-me ka li=thamo le=môô wailu
// \glb in moment-\gl{dem} 3.\gl{pl}=arrive-stand-\gl{mwa}-\gl{me} \gl{agt} \gl{def}.\gl{pl}=woman \gl{com} 3.\gl{pl}=go.down-\gl{mwa}-\gl{me} \gl{agt} \gl{def}.\gl{pl}=woman 3.\gl{pl}=stay H.
// \glft \qu{At this moment the women arrived there so that they would go back, they stayed in Wailu}
// \endgl

\a
\begingl
\gla a=vi ka Leenhardt hapi i=apuli a=vwa ka i=thamo go ya a=fa-thuati i=apuli
// \glb 3.\gl{sg}=say \gl{agt} L. that \gl{def}.\gl{sg}=man 3.\gl{sg}\textsubscript{1}=do \gl{agt} \gl{def}.\gl{sg}=woman\textsubscript{1} then 3.\gl{sg} 3.\gl{sg}=\gl{caus}-come.out \gl{def}.\gl{sg}=man
// \glft \qu{Leenhardt said that the man was made by the woman and he made the man come out [of prison in Houaïlou]}
// \endgl

\a
\begingl
\gla 
nyasi Leenhardt e=bwa ju tena i=yata-n cipe xa-xale-a
// \glb for L. 1.\gl{sg}=\gl{bwa} truly hear \gl{def}.\gl{sg}=name-3.\gl{sg}.\gl{poss} \gl{neg}.1.\gl{sg} \gl{hab}-see-3.\gl{sg}
// \glft \qu{Concerning Leenhardt, I have well heard his name but never seen him}
// \endgl
\xe
 
\glossingAbbrevsList

\end{document}
