\chapter{The Story of the Dance Beat (Yvonne Sahilé)}

This is a legend written by Yvonne Sahilé, a writer and culture activist living in Tiouandé. She drew on traditional motifs, e.g. the Cannibal of Tipije, the knowledge transmitted by bloodlines (the dance is recognized by the father), and everyday details. Mrs. Sahilé wrote this story in French, and it was translated into Vamale by André Kalen, Jacob Oué and Christophe Pei. Her names for the protagonists are in Pije, and were kept that way in the translations.

\label{text:beat}
\ea %\label{ex:t1}

%\langinfo{}{}{} 1
\gll 	Habu Can-Vije vwa i=apemoo a=pwa-n jela i=jahoot a=xhopwen.	\\	long.ago Tipije \gl{exist} \gl{def}.\gl{sg}=tribe \gl{rel}=on-\gl{nspec} side \gl{def}.\gl{sg}=river \gl{rel}=big	\\ \glt  \qu{Long ago in the Tipije valley, there was a tribe on the bank of this great river.} (following the French original text)		


\z \ea %\label{ex:t2}
%\langinfo{}{}{} 2
\gll 	Ca i=apemoo-ca le=vacuti \textbf{ca} daahma a=bwa xawe ka yata-n Thêa Xa-vila.	\\	in \gl{def}.\gl{sg}=tribe-\gl{prox} 3\gl{pl}=erect some chief \gl{rel}=\gl{ipfv} young and name-3\gl{sg}.\gl{poss} T. \gl{nmlz}-dance	\\ 
\glt  \qu{In this tribe they erected some chief who was still young, and his name was Firstborn the Dancer.}		
\z 

\ea %\label{ex:t3}
%\langinfo{}{}{} 3
\gll Ca \textbf{i}=apemoo vwa \textbf{li}=xawe thamo.
\\ in \gl{def}.\gl{sg}=tribe \gl{exist} \gl{def}.\gl{pl}=young woman	
\\ \glt \qu{In the tribe were young women.} (indefinite, non-specific in the French original)
\z 

\ea %\label{ex:t4-6}
%\langinfo{}{}{} 4,5,6
\gll Tha fe nyamaa-n can dawee-le \textbf{eca} thamo a=en maa-n.\\ 
     \gl{ass}.3\gl{sg} take eye-3\gl{sg}.\gl{poss} in between-3\gl{pl}.\gl{poss} \textbf{some} woman \gl{rel}=first face-3\gl{sg}.\gl{poss}\\ 
\glt \qu{The most beautiful woman between them ``took his eye".}
\z

\ea
\gll Ka a=xhani ma mwada-n. Yata-n In Fwe.\\ 
and 3\gl{sg}=choose \gl{subr} wife-3\gl{sg}.\gl{poss} name-3\gl{sg}.\gl{poss} skin guettarda.speciosa\\ 
\glt \qu{And he chose her as his wife. Her name was Figtree Bark.}
\z

\ea
({Note the \textit{i a yatan} construction, \qu{she who name-her}})\\
\gll Le=kiica ka meeka \textbf{li-been} thamo, ma can e-dawee-le \textbf{i}=a=yata-n I-n Thu.\\ 
3\gl{pl}=jealous and all \gl{def}.\gl{pl}-peer woman \gl{com} in.\gl{nspec} \gl{refl}-between-3\gl{pl} \gl{def}.\gl{sg}=\gl{rel}= name-3\gl{sg}.\gl{poss} skin banyan\\ 
\glt \qu{All the other women were jealous, with among them all the one who was called Banyan Bark.}
\z

\ea
\gll A=vwa hmwakan cama bee-n.\\ 
3\gl{sg}=do like \gl{subr} peer-3\gl{sg}.\gl{poss}\\ 
\glt  \qu{She pretends to be her friend.}
\z

\ea
\gll Ena a=mu saxhut cai-n ca li=ape-hân-ea, koma nya i=ya-n ma udoo-n, na i=a=vi nya-thuan nyasi-n.\\ 
\gl{dem} \gl{rel}=\gl{freq} follow behind-\gl{ana} in \gl{def}.\gl{pl}=\gl{loc}.\gl{nmlz}-go-3\gl{sg}.\gl{poss} \gl{subr} give \gl{def}.\gl{sg}=starchy.food-3\gl{sg}.\gl{poss} \gl{com} cold.drink-3\gl{sg}.\gl{poss} \gl{dem} \gl{def}.\gl{sg}=\gl{rel}=say put-do.well for-3\gl{sg}\\ 
\glt \qu{The one who follows her everywhere, waits on her with food and water, her confidante as well.}
\z 

\ea
\gll Ca i=se thoatit In Fwe a=vi nyako In Thu hapi ya xhaapwe-jii-a, kona a=ja nya-bee i=kiica-a koo-n.\\ 
in \gl{def}.\gl{sg}=other day skin figtree 3\gl{sg}=say to skin banyan \gl{comp} 3\gl{sg} fruit-belly-3\gl{sg} \gl{cnj} 3\gl{sg}=\gl{prf} give-peer \gl{def}.\gl{sg}=jealous-3\gl{sg} on-\gl{ana}\\ 
\glt \qu{One day, Ciin Fwe announces to Ciin Thilic that she is pregnant, which adds to the latter's jealousy.}
\z

\ea
\gll Jahnga li=phwê pwecake-n a=mu sibu i=xhaapwe jia In Fwe, ka a=mu vwa wanyima In Thu.\\ 
all.along \gl{def}.\gl{pl}=month after-\gl{ana} 3\gl{sg}=\gl{iter} \gl{def}.\gl{sg}=fruit swell belly skin figtree \gl{cnj} 3\gl{sg}=\gl{iter} do angry skin banyan\\ 
\glt  \qu{During the months after, Ciin Fwe's belly grows, which makes Ciin Thilic angry.}
\z

\ea
\gll Fine thoatit pwecake goon-xat In Thu xa-hân ma In Fwe a=hut thuup can jaahoot, ka a=naami i=jia i=se.\\ 
every day after enough-sun skin banyan \gl{nmlz}-go \gl{com} skin figtree 3\gl{sg}=go.down bathe in river \gl{cnj} 3\gl{sg}=massage \gl{def}.\gl{sg}=belly \gl{def}.\gl{sg}=other\\ 
\glt \qu{Every day in the afternoon Ciin Thilic goes down with Ciin Fwe to bathe in the river, and massage her belly.}
\z

\section{The Flood}

\ea
\gll Ehni a=ja hma-sa-me ko e-nim na-bwa fava ka-n phwêê-n.\\ \gl{dem} 3\gl{sg}=\gl{prf} arrive-go-\gl{dir.cp} on \gl{ord}-five and four \gl{clf}.\gl{poss}-\gl{nspec} month-\gl{ana}\\ \glt  \qu{There arrived the ninth month.}
\z

\ea
\gll Cama a=xaleke pwa-n bwan can un jaahoot a=xaleke hmape-thoatit a=fiing a=kon nyasipoke nya-pwan yee-n bwan.\\ when 3\gl{sg}=look on-\gl{nspec} mountain in bottom valley 3\gl{sg}=see flesh-sky \gl{rel}=dark 3\gl{sg}=\gl{prog} gather put-on tree-\gl{link} mountain
\\ \glt \qu{Looking at the mountains upriver, she saw dark clouds gathering on the summits.}
\z

\ea
\gll a=vwa yathôa ma a=vi nyako In Fwe.\\ 3\gl{sg}=do be.hurried \gl{subr} 3\gl{sg}=say to skin figtree\\ \glt \qu{She hurried to tell Ciin Fwe:}
\z

\ea
\gll ``In Fwe goon ma go=hut takee-ko ca-n jahoot ko xhwat-thuang ma go=tipwa xawe."\\ skin figtree enough \gl{subr} 2\gl{sg}=go.down stretch-2\gl{sg}.\gl{obj} in-\gl{nspec} river because almost \gl{subr} 2\gl{sg}=drop young\\ \glt  \qu{``Ciin Fwe you must go relax in the river, because you will soon give birth."} 
\z 

\ea
\gll Cahma In Fwe cipa a=e-vwa-nyakoo-n, a=vwa ma yathôa hut cai-n ca-n jahoot.\\ 
     whereas skin figtree \gl{neg} 3\gl{sg}=\gl{refl}-\gl{exist}-for-3\gl{sg} 3\gl{sg}=do \gl{subr} be.hurried go.down behind-\gl{ana} in-\gl{nspec} river\\ 
\glt \qu{But Ciin Fwe did not suspect anything (in the original version: did not know of jealousy) and hurriedly followed her down to the river.}
\z 

\ea
\gll Cama lu=hma-cut pwa-n jelan we a=vi nyako In Fwe ``Hut ca-n we."\\ when 3\gl{du}=arrive-go.down on-\gl{nspec} side-\gl{link} water 3\gl{sg}=say to skin figtree go.down in-\gl{nspec} water\\ 
\glt \qu{When the two arrived on the riverbank she told Ciin Fwe ``Go into the water."}
\z

\ea
\gll In Fwe a=hut can we seen we ko li=bucila xha-n.\\ skin figtree 3\gl{sg}=go.down in water border water on \gl{def}.\gl{pl}=main.joint leg-3\gl{sg}.\gl{poss}\\ \glt  \qu{Ciin Fwe goes into the water until her knees.}
\z

\ea
\gll A=vi mwa ``Xhose hut."\\ 3\gl{sg}=say again again go.down\\ \glt  \qu{She tells her ``Go in even further."}
\z

\ea
\gll In Fwe a=hut seen we ko i=goo-n.\\ skin figtree 3\gl{sg}=go.down border water on \gl{def}.\gl{sg}=waist-3\gl{sg}.\gl{poss}\\ \glt  \qu{Ciin Fwe goes into the water until her waist.}
\z

\ea
\gll In Thu a=vi ``xhwan vaa hup-wa, ko-ma naami i=jia-m ma i=duu-n we."\\ skin banyan 3\gl{sg}=say bit much go.down-\gl{rep} because-\gl{subr} massage \gl{def}.\gl{sg}=belly-2\gl{sg}.\gl{poss} \gl{com} \gl{def}.\gl{sg}=bone-\gl{link} water\\ \glt \qu{Ciin Thilic says ``A bit more, because the current needs to massage your belly".}
\z

\ea
\gll In Fwe vaa hup-wa seen we ko-n bwaangan, vaa hân ca-n nya-n we.\\ skin figtree much go.down-\gl{rep} border water on-\gl{nspec} chest much go in-\gl{nspec} inside-\gl{link} water\\ \glt \qu{Ciin Fwe goes into the water until her chest, goes on into the middle of the water.}
\z

\ea
\gll Cama ca i=wadan-aen, xada can un vwa i=wîîn uta ka a=vwa thêên java ca li=si-n dingan a=le=mu hut thaloot ca-n jahoot.\\  when in \gl{def}.\gl{sg}=time-\gl{dem} up.there in-\gl{nspec} bottom \gl{exist} \gl{def}.\gl{sg}=strength rain \gl{cnj} 3\gl{sg}=do flow flood in \gl{def}.\gl{pl}=arm-\gl{link} creek \gl{rel}=3\gl{pl}=\gl{freq} go.down appear in-\gl{nspec} river\\ 
\glt  \qu{But during that time, up there in the background a heavy rain fell and overflowed the little creeks which flow into the river.} 
\z

\ea
\gll In Thu a=caihna-n hapi la ca la a=la ka i=xhoogo, cama xahnang i=thoatit ma cala la li=yee-n bwan xahnuuda ca-n un le=coopwi ka li=hmape-thoatit, tha sinan hapi vwa java.\\ skin banyan 3\gl{sg}=know-\gl{ana} \gl{comp} be.here in be.here 3\gl{sg}=be.here \gl{sbj} \gl{def}.\gl{sg}=home when good \gl{def}.\gl{sg}=sky \gl{subr} when be.here \gl{def}.\gl{pl}=tree-\gl{link} mountain downstream.there in-\gl{nspec} bottom 3\gl{pl}=bury \gl{sbj} \gl{def}.\gl{pl}=flesh-sky \gl{ass} sign \gl{comp} \gl{exist} flood\\ \glt \qu{Ciin Thilic knew that where the tribe was, if the weather is nice and the mountains in the back are covered with clouds, there will be a flood.} 
\z

\ea
\gll A=hmwaan bune can a=vi nyako In Fwe:\\ 3\gl{sg}=laugh steal \gl{subr} 3\gl{sg}=say to skin figtree\\ \glt \qu{While laughing to herself she said to Ciin Fwe:} 
\z

\ea
\gll ``Tha go=caihnan, cipa goon go=hu-pwa, ma ta seen ko-n nyoo-m"\\ \gl{ass} 2\gl{sg}=know \gl{neg} enough 2\gl{sg}=go.down-\gl{rep} \gl{subr} go.up border on-\gl{link} neck-2\gl{sg}.\gl{poss}\\ \glt  \qu{``You know, that's not enough. Come in again, we have to get the water up to your neck."}
\z

\ea
\gll In Fwe a=uhut seen we ko-n nyoo-n cahma i=java a=ja gun hu-pe thaloot.\\ 
     skin figtree 3\gl{sg}=enter.water border water on-\gl{nspec} neck-3\gl{sg}.\gl{poss} whereas \gl{def}.\gl{sg}=flood 3\gl{sg}=\gl{prf} loud.noise go.down-\gl{dir.cp} appear\\ 
\glt \qu{Ciin Fwe was entering the water up to the neck, whereas the flood suddenly crashed down on them.} 
\z

\ea
\gll In Fwe a=ja cie-a ca i=java.\\ skin figtree 3\gl{sg}=\gl{prf} disappear-3\gl{sg} in \gl{def}.\gl{sg}=flood\\ \glt \qu{Ciin Fwe disappeared in the flood.} 
\ex
\gll Xahnang nyima In Thu ma siiwa mwa ca xhoogo ma a=ta vi nyako i=xatoo-n hapi i=see paa cie-a.\\ good heart skin banyan \gl{subr} return \gl{rep} in home \gl{subr} 3\gl{sg}=go.up say for \gl{def}.\gl{sg}=husband-3\gl{sg}.\gl{poss} \gl{comp} \gl{def}.\gl{sg}=other \gl{prf} disappear-3\gl{sg}\\ \glt \qu{Ciin Thilic is delighted to return to the tribe to break the bad news to her husband.} 
\ex
\gll Thêa Xa-vila a=faphâke nyako i=a= vi ka bwa taaduaa jahnga-n li=wadan a=goon.\\ firstborn \gl{nmlz}-dance 3\gl{sg}=believe for \gl{def}.\gl{sg}=\gl{rel}= say \gl{cnj} \gl{ipfv} mourn length-\gl{ana} \gl{def}.\gl{pl}=time \gl{rel}=enough\\ \glt  \qu{Téin Kapila believed what she said and went into mourning for the necessary time.}
\ex
\gll Koin jahnga i=wadan-aen In Thu a=vwa meekan ma e-xahnang-ea nyasi Thêa Xa-vila, koma a=vaadan In Fwe.\\ while length \gl{def}.\gl{sg}=time-\gl{dem} skin banyan 3\gl{sg}=do all \gl{subr} \gl{refl}-good-3\gl{sg} for firstborn \gl{nmlz}-dance \gl{subr} 3\gl{sg}=forget skin figtree\\ \glt  \qu{During all this time Ciin Thilic was doing everything to please Téin Kapila and make him forget Ciin Fwe.}
\ex
\gll Ka a=ja moo ma mwada-n.\\ \gl{cnj} 3\gl{sg}=\gl{prf} stay \gl{subr} wife-3\gl{sg}.\gl{poss}\\ \glt  \qu{And she ends up becoming his wife.}
\z

\section{The Cannibal}\largerpage[2]

\ea
\gll Cipa a=caihnan ko cipa hoot ko mu=si-n bwan, a=mu moo la ka i=apuli a=yata-n Xa-xhwi Apuli ko i=Vai Ta-gulo,\\ \gl{neg} 3\gl{sg}=know \gl{obl} \gl{neg} be.far \gl{obl} \gl{def}.\gl{du}=arm-\gl{link} mountain 3\gl{sg}=\gl{iter} stay be.here \gl{sbj} \gl{def}.\gl{sg}=person \gl{rel}=name-3\gl{sg}.\gl{poss} \gl{agt}.\gl{nmlz}-eat person on \gl{def}.\gl{sg}=rock cut?-umbilical.cord\\ \glt \qu{Without knowing that not far from there after two mountain bells there used to live a man who was a cannibal and called Cauri Kayuk o Paik Tegulo,}
\z

\ea
\gll vwa i=mwa-n-ea patemwano i=jaahoot\\ \gl{exist} \gl{def}.\gl{sg}=house-\gl{link}-3\gl{sg}.\gl{poss} next.to \gl{def}.\gl{sg}=river\\ \glt \qu{who had his house not far from the river.}
\ex
\gll La ca la ka i=jahoot a=mu ta-me nyasipoke nya pwan jelan si-n ye, vukin ye, xhaapwe ye, doon ye ka apuli hman.\\ be.here in be.here \gl{sbj} \gl{def}.\gl{sg}=river 3\gl{sg}=\gl{freq} go.up-\gl{dir.cp} gather put on side arm-\gl{link} tree trunk-\gl{link} tree fruit tree leaf tree \gl{cnj} person also\\ \glt \qu{There where the river brings up on the edge all that it charms: branches, trunks, fruits, and plants or men.}
\ex
\gll Kavi pwa i=jelan we vwa i=yavi a=mu vwa sinan aman nyakoo-n.\\ but on \gl{def}.\gl{sg}=side water \gl{exist} \gl{def}.\gl{sg}=syzygium.malaccense 3\gl{sg}=\gl{freq} do sign thing for-3\gl{sg}.\gl{poss}\\ \glt \qu{But where the edge of this water hole was, there was a mountain apple stalk which served as an alarm.}
\ex
\gll In Fwe a=cavi-a ka i=java a=ja titoot ko i=vuki ye ma vatipwe-a xala i=vuki yavi.\\ skin figtree \gl{rel}=carry.away-3\gl{sg}.\gl{obj} \gl{sbj} \gl{def}.\gl{sg}=flood 3\gl{sg}=\gl{prf} latch.onto \gl{obl} \gl{def}.\gl{sg}=stem tree \gl{subr} drop-3\gl{sg}.\gl{obj} under \gl{def}.\gl{sg}=stem syzygium.malaccense\\ \glt  \qu{Ciin Fwe who had been carried away managed to grab hold of a trunk which deposited it under the mountain apple tree.}
\ex
\gll A=ja fe mwa li=wîî-n ka a=ja xaleke hapi tana li=xhaapwe yavi.\\ 3\gl{sg}=\gl{prf} take \gl{rep} \gl{def}.\gl{pl}=strength-3\gl{sg}.\gl{poss} \gl{cnj} 3\gl{sg}=\gl{prf} see \gl{comp} ripe \gl{def}.\gl{pl}=fruit syzygium.malaccense\\ \glt  \qu{While regaining her strength, she saw that the fruits of the mountain apple were ripe.} 
\z

\ea
\gll A=e-wago ma a=cuut ka a=tabwe thien xhaapwe yavi tana.\\ 3\gl{sg}=\gl{refl}-encourage \gl{subr} 3\gl{sg}=stand \gl{cnj} 3\gl{sg}=pick.fruit three fruit syzygium.malaccense ripe\\ \glt \qu{She got to her feet as best she could and picked up three very ripe fruits.}
\z

\ea
\gll Xa-xhwi Apuli ko i=Vai Tegulo a=kon majit can mwa-n-ea a=tada nyoot ka a=vi :\\ \gl{agt}.\gl{nmlz}-eat person \gl{obl} \gl{def}.\gl{sg}=rock T 3\gl{sg}=\gl{prog} rest in house-\gl{link}-3\gl{sg}.\gl{poss} 3\gl{sg}=be.startled wake.up \gl{cnj} 3\gl{sg}=say\\ \glt  \qu{Kawi Kayuk o Paitegulo who was sleeping in his hut woke up with a start saying:}
\z

\ea
\gll ``Thien li=apuli a=le=xala i=vuki yavi."\\ three \gl{def}.\gl{pl}=person \gl{rel}=3\gl{pl}=under \gl{def}.\gl{sg}=stem syzygium.malaccense\\ \glt \qu{``Three people are at the foot of the mountain apple."}
\z

\ea
\gll ``E=caihnan i=xhe i=wee li=xhapwe yavi ma le=tipwa pu, e=hut xaleke."\\ 1\gl{sg}=know \gl{def}.\gl{sg}=sound \gl{def}.\gl{sg}=water \gl{def}.\gl{pl}=fruit syzygium.malaccense \gl{subr} 3\gl{pl}=fall ground 1\gl{sg}=go.down see\\ \glt  \qu{``I recognize by the noise made by the sap of the three fruits falling to the ground, I will see."}
\z

\ea
\gll Ca i=wadan-aen In Fwe a=hnyame-ke li=thien xhaapwe yavi ka a=ja vwa ma tipwa xawe.\\ in \gl{def}.\gl{sg}=time-\gl{dist} skin figtree 3\gl{sg}=swallow-\gl{tr} \gl{def}.\gl{pl}=three fruit syzygium.malaccense \gl{cnj} 3\gl{sg}=\gl{prf} do \gl{subr} drop young\\ \glt  \qu{During this time Ciin Fwe swallowing the three fruit triggered her childbirth.}
\z

\ea
\gll Cama ja hma-san ca i=goakan ka Xa-whi Apuli ko Vai Tegulo a=hmaako i=thamo ma mu=xhwaawe xa-xahnang maa-lu lu=kon mi-xala-da\\ when \gl{prf} arrive-go in \gl{def}.\gl{sg}=place \gl{sbj} \gl{agt}.\gl{nmlz}-eat person \gl{obl} \gl{def}.\gl{sg}=rock T 3\gl{sg}=find \gl{def}.\gl{sg}=woman \gl{com} \gl{def}.\gl{du}=child \gl{agt}.\gl{nmlz}-bien face-3\gl{du}.\gl{poss} 3\gl{du}=\gl{prog} lying-look-go.up\\ \glt  \qu{Arriving at the place Kawi Kayuk o Paitegulo found a woman with two beautiful babies.}
\z

\ea
\gll A=fwajimwe-a ``na go=kai, thamo, ma go=ha-me muja-ke mu=nyai-m nye-cahni ka go=hma-sa-me tabwe li=xhaapwe yee-n-eong?"\\ 3\gl{sg}=ask-3\gl{sg}.\gl{obj} \gl{dem} 2\gl{sg}=who woman \gl{subr} 2\gl{sg}=go-\gl{dir.cp} vomit-\gl{tr} \gl{def}.\gl{du}=child-2\gl{sg}.\gl{poss} put-here \gl{cnj} 2\gl{sg}=arrive-go-\gl{dir.cp} pick.fruit \gl{def}.\gl{pl}=fruit tree-\gl{link}-1\gl{sg}.\gl{poss}\\ \glt \qu{And he asked her ``Who are you woman to come and vomit your children here and that you to collect my fruit?"}
\z

\ea
\gll A=vi nyakoon can taemwi ca-n si-n mu=nyai-n:\\ 3\gl{sg}=say for-3\gl{sg} while catch in-\gl{nspec} arm-3\gl{sg}.\gl{poss} \gl{def}.\gl{du}=child-3\gl{sg}.\gl{poss}\\ \glt  \qu{Holding her children in her arms she said to him:}
\z

\ea
\gll ``E=bwadut nyakoo-m ko go=apuli a= xhopwen ko cipa goon me tabwe li=xhaapwe yeeca me xaje,\\  
     1\gl{sg}=bow for-3\gl{sg} because 2\gl{sg}=man \gl{rel}= big because \gl{neg} enough \gl{subr}=1\gl{sg} pick.fruit \gl{def}.\gl{pl}=fruit tree-\gl{prox} \gl{subr}=1\gl{sg} eat.juicy\\ 
\glt \qu{``I lower myself before you great man because I am not worthy to collect these fruit and to eat them,}
\z

\ea
\gll kavi vukin, a=cavi-o ka i=java ka e=ta-me vatipwe mu=xawe-ca xala i=yee-n-go.\\ but reason-\gl{ana} 3\gl{sg}=sweep.away-1\gl{sg} \gl{sbj} \gl{def}.\gl{sg}=flood \gl{cnj} 1\gl{sg}=go.up-\gl{dir.cp} drop \gl{def}.\gl{du}=young-\gl{prox} under \gl{def}.\gl{sg}=tree-\gl{link}-2\gl{sg}.\gl{poss}\\ \glt \qu{...but here I am carried away by the flood and that I am come to give birth to these two children at the foot of your tree.}
\z

\ea
\gll Cama xahnang nyasi-m ce-o m=e silaa mu=nyae-ung nya-cahni ma abe=vwa thuan mwa i=a= abe=vwa nyako i=yee-n-go, kona mu=xhwaawe-ca xayu.\\ when good for-2\gl{sg} leave-1\gl{sg}.\gl{obj} \gl{subr}=1\gl{sg} raise \gl{def}.\gl{du}=child-1\gl{sg}.\gl{poss} put-here \gl{subr} 1\gl{pl}.\gl{excl}=do good \gl{rep} \gl{def}.\gl{sg}=\gl{rel}= 1\gl{pl}.\gl{excl}=do for \gl{def}.\gl{sg}=tree-\gl{link}-2\gl{sg}.\gl{poss} \gl{cnj} \gl{def}.\gl{du}=child-\gl{prox} boy\\ 
\glt  \qu{If it's okay with you let me raise my children here and fix what we did to your tree because those two children are boys.}
\z

\ea
\gll Abe=bo  yabwen-go ka mu=nyae-ung lu=bo xa-vwa-vaa-n-go."\\ 1\gl{pl}.\gl{excl}=\gl{irr} subject-2\gl{sg}.\gl{poss} \gl{cnj} \gl{def}.\gl{du}=child-1\gl{sg}.\gl{poss} 3\gl{du}=\gl{irr} \gl{agt}.\gl{nmlz}-do-war-\gl{link}-2\gl{sg}.\gl{poss}\\ \glt \qu{We will serve you and my children will be your warriors.”}
\z\largerpage

\ea
\gll A=fe nyakoo-n ma xhwatin li=fatii In Fwe, Xa-xhwi Apuli ko Vai Thegulu a=thagavi li=buloo mu=bifidu ka coopwi xala i=yavi cela i=vai,\\ 
     3\gl{sg}=take for-3\gl{sg} \gl{subr} small \gl{def}.\gl{pl}=word skin figtree \gl{agt}.\gl{nmlz}-eat person on rock T 3\gl{sg}=cut \gl{def}.\gl{pl}=umbilical.cord \gl{def}.\gl{du}=twin \gl{cnj} bury under \gl{def}.\gl{sg}=syzygium.malaccense next.to \gl{def}.\gl{sg}=rock\\
 \glt  \qu{Softened by Ciin Fwe's words, Kawi Kayuk o Paitegulo cut the umbilical cords of the twins there and buried them under the mountain apple next to a rock, } 
 \z
 
 \ea
 \gll ka bwa i=goakan-aen a=bwa yata-n Vai Tehnang.\\ \gl{cnj} \gl{ipfv} \gl{def}.\gl{sg}=place-\gl{dist} \gl{rel}=\gl{ipfv} name-3\gl{sg}.\gl{poss} rock sharp\\ \glt \qu{...this place will be Pai-tea.}
 \z

\ea
\gll Ka a=fwajimwake nyako i=inya-lu ``Hmwaeke i=yata-m?"\\ 
     \gl{cnj} 3\gl{sg}=ask for \gl{def}.\gl{sg}=mother-3\gl{du}.\gl{poss} how \gl{def}.\gl{sg}=name-2\gl{sg}.\gl{poss}\\ 
     \glt \qu{And he says to the mother ``What's your name?"}
\z

\ea
\gll ka a=sabe nyasi-n ka In Fwe hapi ``In Fwe ka e=hupe nya hnya-da xada" kavi cipa a=vi i=goakan.\\ \gl{cnj} 3\gl{sg}=answer for-3\gl{sg}.\gl{poss} \gl{sbj} skin figtree \gl{comp} skin figtree \gl{cnj} 1\gl{sg}=go.down-\gl{dir.cp} in.direction send-up up.there but \gl{neg} 3\gl{sg}=say \gl{def}.\gl{sg}=place\\ \glt \qu{And she answers him ``Ciin Fwe and I come from a little higher" without specifying the place to him.}
\z

\ea
\gll Ka a=vi ka i=apuli  ``go=bo thovii hmwaeke nyakoo-lu, ma cipa boje-lu."\\ 
     \gl{cnj} 3\gl{sg}=say \gl{sbj} \gl{def}.\gl{sg}=man 2\gl{sg}=\gl{irr} call how for-3\gl{du}.\gl{poss} \gl{subr} \gl{neg} deaf-3\gl{du}\\ 
     \glt  \qu{And the man tells her ``What are you going to give them as a name so that they are not deaf?" }
\z

\ea
\gll Ka a=sabe nyasi-n ``ca=ape-hnyimake-aman-go"\\ \gl{cnj} 3\gl{sg}=answer for-3\gl{sg}.\gl{poss} \gl{indf}.\gl{sg}=\gl{loc}.\gl{nmlz}-think-thing-2\gl{sg}.\gl{poss}\\ \glt \qu{And she answers ``Whatever you think".}
\z

\ea
\gll Kona thovi nyako i=see Ngeein ka i=se Cada\\ \gl{cnj} call for \gl{def}.\gl{sg}=other cycas.rattle \gl{cnj}  \gl{def}.\gl{sg}=other beat\\ \glt \qu{so he gave one the first name of Neen (Seed Rattle) and the other Cada (Bark Clapper Beat).}
\z

\ea
\gll Ka a=vi nyako i=inya-lu ma a=fee-lu ma le=hân cai-n pala-n.\\ \gl{cnj} 3\gl{sg}=say for \gl{def}.\gl{sg}=mother-3\gl{du}.\gl{poss} \gl{subr} 3\gl{sg}=take-3\gl{du}.\gl{obj} \gl{subr} 3\gl{pl}=go behind-3\gl{sg}.\gl{poss} home-3\gl{sg}.\gl{poss}\\ \glt \qu{And he told their mother to take them and follow him home.}
\z

\ea
\gll Ma le=hma-san a=feanake si-le joakan juu mwa,\\ 
     \gl{subr} 3\gl{pl}=arrive-go 3\gl{sg}=show for-3\gl{pl} big real house\\ 
     \glt \qu{When he arrived he would designate a large hut for her, }
\z

\ea
\gll  vwa can vi ``In Fwe ta ca i=juu mwa-ca ma go=silaa mu=nyai-m."\\ 
    do while say skin figtree go.up in \gl{def}.\gl{sg}=real house-\gl{prox} \gl{subr} 2\gl{sg}=raise \gl{def}.\gl{du}=child-2\gl{sg}.\gl{poss}\\ 
    \glt \qu{...saying “Ciin Fwe enter this hut and bring up your children.”}
\z

\ea
\gll a=vi nyakoo-n ``hole ka ma bo vwa xahnang nyasi-m meekan"\\ 3\gl{sg}=say for-3\gl{sg} thanks \gl{cnj} \gl{subr} \gl{irr} do good for-2\gl{sg} all\\ \glt  \qu{She says “Thank you and that everything be returned to you.”}
\z

\ea
\gll Fine bwaabwen pathaabua i=juu mwa a=mu hmaa-ko xeen uvu, xeen xhwaeo, nyu ma juu-mani.\\ 
    every morning before \gl{def}.\gl{sg}=real house 3\gl{sg}=\gl{freq} arrive-on basket yam basket taro fish \gl{com} real-bird\\ 
    \glt \qu{Every morning, in front of her house, she finds baskets of yam and taro, fish and goliath imperial pigeons.}
\z

\ea
\gll Ma cika vukin ma a=xaahni ca aman ma a=vwa tââ-n\\ \gl{subr} \gl{neg}.\gl{exist} reason \gl{subr} 3\gl{sg}=search some thing \gl{subr} 3\gl{sg}=do oven-3\gl{sg}.\gl{poss}\\ \glt  \qu{So that she didn't have to worry about what she put in her oven.}
\z

\section{The Boys Grow Up}
\ea
\gll Le=ha-me ka li=jo, mu=xayu ja xhopwe-lu mwa, ja bwahliia-lu mwa, na a=ja tha-xaleke ka Xa-xhwi Apuli.\\ 3\gl{pl}=go-\gl{dir.cp} \gl{sbj} \gl{def}.\gl{pl}=year \gl{def}.\gl{du}=boy \gl{prf} be.big-3\gl{du} \gl{deict} \gl{prf} be.long-3\gl{du} \gl{deict} \gl{dem} 3\gl{sg}=\gl{prf} strongly-see \gl{sbj} \gl{agt}.\gl{nmlz}-eat person\\ \glt \qu{Years have passed, the boys are growing up, gaining height, which has not escaped Kawi Kayuk.}
\z

\ea
\gll Ca i=se thoatit a=vi nyako In Fwe ``In Fwe ta-me xaleke, ta-me cuut cel-ong pwa i=vai-ca".\\ in \gl{def}.\gl{sg}=other day 3\gl{sg}=say for skin figtree skin figtree go-\gl{dir.cp} see go-\gl{dir.cp} stand next-1\gl{sg}.\gl{poss} on \gl{def}.\gl{sg}=stone-\gl{prox}\\ \glt \qu{One day he said to Ciin Fwe: ``Ciin Fwe come and see, come and stand next to me on this rock."}
\z

\ea
\gll A=ja vwa ka In Fwe.\\ 3\gl{sg}=\gl{prf} do \gl{sbj} skin figtree\\ \glt \qu{This Ciin Fwe does.}
\z

\ea
\gll A=vi nyakoo-n ``xaleke pathabuaa-m, meekan i=pal-ong"\\ 3\gl{sg}=say for-3\gl{sg}.\gl{poss} see before-2\gl{sg}.\gl{poss} all \gl{def}.\gl{sg}=home-1\gl{sg}.\gl{poss}\\ \glt  \qu{He says to her ``Look before you, this is my territory.} 
\z

\ea
\gll Xaahni cama go=vwa nya-la i=nyangan-aman-go.\\ watch.with.purpose if 2\gl{sg}=do put-be.here \gl{def}.\gl{sg}=garden-thing-2\gl{sg}.\gl{poss}\\ \glt  \qu{Choose where you want to do your field."}
\z

\ea
\gll a=sabe nyasi-n ``vwasoon m=e hân ca-n nyangan-aman ko mu=xhwaawe bwa cika wîî-lu"\\ 3\gl{sg}=answer for-3\gl{sg}.\gl{poss} impossible \gl{subr}=1\gl{sg} go in-\gl{nspec} garden-thing because \gl{def}.\gl{du}=child \gl{ipfv} \gl{neg}.\gl{exist} strength-3\gl{du}.\gl{poss}\\ \glt \qu{She answers ``But I cannot go to the field because the children are not strong enough!"}
\z

\ea
\gll A=vi nyakoo-n hapi ``hân, e=bo xaahni-lu."\\ 3\gl{sg}=say for-3\gl{sg}.\gl{poss} \gl{comp} go 1\gl{sg}=\gl{irr} watch.with.purpose-3\gl{du}.\gl{obj}\\ \glt  \qu{He tells her ``Go. I will take care of them”}
\z

\ea
\gll I=bwaabwenan a=ja hân fwadai ca=ma a=vwa nyangan-aman la.\\ \gl{def}.\gl{sg}=next.day 3\gl{sg}=\gl{prf} go look.for \gl{indf}.\gl{sg}=\gl{subr} 3\gl{sg}=do garden-thing be.here\\ \glt \qu{The next day she left to see where she will do the field.}
\z

\ea
\gll Ka a=siiwa can batebwen\\ \gl{cnj} 3\gl{sg}=return in evening\\ \glt \qu{And she returns in the evening.}
\z

\ea
\gll Cama a=ta ca i=juu mwa a=tadake ma a=tena mu=nyain a=lu=kon pala.\\ \gl{subr} 3\gl{sg}=go.up in \gl{def}.\gl{sg}=real house 3\gl{sg}=be.surprised \gl{subr} 3\gl{sg}=hear \gl{def}.\gl{du}=child-3\gl{sg} \gl{rel}=\gl{du}=\gl{prog} talk\\ \glt  \qu{As she enters the hut she is surprised to hear her children speak.}
\z

\ea
\gll A=fwajimwa-lu ``a=ecaa-kau ko-n pala ka=kai?"\\ 3\gl{sg}=ask-3\gl{du} 3\gl{sg}=teach-2\gl{du} \gl{obl}-\gl{nspec} talk \gl{sbj}=who\\ \glt \qu{She asks them ``Who taught you to speak?"}
\z

\ea
\gll Lu=sabe nyasi-n ``ko na Xa-xhwi Apuli"\\ 3\gl{du}=answer for-3\gl{sg} \gl{cnj} \gl{dem} \gl{agt}.\gl{nmlz}-eat person\\ \glt \qu{They reply ``It was Kawi Kayuk"}
\z

\ea
\gll Ca i=se thoatit a=ja=hnyimake ma hân sanya.\\ in \gl{def}.\gl{sg}=other day 3\gl{sg}=\gl{prf}=think \gl{subr} go clear.yam.field\\ \glt \qu{One day she decides to go and prepare the yam field.}
\z

\ea
\gll A=fwajimwa Xa-xhwi Apuli ``goon m=e=hân can nyangan-aman?"\\ 3\gl{sg}=ask \gl{agt}.\gl{nmlz}-eat person enough \gl{subr}=1\gl{sg}=go in garden-something\\ \glt \qu{She asks Kawi Kayuk ``Can I go and do the field?"} 
\z

\ea
\gll A=vi nyakoo-n ``Hân go=bwa=ha-mwa-me cama go=bwa=koin."\\ 3\gl{sg}=say to-3\gl{sg} go 2\gl{sg}=\gl{ipfv}=go-\gl{rep}-\gl{dir.cp} when 2\gl{sg}=\gl{ipfv}=finish\\ \glt \qu{He says to her ``Go and come back when you are done."}
\z

\ea
\gll Koin a=ta can nyangan-aman Xa-xhwi Apuli a=ecaa mu=xawe ko li=yata li=phwê ma i=wadan a=go=vwa i=siya"\\ while 3\gl{sg}=go.up in garden-something \gl{agt}.\gl{nmlz}-eat person 3\gl{sg}=learn \gl{def}.\gl{du}=young \gl{obl} \gl{def}.\gl{pl}=name \gl{def}.\gl{pl}=moon \gl{com} \gl{def}.\gl{sg}=time \gl{rel}=2\gl{sg}=do \gl{def}.\gl{sg}=yam.field\\ \glt  \qu{While she goes to the field, Kawi Kayuk teaches the children the names of the months and the seasons, the yam calendar.}
\ex
\gll Can bate-bwen a=ja=siiwa mwa ka i=nyanyanlu, mu=nyai-n, lu=vi nyakoo-n:\\ in tip-night 3\gl{sg}=\gl{prf}=return \gl{rep} \gl{sbj} \gl{def}.\gl{sg}=mother-3\gl{du}.\gl{poss} \gl{def}.\gl{du}=child-3\gl{sg}.\gl{poss} 3\gl{du}=say to-3\gl{sg}.\gl{poss}\\ \glt \qu{In the evening when the mother comes home the children tell her:}
\ex
\gll ``Nyanya tabo la ma go=tena"\\ mom sit be.here \gl{subr} 2\gl{sg}=hear\\ \glt  \qu{``Mom sit down and listen to this!"}
\ex
\gll ka lu=thapoke saxhuti li=a= lu=ecaa jahnga i=thoatit\\ \gl{cnj} 3\gl{du}=begin tell.story \gl{def}.\gl{pl}=\gl{rel}= 3\gl{du}=learn length \gl{def}.\gl{sg}=day\\ \glt  \qu{And they start to tell her what they learned during the day:}
\ex
\gll i=jo, li=wadan, li=thoatit ma li=phwê.\\ \gl{def}.\gl{sg}=year \gl{def}.\gl{pl}=time \gl{def}.\gl{pl}=day \gl{com} \gl{def}.\gl{pl}=moon\\ \glt \qu{the yam calendar, the seasons, the days and the months.}
\ex
\gll ``Ko na kai a=ecaa-kau ko nien-aen?" a=fwajimwa-lu ka In Fwe.\\ \gl{cnj} \gl{dem} who \gl{rel}=learn-2\gl{du}.\gl{obj} \gl{obl} \gl{def}.\gl{pl}-\gl{prox} 3\gl{sg}=ask-3\gl{du}.\gl{obj} \gl{sbj} skin figtree\\ \glt  \qu{“But who was it that taught you all this?" Ciin Fwe asks her boys.}
\ex\largerpage
\gll ``Na Xa-xhwi Apuli !"\\ \gl{dem} \gl{agt}.\gl{nmlz}-eat person\\ \glt \qu{“It was Kawi Kayuk"} 
\ex
\gll A=vi nyakoo-lu ``mu=xayu-n-eong, vwa juu xahnang nyim-ong koo-u!"\\ 3\gl{sg}=say to-3\gl{du} \gl{def}.\gl{du}=male-\gl{link}-1\gl{sg}.\gl{poss} \gl{exist} real good heart-1\gl{sg}.\gl{poss} \gl{obl}-2\gl{du}\\ \glt \qu{She tells them ``My boys, I am very proud of you!"}
\z 

\ea
\gll Hâ-mwa-me i=wadan ma a=hâ-mwa saavi i=hnea-n.\\ go-\gl{rep}-\gl{dir.cp} \gl{def}.\gl{sg}=time \gl{subr} 3\gl{sg}=go-\gl{rep} visit \gl{def}.\gl{sg}=field-3\gl{sg}.\gl{poss}\\ \glt \qu{There comes the time when she must go and visit her field.}
\z 

\ea
\gll A=ce mu=xayu pala i=Xa-xhwi Apuli.\\ 3\gl{sg}=leave \gl{def}.\gl{du}=boy at.home \gl{def}.\gl{sg}=\gl{agt}.\gl{nmlz}-eat person\\ \glt \qu{She entrusts her boys to Kawi Kayuk.} 
\z 

\ea
\gll Cala ha-mwa ca-n batebwen a=hmaa-ko mu=nyai-n lu=kon=jala.\\ when go-\gl{rep} in-\gl{nspec} night 3\gl{sg}=arrive-on \gl{def}.\gl{du}=child-3\gl{sg}.\gl{poss} 3\gl{du}=\gl{prog}=play\\ \glt  \qu{When she comes home in the evening, she finds her children playing {[}skill] games.}
\z

\ea
\gll A=tadake ``gau=kon=vwa da?"\\ 3\gl{sg}=be.surprised 2\gl{du}=\gl{prog}=do what\\ \glt  \qu{She wonders, ``What are you doing?"}
\z

\ea
\gll Lu=vi nyakoo-n: ``go=caihna-n naen goon ma=abu= ja=vap ma vwa-tau, pa=ja=tehnang li=si-bu"\\ 3\gl{du}=say to-3\gl{sg} 2\gl{sg}=know-\gl{nspec} now enough \gl{subr} 1\gl{du}.\gl{excl}=\gl{prf}=hunt \gl{com} make-impact \gl{prf}=\gl{prf}=sharp \gl{def}.\gl{pl}=arm-1\gl{du}.\gl{incl}.\gl{poss}\\ \glt \qu{They say to her ``You know, now we can go hunting and fishing, we are now skillful."}
\z 

\ea
\gll Kona a-vi nyakoo-lu ``xahmaen gase=bo ta ca-n siya ma gau=bo=balan-thaap"\\ \gl{cnj} 3\gl{sg}=say for-3\gl{du} tomorrow 1\gl{pl}.\gl{incl}=\gl{irr} go.up in-\gl{nspec} field \gl{subr} 2\gl{du}=\gl{irr}=walk-nyaouli\\ \glt \qu{So she says to them “Tomorrow we will go to the field so that you can hunt (lit. walk around among the nyaouli trees)."}
\z

\ea
\gll I=bwaabwenan le=saahma mu-bwen-nyoot, lu=fe li=daa-lu ma li=jikela-lu,\\ 
    \gl{def}.\gl{sg}=next.day 3\gl{pl}=rise little-night-emerge 3\gl{du}=take \gl{def}.\gl{pl}=spear-3\gl{du}.\gl{poss} \gl{com} \gl{def}.\gl{pl}=arrow-3\gl{du}.\gl{poss}\\ 
    \glt  \qu{The next day, getting up early, they took their spears and arrows}
\z

\ea
\gll ka cahma i=nyanya a=fe i=xee-n japita-le\\ \gl{cnj} \gl{subr} \gl{def}.\gl{sg}=mom 3\gl{sg}=take \gl{def}.\gl{sg}=basket-\gl{link} provision-3\gl{pl}.\gl{poss}\\ \glt \qu{...and as to the mom, she takes her basket of provisions.}
\z
 
\ea
\gll Lu=ce i=inya-lu ca-n siya ma lu=hân vap pwan bwan fwadai li=juu-mani hman masoo.\\ 3\gl{du}=leave \gl{def}.\gl{sg}=mother-3\gl{du}.\gl{poss} in-\gl{nspec} field \gl{subr} 3\gl{du}=go hunt on-\gl{nspec} mountain look.for \gl{def}.\gl{pl}=real-bird also fruitbat\\ \glt  \qu{Leaving their mother in the field they go hunting in the heights to find goliath doves and fruitbats.}
\z

\ea
\gll Cala lu=hma-san vwa i=ape-mapeke lu=tabo ma lu=bwaa=hmwet\\ when 3\gl{du}=arrive-move.same.level \gl{exist} \gl{def}.\gl{sg}=\gl{loc}.\gl{nmlz}-bright 3\gl{du}=sit \gl{subr} 3\gl{du}=\gl{ipfv}=tired\\ \glt \qu{Once they arrive in a clearing the two sit down to rest.} 
\z

\ea
\gll Ngein a=vi nyako i=been ``Cada, xaleke, moo cahni gasu=xaleke i=jati"\\ N. 3\gl{sg}=say for \gl{def}.\gl{sg}=brother-3\gl{sg}.\gl{poss} C. see stay here 1\gl{du}.\gl{incl}=see \gl{def}.\gl{sg}=sea\\ \glt \qu{Neen tells his brother ``Cada. Look, from here we see the sea."}
\z


\ea
\gll A=vi ka Cada ``Ma cai-ju gasu=xaleke i=yeen-bwan"\\ 3\gl{sg}=say \gl{sbj} C \gl{subr} behind-1\gl{du}.\gl{incl} 1\gl{du}.\gl{incl}=see \gl{def}.\gl{sg}=tree-mountain\\ \glt \qu{Cada says ``Whereas behind us we see the mountain top."}
\z


\ea
\gll 	``Kavi tha=go=xaleke i\textsubscript{{\upshape SPEC}}={\ob}e kon xaleke{\cb} na hmuun mae!"\\	but \gl{ass}=2\gl{sg}=see \gl{def}.\gl{sg}=\gl{rel}.1\gl{sg} \gl{prog} see \gl{dem} smoke fire	\\ \glt  \qu{``But do you see what I'm seeing, that's smoke!"}		
\z 

\ea
\gll 	``Hmwakan vwa apuli\textsubscript{{\upshape NSPEC}} xada, gasu=bo=fwajimwa nyaanya"\\	maybe \gl{exist} person up.there 1\gl{du}.\gl{incl}=\gl{fut}=ask mum	\\ \glt  \qu{``Maybe there are people up there, we'll ask Mum."}	
\z 

\ea
\gll 	Cala lu=fwajimwe-a, a=vwa e-vwaseekan ka In Fwe ko a=caihnan hapi na la i\textsubscript{{\upshape SPEC}}=xhoogo-nea.\\ when 3\gl{du}=ask-3\gl{sg} 3\gl{sg}=do \gl{ins}-sad \gl{sbj} skin figtree because 3\gl{sg}=know.\gl{inan} \gl{comp} \gl{dem} be.here \gl{def}.\gl{sg}=home-3\gl{sg}.\gl{poss}	\\ \glt  \qu{When they ask her, Fig Bark is sad because she knows that this is her home.}		
\z 

\ea
\gll 	Ka a=sabe nya.sii-lu ``hmwakan vwa apuli\textsubscript{{\upshape NSPEC}} xada."	\\	and 3\gl{sg}=lift.to.mouth give.hand-3\gl{du} maybe \gl{exist} people up.there	\\ \glt  \qu{And she answers them ``Maybe there are people up there."}	
\z 

\ea
%\langinfo{}{}{} Ungrammatical: *\textit{ca hmuu-n mae-aen}
\gll Ca i=wadan a=e-thaloo-kan tha see ma i=a vi,\\
     in.\gl{spec} \gl{def}.\gl{sg}=time \gl{rel}=\gl{ord}-two-\gl{ord} \gl{ass}.3\gl{sg} cry \gl{com} \gl{def}.\gl{sg}=\gl{rel}.3\gl{sg} say,\\ 
     \glt  \qu{The second time she cries about what he said,...}		
\z

\ea
\gll kavi a=vi nyakolu hapi ``ca \textbf{eca} se wadan gase=bo ta xaleke i=hmuu-n mae-aen."\\ \gl{cnj} 3\gl{sg}=say to-3\gl{du} that in some other time 1\gl{pl}.\gl{incl}=\gl{fut} go.up see	\gl{def}.\gl{sg}=smoke-\gl{link} fire-\gl{dem}\\ \glt \qu{...but she tells them that ``Some other time (singular) we'll go up and/to look at that smoke."}
\z


\section{Mathila}

\ea
\gll Ko i=si-n thu cipa le=xaleke \textbf{eca} mani a=xhwatiin yata-n mathila\\ 
     on \gl{def}.\gl{sg}=arm-\gl{link} banyan \gl{neg} 3\gl{pl}=see some bird \gl{rel}=small name-3\gl{sg}.\gl{poss} M.\\ 
     \glt \qu{They didn't see some small bird on the banyan branch called Mathila [streaked fantail]...}	
\z

\ea
\gll ko a=kon fwajimwake hapi na kai ni-ehni a=le=kon bwaa hma.\\ but 3\gl{sg}=\gl{prog} ask that \gl{dem}  who \gl{def}.\gl{pl}-\gl{prox} \gl{rel}=3\gl{pl}=\gl{prog} \gl{ipfv} arrive\\ \glt \qu{...but it was wondering who these people were who had just arrived.}
\z

\ea
\gll I=mani-ca, fine li=thoatit pwecake-n goon-xat, thapoke i=wadan a=cavi-a ka In-Fwe, a=mu ha-me xhaavwa koo-n a=xaahni i=hun-siiwa ka-n.\\ \gl{def}.\gl{sg}=bird-\gl{prox} count \gl{def}.\gl{pl}=day-\gl{nspec} after enough-sun begin \gl{def}.\gl{sg}=time \gl{rel}=carry.away-3\gl{sg}.\gl{obj} \gl{sbj} I. 3\gl{sg}=\gl{freq} go-\gl{dir.cp} wait.around \gl{obl}-\gl{ana} 3\gl{sg}=see  \gl{def}.\gl{sg}=\gl{nmlz}-return \gl{link}-\gl{ana}\\ \glt \qu{This bird, every day at noon, since the time that In Fwe was swept away, it waited for when it would see her return.}
\z

\ea
\gll I=mani-aen, i=xa-vuki-n i=daahma Thêa Xa-vila\\ 
    \gl{def}.\gl{sg}=bird-\gl{dist} \gl{def}.\gl{sg}=\gl{agt}.\gl{nmlz}-stem-3\gl{sg}.\gl{poss} \gl{def}.\gl{sg}=chief Firstborn \gl{agt}.\gl{nmlz}-dance\\ 
    \glt \qu{This bird, its owner was the chief Firstborn Dancer,...}
\z

\ea
\gll  a=fa-tena nyakoo-n meekan li=aman le=mu hmaa koo-le nya-ca i=ape-moo.\\ 3\gl{sg}=\gl{caus}-hear for-3\gl{sg}.\gl{poss} all \gl{def}.\gl{pl}=thing 3\gl{pl}=\gl{iter} arrive in-3\gl{pl}.\gl{poss} inside-\gl{loc} \gl{def}.\gl{sg}=\gl{loc}.\gl{nmlz}-stay\\ \glt \qu{... it let him know all the things which would happen to them in the tribe}
\z

\ea
\gll a=hân tatu vii nyako i=daahma hapi le=hma ka li=xaaya ca i=ape-moo.\\ 3\gl{sg}=go do.quickly say for \gl{def}.\gl{sg}=chief \gl{comp} 3\gl{pl}=arrive \gl{sbj} \gl{def}.\gl{pl}=stranger in \gl{def}.\gl{sg}=\gl{loc}.\gl{nmlz}-stay\\ \glt \qu{It quickly goes to tell the chief that strangers had arrived in the tribe.}
\z

\ea
\gll I=daahma a=vi nyakoo-n ``hân xaleke go bwa ha-me vi hapi le=kon vwa da, nievit jevwa-le"\\ \gl{def}.\gl{sg}=chief 3\gl{sg}=say for-3\gl{sg} go see then \gl{ipfv} go-\gl{dir.cp} say \gl{comp} 3\gl{pl}=\gl{prog} do what how.many number-3\gl{pl}.\gl{poss}\\ \glt \qu{The chief, he tells it ``Go see, then return to tell me what they are doing, and how many they are."}
\z 

\ea
\gll A=sabe nyasi-n ``thaloo apuli ma thamo"\\ 3\gl{sg}=pick.up for-3\gl{sg} two man \gl{com} woman\\ \glt \qu{It answers him ``Two men and a woman."}
\z

\ea
\gll ``Hân xhaavwa koo-le!"\\ go wait \gl{obl}-3\gl{pl}\\ \glt \qu{Go wait for them!}
\z

\ea
\gll I=bwaabwenan pwecaken goon-xat a=hu-pwa\\ \gl{def}.\gl{sg}=next.day after enough-sun 3\gl{sg}=go.down-\gl{rep}\\ \glt \qu{The next day after noon it goes down.}
\z

\ea
\gll A=xaleke hapi tehnang si-lu ko i=vwa tau ma i=vap, juu vaa vwa wîî-lu\\ 3\gl{sg}=see \gl{comp} sharp hand-3\gl{du}.\gl{poss} \gl{obl} \gl{def}.\gl{sg}=do impact \gl{com} \gl{def}.\gl{sg}=hunt real much \gl{exist} strength-3\gl{du}\\ \glt \qu{It sees that their hands are deft at fishing and hunting, that they are very strong.}
\z

\ea
\gll Tha lu=xaleke i=mani\\ \gl{ass} 3\gl{du}=see \gl{def}.\gl{sg}=bird\\ \glt \qu{They see the bird.}
\z

\ea
\gll A=fwajiwa-lu ``tha gau=xale-o?" Lu=vi hapi ``hê"\\ 3\gl{sg}=ask-3\gl{du} \gl{ass} 2\gl{du}=see-1\gl{sg} 3\gl{du}=say \gl{comp} yes\\ \glt \qu{It asks them ``Can you see me?" They say ``Yes".}
\z

\ea
\gll ``Ka gau=tena-o?" Lu=vi hapi ``hê"\\ \gl{cnj} 2\gl{du}=hear-1\gl{sg} 3\gl{du}=say \gl{comp} yes\\ \glt \qu{``And can you hear me?" They say ``Yes".}
\z

\ea 
\gll i=mani a=fwajimwa-lu ``kavi gau=kai ka gau=hân heeve? Gau=hâ-me moo ve?"\\ \gl{def}.\gl{sg}=bird 3\gl{sg}=ask-3\gl{du} but 2\gl{du}=who \gl{cnj} 2\gl{du}=go where.mobile 2\gl{du}=go-\gl{dir.cp} stay where.immobile\\ \glt \qu{The bird asks them ``But who are you, where are you going? Where do you come from?"}
\z

\ea
\gll lu=saxhut i=jaxhut ko i=inya-lu\\ 3\gl{du}=tell \gl{def}.\gl{sg}=story on \gl{def}.\gl{sg}=mother-3\gl{du}.\gl{poss}\\ \glt \qu{They tell the story of their mother.}
\z

\ea
\gll i=mani a=fwajimwa-lu ``ka heeve i=inya-u?"\\ \gl{def}.\gl{sg}=bird 3\gl{sg}=ask-3\gl{du} \gl{cnj} where \gl{def}.\gl{sg}=mother-2\gl{du}.\gl{poss}\\ \glt \qu{The bird asks them ``And where is your mother?"}
\z

\ea
\gll a=vi ka In fwe hapi ``na yo"\\ 3\gl{sg}=say \gl{sbj} skin figtree \gl{comp} \gl{dem} 1\gl{sg}\\ \glt \qu{Figtree Bark says ``That's me".}
\z

\ea
\gll In Fwe tha a=pa mu wanke ko li=jo tha pa hân\\ skin figtree \gl{ass} 3\gl{sg}=\gl{prf} \gl{freq} change because \gl{def}.\gl{pl}=year \gl{ass} \gl{prf} go\\ \glt \qu{Figtree Bark has changed because years have passed.}
\z

\ea
\gll a=vi ka i=mani hapi ``e=bo hân vi nyako i=daahma li=aman a=gavwe saxhuti"\\ 3\gl{sg}=say \gl{sbj} \gl{def}.\gl{sg}=bird \gl{comp} 1\gl{sg}=\gl{irr} go say for \gl{def}.\gl{sg}=chief \gl{def}.\gl{pl}=thing 3\gl{sg}=2\gl{pl} tell.story\\ \glt \qu{The bird says ``I will go tell the chief what you told me."}
\z

\ea
\gll a=vi nyako i=daahma ka i=mani kavi saxhwe-a ka i=see ma faphâke\\ 
    3\gl{sg}=say for \gl{def}.\gl{sg}=chief \gl{sbj} \gl{def}.\gl{sg}=bird but refuse-3\gl{sg} \gl{sbj} \gl{def}.\gl{sg}=other \gl{subr} believe\\ 
    \glt \qu{It tells the chief but the other refuses to believe the bird,...}
\z

\ea
\gll ko i=mani kona a=hnyimake hapi me-a ka In Fwe ca i=java.\\ \gl{obl} \gl{def}.\gl{sg}=bird \gl{cnj} 3\gl{sg}=think \gl{comp} die-3\gl{sg} \gl{sbj} skin figtree in \gl{def}.\gl{sg}=flood\\ \glt \qu{...since he thinks that Figtree Bark died in the flood.}
\z

\ea
\gll a=vi nyako i=mani ``fê-â-me-le m=e faphâke nyakoo-m"\\ 3\gl{sg}=say for \gl{def}.\gl{sg}=bird take-go-\gl{dir.cp}-3\gl{pl}.\gl{obj} \gl{subr}=1\gl{sg} believe for-2\gl{sg}.\gl{poss}\\ \glt \qu{He tells the bird ``Bring them here so that I may believe you."}
\z

\ea
\gll Koin i=mani a=siiwa hut ca i=jahoot, a=nyasipoke meeka li=apuli ca i=ape-moo ka i=daahma ka le=xhavwaleke ma le=fe-â-me li=apuli-ca.\\ while \gl{def}.\gl{sg}=bird 3\gl{sg}=return go.down in \gl{def}.\gl{sg}=river 3\gl{sg}=gather all \gl{def}.\gl{pl}=person in \gl{def}.\gl{sg}=\gl{loc}.\gl{nmlz}-stay \gl{sbj} \gl{def}.\gl{sg}=chief \gl{cnj} 3\gl{pl}=wait \gl{subr} 3\gl{pl}=take-go-\gl{dir.cp} \gl{def}.\gl{pl}=person-\gl{prox}\\ \glt \qu{After the bird returns to the river, the chief meets all the people in village and they wait for them to bring these people to them.}
\z

\section{Going Home}

\ea
\gll Cala a=hma-san ca i=ape-moo a=e-imwi i=hun-see ka In Fwe ko a=caihna mwa li=bee-n.\\ when 3\gl{sg}=arrive-go in \gl{def}.\gl{sg}=\gl{loc}.\gl{nmlz}-stay 3\gl{sg}=\gl{refl}-hold \gl{def}.\gl{sg}=\gl{nmlz}-cry \gl{sbj} skin figtree because 3\gl{sg}=know \gl{rep} \gl{def}.\gl{pl}=peer-3\gl{sg}.\gl{poss}\\ \glt \qu{When she arrives in the village, Figtree Bark holds back her tears, for she recognizes her family.}
\z

\ea
\gll i=daahma a=hnya-le can hâwân.\\ 
    \gl{def}.\gl{sg}=chief 3\gl{sg}=send-3\gl{pl} in front\\ 
    \glt \qu{The chief sends them before him.}
\z

\ea
\gll a=fwajimwa-le na kai niehni.\\ 
    3\gl{sg}=ask-3\gl{pl} \gl{dem} who \gl{dem}.\gl{pl}\\ 
    \glt \qu{He asks them who they are.}
\z

\ea
\gll lu=sabe nyasi-n ``yo Ngeein ka yo Cada, ka cahma i=thamo-ca na i=inya-bu In Fwe"\\ 3\gl{du}=answer for-3\gl{sg} 1\gl{sg} cycas.rattle \gl{cnj} 1\gl{sg} beat \gl{cnj} whereas \gl{def}.\gl{sg}=woman-\gl{prox} \gl{dem} \gl{def}.\gl{sg}=mother-1\gl{du}.\gl{excl}.\gl{poss} skin figtree\\ \glt \qu{They answer him ``I am Cycas Rattle and I am Beat, and as to this woman, that is our mother Figtree Bark."}
\z

\ea
\gll a=vi nyakoo-le ``ûhû! Me-a ka In Fwe paa hmain li=jo, a=thêên fe-a ka i=we"\\ 3\gl{sg}=say for-3\gl{pl} no die-3\gl{sg} \gl{sbj} skin figtree \gl{prf} many \gl{def}.\gl{pl}=year 3\gl{sg}=run take-3\gl{sg}.\gl{obj} \gl{sbj} \gl{def}.\gl{sg}=water\\ \glt \qu{He tells them ``No! Figtree Bark died many years ago, the water rushed her away."}
\z

\ea
\gll goon ma gase=fêanake hmwaeke i=juuju?\\ enough \gl{subr} 1\gl{pl}.\gl{excl}=show how \gl{def}.\gl{sg}=truth\\ \glt \qu{``How can we show you the truth?"}
\z

\ea
\gll a=vi nyakoo-lu hapi\\ 
    3\gl{sg}=say for-3\gl{du} \gl{comp}\\ 
    \glt \qu{He tells them that ...}
\z

\ea
\gll ``gau=fêanake s-ung hapi tehnang li=si-u ko i=vwa tau ma vap vwa can fê-a-mwa-me le=ilake si-u"\\ 2\gl{du}=show \gl{ben}-1\gl{sg}.\gl{poss} \gl{comp} sharp \gl{def}.\gl{pl}=hand-2\gl{du}.\gl{poss} \gl{obl} \gl{def}.\gl{sg}=do impact \gl{com} hunt do while take-go-\gl{rep}-\gl{dir.cp} 3\gl{pl}=ask \gl{ben}-2\gl{du}\\ \glt \qu{``You show me that your hands are deft in fishing and hunting by bringing something they ask of you."}
\z

\ea
\gll a=ilake juu-mani nyasi Ngeein ka veena nyasi Cada\\ 3\gl{sg}=ask real-bird for rattle \gl{cnj} eel for beat\\ \glt \qu{He asks a goliath imperial pigeon of Rattle and an eel of Beat.}
\z

\ea
\gll Ka lu=the-hân ka lu=siiwa fe mwa li=aman a=a=ilake\\ \gl{cnj} 3\gl{du}=do.running-go \gl{cnj} 3\gl{du}=return take \gl{rep} \gl{def}.\gl{pl}=thing \gl{rel}=3\gl{sg}=demand\\ \glt \qu{And they rush off and they bring back the things he demanded.}
\z

\ea
\gll a=vi nyakoolu hapi\\ 
      3\gl{sg}=say for-3\gl{du}.\gl{obj} \gl{comp}\\ 
      \glt \qu{He tells them that...}
\z

\ea
\gll ``naen gau=feânake s-ung hapi vwa li=nyamaa-u ma gau=xaleke ka li=jenau ma gau=tena i=xa-thêên-fe-fati-n-eong mathila,..."\\ now 2\gl{du}=show \gl{ben}-1\gl{sg} \gl{comp} \gl{exist} \gl{def}.\gl{pl}=eye-2\gl{du}.\gl{poss} \gl{subr} 2\gl{du}=see \gl{cnj} \gl{def}.\gl{pl}=ear-2\gl{du}.\gl{poss} \gl{subr} 2\gl{du}=hear \gl{def}.\gl{sg}=\gl{agt}.\gl{nmlz}-run-take-speech-\gl{link}-1\gl{sg}.\gl{poss} streaked.fantail\\ \glt \qu{``Now you two show me that you have the eyes to see and the ears to hear my messenger Mathila..."}
\z

\ea
\gll  ``ko i=mani-ca cipa le=xaleke ka meekan apuli"\\ because \gl{def}.\gl{sg}=bird-\gl{prox} \gl{neg} 3\gl{pl}=see \gl{sbj} all person\\ \glt \qu{``..., for this bird is not seen by all people."}
\z

\ea
\gll cahma i=mani a=hân tabo ko i=sin yee a=hoot a=cuut nyeca i=xhoogo ka a=fwajimwa-ke ``yo kai ka heeve-o"\\ whereas \gl{def}.\gl{sg}=bird 3\gl{sg}=go sit on \gl{def}.\gl{sg}=branch tree \gl{rel}=far \gl{rel}=stand inside \gl{def}.\gl{sg}=home \gl{cnj} 3\gl{sg}=ask-\gl{tr} 1\gl{sg} who \gl{cnj} where-1\gl{sg}\\ \glt \qu{As for the bird, it goes to sit on a tree branch far away, and stands in the village and asks ``Who am I, and where?"}
\z

\ea
\gll lu=sabe ``go=tabo pwa i=sin yee a=mwa hoot ka go=i=mathila."\\ 3\gl{du}=pick.up 2\gl{sg}=sit on \gl{def}.\gl{sg}=branch tree \gl{rel}.3\gl{sg}=\gl{deict} far \gl{cnj} 2\gl{sg}=\gl{def}.\gl{sg}=streaked.fantail\\ \glt \qu{They answer ``You sit on the tree branch that is far, and you are the streaked fantail."}
\z

\ea
\gll a=vi ka matila hapi ``thake bwajep m=e vila"\\ 3\gl{sg}=say \gl{sbj} M. \gl{comp} throw bark.clapper \gl{subr}=1\gl{sg} dance\\ \glt \qu{Mathila says ``Beat the bark clappers so that I may dance."}
\z

\ea
\gll Cala lu=thake bwajep i=jamwa-lu a=tena i=xheen ma i=geen a=fa-hnyimake nyako li=hun-vila ka-n habu\\ 
     when 3\gl{du}=strike bark.clapper \gl{def}.\gl{sg}=father-3\gl{du}.\gl{poss} 3\gl{sg}=hear \gl{def}.\gl{sg}=sound \gl{com} \gl{def}.\gl{sg}=voice 3\gl{sg}=\gl{caus}-think for \gl{def}.\gl{pl}=\gl{nmlz}-dance \gl{link}-\gl{nspec} long.ago\\ \glt \qu{When the two beat the bark clappers, [the chief] hears the sound and the voice that reminds him of the dances of yore,}
\z

\ea
\gll ka a=vi hapi ``na Ngeein ma Cada, mu=nyae-ung!"\\  
    \gl{cnj} 3\gl{sg}=say \gl{comp} \gl{dem} rattle \gl{com} beat \gl{def}.\gl{du}=child-1\gl{sg}.\gl{poss}\\ 
    \glt \qu{...and he says ``They are Rattle and Beat, my sons!"}
\z

\ea
\gll Le=e-imwi-le ka le=e-see ma i=inya-lu\\ 
    3\gl{pl}=\gl{recp}-grasp-3\gl{pl} \gl{cnj} 3\gl{pl}=\gl{recp}-one \gl{com} \gl{def}.\gl{sg}=mother-3\gl{du}.\gl{poss}\\ 
\glt \qu{They embrace each other and they unite with their mother.}
\z

\ea
\gll A=fwajimwake nyasi In Fwe ma a=saxhuti i=jaxhut-ea ca i=ape-thapoke-ka-n\\ 3\gl{sg}=ask for skin figtree \gl{com} 3\gl{sg}=tell.story \gl{def}.\gl{sg}=story-3\gl{sg}.\gl{poss} in \gl{def}.\gl{sg}=\gl{loc}.\gl{nmlz}-begin \gl{link}-\gl{nspec}\\ \glt \qu{He asks Figtree Bark to tell her story at the beginning.}
\z

\section{Conclusion}

\ea
\gll Cala koin i=jaxhut-ea i=daahma a=tho nyako In Thu ma a=vi nyakoo-n ma a=uti li=naamuu-n, ka In Fwe hman.\\ when finish \gl{def}.\gl{sg}=story-3\gl{sg}.\gl{poss} \gl{def}.\gl{sg}=chief 3\gl{sg}=call for skin banyan \gl{com} 3\gl{sg}=say for-3\gl{sg}.\gl{poss} \gl{subr} 3\gl{sg}=take.off \gl{def}.\gl{pl}=clothing-3\gl{sg}.\gl{poss} \gl{cnj} skin figtree also\\ \glt \qu{Once she finishes her story the chief calls Banyan Bark to tell her to take off her clothes, and to Figtree Bark as well.}
\z

\ea
\gll A=ilake nyasi li=been apuli ma le=vacuti yeen-xhwâi, ma le=siiva In Thu hnya-an koo-n.\\ 3\gl{sg}=ask for \gl{def}.\gl{pl}=other person \gl{subr} 3\gl{pl}=raise tree-dance \gl{subr} 3\gl{pl}=tie skin banyan put-go on-\gl{ana}\\ \glt \qu{He asks the others to raise the dancing pole, to tie Banyan Bark to it.}
\z

\ea
\gll A=ilake nyasi li=apuli-n-ea ma le=hnya li=naamuu-n In Thu nyasi In Fwe.\\ 3\gl{sg}=ask for \gl{def}.\gl{pl}=person-\gl{link}-3\gl{sg}.\gl{poss} \gl{subr} 3\gl{pl}=put \gl{def}.\gl{pl}=clothing-\gl{link} skin banyan for skin figtree\\ \glt \qu{He asks his people to put Banyan Bark's clothes on Figtree Bark.}
\z

\ea
\gll Ca i=wadan cahma In Fwe a=hnya li=mween a= xahnang koo-n ka a=tabo cela i=daahma,\\ 
     in \gl{def}.\gl{sg}=time whereas skin figtree 3\gl{sg}=send \gl{def}.\gl{pl}=clothing \gl{rel}= good on-3\gl{sg} \gl{cnj} 3\gl{sg}=sit next.to \gl{def}.\gl{sg}=chief\\ 
\glt \qu{At that moment, Figtree Bark put on the good clothes and sat next to the chief, ...}
\z

\ea
\gll a=ilake nyasi li=apuli ma le=cini In Thu, ma a=vwa suki li=gat-ea.\\ 
     3\gl{sg}=ask for \gl{def}.\gl{pl}=person \gl{subr} 3\gl{pl}=burn skin banyan \gl{subr} 3\gl{sg}=do price \gl{def}.\gl{pl}=lie-3\gl{sg}.\gl{poss}\\ 
     \glt \qu{...who asked the people to burn Banyan Bark, so that she may pay for her lies.}
\z

\ea
\gll Cala In Thu a=kon e-sunixhiit, a=ilake ka i=daahma nyasi mu=xhwaawe ma lu=fe li=sau i=inya-lu ma lu=thake nya-koo-n ma lu=vila bitake i=yee-n vila.\\ when skin banyan 3\gl{sg}=\gl{prog} \gl{mid}-suffer 3\gl{sg}=ask \gl{sbj} \gl{def}.\gl{sg}=chief for \gl{def}.\gl{du}=child \gl{subr} 3\gl{du}=take \gl{def}.\gl{pl}=clothing \gl{def}.\gl{sg}=mother-3\gl{du}.\gl{poss} \gl{subr} throw put-on-3\gl{sg} \gl{subr} 3\gl{du}=dance go.around \gl{def}.\gl{sg}=tree-\gl{link} dance\\ \glt \qu{While Banyan Bark is suffering, the chief asks the two children to fetch their mother's dress, to dress her and to dance around the dancing pole.}
\z

\ea
\gll Ka jahngan i=bwen, meeka li=apuli le=vila i=ngeein ma i=cada bitake i=yee-n vila hma-san ca i=wadan a=me ka In Thu.\\ \gl{cnj} length \gl{def}.\gl{sg}=night all \gl{def}.\gl{pl}=person 3\gl{sg}=dance \gl{def}.\gl{sg}=rattle \gl{com} \gl{def}.\gl{sg}=beat go.around \gl{def}.\gl{sg}=tree-\gl{link} dance arrive-go in \gl{def}.\gl{sg}=time \gl{rel}=dance \gl{sbj} skin banyan\\ \glt \qu{And all night long, all the people danced the Rattle and the Beat around the dancing pole until Banyan Bark had died.}
\z
