\documentclass[ number=1
,series=labphon
,output=long
% ,draftmode
,url=http://langsci-press.org/catalog/book/16
,isbn=978-3-944675-01-5
]{LSP/langsci}
\usepackage{graphicx}	% image insertion
\usepackage{float}	% figure positioning
\usepackage{lscape}	% table/figure rotation
\usepackage{multirow}	% specific kind of tables
\usepackage[british,italian]{babel}	% hyphenation
\usepackage{LSP/lsp-styles/lsp-gb4e}	% leipzig glosses
\usepackage{setspace}	% and following: epigraphs
\renewcommand\epigraph[4]{
\vspace{1em}\hfill{}\begin{minipage}{#1}{\begin{spacing}{0.9}
\noindent\textit{#2}\end{spacing}
\vspace{1em}\small\hfill{}{#3}\\
\vspace{-2em}\begin{flushright}{#4}\end{flushright}}\vspace{2em}
\end{minipage}}
\title{Prosodic detail in Neapolitan Italian}
\author{Francesco Cangemi}
\typesetter{Francesco Cangemi}
\proofreader{Tom Gardner, Martin Hilpert, Michelle Natolo, Stephanie Natolo, Benedikt Singpiel, Siri Tuttle, Tamara Schmidt}
\BackTitle{Prosodic detail in Neapolitan Italian}
\BackBody{Recent findings on phonetic detail have been taken as supporting exemplar-based approaches to prosody. Through four experiments on both production and perception of both melodic and temporal detail in Neapolitan Italian, we show that prosodic detail is not incompatible with abstractionist approaches either. Specifically, we suggest that the exploration of prosodic detail leads to a refined understanding of the relationships between the richly specified and continuously varying phonetic information on one side, and coarse phonologically structured contrasts on the other, thus offering insights on how pragmatic information is conveyed by prosody.}
\lccode`\'=`\'
\hyphenation{
D'Im-pe-rio
}
\usepackage{booktabs}
\usepackage{floatrow}
\floatsetup[table]{capposition=top}
\newcommand{\mytoprule}{\midrule\toprule}
\newcommand{\mybottomrule}{\bottomrule\midrule}
\begin{document}
\selectlanguage{british}
\maketitle
\tableofcontents\enlargethispage{2em}
\mainmatter
\include{chapters/chapter1}
\include{chapters/chapter2}
\include{chapters/chapter3}
\include{chapters/chapter4}
\include{chapters/chapter5}
\include{chapters/chapter6}
\backmatter
\ohead{Bibliography}


%\bibliography{biblio}

\begin{thebibliography}{284}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{#1}
\providecommand{\urlprefix}{}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
  \urlstyle{rm}\Url}\fi

\bibitem[{Abercrombie(1967)}]{abercrombie1967elements}
Abercrombie, David. 1967.
\newblock \emph{Elements of general phonetics}.
\newblock Edinburgh: Edinburgh University Press.

\bibitem[{Adjarian(1899)}]{adjarian1899explosives}
Adjarian, Hrachia. 1899.
\newblock Les explosives de l'ancien arménien étudiées dans les dialectes
  modernes.
\newblock \emph{La Parole: Revue internationale de Rhinologie, Otologie,
  Laryngologie et Phonétique expérimentale} 119--127.

\bibitem[{{Albano Leoni}(2006)}]{albanoleoni2006statuto}
{Albano Leoni}, Federico. 2006.
\newblock Lo statuto del fonema.
\newblock In Stefano Gensini \& Martone Arturo (eds.), \emph{Il linguaggio:
  {T}eorie e storia delle teorie. {I}n onore di {L}ia {F}ormigari}, 281--303.
  Napoli: Liguori.

\bibitem[{Anderson et~al.(1984)Anderson, Pierrehumbert \&
  Liberman}]{anderson1984synthesis}
Anderson, Mark, Janet Pierrehumbert \& Mark Liberman. 1984.
\newblock Synthesis by rule of {E}nglish intonation patterns.
\newblock In \emph{Proceedings of 9th {I}nternational {C}onference of
  {A}coustics, {S}peech and {S}ignal {P}rocessing}, vol.~9, 77--80. San Diego.

\bibitem[{Andr\'{e} et~al.(2003)Andr\'{e}, Ghio, Cav\'{e} \&
  Teston}]{andre2003perceval}
Andr\'{e}, Carine, Alain Ghio, Christian Cav\'{e} \& Bernard Teston. 2003.
\newblock Perceval: {A} computer-driven system for experimentation on auditory
  and visual perception.
\newblock In Daniel Recasens, Maria~Josep Sol\'{e} \& Joaquín Romero (eds.),
  \emph{Proceedings of the 15th {I}nternational {C}ongress of {P}honetic
  {S}ciences}, 1421--1424. Barcelona.

\bibitem[{Angelini et~al.(1993)Angelini, Brugnara, Falavigna, Giuliani, Gretter
  \& Omologo}]{angelini1993baseline}
Angelini, Bianca, Fabio Brugnara, Daniele Falavigna, Diegl Giuliani, Roberto
  Gretter \& Maurizio Omologo. 1993.
\newblock A baseline of a speaker independent continuous speech recognizer of
  {I}talian.
\newblock In \emph{Proceedings of the 3rd {European Conference on Speech
  Communication and Technology}}, 847--850. Berlin.

\bibitem[{Arndt(1960)}]{arndt1960modal}
Arndt, Walter. 1960.
\newblock Modal particles in {R}ussian and {G}erman.
\newblock \emph{Word} 16. 323--336.

\bibitem[{Atterer \& Ladd(2004)}]{atterer2004phonetics}
Atterer, Michaela \& Robert Ladd. 2004.
\newblock On the phonetics and phonology of segmental anchoring of {F}0:
  {E}vidence from {G}erman.
\newblock \emph{Journal of Phonetics} 32. 177--197.

\bibitem[{{Audacity Development Team}(2006)}]{audacity2006audacity}
{Audacity Development Team}. 2006.
\newblock Audacity: {F}ree audio editor and recorder.
\newblock Computer program, retrieved from http://audacity.sourceforge.net/.

\bibitem[{Avesani(1990)}]{avesani1990contribution}
Avesani, Cinzia. 1990.
\newblock A contribution to the synthesis of {I}talian intonation.
\newblock In \emph{Proceedings of the 1st {International Conference on Spoken
  Language Processing}}, 833--836. Kobe.

\bibitem[{Balota(1994)}]{balota1994visual}
Balota, David. 1994.
\newblock Visual word recognition.
\newblock In Matthew Traxler \& Morton Gernsbacher (eds.), \emph{Handbook of
  psycholinguistics}, 334--357. San Diego: Academic Press.

\bibitem[{Batliner \& M{\"o}bius(2006)}]{batliner2005prosodic}
Batliner, Anton \& Batliner M{\"o}bius. 2006.
\newblock Prosodic models, automatic speech understanding, and speech
  synthesis: {T}owards the common ground?
\newblock In William Barry, Wim {van Dommelen} \& Jacques Koreman (eds.),
  \emph{The integration of phonetic knowledge in speech technology}, 21--44.
  Dordrecht: Kluwer.

\bibitem[{Batliner et~al.(2001)Batliner, M{\"o}bius, M{\"o}hler, Schweitzer \&
  N{\"o}th}]{batliner2001prosodic}
Batliner, Anton, Bernd M{\"o}bius, Gregor M{\"o}hler, Antje Schweitzer \& Elmar
  N{\"o}th. 2001.
\newblock Prosodic models, automatic speech understanding, and speech
  synthesis: {T}owards the common ground.
\newblock In Paul Dalsgaard, Børge Lindberg \& Henrik Benner (eds.),
  \emph{Proceedings of the 7th {European Conference on Speech Communication and
  Technology}}, vol.~4, 2285--2288. Aalborg.

\bibitem[{Beckman(1996)}]{beckman1996parsing}
Beckman, Mary. 1996.
\newblock The parsing of prosody.
\newblock \emph{Language and Cognitive Processes} 11(1-2). 17--68.

\bibitem[{Beckman(1997)}]{beckman1997typology}
Beckman, Mary. 1997.
\newblock A typology of spontaneous speech.
\newblock In Yoshinori Sagisaka, Nick Campbell \& Norio Higuchi (eds.),
  \emph{Computing prosody: {C}omputational models for processing spontaneous
  speech}, 7--26. Dordrecht, Heidelberg, London, New York: Springer.

\bibitem[{Bigi \& Hirst(2012)}]{bigi2012speech}
Bigi, Brigitte \& Daniel Hirst. 2012.
\newblock {SP}eech {P}honetization {A}lignment and {S}yllabification ({SPPAS}):
  {A} tool for the automatic analysis of speech prosody.
\newblock In Qiuwu Ma, Hongwei Ding \& Daniel Hirst (eds.), \emph{Proceedings
  of the 5th {I}nternational {C}onference on {S}peech {P}rosody}, vol.~1,
  19--22. Shanghai: Tongji University Press.

\bibitem[{Black \& Hunt(1996)}]{black1996generating}
Black, Alan \& Andrew Hunt. 1996.
\newblock Generating f0 contours from {ToBI} labels using linear regression.
\newblock In \emph{Proceedings of the 4th {International Conference on Spoken
  Language Processing}}, vol.~3, 1385--1388. Philadelphia.

\bibitem[{Blesser(1972)}]{blesser1972speech}
Blesser, Barry. 1972.
\newblock Speech perception under conditions of spectral transformation: {I}.
  {P}honetic characteristics.
\newblock \emph{Journal of Speech and Hearing Research} 15(1). 5--41.

\bibitem[{Boersma \& Weenink(2008)}]{boersma2008praat}
Boersma, Paul \& David Weenink. 2008.
\newblock Praat: {D}oing phonetics by computer.
\newblock Computer program, retrieved from http://www.praat.org/.

\bibitem[{Bolinger(1951)}]{bolinger1951intonation}
Bolinger, Dwight. 1951.
\newblock Intonation: {L}evels versus configurations.
\newblock \emph{Word} 7. 199--210.

\bibitem[{Bolinger(1964)}]{bolinger1964intonation}
Bolinger, Dwight. 1964.
\newblock Intonation as a universal.
\newblock In Horace Lunt (ed.), \emph{Proceedings of the 9th {I}nternational
  {C}ongress of {L}inguists}, 833--848. The Hague: Mouton.

\bibitem[{Bolinger(1989)}]{bolinger1989intonation}
Bolinger, Dwight. 1989.
\newblock \emph{Intonation and its uses: {M}elody in grammar and discourse}.
\newblock Palo Alto: Stanford University Press.

\bibitem[{Brooks(1978)}]{brooks1978nonanalytic}
Brooks, Lee. 1978.
\newblock Nonanalytic concept formation and memory for instances.
\newblock In Eleanor Rosch \& Barbara Lloyd (eds.), \emph{Cognition and
  categorization}, 170--211. Hillsdale: Erlbaum.

\enlargethispage{\baselineskip}
\bibitem[{Browman \& Goldstein(1986)}]{browman1986articulatory}
Browman, Catherine \& Louis Goldstein. 1986.
\newblock Towards an articulatory phonology.
\newblock \emph{Phonology Yearbook} 3. 219--252.

\bibitem[{Brunetti et~al.(2010)Brunetti, D'Imperio \&
  Cangemi}]{brunetti2010prosodic}
Brunetti, Lisa, Mariapaola D'Imperio \& Francesco Cangemi. 2010.
\newblock On the prosodic marking of contrast in {R}omance sentence topic:
  {E}vidence from {N}eapolitan {I}talian.
\newblock In \emph{Proceedings of the 5th {I}nternational {C}onference on
  {S}peech {P}rosody}, Chicago.

\bibitem[{Bruni(1992)}]{bruni1992italiano}
Bruni, Francesco. 1992.
\newblock \emph{L'italiano nelle regioni. {L}ingua nazionale e identità
  regionali}.
\newblock Torino: Utet.

\bibitem[{B\"{u}ring(1997)}]{buring1997meaning}
B\"{u}ring, Daniel. 1997.
\newblock \emph{The meaning of topic and focus: {T}he 59th street bridge
  accent}.
\newblock London, New York: Routledge.

\bibitem[{Bybee(2001)}]{bybee2001phonology}
Bybee, Joan. 2001.
\newblock \emph{Phonology and language use}.
\newblock Cambridge: Cambridge University Press.

\bibitem[{Bybee(2006)}]{bybee2006usage}
Bybee, Joan. 2006.
\newblock From usage to grammar: {T}he mind's response to repetition.
\newblock \emph{Language} 82(4). 711--733.

\bibitem[{Campbell \& Mokhtari(2003)}]{campbell2003voice}
Campbell, Nick \& Parham Mokhtari. 2003.
\newblock Voice quality: {T}he 4th prosodic dimension.
\newblock In Daniel Recasens, Maria~Josep Sol\'{e} \& Joaquín Romero (eds.),
  \emph{Proceedings of the 15th {I}nternational {C}ongress of {P}honetic
  {S}ciences}, 2417--2420. Barcelona.

\bibitem[{Cangemi(2009)}]{cangemi2009phonetic}
Cangemi, Francesco. 2009.
\newblock Phonetic detail in intonation contour dynamics.
\newblock In Stephan Schmid, Michael Schwarzenbach \& Dieter Studer (eds.),
  \emph{La dimensione temporale del parlato: {P}roceedings of the 5th
  {C}onference of {A}ssociazione {I}taliana di {S}cienze della {V}oce},
  325--334. Torriana: EDK.

\bibitem[{Cangemi et~al.(2011)Cangemi, Cutugno, Ludusan, Seppi \&
  Van~Compernolle}]{cangemi2011automatic}
Cangemi, Francesco, Francesco Cutugno, Bogdan Ludusan, Dino Seppi \& Dirk
  Van~Compernolle. 2011.
\newblock Automatic {S}peech {S}egmentation for {I}talian ({ASSI}): {T}ools,
  models, evaluation and application.
\newblock In Barbara Gili~Fivela, Antonio Stella, Luigia Garrapa \& Mirko
  Grimaldi (eds.), \emph{Contesto comunicativo e variabilità nella produzione
  e percezione della lingua: {P}roceedings of the 7th {C}onference of
  {A}ssociazione {I}taliana di {S}cienze della {V}oce}, Roma: Bulzoni.

\bibitem[{Cangemi \& D'Imperio(2011{\natexlab{a}})}]{cangemi2011local}
Cangemi, Francesco \& Mariapaola D'Imperio. 2011{\natexlab{a}}.
\newblock Local speech rate differences between questions and statements in
  italian.
\newblock In Wai-Sum Lee \& Eric Zee (eds.), \emph{Proceedings of the 17th
  {I}nternational {C}ongress of {P}honetic {S}ciences}, 392--395. Hong Kong:
  City University of Hong Kong.

\bibitem[{Cangemi \& D'Imperio(2011{\natexlab{b}})}]{cangemi2011prosodia}
Cangemi, Francesco \& Mariapaola D'Imperio. 2011{\natexlab{b}}.
\newblock Prosodia oltre la f0: {T}empo e modalità.
\newblock In Barbara Gili~Fivela, Antonio Stella, Luigia Garrapa \& Mirko
  Grimaldi (eds.), \emph{Contesto comunicativo e variabilità nella produzione
  e percezione della lingua: {P}roceedings of the 7th {C}onference of
  {A}ssociazione {I}taliana di {S}cienze della {V}oce}, Roma: Bulzoni.

\bibitem[{Cangemi \& D'Imperio(2013)}]{cangemiFORTHtempo}
Cangemi, Francesco \& Mariapaola D'Imperio. 2013.
\newblock Tempo and the perception of sentence modality.
\newblock \emph{Laboratory Phonology} 4(1). 191--219.

\enlargethispage{\baselineskip}
\bibitem[{Cangemi \& D'Imperio(forthcoming)}]{cangemiFORTHbeyond}
Cangemi, Francesco \& Mariapaola D'Imperio. forthcoming.
\newblock Beyond f0: {S}entence modality and speech rate.
\newblock In Joaquín Romero \& Maria Riera (eds.), \emph{Selected papers from
  the 5th {C}onference on {P}honetics and {P}honology in {I}beria}, Amsterdam:
  John Benjamins.

\bibitem[{Caputo(1994)}]{caputo1994intonazione}
Caputo, Maria~Rosaria. 1994.
\newblock L'intonazione delle domande s{\`i}/no in un campione di italiano
  parlato.
\newblock In \emph{Proceedings of the 4th {Gruppo di Fonetica Sperimentale
  Workshop}}, 9--18. Torino.

\bibitem[{Caputo(1996)}]{caputo1996presupposizione}
Caputo, Maria~Rosaria. 1996.
\newblock Presupposizione, fuoco, modalità e schemi melodici.
\newblock In \emph{Proceedings of the 24th {National Congress of Associazione
  Italiana di Acustica}}, 49--54. Trento.

\bibitem[{Caputo \& D'Imperio(1995)}]{caputo1995possibile}
Caputo, Maria~Rosaria \& Mariapaola D'Imperio. 1995.
\newblock Verso un possibile sistema di trascrizione prosodica dell’italiano:
  {C}enni preliminari.
\newblock In \emph{Proceedings of the 5th workshop of {Gruppo di Fonetica
  Sperimentale}}, 71--83. Trento.

\bibitem[{Charles-Luce(1985)}]{charlesluce1985word}
Charles-Luce, Jan. 1985.
\newblock Word-final devoicing in {G}erman and the effects of phonetic and
  sentential contexts.
\newblock \emph{Journal of Phonetics} 13. 309--324.

\bibitem[{Charles-Luce \& Dinnsen(1987)}]{charlesluce1987reanalysis}
Charles-Luce, Jan \& Daniel Dinnsen. 1987.
\newblock A reanalysis of {C}atalan devoicing.
\newblock \emph{Journal of Phonetics} 15(2). 187--190.

\bibitem[{Chomsky(1965)}]{chomsky1965aspects}
Chomsky, Noam. 1965.
\newblock \emph{Aspects of the theory of syntax}.
\newblock Cambridge: MIT Press.

\bibitem[{Church \& Schacter(1994)}]{church1994perceptual}
Church, Barbara \& Daniel Schacter. 1994.
\newblock Perceptual specificity of auditory priming: {I}mplicit memory for
  voice intonation and fundamental frequency.
\newblock \emph{Journal of Experimental Psychology: Learning, Memory, and
  Cognition} 20(3). 521--533.

\bibitem[{Cole \& Shattuck-Hufnagel(2011)}]{cole2011phonology}
Cole, Jennifer \& Stefanie Shattuck-Hufnagel. 2011.
\newblock The phonology and phonetics of perceived prosody: {W}hat do listeners
  imitate?
\newblock In \emph{Proceedings of the 12th {Annual Conference of the
  International Speech Communication Association}}, 969--972. Firenze.

\bibitem[{Coleman(2003)}]{coleman2003discovering}
Coleman, John. 2003.
\newblock Discovering the acoustic correlates of phonological contrasts.
\newblock \emph{Journal of Phonetics} 31(3-4). 351--372.

\bibitem[{Cooper et~al.(1952)Cooper, Delattre, Liberman, Borst \&
  Gerstman}]{cooper1952experiments}
Cooper, Franklin, Pierre Delattre, Alvin Liberman, John Borst \& Louis
  Gerstman. 1952.
\newblock Some experiments on the perception of synthetic speech sounds.
\newblock \emph{Journal of the Acoustical Society of America} 24(6). 597--606.

\bibitem[{Cooper \& Paccia-Cooper(1980)}]{cooper1980syntax}
Cooper, William \& Jeanne Paccia-Cooper. 1980.
\newblock \emph{Syntax and speech}.
\newblock Cambridge: Harvard University Press.

\bibitem[{Dahan et~al.(2002)Dahan, Tanenhaus \& Chambers}]{dahan2002accent}
Dahan, Delphine, Michael Tanenhaus \& Craig Chambers. 2002.
\newblock Accent and reference resolution in spoken-language comprehension.
\newblock \emph{Journal of Memory and Language} 47(2). 292--314.

\bibitem[{De~Dominicis(2010)}]{dedominicis2010interrogative}
De~Dominicis, Amedeo. 2010.
\newblock Interrogative e assertive in un corpus dialettale recuperato
  ({B}omarzo).
\newblock In Francesco Cutugno, Pietro Maturi, Renata Savy, Giovanni Abete \&
  Iolanda Alfano (eds.), \emph{Parlare con le persone, parlare alle macchine:
  {L}a dimensione interazionale della comunicazione verbale: {P}roceedings of
  the 6th {C}onference of {A}ssociazione {I}taliana di {S}cienze della {V}oce},
  Torriana: EDK.


\enlargethispage{2\baselineskip}
\bibitem[{De~Mauro(1970)}]{demauro1970storia}
De~Mauro, Tullio. 1970.
\newblock \emph{Storia linguistica dell'{I}talia unita (nuova edizione)}.
\newblock Laterza: Laterza.

\bibitem[{Del~Giudice et~al.(2007)Del~Giudice, Shosted, Davidson, Salihie \&
  Arvaniti}]{delgiudice2007comparing}
Del~Giudice, Alex, Ryan Shosted, Kathryn Davidson, Mohammad Salihie \& Amalia
  Arvaniti. 2007.
\newblock Comparing methods for locating pitch ``elbows''.
\newblock In Jürgen Trouvain \& William Barry (eds.), \emph{Proceedings of the
  16th {I}nternational {C}ongress of {P}honetic {S}ciences}, 1117--1120.
  Saarbr\"{u}cken.

\bibitem[{Delattre(1966)}]{delattre1966dix}
Delattre, Pierre. 1966.
\newblock Les dix intonations de base du fran\c{c}ais.
\newblock \emph{The French Review} 40(1). 1--14.

\bibitem[{Delattre et~al.(1955)Delattre, Liberman \&
  Cooper}]{delattre1955acoustic}
Delattre, Pierre, Alvin Liberman \& Franklin Cooper. 1955.
\newblock Acoustic loci and transitional cues for consonants.
\newblock \emph{Journal of the Acoustical Society of America} 27(4). 769--773.

\bibitem[{D'Imperio(1995)}]{dimperio1995timing}
D'Imperio, Mariapaola. 1995.
\newblock Timing differences between prenuclear and nuclear pitch accents in
  {I}talian.
\newblock \emph{Journal of the Acoustical Society of America} 98(5). 2894.

\bibitem[{D'Imperio(1996)}]{dimperio1996caratteristiche}
D'Imperio, Mariapaola. 1996.
\newblock Caratteristiche di timing degli accenti nucleari in parlato italiano
  letto.
\newblock In \emph{Proceedings of the 24th {National Congress of Associazione
  Italiana di Acustica}}, 55--60. Trento.

\bibitem[{D'Imperio(1997{\natexlab{a}})}]{dimperio1997breadth}
D'Imperio, Mariapaola. 1997{\natexlab{a}}.
\newblock Breadth of focus, modality, and prominence perception in {N}eapolitan
  {I}talian.
\newblock \emph{Working Papers in Linguistics -- Ohio State University} 50.
  19--39.

\bibitem[{D'Imperio(1997{\natexlab{b}})}]{dimperio1997narrow}
D'Imperio, Mariapaola. 1997{\natexlab{b}}.
\newblock Narrow focus and focal accent in the {N}eapolitan variety of
  {I}talian.
\newblock In Antonis Botinis, Georgios Kouroupetroglou \& George Carayiannis
  (eds.), \emph{Intonation: {T}heory, models and applications. {Proceedings of
  an European Conference on Speech Communication and Technology Workshop}},
  87--90. Athens.

\bibitem[{D'Imperio(1999)}]{dimperio1999tonal}
D'Imperio, Mariapaola. 1999.
\newblock Tonal structure and pitch targets in {I}talian focus constituents.
\newblock In John Ohala (ed.), \emph{Proceedings of the 14th {I}nternational
  {C}ongress of {P}honetic {S}ciences}, 1757--1760. San Francisco: University
  of California.

\bibitem[{D'Imperio(2000)}]{dimperio2000role}
D'Imperio, Mariapaola. 2000.
\newblock \emph{The role of perception in defining tonal targets and their
  alignment}: Columbus: The Ohio State University dissertation.

\bibitem[{D'Imperio(2001)}]{dimperio2001focus}
D'Imperio, Mariapaola. 2001.
\newblock Focus and tonal structure in neapolitan italian.
\newblock \emph{Speech Communication} 33(4). 339--356.

\bibitem[{D'Imperio(2002)}]{dimperio2002italian}
D'Imperio, Mariapaola. 2002.
\newblock Italian intonation: {A}n overview and some questions.
\newblock \emph{Probus} 14(1). 37--69.

\bibitem[{D'Imperio(2003)}]{dimperio2003tonal}
D'Imperio, Mariapaola. 2003.
\newblock Tonal structure and pitch targets in {I}talian focus constituents.
\newblock \emph{Catalan Journal of Linguistics} 2. 55--65.


%%\enlargethispage{\baselineskip}
\bibitem[{D'Imperio \& Cangemi(2009)}]{dimperio2009interplay}
D'Imperio, Mariapaola \& Francesco Cangemi. 2009.
\newblock The interplay between tonal alignment and rise shape in the
  perception of two {N}eapolitan rising accents.
\newblock Talk presented at the 4th Conference on Phonetics and Phonology in
  Iberia, Gran Canaria, Spain.

\enlargethispage{\baselineskip}
\bibitem[{D'Imperio \& Cangemi(2011)}]{dimperio2011phrasing}
D'Imperio, Mariapaola \& Francesco Cangemi. 2011.
\newblock Phrasing, register level downstep and partial topic constructions in
  {N}eapolitan {I}talian.
\newblock In Christoph Gabriel \& Conxita Lle\'{o} (eds.), \emph{Intonational
  phrasing in {R}omance and {G}ermanic: {C}ross-linguistic and bilingual
  studies}, 75--94. Amsterdam: John Benjamins.

\bibitem[{D'Imperio et~al.(2008)D'Imperio, Cangemi \&
  Brunetti}]{dimperio2008phonetics}
D'Imperio, Mariapaola, Francesco Cangemi \& Lisa Brunetti. 2008.
\newblock The phonetics and phonology of contrastive topic constructions in
  {I}talian.
\newblock Poster presented at the 3rd Conference on Tone and Intonation in
  Europe, Lisbon, Portugal.

\bibitem[{D'Imperio et~al.(2005)D'Imperio, Elordieta, Frota, Prieto \&
  Vig\`{a}rio}]{dimperio2005intonational}
D'Imperio, Mariapaola, Gorka Elordieta, Sónia Frota, Pilar Prieto \& Marina
  Vig\`{a}rio. 2005.
\newblock Intonational phrasing in {R}omance: {T}he role of syntactic and
  prosodic structure.
\newblock In Sónia Frota, Marina Vig\`{a}rio \& Maria Freitas (eds.),
  \emph{Prosodies}, 59--97. Berlin, New York: Mouton de Gruyter.

\bibitem[{D'Imperio \& Gili~Fivela(2003)}]{dimperio2003levels}
D'Imperio, Mariapaola \& Barbara Gili~Fivela. 2003.
\newblock How many levels of phrasing? {E}vidence from two varieties of
  {I}talian.
\newblock In John Local, Richard Ogden \& Rosalind Temple (eds.), \emph{Papers
  in {L}aboratory {P}honology}, vol.~6, 38--57. Cambridge: Cambridge University
  Press.

\bibitem[{D'Imperio \& House(1997)}]{dimperio1997perception}
D'Imperio, Mariapaola \& David House. 1997.
\newblock Perception of questions and statements in {N}eapolitan {I}talian.
\newblock In George Kokkinakis, Nikos Fakotakis \& Evangelos Dermatas (eds.),
  \emph{Proceedings of the 5th {European Conference on Speech Communication and
  Technology}}, 251--254. Rhodes.

\bibitem[{D'Imperio et~al.(2007)D'Imperio, Petrone \&
  Nguyen}]{dimperio2007effects}
D'Imperio, Mariapaola, Caterina Petrone \& Noël Nguyen. 2007.
\newblock Effects of tonal alignment on lexical identification in {I}talian.
\newblock In Tomas Riad \& Carlos Gussenhoven (eds.), \emph{Tones and tunes:
  {E}xperimental studies in word and sentence prosody}, vol.~2, 79--106.
  Berlin: de Gruyter.

\bibitem[{Dinnsen \& Charles-Luce(1984)}]{dinnsen1984phonological}
Dinnsen, Daniel \& Jan Charles-Luce. 1984.
\newblock Phonological neutralization, phonetic implementation and individual
  differences.
\newblock \emph{Journal of Phonetics} 12(1). 49--60.

\bibitem[{Dinnsen \& {Garcia Zamor}(1971)}]{dinnsen1971three}
Dinnsen, Daniel \& Maria {Garcia Zamor}. 1971.
\newblock The three degrees of vowel length in {G}erman.
\newblock \emph{Research on Language \& Social Interaction} 4(1). 111--126.

\bibitem[{Dmitrieva et~al.(2010)Dmitrieva, Jongman \&
  Sereno}]{dmitrieva2010phonological}
Dmitrieva, Olga, Allard Jongman \& Joan Sereno. 2010.
\newblock Phonological neutralization by native and non-native speakers: {T}he
  case of {R}ussian final devoicing.
\newblock \emph{Journal of Phonetics} 38(3). 483--492.

\bibitem[{Dombrowski \& Niebuhr(2005)}]{dombrowski2005acoustic}
Dombrowski, Ernst \& Oliver Niebuhr. 2005.
\newblock Acoustic patterns and communicative functions of phrase-final f0
  rises in {G}erman: {A}ctivating and restricting contours.
\newblock \emph{Phonetica} 62(2-4). 176--195.

\bibitem[{Dryer(2011)}]{wals-2011-116}
Dryer, Matthew. 2011.
\newblock Polar questions.
\newblock In Matthew Dryer \& Martin Haspelmath (eds.), \emph{{The World Atlas
  of Language Structures Online}}, Munich: Max Planck Digital Library.

\bibitem[{Duncan(1972)}]{duncan1972signals}
Duncan, Starkey. 1972.
\newblock Some signals and rules for taking speaking turns in conversations.
\newblock \emph{Journal of Personality and Social Psychology} 23(2). 283--292.

\enlargethispage{\baselineskip}
\bibitem[{Eefting(1991)}]{eefting1991effect}
Eefting, Wieke. 1991.
\newblock The effect of ‘‘information value’’and
  ‘‘accentuation’’ on the duration of {D}utch words, syllables, and
  segments.
\newblock \emph{Journal of the Acoustical Society of America} 89(1). 412--424.

\bibitem[{Elman \& McClelland(1988)}]{elman1988cognitive}
Elman, Jeffrey \& James McClelland. 1988.
\newblock Cognitive penetration of the mechanisms of perception: {C}ompensation
  for coarticulation of lexically restored phonemes.
\newblock \emph{Journal of Memory and Language} 27(2). 143--165.

\bibitem[{Ernestus(2014)}]{ernestusacoustic}
Ernestus, Mirjam. 2014.
\newblock Acoustic reduction and the roles of abstractions and exemplars in
  speech processing.
\newblock \emph{Lingua} 142. 27--41.

\bibitem[{Ernestus \& Baayen(2006)}]{ernestus2006functionality}
Ernestus, Mirjam \& Harald Baayen. 2006.
\newblock The functionality of incomplete neutralization in {D}utch: {T}he case
  of past-tense formation.
\newblock In Louis Goldstein, Douglas Whalen \& Catherine Best (eds.),
  \emph{Papers in {L}aboratory {P}honology}, vol.~8, 27--49. Cambridge:
  Cambridge University Press.

\bibitem[{Faber(1992)}]{faber1992phonemic}
Faber, Alice. 1992.
\newblock Phonemic segmentation as epiphenomenon: {E}vidence from the history
  of alphabetic writing.
\newblock In Pamela Downing, Susan Lima \& Michael Noonan (eds.), \emph{The
  linguistics of literacy}, 111--134. Amsterdam: John Benjamins.

\bibitem[{Face(2001)}]{face2001focus}
Face, Timothy. 2001.
\newblock Focus and early peak alignment in {S}panish intonation.
\newblock \emph{Probus} 13(2). 223--246.

\bibitem[{Farnetani \& Kori(1991)}]{farnetani1990rhytmic}
Farnetani, Edda \& Shiro Kori. 1991.
\newblock Rhytmic structure in {I}talian noun phrases: {A} study on vowel
  durations.
\newblock \emph{Phonetica} 47. 50--65.

\bibitem[{Firth(1948)}]{firth1948prosodies}
Firth, John. 1948.
\newblock Sounds and prosodies.
\newblock \emph{Transactions of the Philological Society} 47(1). 127--152.

\bibitem[{Flemming(1997)}]{flemming1997phonetic}
Flemming, Edward. 1997.
\newblock Phonetic detail in phonology: {T}owards a unified account of
  assimilation and coarticulation.
\newblock In Keiichiro Suzuki \& Dirk Elzinga (eds.), \emph{Proceedings of the
  1995 {Southwestern Workshop in Optimality Theory (SWOT)}}, Tucson.

\bibitem[{Flemming(2001)}]{flemming2001scalar}
Flemming, Edward. 2001.
\newblock Scalar and categorical phenomena in a unified model of phonetics and
  phonology.
\newblock \emph{Phonology} 18(1). 7--44.

\bibitem[{Fourakis \& Iverson(1984)}]{fourakis1984incomplete}
Fourakis, Marios \& Gregory Iverson. 1984.
\newblock On the ‘incomplete neutralization’ of {G}erman final obstruents.
\newblock \emph{Phonetica} 41(3). 140--149.

\bibitem[{Frick(1995)}]{frick1995accepting}
Frick, Robert. 1995.
\newblock Accepting the null hypothesis.
\newblock \emph{Memory \& Cognition} 23(1). 132--138.

\bibitem[{Frota et~al.(2007)Frota, D'Imperio, Elordieta, Prieto \&
  Vig\`{a}rio}]{frota2007phonetics}
Frota, Sónia, Mariapaola D'Imperio, Gorka Elordieta, Pilar Prieto \& Marina
  Vig\`{a}rio. 2007.
\newblock The phonetics and phonology of intonational phrasing in {R}omance.
\newblock In Pilar Prieto, Joan Mascar\`{o} \& Maria~Josep Sol\'{e} (eds.),
  \emph{Segmental and prosodic issues in {R}omance phonology}, 131--154.
  Amsterdam: John Benjamins.

%\enlargethispage{\baselineskip}
\bibitem[{Fujisaki \& Hirose(1982)}]{fujisaki1982modelling}
Fujisaki, Hiroya \& Keikichi Hirose. 1982.
\newblock Modelling the dynamic characteristics of voice fundamental frequency
  with application to analysis and synthesis of intonation.
\newblock In Shiro Hattori \& Kazuko Inoue (eds.), \emph{Proceedings of 13th
  {International Congress of Linguists}}, 57--70. Tokyo.

\bibitem[{Gili~Fivela(2004)}]{gilifivela2004phonetics}
Gili~Fivela, Barbara. 2004.
\newblock \emph{The phonetics and phonology of intonation: {T}he case of {P}isa
  {I}talian}: Pisa: Scuola Normale Superiore dissertation.

\bibitem[{Gili~Fivela(2006)}]{gilifivela2006coding}
Gili~Fivela, Barbara. 2006.
\newblock The coding of target alignment and scaling in pitch accent
  transcription.
\newblock \emph{Italian Journal of Linguistics} 18(1). 189--221.

\bibitem[{Gili~Fivela(2008)}]{gilifivela2008intonation}
Gili~Fivela, Barbara. 2008.
\newblock \emph{Intonation in production and perception: {T}he case of {P}isa
  {I}talian}.
\newblock Alessandria: Edizioni dell'{O}rso.

\bibitem[{Gili~Fivela \& D'Imperio(2003)}]{gilifivela2003tonal}
Gili~Fivela, Barbara \& Mariapaola D'Imperio. 2003.
\newblock Tonal alignment of prenuclear accents in {I}talian.
\newblock Poster presented at the 2nd Conference on Tone and Intonation in
  Europe, Santorini, Greece.

\bibitem[{Goldinger(1996)}]{goldinger1996words}
Goldinger, Stephen. 1996.
\newblock Words and voices: {E}pisodic traces in spoken word identification and
  recognition memory.
\newblock \emph{Journal of Experimental Psychology: Learning, Memory, and
  Cognition} 22(5). 1166--1183.

\bibitem[{Goldinger et~al.(1991)Goldinger, Pisoni \&
  Logan}]{goldinger1991nature}
Goldinger, Stephen, David Pisoni \& John Logan. 1991.
\newblock On the nature of talker variability effects on recall of spoken word
  lists.
\newblock \emph{Journal of Experimental Psychology: Learning, Memory, and
  Cognition} 17(1). 152--162.

\bibitem[{Goldman(2011)}]{goldman2011easyalign}
Goldman, Jean-Philippe. 2011.
\newblock Easy{A}lign: {A}n automatic phonetic alignment tool under {P}raat.
\newblock In \emph{Proceedings of the 12th {Annual Conference of the
  International Speech Communication Association}}, Firenze.

\bibitem[{Grice(1991)}]{grice1991intonation}
Grice, Martine. 1991.
\newblock The intonation of interrogation in two varieties of {S}icilian
  {I}talian.
\newblock In \emph{Proceedings of the 12th {I}nternational {C}ongress of
  {P}honetic {S}ciences}, vol.~5, 210--213. Aix-en-Provence.

\bibitem[{Grice(1995)}]{grice1995intonation}
Grice, Martine. 1995.
\newblock \emph{The intonation of interrogation in {P}alermo {I}talian:
  {I}mplications for intonation theory}.
\newblock T\"{u}bingen: Niemeyer.

\bibitem[{Grice \& Baumann(2002)}]{grice2002deutsche}
Grice, Martine \& Stefan Baumann. 2002.
\newblock Deutsche {I}ntonation und {GToBI}.
\newblock \emph{Linguistische Berichte} 191. 267--298.

\bibitem[{Grice et~al.(2005{\natexlab{a}})Grice, Baumann \&
  Benzm\"{u}lller}]{grice2005german}
Grice, Martine, Stefan Baumann \& Ralf Benzm\"{u}lller. 2005{\natexlab{a}}.
\newblock German intonation in autosegmental-metrical phonology.
\newblock In Sun-Ah Jun (ed.), \emph{Prosodic typology: {T}he phonology of
  intonation and phrasing}, 55--83. Oxford: Oxford University Press.

\bibitem[{Grice et~al.(2005{\natexlab{b}})Grice, D'Imperio, Savino \&
  Avesani}]{grice2005strategy}
Grice, Martine, Mariapaola D'Imperio, Michelina Savino \& Cinzia Avesani.
  2005{\natexlab{b}}.
\newblock Towards a strategy for labelling varieties of italian.
\newblock In Sun-Ah Jun (ed.), \emph{Prosodic typology: {T}he phonology of
  intonation and phrasing}, 362--389. Oxford: Oxford University Press.

\bibitem[{Grice et~al.(2000)Grice, Ladd \& Arvaniti}]{grice2000place}
Grice, Martine, Robert Ladd \& Amalia Arvaniti. 2000.
\newblock On the place of phrase accents in intonational phonology.
\newblock \emph{Phonology} 17(2). 143--185.

\bibitem[{Gubian et~al.(2010)Gubian, Cangemi \& Boves}]{gubian2010automatic}
Gubian, Michele, Francesco Cangemi \& Lou Boves. 2010.
\newblock Automatic and data driven pitch contour manipulation with functional
  data analysis.
\newblock In \emph{Proceedings of the 5th {I}nternational {C}onference on
  {S}peech {P}rosody}, Chicago.

\enlargethispage{\baselineskip}
\bibitem[{Gubian et~al.(2011)Gubian, Cangemi \& Boves}]{gubian2011joint}
Gubian, Michele, Francesco Cangemi \& Lou Boves. 2011.
\newblock Joint analysis of f0 and speech rate with functional data analysis.
\newblock In \emph{Proceedings of 36th {I}nternational {C}onference of
  {A}coustics, {S}peech and {S}ignal {P}rocessing}, 4972--4975. Prague.

\bibitem[{Gussenhoven(1984)}]{gussenhoven1984grammar}
Gussenhoven, Carlos. 1984.
\newblock \emph{On the grammar and semantics of sentence accents}.
\newblock Dordrecht: Foris.

\bibitem[{Gussenhoven(2004)}]{gussenhoven2004phonology}
Gussenhoven, Carlos. 2004.
\newblock \emph{The phonology of tone and intonation}.
\newblock Cambridge: Cambridge University Press.

\bibitem[{Gussenhoven(2006)}]{gussenhoven2006experimental}
Gussenhoven, Carlos. 2006.
\newblock Experimental approaches to establishing discreteness of intonational
  contrasts.
\newblock In Stefan Sudhoff, Denisa Lenertov\'{a}, Roland Meyer, Sandra
  Pappert, Petra Augurzky, Ina Mleinek, Nicole Richter \& Johannes
  Schlie\ss{}er (eds.), \emph{Methods in empirical prosody research}, 321--334.
  Berlin: De Gruyter.

\bibitem[{Haan(2002)}]{haan2002speaking}
Haan, Judith. 2002.
\newblock \emph{Speaking of questions}.
\newblock Utrecht: LOT.

\bibitem[{Harris(1955)}]{harris1955phoneme}
Harris, Zellig. 1955.
\newblock From phoneme to morpheme.
\newblock \emph{Language} 31(2). 190--222.

\bibitem[{Hawkins(2003)}]{hawkins2003roles}
Hawkins, Sarah. 2003.
\newblock Roles and representations of systematic fine phonetic detail in
  speech understanding.
\newblock \emph{Journal of Phonetics} 31(3-4). 373--405.

\bibitem[{Hawkins(2010)}]{hawkins2010phonetic}
Hawkins, Sarah. 2010.
\newblock Phonetic variation as communicative system: {P}erception of the
  particular and the abstract.
\newblock In C\'{e}cile Fougeron, Barbara K\"{u}hnert, Mariapaola D'Imperio \&
  Nathalie Vall\'{e}e (eds.), \emph{Papers in {L}aboratory {P}honology},
  vol.~10, 479--510. Cambridge: Cambridge University Press.

\bibitem[{Hawkins(2011)}]{hawkins2011phonetic}
Hawkins, Sarah. 2011.
\newblock Does phonetic detail guide situation-specific speech recognition?
\newblock In Wai-Sum Lee \& Eric Zee (eds.), \emph{Proceedings of the 17th
  {I}nternational {C}ongress of {P}honetic {S}ciences}, 9--18. Hong Kong: City
  University of Hong Kong.

\bibitem[{Hawkins \& Nguyen(2003)}]{hawkins2003effects}
Hawkins, Sarah \& Noël Nguyen. 2003.
\newblock Effects on word recognition of syllable-onset cues to syllable-coda
  voicing.
\newblock In John Local, Richard Ogden \& Rosalind Temple (eds.), \emph{Papers
  in {L}aboratory {P}honology}, vol.~6, 38--57. Cambridge: Cambridge University
  Press.

\bibitem[{Hawkins \& Nguyen(2004)}]{hawkins2004influence}
Hawkins, Sarah \& Noël Nguyen. 2004.
\newblock Influence of syllable-coda voicing on the acoustic properties of
  syllable-onset /l/ in {E}nglish.
\newblock \emph{Journal of Phonetics} 32(2). 199--231.

\bibitem[{Hawkins \& Smith(2001)}]{hawkins2001polysp}
Hawkins, Sarah \& Rachel Smith. 2001.
\newblock Polysp: {A} polysystemic, phonetically-rich approach to speech
  understanding.
\newblock \emph{Italian Journal of Linguistics} 13. 99--188.

\bibitem[{Hawkins \& Warren(1991)}]{hawkins1991factors}
Hawkins, Sarah \& Paul Warren. 1991.
\newblock Factors affecting the given-new distinction in speech.
\newblock In \emph{Proceedings of the 12th {I}nternational {C}ongress of
  {P}honetic {S}ciences}, 66--69. Aix-en-Provence: Universit\'{e} de Provence.

\bibitem[{Heinrich et~al.(2010)Heinrich, Flory \&
  Hawkins}]{heinrich2010influence}
Heinrich, Antje, Yvonne Flory \& Sarah Hawkins. 2010.
\newblock Influence of english r-resonances on intelligibility of speech in
  noise for native {E}nglish and {G}erman listeners.
\newblock \emph{Speech Communication} 52(11). 1038--1055.

\bibitem[{Henriksen(2012)}]{henriksen2012intonation}
Henriksen, Nicholas. 2012.
\newblock The intonation and signaling of declarative questions in {M}anchego
  {P}eninsular {S}panish.
\newblock \emph{Language and Speech} 55(4).

\enlargethispage{\baselineskip}
\bibitem[{Hintzman(1986)}]{hintzman1986schema}
Hintzman, Douglas. 1986.
\newblock Schema abstraction in a multiple-trace memory model.
\newblock \emph{Psychological Review} 93(4). 411--428.

\bibitem[{Hirschberg \& Ward(1992)}]{hirschberg1992influence}
Hirschberg, Julia \& Gregory Ward. 1992.
\newblock The influence of pitch range, duration, amplitude and spectral
  features on the interpretation of the rise-fall-rise intonation contour in
  {E}nglish.
\newblock \emph{Journal of {P}honetics} 20. 241--251.

\bibitem[{Hooper(1976)}]{hooper1976word}
Hooper, Joan. 1976.
\newblock Word frequency in lexical diffusion and the source of
  morphophonological change.
\newblock In William Christie (ed.), \emph{Current progress in historical
  linguistics. {P}roceedings of the 2nd {I}nternational {C}onference on
  {H}istorical {L}inguistics}, 96--105. Amsterdam: North-Holland.

\bibitem[{House(1990)}]{house1990tonal}
House, David. 1990.
\newblock \emph{Tonal perception in speech}.
\newblock Lund: Lund University Press.

\bibitem[{House(1997)}]{house1997perceptual}
House, David. 1997.
\newblock Perceptual thresholds and tonal categories.
\newblock \emph{Phonum} 4. 179--182.

\bibitem[{Hualde(2002)}]{hualde2002intonation}
Hualde, José. 2002.
\newblock Intonation in {S}panish and the other {Ibero-Romance} languages:
  {O}verview and status quaestionis.
\newblock In Caroline Wiltshire \& Joaquim Camps (eds.), \emph{Romance
  phonology and variation}, 101--116. Amsterdam: John Benjamins.

\bibitem[{Huddleston(1994)}]{huddleston1994contrast}
Huddleston, Rodney. 1994.
\newblock The contrast between interrogatives and questions.
\newblock \emph{Journal of Linguistics} 30(2). 411--39.

\bibitem[{{Institut f\"{u}r Phonetik und digitale
  Sprachverarbeitung}(1994)}]{ipds1994kcrs}
{Institut f\"{u}r Phonetik und digitale Sprachverarbeitung}. 1994.
\newblock The {K}iel {C}orpus of {R}ead {S}peech.
\newblock CD rom.

\bibitem[{Isa{\v{c}}enko \& Sch{\"a}dlich(1970)}]{isavcenko1970untersuchungen}
Isa{\v{c}}enko, Aleksandr \& Hans~Joachim Sch{\"a}dlich. 1970.
\newblock \emph{Untersuchungen {\"u}ber die deutsche {S}atzintonation}.
\newblock Berlin: Mouton.

\bibitem[{Jackendoff(1972)}]{jackendoff1972semantic}
Jackendoff, Ray. 1972.
\newblock \emph{Semantic interpretation in generative grammar}.
\newblock Cambridge: MIT Press.

\bibitem[{Jakobson et~al.(1952)Jakobson, Fant \&
  Halle}]{jakobson1952preliminaries}
Jakobson, Roman, Gunnar Fant \& Morris Halle. 1952.
\newblock Preliminaries to speech analysis: {T}he distinctive features.
\newblock MIT {A}coustics {L}aboratory technical report.

\bibitem[{Jassem \& Richter(1989)}]{jassem1989neutralization}
Jassem, Wiktor \& Lutosława Richter. 1989.
\newblock Neutralization of voicing in {P}olish obstruents.
\newblock \emph{Journal of Phonetics} 17(4). 317--326.

\bibitem[{Johnson(1997)}]{johnson1997speech}
Johnson, Keith. 1997.
\newblock Speech perception without speaker normalization.
\newblock In Keith Johnson \& John Mullennix (eds.), \emph{Talker variability
  in speech processing}, 145--165. San Diego: Academic Press.

\bibitem[{Johnson \& Mullennix(1997)}]{johnson1997complex}
Johnson, Keith \& John Mullennix. 1997.
\newblock Complex representations used in speech processing.
\newblock In Keith Johnson \& John Mullennix (eds.), \emph{Talker variability
  in speech processing}, 1--8. San Diego: Academic Press.

\bibitem[{Jun(1993)}]{jun1993phonetics}
Jun, Sun-Ah. 1993.
\newblock \emph{The phonetics and phonology of {K}orean prosody}: Columbus: The
  Ohio State University dissertation.

\bibitem[{Jusczyk \& Luce(2002)}]{jusczyk2002speech}
Jusczyk, Peter \& Paul Luce. 2002.
\newblock Speech perception and spoken word recognition: {P}ast and present.
\newblock \emph{Ear and Hearing} 23(1). 2--40.

\enlargethispage{\baselineskip}
\bibitem[{Kelly \& Local(1986)}]{kelly1986long}
Kelly, John \& John Local. 1986.
\newblock Long-domain resonance patterns in {E}nglish.
\newblock In \emph{Proceedings of the {International Conference on Speech
  Input/Output, Techniques and Applications}}, 304--308. London.

\bibitem[{Kirsner et~al.(1994)Kirsner, {van Heuven} \& {van
  Bezooijen}}]{kirsner1994interaction}
Kirsner, Robert, Vincent {van Heuven} \& Renée {van Bezooijen}. 1994.
\newblock Interaction of particle and prosody in the interpretation of factual
  {D}utch sentences.
\newblock In Reineke Bok-Bennema \& Crit Cremers (eds.), \emph{Linguistics in
  the {N}etherlands}, 107--118. Amsterdam: John Benjamins.

\bibitem[{Klatt(1973)}]{klatt1973discrimination}
Klatt, Dennis. 1973.
\newblock Discrimination of fundamental frequency contours in synthetic speech:
  {I}mplications for models of pitch perception.
\newblock \emph{Journal of the Acoustical Society of America} 53(1). 8--16.

\bibitem[{Klatt(1979)}]{klatt1979speech}
Klatt, Dennis. 1979.
\newblock Speech perception: {A} model of acoustic-phonetic analysis and
  lexical access.
\newblock \emph{Journal of Phonetics} 7. 279--312.

\bibitem[{Kleber et~al.(2010)Kleber, John \&
  Harrington}]{kleber2010implications}
Kleber, Felicitas, Tina John \& Jonathan Harrington. 2010.
\newblock The implications for speech perception of incomplete neutralization
  of final devoicing in {G}erman.
\newblock \emph{Journal of Phonetics} 38(2). 185--196.

\bibitem[{Kohler(1987)}]{kohler1987categorical}
Kohler, Klaus. 1987.
\newblock Categorical pitch perception.
\newblock In \emph{Proceedings of the 11th {I}nternational {C}ongress of
  {P}honetic {S}ciences}, vol.~5, 331--333. Tallin: Academy of Sciences.

\bibitem[{Kohler(1991)}]{kohler1991model}
Kohler, Klaus. 1991.
\newblock A model of {G}erman intonation.
\newblock \emph{Arbeitsberichte des Instituts f\"{u}r Phonetik der
  Universit\"{a}t Kiel} 25. 295--360.

\bibitem[{Kopkalli(1993)}]{kopkalli1993phonetic}
Kopkalli, Handan. 1993.
\newblock \emph{A phonetic and phonological analysis of final devoicing in
  {T}urkish}: Ann Arbor: University of Michigan dissertation.

\bibitem[{Kretschmer(1938)}]{kretschmer1938ursprung}
Kretschmer, Paul. 1938.
\newblock Der {U}rsprung des {F}ragetons \& {F}ragesatzes.
\newblock In \emph{Scritti in onore di alfredo trombetti}, 27--50. Milano:
  Hoepli.

\bibitem[{Kruschke(1992)}]{kruschke1992alcove}
Kruschke, John. 1992.
\newblock {ALCOVE}: {A}n exemplar-based connectionist model of category
  learning.
\newblock \emph{Psychological Review} 99(1). 22--44.

\bibitem[{Ladd(1996)}]{ladd1996intonational}
Ladd, Robert. 1996.
\newblock \emph{Intonational phonology}.
\newblock Cambridge: Cambridge University Press.

\bibitem[{Ladd(2008)}]{ladd2008intonational}
Ladd, Robert. 2008.
\newblock \emph{Intonational phonology (2nd edition)}.
\newblock Cambridge: Cambridge University Press.

\bibitem[{Ladd \& Schepman(2003)}]{ladd2003sagging}
Ladd, Robert \& Astrid Schepman. 2003.
\newblock ``{S}agging transitions'' between high pitch accents in {E}nglish:
  {E}xperimental evidence.
\newblock \emph{Journal of Phonetics} 31(1). 81--112.

\bibitem[{Ladefoged(2000)}]{ladefoged2000course}
Ladefoged, Peter. 2000.
\newblock \emph{A course in phonetics (4th edition)}.
\newblock Boston: Heinle \& Heinle.

\bibitem[{Ladefoged \& Broadbent(1957)}]{ladefoged1957information}
Ladefoged, Peter \& Donald Broadbent. 1957.
\newblock Information conveyed by vowels.
\newblock \emph{Journal of the Acoustical Society of America} 29(1). 98--104.

\bibitem[{Lehiste(1970)}]{lehiste1970suprasegmentals}
Lehiste, Ilse. 1970.
\newblock \emph{Suprasegmentals}.
\newblock Cambridge: MIT Press.

\bibitem[{Levi \& Bruno(2010)}]{levi2010priming}
Levi, Susannah \& Jennifer Bruno. 2010.
\newblock Priming at the level of phonetic detail: {E}vidence from voice onset
  time.
\newblock \emph{Journal of the Acoustical Society of America} 127(3). 1853.

\bibitem[{Liberman(1979)}]{liberman1979intonational}
Liberman, Mark. 1979.
\newblock \emph{The intonational system of {E}nglish}.
\newblock New York: Garland.

\bibitem[{Licklider(1952)}]{licklider1952process}
Licklider, Joseph. 1952.
\newblock On the process of speech perception.
\newblock \emph{Journal of the Acoustical Society of America} 24(6). 590--594.

\bibitem[{Lindgren(1965)}]{lindgren1965machine}
Lindgren, Nilo. 1965.
\newblock Machine recognition of human language.
\newblock \emph{Spectrum} 2(3). 114--136.

\bibitem[{Lisker(1986)}]{lisker1986voicing}
Lisker, Leigh. 1986.
\newblock Voicing in {E}nglish: {A} catalogue of acoustic features signaling
  /b/ versus /p/ in trochees.
\newblock \emph{Language and Speech} 29. 3--11.

\bibitem[{Lisker \& Abramson(1964)}]{lisker1964crosslanguage}
Lisker, Leigh \& Arthur Abramson. 1964.
\newblock A cross-language study of voicing in initial stops: {A}coustical
  measurements.
\newblock \emph{Word} 20(3). 384--422.

\bibitem[{Local(2003{\natexlab{a}})}]{local2003phonetics}
Local, John. 2003{\natexlab{a}}.
\newblock Phonetics and talk-in-interaction.
\newblock In Daniel Recasens, Maria~Josep Sol\'{e} \& Joaquín Romero (eds.),
  \emph{Proceedings of the 15th {I}nternational {C}ongress of {P}honetic
  {S}ciences}, 115--118. Barcelona.

\bibitem[{Local(2003{\natexlab{b}})}]{local2003variable}
Local, John. 2003{\natexlab{b}}.
\newblock Variable domains and variable relevance: {I}nterpreting phonetic
  exponents.
\newblock \emph{Journal of Phonetics} 31(3-4). 321--339.

\bibitem[{Luce \& McLennan(2005)}]{luce2005spoken}
Luce, Paul \& Conor McLennan. 2005.
\newblock Spoken word recognition: {T}he challenge of variation.
\newblock In David Pisoni \& Robert Remez (eds.), \emph{The handbook of speech
  perception}, 590--609. Hoboken: Wiley-Blackwell.

\bibitem[{Luce et~al.(2003)Luce, McLennan \&
  Charles-Luce}]{luce2003abstractness}
Luce, Paul, Conor McLennan \& Jan Charles-Luce. 2003.
\newblock Abstractness and specificity in spoken word recognition: {I}ndexical
  and allophonic variability in long-term repetition priming.
\newblock In Jeffrey Bowers \& Chad Marsolek (eds.), \emph{Rethinking implicit
  memory}, 145--165. Oxford: Oxford University Press.

\bibitem[{Lyberg(1981)}]{lyberg1981observations}
Lyberg, Bertil. 1981.
\newblock Some observations on the vowel duration and the fundamental frequency
  contour in {S}wedish utterances.
\newblock \emph{Journal of Phonetics} 9. 261--272.

\bibitem[{Lyons(1977)}]{lyons1977semantics}
Lyons, John. 1977.
\newblock \emph{Semantics}.
\newblock Cambridge: Cambridge University Press.

\bibitem[{Magno~Caldognetto et~al.(1978)Magno~Caldognetto, Ferrero, Lavagnoli
  \& Vagges}]{magnocaldognetto1978f0}
Magno~Caldognetto, Emanuela, Franco Ferrero, Carlo Lavagnoli \& Kyriaki Vagges.
  1978.
\newblock F0 contours of statements, yes-no questions, and wh-questions of two
  regional varieties of {I}talian.
\newblock \emph{Journal of Italian Linguistics} 3(1). 57--66.

\bibitem[{{Manaster Ramer}(1996)}]{manaster1996letter}
{Manaster Ramer}, Alexis. 1996.
\newblock A letter from an incompletely neutral phonologist.
\newblock \emph{Journal of Phonetics} 24(4). 477--489.

\bibitem[{Marchese(1978)}]{marchese1978atlas}
Marchese, Lynell. 1978.
\newblock \emph{Atlas linguistique {K}ru: {E}ssai de typologie}.
\newblock Abidjan: Institut de Linguistique Appliqu\'{e}e.

\bibitem[{Marslen-Wilson \& Tyler(1980)}]{marslen1980temporal}
Marslen-Wilson, William \& Lorraine Tyler. 1980.
\newblock The temporal structure of spoken language understanding.
\newblock \emph{Cognition} 8(1). 1--71.

\bibitem[{Marslen-Wilson \& Welsh(1978)}]{marslen1978processing}
Marslen-Wilson, William \& Alan Welsh. 1978.
\newblock Processing interactions and lexical access during word recognition in
  continuous speech.
\newblock \emph{Cognitive Psychology} 10(1). 29--63.

\bibitem[{Mascar{\'o}(1987)}]{mascaro1987underlying}
Mascar{\'o}, Joan. 1987.
\newblock Underlying voicing recoverability of finally devoiced obstruents in
  {C}atalan.
\newblock \emph{Journal of Phonetics} 15. 183--186.

\bibitem[{Matthews(1993)}]{matthews1993grammatical}
Matthews, Peter. 1993.
\newblock \emph{Grammatical theory in the {U}nited {S}tates from {B}loomfield
  to {C}homsky}.
\newblock Cambridge: Cambridge University Press.

\bibitem[{Maturi(1988)}]{maturi1988intonazione}
Maturi, Pietro. 1988.
\newblock L'intonazione delle frasi dichiarative ed interrogative nella
  variet\`{a} napoletana dell'italiano.
\newblock \emph{Rivista Italiana di Acustica} 12. 13--30.

\bibitem[{McClelland(1981)}]{mcclelland1981retrieving}
McClelland, James. 1981.
\newblock Retrieving general and specific information from stored knowledge of
  specifics.
\newblock In \emph{Proceedings of the 3rd {Annual Conference of the Cognitive
  Science Society}}, 170--172. Berkeley.

\bibitem[{McClelland \& Elman(1986)}]{mcclelland1986trace}
McClelland, James \& Jeffrey Elman. 1986.
\newblock The {TRACE} model of speech perception.
\newblock \emph{Cognitive Psychology} 18(1). 1--86.

\bibitem[{McClelland \& Rumelhart(1985)}]{mcclelland1985distributed}
McClelland, James \& David Rumelhart. 1985.
\newblock Distributed memory and the representation of general and specific
  information.
\newblock \emph{Journal of Experimental Psychology: General} 114(2). 159--197.

\bibitem[{Medin \& Schaffer(1978)}]{medin1978context}
Medin, Douglas \& Marguerite Schaffer. 1978.
\newblock Context theory of classification learning.
\newblock \emph{Psychological Review} 85(3). 207--238.

\bibitem[{Michelas(2011)}]{michelas2011caracterisation}
Michelas, Amandine. 2011.
\newblock \emph{Caract\'{e}risation phon\'{e}tique et phonologique du syntagme
  interm\'{e}diaire en fran\c{c}ais: {D}e la production \`{a} la perception}:
  Aix-en-Provence: Universit\'{e} de Provence dissertation.

\bibitem[{M{\"o}hler(2001)}]{mohler2001improvements}
M{\"o}hler, Gregor. 2001.
\newblock Improvements of the {PaIntE} model for f0 parametrization.
\newblock Research Papers from the Phonetics Lab, IMS Universit\"{a}t
  Stuttgart.

\bibitem[{M{\"o}hler \& Conkie(1998)}]{mohler1998parametric}
M{\"o}hler, Gregor \& Alistair Conkie. 1998.
\newblock Parametric modeling of intonation using vector quantization.
\newblock In \emph{Proceedings of the 3rd {ESCA Workshop on Speech Synthesis}},
  311--316. Jenolan Caves.

\bibitem[{Moulines \& Charpentier(1990)}]{moulines1990pitchsyncronous}
Moulines, Eric \& Francis Charpentier. 1990.
\newblock Pitch-synchronous waveform processing techniques for text-to-speech
  synthesis using diphones.
\newblock \emph{Speech Communication} 9(5). 453--467.

\bibitem[{Nash \& Mulac(1980)}]{nash1980intonation}
Nash, Rose \& Anthony Mulac. 1980.
\newblock The intonation of verifiability.
\newblock In Linda Waugh \& Cornelis {van Schoonevenld} (eds.), \emph{The
  melody of language: {I}ntonation and prosody}, 219--242. Baltimore:
  University Park Press.

\bibitem[{Neukom(1995)}]{neukom1995description}
Neukom, Lukas. 1995.
\newblock \emph{Description grammaticale du nateni ({B}{\'e}nin): {S}yst{\`e}me
  verbal, classification nominale, phrases complexes, textes}.
\newblock Z\"{u}rich: Universit\"{a}t Z\"{u}rich.

\bibitem[{Nguyen et~al.(2009)Nguyen, Wauquier \& Tuller}]{nguyen2009dynamical}
Nguyen, Noël, Sophie Wauquier \& Betty Tuller. 2009.
\newblock The dynamical approach to speech perception: {F}rom fine phonetic
  detail to abstract phonological categories.
\newblock In François Pellegrino, Egidio Marsico, Ioana Chitoran \& Christophe
  Coup{\'e} (eds.), \emph{Approaches to phonological complexity}, 193--217.
  Berlin: Mouton de Gruyter.

\enlargethispage{\baselineskip}
\bibitem[{Niebuhr(2007)}]{niebuhr2007categorical}
Niebuhr, Oliver. 2007.
\newblock Categorical perception in intonation: {A} matter of signal dynamics?
\newblock In Cyril Goutte, Nicola Cancedda, Marc Dymetman \& George Foster
  (eds.), \emph{Proceedings of the 8th {Annual Conference of the International
  Speech Communication Association}}, 642--645. Antwerp.

\bibitem[{Niebuhr \& Ambrazaitis(2006)}]{niebuhr2006alignment}
Niebuhr, Oliver \& Gilbert Ambrazaitis. 2006.
\newblock Alignment of medial and late peaks in {G}erman spontaneous speech.
\newblock In Rüdiger Hoffmann \& Hansjörg Mixdorff (eds.), \emph{Proceedings
  of the 3rd {I}nternational {C}onference on {S}peech {P}rosody}, 161--164.
  Dresden.

\bibitem[{Niebuhr et~al.(2011)Niebuhr, D'Imperio, Gili~Fivela \&
  Cangemi}]{niebuhr2011shapers}
Niebuhr, Oliver, Mariapaola D'Imperio, Barbara Gili~Fivela \& Francesco
  Cangemi. 2011.
\newblock Are there ``shapers'' and ``aligners''? {I}ndividual differences in
  signalling pitch accent category.
\newblock In Wai-Sum Lee \& Eric Zee (eds.), \emph{Proceedings of the 17th
  {I}nternational {C}ongress of {P}honetic {S}ciences}, 120--123. Hong Kong:
  City University of Hong Kong.

\bibitem[{Niebuhr \& Pfitzinger(2010)}]{niebuhr2010pitchaccent}
Niebuhr, Oliver \& Hartmut Pfitzinger. 2010.
\newblock On pitch-accent identification: {T}he role of syllable duration and
  intensity.
\newblock In \emph{Proceedings of the 5th {I}nternational {C}onference on
  {S}peech {P}rosody}, Chicago.

\bibitem[{Nosofsky(1986)}]{nosofsky1986attention}
Nosofsky, Robert. 1986.
\newblock Attention, similarity, and the identification--categorization
  relationship.
\newblock \emph{Journal of Experimental Psychology: General} 115(1). 39--61.

\bibitem[{Nosofsky(1988)}]{nosofsky1988exemplar}
Nosofsky, Robert. 1988.
\newblock Exemplar-based accounts of relations between classification,
  recognition, and typicality.
\newblock \emph{Journal of Experimental Psychology: Learning, Memory, and
  Cognition} 14(4). 700--708.

\bibitem[{Noth et~al.(2000)Noth, Batliner, Kie\ss{}ling, Kompe \&
  Niemann}]{noth2000verbmobil}
Noth, Elmar, Anton Batliner, Andreas Kie\ss{}ling, Ralf Kompe \& Heinrich
  Niemann. 2000.
\newblock Verbmobil: {T}he use of prosody in the linguistic components of a
  speech understanding system.
\newblock \emph{Speech and Audio Processing} 8(5). 519--532.

\bibitem[{{O'Dell} \& Port(1983)}]{odell1983discrimination}
{O'Dell}, Michael \& Robert Port. 1983.
\newblock Discrimination of word-final voicing in {G}erman.
\newblock \emph{Journal of the Acoustical Society of America} 73. S31.

\bibitem[{Ohala(1983)}]{ohala1983cross}
Ohala, John. 1983.
\newblock Cross-language use of pitch: {A}n ethological view.
\newblock \emph{Phonetica} 40(1). 1--18.

\bibitem[{Ohala(1984)}]{ohala1984ethological}
Ohala, John. 1984.
\newblock An ethological perspective on common cross-language utilization of f0
  of voice.
\newblock \emph{Phonetica} 41(1). 1--16.

\bibitem[{Ohala(1990)}]{ohala1990interface}
Ohala, John. 1990.
\newblock There is no interface between phonology and phonetics: {A} personal
  view.
\newblock \emph{Journal of Phonetics} 18(2). 153--172.

\bibitem[{Oldfield(1966)}]{oldfield1966things}
Oldfield, Richard. 1966.
\newblock Things, words and the brain.
\newblock \emph{The Quarterly Journal of Experimental Psychology} 18(4).
  340--353.

%\enlargethispage{\baselineskip}
\bibitem[{Osgood et~al.(1957)Osgood, Suci \&
  Tannenbaum}]{osgood1957measurement}
Osgood, Charles, George Suci \& Percy Tannenbaum. 1957.
\newblock \emph{The measurement of meaning}.
\newblock Urbana: University of Illinois Press.

\bibitem[{Palmeri et~al.(1993)Palmeri, Goldinger \&
  Pisoni}]{palmeri1993episodic}
Palmeri, Thomas, Stephen Goldinger \& David Pisoni. 1993.
\newblock Episodic encoding of voice attributes and recognition memory for
  spoken words.
\newblock \emph{Journal of Experimental Psychology: Learning, Memory, and
  Cognition} 19(2). 309--328.

\bibitem[{Pandelaere \& Dewitte(2006)}]{pandelaere2006question}
Pandelaere, Mario \& Siegfried Dewitte. 2006.
\newblock Is this a question? {N}ot for long: {T}he statement bias.
\newblock \emph{Journal of Experimental Social Psychology} 42(4). 525--531.

\bibitem[{Peterson(1952)}]{peterson1952information}
Peterson, Gordon. 1952.
\newblock The information-bearing elements of speech.
\newblock \emph{Journal of the Acoustical Society of America} 24(6). 629--637.

\bibitem[{Peterson \& Barney(1952)}]{peterson1952control}
Peterson, Gordon \& Harold Barney. 1952.
\newblock Control methods used in a study of the vowels.
\newblock \emph{Journal of the Acoustical Society of America} 24(2). 175--184.

\bibitem[{Petrone(2008)}]{petrone2008role}
Petrone, Caterina. 2008.
\newblock \emph{Le r{\^o}le de la variabilit{\'e} phon{\'e}tique dans la
  repr{\'e}sentation des contours intonatifs et de leur sens}: Aix-en-Provence:
  Universit{\'e} de Provence dissertation.

\bibitem[{Petrone \& D'Imperio(2008)}]{petrone2008tonal}
Petrone, Caterina \& Mariapaola D'Imperio. 2008.
\newblock Tonal structure and constituency in {N}eapolitan {I}talian:
  {E}vidence for the accentual phrase in statements and questions.
\newblock In Plinio Barbosa, Sandra Madureira \& César Reis (eds.),
  \emph{Proceedings of 4th {I}nternational {C}onference on {S}peech {P}rosody},
  301--304. Campinas.

\bibitem[{Petrone \& D'Imperio(2009)}]{petrone2009tonal}
Petrone, Caterina \& Mariapaola D'Imperio. 2009.
\newblock Is tonal alignment interpretation independent of methodology?
\newblock In Maria Uther, Roger Moore \& Stephen Cox (eds.), \emph{Proceedings
  of the 10th {Annual Conference of the International Speech Communication
  Association}}, 2459--2462. Brighton.

\bibitem[{Petrone \& D'Imperio(2011)}]{petrone2011tones}
Petrone, Caterina \& Mariapaola D'Imperio. 2011.
\newblock From tones to tunes: {E}ffects of the f0 prenuclear region in the
  perception of {N}eapolitan statements and questions.
\newblock In Sónia Frota, Gorka Elordieta \& Pilar Prieto (eds.),
  \emph{Prosodic categories: {P}roduction, perception and comprehension},
  207--230. Dordrecht, Heidelberg, London, New York: Springer.

\bibitem[{Petrone \& Niebuhr(2014)}]{petrone2014intonation}
Petrone, Caterina \& Oliver Niebuhr. 2014.
\newblock On the intonation in {G}erman intonation questions: {T}he role of the
  prenuclear region.
\newblock \emph{Language and Speech} 57(1). 108--46.

\bibitem[{Pfitzinger(2001)}]{pfitzinger2001phonetische}
Pfitzinger, Hartmut. 2001.
\newblock \emph{Phonetische {A}nalyse der {S}prechgeschwindigkeit}: Munich:
  Ludwig-Maximilians-Universit{\"a}t M{\"u}nchen dissertation.

\bibitem[{Pierrehumbert(1980)}]{pierrehumbert1980phonology}
Pierrehumbert, Janet. 1980.
\newblock \emph{The phonology and phonetics of {E}nglish intonation}:
  Cambridge, MA: Massachussets Institut of Technology dissertation.

\bibitem[{Pierrehumbert(1981)}]{pierrehumbert1981synthesizing}
Pierrehumbert, Janet. 1981.
\newblock Synthesizing intonation.
\newblock \emph{Journal of the Acoustical Society of America} 70(4). 985--995.

\bibitem[{Pierrehumbert(1990)}]{pierrehumbert1990phonological}
Pierrehumbert, Janet. 1990.
\newblock Phonological and phonetic representation.
\newblock \emph{Journal of Phonetics} 18(3). 375--394.

%\enlargethispage{\baselineskip}
\bibitem[{Pierrehumbert(2001)}]{pierrehumbert2001exemplar}
Pierrehumbert, Janet. 2001.
\newblock Exemplar dynamics: {W}ord frequency, lenition and contrast.
\newblock In Joan Bybee \& Paul Hopper (eds.), \emph{Frequency and the
  emergence of linguistic structure}, 137--158. Amsterdam: John Benjamins.

\bibitem[{Pierrehumbert \& Beckman(1988)}]{pierrehumbert1988japanese}
Pierrehumbert, Janet \& Mary Beckman. 1988.
\newblock \emph{Japanese tone structure}.
\newblock Cambridge: MIT Press.

\bibitem[{Pierrehumbert et~al.(2000)Pierrehumbert, Beckman \&
  Ladd}]{pierrehumbert2000conceptual}
Pierrehumbert, Janet, Mary Beckman \& Robert Ladd. 2000.
\newblock Conceptual foundations of phonology as a laboratory science.
\newblock In Noel Burton-Roberts, Philip Carr \& Gerard Docherty (eds.),
  \emph{Phonological knowledge: {C}onceptual and empirical issues}, 273--303.
  Oxford: Oxford University Press.

\bibitem[{Pierrehumbert \& Hirschberg(1990)}]{pierrehumbert1990meaning}
Pierrehumbert, Janet \& Julia Hirschberg. 1990.
\newblock The meaning of intonational contours in the interpretation of
  discourse.
\newblock In Philip Cohen, Jerry Morgan \& Martha Pollack (eds.),
  \emph{Intentions in communication}, 271--311. Cambridge: MIT Press.

\bibitem[{Pierrehumbert \& Steele(1989)}]{pierrehumbert1989categories}
Pierrehumbert, Janet \& Shirley Steele. 1989.
\newblock Categories of tonal alignment in {E}nglish.
\newblock \emph{Phonetica} 46(3). 181--196.

\bibitem[{Pike(1945)}]{pike1945intonation}
Pike, Kenneth. 1945.
\newblock \emph{The intonation of {A}merican {E}nglish}.
\newblock Ann Arbor: University of Michigan Press.

\bibitem[{Podi(1995)}]{podi1995esquisse}
Podi, Napo. 1995.
\newblock \emph{Esquisse comparative de l'akasilimi et du basaal}: Grenoble:
  Universit\'{e} de Grenoble III dissertation.

\bibitem[{Port(1996)}]{port1996discreteness}
Port, Robert. 1996.
\newblock The discreteness of phonetic elements and formal linguistics:
  {R}esponse to {A}. {M}anaster {R}amer.
\newblock \emph{Journal of Phonetics} 24(4). 491--512.

\bibitem[{Port(2006)}]{port2006graphical}
Port, Robert. 2006.
\newblock The graphical basis of phones and phonemes.
\newblock In Murray Munro \& Ocke-Schwen Bohn (eds.), \emph{Second language
  speech learning: {T}he role of language experience in speech perception and
  production}, 349--365. Amsterdam: John Benjamins.

\bibitem[{Port \& Crawford(1989)}]{port1989incomplete}
Port, Robert \& Penny Crawford. 1989.
\newblock Incomplete neutralization and pragmatics in {G}erman.
\newblock \emph{Journal of Phonetics} 17(4). 257--282.

\bibitem[{Port et~al.(1981)Port, Mitleb \& O'Dell}]{port1981neutralization}
Port, Robert, Fares Mitleb \& Michael O'Dell. 1981.
\newblock Neutralization of obstruent voicing in {G}erman is incomplete.
\newblock \emph{Journal of the Acoustical Society of America} 70. S13.

\bibitem[{Port \& O'Dell(1985)}]{port1985neutralization}
Port, Robert \& Michael O'Dell. 1985.
\newblock Neutralization of syllable-final voicing in {G}erman.
\newblock \emph{Journal of Phonetics} 13(4). 455--471.

\bibitem[{Prieto et~al.(2005)Prieto, D'Imperio \&
  Gili~Fivela}]{prieto2005pitch}
Prieto, Pilar, Mariapaola D'Imperio \& Barbara Gili~Fivela. 2005.
\newblock Pitch accent alignment in {R}omance: {P}rimary and secondary
  associations with metrical structure.
\newblock \emph{Language and Speech} 48(4). 359--396.

\bibitem[{Pye(1986)}]{pye1986word}
Pye, Susan. 1986.
\newblock Word-final devoicing of obstruents in {R}ussian.
\newblock \emph{Cambridge Papers in Phonetics and Experimental Linguistics} 5.
  1--10.

\bibitem[{{R Development Core Team}(2008)}]{r2008r}
{R Development Core Team}. 2008.
\newblock R: {A} language and environment for statistical computing.
\newblock Computer program, retrieved from http://www.R-project.org/.

\bibitem[{Repp(1979)}]{repp1979relative}
Repp, Bruno. 1979.
\newblock Relative amplitude of aspiration noise as a voicing cue for
  syllable-initial stop consonants.
\newblock \emph{Language and Speech} 22(2). 173--189.

\bibitem[{Rialland(1984)}]{rialland1984fini}
Rialland, Annie. 1984.
\newblock Le fini/l'infini ou l'affirmation/l'interrogation en moba (langue
  volta{\"i}que parl{\'e}e au {N}ord-{T}ogo).
\newblock \emph{Studies in African Linguistics} supp. 9. 258--261.

\enlargethispage{2\baselineskip}
\bibitem[{Rialland(2007)}]{rialland2007question}
Rialland, Annie. 2007.
\newblock Question prosody: {A}n {A}frican perspective.
\newblock In Tomas Riad \& Carlos Gussenhoven (eds.), \emph{Tones and tunes:
  {E}xperimental studies in word and sentence prosody}, vol.~2, 35--62. Berlin:
  de Gruyter.

\bibitem[{Rietveld \& Gussenhoven(1987)}]{rietveld1987perceived}
Rietveld, Toni \& Carlos Gussenhoven. 1987.
\newblock Perceived speech rate and intonation.
\newblock \emph{Journal of Phonetics} 15(3). 273--285.

\bibitem[{Rossi(1971)}]{rossi1971seuil}
Rossi, Mario. 1971.
\newblock Le seuil de glissando ou seuil de perception des variations tonales
  pour les sons de la parole.
\newblock \emph{Phonetica} 23(1). 1--33.

\bibitem[{R{\"o}ttger et~al.(2011)R{\"o}ttger, Winter \&
  Grawunder}]{rottger2011robustness}
R{\"o}ttger, Timo, Bodo Winter \& Sven Grawunder. 2011.
\newblock The robustness of incomplete neutralization in {G}erman.
\newblock In Wai-Sum Lee \& Eric Zee (eds.), \emph{Proceedings of the 17th
  {I}nternational {C}ongress of {P}honetic {S}ciences}, 1722--1725. Hong Kong:
  City University of Hong Kong.

\bibitem[{Ryalls et~al.(1994)Ryalls, Le~Dorze, Lever, Ouellet \&
  Larfeuil}]{ryalls1994effects}
Ryalls, John, Guylaine Le~Dorze, Nathalie Lever, Lisa Ouellet \& Céline
  Larfeuil. 1994.
\newblock The effects of age and sex on speech intonation and duration for
  matched statements and questions in {F}rench.
\newblock \emph{Journal of the Acoustical Society of America} 95(4).
  2274--2276.

\bibitem[{Sabatini(1985)}]{sabatini1985italiano}
Sabatini, Francesco. 1985.
\newblock L'italiano dell'uso medio: {U}na realt{\`a} tra le variet{\`a}
  linguistiche italiane.
\newblock In Günter Holtus \& Edgar Radtke (eds.), \emph{Gesprochenes
  {I}talienisch in {G}eschichte und {G}egenwart}, 154--184. T{\"u}bingen:
  Gunter Narr.

\bibitem[{Sadock \& Zwicky(1985)}]{sadock1985speech}
Sadock, Jerrold \& Arnold Zwicky. 1985.
\newblock Speech act distinctions in syntax.
\newblock In Timothy Shopen (ed.), \emph{Language typology and syntactic
  description}, 155--196. Cambridge: Cambridge University Press.

\bibitem[{Savino(1997)}]{savino1997ruolo}
Savino, Michelina. 1997.
\newblock \emph{Il ruolo dell'intonazione nell'interazione comunicativa:
  {A}nalisi strumentale delle domande polari in un corpus di dialoghi spontanei
  (varieta' di {B}ari)}: Bari: Universit{\`a}/{P}olitecnico di {B}ari
  dissertation.

\bibitem[{Savino(2012)}]{savino2012intonation}
Savino, Michelina. 2012.
\newblock The intonation of polar questions in {I}talian: {W}here is the rise?
\newblock \emph{Journal of the International Phonetic Association} 42. 23--48.

\bibitem[{Savy \& Cutugno(2009)}]{savy2009clips}
Savy, Renata \& Francesco Cutugno. 2009.
\newblock {CLIPS: D}iatopic, diamesic and diaphasic variations in spoken
  {I}talian.
\newblock In \emph{Proceedings of the 5th {Corpus Linguistics Conference}},
  Liverpool.

\bibitem[{Schacter \& Church(1992)}]{schacter1992auditory}
Schacter, Daniel \& Barbara Church. 1992.
\newblock Auditory priming: {I}mplicit and explicit memory for words and
  voices.
\newblock \emph{Journal of Experimental Psychology: Learning, Memory, and
  Cognition} 18(5). 915--930.

\bibitem[{Scherer et~al.(1984)Scherer, Ladd \& Silverman}]{scherer1984vocal}
Scherer, Klaus, Robert Ladd \& Kim Silverman. 1984.
\newblock Vocal cues to speaker affect: {T}esting two models.
\newblock \emph{Journal of the Acoustical Society of America} 76(5).
  1346--1356.

\bibitem[{Schouten(1985)}]{schouten1985identification}
Schouten, Marten Egbertus~Hendrik. 1985.
\newblock Identification and discrimination of sweep tones.
\newblock \emph{Attention, Perception, \& Psychophysics} 37(4). 369--376.

\enlargethispage{\baselineskip}
\bibitem[{Schweitzer et~al.(2010{\natexlab{a}})Schweitzer, Calhoun,
  Sch{\"u}tze, Schweitzer \& Walsh}]{schweitzer2010relative}
Schweitzer, Katrin, Sasha Calhoun, Hinrich Sch{\"u}tze, Antje Schweitzer \&
  Michael Walsh. 2010{\natexlab{a}}.
\newblock Relative frequency affects pitch accent realisation: {E}vidence for
  exemplar storage of prosody.
\newblock In Marija Tabain, Janet Fletcher, David Grayden, John Hajek \& Andy
  Butcher (eds.), \emph{Proceedings of the 13th {Australasian International
  Conference on Speech Science and Technology}}, 62--65. Melbourne.

\bibitem[{Schweitzer et~al.(2011)Schweitzer, Walsh, Calhoun \&
  Sch{\"u}tze}]{schweitzer2011prosodic}
Schweitzer, Katrin, Michael Walsh, Sasha Calhoun \& Hinrich Sch{\"u}tze. 2011.
\newblock Prosodic variability in lexical sequences: {I}ntonation entrenches
  too.
\newblock In Wai-Sum Lee \& Eric Zee (eds.), \emph{Proceedings of the 17th
  {I}nternational {C}ongress of {P}honetic {S}ciences}, 1778--1781. Hong Kong:
  City University of Hong Kong.

\bibitem[{Schweitzer et~al.(2009)Schweitzer, Walsh, M{\"o}bius, Riester,
  Schweitzer \& Sch{\"u}tze}]{schweitzer2009frequency}
Schweitzer, Katrin, Michael Walsh, Bernd M{\"o}bius, Arndt Riester, Antje
  Schweitzer \& Hinrich Sch{\"u}tze. 2009.
\newblock Frequency matters: {P}itch accents and information status.
\newblock In Diana McCarthy \& Shuly Wintner (eds.), \emph{Proceedings of the
  12th {Conference of the European Chapter of the Association for Computational
  Linguistics}}, 728--736. Athens.

\bibitem[{Schweitzer et~al.(2010{\natexlab{b}})Schweitzer, Walsh, M{\"o}bius \&
  Sch{\"u}tze}]{schweitzer2010frequency}
Schweitzer, Katrin, Michael Walsh, Bernd M{\"o}bius \& Hinrich Sch{\"u}tze.
  2010{\natexlab{b}}.
\newblock Frequency of occurrence effects on pitch accent realisation.
\newblock In Takao Kobayashi, Keikichi Hirose \& Satoshi Nakamura (eds.),
  \emph{Proceedings of the 11th {Annual Conference of the International Speech
  Communication Association}}, 138--141. Makuhari.

\bibitem[{Sekiguchi(2006)}]{sekiguchi2006effects}
Sekiguchi, Takahiro. 2006.
\newblock Effects of lexical prosody and word familiarity on lexical access of
  spoken {J}apanese words.
\newblock \emph{Journal of Psycholinguistic Research} 35(4). 369--384.

\bibitem[{Sergeant \& Harris(1962)}]{sergeant1962sensitivity}
Sergeant, Russell \& Donald Harris. 1962.
\newblock Sensitivity to unidirectional frequency modulation.
\newblock \emph{Journal of the Acoustical Society of America} 34(10).
  1625--1628.

\bibitem[{Shattuck-Hufnagel \& Turk(1996)}]{shattuck1996prosody}
Shattuck-Hufnagel, Stefanie \& Alice Turk. 1996.
\newblock A prosody tutorial for investigators of auditory sentence processing.
\newblock \emph{Journal of Psycholinguistic Research} 25(2). 193--247.

\bibitem[{Sheffert et~al.(2002)Sheffert, Pisoni, Fellowes \&
  Remez}]{sheffert2002learning}
Sheffert, Sonya, David Pisoni, Jennifer Fellowes \& Robert Remez. 2002.
\newblock Learning to recognize talkers from natural, sinewave, and reversed
  speech samples.
\newblock \emph{Journal of Experimental Psychology: Human Perception and
  Performance} 28(6). 1447--1469.

\bibitem[{Shriberg et~al.(1998)Shriberg, Stolcke, Jurafsky, Coccaro, Meteer,
  Bates, Taylor, Ries, Martin \& Van Ess-Dykema}]{shriberg1998can}
Shriberg, Elizabeth, Andreas Stolcke, Daniel Jurafsky, Noah Coccaro, Marie
  Meteer, Rebecca Bates, Paul Taylor, Klaus Ries, Rachel Martin \& Carol Van
  Ess-Dykema. 1998.
\newblock Can prosody aid the automatic classification of dialog acts in
  conversational speech?
\newblock \emph{Language and Speech} 41(3-4). 443--492.

%\enlargethispage{\baselineskip}
\bibitem[{Slowiaczek \& Dinnsen(1985)}]{slowiaczek1985neutralizing}
Slowiaczek, Louisa \& Daniel Dinnsen. 1985.
\newblock On the neutralizing status of {P}olish word-final devoicing.
\newblock \emph{Journal of Phonetics} 13(3). 325--341.

\bibitem[{Slowiaczek \& Szymanska(1989)}]{slowiaczek1989perception}
Slowiaczek, Louisa \& Helena Szymanska. 1989.
\newblock Perception of word-final devoicing in {P}olish.
\newblock \emph{Journal of Phonetics} 17(3). 205--212.

\bibitem[{Smith(2002)}]{smith2002prosodic}
Smith, Caroline. 2002.
\newblock Prosodic finality and sentence type in {F}rench.
\newblock \emph{Language and Speech} 45(2). 141--178.

\bibitem[{Smith \& Medin(1981)}]{smith1981categories}
Smith, Edward \& Douglas Medin. 1981.
\newblock \emph{Categories and concepts}.
\newblock Cambridge: Harvard University Press.

\bibitem[{Smith et~al.(2012)Smith, Baker \& Hawkins}]{smith2012phonetic}
Smith, Rachel, Rachel Baker \& Sarah Hawkins. 2012.
\newblock Phonetic detail that distinguishes prefixed from pseudo-prefixed
  words.
\newblock \emph{Journal of Phonetics} 40(5). 689--705.

\bibitem[{Smith \& Hawkins(2000)}]{smith2000allophonic}
Smith, Rachel \& Sarah Hawkins. 2000.
\newblock Allophonic influences on word-spotting experiments.
\newblock In Anne Cutler, James McQueen \& Rian Zondervan (eds.), \emph{{ISCA
  Tutorial and Research Workshop on Spoken Word Access Processes}}, 139--142.
  Nijmegen: Max-Planck-Gesellschaft zur Förderung der Wissenschaften.

\bibitem[{Sobrero(1992)}]{sobrero1992italiano}
Sobrero, Alberto. 1992.
\newblock \emph{L'italiano di oggi}.
\newblock Roma: Istituto della Enciclopedia Italiana.

\bibitem[{Standing et~al.(1970)Standing, Conezio \&
  Haber}]{standing1970perception}
Standing, Lionel, Jerry Conezio \& Ralph Haber. 1970.
\newblock Perception and memory for pictures: {S}ingle-trial learning of 2500
  visual stimuli.
\newblock \emph{Psychonomic Science} 19(2). 73--74.

\bibitem[{Stevens(1960)}]{stevens1960model}
Stevens, Kenneth. 1960.
\newblock Toward a model for speech recognition.
\newblock \emph{Journal of the Acoustical Society of America} 32(1). 47--55.

\bibitem[{Stevens(2004)}]{stevens2004invariance}
Stevens, Kenneth. 2004.
\newblock Invariance and variability in speech: {I}nterpreting acoustic
  evidence.
\newblock In Janet Slifka, Sharon Manuel \& Melanie Matthies (eds.),
  \emph{Proceedings of {From Sound to Sense Workshop}}, vol.~B, 77--85.
  Cambridge: MIT Press.

\bibitem[{Swerts et~al.(1999)Swerts, Avesani \&
  Krahmer}]{swerts1999reaccentuation}
Swerts, Marc, Cinzia Avesani \& Emiel Krahmer. 1999.
\newblock Reaccentuation or deaccentuation: {A} comparative study of {I}talian
  and {D}utch.
\newblock In John Ohala (ed.), \emph{Proceedings of the 14th {I}nternational
  {C}ongress of {P}honetic {S}ciences}, 1541--144. San Francisco: University of
  California.

\bibitem[{{'t Hart}(1976)}]{thart1976psychoacoustic}
{'t Hart}, Johan. 1976.
\newblock Psychoacoustic backgrounds of pitch contour stylisation.
\newblock \emph{IPO -- Annual Progress Report} 11. 11--19.

\bibitem[{{'t Hart} et~al.(1990){'t Hart}, Collier \&
  Cohen}]{tHart1990perceptual}
{'t Hart}, Johan, Rene Collier \& Antonie Cohen. 1990.
\newblock \emph{A perceptual study of intonation: {A}n experimental-phonetic
  approach}.
\newblock Cambridge: Cambridge University Press.

\bibitem[{Taljaard \& Bosch(1988)}]{taljaard1988handbook}
Taljaard, Petrus \& Sonja Bosch. 1988.
\newblock \emph{Handbook of {I}sizulu}.
\newblock Hatfield, Pretoria: J.L. van Schaik.

\bibitem[{Tenpenny(1995)}]{tenpenny1995abstractionist}
Tenpenny, Patricia. 1995.
\newblock Abstractionist versus episodic theories of repetition priming and
  word identification.
\newblock \emph{Psychonomic Bulletin \& Review} 2(3). 339--363.

\bibitem[{Theodore(2009)}]{theodore2009characteristics}
Theodore, Rachel. 2009.
\newblock \emph{Some characteristics of talker-specific phonetic detail}:
  Boston: Northeastern University dissertation.

\bibitem[{Trager \& Smith(1951)}]{trager1951outline}
Trager, Leonard \& Henry Smith. 1951.
\newblock \emph{An outline of {E}nglish structure}.
\newblock Norman: Battenburg Press.

\bibitem[{Trouvain(2004)}]{trouvain2004tempo}
Trouvain, Jürgen. 2004.
\newblock \emph{Tempo variation in speech production: {I}mplications for speech
  synthesis}: Saarbr{\"u}cken: Saarland University dissertation.

\bibitem[{Trubetzkoy(1939)}]{trubeckoj1939grundzuege}
Trubetzkoy, Nikolaus. 1939.
\newblock \emph{Grundz\"{u}ge der {P}honologie}.
\newblock Prag: Travaux du cercle linguistique de Prague.

\bibitem[{Turk et~al.(2006)Turk, Nakai \& Sugahara}]{turk2006acoustic}
Turk, Alice, Satsuki Nakai \& Mariko Sugahara. 2006.
\newblock Acoustic segment durations in prosodic research: {A} practical guide.
\newblock In Stefan Sudhoff, Denisa Lenertov\'{a}, Roland Meyer, Sandra
  Pappert, Petra Augurzky, Ina Mleinek, Nicole Richter \& Johannes
  Schlie\ss{}er (eds.), \emph{Methods in empirical prosody research}, 1--28.
  Berlin: De Gruyter.

\bibitem[{Uldall(1964)}]{uldall1964dimensions}
Uldall, Elizabeth. 1964.
\newblock Dimensions of meaning in intonation.
\newblock In David Abercrombie (ed.), \emph{In honour of {D}aniel {J}ones},
  271--279. London: Longmans.

\bibitem[{{van Alphen} \& McQueen(2006)}]{vanalphen2006effect}
{van Alphen}, Petra \& James McQueen. 2006.
\newblock The effect of voice onset time differences on lexical access in
  {D}utch.
\newblock \emph{Journal of Experimental Psychology: Human Perception and
  Performance} 32(1). 178--196.

\bibitem[{{van Heerden} \& Barnard(2007)}]{vanheerden2007speech}
{van Heerden}, Charl~Johannes \& Etienne Barnard. 2007.
\newblock Speech rate normalization used to improve speaker verification.
\newblock \emph{South African Institute of Electrical Engineers} 98(4).
  136--140.

\bibitem[{{van Heuven} \& Haan(2000)}]{vanheuven2000phonetic}
{van Heuven}, Vincent \& Judith Haan. 2000.
\newblock Phonetic correlates of statement versus question intonation in
  {D}utch.
\newblock In Antonis Botinis (ed.), \emph{Intonation: {A}nalysis, modelling and
  technology}, 119--144. Dordrecht: Kluwer.

\bibitem[{{van Heuven} \& Haan(2002)}]{vanheuven2002temporal}
{van Heuven}, Vincent \& Judith Haan. 2002.
\newblock Temporal distribution of interrogativity markers in {D}utch: {A}
  perceptual study.
\newblock In Carlos Gussenhoven \& Natasha Warner (eds.), \emph{Papers in
  {L}aboratory {P}honology}, vol.~7, 61--86. Berlin: Mouton de Gruyter.

\bibitem[{{van Heuven} \& {van Zanten}(2005)}]{vanheuven2005speech}
{van Heuven}, Vincent \& Ellen {van Zanten}. 2005.
\newblock Speech rate as a secondary prosodic characteristic of polarity
  questions in three languages.
\newblock \emph{Speech Communication} 47(1). 87--99.

\bibitem[{{van Santen} \& M{\"o}bius(2000)}]{vansanten2000quantitative}
{van Santen}, Jan \& Bernd M{\"o}bius. 2000.
\newblock A quantitative model of f0 generation and alignment.
\newblock In Antonis Botinis (ed.), \emph{Intonation: {A}nalysis, modelling and
  technology}, 269--288. Dordrecht: Kluwer.

\bibitem[{Walsh et~al.(2008)Walsh, Schweitzer, M{\"o}bius \&
  Sch{\"u}tze}]{walsh2008examining}
Walsh, Michael, Katrin Schweitzer, Bernd M{\"o}bius \& Hinrich Sch{\"u}tze.
  2008.
\newblock Examining pitch-accent variability from an exemplar-theoretic
  perspective.
\newblock In Janet Fletcher, Deborah Loakes, Roland Gocke, Denis Burnham \&
  Michael Wagner (eds.), \emph{Proceedings of the 9th {Annual Conference of the
  International Speech Communication Association}}, 877--880. Brisbane.

\newpage
\bibitem[{Warner et~al.(2006)Warner, Good, Jongman \&
  Sereno}]{warner2006orthographic}
Warner, Natasha, Erin Good, Allard Jongman \& Joan Sereno. 2006.
\newblock Orthographic vs. morphological incomplete neutralization effects.
\newblock \emph{Journal of Phonetics} 34(2). 285--293.

\bibitem[{Warner et~al.(2004)Warner, Jongman, Sereno \&
  Kemps}]{warner2004incomplete}
Warner, Natasha, Allard Jongman, Joan Sereno \& Rachèl Kemps. 2004.
\newblock Incomplete neutralization and other sub-phonemic durational
  differences in production and perception: {E}vidence from {D}utch.
\newblock \emph{Journal of Phonetics} 32(2). 251--276.

\bibitem[{Wells(1945)}]{wells1945pitch}
Wells, Rulon. 1945.
\newblock The pitch phonemes of {E}nglish.
\newblock \emph{Language} 21. 27--39.

\bibitem[{West(1999)}]{west1999perception}
West, Paula. 1999.
\newblock Perception of distributed coarticulatory properties of english /l/
  and /r/.
\newblock \emph{Journal of Phonetics} 27(4). 405--426.

\bibitem[{Wheeler(1970)}]{wheeler1970processes}
Wheeler, Daniel. 1970.
\newblock Processes in word recognition.
\newblock \emph{Cognitive Psychology} 1(1). 59--85.

\bibitem[{Williams \& Stevens(1972)}]{williams1972emotions}
Williams, Carl \& Kenneth Stevens. 1972.
\newblock Emotions and speech: {S}ome acoustical correlates.
\newblock \emph{Journal of the Acoustical Society of America} 52(4).
  1238--1250.

\bibitem[{Winter \& R{\"o}ttger(forthcoming)}]{winterFORTHnature}
Winter, Bodo \& Timo R{\"o}ttger. forthcoming.
\newblock The nature of incomplete neutralization in {G}erman: {I}mplications
  for laboratory phonology.
\newblock \emph{Grazer Linguistische Studien} .

\bibitem[{Wood(1973)}]{wood1973speech}
Wood, Sidney. 1973.
\newblock Speech tempo.
\newblock In \emph{Working papers}, vol.~9, 99--147. Lund: Department of
  Linguistics, Lund University.

\bibitem[{Xu(2005)}]{xu2005speech}
Xu, Yi. 2005.
\newblock Speech melody as articulatorily implemented communicative functions.
\newblock \emph{Speech Communication} 46(3). 220--251.

\bibitem[{Zeng et~al.(2004)Zeng, Martin \& Boulakia}]{zeng2004tones}
Zeng, XiaoLi, Philippe Martin \& Georges Boulakia. 2004.
\newblock Tones and intonation in declarative and interrogative sentences in
  {M}andarin.
\newblock In \emph{Proceedings of the {International Symposium on Tonal Aspects
  of Languages}}, 235--238. Beijing.

\end{thebibliography}



\clearpage

\phantomsection%this allows hyperlink in ToC to work
\addcontentsline{toc}{chapter}{Name index}
\ohead{Name index}
% \pdfbookmark[0]{Index}{Index}

%\pdfbookmark[1]{Name index}{Name index}
\printindex[aut]

% 
% \phantomsection%this allows hyperlink in ToC to work
% \addcontentsline{toc}{chapter}{Language index}
% \ohead{Language index}
%\pdfbookmark[1]{Language index}{Language index}
% \printindex[lan]


\phantomsection%this allows hyperlink in ToC to work
\addcontentsline{toc}{chapter}{Subject index}
\ohead{Subject index}
%\pdfbookmark[1]{Subject index}{Subject index}
\printindex
\end{document}
