\section{Corpus studies on \emph{dont}}
\label{ch:dont-corpus}
This section summarizes the results of three corpus studies previously published in \citet{Abeille.2016} and \citet{Abeille.2020.JFLS} looking at relative clauses introduced by the relative word (and complementizer) \emph{dont}. I call them \emph{dont} relative clauses in this work. Previously, \citet{Godard.1988} claimed that these relative clauses allow for extraction out of the NP subject. The corpus studies support \citeauthor{Godard.1988}'s claim and go beyond it, showing that this is a common and frequently attested phenomenon in French.

\input{chapters/part2-Empirical/dont/motivation}
\input{chapters/part2-Empirical/dont/Candito-Saddah}
\input{chapters/part2-Empirical/dont/procedure}
\input{chapters/part2-Empirical/dont/results}

\subsection{General conclusion on the corpus studies on \emph{dont}}

All these corpus results converge on one major point: extracting out of the subject with \emph{dont} is not only possible, but is in fact its most common usage. In Frantext 2000--2013, it does not significantly differ from extraction out of the object, but it is significantly different from all other kinds of extractions in FTB and Frantext 1900--1913. This corroborates \citet{Godard.1988}, but also any account based on minimizing the distance between the relative word and the gap. 
In general, extraction out of NPs is very frequent (over 3/4 of all extractions in all three corpora). 

Extraction out of subjects of transitive verbs is attested in all corpora, and is actually a very frequent case. This means that extraction out of subject cannot be claimed to be restricted to cases of underlying objects, but is also found for what all theories on syntax consider as ``real'' subjects. Even if one assumes subject movement as in Minimalism, the data show clear evidence that there is no freezing effect blocking the extraction.

However, we can also observe in all three corpora that passives are more frequent among extractions out of subjects than among all other kinds of extractions. We can also see that transitive are less frequent among extractions out of subjects than among all other kinds of extractions. This does not mean that there are more passives than transitives among the extractions out of the subject: in fact, Frantext 1900--1913 has significantly more transitives, and in the two other corpora, the proportions of passives and transitives are very similar. This may however explain French scholars' intuitions that extractions out of the subject are especially good with passives, and especially bad with transitives.\footnote{It is also important to take into account some parameters that may affect our results: 
\begin{itemize}
    \item With a passive verb, there are few possible gap sites for a \emph{de}-PP besides one inside the subject itself. 
    %The only other possibilities are to extract an adverb (for \emph{dont} relatives with an inanimate antecedent only) or extract out of the oblique object (this leads to an extraction out of a PP, that is syntactically not impossible, but seem subject to some strong constraints). 
    This fact alone can explain why there are more passives in this group than in other kinds of extractions.
    \item Extraction out of an object NP necessarily involves transitive verbs, no other verb type is possible. This fact alone can explain why there are more transitives in the ``other kinds of extractions'' group.
    \item Extraction of the \emph{de}-complement of a verb can only involve unergative verbs, and no other verb type. This fact alone can explain why there are so much unergatives in the ``other kinds of extractions'' group.
\end{itemize}
In this respect the factors are not completely independent of each other. This weakens the conclusion that we can draw from the comparison.}

In \citet{Godard.1988}, all felicitous examples of extraction out of the subject contain the state verb \emph{être} (`be'). Indeed, extraction out of the subject of state verbs tends to be frequent, but possibly \citeauthor{Godard.1988} used these in order to construct simple sentences with a very straightforward semantic content. Similarly, the reason for using passives to illustrate extraction from subjects in the literature may be that such sentences are relatively simple. No other conclusion should be drawn from this choice, at least as far as production data are concerned.

In Frantext 2000--2013, we find evidence that material can stand between \emph{dont} and the subject NP it is extracted from. This contradicts \citet{Heck.2009}. The fact that intervening material rarely appears after \emph{dont} is easy to understand under accounts assuming that speakers tend to minimize the linear distance between \emph{dont} and the gap. These eccounts can also explain why constructed counterexamples have been judged as degraded in the literature \citep[e.g.,][]{Tellier.1990}. But the corpus data provide evidence that extractions out of subjects with \emph{dont} are ``real'' extractions.  
