\addchap{\lsAcknowledgementTitle} 

This book is a labor of love and collaboration, made possible through the generosity, wisdom, and dedication of many. I am profoundly grateful to the Ende people, whose vibrant language and stories form the heart of this collection. \iai{The Ende Language Committee} and the countless community authors, illustrators, and translators have shared their cultural heritage with grace and trust, allowing these 19 texts to shine as a testament to their enduring traditions. It has been an honor to collaborate in preserving and celebrating the richness of Ende life.

My deepest thanks go to the many linguistic collaborators who meticulously assisted in collecting, translating, and editing the Ende language corpus, ensuring fidelity to the Ende language while making its beauty accessible to a wider audience. Their expertise and commitment have been invaluable in shaping this anthology. I also extend my gratitude to the funders, especially the Firebird Foundation, whose generous support made this project possible, enabling the preservation and dissemination of Ende stories and songs for future generations.

To the editors and reviewers at Language Science Press, thank you for your guidance and belief in this work. Your insights have strengthened this book and amplified its reach. Finally, I am endlessly grateful to my family, whose unwavering encouragement and patience have supported my work.

This collection stands as a shared achievement, a bridge between cultures and generations, and I am deeply thankful to everyone who has made it possible.