\addchap{\lsAcknowledgementTitle} 

Writing this book has been a long journey, and I am forever grateful for and to all those who have accompanied and helped me with their continued support, academic and otherwise. 

First and foremost, I would like to thank my supervisor, Ingo Plag. In 2013, he provided me with the research idea to model derivational semantics in frames, and the demand that I know more about the topic than him at the end of my PhD. I remember well the feeling of disbelief and shock I had back then. 
Ingo, since then you have enabled me to grow from an intimidated linguistics student into a confident and independent researcher. You have provided invaluable support during the academic and personal challenges that I encountered in the process. Thank you. 

Thank you also to Shelly Lieber, who I was very lucky to have had as my co-supervisor. You accompanied me throughout the project with priceless and always timely feedback, no matter how engaged you were in your own projects. 

Thank you to the other members of my dissertation committee, Heidrun Dor\-ge\-loh and Wiebke Petersen. Both of you have accompanied me for a long time, and I am grateful and proud to have been able to complete this important step with you by my side. 

It seems almost impossible to give credit to all members of my academic family, past and present, at Heinrich Heine University and at other universities. 
There are all the members of the DFG collaborative research center \textit{The Structure of Representations in Language, Cognition, and Science}: 
Rainer Osswald and Thomas Gamerschlag, thank you for your input in countless meetings. 
My fellow project members Marios Andreou, Sven Kotowski and Viktoria Schneider, I hope you enjoyed our collaboration as much as I did.
Everyone in the CRC graduate training program, thank you for being not only my colleagues but my companions. 
Koen and Vicky, Katja, thank you for all the happy memories and for still being around. 

Then, the department of English language and linguistics. I took my first class here in 2007. Many colleagues have since moved on, new ones have joined, and some I can now call my friends. Thank you to the Hiwis who spent hours upon hours annotating and discussing my data, especially Fabian and Tobias. Julia M., Katja, Sonia, Simon, Dominic, Marie, Ghattas, Viktoria \textendash{} finishing a PhD is much easier when you have great company. Christian, Christine, Gero, Sabine \textendash{} finishing a PhD is also much easier when you have great role models. 
Ulrike, you are the heart of the department. Thank you for your help and patience.
The Ang3 is not just a work place, this is a home away from home, thanks to all of you.  

Thank you to all colleagues from other universities who have worked and published with me, especially Gabriella and Sebastian. You have opened my eyes to new possibilities.  

Thank you to the Deutsche Forschungsgemeinschaft for funding the CRC, which allowed me to focus on my PhD project and do what I love. 

Thank you to everyone who has proofread this book at some point, especially Benjamin and Sarah, who are obviously blessed with the patience of a saint (and now probably know more about derivational semantics than they ever thought they might). 

Thank you to my friends, for providing moral support, distraction, childcare, and snacks, and for making me feel valued and loved. 

Thank you to my family, who has always stood by me through thick and thin. Mama, Papa, and Jenni, I am truly grateful for your unwavering, unconditional love and support. Þetta reddast!

Gregor, the support, patience and love I have received from you over the years and decades is immeasurable. I cannot thank you enough for being by my side always.

Joshi, du bist mein Sonnenschein. Danke, dass du du bist. 
