\documentclass{article}
\usepackage[a4paper, margin=1.3cm, twocolumn, columnsep=.3cm, driver=dvips]{geometry}
\usepackage{fontspec}
%\usepackage{dict}%use the style file
\usepackage{paralist}%for the \compactdesc environment
\usepackage{tgtermes}
\setdefaultleftmargin{1em}{}{}{}{}{}
 %for illustrations
\usepackage{pstricks}
\psset{arrows=c-c}
\usepackage{overpic}
\usepackage{graphicx}
\newlength\cus
\sloppy
%\fontspec[AutoFakeBold=1.5]{Charis SIL}
%\setmainfont{DoulosSIL}
%\newfontfamily\doulos{Doulos SIL}
%\DeclareTextFontCommand{\textIPA}{\doulos}

\setmainfont{Charis SIL}
\DeclareTextFontCommand{\textipa}{\doulos}

 \begin{document} 


\setlength\cus{0.08\linewidth}
\psset{linewidth=.5pt}

\noindent
{\bf (mùdì wà) zìmbà}  {\it n} 1/2 soldier {\it pl.} (bùdì bá) bàzìmbà         \\ 
{\bf á} d {\it n} 5/6 crab {\it pl.} má         \\ 
{\bf áá} m {\it n} 0/6 chance, luck         \\ 
{\bf àfríkà}  {\it n} 1/0 Africa         \\ 
{\bf ákÈ} d {\it n} 5/6 nest {\it pl.} mákÈ         \\ 
{\bf álÈ } bw {\it n} 8/6 canoe {\it pl.} málÈ         \\ 
{\bf àmbÒ} m {\it n} 0/6 thing         \\ 
{\bf ámÓ} d {\it n} 5/6 tucan {\it pl.} mámÓ         \\ 
{\bf ányÒnè}  {\it n} 1/2  onion {\it pl.} bànányÒnè         \\ 
{\bf áwÈ} j {\it n} 5/6 goliath frog {\it pl.} máwè         \\ 
{\bf b°â} lè {\it n} 5/6 loutre à cou tacheté {\it pl.} màb°â         \\ 
{\bf b°â}  {\it n} 3/4 pit, stone {\it pl.} mìb°â         \\ 
{\bf b°ä}  {\it n} 7/8 word {\it pl.} bèb°ä         \\ 
{\bf b°áááá}  {\it ideo} depiction of walking a long distance fast         \\ 
{\bf b°É°È} bè {\it n} 0/8 beauty         \\ 
{\bf b°Ó} lè {\it n} 5/6 knee {\it pl.} màb°Ô         \\ 
{\bf bà}  {\it v} smoke (tr.) (cigarette)   mbàyá   bàyàlà bàyàgà  \\ 
{\bf bâ}  {\it v} marry   mbánâ  bálÈsÈ bánàlà   \\ 
{\bf báà} * {\it mod} two         \\ 
{\bf báà} lè {\it n} 5/6 trébuchement, action de trébucher {\it pl.} màbáà         \\ 
{\bf bàà} lè {\it n} 5/6 view {\it pl.} màbàà         \\ 
{\bf báàlà nà}  {\it v} repeat   mbàálâ      \\ 
{\bf báàlÈ}  {\it v} protect, guard, keep   mbàálá   báàlà   \\ 
{\bf bààm}  {\it ideo} depiction of closing or finishing something         \\ 
{\bf bàdà} lè {\it n} 5/6 ground {\it pl.} màbàdà         \\ 
{\bf bàdò} lè {\it n} 5/6 skin disease with blisters under skin, caused by lack of hygiene {\it pl.} màbàdò         \\ 
{\bf bágá} lè {\it n} 5/6 Flicken (zum Stopfen von Kleidung) {\it pl.} màbágá         \\ 
{\bf bágà nà}  {\it v} do sth. for last time, stop, separate   mbágâ   bágàlà   \\ 
{\bf bàgò}  {\it n} 7/8 hoe {\it pl.} bèbàgò         \\ 
{\bf bàkÈ}  {\it v} anheften, ankleben   mbàgá      \\ 
{\bf bàlándè}  {\it n} 7/8 larva, caterpillar {\it pl.} bàbàlándè         \\ 
{\bf bálÈ}  {\it v} surpass, overtake, conquer   mbálâ      \\ 
{\bf bálÒwÒ}  {\it v} bend down, se courber   mbálÓwÔ      \\ 
{\bf bàmbèyè}  {\it n} 7/8 prostitution {\it pl.} bèbàmbèyè         \\ 
{\bf bámíwálÉ}  {\it n} 7/8 scorpion {\it pl.} bèbámíwálÉ         \\ 
{\bf bámÒ}  {\it v} scold   mbámâ tu as grondé bámÈlÈ  bámàlà   \\ 
{\bf bàmò}  {\it n} 7/8 scar {\it pl.} bèbàmò         \\ 
{\bf bándá}  {\it n} 7/8 kingfisher {\it pl.} bèbándá         \\ 
{\bf bándí} lè {\it n} 5/6 protecting fetish (in house, not on body) {\it pl.} màbándí         \\ 
{\bf bándÓwÓ lé mpÒmbÓ} lè {\it n} 5/6 forehead {\it pl.} màbándÓwÓ má mámpÒmbÓ         \\ 
{\bf bándyÈ (wà lèkÓÒ)}  {\it n} 1/2 cave (of stone) {\it pl.} bàbándyÈ         \\ 
{\bf bándyì} lè {\it n} 5/6 gifle {\it pl.} màbándyì         \\ 
{\bf básí}  {\it n} 7/8 shoulder blade {\it pl.} bèbásí         \\ 
{\bf báßÈ}  {\it n} 7/6 disease {\it pl.} màbáßÈ         \\ 
{\bf báwÈ}  {\it v} injure (oneself)   mbáwâ  báwÈsÈ báwàlà   \\ 
{\bf bàwÈ}  {\it v} carry   mbàwá  bàwÈsÈ bàwàlà   \\ 
{\bf bé}  {\it n} 7/8 well, pit, hole {\it pl.} bèbé         \\ 
{\bf bÈ}  {\it v} be         \\ 
{\bf bÈ}  {\it v} sow, plant, cultivate   mbÈyá   bèyàlà   \\ 
{\bf bÈ'È}  {\it n} 7/6 shoulder {\it pl.} màbÈ'È         \\ 
{\bf bè(bè)} ná {\it qual} red         \\ 
{\bf bÉdÈ}  {\it v} light   mbÉdâ   bÉdàlà bÉdÈgà  \\ 
{\bf bédÒ}  {\it v} go up, mount   mbédâ bédèlÈ bédèsÈ bédàlà bédègà ascend  \\ 
{\bf bédò}  {\it v} ferment   mbédálâ      \\ 
{\bf bÈlànÈ}  {\it v} use   mbèlánÊ      \\ 
{\bf bÉlÉ}  {\it n} 7/8 handicap {\it pl.} bèbÉlÉ         \\ 
{\bf bÉlÈ} lè {\it n} 5/6 breast {\it pl.} màbÉlÈ         \\ 
{\bf bèlé} lè {\it n} 5/6 kola nut {\it pl.} màbèlé         \\ 
{\bf bénÈlÈ}  {\it v} lift, raise      bènala bÉnègà  \\ 
{\bf bÈngvùdÈ}  {\it n} 1/2 angwantibo doré {\it pl.} bàbÈngvùdÈ         \\ 
{\bf bénó}  {\it n} 7/8 buttock {\it pl.} bèbénó         \\ 
{\bf bènÒ}  {\it v} refuse   mbèná   bènàlà   \\ 
{\bf bÉwÒ}  {\it ?} bad         \\ 
{\bf béyÒ}  {\it v} ripen   mbéyâ  bélèsÈ  béyàgà en train de murir  \\ 
{\bf bí} lè {\it n} 5/6 excrements {\it pl.} màbí         \\ 
{\bf bí'ì} lè {\it n} 5/6 leech {\it pl.} màbí'ì         \\ 
{\bf bíá}  {\it n} 1/2 beer (French loan) {\it pl.} bàbíá         \\ 
{\bf bígÈ}  {\it v} become rich, develop, emerge   mbígâ  bígÈsÈ    \\ 
{\bf bímbú}  {\it n} 7/6 quantity {\it pl.} màbímbú         \\ 
{\bf bìndì} lè {\it n} 5/6 testicule {\it pl.} màbìndì         \\ 
{\bf bìnÓ}  {\it n} 7/8 louse {\it pl.} bèbìnÓ         \\ 
{\bf bísì nà}  {\it v} pay attention, consider         \\ 
{\bf bíßÒ}  {\it n} 7/8 thickness {\it pl.} bèbíßÒ         \\ 
{\bf bíwÒ}  {\it n} 3/0 bad luck, malheur         \\ 
{\bf bíwÒ}  {\it qual} bad         \\ 
{\bf bíyálá}  {\it n} 7/8 awful, hysterical, terribel (positive or negative) {\it pl.} bèbíyálá         \\ 
{\bf bíyÒ}  {\it v} hit, beat   mbílâ bìyÈlÈ faire qc de très mal, activer bílÈsÈ bínàlà   \\ 
{\bf bÓ} lè {\it n} 5/6 sole, footprint, hoof {\it pl.} màbÓ         \\ 
{\bf bÒ}  {\it v} rot   mbÒyá  bÒyÈsÈ    \\ 
{\bf bÔ} m {\it n} 3/6 arm {\it pl.} màbÔ         \\ 
{\bf bô}  {\it v} lie down (intr)   mbúgâ     búgÈ (tr) \\ 
{\bf bÓ'Ò} lè {\it n} 5/6 bread fruit, bread fruit tree (treculia africana) {\it pl.} màbÓ'Ò         \\ 
{\bf bódé}  {\it n} 1/2 boot {\it pl.} bàbódé         \\ 
{\bf bÒgÈ}  {\it v} enlarge   mbÒgá  bÒgÈsÈ bÒgàlà   \\ 
{\bf bÒlÉ}  {\it n} 7/8 pourriture (food) {\it pl.} bèbÒlÉ         \\ 
{\bf bÓmÈlÈ}  {\it v} wrinkle    mbÓmálâ   bÓmàlà   \\ 
{\bf bÒndì}  {\it n} 7/8 colobe noir {\it pl.} bèbÒndì         \\ 
{\bf bóndó} lè {\it n} 5/6 toad {\it pl.} màbóndó         \\ 
{\bf bòtù} mà {\it n} 0/6 teigne (white things in head)         \\ 
{\bf bû}  {\it v} destroy   mbúyâ   búyàlà   \\ 
{\bf bùábùá}  {\it n} 7/8 when smoking an animal or fish, the meat is not yet dry         \\ 
{\bf bùdÉ}  {\it n} 7/8 shell (sea, turtle, nut), skin of fruit {\it pl.} bèbùdÉ         \\ 
{\bf bùgù}  {\it n} 7/8 place {\it pl.} bèbùgù         \\ 
{\bf búkÉ}  {\it n} 7/8 crazy person; tsetse fly {\it pl.} bèbúkÉ         \\ 
{\bf búlÈ}  {\it v} burst   mbúlâ      \\ 
{\bf búlÒ}  {\it v} fish   mbúlâ   búlàlà   \\ 
{\bf búlÒ m°â} m {\it n} 1/2 fisherman {\it pl.} bàbúlÒ m°â         \\ 
{\bf búmÈ}  {\it v} bark      búmàlà   \\ 
{\bf bùmÈ}  {\it v} announce sth   mbùmá   bùmàlà   \\ 
{\bf bùmè màpÔ} m {\it n} 1/2 announcer, messenger {\it pl.} bàbùmÈ bá mápÔ         \\ 
{\bf bùmÈlÈ}  {\it v} hit (nail)   mbùmálâ   bùmàlà   \\ 
{\bf búndÓ}  {\it n} 7/8 bride price {\it pl.} bèbúndÓ         \\ 
{\bf búndÒ}  {\it v} pay brideprice   mbúndâ  búndÈsÈ búndàlà   \\ 
{\bf búnj°â} ná {\it n} 1/2 bed bug {\it pl.} bànàbúnj°â         \\ 
{\bf búÒ}  {\it n} 1/2 mute person {\it pl.} bàbúÒ         \\ 
{\bf búÒ}  {\it n} 7/8 mortar {\it pl.} bèbúÒ         \\ 
{\bf bùÓ} lè {\it n} 5/6 cripple {\it pl.} màbùÓ         \\ 
{\bf búùlÈ}  {\it n} 7/8 old settlement {\it pl.} bèbúùlÈ         \\ 
{\bf búwà} lè {\it n} 5/6 lung {\it pl.} màbúwà         \\ 
{\bf búwÈlÈ}  {\it v} tâter, drücken (if mango ripe)   mbúwálâ      \\ 
{\bf bvû}  {\it v} think, believe      bvúàlà   \\ 
{\bf bvúàlà}  {\it v} believe   mbvúálâ      \\ 
{\bf bvúbvù}  {\it n} 9/0 multitude         \\ 
{\bf bvùbvù}  {\it inv} (too) many, (too) much         \\ 
{\bf bvúdà nà}  {\it v} quarrel   mbvúdâ   bvúdàlà   \\ 
{\bf bvùdÈ}  {\it n} 7/6 clearing (in forest) {\it pl.} màbvùdÈ         \\ 
{\bf bvúlÈ}  {\it n} 0/2 Bulu people         \\ 
{\bf bvùlÉ}  {\it n} 8/8 night {\it pl.} bèbvùlÉ         \\ 
{\bf bvùmá}  {\it n} 7/8 1) fruit 2) ball {\it pl.} bèbvùmá         \\ 
{\bf bvùmà}  {\it v} donnern (tonnerre gronde)       bvùmàgà flock of birds flys away suddenly  \\ 
{\bf bvùmá bé mábÉlÈ} bè {\it n} 7/8 female breast         \\ 
{\bf bvùmá yá ngÒndÈ}  {\it n} 7/8 full moon (ball of moon) {\it pl.} bèbvùmá bé ngÒndÈ         \\ 
{\bf bvùmbà}  {\it v} surprise sb, chasser (voleur)   mbvùmbá   bvùmbàlà   \\ 
{\bf bvúÓ}  {\it n} 8/8 elephant trunk {\it pl.} bèbvúÓ         \\ 
{\bf bvúÒ}  {\it v} break (tr), harvest mais   mbvúgâ   bvúgàlà  bvúkÈ break (intr) \\ 
{\bf bvúú} lè {\it n} 5/6 anger, being annoyed, unhappiness         \\ 
{\bf bw°â}  {\it n} 8/6 medicine {\it pl.} màbw°â         \\ 
{\bf bw°ásà} mà {\it n} 0/6 thoughts         \\ 
{\bf bw°âsà}  {\it v} think, remember         \\ 
{\bf bw°ÉÈlÈ}  {\it v} wait      bw°áàlà   \\ 
{\bf bw°O} lè {\it n} 5/6 beehive {\it pl.} màbw°O         \\ 
{\bf bwà}  {\it v} give birth   mbwàlá bwàlÈ be born bwàlèsÈ    \\ 
{\bf bwà}  {\it v} become big   mbògá   bÒgàlà  bÒgÈ fatten, make fat \\ 
{\bf bwá mákí}  {\it v} lay eggs (give borth to eggs)         \\ 
{\bf bwàà}  {\it v} become, have         \\ 
{\bf bwádÒ}  {\it v} dress, wear   mbwádâ  bÓdÈsÈ bÓdàlà   \\ 
{\bf bwálÉ} m {\it n} 1/2 parent {\it pl.} bàbwálÉ         \\ 
{\bf bwálÉ} mà {\it n} 0/6 birth         \\ 
{\bf bwálÈsÈ bùd°â} m {\it n} 1/2 midwife {\it pl.} bàbwálèsè bá bùd°â         \\ 
{\bf bwámÒ}  {\it v} 1) leave, go out 2) receive, obtain 3) become   mbwámâ   bwámàlà   \\ 
{\bf bwàndÒ}  {\it v} peel (pellen) (mais, mango)   mbwàndá   bwàndàlà   \\ 
{\bf bwàndyá}  {\it n} 7/8 disdain, adultery {\it pl.} bèbwàndyá         \\ 
{\bf bwàndyà}  {\it v} despise   mbwàndyá   bwàndyàlà   \\ 
{\bf bwè}  {\it v} catch, arrest   mbùlá   bèyàlà   \\ 
{\bf bwèdÒwÒ}  {\it n} 7/6 taste {\it pl.} màbwèdÒwÒ         \\ 
{\bf bwÈdÒwÒ}  {\it v} be sweet, be tasty     bòdÈsÈ make sweet    \\ 
{\bf bwi} lè {\it n} 5/6 hyena {\it pl.} màbwi         \\ 
{\bf bwímÒ}  {\it n} 7/8 net hunting {\it pl.} bèbwímÒ         \\ 
{\bf bwÔ(bwÒ)}  {\it n} 7/8 brain {\it pl.} bèbwÔ         \\ 
{\bf bwúyà}  {\it n} 7/8 hundred {\it pl.} bèbwúyà         \\ 
{\bf byáàdà}  {\it v} answer, respond   mbyáàdá      \\ 
{\bf d°à}  {\it v} draw water   nd°ààlá d°ààlÈ  d°àngàlà   \\ 
{\bf d°â lé báfú} lè {\it n} 5/6 fish pont, source {\it pl.} màd°â má báfú         \\ 
{\bf d°Ê}  {\it adv} today         \\ 
{\bf dá lé tíndí}  {\it n} 5/6 giftiger Krebs im Wald         \\ 
{\bf dè}  {\it v} eat   ndíyâ  dílèsÈ díyàlà   \\ 
{\bf dèlÉmÓÒ} lè {\it n} 5/6 mud wasp {\it pl.} màdèlÈmÓÒ         \\ 
{\bf dÉndÈ}  {\it v} set (trap)   ndÉndâ   dÉndàlà   \\ 
{\bf déwÒ} bè {\it n} 0/8 food         \\ 
{\bf dígà} mà {\it n} 0/6 vision, apparition         \\ 
{\bf dìlá} mà {\it n} 0/6 funeral         \\ 
{\bf dìlÈ}  {\it v} bury   ndìlá   dìlàlà   \\ 
{\bf dísì}  {\it n} 7/8 bowl {\it pl.} bèdísì         \\ 
{\bf dísì lé bénó}  {\it n} 5/6 anus (eye of the buttock) {\it pl.} mísì mí bénó         \\ 
{\bf díyÈ}  {\it qual} expensive         \\ 
{\bf dj°á°à(-sà)}  {\it v} disappear suddenly (slowly)   ndj°áásá   dj°á°àlà   \\ 
{\bf djàngàlà}  {\it v} have sex         \\ 
{\bf djí}  {\it n} 7/8 place (where someone stays) {\it pl.} bèdjí         \\ 
{\bf djì}  {\it v} open   ndjìyá   djìyàlá   \\ 
{\bf djí yá más°Ô}  {\it n} 7/8 cemetery (forest of graves) {\it pl.} bèdjí bé más°Ô         \\ 
{\bf djì(yÒ) (sí)}  {\it v} sit (down), habiter, stay   ndjìlá djìlÈ seat sb, stay  djìlàlà   \\ 
{\bf djìbÒ}  {\it v} close   ndjìßá   djìßàlà on se ferme les portes   \\ 
{\bf djígà}  {\it v} be angry         \\ 
{\bf djíì} bè {\it n} 0/8 anger         \\ 
{\bf djíì}  {\it v} ask (a question)         \\ 
{\bf djìí}  {\it n} 7/8 forest, brousse {\it pl.} bèdjìí         \\ 
{\bf djíkÈsÈ}  {\it v} make sb. angry         \\ 
{\bf djílÒ}  {\it v} be satisfied (not hungry)   ndjílâ  djílèsÈ    \\ 
{\bf djìlÒ} lè {\it n} 5/6 1) weight 2) dignity {\it pl.} màdjìlÒ         \\ 
{\bf djìlÒ}  {\it v} be heavy   ndjìlá  djìlÈsÈ    \\ 
{\bf djímbÈ}  {\it v} get lost   ndjímbâ djímbÈlÈ lose djímbÈsÈ make forget djímbàlà forget each other   \\ 
{\bf djímÈsÈ}  {\it v} extinguish   ndjímâ      \\ 
{\bf djímÒ}  {\it v} be deep         \\ 
{\bf djìnà}  {\it v} dive, disappear in water   ndjìnâ  djìnÈsÈ  djìnÈgà sink (intr), melt  \\ 
{\bf djínÒ}  {\it n} 5/6 name {\it pl.} mínÒ         \\ 
{\bf djísÒwÒ}  {\it n} 7/8 patience {\it pl.} bèdjísÒwÒ         \\ 
{\bf djíwÒ}  {\it v} steal, plunder   ndjíßâ   djíßàlà   \\ 
{\bf djíyÈ}  {\it n} burn (intr)         \\ 
{\bf djíyÈ}  {\it v} burn (intr)   ndjígâ  djígÈsÈ make angry djígàlà burn recip  djígÈ burn sth \\ 
{\bf djìyÓ}  {\it n} 7/8 chair {\it pl.} bèdjìyÓ         \\ 
{\bf djÓwÒ}  {\it n} 7/8 day work {\it pl.} bèdjÓwÒ         \\ 
{\bf dó}  {\it n} 7/8 lie {\it pl.} bèdó         \\ 
{\bf dÒ} mà {\it n} 0/6 negotiation for price         \\ 
{\bf dÒ}  {\it v} negotiate (for price), discuss         \\ 
{\bf dómÈ}  {\it n} 7/8 laziness {\it pl.} bèdómÈ         \\ 
{\bf dÒÒ}  {\it n} 7/8 puddle {\it pl.} bèdÒÒ         \\ 
{\bf dówó} bè {\it n} 0/8 sweat         \\ 
{\bf dù}  {\it n} 7/6 thigh {\it pl.} màdù         \\ 
{\bf dùlÈ}  {\it n} 7/6 bitterness {\it pl.} màdùlÈ         \\ 
{\bf dùlè mákímbÓ}  {\it n} 7/6 salty (bitterness of salt) {\it pl.} màdùlÈ         \\ 
{\bf dúmbÓ}  {\it n} 7/8 package, packet {\it pl.} bèdúmbÓ         \\ 
{\bf dúngìlà}  {\it n} 7/8 hérisson à grandes oreilles, small with rabbit like ears {\it pl.} bèdúngìlà         \\ 
{\bf dúÒdù}  {\it n} 7/8 termite queen, Raupe {\it pl.} bèdúÒdù         \\ 
{\bf dúù}  {\it v} must not         \\ 
{\bf dv°ü}  {\it n} 7/8 noise {\it pl.} bèdv°ü         \\ 
{\bf dv°ú°Ò}  {\it n} 7/8 great hornbill {\it pl.} bèdv°ú°Ò         \\ 
{\bf dvùbÒ}  {\it v} soak, dip   ndvùbá dvùbÈlÈ bèkà Gras jäten mit Hake dvùbÈsÈ (wÊ màjíwÓ) dvùbàlà   \\ 
{\bf dvùdÒ}  {\it v} drive   ndvùdá   dvùdàlà   \\ 
{\bf dvúmá}  {\it n} 7/8 honour {\it pl.} bèdvúmá         \\ 
{\bf dvúmÈlÈ}  {\it v} praise sb.   ndvùmálá   dvùmàlà   \\ 
{\bf dvúmÒ}  {\it n} 7/8 baobab tree {\it pl.} bèdvúmÒ         \\ 
{\bf dvùmÒ}  {\it v} fall down (tree)   ndvùmá  dvùmÈsÈ dvùmàlà   \\ 
{\bf dvùÒ}  {\it v} hurt (intr. refl)   ndvùgá  dvùgÈsÈ dvùgàlà  dvùgè \\ 
{\bf dvùwÒ}  {\it v} stuff sth.         \\ 
{\bf dwàmbò}  {\it v} ask for sth   ndwàmbá   dwàmbàlà   \\ 
{\bf dy°á°à}  {\it v} chase, drive away   ndyángâ   dyángàlà   \\ 
{\bf dy°Ò°Ó}  {\it n} 7/8 bed {\it pl.} bèdy°Ò°Ó         \\ 
{\bf dy°ù}  {\it v} be hot   ndyúngâ, ndúngálâ dyúngÈlÈ heat, boil sth.  dyúngàlà (warm body around fire)   \\ 
{\bf dy°û} lè {\it n} 5/6 heat (from sun), fever {\it pl.} màdy°û         \\ 
{\bf dyà}  {\it v} sing   ndyàyâ   dyààlà   \\ 
{\bf dyâ (sí)}  {\it v} lie (down), live (habiter)   ndyáyâ   dyáàlà have sex   \\ 
{\bf dyáàlà}  {\it v} have sex         \\ 
{\bf dyágÓ}  {\it n} 7/8 sleeping place {\it pl.} bèdyágÓ         \\ 
{\bf dyàmbò}  {\it v} copulate         \\ 
{\bf dyÊ} lè {\it n} 5/6 pincers (insect) {\it pl.} màdyÊ         \\ 
{\bf dyÉkÈ}  {\it v} lean sth, incline sth   ndyÉkâ   dyékàlà aneinander lehnen   \\ 
{\bf dyÒ}  {\it n} 7/8 smile, laughter {\it pl.} bèdyÒ         \\ 
{\bf dyÒ}  {\it v} laugh, smile   ndyòlàsà  dyòlÈsÈ dyÒàlà   \\ 
{\bf dyÔ}  {\it n} 7/8 sleep {\it pl.} bèdyÔ         \\ 
{\bf dyÒdálà} mà {\it n} 0/6 deception, tromperie         \\ 
{\bf dyÒdálà} mà {\it n} 0/6 tromperie         \\ 
{\bf dyÒdÈ}  {\it v} deceive, cheat   ndyÒdá   dyÒdàlà   \\ 
{\bf dyúà}  {\it v} swim         \\ 
{\bf dyúàdà}  {\it v} feel, hear, perceive sensually         \\ 
{\bf dyùlÈ}  {\it v} be bitter   ndyùlá versalzen  dyùlÈsÈ dyùlàlà   \\ 
{\bf dyúmÒ}  {\it n} 1/2 spouse {\it pl.} bàdyúmÒ         \\ 
{\bf dyùmÒ}  {\it v} heal (se soigner), get well   ndyùmá      \\ 
{\bf dyúmù} mà {\it n} 0/6 sperm {\it pl.} màdyúmù         \\ 
{\bf dyúná}  {\it n} 7/8 quarrel, dispute {\it pl.} bèdyúná         \\ 
{\bf dyúnà}  {\it v} quarrel   ndúnâ      \\ 
{\bf dyúngúlÈ}  {\it n} 7/8 chameleon {\it pl.} bèdyúngúlÈ         \\ 
{\bf dyùù}  {\it v} kill   ndyúwâ   dyúwàlà   \\ 
{\bf dyùwá}  {\it n} 7/6, 8 thorn {\it pl.} bèdyùwá, màdyùwá         \\ 
{\bf dyúwÒ}  {\it n} 5/0 sky         \\ 
{\bf dyúwÒ}  {\it post} on top, above         \\ 
{\bf dyúwÒ}  {\it v} hear   ndyùgá dyúwÈlÈ listen dyúgÈsÈ make feel dyúwàlà   \\ 
{\bf dzámÈ}  {\it v} excuse, forgive         \\ 
{\bf É}  {\it prep} at, on, by         \\ 
{\bf É ná}  {\it ?} how         \\ 
{\bf É vÉ}  {\it ?} where         \\ 
{\bf Éndì} d {\it n} 5/6 courtyard {\it pl.} mÉndì         \\ 
{\bf ès°âs}  {\it n} 7/8 gaz, fuel {\it pl.} bès°âs         \\ 
{\bf ÉsÈ} * {\it mod} all, every         \\ 
{\bf fàlà}  {\it n} 7/0 France         \\ 
{\bf fàmí}  {\it n} 1/2 family {\it pl.} bàfàmí         \\ 
{\bf fàrínì}  {\it n} 1/2 flour {\it pl.} bàfàrínì         \\ 
{\bf fû}  {\it n} 1/2 fish {\it pl.} bàfû         \\ 
{\bf fû} lè {\it n} 5/6 day {\it pl.} máfû         \\ 
{\bf fù'ú}  {\it n} 1/2 rainy season(Aug-Nov), (Kwasio) {\it pl.} bàfù'ú         \\ 
{\bf fùÈsÈ}  {\it v} shake   fùásâ      \\ 
{\bf fúgÈ}  {\it v} end   mfúgâ   fúàlà   \\ 
{\bf fúkÈ}  {\it n} 1/2 driver ants (Hymenoptera) {\it pl.} bàfúkÈ         \\ 
{\bf fùláwà}  {\it n} 7/8 flower, hedge, bush {\it pl.} bèfùláwà         \\ 
{\bf fùlÈ}  {\it v} miss, escape   mfùlâ  fùlÈsÈ fùlàlà   \\ 
{\bf fùlÒ}  {\it v} descend, go down   mfùlâ  fùlèsÈ    \\ 
{\bf fúmbélé}  {\it n} 3/4 shin {\it pl.} mìmfùmbélé         \\ 
{\bf fùmbí}  {\it n} 7/8 orange {\it pl.} bèfùmbí         \\ 
{\bf fùÒ} lè {\it n} 5/6 stem, plant {\it pl.} màfùÒ         \\ 
{\bf fúsì} * {\it mod} different         \\ 
{\bf fwálá} lè {\it n} 5/6 end {\it pl.} màfwálá         \\ 
{\bf fwálá má bétúmbÓ} mà {\it  n} 5/6 border (between countries)         \\ 
{\bf fwálá má nkùlÉ} mà {\it n} 0/6 summit         \\ 
{\bf g°â}  {\it n} 1/2 gown {\it pl.} bàg°â         \\ 
{\bf gb°í gb°ì gb°í gb°ì gb°í}  {\it ideo} dpeiction of small objects moving in space, e.g. bacteria roaming in body         \\ 
{\bf gbìm}  {\it ideo} depiction of putting or falling down of person or object         \\ 
{\bf gíndó'ó}  {\it n} 7/8 angwantibo  doré + de Calabar, potto? {\it pl.} bègíndó'ó         \\ 
{\bf gìyÒ}  {\it v} cry     gìlÈsÈ gìlàlà   \\ 
{\bf gÓlÈ}  {\it n} 7/8 gold {\it pl.} bègÓlÈ         \\ 
{\bf gwámbÒ}  {\it v} ask for sth, beg         \\ 
{\bf gwàwÓ}  {\it n} 7/8 Lycaon, civet {\it pl.} bègwàwÓ         \\ 
{\bf gwÉmbÈ}  {\it n} 7/8 cloth {\it pl.} bègbÉmbÈ         \\ 
{\bf gy°à}  {\it v} paint, draw   ngyàngâ      \\ 
{\bf gy°á°à}  {\it n} 1/2 side {\it pl.} bàgy°á°à         \\ 
{\bf gy°âlÈ}  {\it v} roast   ngy°áàlâ      \\ 
{\bf gy°Ê} n {\it n} 1/2 stranger, guest {\it pl.} bàgy°Ê         \\ 
{\bf gyà}  {\it n} 7/8 music, song {\it pl.} bègyà         \\ 
{\bf gyà}  {\it n} 7/8 distance, length {\it pl.} bèdyà         \\ 
{\bf gyà}  {\it v} be long         \\ 
{\bf gyâ} lè {\it n} 5/6 charcoal {\it pl.} màgyâ         \\ 
{\bf gyà wá lètÓlÈ} n {\it n} 3/4 hernia {\it pl.} mìngyà mí mátÓlÈ         \\ 
{\bf gyá yá nyúmbù}  {\it n} 7/8 lip {\it pl.} bègyá bé nyúmbù         \\ 
{\bf gyàgà}  {\it v} buy   ngyàgá   gyàgàlà   \\ 
{\bf gyàgÈsì bèsâ} n {\it n} 1/2 merchant, vendor {\it pl.} bàgyàgÈsì bá bésâ         \\ 
{\bf gyàlÉ}  {\it n} 7/8 period after giving birth (about a month) {\it pl.} bègyàlÉ         \\ 
{\bf gyàlÉ} ná {\it n} 1/2 breastfeeding woman {\it pl.} bànágyàlÉ         \\ 
{\bf gyámbÒ}  {\it v} prepare, cook   ngyámbâ gyámbÈlÈ prepare for  gyámbàlà   \\ 
{\bf gyángyà}  {\it v} work   ngyángyâ  gyángyÈsÈ make work gyángyàlà   \\ 
{\bf gyÉ} lè {\it n} 5/6 tooth {\it pl.} màgyÉ         \\ 
{\bf gyÊ}  {\it n} 7/8 loutre à joues blanches de Congo {\it pl.} bègyÊ         \\ 
{\bf gyÉ'È}  {\it v} block   ngyÉgâ   gyÉgàlà   \\ 
{\bf gyÈ'ÈlÈ} mà {\it n} 0/6 prayer         \\ 
{\bf gyÈ'ÈlÈ}  {\it v} pray, beg, demand   ngyàálâ      \\ 
{\bf gyÉlÈ}  {\it v} jump, fly   ngyÉlâ  gyÉlÈsÈ gyÉlàlà   \\ 
{\bf gyÈlì} n {\it n} 1/2 Gyeli person {\it pl.} bàgyÈlì         \\ 
{\bf gyÉmÒ}  {\it n} 7/8 habit, manner {\it pl.} bègyÉmÒ         \\ 
{\bf gyÈndÒ}  {\it v} slip   ngyÈndá      \\ 
{\bf gyÉsÒ}  {\it v} look for, search, lack   ngyÉsâ   gyÉsàlà   \\ 
{\bf gyí}  {\it pro} what         \\ 
{\bf gyíbÒ}  {\it v} call   ngyíbâ   gyíbàlà   \\ 
{\bf gyìbÒ}  {\it v} sharpen   ngyìbá   gyìbàlà   \\ 
{\bf gyìdÈ}  {\it v} forgive   ngyìdá      \\ 
{\bf gyíkà (nà)}  {\it v} resemble         \\ 
{\bf gyíkÈ}  {\it v} learn   ngyíkâ be intelligent  gyíkÈsÈ teach    \\ 
{\bf gyímbÒ}  {\it v} dance   ngyímbáà  gyímbÈsÈ gyímbàlà   \\ 
{\bf gyìmbÒ} n {\it n} 1/2 sorcerer {\it pl.} bàgyìmbÒ         \\ 
{\bf gyìmbÒ} lè {\it n} 5/6 magic (some people born with) {\it pl.} màgyìmbÒ         \\ 
{\bf gyímÈ}  {\it v} wake sb. up   ngyímâ  gyímÈsÈ  gyímàgà wake up  \\ 
{\bf gyímù}  {\it n} 7/8 tongue {\it pl.} bègyímù         \\ 
{\bf gyÓlÉ} lè {\it n} 5/6 bushbaby (galago alleni) {\it pl.} màgyÓlÉ         \\ 
{\bf hámà}  {\it n} 1/2 hammer {\it pl.} bàhámà         \\ 
{\bf í} m {\it n} 1/2 non-Pygmy people {\it pl.} bí         \\ 
{\bf ímbÓ} dj {\it n} 5/6 raffia palm {\it pl.} mímbÓ         \\ 
{\bf ísì} d {\it n} 5/6 1) eye 2) kernel, seed {\it pl.} mísì         \\ 
{\bf ìtálíyÈn}  {\it n} 7/0 Italy         \\ 
{\bf ji}  {\it n} 7/8 bench {\it pl.} bèji         \\ 
{\bf jíbí} n {\it n} 1/2 thief {\it pl.} bàjíbí         \\ 
{\bf jíwÓ}  {\it n} 7/8 river {\it pl.} bìjíwÓ         \\ 
{\bf jíwÓ} mà {\it n} 0/6 water         \\ 
{\bf k°â}  {\it n} 7/8 régime of palm nuts {\it pl.} bèk°â         \\ 
{\bf k°â}  {\it v} wrap   nk°âlá   k°áàlà   \\ 
{\bf k°è}  {\it v} shave   nkèngá   kèngàlà   \\ 
{\bf k°ê nlô} n {\it n} 1/2 gecko (shave-head) {\it pl.} bak°ê mínlô         \\ 
{\bf k°É°ÉsÓ}  {\it n} 7/8 equal/next person, neighbor {\it pl.} bèk°É°ÉsÓ         \\ 
{\bf k°ÔlÈ}  {\it v} snore         \\ 
{\bf k°û}  {\it n} 1/2 leopard {\it pl.} bàk°û         \\ 
{\bf ká} lè {\it n} 5/6 leaf {\it pl.} màká         \\ 
{\bf ká}  {\it n} 7/8 grass {\it pl.} bèká         \\ 
{\bf kà}  {\it v} catch (object in air)   nkàsá kàsÈlÈ light     \\ 
{\bf ká'à}  {\it v} role up (mattress, paper), einrollen, einwickeln (bandage), verbinden   nkágâ   kágàlà   \\ 
{\bf kà'á} lè {\it n} 5/6 clan, tribe, kind {\it pl.} màkà'á         \\ 
{\bf kábálá}  {\it n} 7/8 horse {\it pl.} bèkábálá         \\ 
{\bf kàbÒ}  {\it v} share, divide, serve   nkàßá   kàßàlà   \\ 
{\bf kàdÈ}  {\it v} detach (baton de manioc)   nkàdá  kàdÈsÈ kàdàlà kàdègà detach by itself  \\ 
{\bf kádÒ}  {\it v} depasser, be too much   nkádâ   kádàlà   \\ 
{\bf kàdó}  {\it n} 1/2 gift, present {\it pl.} bàkàdó         \\ 
{\bf kàdÔ}  {\it n} 1/2 ladder {\it pl.} bàkàdÔ         \\ 
{\bf kàgá}  {\it n} 7/8 defect giving birth {\it pl.} bèkàgá         \\ 
{\bf kàgà} lè {\it n} 5/6 bewitched woman {\it pl.} màkàgà         \\ 
{\bf kágÉ} lè {\it n} 5/6 promise {\it pl.} màkágÉ         \\ 
{\bf kàgÒ}  {\it v} promise      kàgàlà  kàgÈ mÊ  \\ 
{\bf kákà}  {\it v} shiver         \\ 
{\bf kàká}  {\it n} 7/8 cocoa (theobroma cacao) {\it pl.} bèkàká         \\ 
{\bf kálá}  {\it n} 7/8 chili paste seasoning {\it pl.} bèkálá         \\ 
{\bf kàlà} lè {\it n} 5/6 doughnuts {\it pl.} màkàlà         \\ 
{\bf kàlà}  {\it n} 7/8 strawmat {\it pl.} bèkàlà         \\ 
{\bf kálàdÈ}  {\it n} 7/8 book {\it pl.} bèkálàdÈ         \\ 
{\bf kàlànÈ}  {\it v} transmit, translate   nkálánÊ      \\ 
{\bf kálé}  {\it n} 1/2 sister (older and younger) {\it pl.} bàkálé         \\ 
{\bf kàlÈgà}  {\it v} stop over, go over with stops         \\ 
{\bf kámbÈ}  {\it n} 1/2 weaver ants (Oecophylla) {\it pl.} bàkámbÈ         \\ 
{\bf kámbÒ}  {\it v} chew   nkámbâ   kámbàlà   \\ 
{\bf kàmbÒ nà}  {\it v} defend   nkàmbá   kàmbàlà   \\ 
{\bf kàmÈrún}  {\it n} 1/0 Cameroon         \\ 
{\bf kándà}  {\it v} crack (bottle, cup, glass)   nkándâ  kándÈsÈ    \\ 
{\bf kàndá}  {\it n} 7/8 proverb {\it pl.} bèkàndá         \\ 
{\bf kàNgá}  {\it n} 7/8 proverb {\it pl.} bèkàNgá         \\ 
{\bf kàsà}  {\it n} 7/8 bridge {\it pl.} bèkàsà         \\ 
{\bf kàsÈlÈ}  {\it v} light   nkàsálâ   kàsàlà   \\ 
{\bf kásÒ}  {\it v} become thin, maigrir   nkásâ kásÈlÈ  kásàlà kásègà get suddenly angry  \\ 
{\bf kàwà}  {\it n} 7/8 long dress {\it pl.} bèkàwà         \\ 
{\bf kÉ}  {\it n} 7/8 fish scale {\it pl.} bèkÉ         \\ 
{\bf kÈ}  {\it v} go, walk         \\ 
{\bf kÉ kÉ kÉ kÉ kÉ}  {\it ideo} depiction of placing objects in a row         \\ 
{\bf kÉ mpfúndÓ}  {\it v} run, go fast         \\ 
{\bf kÈ'È} lè {\it n} 5/6 molar tooth {\it pl.} màkÈ'È         \\ 
{\bf kÉ'È (mákí)}  {\it v} hatch   nkÉgâ      \\ 
{\bf kÈdÈlÈ}  {\it v} gnaw, knabbern   nkÈdálâ   kÈdàlà   \\ 
{\bf kÉlÈ} lè {\it n} 5/6 language {\it pl.} màkÉlÉ         \\ 
{\bf kÈlÈ}  {\it v} hang   nkÈlá   kÈlàlà   \\ 
{\bf kÈmbÈ}  {\it n} 7/8 phlegm {\it pl.} bèkÈmbÈ         \\ 
{\bf kÈndÈ}  {\it n} 7/8 1) voyage, traveling 2) time {\it pl.} bèkÈndÈ         \\ 
{\bf kÈndÈ vúd°û}  {\it n} 7/8 1) once, one time 2) in one go, immediately {\it pl.} bèkÈndÈ bébáà         \\ 
{\bf kénó} lè {\it n} 5/6 blue duiker (Cephalophus monticola) {\it pl.} màkénó         \\ 
{\bf kf°ù} lè {\it n} 5/6 owl {\it pl.} màkf°ù         \\ 
{\bf kfùbàlà}  {\it v} move         \\ 
{\bf kfùbÈ}  {\it v} provoke   nkfùbálâ kfùbÈlÈ provoke  kfùbàlà move   \\ 
{\bf kfúbÓ}  {\it n} 7/8 epilepsy {\it pl.} bèkfúbÓ         \\ 
{\bf kfúbÒ}  {\it n} 1/2 chicken {\it pl.} bàkúbÒ         \\ 
{\bf kfúdÈ} lè {\it n} 5/6 mad person (imbécile) {\it pl.} màkúdÈ         \\ 
{\bf kfúdÈ}  {\it v} cover, put a lid   nkfúdâ   kfúdàlà kfúdègà close oneself  \\ 
{\bf kfúdÓ máb°Ô}  {\it v} kneel         \\ 
{\bf kfùdÓ yá ntÉlÉ}  {\it n} 7/8 old tissue, rag         \\ 
{\bf kfúdÒwÒ}  {\it n} 7/8 chest {\it pl.} bèkúdÒwÒ         \\ 
{\bf kfúl°ÉÉ}  {\it n} 7/8 raffia {\it pl.} bèkfúl°ÉÉ         \\ 
{\bf kfúlà}  {\it n} 7/8 two sticks in monkey trap that hold the trigger {\it pl.} bèkfúlà         \\ 
{\bf kfúlé} lè {\it n} 5/6 paw, sole {\it pl.} màkúlé         \\ 
{\bf kfúlè} lè {\it n} 5/6 hump {\it pl.} màkfúlÈ         \\ 
{\bf kfúlÈ (wà djí)}  {\it n} 1/2 tortoise (of forest) {\it pl.} bàkfúlÈ         \\ 
{\bf kfúlè bìpÈbÈ}  {\it n} 1/2 sea turtle         \\ 
{\bf kfùlÒ}  {\it v} schaben, skin porcupine (soak in hot water, then remove Stacheln)   nkfùlá   kfùlàlà kfùlègà  \\ 
{\bf kfúmá}  {\it n} 1/2 chef, rich {\it pl.} bàkfúmá         \\ 
{\bf kfùmàlà}  {\it v} find   nkfùmá      \\ 
{\bf kfúmbÓ}  {\it n} 7/8 bragging, showing-off {\it pl.} bèkfúmbÓ         \\ 
{\bf kfùmÓ} lè {\it n} 5/6 heap {\it pl.} màkùmÓ         \\ 
{\bf kfùmÓ}  {\it n} 7/8 stump {\it pl.}  bèkfùmÓ         \\ 
{\bf kfùndÈ} lè {\it n} 5/6 garbage {\it pl.} màkfùndÈ         \\ 
{\bf kfúnÓ}  {\it n} 7/8 hornbill {\it pl.} bèkfúnÓ         \\ 
{\bf kfùÒ}  {\it n} 7/8 alstonia tree (alstonia congensis) {\it pl.} bèkfùÒ         \\ 
{\bf kfúzá}  {\it n} 7/8 fist {\it pl.} bèkfúzá         \\ 
{\bf kí} lè {\it n} 5/6 egg {\it pl.} màkí         \\ 
{\bf kì}  {\it v} say         \\ 
{\bf kìkùù} mà {\it n} 0/6 exam         \\ 
{\bf kílì} bè {\it n} 0/8 slyness, guile         \\ 
{\bf kílÒwÒ}  {\it v} be vigilant   nkílásâ be warned  kílÈsÈ make vigilant    \\ 
{\bf kímbÓ} mà {\it n} 0/6 salt         \\ 
{\bf kímì}  {\it n} 1/2 monkey (generic) {\it pl.} bàkímì         \\ 
{\bf kìndá}  {\it n} 7/8 sugar ant {\it pl.} bèkìndá         \\ 
{\bf kíngÈlÈ}  {\it v} become stiff   nkíngálâ      \\ 
{\bf kísínì}  {\it n} 1/2 kitchen {\it pl.} bàkísínì         \\ 
{\bf kìyà}  {\it v} give   nkìyá  kìyÈsÈ chase sb. kìyàlà   \\ 
{\bf kíyÉ}  {\it n} 7/8 iron {\it pl.} bèkíyÉ         \\ 
{\bf kìyÈ}  {\it v} 1) try 2) tempt   -0 kìyÈlÈ taste  kìyàlà taste each other   \\ 
{\bf kó}  {\it n} 1/2 uncle (mother's brother) {\it pl.} bàkó         \\ 
{\bf kÓ} lè {\it n} 5/6 stone {\it pl.} màkÓ         \\ 
{\bf kÔ}  {\it adv} always         \\ 
{\bf kÔ}  {\it v} gather, pluck, pick   nkÓyâ   kÓyàlà kòyàgà  \\ 
{\bf kÓ lé tS°í} lè {\it n} 5/6 nape of neck {\it pl.} màkÓ má ts°í         \\ 
{\bf kÒ'Ò}  {\it n} 7/8 plant species (gnetum africanum) {\it pl.} bèkÒ'Ò         \\ 
{\bf kÓbÈ}  {\it n} 1/2 cup {\it pl.} bàkÓbÈ         \\ 
{\bf kÓbÈ}  {\it v} violate, break (rule)   nkÓbâ   kÓbàlà   \\ 
{\bf kÓdÉ} lè {\it n} 5/6 bend, curve {\it pl.} màkÓdÉ         \\ 
{\bf kÓdÈ}  {\it v} turn sth. (with vehicle)   nkÓdâ  kÓdÈsÈ kÓdàlà kÓdÈgà turn oneself  \\ 
{\bf kÓdÒ}  {\it n} 7/8 yam {\it pl.} bèkÓdÒ         \\ 
{\bf kÒfí}  {\it n} 7/8 coffee {\it pl.} bèkÒfí         \\ 
{\bf kÓgÈ}  {\it v} straighten   nkÓgâ  kÓgÈsÈ kÓgàlà   \\ 
{\bf kÓkÓ}  {\it n} 7/8 emptiness {\it pl.} bèkÓkÓ         \\ 
{\bf kókò} bà {\it n} 0/2 Bakoko         \\ 
{\bf kÓkÓ (yá nlô)}  {\it n} 7/8 skull {\it pl.} bèkÓkÓ bé mínlô         \\ 
{\bf kÓkÓ yá ngwálà}  {\it n} 7/8 snail house {\it pl.} bèkÓkÓ bé bángwálà         \\ 
{\bf kòkù}  {\it n} 7/8 albino {\it pl.} bèkòkù         \\ 
{\bf kòlà}  {\it v} add, lengthen   nkòlá   kòlàlà   \\ 
{\bf kòlÈ}  {\it v} help   nkòlá   kòlàlà   \\ 
{\bf kÒlÈ (mábÓÒ)}  {\it v} stumble      kÒlàlà   \\ 
{\bf kÓndà} lè {\it n} 5/6 sap {\it pl.} màkÓndà         \\ 
{\bf kÓndyì} lè {\it n} 5/6 palm (of hand) {\it pl.} màkÓndyì         \\ 
{\bf kÓsÈ}  {\it v} cough    kÓsÈlÈ make cough  kÓsàlà   \\ 
{\bf kòyà}  {\it n} 7/8 rope Strick {\it pl.} bèkòyà         \\ 
{\bf kpàdà kpàdà}  {\it ideo} depiction of drumming on bamboo         \\ 
{\bf kpÈmÈ}  {\it n} 7/8 manioc leaves {\it pl.} bèkpÈmÈ         \\ 
{\bf kpúdùm kpúdùm}  {\it ideo} depiction of drumming         \\ 
{\bf kù}  {\it n} 1/2 rat {\it pl.} bàkù         \\ 
{\bf kúdÉ}  {\it n} 7/8 skin {\it pl.} bèkúdÉ         \\ 
{\bf kúÈlÈ}  {\it v} mock, make fun of      kúàlà   \\ 
{\bf kùgà}  {\it v} spread, fit, be.enough   nkùgá      \\ 
{\bf kùgúù}  {\it n} 7/8 evening {\it pl.} bèkùgúù         \\ 
{\bf kùgúù bvúÒ}  {\it n} 7/ day before yesterday         \\ 
{\bf kùgúù mgbágà}  {\it n} 7/0 day before yesterday         \\ 
{\bf kúkú}  {\it n} 7/8 mushroom {\it pl.} bèkúkú         \\ 
{\bf kùlÈ}  {\it v} borrow   nkùlá  kùlÈsÈ lend kùlàlà   \\ 
{\bf kúlí}  {\it n} 9/6 "deuil" funeral ceremony from death to burying {\it pl.} màkúlí         \\ 
{\bf kùlì}  {\it n} 7/8 pimpel {\it pl.} bèkùlì         \\ 
{\bf kúlúú} ná {\it n} 1/2 forest tortoise (Kinixys homeana) {\it pl.} bànákúlúú         \\ 
{\bf kùmàsà}  {\it v} prepare   nkùmásâ      \\ 
{\bf kúmbÉ}  {\it n} 7/8 tin {\it pl.} bèkúmbÉ         \\ 
{\bf kùmbÓ}  {\it n} 7/8 womb {\it pl.} bèkùmbÓ         \\ 
{\bf kùmbÒ}  {\it v} repair, reconciliate, arrange, fix   nkùmbá   kùmbàlà   \\ 
{\bf kùnàà}  {\it inv} good         \\ 
{\bf kùndá}  {\it n} 7/8 shoe {\it pl.} bèkùndá         \\ 
{\bf kúndí} lè {\it n} 5/6 mat {\it pl.} màkúndí         \\ 
{\bf kúÓ} lè {\it n} 5/6 Azobé tree, Ironwood tree (lophira alata) {\it pl.} màkúÓ         \\ 
{\bf kùr°â}  {\it n} 7/0 electricity         \\ 
{\bf kùsì}  {\it n} 1/2 parrot {\it pl.} bàkùsì         \\ 
{\bf kùßÊ}  {\it n} 7/6 heritage {\it pl.} màkùßÊ         \\ 
{\bf kúwÓ} lè {\it n} 5/6 flea {\it pl.} màkúwÓ         \\ 
{\bf kw°a} lè {\it n} 5/6 spear, arrow {\it pl.} màkw°a         \\ 
{\bf kw°â}  {\it v} cut raffia leaves in tree   nkwángâ   kwángàlà   \\ 
{\bf kw°â}  {\it v} trahir   kwángâ  kwángÈsÈ kwángàlà   \\ 
{\bf kw°á°àlÈ}  {\it v} spy   nkw°áálâ   kw°álàlà   \\ 
{\bf kw°î}  {\it n} 7/8 Peter's duiker (Cephalophus callipygus) {\it pl.} bèkw°î         \\ 
{\bf kwà}  {\it v} grind (with stone), hollow out canoe   nkwàgá   kwàgàlà   \\ 
{\bf kwádÓ}  {\it n} 7/6 village {\it pl.} màkwádÓ         \\ 
{\bf kwádÒ}  {\it v} twist, bend   nkwádâ    kwádègà turn (vehicle)  \\ 
{\bf kwádÓ písÈ}  {\it n} 7/6 countryside, rural area {\it pl.} màkwádÓ písÈ         \\ 
{\bf kwàlÉ}  {\it n} 7/8 1) love, desire 2) partridge {\it pl.} bèkwàlÉ         \\ 
{\bf kwàlÈ}  {\it v} love, like   nkwàlá   kwàlàlà love the same people   \\ 
{\bf kwálówó} lè {\it n} 5/6 knuckle (hand, foot) {\it pl.} màkwálówó         \\ 
{\bf kwámÓ}  {\it n} 9/6 bag {\it pl.} màkwámÓ         \\ 
{\bf kwàndÒ}  {\it n} 7/8 plantain {\it pl.} bèkwàndÒ         \\ 
{\bf kwánÉ}  {\it n} 7/8 meeting, party {\it pl.} bèkwánÉ         \\ 
{\bf kwànÈ}  {\it v} sell   nkwàná      \\ 
{\bf kwásì} mà {\it n} 0/6 clapping (with hands)         \\ 
{\bf kwàsyó}  {\it n} 0/2 Kwasio people         \\ 
{\bf kwÊ}  {\it n} 7/8 cough {\it pl.} bèkwÊ         \\ 
{\bf kwê}  {\it v} fall, fail (trans.)   nkwéyâ  kùsÈsÈ make fall kwéyàlà   \\ 
{\bf kwÊlÈ}  {\it v} bite   nkwáálâ   kwáàlà   \\ 
{\bf kwÈlÒ}  {\it v} 1) cut down, abattre 2) injure someone   nkwÈlá   kwÈlàlà kwÈlègà (nà só)  \\ 
{\bf kyàlè}  {\it v} start an engine   nkyàlá      \\ 
{\bf kyÉgÈ}  {\it n} 7/8 Basaa {\it pl.} bèkyÉgÈ         \\ 
{\bf kyÈlègà}  {\it v} fall from tree when branch breaks   nkyÈlégâ   kyÈlàlà   \\ 
{\bf kyÈlí}  {\it n} 7/8 bird trap {\it pl.} bèkyÈlí         \\ 
{\bf l°à}  {\it v} pass, overtake, pass by   nlàngá làngÈlÈ let pass, time  làngàlà   \\ 
{\bf l°â}  {\it v} read, count   nlángâ   lángàlà   \\ 
{\bf l°á (yá nyúà)}  {\it n} 7/8 green mamba {\it pl.} bèl°á (bé nyúà)         \\ 
{\bf l°â mímbvû}  {\it n} 1/2 larvaes on a tree {\it pl.} bàl°â mímbvû         \\ 
{\bf l°È}  {\it v} pour in (eingießen)   nlÈngá   lÈngàlà   \\ 
{\bf l°ù} mà {\it n} 0/6 sexual intercourse         \\ 
{\bf l°û} lè {\it n} 5/6 insult {\it pl.} màl°û         \\ 
{\bf l°û}  {\it v} insult   nlúngâ   lúngàlà and lúwàlà   \\ 
{\bf l°ú°à yÉlÈ}  {\it v} whistle   nlÓngâ   lÓngàlà scream  lÓngÈ yÉlÈ  \\ 
{\bf l°ú°Ò}  {\it v} build, construct   nlúngâ  lúngÈsÈ lúngàlà   \\ 
{\bf lá} lè {\it n} 5/6 fish trap {\it pl.} màlá         \\ 
{\bf là} mà {\it n} 0/6 1) meaning 2) support (material, financial) 3) importance         \\ 
{\bf lâ} lè {\it n} 5/6 antenna {\it pl.} màlâ         \\ 
{\bf lâ}  {\it v} harvest, collect honey   nláyâ   léyàlà léyègà  \\ 
{\bf láà}  {\it v} tell    nláwâ   láàlà   \\ 
{\bf láálÈ} * {\it mod} three         \\ 
{\bf làdò nà}  {\it v} meet   nlàdá  làdÈsÈ làdàlà   \\ 
{\bf lágà}  {\it v} contaminate (intr) (disease)   nlágâ  légÈsÈ légàlà   \\ 
{\bf lámbÒ} lè {\it n} 5/6 trap {\it pl.} màlámbÒ         \\ 
{\bf lámbÒ}  {\it n} 7/6 lamp {\it pl.} màlámbÒ         \\ 
{\bf lámbÒ}  {\it v} trap   nlámbâ   lámbàlà   \\ 
{\bf lánà}  {\it v} distribute, unlimited offer   nlánâ   lánàlà   \\ 
{\bf lándè}  {\it n} 7/6 badamier {\it pl.} màlándè         \\ 
{\bf láßÈ}   {\it n} 1/2 big rainy season {\it pl.} bàláßÈ         \\ 
{\bf lé}  {\it n} 7/8 tree, bush {\it pl.} bèlé         \\ 
{\bf lÉ}  {\it n} 7/8 glass {\it pl.} bèlÉ         \\ 
{\bf lÊ}  {\it v} offer   nlÉyâ   lÉyàlà   \\ 
{\bf lÈbÈlÈ}  {\it v} follow, chase   nlÈbálâ   lÈbàlà   \\ 
{\bf lèbvùá}  {\it inv} nine         \\ 
{\bf lÈÈ}  {\it } uproot, disroot   nlÈyá   lÈyàlà   \\ 
{\bf lÉgÈ}  {\it v} singe   nlÉgâ  lÉgÈsÈ lÉgàlà lÉgègà  \\ 
{\bf lèmbò}  {\it v} 1) know 2) flee, escape   nlèmbá escaped  lèmbÈsÈ lèmbàlà   \\ 
{\bf lÉndÉ} lè {\it n} 5/6 palm tree {\it pl.} màlÉndÉ         \\ 
{\bf lÈndÒ}  {\it v} flow   nlÈndá  lÈndÈsÈ  lÈndègà  \\ 
{\bf lÉnÈ}  {\it n} 7/8 offer {\it pl.} bèlÉnÈ         \\ 
{\bf lÉngÒ}  {\it n} 7/8 fun, amusement, joke {\it pl.} bèlÉngÒ         \\ 
{\bf líbÈlà}  {\it v} show up, appear (e.g. moon), auftreten   nlíßálâ      \\ 
{\bf líbÈlÈ}  {\it v} show   nlíbálâ   líbàlà   \\ 
{\bf líÈ}  {\it v} leave (to sb), cede, let   nlígâ   lígàlà   \\ 
{\bf límà}  {\it n} 7/8 stupidity {\it pl.} bèlímà         \\ 
{\bf límbÈ}  {\it v} pull   nlímbâ   límbàlà límbègà se faire tirer  \\ 
{\bf línâ}  {\it inv} since         \\ 
{\bf líndÈ}  {\it inv} when         \\ 
{\bf líßÉlá (má ngÒndÈ)} mà {\it n} 0/6 Aufgang, apparition (of moon)         \\ 
{\bf lìvrÉ}  {\it n} 1/2 book {\it pl.} bàlìvrÉ         \\ 
{\bf líyÈlÈ}  {\it v} accompany   nlíyálâ   líyàlà   \\ 
{\bf líyÒ}  {\it v} clear land   nléyâ   líyàlà líyàgà  \\ 
{\bf lÒ}  {\it v} sew, weave, weave nest   nlÒyá  lÒyÈsÈ lÒyàlà   \\ 
{\bf lÔ} lè {\it n} 5/6 ear {\it pl.} màlÔ         \\ 
{\bf lòá}  {\it n} 7/8 slave, servant {\it pl.} bèlòá         \\ 
{\bf lÒgÒ}  {\it n} 7/8 curse {\it pl.} bèlÒgÒ         \\ 
{\bf lÓlÈ mìnkÒlÉ} n {\it n} 1/2 weaver, taylor {\it pl.} bàlÓlÈ bá mínkÒlÉ         \\ 
{\bf lÒlÒ}  {\it n} 7/8 duck {\it pl.} bèlÒlÒ         \\ 
{\bf lÒmbì}  {\it inv} eight         \\ 
{\bf lÒndÓ}  {\it n} 7/8 ring {\it pl.} bèlÒndÓ         \\ 
{\bf lÓngá}  {\it n} 7/8 group, swarm, flock {\it pl.} bèlÓngá         \\ 
{\bf lÒÓ} lè {\it n} 5/6 dew {\it pl.} màlÒÓ         \\ 
{\bf lùà}  {\it v} curse   nlÒgá  lÒgÈsÈ lÒgàlà   \\ 
{\bf lúmÈ}  {\it v} send   nlúmâ, nlúmálâ lúmÈlÈ  lúmàlà   \\ 
{\bf lùmÓ}  {\it n} 7/8 yellow fever mosquito (Aedes aegypti) {\it pl.} bèlùmÓ         \\ 
{\bf lùndá}  {\it n} 7/8 small forest, grove between villages and houses {\it pl.} bèlùndá         \\ 
{\bf lúndÉ}  {\it n} 7/8 tree species (afzelia bipidensis) {\it pl.} bèlúndÉ         \\ 
{\bf lúndÒ}  {\it v} fill oneself (with food)   nlúndá lúndÈlÈ fill sth lúndÈsÈ lúndàlà   \\ 
{\bf lùNgà}  {\it v} grow   nlùngá  lùngÈsÈ mwánÒ    \\ 
{\bf lùngÈlÈ}  {\it v} aim (at)   nlùngálâ   lùngàlà   \\ 
{\bf lúwÒ}  {\it v     } bite   nlùwá  lúwÈsÈ lúwàlà   \\ 
{\bf lvùgà} mà {\it n} 0/6 animation, liveliness         \\ 
{\bf lvúmÓ}  {\it n} 7/8 maggot {\it pl.} bèlvúmÓ         \\ 
{\bf lvúmÒ}  {\it v} sting   nlvúmâ  lvúmÈsÈ lvúmàlà fight in war   \\ 
{\bf lvùúgÒ}  {\it v} animate, excite mvàgá        \\ 
{\bf m°â}  {\it n} 0/6 sea         \\ 
{\bf má'à}  {\it v} accuse   mágâ mándÈlÈ  mágàlà  mágÈ \\ 
{\bf máá} lè {\it n} 5/6 cheek {\it pl.} màmáá         \\ 
{\bf màbè}  {\it n} 0/6 Mabi people         \\ 
{\bf màbùnzò}  {\it n} 1/2 lion (Kwasio word) {\it pl.} bàmábùnzò         \\ 
{\bf màkítì}  {\it n} 0/6 market         \\ 
{\bf mámé}  {\it n} 1/2 aunt (ather's sister) {\it pl.} bàmámé         \\ 
{\bf mándÈlÈ}  {\it v} accuse (interchangeable with máá)         \\ 
{\bf màndjìmÒ}  {\it inv} whole, entire         \\ 
{\bf mándÒ}  {\it v} Mund vollstopfen   mándâ mándÈlÈ accuse repeatedly  mándàlà accuse   \\ 
{\bf mángÒ(mángÒ)} ná {\it n} 1/2 male lizard {\it pl.} bànámángÒ(mángÒ)         \\ 
{\bf màNgÓlÓ} lè {\it n} 5/6 ankle {\it pl.} màmàNgÓlÓ         \\ 
{\bf mânù}  {\it n} 0/6 junction, crossroad         \\ 
{\bf másà}  {\it n} 1/2 boss {\it pl.} bàmásà         \\ 
{\bf másÉgá}  {\it adv} suddenly, unexpectedly         \\ 
{\bf màtèlà}  {\it n} 1/2 mattress {\it pl.} bàmàtèlà         \\ 
{\bf mátʃà màtʃà}  {\it ideo} depiction of eating in little bits         \\ 
{\bf màtúà}  {\it n} 1/2 car {\it pl.} bàmàtúà         \\ 
{\bf mb°è}  {\it n} 1/2 flood {\it pl.} bàmb°è         \\ 
{\bf mb°ì} lè {\it n} 5/6 pillar {\it pl.} màmb°ì         \\ 
{\bf mb°ù}  {\it n} 7/8 Ochsenfrosch {\it pl.} bèmb°ù         \\ 
{\bf mbáálÓ}  {\it n} 3/4 jaw {\it pl.} mìmbáálÓ         \\ 
{\bf mbàdÓ}  {\it n} 3/4 lake {\it pl.} mìmbàdÓ         \\ 
{\bf mbàfùmbÒ}  {\it n} 3/4 shrew  {\it pl.} mìmbàfùmbÒ         \\ 
{\bf mbágò}  {\it n} 3/4 package, envelope {\it pl.} mìmbágò         \\ 
{\bf mbàmbà}  {\it n} 3/4 co-wife {\it pl.} mìmbàmbà         \\ 
{\bf mbàmbàlà(mbàmbàlà)} ná {\it qual} white         \\ 
{\bf mbámbÉ}  {\it n} 1/2 grand-parent, ancestor {\it pl.} bàmbámbÉ         \\ 
{\bf mbàmbìlì}  {\it n} 1/2 father-in-law {\it pl.} bàmbàmbìlì         \\ 
{\bf mbàNgá}  {\it n} 3/4 nut, pit, stone {\it pl.} mìmbàNgá         \\ 
{\bf mbàNgá lÉ lÉndÍ}  {\it n} 3/4 coconut {\it pl.} mìmbàNgá mí màlÉndí         \\ 
{\bf mbÈ}  {\it n} 3/4 drum {\it pl.} mìmbÈ         \\ 
{\bf mbÊ}  {\it n} 3/4 door {\it pl.} mìmbÊ         \\ 
{\bf mbÉÉ}  {\it n} 3/4 metal oven {\it pl.} mìmbÉÉ         \\ 
{\bf mbèlè}  {\it n} 3/4 African Padauk, African Coralwood (Pterocarpus soyauxii ) {\it pl.} mìmbèlè         \\ 
{\bf mbènÈ}  {\it n} 9/6  bad sign, omen {\it pl.} màmbènÈ         \\ 
{\bf mbéwÒ}  {\it n} 3/4 selfishness, sin {\it pl.} mìmbéwÒ         \\ 
{\bf mbìmbó}  {\it n} 3/4 corps {\it pl.} mìmbìmbó         \\ 
{\bf mbÒlÈ}  {\it n} 3/4 okra {\it pl.} mìmbÒlÈ         \\ 
{\bf mbÒlÉ kfúnó}  {\it n} 7/8 slime {\it pl.} bèmbÒlÉ kfúnÓ         \\ 
{\bf mbÒmbÓ}  {\it n} 9/6 face {\it pl.} màmbÒmbÓ         \\ 
{\bf mbÒmbÓ}  {\it n} 3/4 daughter-in-law {\it pl.} mìmbÒmbÓ         \\ 
{\bf mbómò}  {\it n} 3/4 eldest (in village) {\it pl.} mìmbómò         \\ 
{\bf mbóndí}  {\it n} 3/4 oil (for cooking) {\it pl.} mìmbóndí         \\ 
{\bf mbÒngÒ}  {\it n} 7/6 plant         \\ 
{\bf mbÒÒ}  {\it n} 3/4 fatness {\it pl.} mìmbÒÒ         \\ 
{\bf mbÒsàwà}  {\it n} 3/4 wetness {\it pl.} mìmbÒsàwà         \\ 
{\bf mbúlá}  {\it n} 7/6 debt {\it pl.} màmbúlá         \\ 
{\bf mbúlÈ wá sí}  {\it n} 3/4 blister {\it pl.} mìmbúlÈ mí sí         \\ 
{\bf mbúlÒ}  {\it n} 3/4 migratory locust (Locusta migratoria) {\it pl.} mìmbúlÒ         \\ 
{\bf mbúmbá}  {\it n} 3/4 froissement (wrinkled clothes) {\it pl.} mìmbúmbá         \\ 
{\bf mbúmbù}  {\it n} 1/2 namesake {\it pl.} bàmbúmbù         \\ 
{\bf mbùNgá}  {\it n} 7/8 earring {\it pl.} bèmbùNgá         \\ 
{\bf mbùngù}  {\it n} 0/2 Yassa         \\ 
{\bf mbvú}  {\it n} 3/0 white/grey hair         \\ 
{\bf mbvû}  {\it n} 3/4 year {\it pl.} mìmbvû         \\ 
{\bf mbvùlÈ}  {\it n} 7/8 bushbuck (Tragelaphus scriptus) {\it pl.} bèmbvùlÈ         \\ 
{\bf mbvúlÈ síyÈ}  {\it n} 7/8 black stuff of fire: Ruß {\it pl.} bèmbvúlÈ bé síyÈ         \\ 
{\bf mbvúndá}  {\it n} 9/6 trouble, error, mistake {\it pl.} màmbvúndá         \\ 
{\bf mbvúndyÈ} lè {\it n} 5/6 débris de feuille (used for hiding trap) {\it pl.} màmbvúndyÈ         \\ 
{\bf mbvúÒ}  {\it n} 1/2 rain {\it pl.} bàmbvúÒ         \\ 
{\bf mbvúÒ wà mbvú}  {\it  n} 1/2 drizzle (rain of white hair) {\it pl.} bàmbvúÒ bá mbvú         \\ 
{\bf mbvúÒ wà nÉnÈ}  {\it n} 1/2 strong rain {\it pl.} bàmbvúÒ bá nÉnÈ         \\ 
{\bf mbwâ}  {\it n} 3/4 tuber, bulb {\it pl.} mìmbwâ         \\ 
{\bf mbwàmbÒ}  {\it n} 3/4 bundle, package, etwas Zusammengefaltetes {\it pl.} mìmbwàmbÒ         \\ 
{\bf mbwàmÒ}  {\it n} 3/4 stay with woman in other compound, adultery {\it pl.} mìmbwàmÒ         \\ 
{\bf mbwàmÒ}  {\it n} 1/2 python {\it pl.} bàmbwàmÒ         \\ 
{\bf mbwE}  {\it n} 1/2 dog {\it pl.} bàmbwE         \\ 
{\bf mbyÊ}  {\it n} 3/4 high, up-stream {\it pl.} mìmbyÊ         \\ 
{\bf mÉÈlÈ}  {\it v} accept, respond, reply   mÉÉlâ   mÉàlà   \\ 
{\bf mÉmÉdÉ wà bèsâ}  {\it n} 1/2 owner {\it pl.} bÉbÉdÉ bá bésâ         \\ 
{\bf mÈmÒ}  {\it v} admit   mèmá   mÈmàlà   \\ 
{\bf mÉnÓ}  {\it n} 7/8 morning {\it pl.} bèmÉnÓ         \\ 
{\bf mÈsÒ}  {\it v} wave (greeting)   mÈsá   mÈsàlà   \\ 
{\bf mÈvâ}  {\it n} 7/0 pride         \\ 
{\bf mfû}  {\it n} 3/4 poison {\it pl.} mìmfû         \\ 
{\bf mfùlÈ}  {\it n} 3/4 fart {\it pl.} mìmfùlÈ         \\ 
{\bf mgb°à}  {\it n} 7/8 crow {\it pl.} bèmgb°à         \\ 
{\bf mgbámàlà} mà {\it n} 0/6 acidity         \\ 
{\bf mgbámàlà}  {\it v} be sour         \\ 
{\bf mgbásá}  {\it n} 7/8 hunting with spear and dogs {\it pl.} bèmgbásá         \\ 
{\bf mgbÈNmbÈmÈ}  {\it n} 7/8 lion (Bagyeli word) {\it pl.} bèmgbÈmgbÈmÈ         \\ 
{\bf mgbísì}  {\it n} 3/4 freshness, rawness, living {\it pl.} mìmgbísì         \\ 
{\bf mímbà}  {\it v} brag      mímbàlà   \\ 
{\bf mìnk°í}  {\it n} 1/2 pot, casserole {\it pl.} bàmìnk°í         \\ 
{\bf mìnò}  {\it v} swallow   mìná  mìnÈsÈ mìnàlà   \\ 
{\bf mínsÒgÈ} nà {\it n} 1/2 palm rat {\it pl.} bànàmínsógÈ         \\ 
{\bf mìntùlí}  {\it n} 1/2 mouse {\it pl.} bàmìntùlí         \\ 
{\bf mìnú} lè {\it n} 5/6 gill {\it pl.} màmìnú         \\ 
{\bf míyù}  {\it n} 1/2 brother, cousin, close friend (younger or same age)         \\ 
{\bf mkpámá}  {\it n} 3/4 novelty {\it pl.} mìnkpámá         \\ 
{\bf mÒ}  {\it n} 3/4 stomach {\it pl.} mìmÒ         \\ 
{\bf mÒnÉ}  {\it n} 1/2 money {\it pl.} bàmÒnÉ         \\ 
{\bf mÓngÒ} lè {\it n} 5/6 male driver ant (Dorylinae) {\it pl.} màmÓngÒ         \\ 
{\bf mp°û}  {\it n} 3/4 payment {\it pl.} mìmp°û         \\ 
{\bf mpá}  {\it n} 3/4 island {\it pl.} mìmpá         \\ 
{\bf mpà}  {\it n} 1/2 1) galago de Thomas 2) virgin {\it pl.} bàmpà         \\ 
{\bf mpà'à}  {\it n} 9/0 vapor, fog         \\ 
{\bf mpá'à wá nyúlÊ}  {\it n} 3/4 side of the body {\it pl.} mìmpá 'à mí nyúlÊ         \\ 
{\bf mpà(mpà)}  {\it qual} good         \\ 
{\bf mpàálé}  {\it n} 9/6 news {\it pl.} màmpàálé         \\ 
{\bf mpàgó}  {\it n} 3/4 road {\it pl.} mìmpàgó         \\ 
{\bf mpàmbìlì}  {\it n} 3/4 Sturz {\it pl.} mìmpàmbìlì         \\ 
{\bf mpàndà}  {\it n} 1/2 bug (all larger, rounder insects) {\it pl.} bàmpàndà         \\ 
{\bf mpàndyè}  {\it n} 7/8 bamboo {\it pl.} bèmpàndyè         \\ 
{\bf mpàndyì}  {\it n} 7/6 rib {\it pl.} màmpàndyí         \\ 
{\bf mpèlÈ}  {\it n} 1/2 eagle {\it pl.} bàmpèlÈ         \\ 
{\bf mpèndÈ}  {\it n} 3/4 root {\it pl.} mìmpèndÈ         \\ 
{\bf mpÈwÓ}  {\it n} 3/4 wind {\it pl.} mìmpÈwÓ         \\ 
{\bf mpfùmbÒ}  {\it n} 3/4 dead tree (without leaves) {\it pl.} mìmpfùndÒ         \\ 
{\bf mpfùmÒ}  {\it n} 3/4 midnight {\it pl.} mìmpfùmÒ         \\ 
{\bf mpfúndÓ}  {\it n} 3/4 running, race {\it pl.} mìmpfúndÓ         \\ 
{\bf mpfùngyá'à}  {\it n} 3/4 dust {\it pl.} mìmpfùngyá'à         \\ 
{\bf mpfùÒ}  {\it n} 3/4 last meal with medicine in a healing session {\it pl.} mìmpfùÒ         \\ 
{\bf mpí'ìdì}  {\it n} 9/6 heat (from fire, pot, people) {\it pl.} màmpí'ìdì         \\ 
{\bf mpíì}  {\it n} 3/4 kidney {\it pl.} mìmpíì         \\ 
{\bf mpìmbá}  {\it n} 7/8 pancreas {\it pl.} bèmpìmbá         \\ 
{\bf mpìnàgà}  {\it n} 3/4 obligation, duty {\it pl.} mìmpìnàgà         \\ 
{\bf mpíndá}  {\it n} 9/6 law, prohibition {\it pl.} màmpíndá         \\ 
{\bf mpìndí}  {\it n} 3/4 non-ripeness {\it pl.} mìmpìndí         \\ 
{\bf mpìndì}  {\it n} 9/6 dirt {\it pl.} màpìndì         \\ 
{\bf mpìndyÓ}  {\it n} 3/4 trigger in trap {\it pl.} mìmpìndyÓ         \\ 
{\bf mpìNgá}  {\it n} 3/4 sweet cassava {\it pl.} mìmpìNgá         \\ 
{\bf mpö}  {\it n} 1/2 héloisciure {\it pl.} bàmpö         \\ 
{\bf mpÒ'Ò}  {\it n} 3/4 Zahnlücke {\it pl.} mìmpÒ'Ò         \\ 
{\bf mpÓndÓ}  {\it n} 3/4 shirt {\it pl.} mìmpÓndÓ         \\ 
{\bf mpÒNgÓ}  {\it n} 9/6 seedling {\it pl.} màmpÒNgÓ         \\ 
{\bf mpòngóló}  {\it n} 7/8 ginger plant (aframomum) {\it pl.} bèmpòngóló         \\ 
{\bf mpù}  {\it ?} like (this)         \\ 
{\bf mpúbÉlÈ}  {\it n} 1/2 current, rip tide {\it pl.} bàmpúbÉlÈ         \\ 
{\bf mpúdÉ}  {\it n} 3/4 mais {\it pl.} mìmp´dÉ         \\ 
{\bf mpúÈrÉ}  {\it inv} seven         \\ 
{\bf mpùlÉ}  {\it n} 3/4 tree with yellow bark (enantia chlorantha); yellow color {\it pl.} mìmpùlÉ         \\ 
{\bf mpúmbú}  {\it n} 3/4 calf {\it pl.} mìmpúmbú         \\ 
{\bf mpwá}  {\it n} 3/4 bouillon (water, salt, chili) {\it pl.} mìmpwá         \\ 
{\bf mtʃà mtʃà mtʃà}  {\it ideo} depiction of picky eating (only taking certain items off the plate)         \\ 
{\bf múÈlÈ}  {\it v} nibble   múálâ  múÈsÈ múàlà   \\ 
{\bf múNgÈ} lè {\it n} 5/6 beetle (Buprestidae) {\it pl.} màmúNgÈ         \\ 
{\bf músÓ}  {\it n} 7/8 midday, noon {\it pl.} bèmúsÓ         \\ 
{\bf mvÉbÉ}  {\it n} 7/8 hedgehog {\it pl.} bèmvÉbÉ         \\ 
{\bf mvíndÓ}  {\it n} 3/4 sweet water turtle {\it pl.} mìmvíndÓ         \\ 
{\bf mwádèk°â}  {\it n} 7/8 other side {\it pl.} bèmwádèk°â         \\ 
{\bf mwálÉ}  {\it n} 3/4 female {\it pl.} mìmwálÉ         \\ 
{\bf mwàNgÓlÓ} lè {\it n} 5/6 joint {\it pl.} màmwàNgÓlÓ         \\ 
{\bf mwàsÒ}  {\it n} 3/4 long bendable stick in trap that holds animal {\it pl.} mìmwàsÒ         \\ 
{\bf mwàsÒ}  {\it v} throw   mwàsá   mwàsàlà   \\ 
{\bf mwÉ}  {\it n} 3/4 dam, barrage {\it pl.} mìmwÉ         \\ 
{\bf myàkÈ}  {\it v} sprinkle   myàká  myàkèsÈ myàkàlà   \\ 
{\bf myámàtà}  {\it v} be narrow, narrow sth.   myámátâ      \\ 
{\bf myámÒ}  {\it v} knead, press (dough or fruit), press between fingers      myámàlà sich kneten   \\ 
{\bf myángálÈ}  {\it n} 3/4 rust {\it pl.} mìmyángálÈ         \\ 
{\bf myÉ}  {\it n} 0/4 fur         \\ 
{\bf n°â} * {\it mod} four         \\ 
{\bf ná}  {\it adv} still, again         \\ 
{\bf nà}  {\it com} and, with         \\ 
{\bf nâ}  {\it comp} that         \\ 
{\bf nà kùgúù}  {\it ?} yesterday         \\ 
{\bf nà mÉnÓ}  {\it ?} tomorrow         \\ 
{\bf náàtà nà}  {\it v} stick (sth.), be sticky   nátâ      \\ 
{\bf nábànkúdí} ná {\it n} 1/2 female lizard {\it pl.} bànábànkúdí         \\ 
{\bf nákÓ}  {\it n} 3/6 earwax {\it pl.} màkÓ         \\ 
{\bf náNgá} lè {\it n} 5/6 star {\it pl.} mànáNgá         \\ 
{\bf nd°ÉÉÉÉ}  {\it ideo} depiction of staring         \\ 
{\bf ndà}  {\it v} cross   ndàngá   ndàngàlà   \\ 
{\bf ndáà}  {\it adv} also, too         \\ 
{\bf ndàlò}  {\it n} 1/2 tobacco {\it pl.} bàndàlò         \\ 
{\bf ndáwÒ}  {\it n} 9/6 house {\it pl.} màndáwÒ         \\ 
{\bf ndÈ}  {\it n} 3/4 bait {\it pl.} mìndÈ         \\ 
{\bf ndÈmó}  {\it n} 9/6 dream {\it pl.} màntÈmó         \\ 
{\bf ndísì}  {\it n} 3/4 rice {\it pl.} mìndísì         \\ 
{\bf ndjímbà}  {\it n} 3/4 ignorance {\it pl.} mìndjìmbá         \\ 
{\bf ndjìmÒ}  {\it n} 3/4 some, irgendein, any {\it pl.} mìndjìmÒ mí bùdì         \\ 
{\bf ndjù} lè {\it n} 5/6 banana (sweet) {\it pl.} màndjù         \\ 
{\bf ndjw°â} lè {\it n} 5/6 eggplant {\it pl.} màndjw°â         \\ 
{\bf ndtÉndíbù}  {\it n} 1/2 spider, spider web {\it pl.} bàndtÉndíbù         \\ 
{\bf ndtúá}  {\it n} 7/8 clitoris {\it pl.} bèndúá         \\ 
{\bf ndùwÓ}  {\it n} 3/4 roof {\it pl.} mìndùwÓ         \\ 
{\bf ndvùÓ}  {\it n} 7/8 suffering, difficulty {\it pl.} bèndvùÓ         \\ 
{\bf ndvùù}  {\it n} 3/4 bad luck, bad event {\it pl.} mìndvùù         \\ 
{\bf ndwàmbÈLÈ}  {\it n} 3/4 demande exxagéré {\it pl.} mìndwàmbÈlÈ         \\ 
{\bf ndyándyà wá !múdÌ}  {\it n} 3/4 giant (tall person) {\it pl.} mìndyándyà mí búdÌ         \\ 
{\bf ndyàwÒ}  {\it n} 7/6 chisel {\it pl.} màndyàwÒ         \\ 
{\bf ndyúà}  {\it n} 3/4 swimming {\it pl.} mìndyúà         \\ 
{\bf ndz°à}  {\it n} 9/6 dance {\it pl.} màndz°à         \\ 
{\bf ndz°í}  {\it n} 9/6 jealousy, envy {\it pl.} mándz°í         \\ 
{\bf ndz°ì}  {\it n} 1/2 fly {\it pl.} bàndz°ì         \\ 
{\bf ndzà}  {\it n} 9/6 hunger {\it pl.} màndzà         \\ 
{\bf ndzààlÉ}  {\it n} 1/2 tree pangolin (Manis tricuspis) {\it pl.} bàndzààlÉ         \\ 
{\bf ndzámbÒ}  {\it n} 7/6 upper arm {\it pl.} màndzámbÒ         \\ 
{\bf ndzàmbÒ}  {\it n} 7/8 mud {\it pl.} bèndzàmbÒ         \\ 
{\bf ndzE}  {\it n} 1/2 panther, leopard {\it pl.} bàndzE         \\ 
{\bf ndzÉlì (yá mísì)}  {\it n} 7/8 hair in face (beard, around eyes) {\it pl.} bèndzÉlì (bé mísì)         \\ 
{\bf ndzi}  {\it n} 9/6 path {\it pl.} màndzi         \\ 
{\bf ndzìÈ}  {\it n} 1/2 gorilla {\it pl.} bàndzìÈ         \\ 
{\bf ndzìlì}  {\it n} 1/2 guard {\it pl.} bàndzìlì         \\ 
{\bf ndzílí yá mbÔ}  {\it n} 7/8 elbow {\it pl.} bèndzílí mábÔ         \\ 
{\bf ndzìmózó}  {\it n} 1/2 guard {\it pl.} bàndìmózó         \\ 
{\bf ndzìwÒ}  {\it n} 1/2 céphalophe géant {\it pl.} bàndzìwÒ         \\ 
{\bf ndzólÈ} lè {\it n} 5/6 tear {\it pl.} màndzólÈ         \\ 
{\bf nÉnÈ}  {\it qual} big         \\ 
{\bf ng°à(ng°á)}  {\it n} 1/2 healer {\it pl.} bàNg°à(Ng°á)         \\ 
{\bf ng°Ê}  {\it n} 9/6 field, garden {\it pl.} màng°Ê         \\ 
{\bf ng°Òlíng°Ólì}  {\it n} 7/8 throat, larynx {\it pl.} bèng°Òlíng°Ólì         \\ 
{\bf ng°ù°Ò}  {\it n} 7/8 tomato {\it pl.} bèng°ù°Ò         \\ 
{\bf ngálÈ}  {\it n} 1/2 thunder, lightning, melmel {\it pl.} bàngálÈ         \\ 
{\bf ngàmbàlà}  {\it n} 7/6 rarity, difficulty {\it pl.} màngàmbàlà         \\ 
{\bf ngámbÉ}  {\it n} 7/6 vision, oracle {\it pl.} màngámbÉ         \\ 
{\bf ngàtà}  {\it n} 9/6 bandage, action d'enrouler {\it pl.} màngàtà         \\ 
{\bf ngÈ'È}  {\it n} 7/8 eyebrow {\it pl.} bènkÈ'È         \\ 
{\bf ngÈlÉnÈ}  {\it n} 1/2 English person {\it pl.} bàngÈlÉnÈ         \\ 
{\bf ngÒ}  {\it n} 9/6 grinding stone plate {\it pl.} màngÒ         \\ 
{\bf ngö}  {\it n} 1/2 pig {\it pl.} bàngö         \\ 
{\bf ngó wà djí}  {\it n} 1/2 phacochère commun, sanglier {\it pl.} bàngó bá djí         \\ 
{\bf ngÓkÒbÉ}  {\it n} 7/8 bracelet {\it pl.} bèngÓkÒbÉ         \\ 
{\bf ngòmbáà}  {\it n} 1/2 lemon {\it pl.} bàngòmbáà         \\ 
{\bf ngÒmbÒ}  {\it n} 1/2 monitor lizard {\it pl.} bàngÒmbÒ         \\ 
{\bf ngÒmÒ}  {\it n} 9/6 tam tam (small drum) {\it pl.} màngÒmÒ         \\ 
{\bf ngÒndÈ}  {\it n} 1/2 moon, month {\it pl.} bàngÒndÈ         \\ 
{\bf ngÒNgÒlÈ}  {\it n} 7/0 sadness (about lack), compassion         \\ 
{\bf ngóvìnà}  {\it n} 1/2 government {\it pl.} bàngóvìnà         \\ 
{\bf ngùlá}  {\it n} 3/4 headscarf {\it pl.} mìngùlá         \\ 
{\bf ngùmbÓ}  {\it n} 1/2 porcupine {\it pl.} bàngùmbÓ         \\ 
{\bf ngùndyá}  {\it n} 9/6 raffia leaf when used for weaving {\it pl.} màngùndyá         \\ 
{\bf ngùÓ}  {\it n} 7/8 sugar (cane) {\it pl.} bèngùÓ         \\ 
{\bf ngv°ú°Ò}  {\it n} 1/2 storm, tornado {\it pl.} bàngv°ú°Ò         \\ 
{\bf ngvù}  {\it n} 1/2 hypsingathe monstrueux, flying squirrel {\it pl.} bàngvù         \\ 
{\bf ngvùbÓ}  {\it n} 1/2 hippopotamus {\it pl.} bàngvùbÓ         \\ 
{\bf ngvúlÈ}  {\it n} 9/6 strength, force {\it pl.} màngvúlÈ         \\ 
{\bf ngvúmà}  {\it n} 1/2 some, irgendein (unspecified, unknown) {\it pl.} bàngvúmà         \\ 
{\bf ngvúmbÒ} mà {\it n} 0/6 flirt, attention seeking         \\ 
{\bf ngvùmbÒ}  {\it n} 0/2 Ngumba people         \\ 
{\bf ngvùndÈ}  {\it n} 7/8 mask {\it pl.} bèngvùndÈ         \\ 
{\bf ngvùndÒ}  {\it n} 9/6 vengence {\it pl.} màngvùndÒ         \\ 
{\bf ngvúngvúlÓ}  {\it n} 3/4 bush cricket (Tettigoniidae) + grasshopper (Zonocerus) {\it pl.} mìngvúngvúlÓ         \\ 
{\bf ngvúú}  {\it n} 7/8 shyness {\it pl.} bèngvúú         \\ 
{\bf ngw°E}  {\it n} 1/2 millipede {\it pl.} bàngw°E         \\ 
{\bf ngwálà}  {\it n} 1/2 snail {\it pl.} bàngwálà         \\ 
{\bf ngwálÒ}  {\it n} 7/6 side, next, corner {\it pl.} màngwálÒ         \\ 
{\bf ngwámé}  {\it n} 7/0 danger         \\ 
{\bf ngwàndÓ}  {\it n} 3/4 melon seed (pistache) {\it pl.} mìngwàndÓ         \\ 
{\bf ngwáwà}  {\it n} 7/8 guava {\it pl.} bèngwáwà         \\ 
{\bf ngwáwÒ}  {\it v} bend (only animate), bow   ngwáwâ  ngwàngÈsÈ     \\ 
{\bf ngwÉlÈ}  {\it n} 9/0 withcraft         \\ 
{\bf ngy°Ê}  {\it n} 3/4 visit {\it pl.} mìngy°Ê         \\ 
{\bf ngyà}  {\it n} 3/4 intestines {\it pl.} mìngyà         \\ 
{\bf ngyámànÈ}  {\it n} 7/0 Germany         \\ 
{\bf ngyàngÓ}  {\it n} 7/8 hunt (with gun) {\it pl.} bèngyàngÓ         \\ 
{\bf ngyE} mì {\it n} 0/4 hunting rats (in holes)         \\ 
{\bf ngyÉmÒ}  {\it n} 3/4 fruit bat  {\it pl.} mìngyÉmÒ         \\ 
{\bf ngyÉsá}  {\it n} 7/8 cake (gateau) {\it pl.} bèngyÉsá         \\ 
{\bf ngyÓwÒ}  {\it n} 3/4 hook {\it pl.} mìngyÓwÒ         \\ 
{\bf ngyùlè}  {\it n} 3/4 light {\it pl.} mìngyùlè         \\ 
{\bf ngyùlè wá vísÓ}  {\it n} 3/4 sunlight {\it pl.} mìngùlè mí vísÓ         \\ 
{\bf níÈ}  {\it v} be beautiful     níngÈsÈ    \\ 
{\bf níí}  {\it n} 7/8 vagina {\it pl.} bèníí         \\ 
{\bf níndyà}  {\it v} urinate     níndyÈsÈ (fait pisser like beer) níndyàlà   \\ 
{\bf níyÈ}  {\it inv} how many         \\ 
{\bf nj°û}  {\it n} 7/8 gall bladder, gall {\it pl.} bènj°û         \\ 
{\bf njì}  {\it v} come   njìyá      \\ 
{\bf njí nà}  {\it v} bring (come with)         \\ 
{\bf njímí}  {\it n} 1/2 blind person {\it pl.} bànjímí         \\ 
{\bf njÓ'Ò}  {\it n} 1/2 elephant {\it pl.} bàndjÓ'Ò         \\ 
{\bf njú}  {\it n} 7/8 gap between incisor teeth {\it pl.} bènjú         \\ 
{\bf nk°â}  {\it n} 3/4 guinea fowl {\it pl.} mìnk°â         \\ 
{\bf nk°Ô}  {\it n} 3/4 back {\it pl.} mìnk°Ô         \\ 
{\bf nk°ú°Ò}  {\it n} 3/4 trahison {\it pl.} mìnk°ú°Ò         \\ 
{\bf nk°ú°Ò bùdì}  {\it n} 1/2 traître {\it pl.} bàk°ú°Ò bá bùdì         \\ 
{\bf nká}  {\it n} 3/4 line, row {\it pl.} mìnká         \\ 
{\bf nká'à}  {\it n} 3/4 colobe bai d'Afrique centrale, colobe bai de Tollon, Mangabey noir {\it pl.} mìnkâ         \\ 
{\bf nkáálè}  {\it n} 3/4 vertebrate {\it pl.} mìnkáálè         \\ 
{\bf nkáálÓ}  {\it n} 3/4 fence {\it pl.} mìnkáálÓ         \\ 
{\bf nkáálÓ}  {\it n} 3/4 African/Guinea pepper tree (Xylopia aethiopica) {\it pl.} mìnkáálÓ         \\ 
{\bf nkábÉ}  {\it n} 9/6 paddle {\it pl.} mànkábÉ         \\ 
{\bf nkàdÈ}  {\it n} 3/4 provocation {\it pl.} mìnkàdÈ         \\ 
{\bf nkágá}  {\it n} 3/4 side of an animal         \\ 
{\bf nkámbílí}  {\it n} 3/4 gekautes (Gräten, Knochen), das überbleibt {\it pl.} mìnkámbílí         \\ 
{\bf nkàmÈ}  {\it n} 3/4 sticky sap (from vein, used for birdlime) {\it pl.} mìnkàmÈ         \\ 
{\bf nkàmÒ}  {\it n} 9/0 reason         \\ 
{\bf nkándâ}  {\it n} 3/4 crack {\it pl.} mìnkándâ         \\ 
{\bf nkàndÉ}  {\it n} 1/2 African dwarf crocodile (Osteolaemus tetraspis) {\it pl.} bànkàndÉ         \\ 
{\bf nkándÒ}  {\it n} 3/4 beer {\it pl.} mìnkándÒ         \\ 
{\bf nkângà}  {\it n} 1/2 weaver bird {\it pl.} bànkângà         \\ 
{\bf nkázá}  {\it n} 3/4 whip {\it pl.} mìnkázá         \\ 
{\bf nkÈ}  {\it n} 3/4 low, down-stream {\it pl.} mìnkÈ         \\ 
{\bf nkÉ'É}  {\it n} 3/4 jaw {\it pl.} mìnkÉ'É         \\ 
{\bf nké'é}  {\it n} 7/8 scream {\it pl.} bènké'é         \\ 
{\bf nkÈ'È}  {\it n} 3/4 chin {\it pl.} mìnkÈ'È         \\ 
{\bf nkÉdÉ} lè {\it n} 5/6 hip, waist {\it pl.} mànkÉdÉ         \\ 
{\bf nkÉdÉ}  {\it n} 9/6 courage {\it pl.} mànkÉdÉ         \\ 
{\bf nkÈlÈ yá dìsí}  {\it n} 7/8 eyebrow {\it pl.} bènkÈlÈ bí mísì         \\ 
{\bf nkfù lé lÔ}  {\it n} 3/4 whole in ear {\it pl.} mìnkù mí màlÔ         \\ 
{\bf nkfúdÉ}  {\it n} 7/8 cloud, fog {\it pl.} bènkfúdÉ         \\ 
{\bf nkfùndÉ}  {\it n} 3/4 barren woman {\it pl.} mìnkfùndÉ         \\ 
{\bf nkfùßÓ}  {\it n} 3/4 trunk (body) {\it pl.} mìnkfùßÓ         \\ 
{\bf nkfúù}  {\it n} 3/4 ghost {\it pl.} mìnkfúù         \\ 
{\bf nkfùwÓ}  {\it n} 3/4 torso  {\it pl.} mìnkvùwÓ         \\ 
{\bf nkìngù}  {\it n} 3/4 hem {\it pl.} mìnkìngù         \\ 
{\bf nkìngù}  {\it n} 3/4 edge {\it pl.} mìnkìngù         \\ 
{\bf nkìyÓ}  {\it n} 3/4 wave {\it pl.} mìnkìyÓ         \\ 
{\bf nkÒlÉ}  {\it n} 3/4 vein, rope, line {\it pl.} minkÒlÉ         \\ 
{\bf nkÓlÒ}  {\it n} 3/4 watch, clock {\it pl.} mìnkÓlÒ         \\ 
{\bf nkÓngÓ}  {\it n} 3/4 frog (general term) {\it pl.} mìnkÓngÓ         \\ 
{\bf nkÓsâ}  {\it n} 3/4 manner of coughing {\it pl.} mìnkÓsâ         \\ 
{\bf nkÓzì}  {\it n} 7/8 part of throat of animal that gets removed after killing {\it pl.} bènkÓzì         \\ 
{\bf nkù}  {\it n} 3/4 hole, animal den {\it pl.} mìnkù         \\ 
{\bf nkû}  {\it n} 1/2 orytérope {\it pl.} bànkû         \\ 
{\bf nkû}  {\it n} 3/4 leg, foot {\it pl.} mìnkû         \\ 
{\bf nkùá}  {\it n} 3/4 trunk (tree) {\it pl.} mìnkùá         \\ 
{\bf nkùlÉ}  {\it n} 3/4 hill, mountain {\it pl.} mìnkùlÉ         \\ 
{\bf nkúlÓ}  {\it n} 3/4 "dead" season (May-Aug) {\it pl.} mìnkúlÓ         \\ 
{\bf nkùmàsà}  {\it n} 3/4 preparation {\it pl.} mìnkùmàsà         \\ 
{\bf nkùmbÓ}  {\it n} 1/2 porcupine {\it pl.} bànkùmbÓ         \\ 
{\bf nkùmbò}  {\it n} 3/4 Nile crocodile (Crocodylus niloticus) {\it pl.} mìnkùmbò         \\ 
{\bf nkùmbÓ wá dúú}  {\it n} 3/4 Nasenflügel {\it pl.} mìnkùmbÓ mí múú         \\ 
{\bf nkúmbÒlÓ}  {\it n} 3/4 diarrhea {\it pl.} mìnkúmbÒlÓ         \\ 
{\bf nkùmù}  {\it n} 3/4 prison {\it pl.} mìnkùmù         \\ 
{\bf nkùndé}  {\it n} 3/4 tail {\it pl.} mìnkùndé         \\ 
{\bf nkúnkúmbÉ}  {\it n} 3/4 bow {\it pl.} mìnkúnkúmbÉ         \\ 
{\bf nkùù}  {\it n} 3/4 evil spirit {\it pl.} mìnkùù         \\ 
{\bf nkùzÓ}  {\it n} 3/4 widow/er {\it pl.} mìnkùzÓ         \\ 
{\bf nkw°á°àlÈ bùdì}  {\it n} 1/2 spy {\it pl.} bàkw°á°àlÈ bá bùdì         \\ 
{\bf nkwálá}  {\it n} 3/4 machete {\it pl.} mìnkwálá         \\ 
{\bf nkwànò}  {\it n} 3/4 honey {\it pl.} mìnkwànò         \\ 
{\bf nkwásá}  {\it n} 3/4 fishing pole {\it pl.} mìnkwásá         \\ 
{\bf nkwE}  {\it n} 3/4 basket {\it pl.} mìnkwE         \\ 
{\bf nkwÉlì}  {\it n} 9/6 witchcraft, sorcellerie {\it pl.} mànkwÉlì         \\ 
{\bf nky°á}  {\it n} 3/4 shrimp {\it pl.} mìnky°á         \\ 
{\bf nky°â}  {\it n} 3/4 scabies {\it pl.} mìnky°â         \\ 
{\bf nkyàálÉ} ná {\it n} 1/2 termite mound {\it pl.} bànánkyàálÉ         \\ 
{\bf nl°á}  {\it n} 3/4 anus {\it pl.} mìnl°á         \\ 
{\bf nl°â}  {\it n} 3/4 story, tale, problem {\it pl.} mìnl°â         \\ 
{\bf nlàà}  {\it n} 3/4 antenna, horn {\it pl.} mìnlàà         \\ 
{\bf nlàwÓ}  {\it n} 3/84 branch {\it pl.} mìnlàwÓ         \\ 
{\bf nlémò}  {\it n} 3/4 heart {\it pl.} mìnlémò         \\ 
{\bf nlô}  {\it n} 3/4 head {\it pl.} mìnlô         \\ 
{\bf nlùdÈ}  {\it n} 3/4 scale (for weighing) {\it pl.} mìnlùdÈ         \\ 
{\bf nlùNgá}  {\it n} 3/4 bucket {\it pl.} mìnlùNgá         \\ 
{\bf nlvúmá}  {\it n} 3/4 fork {\it pl.} mìnlvúmá         \\ 
{\bf nÒ:nÉ}  {\it n} 7/8 bird {\it pl.} bènÒ:nÉ         \\ 
{\bf nÓÓ}  {\it n} 1/2 deaf person {\it pl.} bànÓÓ         \\ 
{\bf nÒÒ}  {\it v} take   nÒngá   nÒngàlà   \\ 
{\bf ns°î}  {\it n} 3/4 linsang de Leighton, wild cat, mangoost {\it pl.} mìns°î         \\ 
{\bf ns°Ô}  {\it n} 3/4 (intestinal) worm {\it pl.} mìns°Ô         \\ 
{\bf ns°Ô}  {\it n} 3/4 beak {\it pl.} mìns°Ô         \\ 
{\bf nsá}  {\it n} 3/4 shore {\it pl.} mìnsá         \\ 
{\bf nsá wá m°â}  {\it n} 3/4 beach, shore (bord de la mer) {\it pl.} mìnsá mí m°â         \\ 
{\bf nsá'à}  {\it n} 3/4 Staude (Banane) {\it pl.} mìnsá'à         \\ 
{\bf nsà'á}  {\it n} 3/4 colobe bai d'Iringa  {\it pl.} mìnsà'á         \\ 
{\bf nsá'àwà}  {\it n} 3/4 Zappeln, repeated movement (e.g. leaves) {\it pl.} mìnsá'àwà         \\ 
{\bf nsàlá}  {\it n} 3/4 crevice, fissure {\it pl.} mìnsàlá         \\ 
{\bf nsálÈ mànk°Ê}  {\it n} 1/2 farmer {\it pl.} bàsálÈ bá mánk°Ê         \\ 
{\bf nsámbò}  {\it n} 3/4 penis {\it pl.} mìnsámbò         \\ 
{\bf nsÉ}  {\it n} 3/4 sand {\it pl.} mìnsÉ         \\ 
{\bf nsÉlÓ}  {\it n} 3/4 plant with thorns {\it pl.} mìnsÉlÓ         \\ 
{\bf nsíngÓ}  {\it n} 3/4 fast {\it pl.} mìnsíngÓ         \\ 
{\bf nsínó}  {\it n} 3/4 color, paint {\it pl.} mìnsínó         \\ 
{\bf nsìsÓ}  {\it n} 3/4 vein {\it pl.} mìnsìsÓ         \\ 
{\bf nsìyÈ}  {\it n} 3/4 string {\it pl.} mìnsìyÈ         \\ 
{\bf nsÔ wá dísì}  {\it n} 3/4 pupil {\it pl.} mìnsÔ mí mísì         \\ 
{\bf nsÓnsÓ}  {\it n} 3/4 bone marrow {\it pl.} mìnsÓnsÓ         \\ 
{\bf nsùlÈ}  {\it n} 3/4 ripeness {\it pl.} mìnsùlÈ         \\ 
{\bf nsùmbÓ}  {\it n} 3/4 hunt (with dogs and spears) {\it pl.} mìnsùmbÓ         \\ 
{\bf nt°á°à}  {\it v} climb over, overcome, suceed   ntàngá  ntàngÈsÈ help traverse ntàngàlà   \\ 
{\bf ntá}  {\it n} 3/4 niece, nephew (children of the sister, i.e. children who do not belong to the house, but have their father elsewhere) {\it pl.} mìntá         \\ 
{\bf ntà}  {\it n} 1/2 grand-child {\it pl.} bàntà         \\ 
{\bf ntámànÈ}  {\it v} ruin, destroy, be ruined         \\ 
{\bf ntàmbÈ}  {\it n} 1/2 rubber {\it pl.} bàntàmbÈ         \\ 
{\bf ntàmbÊ}  {\it n} 1/2 stick {\it pl.} bàntàmbÊ         \\ 
{\bf ntàNgànÈ}  {\it n} 3/4 white person {\it pl.} mìntàNgànÈ         \\ 
{\bf ntányá}  {\it n} 3/4 cleanliness {\it pl.} mìntányá         \\ 
{\bf ntÉ}  {\it n} 3/4 tallness, size {\it pl.} mìntÉ         \\ 
{\bf ntÈgá}  {\it n} 3/4 weakness, softness         \\ 
{\bf ntÉgÈlÈ}  {\it v} threaten, annoy, disturb   ntÉgálâ   ntÉgàlà   \\ 
{\bf ntÉlÉ}  {\it n} 7/8 clothing, fabric {\it pl.} bèntÉlÉ         \\ 
{\bf ntÈmbÓ}  {\it n} 1/2 younger siblings and cousins {\it pl.} bàntÈmbÓ         \\ 
{\bf ntÈmbwà} lè {\it n} 5/6 wrinkle (in skin) {\it pl.} màntÉmbwá         \\ 
{\bf ntÈndá}  {\it n} 3/4 déchirure {\it pl.} mìntÈndá         \\ 
{\bf ntÈndì} lè {\it n} 5/6 saliva, drool {\it pl.} màntÈndì         \\ 
{\bf ntfùgà}  {\it n} 7/8 lid (of bottle) {\it pl.} bèntfùgà         \\ 
{\bf ntfúmò}  {\it n} 3/4 knife {\it pl.} mìntfúmò         \\ 
{\bf ntÒgÒ}  {\it n} 7/8 sweet potato {\it pl.} bèntÒgÒ         \\ 
{\bf ntÒndÒgÈ}  {\it n} 7/8 needle {\it pl.} bèntÒndÒgÈ         \\ 
{\bf ntÒndÒm}  {\it ideo} depiction of monkeys jumping in trees         \\ 
{\bf ntÒNgÈ}  {\it n} 1/2 hornet, wasp, mantispid {\it pl.} bàntÒNgÈ         \\ 
{\bf ntS°ántSùgÈ}  {\it n} 3/4 dragon fly (Odonata) {\it pl.} mìntS°ántSùgÈ         \\ 
{\bf ntúà}  {\it n} 7/6 mango (fruit),  mango tree, wild mango (irvingia gabonensis) {\it pl.} màntúà         \\ 
{\bf ntúdégá} lè {\it n} 5/6 bruise {\it pl.} màntúdégá         \\ 
{\bf ntúlé}  {\it n} 3/4 old person {\it pl.} mìntúlé         \\ 
{\bf ntúmÉ}  {\it n} 3/4 walking stick {\it pl.} mìntúmÉ         \\ 
{\bf ntúmÒ}  {\it n} 0/2 Mvai (Campo, Guinée, Mbam)         \\ 
{\bf ntùngù}  {\it n} 3/4 manner, behavior {\it pl.} mìntùngù         \\ 
{\bf ntùÓ}  {\it inv} six         \\ 
{\bf ntúßí}  {\it n} 3/4 savannah {\it pl.} mìntúßí         \\ 
{\bf númbá}  {\it n} 7/8 place {\it pl.} bènúmbá         \\ 
{\bf nùmbà}  {\it n} 1/2 logger {\it pl.} bànùmbà         \\ 
{\bf nvÈwÒ}  {\it n} 3/4 breath {\it pl.} mìnvÈwÒ         \\ 
{\bf NwándÓ}  {\it n} 3/4 bitter manioc {\it pl.} mìNwándÓ         \\ 
{\bf NwándÓ}  {\it n} 9/6 cassava stick {\it pl.} mànwándÓ         \\ 
{\bf ny°â}  {\it n} 1/2 mother {\it pl.} bàny°â         \\ 
{\bf nyá}  {\it inv} really         \\ 
{\bf nyâ}  {\it n} 7/8 nail (finger, toe), claw {\it pl.} bènyâ         \\ 
{\bf nyâ} mà {\it n} 0/6 milk         \\ 
{\bf nyâ}  {\it v} lick, suckle (babies)   nyángâ  nyángÈsÈ nyángàlà   \\ 
{\bf nyàà}  {\it v} defecate   nyàgâ been shitted  nyàgÈsÈ nyàgàlà   \\ 
{\bf nyáàlÈ}  {\it n} 1/2 beggar {\it pl.} bànyáàlÈ         \\ 
{\bf nyádÈ}  {\it n} 1/2 buffalo {\it pl.} bànyádÈ         \\ 
{\bf nyàgà}  {\it n} 7/8 cow {\it pl.} bènyágà         \\ 
{\bf nyàlÉ}  {\it n} 1/2 son/brother -in-law {\it pl.} bànyàlÉ         \\ 
{\bf nyàlÈ}  {\it v} scratch   nyàlá   nyàlàlà   \\ 
{\bf nyàmá}  {\it n} 3/4 kaputt (things) {\it pl.} mìnyàmá         \\ 
{\bf nyámbá}  {\it n} 9/6 armpit {\it pl.} mànyámbá         \\ 
{\bf nyàmÈ}  {\it n} 7/8 poverty {\it pl.} bènyàmÈ         \\ 
{\bf nyàmÒ}  {\it v} se gâter (house, fruit)   nyàmá  nyàmÈsÈ faire gâter nyàmàlà   \\ 
{\bf nyánÈ}  {\it n} 7/8 war {\it pl.} bènyánÈ         \\ 
{\bf nyánÒ} mà {\it n} 0/6 pain         \\ 
{\bf nyànò}  {\it v} hurt (faire mal)         \\ 
{\bf nyànò}  {\it v} hurt (faire mal)         \\ 
{\bf nyè}  {\it v} return   nyìgá   nyìgàlà   \\ 
{\bf nyÊ}  {\it v} see, look      nyÉnàlà   \\ 
{\bf nyÈmbÉ}  {\it n} 7/8 gun {\it pl.} bènyÈmbÉ         \\ 
{\bf nyÈsÈlÈ}  {\it v} auf etwas runterdrücken, deepen   nyÈsá abgesenkt      \\ 
{\bf nyî}   {\it v} enter   nyíngâ nyíngÈlÈ faire entrer, insert  nyíngàlà   \\ 
{\bf nyígÈ}  {\it v} beg         \\ 
{\bf nyìkà (yá mbÔ)}  {\it n} 7/8 Arm(-Beuge) {\it pl.} bènyìkà bé mábÔ         \\ 
{\bf nyímÈ}  {\it v} refuse   nyímâ  nyímÈsÈ nyímàlà   \\ 
{\bf nyímÈlÈ}  {\it v} tighten   nyímálâ   nyímàlà   \\ 
{\bf nyÒmbÈlÈ}  {\it v} tickle      nyÒmbàlà   \\ 
{\bf nyÓny°Ô}  {\it n} 7/8 yawn {\it pl.} bènyÓny°Ô         \\ 
{\bf nyÓÒ} mà {\it n} 0/6 wine, general term for alcohol         \\ 
{\bf nyÓÒ má lÉndÉ} mà {\it  n} 5/6 palm wine         \\ 
{\bf nyú (wá nkwànò)}  {\it n} 1/2 bee {\it pl.} bànyú (bá nkwànò)         \\ 
{\bf nyú:lé}  {\it n} 7/8 insect {\it pl.} bènyú:lé         \\ 
{\bf nyú°à}  {\it n} 1/2 snake {\it pl.} bànyú°à         \\ 
{\bf nyúlÉ}  {\it n} 3/4 orphan {\it pl.} mìnyúlÉ         \\ 
{\bf nyúlÈ}  {\it n} 9/6 body {\it pl.} mànyúlÈ         \\ 
{\bf nyùlÈ}  {\it n} 3/4 flame {\it pl.} mìnyùlÈ         \\ 
{\bf nyùlÈ}  {\it v} drink   nyùlá  nyùlÈsÈ nyùlàlà   \\ 
{\bf nyùmbò}  {\it n} 3/4 mouth {\it pl.} mìnyùmbò         \\ 
{\bf nyùmbÒ}  {\it v} smell intr (good or bad)   nyùmbá nyùmbÈlÈ smell sth nyùmbÈsÈ nyùmbàlà   \\ 
{\bf nyùngù}  {\it n} 1/2 rainbow {\it pl.} bànyùngù         \\ 
{\bf nyùùlÈ}  {\it n} 1/2 mosquito {\it pl.} bànyùùlÈ         \\ 
{\bf nyw°á°í}  {\it adv} early (in the day, before sunset)         \\ 
{\bf nz°á°à}  {\it n} 7/0 appetite for meat or fish         \\ 
{\bf nzá}  {\it pro} who         \\ 
{\bf nzá} lè {\it n} 5/6 dead leaves in water {\it pl.} mànzá         \\ 
{\bf nzálè} mà {\it n} 0/6 urine {\it pl.} mànzálè         \\ 
{\bf nzàmbí}  {\it n} 1/2 god, good spirit {\it pl.} bànzàmbí         \\ 
{\bf nzàmbÒ}  {\it n} 7/8 marsh {\it pl.} bènzàmbÒ         \\ 
{\bf nzámù}  {\it n} 1/2 appetite {\it pl.} bànzámù         \\ 
{\bf nzÈlÈ}  {\it n} 7/8 beard {\it pl.} bènzÈlÈ         \\ 
{\bf nzí nzálÈ}  {\it n} 7/8 bladder (place of urine) {\it pl.} bènzí nzálÈ         \\ 
{\bf nzìl°û}  {\it n} 7/8 swallow {\it pl.} bènzìl°û         \\ 
{\bf nzímÒ} lè {\it n} 5/6 termite (Isoptera) {\it pl.} mànzímÒ         \\ 
{\bf nzÒmÉ}  {\it n} 7/8 splinter {\it pl.} bènzÒmÉ         \\ 
{\bf Ó(nÉ)gá} * {\it mod} (an)other         \\ 
{\bf p°â}  {\it v} first, begin?         \\ 
{\bf p°â}  {\it v} reign, govern, command   mpángâ   pángàlà recip   \\ 
{\bf p°é}  {\it n} 9/6 injury {\it pl.} màp°é         \\ 
{\bf p°Ô}  {\it n} 0/2 Fang (Kribi)         \\ 
{\bf pà} lè {\it n} 5/6 paw {\it pl.} màpà         \\ 
{\bf pá'á}  {\it n} 7/8 1) bark (tree) 2) coin {\it pl.} bèpá'á         \\ 
{\bf pá'à}  {\it v} dig, hollow out (e.g. drum)   mpágâ   págàlà   \\ 
{\bf pà'à}  {\it v} grow (plants)   mpàgá   pàgàlà   \\ 
{\bf páàlà}  {\it n} 9/6 valley {\it pl.} màpáàlà         \\ 
{\bf pádÒ}  {\it v} zupfen, pflücken (prunes, kala), wring out   mpádâ   pádàlà   \\ 
{\bf pálàbà}  {\it v} blink (eye)         \\ 
{\bf pálÒ}  {\it v} sort   mpálâ   pálàlà   \\ 
{\bf pámó} mà {\it n} 0/6 rise, arrival         \\ 
{\bf pámò}  {\it v} appear   mpámâ   pámàlà   \\ 
{\bf pàmpÉlÈ}  {\it n} 7/8 grapefruit {\it pl.} bèpàmpÉlÈ         \\ 
{\bf pándÈ}  {\it v} arrive   mpándâ   pándàlà   \\ 
{\bf pándyì}  {\it n} 1/2 plate {\it pl.} bàpándyì         \\ 
{\bf pándyì wà dô}  {\it  n} 1/2 deep plate         \\ 
{\bf pánÈ}  {\it v} hang up, accrocher   mpánâ  pánÈsÈ pánàlà   \\ 
{\bf pànò}  {\it v} shine (sun, fireflies, stars, moon, light, lamp)   mpàná c'est brillant      \\ 
{\bf pÉ}  {\it prep} to, for         \\ 
{\bf pÊ}  {\it v} choose   mpÉyâ   pÉyàlà   \\ 
{\bf pè'è}   {\it n} 9/6 wisdom {\it pl.} màpè'è         \\ 
{\bf pébà} lè {\it n} 5/6 fin {\it pl.} mápébà         \\ 
{\bf péè}  {\it n} 7/8 avocado (tree and fruit) {\it pl.} bèpéè         \\ 
{\bf pèè}  {\it n} 9/6 conscience {\it pl.} màpèè         \\ 
{\bf pÉÉpÉÈ}  {\it n} 1/2 cockroach {\it pl.} bàpÉÉpÉÈ         \\ 
{\bf pÉlÈ}  {\it n} 7/8 side {\it pl.} bèpÉlÈ         \\ 
{\bf pÉlÉ (bé bénó)} bè {\it n} 0/8 two halfs (things that have been cut in two), Pobacken         \\ 
{\bf pÉmbÓ}  {\it n} 7/8 clay, bread {\it pl.} bèpÈmbÓ         \\ 
{\bf pÈndÈlÈ}  {\it v} auslecken mit Finger   mpÈndálâ ausgeleckt Teller   pÈndàlà   \\ 
{\bf péndÒ}  {\it v} braid   mpéndâ   péndàlà   \\ 
{\bf pÉpÉ}  {\it n} 1/2 leaf-hopper bug (Cicadellidae) {\it pl.} bàpÉpÉ         \\ 
{\bf pÉßá} lè {\it n} 5/6 wing {\it pl.} màpÉßá         \\ 
{\bf pÉwÓ}  {\it n} 7/8 scar {\it pl.} bèpÉwÓ         \\ 
{\bf péyà}  {\it v} booze, drunk     péyÈsÈ make drunk péyàlà   \\ 
{\bf pf'ö} bà {\it n} 0/2 Bapoko (rather Kwasio)         \\ 
{\bf pf°û}  {\it n} 7/8 colobe bai de l'Ouest, de camail {\it pl.} bèpf°ù         \\ 
{\bf pfáááá}  {\it ideo} depiction of flinging a long object or slinging         \\ 
{\bf pfû(phû)} ná {\it qual} darkened color         \\ 
{\bf pfùdé}  {\it n} 9/6 mold {\it pl.} màpfùdé         \\ 
{\bf pfùdÓ}  {\it n} 7/8 abandonment {\it pl.} bèkfùdÓ         \\ 
{\bf pfúÈlÈ}  {\it v} crunch   mpfúálâ   pfúàlà   \\ 
{\bf pfùmbÈ}  {\it v} pull out (groundnuts)   mpfùmbá   pfùmbàlà   \\ 
{\bf pfúndÒ}  {\it v} be frightened, avoir peur     pfúndÈsÈ pfúndàlà   \\ 
{\bf pfùNgà}  {\it n} 7/8 lid (pot, eye) {\it pl.} bèpfùNgà         \\ 
{\bf pfúßánÉ}  {\it n} 3/4 cleanliness {\it pl.} mìpfúßánÉ         \\ 
{\bf pfùßÈlÈ}  {\it v} blow (tr), blow down   mpfúßálâ   pfùßàlà   \\ 
{\bf pfùtùm}  {\it ideo} depiction of sound when jumping into water         \\ 
{\bf pfùwÒ}  {\it v} dust   mpfùwâ   pfùwàlà   \\ 
{\bf pfùyá} bè {\it n} 0/8 ashes, powder         \\ 
{\bf pî:pî:}  {\it n} 1/2 butterfly, moth {\it pl.} bàpî:pî:         \\ 
{\bf pílá ngàndÉ } bè {\it n} 0/8 übereinanderstehende Zähne (ngande=Krokodil)         \\ 
{\bf pílì}  {\it n} 7/6 moment, season {\it pl.} màpílì         \\ 
{\bf pílÒ}  {\it adv} when         \\ 
{\bf pìmáá}  {\it n} 7/8 wall {\it pl.} bìpìmáá         \\ 
{\bf pímbÈ}  {\it v} wipe   mpímbâ   pímbàlà   \\ 
{\bf pímù}  {\it n} 9/6 force, power {\it pl.} màpímù         \\ 
{\bf píndyÓ}  {\it n} 7/8 piece, part (broken off) {\it pl.} bèpíndyÓ         \\ 
{\bf pínÈsÈ}  {\it v} squeeze   mpínâ   pínàlà   \\ 
{\bf písÈ}  {\it adv, post} last, late         \\ 
{\bf písÈ}  {\it n} 7/8 back (spatial) {\it pl.} bèpísÈ         \\ 
{\bf píyÒ}  {\it qual} small, thin         \\ 
{\bf pìyù pìyù}  {\it n} 1/2 small rain, small rainy season (Mar - May) {\it pl.} bàpìyù pìyù         \\ 
{\bf pÓ}  {\it n} 9/6 news, prophecy {\it pl.} màpÓ         \\ 
{\bf pÓdÈ}  {\it n} 1/2 port, harbour {\it pl.} bàpÓdÈ         \\ 
{\bf pÓm}  {\it n} 1/2 potato {\it pl.} bàpÓm         \\ 
{\bf pÓndÈsÈ}  {\it v} punish   mpÓndásâ      \\ 
{\bf pÓnÉ}  {\it n} 7/0 truth         \\ 
{\bf pÒpÓ}  {\it n} 7/8 papaya {\it pl.} bèpÒpÓ         \\ 
{\bf pÒtÒ}  {\it n} 7/8 clay (for building houses) {\it pl.} bèpÒtÒ         \\ 
{\bf pówàlà}  {\it qual} tranquille, calm         \\ 
{\bf pú°Ò}  {\it v} pay   mpúngâ   púngàlà   \\ 
{\bf pùdùm}  {\it ideo} depiction of falling into mud or throwing stone into water         \\ 
{\bf púndí}  {\it n} 1/2 cercophitèque ascague, colobe vert? {\it pl.} bàpúndí         \\ 
{\bf púndì}  {\it v} polish   mpúndâ   púndàlà   \\ 
{\bf pùsÈ}  {\it v} push   mpùsá   pùsàlà   \\ 
{\bf púsí}  {\it n} 7/8 bottle {\it pl.} bèpúsí         \\ 
{\bf púù}  {\it n} 7/0 1) reason 2) púù ATT/GEN for, because         \\ 
{\bf pùúlì}  {\it n} 7/8 hat {\it pl.} bèpùúlì         \\ 
{\bf pwápwâ}  {\it n} 1/2 truth, honesty {\it pl.} bàpwàpwâ         \\ 
{\bf pwàsÒwÒ}  {\it v} stretch (animal with sticks for smoke), stretch oneself intr.   mpwàsá   pwàsàlà   \\ 
{\bf pwÈdà}  {\it n} 1/2 grass {\it pl.} bàpwÈdà         \\ 
{\bf pyàgá}  {\it n} 7/6 paper {\it pl.} màpyàgá         \\ 
{\bf s°á}  {\it n} 1/2 father, male {\it pl.} bàs°á         \\ 
{\bf s°â} mà {\it n} 0/6 game (playing)         \\ 
{\bf s°â}  {\it v} vomit   nsángâ  sángÈsÈ sángàlà   \\ 
{\bf s°á wà kfúbÒ}  {\it n} 1/2 rooster (male of chicken) {\it pl.} bàs°á bá kfúbÒ         \\ 
{\bf s°á°àsà}  {\it v} mix   ns°á°ásâ      \\ 
{\bf s°È} lè {\it n} 5/6 small canoe (Einbaum) {\it pl.} màs°È         \\ 
{\bf s°Ê} lè {\it n} 5/6 umbrella tree (musanga cecropioides) {\it pl.} màs°Ê         \\ 
{\bf s°í°ì}  {\it v} approach (tr.)   nsíngâ sísÈlÈ  síngàlà   \\ 
{\bf s°Ô}  {\it n} 9/6 grave, tomb {\it pl.} màs°Ô         \\ 
{\bf s°Òkìndá}  {\it n} 1/2 biting ants {\it pl.} bàs°Òkìndá         \\ 
{\bf sá}   {\it n} 1/2 earth worm {\it pl.} bàsá         \\ 
{\bf sá} lè {\it n} 5/6 prune {\it pl.} màsá         \\ 
{\bf sà}  {\it n} 7/8 hut {\it pl.} bèsà         \\ 
{\bf sâ}  {\it n} 7/8 thing {\it pl.} bèsâ         \\ 
{\bf sâ} lè {\it n} 5/6 feather {\it pl.} màsâ         \\ 
{\bf sâ}  {\it v} do   nsáyâ   sáàlà   \\ 
{\bf sá'àwà}  {\it v} move repeatedly         \\ 
{\bf sáálÉ}  {\it n} 7/8 work {\it pl.} bÌsáálÉ         \\ 
{\bf sàgà}  {\it v} shock, scare, effrayer, be surprised   nsàgá   sàgàlà   \\ 
{\bf ságÓságÓ}  {\it n} 1/2 comb {\it pl.} bàságÓságÓ         \\ 
{\bf sálá (má kúlí)} mà {\it n} 0/6 ceremony months after a funeral ending the deuil         \\ 
{\bf sàlàgà}  {\it n} 7/8 ditch {\it pl.} bèsàlàgà         \\ 
{\bf sàlÈ}  {\it v} crack (wood, wall)         \\ 
{\bf sálÈ bàmbèyè} n {\it n} 1/3 prostitute {\it pl.} bàsálÈ bá bébàmbèyè         \\ 
{\bf sálÈ màngámbÉ} n {\it n} 1/2 diviner, fortune-teller {\it pl.} bàsálè bá mángámbÉ         \\ 
{\bf sálÈ ngyàngÓ} n {\it n} 1/2 hunter {\it pl.} bàsálÈ bá béngyàngÓ         \\ 
{\bf sálÒ}  {\it v} become lots   nsálâ      \\ 
{\bf sàlÒ}  {\it v} cut längs   nsàlá sàlÈ crack intr (wood, wall)  sàlàlà sich gegenseitig schneiden   \\ 
{\bf sàmbÈsÈ}  {\it n} 7/0 rape         \\ 
{\bf sàndyá}  {\it n} 7/8 raffia mat for house building {\it pl.} bèsàndyá         \\ 
{\bf sàndyà}  {\it n} 1/2 fabric (pagne) {\it pl.} bàsàndyà         \\ 
{\bf sánÈ}  {\it v} decide   nsánâ   sánàlà   \\ 
{\bf sàsàmbÉ (yá mwánÒ)}  {\it n} 7/8 miscarriage {\it pl.} bèsàsàmbÉ         \\ 
{\bf sÉ'È}  {\it n} 7/8 liver {\it pl.} bèsÉ'È         \\ 
{\bf sÉ'È}  {\it n} 7/8 mandrill (Mandrillus sphinx) {\it pl.} bèsÉ'È         \\ 
{\bf sÈgÈsÈ}  {\it n} 7/8 sieve {\it pl.} bèsÈgÈsÈ         \\ 
{\bf sÈgÈsÈ}  {\it v} sieve   nsÉgásâ      \\ 
{\bf sÉkÈ}  {\it n} 1/2 termite {\it pl.} bàsÉkÈ         \\ 
{\bf sÉlÒ}  {\it v} peel (schälen)   nsÉlâ   sÉlàlà   \\ 
{\bf sÉmbÒ}  {\it v} arrive, accoster         \\ 
{\bf sÈndÈ}  {\it v} slip   msÈndá  sèndÈsÈ sÈndàlà   \\ 
{\bf sÈngÈ}  {\it v} lower   nsÈngá   sÈngàlà   \\ 
{\bf sí}  {\it n} 9/6 ground, soil, world {\it pl.} màsí         \\ 
{\bf sí}  {\it post} under         \\ 
{\bf síàwà}  {\it v} have a hiccup         \\ 
{\bf sìgá}  {\it n} 1/2 cigarette {\it pl.} bàsìgá         \\ 
{\bf sìlá} lè {\it n} 5/6 mole-cricket (Gryllotalpa africana), tiger beetle (Megacephala) {\it pl.} màsìlá         \\ 
{\bf sìlá} lè {\it n} 5/6 grillon, cicade {\it pl.} màsìlá         \\ 
{\bf sílÈ}  {\it v} finish, end, use up, kill   nsílâ  sílÈsÈ wÊ sálÈ sílàlà   \\ 
{\bf sìlÈgà}  {\it v} descend, fade   nsìlá  sìlÈsÈ sìlàlà gegenseitig abseilen   \\ 
{\bf sìlí}  {\it n} 7/8 1) hair 2) spark (bé béyí) {\it pl.} bèsìlí         \\ 
{\bf sílífàzì}  {\it n} 1/2 sandal {\it pl.} bàsílífàzì         \\ 
{\bf sílÒ}  {\it v} rub, smear, paint   nsílâ   sílàlà   \\ 
{\bf símàsà}  {\it v} regret   nsímásâ      \\ 
{\bf sìmbÒ}  {\it v} drag   nsìmbá   sìmbàlà   \\ 
{\bf símÈ}  {\it v} respect   nsímâ   símàlà   \\ 
{\bf sìmú}  {\it n} 7/8 liquid sauce {\it pl.} bèsìmú         \\ 
{\bf síndyà}  {\it v} change, exchange   nsíndyâ   síndyàlà   \\ 
{\bf síNgí}  {\it n} 7/8 squirrel {\it pl.} bèsíNgí         \\ 
{\bf síNgì}  {\it n} 7/8 cat {\it pl.} bèsíNgì         \\ 
{\bf sìNgì}  {\it n} 7/8 soul, spirit {\it pl.} bèsìNgì         \\ 
{\bf sísà}  {\it n} 3/4 Aidan fruit and tree (tetrapleura tetraptera) {\it pl.} mìnsísà         \\ 
{\bf sísÈlÈ}  {\it v} scare sb   nsísâ I was scared   sísàlà sísÈgà I'm scared  \\ 
{\bf sìsímù}  {\it n} 7/8 shadow (of  person) {\it pl.} bèsìsímù         \\ 
{\bf sísÒ}  {\it v} approach (intr)   nsísâ   sísàlà   \\ 
{\bf sìsÒ}  {\it v} be happy      sìsàlà   \\ 
{\bf sìsùù}  {\it n} 7/8 apparition {\it pl.} bèsìsùù         \\ 
{\bf síyá} bè {\it n} 0/8 imitation         \\ 
{\bf sìyà}  {\it v} wash, bathe   nsìyá   sìyàlà   \\ 
{\bf síyÈ}  {\it n} 7/8 fire (Kwasio loan word) {\it pl.} bèsíyÈ         \\ 
{\bf síyÈ}  {\it v} saw   nsíyâ   síyàlà   \\ 
{\bf síyÈsÈ}  {\it v} swing, shake   nsíyàsâ      \\ 
{\bf síyÒ}  {\it n} 7/8 dry season (Nov-Mar) {\it pl.} bèsíyÒ         \\ 
{\bf síyÒ} lè {\it n} 5/6 tusk {\it pl.} màsíyÒ         \\ 
{\bf sÓ}  {\it n} 1/2 friend {\it pl.} bàsÓ         \\ 
{\bf sÒ}  {\it n} 7/8 saw {\it pl.} bèsÒ         \\ 
{\bf sÓ'Ò}  {\it v} continue   nsÓsàlà sÓsÈlÈ nlâ continue story     \\ 
{\bf sÒ'Ó}  {\it n} 7/8 cynocéphale {\it pl.} bèsÒ'Ó         \\ 
{\bf sòbàlà}  {\it v} accumulate, coagulate   nsòbálá      \\ 
{\bf sÒgá}  {\it n} 7/8 secret {\it pl.} bèsÒgá         \\ 
{\bf sÓlÈ}  {\it v} undress, take off (clothes)   nsÓlâ  sòlÈsÈ sÓlàlà   \\ 
{\bf sÒlÈ}  {\it v} hide sth   nsÒlá   sÒlàlà   \\ 
{\bf sÓlÉ yá gÓlÈ}  {\it n} 7/8 hummingbird (turquoise and red) {\it pl.} bèsÓlÉ bé gÓlÈ         \\ 
{\bf sÓlÈgà}  {\it v} dégringoler, fall   nsÓlÉgâ      \\ 
{\bf sÒmÒnÈ}  {\it n} 7/0 complaint         \\ 
{\bf sÓndÒ}  {\it n} 1/2 week {\it pl.} bàsÓndÒ         \\ 
{\bf sÓndyà}  {\it v} bring to point, sharpen, spitzen   nsÓndyà   sÓndyàlá   \\ 
{\bf sÓnì}  {\it n} 7/0 shame         \\ 
{\bf sÓÒ}  {\it post} before, in front         \\ 
{\bf sÓÒ}  {\it n} 7/0 front (spatial)         \\ 
{\bf sÓsÈlÈ}  {\it v} smoke (fish or animal)   nsÓsálâ      \\ 
{\bf sÒsí} mà {\it n} 0/6 joy         \\ 
{\bf sÓßá}  {\it n} 7/8 mud {\it pl.} bèsÓßá         \\ 
{\bf sÓßÌ}  {\it n} 7/8 soap {\it pl.} bÌsÓßÌ         \\ 
{\bf sÓtì}  {\it n} 1/2 trousers {\it pl.} bàsÓtì         \\ 
{\bf sù} lè {\it n} 5/6 jigger {\it pl.} màsù         \\ 
{\bf sù'ù}  {\it n} 7/8 putty-nosed monkey (Cercopithecus nictitans) {\it pl.} bèsù'ù         \\ 
{\bf sù'ù} lè {\it n} 5/6 waterfall {\it pl.} màsù'ù         \\ 
{\bf sùbÈ}  {\it v} pour out (ausschütten), turn over   nsùbá sùbÈlÈ ejaculate sùbèsÈ turn sth over sùbàlà   \\ 
{\bf súbì}  {\it n} 7/8 sauce, soup {\it pl.} bèsúbì         \\ 
{\bf sùmbÒ}  {\it v} die in a mystical way   nsùmbá   sùmbàlà   \\ 
{\bf súmÈlÈ}  {\it v} greet   nsúmálâ   súmàlà   \\ 
{\bf sùnÉ} n {\it n} 3/4 flesh {\it pl.} mìsùnÈ         \\ 
{\bf súNgú}  {\it n} 7/8 drinking cup made of leaves {\it pl.} bèsúNgú         \\ 
{\bf súngú}  {\it n} 7/8 container made of leaf (for water or medicine) {\it pl.} bèsúngú         \\ 
{\bf súngú}  {\it n} 7/8 container made of leaf (for water or medicine) {\it pl.} bèsúngú         \\ 
{\bf sùngù}  {\it n} 7/8 war  {\it pl.} bèsùngù         \\ 
{\bf súnÓ} lè {\it n} 5/6 doubt {\it pl.} màsúnÓ         \\ 
{\bf súwálá}  {\it n} 7/8 meeting, conference {\it pl.} bèsúwálá         \\ 
{\bf sùwÒ}  {\it v} spill    sùwElE pour sth     \\ 
{\bf swáálÈ}  {\it n} 1/2 bone marrow {\it pl.} bàswááe         \\ 
{\bf swàmbò} lè {\it n} 5/6 going out (for hunting) {\it pl.} màswàmbò         \\ 
{\bf swásÒ}  {\it v} dry (intr.)   nswásâ swásÈlÈ dry sth  swásàlà   \\ 
{\bf swàwÒ}  {\it v} hide (intr)   nswàwá      \\ 
{\bf swî} lè {\it n} 5/6 death {\it pl.} màswî         \\ 
{\bf ʃyÊ ʃyÊ}  {\it ideo} depiction of sneaking         \\ 
{\bf t°â}  {\it n} 9/6 number, price {\it pl.} màt°â         \\ 
{\bf t°â}  {\it n} 7/8 Gestell, grillage for smoking meat {\it pl.} bèt°â         \\ 
{\bf t°á°à}  {\it v} tell (only used for stories, anecdotes fairy tales)         \\ 
{\bf t°áàlà nà}  {\it v} judge         \\ 
{\bf t°Ê}  {\it v} limp      tÉngàlà   \\ 
{\bf t°È°È}  {\it v} abandon   ntÈngá   tÈngàlà   \\ 
{\bf t°î} ná {\it qual   } straight         \\ 
{\bf t°ì°É} lè {\it n} 5/6 knod {\it pl.} màt°ì°É         \\ 
{\bf t°ÓÒlÈ}  {\it v} guide, direct         \\ 
{\bf t°ú°ù}  {\it n} 7/8 axe {\it pl.} bèt°ú°ù         \\ 
{\bf tá} lè {\it n} 5/6 stain {\it pl.} màtá         \\ 
{\bf tá}  {\it n} 1/2 father {\it pl.} bátá         \\ 
{\bf tá'àlÈ}  {\it v} start, begin    ntáálâ      \\ 
{\bf tàbá}  {\it n} 7/8 necklace {\it pl.} bètàbá         \\ 
{\bf tálá} mà {\it n} 0/6 beginning, start         \\ 
{\bf támbí} lè {\it n} 5/6 oyster {\it pl.} màtámbí         \\ 
{\bf tàmbó} lè {\it n} 5/6 bee wax {\it pl.} màtàmbó         \\ 
{\bf tánà} lè {\it n} 5/6 hail {\it pl.} màtánà         \\ 
{\bf tándÓ yá mwánÒ}  {\it n} 7/8 womb (cage, net of child) {\it pl.} bètándÓ bé bwánÒ         \\ 
{\bf tánÈ} * {\it mod} five         \\ 
{\bf tángà} bà {\it n} 0/2 Batanga (Banua and Bapoko)         \\ 
{\bf tàNgò} mà {\it n} 0/6 palm wine (areal term)         \\ 
{\bf tàtànÓs}  {\it n} 1/2 mantis {\it pl.} bàtàtànÓs         \\ 
{\bf tátÒ}  {\it v} take care of, guard   ntátâ   tátàlà   \\ 
{\bf tàtÒ}  {\it v} squeak, scream   ntàdá  tàdÈsÈ tàtàlà   \\ 
{\bf tàwÒ}  {\it n} 7/8 goat, sheep {\it pl.} bètàwÒ         \\ 
{\bf té}  {\it n} 7/8 posture, position {\it pl.} bèté         \\ 
{\bf tÊ}  {\it adv} now         \\ 
{\bf tÊ}  {\it v} create, invent, found   ntÉyâ   tÉyàlà   \\ 
{\bf tÉ'É} lè {\it n} 5/6 fatigue {\it pl.} màtÉ'É         \\ 
{\bf tÉ'È}  {\it v} be soft, be weak   ntÉgâ     tÉgÈ make soft \\ 
{\bf tè'ètè}  {\it n} 7/8 tenderness {\it pl.} bètè'ètè (plr)         \\ 
{\bf tÈbÉ}  {\it n} 7/8 beach, shore {\it pl.} bètÈbÉ         \\ 
{\bf tÉbÒ}  {\it v} get up, rise, stop, stand   ntÉlâ tÉlÈ place sth upright  ntÉlàlà sich gegenseitig hinstellen   \\ 
{\bf tèèèè}  {\it ideo} depiction of waiting         \\ 
{\bf tÉgÈ}  {\it v} make tired   ntÉgâ  tÉgÈsÈ tÉgàlà   \\ 
{\bf tÉlÈ} mà {\it n} 0/6 saliva (spit)         \\ 
{\bf tÈmbÒwÒ}  {\it v} set, go down (only for sun)   ntÈmbá  ntÈmbÈsÈ    \\ 
{\bf tÈmbÓwÓ (má vísÓ)} mà {\it n} 0/6 sunset         \\ 
{\bf tÈndáà} lè {\it n} 5/6 grillon, cicade (lives in holes in ground) {\it pl.} màtÈndáà         \\ 
{\bf tÈndÒ}  {\it v} tear (déchirer)   ntÈndá  tÈndÈsÈ  tÈndàlà   \\ 
{\bf tÈtÈkÈ}  {\it n} 7/8 frogs that fall from sky with rain {\it pl.} bètÈtÈkÈ         \\ 
{\bf tfúàdà}  {\it v} be late, tarder         \\ 
{\bf tfùbó}  {\it n} 7/8 black mamba {\it pl.} bètfùbó         \\ 
{\bf tfùbò}  {\it v} 1) pierce 2) rape   ntfúbâ   tfúbàlà   \\ 
{\bf tfùdáà}  {\it n} 7/8 pinch {\it pl.} bètfùdáà         \\ 
{\bf tfúdÉ}  {\it n} 7/8 bump {\it pl.} bèkfúdÉ         \\ 
{\bf tfùdÒ}  {\it v} pinch   ntfùdá   tfùdàlà   \\ 
{\bf tfúgà}  {\it n} 7/8 suffering {\it pl.} bètfúgà         \\ 
{\bf tfúgà}  {\it v} suffer   ntfúgâ  tfúgÈsÈ tfúgàlà   \\ 
{\bf tfùlÈ} mà {\it n} 0/6 smell         \\ 
{\bf tfúmbÒ}  {\it v} fold, wrinkle   ntfúmbâ  tfúmbÈsÈ tfúmbàlà tfúmbàgà fold by itself  \\ 
{\bf tfùnÈ}   {\it n} 7/8 strap (made of bark or veins), scarf for carrying babies {\it pl.} bètfùnÈ         \\ 
{\bf tíì}  {\it v} start walking, displace oneself   ntíyâ   tíyàlà   \\ 
{\bf tìnÒ}  {\it v} arracher, harvest (tubers) (tr and intr)   ntìná tílÈ  tìnàlà   \\ 
{\bf tísònì}  {\it n} 7/8 town {\it pl.} bètísònì         \\ 
{\bf títímÓ}  {\it n} 7/8 middle {\it pl.} bètítímÓ         \\ 
{\bf tÓ} lè {\it n} 5/6 drop {\it pl.} màtÓ         \\ 
{\bf tÒ}  {\it inv} any         \\ 
{\bf tÒà }  {\it v} boil (intr)   ntÒgá   tÒgàlà  tÒkÈ boil sth \\ 
{\bf tÒdÈ}  {\it n} 7/8 roundness {\it pl.} bètÒdÈ         \\ 
{\bf tÓdyínì}  {\it n} 1/2 thousand {\it pl.} bàtÓdyínì         \\ 
{\bf tÓkÈ}  {\it v} take, pick up   ntÓkâ  tÓkÈsÈ make pick up tÓkàlà   \\ 
{\bf tÓlÈ} lè {\it n} 5/6 navel {\it pl.} màtÓlÈ         \\ 
{\bf tÔmbÓ}  {\it n} 7/8 problem {\it pl.} bètÔmbÓ         \\ 
{\bf tóndí} lè {\it n} 5/6 friend/lover {\it pl.} màtóndí         \\ 
{\bf tÒndÒ}  {\it n} 1/2 nail {\it pl.} bàtÒndÒ         \\ 
{\bf tÒntsá}  {\it n} 7/8 mistletoe plant (Agelanthus djurensis) {\it pl.} bètÒntsá         \\ 
{\bf tÒsâ}  {\it adv} no, never, nothing         \\ 
{\bf tÓwá}  {\it ?} all (used with time only, whole time/night/day/hour)         \\ 
{\bf tÓwà}  {\it v} drip, leak   ntÓwâ      \\ 
{\bf trésì}  {\it n} 1/2 thread {\it pl.} bàtrésì         \\ 
{\bf tS°í}  {\it n} 9/6 1) neck 2) voice {\it pl.} màtS°í         \\ 
{\bf ts°î}  {\it v} untie, unwrap, loosen   ntsíngâ   tsíngàlà   \\ 
{\bf ts°íÈlÈ}  {\it v} make a knod, bind, tie   nts°íyálâ   ts°íyàlà   \\ 
{\bf ts°ìy°É} lè {\it n} 5/6 knot {\it pl.} màts°ìy°É         \\ 
{\bf tsàmÈ}  {\it v} spit   ntsàmá   tsàmàlà   \\ 
{\bf tsì} n {\it n} 1/2 in-law {\it pl.} bàtsì         \\ 
{\bf tsì}  {\it n} 7/8 interdiction {\it pl.} bètsì         \\ 
{\bf tsí wà mùr°á} n {\it n} 1/2 mother/sister -in-law {\it pl.} bàtsí bá búrá         \\ 
{\bf tsíbÒ}  {\it v} grind, trample (in mortar)   ntsíbâ   tsíbàlà   \\ 
{\bf tsìdÈdÈ}  {\it n} 1/2 honesty {\it pl.} bàtsìdÈdÈ         \\ 
{\bf tsídí}  {\it n} 1/2 animal, meat {\it pl.} bàtsídí         \\ 
{\bf tSíÈ}  {\it n} 9/6 blood {\it pl.} màtSíÈ         \\ 
{\bf tsíÈ}  {\it v} cut   ntsíyâ   tsíyàlà   \\ 
{\bf tsìÈ}  {\it v} live, be well   ntsìgá je suis vivant      \\ 
{\bf tsíÈ bènyàgà} n {\it n} 1/2 butcher (cow slaughterer) {\it pl.} bàtsíÈ bá bényàgà         \\ 
{\bf tsíÈsámÈ}  {\it n} 1/2 circumcision {\it pl.} bàtsíÈsámÈ         \\ 
{\bf tsígÈ}  {\it v} take off, start going (only with plural subject)         \\ 
{\bf tsíì}  {\it n} 7/8 life {\it pl.} bètsî         \\ 
{\bf tsílì}  {\it n} 7/8 smallness, part, shortness, half {\it pl.} bètsílì         \\ 
{\bf tsílí yá kàbà}  {\it n} 7/8 long skirt {\it pl.} bètSílí bé kàbà         \\ 
{\bf tsílì yá mùdì}  {\it n} 7/8 dwarf (small person) {\it pl.} bètsílì bé búdì         \\ 
{\bf tsílì yá ndáwÒ}  {\it n} 7/8 room {\it pl.} bètsílì mándáwÒ         \\ 
{\bf tsílí yá sÓtì}  {\it n} 7/8  pants {\it pl.} bètsílí bé sÓtì         \\ 
{\bf tsìlÒ}  {\it v} write   ntsìlá  tsìlÈsÈ tsìlàlà   \\ 
{\bf tsímbÉ}  {\it n} 7/8 plank {\it pl.} bètsímbÉ         \\ 
{\bf tsímÈlÈ}  {\it v} sneeze     tsímÈsÈ tsímàlà   \\ 
{\bf tsíndí}  {\it n} 9/6 Ufer, bord {\it pl.} màtsíndí         \\ 
{\bf tsíndí (lé nkú)} lè {\it n} 5/6 heel (of the foot) {\it pl.} màtsíndí má nkú         \\ 
{\bf tsíndÓ} lè {\it n} 5/6 1) fête, party, festival 2) "neuvène" ceremony nine days after funeral {\it pl.} màtsíndÓ         \\ 
{\bf tsíndÒ}  {\it v} schieben (push lightly), shove   ntsíndâ   tsíndàlà   \\ 
{\bf tsíyà}  {\it n} 1/2 question {\it pl.} bátsíyà         \\ 
{\bf tsùk tsùk tsùk tsùk}  {\it ideo} depiction of noise that mice make         \\ 
{\bf tʃÒp tʃÒ tʃÒ}  {\it ideo} depiction of dripping sound or sound walking in mud         \\ 
{\bf tù}  {\it post} inside         \\ 
{\bf túà}  {\it v} move places/houses   ntÓgâ  tÓgÈsÈ tÓgàlà   \\ 
{\bf túdÈ}  {\it n} 7/8 tumor {\it pl.} bètúdÈ         \\ 
{\bf túmbà} n {\it n} 1/2 older brother, cousin, close friend {\it pl.} bàtúmbà         \\ 
{\bf túmbÓ}  {\it n} 7/8 country {\it pl.} bètúmbÓ         \\ 
{\bf tùndÈ}  {\it v} miss, fallir de blesser   ntùndá   tùndàlà sich gegenseitig leicht verletzten   \\ 
{\bf túnÒwÒ}  {\it v} float         \\ 
{\bf túù}  {\it n} 7/8 spoon {\it pl.} bètúù         \\ 
{\bf túwànÈ nà}  {\it v} meet (on appointment)   ntúwánÊ   túwàlà   \\ 
{\bf twálÒ}  {\it v} peck   ntwálâ   twálàlà   \\ 
{\bf ù} d {\it n} 5/6 oven, hearth {\it pl.} mù         \\ 
{\bf ùd°â} m {\it n} 1/2 woman, wife {\it pl.} bùd°â         \\ 
{\bf ùd°û} m {\it n} 1/2 man, husband {\it pl.} bùd°û         \\ 
{\bf ùdì} m {\it n} 1/2 person {\it pl.} bùdì         \\ 
{\bf ùdì wà wÓngÓ} m {\it n} 1/2 soldier {\it pl.} bùdì bá béwÓngÓ         \\ 
{\bf ùf}  {\it ideo} depiction of sound when something catches fire         \\ 
{\bf úgó} dv {\it n} 5/6 toilet {\it pl.} múgó         \\ 
{\bf úmbÓ} d {\it n} 5/6 wrap {\it pl.} múmbÓ         \\ 
{\bf úmbÓ lé ká} d {\it n} 5/6 fish or meat wrapped and prepared in leaf {\it pl.} múmbó má ká         \\ 
{\bf úmbÓ lé nk°Ê} d {\it n} 5/6  fish or meat prepared in pot, dish with fish in lemon sauce {\it pl.} múmbÓ má nk°Ê         \\ 
{\bf úndÒ} d {\it n} 5/6 galago à queue touffue {\it pl.} múndÒ         \\ 
{\bf úú} d {\it n} 5/6 nose {\it pl.} múú         \\ 
{\bf ùwÒ} d {\it n} 5/6 daytime {\it pl.} mùwÒ         \\ 
{\bf v°áìv°áì}  {\it n} 7/8 generosity {\it pl.} bèv°á°ìv°á°ì         \\ 
{\bf v°ù°ù}  {\it v} worry, be excited         \\ 
{\bf váá} lè {\it n} 5/0 thing         \\ 
{\bf vàà}  {\it v} praise, be proud   mvàgá   vàgàlà   \\ 
{\bf válÓ}  {\it n} 7/8 polygamy {\it pl.} bèválÓ         \\ 
{\bf vàmò kwÈ}  {\it v} knock over         \\ 
{\bf vásÈ}  {\it v} aufgehen (Teig)   mvásâ vásÈlÈ (caus)     \\ 
{\bf vÉ}  {\it inv} which         \\ 
{\bf vÊ}  {\it v} give (more Gyeli)   mvÉyâ   vÉyàlà   \\ 
{\bf vè'è}  {\it v} try on (clothes)   mvègá vè'ÈlÈ make try  vègàlà try on recip   \\ 
{\bf vÉÈlá}  {\it n} 7/8 decoration {\it pl.} bèvÉÈlá         \\ 
{\bf vèkò}  {\it n} 7/8 drawing, painting {\it pl.} bèvèkò         \\ 
{\bf vÉmbÓ} lè {\it n} 5/6 guenon (Cercopithecus) {\it pl.} màvÉmbÓ         \\ 
{\bf vÉmbÒ (kÈmbÈ)}  {\it v} blow nose (phlegm)   mvÉmbâ   vÉmbàlà   \\ 
{\bf vÉsÒ}  {\it v} have desire   mvÉsâ   vÉsàlà   \\ 
{\bf vÉwÒ} lè {\it n} 5/6 cold, malaria {\it pl.} màvÉwÒ         \\ 
{\bf vÈwÒ}  {\it v} breathe (respirer)         \\ 
{\bf vèyÈ}  {\it v} mesure   mvèyá   vèyàlà   \\ 
{\bf ví}  {\it n} 7/8 wooden part in trap hiding the hole in the ground {\it pl.} bèví         \\ 
{\bf vìdÈ}  {\it v} turn, return, roll tr (same as vidElE)   mvìdá and mvìdálâ vìdÈlÈ turn sth  vìdàlà vìdègà turn  \\ 
{\bf vídÉlÈ}  {\it n} 7/8 smoke {\it pl.} bèvídÉlÈ         \\ 
{\bf vídósí} lè {\it n} 5/6 dawn, early morning {\it pl.} màvídósí         \\ 
{\bf vídú} lè {\it n} 5/6 darkness {\it pl.} màvídú         \\ 
{\bf vìlÈ} lè {\it n} 5/6 ginger species (aframomum) {\it pl.} màvìlÈ         \\ 
{\bf vímàlà}  {\it v} groan   mvímálâ      \\ 
{\bf vímù}  {\it n} 7/8 giant pangolin (Manis gigantea) {\it pl.} bèvímù         \\ 
{\bf víndò}  {\it v} hate   mvíndâ   víndàlà   \\ 
{\bf vínó} mà {\it n} 0/6 pus         \\ 
{\bf vìnÓ}  {\it n} 7/8 finger {\it pl.} bìvìnÓ         \\ 
{\bf vìnÓ yá s°á}  {\it n} 7/8 thumb (main finger)         \\ 
{\bf vísÓ}  {\it n} 8/0 sun         \\ 
{\bf vísÒ}  {\it v} cover   mvísâ and mvísálâ vísÈlÈ cover  vísàlà   \\ 
{\bf vìsÓ}  {\it n} 7/8 bone, skeleton, fish bone {\it pl.} bèvìsÓ         \\ 
{\bf vìsÓ yá nkáàlÈ}  {\it n} 7/8 backbone {\it pl.} bèvìsÓ bé mínkáàlÈ         \\ 
{\bf víwÒ}  {\it v} suck   mvíwâ   víwàlà   \\ 
{\bf víy°àsà}  {\it v} be light   mvíy°ásâ      \\ 
{\bf víyàlà}  {\it v} touch   mvíyálâ      \\ 
{\bf vìyÓ}  {\it n} 0/8 fire         \\ 
{\bf vÔ}  {\it v} 1) be calm 2) be cold   mvóyâ  vÓlÈsÈ calm sb down vólàlà   \\ 
{\bf vÒdá} lè {\it n} 5/6 rest, vacation {\it pl.} màvÒdá         \\ 
{\bf vòdà}  {\it v} rest   mvòdá   vòdàlà   \\ 
{\bf vÓlÈ} bè {\it n} 0/8 grief (after sb.' departure/death)         \\ 
{\bf vÓlÈ}  {\it v} help   mvÓlâ   vÓlàlà   \\ 
{\bf vóvÓlÈ}  {\it n} 7/0 freshness, peace, tranquillité         \\ 
{\bf vòwà}  {\it v} wake (up)   mvòwâ  vòlÈsÈ vòwàlà vòlègà wake up  \\ 
{\bf vû}  {\it v} leave   mvúyâ vúlÈ get rid of, take away  vúyàlà   \\ 
{\bf vúbà (nà mùd°â)}  {\it v} hug each other         \\ 
{\bf vúd°ù} * {\it mod} one         \\ 
{\bf vúÈlÈ}  {\it v} blow (with mouth, e.g. into fire)   mvúálâ      \\ 
{\bf vúlÒ} mà {\it n} 0/6 scharfe Seite vom Messer oder Machete         \\ 
{\bf vúlÒ}  {\it v} be sharp   mvúlâ      \\ 
{\bf vúlù} lè {\it n} 5/6 foam {\it pl.} màvúlù         \\ 
{\bf vùlùngù}  {\it n} 7/8 Schlinge in trap {\it pl.} bèvùlùngù         \\ 
{\bf vúsí} lè {\it n} 5/6 hole {\it pl.} màvúsí         \\ 
{\bf vútÒ} mà {\it n} 0/6 oil (for body)         \\ 
{\bf vùvùlÈ}  {\it n} 7/8 baked bread or baguette {\it pl.} bèvùvùlÈ         \\ 
{\bf vùzí}  {\it n} 7/8 abdomen {\it pl.} bèvúzì         \\ 
{\bf vyámbÈlÈ}  {\it v} surround   mvyámbálâ      \\ 
{\bf vyÈ}  {\it v} draw   mvyègá   vyégàlà   \\ 
{\bf vyû(vyû)} ná {\it qual} black         \\ 
{\bf w°â} mà {\it n} 0/6 fat         \\ 
{\bf w°ÉÈ}  {\it v} skin (animals with fur; burn the fur, then scratch fur off)   ngw°Éngâ   wÉngàlà   \\ 
{\bf wa} lè {\it n} 5/6 twin {\it pl.} màwä         \\ 
{\bf wâ ntúà} m {\it n} 1/2 young woman {\it pl.} bwâ bá túà         \\ 
{\bf wàà}  {\it n} 1/2 chimpanzee, bonobo {\it pl.} bàwáa         \\ 
{\bf wáádÓ}  {\it n} 7/6 net {\it pl.} màwáádÓ         \\ 
{\bf wàlÈ}  {\it n} 7/8 bitter kola (fruit and tree) (garcinia kola) {\it pl.} bèwàlÈ         \\ 
{\bf wámÈ}  {\it v} hurry         \\ 
{\bf wámíyÉ}  {\it adv} fast         \\ 
{\bf wánÒ} m {\it n} 1/2 1) child, baby 2) small, few {\it pl.} bwánÒ         \\ 
{\bf wánÒ (wà) múd°â} m {\it n} 1/2 girl (female child), daughter {\it pl.} bwànó búd°â         \\ 
{\bf wánÒ (wà) múd°û} m {\it n} 1/2 boy (male child), son {\it pl.} bwànó búd°û         \\ 
{\bf wánÒ nláwÓ} m {\it n} 3/4 twig (child of branch) {\it pl.} bwánÒ mínláwÒ         \\ 
{\bf wányÈ} lè {\it n} 5/6 young man {\it pl.} màwányÈ         \\ 
{\bf wàwÈ}  {\it v} spread (out)   mwàwá   wàwàlà wàwÈgà spread out by itself  \\ 
{\bf wáwÒ}  {\it v} crawl         \\ 
{\bf wáyà}  {\it n} 7/8 wire {\it pl.} bèwáyà         \\ 
{\bf wÈ}  {\it v} die   mwÈyá      \\ 
{\bf wê} lè {\it n} 5/6 cry {\it pl.} màwê         \\ 
{\bf wÒ} lè {\it n} 5/6 taro, cocoyam {\it pl.} màwÒ         \\ 
{\bf wÓ'Ò}  {\it n} 7/8 broom {\it pl.} bèwÓ'Ò         \\ 
{\bf wÓlÈ}  {\it n} 7/8 hawk {\it pl.} bèwÓlÈ         \\ 
{\bf wÒm}  {\it ideo} depiction of (sudden) silence         \\ 
{\bf wómbÈlÈ}  {\it v} sweep   mwómbálâ   wómbàlà   \\ 
{\bf wÓngÓ}  {\it n} 7/8 helmet {\it pl.} bÈwÓngÓ         \\ 
{\bf wÓÓÓÓ}  {\it ide} depiction of moving by foot or motorbike         \\ 
{\bf wùdÈ} lè {\it n} 5/6 cooking stone {\it pl.} màwùdÈ         \\ 
{\bf wùlà} lè {\it n} 5/6 time, hour {\it pl.} màwùlà         \\ 
{\bf wúmbÉ} lè {\it n} 5/6 wish, desire, want {\it pl.} màwúmbÉ         \\ 
{\bf wúmbÈ}  {\it v} want, wish, need   mwúmbâ   wúmbàlà want each other's things, desire each other   \\ 
{\bf wùmbó} lè {\it n} 5/6 cotton {\it pl.} màwùmbó         \\ 
{\bf wùmÈ (kfubO)}  {\it v} pluck   mwùmá   wùmàlà   \\ 
{\bf wúmÒ} lè {\it n} 5/6 ten {\it pl.} màwúmÒ         \\ 
{\bf wúndÈ}  {\it n} 1/2 window {\it pl.} bàwúndÈ         \\ 
{\bf wùndÈ}  {\it n} 7/8 groundnut {\it pl.} bèwùndÈ         \\ 
{\bf wúngàlà}  {\it v} wander, dangle          \\ 
{\bf wúnjÒÒ}  {\it n} 0/2 Ewondo         \\ 
{\bf wùsà}  {\it n} 7/8 dry banana leaf {\it pl.} bèwùsà         \\ 
{\bf wùsà}  {\it v} forget   mwùsá   wùsàlà   \\ 
{\bf wúsÈ}  {\it n} 7/8 drought {\it pl.} bèsúsÈ         \\ 
{\bf wùù wúú wùù wúú}  {\it ideo} depiction of sound of bees         \\ 
{\bf wùùùù}  {\it ideo} depiction of pouring liquids or granulars         \\ 
{\bf wùwù}  {\it n} 7/8 renard volant de Pemba, small bat {\it pl.} bèwùwù         \\ 
{\bf y°ây°â}  {\it n} 1/2 pan {\it pl.} bày°ây°â         \\ 
{\bf y°Èy°È yá mùdì}  {\it n} 7/8 retarded person {\it pl.} bèyÈyÈ bé bùdì         \\ 
{\bf yákú}  {\it n} 7/8 fire fly {\it pl.} bèyákú         \\ 
{\bf yàlànÈ}  {\it v} respond         \\ 
{\bf yándÓ}  {\it n} 7/8 trace {\it pl.} bèyándÓ         \\ 
{\bf yànÈ}  {\it v} must         \\ 
{\bf yÉ}  {\it n} 7/8 mushroom  {\it pl.} bèyÉ         \\ 
{\bf yÉ'É}  {\it n} 7/8 thirst, desire, envie {\it pl.} bèyÉ'É         \\ 
{\bf yÊ(yÊ)} ná {\it qual} brightened color         \\ 
{\bf yÉdÉlÈ}  {\it n} 7/8 star (also used in Kwasio) {\it pl.} bèyÉdÉlÈ         \\ 
{\bf yÉlÈ}  {\it n} 7/8 whistle (both with mouth and whistle) {\it pl.} bèyÉlÈ         \\ 
{\bf yÉmÈdÈ}  {\it v} tighten   myÉmâ   yÉmàlà   \\ 
{\bf yÉngÈ}  {\it n} 7/8 Getriller at wedding {\it pl.} bèyÉngÈ         \\ 
{\bf yí}  {\it n} 7/8 wood, firewood, fire {\it pl.} bèyí         \\ 
{\bf yíÈ}  {\it v} avoid, dodge   nyéyâ   yéàlà   \\ 
{\bf yílÈ}  {\it n} 7/6 viper {\it pl.} màyílÈ         \\ 
{\bf yìmbá}  {\it n} 7/8 age {\it pl.} bèyìmbá         \\ 
{\bf yímbálî} lè {\it n} 5/6 entrance {\it pl.} màyímbálî         \\ 
{\bf yímbÒ}  {\it v} go for a walk, visit   yímbâ   yímbàlà   \\ 
{\bf yúlÈ}  {\it n} 1/2 decedent, deceaed person {\it pl.} bàyúlÈ         \\ 
{\bf yúngú}  {\it n} 7/8 sea eagle {\it pl.} bèyúngú         \\ 
{\bf yûyû} ná {\it n} 1/2 vertigo {\it pl.} bànáyûyû         \\ 
{\bf zíNgÓ}  {\it n} 7/8 short dress {\it pl.} bèzíNgÓ         \\ 
{\bf zìßí}  {\it n} 7/8 tsetse fly (Glossina) {\it pl.} bèzìßí         \\ 


\end{document}