\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.15394180}
%\bibliography{localbibliography}
%\usepackage{hhline}% http://ctan.org/pkg/hhline

\author{Dorota Klimek-Jankowska\orcid{https://orcid.org/0000-0003-3102-1384}\affiliation{University of Wrocław} and Vesela Simeonova\orcid{https://orcid.org/0000-0002-3841-1265}\affiliation{University of Graz}} 
% replace the above with you and your coauthors
% rules for affiliation: If there's an official English version, use that (find out on the official website of the university); if not, use the foriginal
% orcid doesn't appear printed; it's metainformation used for later indexing

%%% uncomment the following line if you are a single author or all authors have the same affiliation
% \SetupAffiliations{mark style=none}

%% in case the running head with authors exceeds one line (which is the case in this example document), use one of the following methods to turn it into a single line; otherwise comment the line below out with % and ignore it
% \lehead{Šimík, Gehrke, Lenertová, Meyer, Szucsich \& Zaleska}
% \lehead{Radek Šimík et al.}

\title[Secondary imperfectives in Polish and Bulgarian]{Two types of secondary imperfectives: Evidence from Polish and Bulgarian}
% replace the above with your paper title
%%% provide a shorter version of your title in case it doesn't fit a single line in the running head
% in this form: \title[short title]{full title}
\abstract{Secondary imperfective (SI) morphology differs in its productivity in Polish (PL) and  Bulgarian (BG): in PL, the SI morphology combines with some but not all prefixes. By contrast, almost every BG perfective verb has a SI  variant. To our knowledge, there is no research that has attempted to get a closer understanding of the source of this discrepancy. To fill in this niche, we conducted a comparative study of the interaction of SI morphology with different classes of aspectual prefixes in PL and BG and the meaning effects they give rise to. We present novel observations and account for them by proposing that there are two distinct layers at which SI morphemes are generated in BG and only one such layer in PL. 

\keywords{secondary imperfective, aspect, habituality, prefixes, Polish, Bulgarian} %here, come, your, 4 to 6 keywords or keyphrases
}

\begin{document}
\maketitle

\section{Introduction}
There is an ongoing debate in the literature on Slavic aspect concerning the status of aspectual morphemes. Little agreement has been reached as to the status of secondary imperfective (SI) /(y)v/ in Slavic languages. The views vary as to where /(y)v/ is generated in the structure (see \citealt{kli+:schoorlemmer1995,babko1999,istratkova2004,kli+:milicevic2004lexical,svenonius2004b,kli+:svenonius2004a,kli+:romanova2004,kli+:filip2005,sciullo2005,kli+:arsenijevic2006,romanova2007,kli+:ramchand2008a,ramchand2008b,lazorczyk2010,markova2011,tatevosov2011,kli+:tatevosov2015,biskup2012,kli+:biskup2019,wiland2012,zaucer2012,rothstein2020,klimek2022,klimekNLLT,kli+:kwapiszewski2022}). 
In most of these studies, generalizations about the status of secondary imperfective morphology are made based on the data from a single language. However, Slavic languages differ considerably in the productivity of forming SI verbs. For example, the Bulgarian (BG) secondary imperfective is considerably more productive as compared to Polish (PL): almost every BG perfective verb has a SI variant (see \citealt{markova2011,rivero2014}; \citealt{nicolova2017}: 5.3.25), while in PL, SI morphology combines with some but not all prefixed verbs (see \citealt{lazorczyk2010,lazinski11,lazinski20,wiemer-et-al-20,kli+:kwapiszewski2022,klimek2022,klimekNLLT,}). While it is a well-known observation, there are no works that attempt to explain the locus of variation. Our goal in this paper is to fill that gap. We provide two major novel empirical observations: 

\begin{enumerate}
    \item 
     Within Bulgarian: while all previous works on SI in general and in Bulgarian specifically treat it as a single class, we identify two distinct SI classes within Bulgarian with systematically different sets of formal derivational and semantic properties. %One class of SI has only habitual readings, while the other class allows both ongoing and habitual readings.  

     \item 
     Between languages: we identify the locus of cross-linguistic variation in the productivity of SI between Polish and Bulgarian: the two languages share one of the two classes of SI and Polish lacks the second class of SI that Bulgarian has.  
\end{enumerate}


\noindent We conclude that the difference in the productivity of SI in Polish and Bulgarian is not random, but is systematically determined based on the range of derivational possibilities in the two languages, with respective semantic consequences. 

To arrive at these conclusions, we tested the interaction of SI morphology in PL and BG with two classes of prefixes -- lexical and purely perfectivizing prefixes -- and the meaning effects these different combinations of the tested aspectual prefixes with SI morphemes give rise to. In this paper, the empirical scope of the environments tested is limited to past tense contexts. We discuss possible extensions in \S \ref{section:conclusion}.

We show that in BG there are two types of SI morphemes that bear different meanings; PL has only one of these SI morphemes. We propose that the two types of SI in BG are realized in two different syntactic layers, following \citeauthor{cinque1999}'s (\citeyear{cinque1999}) model. 
 
 
 


\section{Background on the secondary imperfective in Polish and Bulgarian}\label{section:background-SI}


\subsection{Similarities between Polish and Bulgarian}\label{subsection:background-SI:similarities}
In both PL and BG, aspectual distinctions are encoded on almost all verbs.\footnote{The encoding of aspectual distinctions can be blocked, e.g. for phonological reasons, %e.g. glasuvam `vote' already has the \textit{-va} morphology in  
or in certain loanwords, e.g. \textit{printiram} `print' only has one form in BG.} The least morphologically complex aspectual forms are primary imperfectives (bare, i.e. ``unprefixed’’ verbs) and they can be perfectivized by means of a prefix, cf. \REF{basics:perfective}. Some prefixes do not change the verb in any way other than its aspectual value; they are called \textsc{purely perfectivizing} or \textsc{empty} prefixes    (\citealt{boguslawski1963,svenonius2004b,kli+:svenonius2004a,mlynarczyk2004,willim2006,kli+:ramchand2008a}).\footnote{\citet{janda2010} emphasize that Russian has at least 16 prefixes forming natural perfectives (those perfectives which are not semantically distinct from the unprefixed base verb), which may suggest that they encode hidden distinctions.
They propose that in the case of natural perfectives there is a semantic overlap between the meaning of the prefix and the meaning of the base verb and the diversity of prefixes used in natural perfectives follows from the fact that the base verbs from which they are derived fall into semantically diverse classes. Building on that, \citet{janda2013} propose that the verbal prefixes act as classifiers in that they select verbs according to broad semantic traits, categorizing them the way numeral classifiers in some languages categorize nouns. We think that irrespective of the terminology used, there is a general consensus that the prefixes in natural perfectives do not modify the meaning of the base verbs but they may only impose selectional restrictions on the base verbs they combine with. Therefore, we will maintain the terminology `purely perfectivizing'.}

%NOTEVES When you add the % sign to something, you delete it from the PDF but you keep it for yourself in the code file.
% We would like to acknowledge that this may very well be the case that purely perfectivizing prefixes are diverse because they attach to different semantic classes of base verbs (and as such they may be treated as classifiers) but even so it is still correct to assume that if the prefixes forming natural perfectives overlap the meaning of the base verb, they do not modify its lexical content and their role is purely aspectual. We think the apparent conflict between these two approaches discussed above is not a real conflict. Both approaches are correct. 

\ea\label{basics:perfective}
\ea  \gll pisać --  napisać  \\
 write.\textsc{ipfv}  {} write.\textsc{pfv} \\ \hfill PL
\ex \gll  piša -- napiša   \\
write.\textsc{ipfv} {} write.\textsc{pfv} \\ \hfill BG
\z % this ends the (sub)example
\z %in this case we need two - one to end the subexample and one to end the example


% (see e.g. BabkoMalaya 1999, Ramchand 2008a,b, Romanova 2004, 2007, Svenonius 2004a,b, Sciullo & Slabakova 2005, Arsenijević 2006, Biskup 2012, 2019, Žaucer 2009, 2012, Markova 2011, Wiland 2012).


\noindent Another class of prefixes are the so-called \textsc{lexical prefixes} (\citealt{babko1999,svenonius2004b,kli+:svenonius2004a,kli+:romanova2004,romanova2007,kli+:ramchand2008a,ramchand2008b,biskup2012,kli+:biskup2019}, a.m.o.): they have an idiosyncratic meaning where the prefix changes the lexical interpretation of the verb, but not in a predictable way, for example the prefix \textit{prze-} in \REF{basics:idiosyncratic:PL} and the prefix \textit{pod-} in \REF{basics:lexical:idiosyncratic-prefixes:BG} have very different meaning in various verbs they participate in.  
Lexical prefixes cause idiosyncratic changes in the meaning of a verbal predicate that is not derivable from either the verb or the prefix, cf. \REF{basics:idiosyncratic:PL} and \REF{basics:lexical:idiosyncratic-prefixes:BG}.\footnote{Because of these properties, \citet{kli+:romanova2004,svenonius2004b,kli+:svenonius2004a,kli+:ramchand2004,kli+:ramchand2008a,ramchand2008b,lazorczyk2010} argue that lexical prefixes are merged vP-internally.}
%in this study, we focus on the interaction of SI morphology with lexical and purely perfectivizing prefixes in PL and BG. 

\ea\label{basics:idiosyncratic:PL} 
\ea
\gll kupić -- \textit{prze}-kupić  \\
buy.\textsc{pfv} {} bribe.\textsc{pfv} \\  \hfill PL
\glt`to buy’ -- `to bribe’
%
\ex \gll grać -- \textit{prze}-grać\\
play.\textsc{ipfv} {} lose.\textsc{pfv} \\ \hfill PL
\glt `play' -- `lose'
%
\ex \gll łączyć -- \textit{prze}-łączyć\\
connect.\textsc{ipfv} {} switch.\textsc{pfv} \\\hfill PL
\glt `connect' -- `switch'\\ 
\z
\z

% \ex \gll pisha -- o-pisha [BG]\\
% write -- o-write.pfv\\
% \glt `write  -- describe'
% \z
% \z 
\ea\label{basics:lexical:idiosyncratic-prefixes:BG}
\ea \gll seštam \minsp{*(} se) -- \textit{pod}-seštam\\
recall {} \textsc{refl} {} remind.{\PFV}\\ \hfill BG
\glt `recall' -- `remind' 
%
\ex \gll budja -- \textit{pod}-budja\\
wake.up.{\IPFV} {} incite.{\PFV}\\ \hfill BG
\glt `wake up' -- `incite, instigate'
%
\ex \gll igraja -- \textit{pod}-igraja\\
play.{\IPFV} {} mock.{\PFV}\\ \hfill BG
\glt `play' -- `mock' \\ 
%
\z
\z

\noindent Furthermore, lexical prefixes (can) alter the argument structure/selectional restrictions of a verb, cf. \REF{basics:argument-structurePL}--\REF{basics:argument-structureBG}.
    

\ea\label{basics:argument-structurePL}
\ea \gll znać \minsp{\{} kogo\'{s} / \minsp{*} uszkodzenia\} \\ 
know.\textsc{ipfv} {} someone {} {} damage \\ \hfill PL
\glt `to know someone'
\ex \gll do-znać \minsp{\{*} kogo\'{s} / uszkodzenia\}\\
suffer.\textsc{pfv} {} someone {} damage\\ 
\glt `to suffer damage'  
\z 
\z 

\ea\label{basics:argument-structureBG} 
\gll mislja \minsp{(*} se) -- za-mislja \minsp{*(} se) \\
think {} \textsc{refl} {} consider {}  \textsc{refl} \\ \hfill BG
\glt `think'  -- `consider'
%
\z 

\iffalse
\ea\label{basics:argument-structure}
\ea \gll znać \minsp{\{} kogo\'{s} / *uszkodzeni\}a -- \textit{do}-znać \minsp{\{*} kogo\'{s} / uszkodzenia\} \\
know.\textsc{ipfv} {} someone {} damage {} suffer.\textsc{pfv} someone {} damage\\ \hfill [PL]
\glt `to know someone' -- `to suffer damage'  
%
\ex \gll mislja \minsp{(*} se) -- \textit{za}-mislja \minsp{*(} se) \\
think {} \textsc{refl} {} consider {}  \textsc{refl} \\ \hfill [BG]
\glt `think'  -- `consider'
%
\z 
\z 
\fi



\noindent Lexically prefixed perfective verbs are imperfectivized by means of an \textit{-yw-} or \textit{-a-} suffix in Polish and by a \textit{-va-} suffix or vowel alternations in Bulgarian, cf. \REF{basics:lexical}. These imperfective forms derived from perfective verbs are called \textsc{secondary imperfective} (SI). Table \ref{table:lexical:PL+BG} shows more verbs from this morphological pattern. 

\ea \label{basics:lexical}
\ea \gll podpisać -- podpisywać \\
sign.\textsc{pfv} {} sign.\textsc{si}\\\hfill PL
\glt `sign' -- `sign'
\ex  \gll  podpiša -- podpisvam   \\
sign.\textsc{pfv} {} sign.\textsc{si}\\ \hfill BG
\glt `sign' -- `sign'
\z 
\z 



\begin{table}
\begin{tabularx}{\textwidth}{ X X  l l  l }
  \lsptoprule
\multicolumn{2}{c}{Polish}  & \multicolumn{2}{c}{Bulgarian} & {English}\\
  \cmidrule(lr){1-2}\cmidrule(lr){3-4}
%
PFV & SI & PFV & SI & \\ 

\midrule
 podpisać  & podpisywać & podpiša & podpisvam & `sign'\\
 %
odpowiedzieć & odpowiadać & otgovorja & otgovarjam  & `reply’ \\
%
naprawić & naprawiać & popravja & popravjam & `repair' \\ 
%
wyjaśnić & wyjaśniać & objasnja & objasnjavam & `explain' \\ 
%
sprzedać & sprzedawać & prodam & prodavam & `sell'\\ 
%
opisać & opisywać & opiša & opisvam & `describe' \\
%
%zarobić & zarabiać & zarabotja & zarabotvam & 'earn' \\ %NoteVES: I am not sure if this is superlexical or lexical. It behaves only habitual in Bg.
\lspbottomrule
\end{tabularx}
\caption{Lexical prefixes and SI in Polish and Bulgarian}
\label{table:lexical:PL+BG}
\end{table}

 




\subsection{A major difference: SI productivity} 



In the previous section, we showed that SI is possible with lexically prefixed verbs both in PL and in BG. 
%\hl{(this needs to be integrated better)} - NoteVes: DONE
However, there is a major difference between PL and BG in that almost every BG verb can form SI (\citealt{dickey-book}: 11; \citealt{nicolova2017}: 5.3.25). Most verbs with empty prefixes also have SI forms, as illustrated in \REF{basics:bg:triplet:possible}. In PL, empty prefixed forms cannot form SI, see the ungrammatical form in \REF{basics:pl:triplet:impossible}. 
 The pattern described in example \REF{basics:pl-bg-distinction} is systematic in the two languages, as demonstrated in  Tables \ref{table:pure:longBG} and~\ref{table:pure:longPL}.\footnote{The fact that one may find some rare instances of these verbs on the internet suggests that someone either used them creatively or mistakenly. Such rare uses may suggest that the two projections that we will argue for in \S\ref{section:cinque} high SI and low SI are universally there in the hierarchy of projections but in some languages such as Polish, for example, the high SI morpheme generally does not merge in this position (it is blocked), but it may exceptionally be unblocked when used creatively or in speech production errors.}



% • have secondary (derived) imperfective counterparts: 

% \ea\label{basics:derived}
% \ea
% \gll pod-pisać -- pod-pis-yw-ać [PL]\\
% pod-write.pfv -- pod-write-SI-inf\\
% \glt `to sign -- to be signing'\\
% %
% \ex \gll pod-pisha -- pod-pis-va-m [BG]\\
% pod-write.pfv -- pod-write-SI-1sg\\
% \glt `to sign -- to be signing'\\
% \z
% \z


% przy-gotow-aćP — przy-gotow-yw-aćSI
% przy-cook.pfv -przy-cook-si.ipfv
% ‘to prepare’ — ‘to be preparing’
% BG examples needed


\ea\label{basics:pl-bg-distinction}	
\ea \gll pisać – napisać -- \minsp{*} napisywać \label{basics:pl:triplet:impossible}\\ \hfill 
 write.\textsc{ipfv}  {} write.\textsc{pfv} {} {} write.\textsc{si} \\\hfill PL 
\ex  \gll piša – napiša -- napisvam \label{basics:bg:triplet:possible} \\
 write.\textsc{ipfv}  {} write.\textsc{pfv} {} write.\textsc{si}\\\hfill BG
 \z
 \z 
\begin{table}
\begin{tabularx}{.8\textwidth}{XXll}
\lsptoprule
\multicolumn{3}{c}{Bulgarian} & {English}\\
\cmidrule{1-3}
%
IPFV & PFV & SI \\ 

\midrule
 stroja & postroja & \cellcolor{gray!15} postrojavam & `build'\\
 piša & napiša & \cellcolor{gray!15} napisvam & `write’ \\ p\v{u}r\v{z}a & izpăr\v{z}a & \cellcolor{gray!15} izpăr\v{z}vam & `fry'\\ gladja & izgladja & \cellcolor{gray!15} izgla\v{z}dam & `iron' \\ broja & prebroja & \cellcolor{gray!15} prebrojavam & `count'\\  molja & pomolja & \cellcolor{gray!15} pomolvam & `ask'\\ četa & pročeta & \cellcolor{gray!15} pročitam & `read' \\gubja & izgubja & \cellcolor{gray!15} izgubvam & `lose' \\ merja & izmerja & \cellcolor{gray!15} izmervam & `measure' \\ zvănja & pozvănja & \cellcolor{gray!15} pozvănjavam & `call'\\ čupja & sčupja & \cellcolor{gray!15} sčupvam & `break' \\
\lspbottomrule
\end{tabularx}
\caption{Purely perfectivizing prefixes and SI in Bulgarian}
\label{table:pure:longBG}
\end{table}

\begin{table}
\begin{tabularx}{.8\textwidth}{XXll}
\lsptoprule
\multicolumn{3}{c}{Polish}  & {English}\\
\cmidrule{1-3}
%
IPFV & PFV & SI\\ 
\midrule
budować  & zbudować & \cellcolor{gray!15}*zbudowywać & `build'\\
pisać & napisać & \cellcolor{gray!15}*napisywać & `write’ \\
smażyć & usmażyć & \cellcolor{gray!15}*usmażywać & `fry'\\
prasować & wyprasować & \cellcolor{gray!15}*wyprasowywać & `iron' \\
liczyć & policzyć & \cellcolor{gray!15}*policzać & `count'\\ 
prosić & poprosić & \cellcolor{gray!15}*popraszać & `ask'\\
czytać & przeczytać & \cellcolor{gray!15}*przeczytywać & `read' \\
gubić & zgubić & \cellcolor{gray!15}*zgubiać & `lose' \\
mierzyć & zmierzyć & \cellcolor{gray!15}*zmierzać & `measure' \\
dzwonić & zadzwonić & \cellcolor{gray!15}*zadzwaniać & `call'\\
łamać & złamać & \cellcolor{gray!15}*złamywać & `break' \\ 
\lspbottomrule
\end{tabularx}
\caption{Purely perfectivizing prefixes and SI in Polish}
\label{table:pure:longPL}
\end{table}


% a paradox - SI is used as a test for lexical prefixes but not in BG
% how to account for the greater productivity of SI forms in Bg, is the meaning of SI of empty prefixed verbs in BG the same as the meaning of SI verbs of lexically prefixed verbs?



%!!! VES: FIX TABLE - ^^FIXED^^
In other words, in BG, there is a morphological triplet for  verbs with purely perfectivizing prefixes and a pair for lexically prefixed forms, as in Tables \ref{table:pure:longBG} and~\ref{table:BG}. 
And in PL, verbs form morphological pairs: either the bare imperfective and a verb with a purely perfectivizing prefix or the lexically prefixed imperfective and the derived SI, as in Tables \ref{table:pure:longPL} and \ref{table:PL}. The crucial difference between the two languages in the two tables is marked with shading.\footnote{
We acknowledge that the type of classification of triplets and pairs that we are using to make this claim is not the only one that exists in the literature. In a very recent study on aspectual triplets in Russian, Czech, Polish, \citet{wiemer-et-al-20} identify triplets based on a different set of criteria. They assume that lexically prefixed verbs also form triplets. We explain how we understand triplets and pairs Section, \S \Ref{subsection:background-SI:similarities} that we use in this paper  are  based on the works cited in \S \ref{section:background-SI}.
  
}
 



\begin{table}
\begin{tabular}{ c c c }
  \lsptoprule
primary imperfective & perfective & secondary imperfective \\ 
  \midrule
 N/A & podpiša `sign' & podpisvam \\
 N/A & poleja `water' & polivam \\
 \tablevspace
 stroja ‘build’ & postroja & \cellcolor{gray!15}postrojavam\\
piša ‘write’ & napiša & \cellcolor{gray!15}napisvam \\
  \lspbottomrule
\end{tabular}
\caption{Purely perfectivizing vs. lexical prefixes and SI in Bulgarian}
\label{table:BG}
\end{table}




\begin{table}
\begin{tabular}{ c c c }
  \lsptoprule
primary imperfective & perfective & secondary imperfective \\ 
  \midrule
 N/A & podpisać `sign' & podpisywać \\  
 N/A & podlać `water' & podlewać \\
 \tablevspace
 budować ‘build’ & zbudować & \cellcolor{gray!15}*zbudowywać\\
pisać ‘write’ & napisać & \cellcolor{gray!15}*napisywać \\
\lspbottomrule
\end{tabular}
\caption{Purely perfectivizing vs. lexical prefixes and SI in Polish}
\label{table:PL}
\end{table}

% . SI is possible in BG but not in PL with prefixes like \textit{na-} in \REF{basics:bg:triplet:possible}. 


While the morphological determinant of the restrictive SI in PL is well-known -- the availability of SI counterparts of perfective verbs depends on the prefix type, as we described above -- and it is also well-known that BG SI is fully productive, these observations raise many questions that remain unanswered to date. The questions we address in this paper are whether there are semantic differences between the SI forms in BG that do not have equivalents in PL and the ones that do have equivalents in PL and why the BG SI forms are not possible in PL.

\section{Novel findings: Two types of SI}


Our first finding is that the pair SI in PL is equivalent to the pair SI in BG and it is ambiguous between the single ongoing and habitual reading. The second finding is that within BG, the triplet SI is qualitatively different from the pair SI. The triplet SI is habitual only, while the pair SI is ambiguous between single ongoing and habitual reading. This means that the properties of SI are not uniform across languages and even within the same language they are not homogeneous. %but we propose, they are uniform across morpho-syntactic constructions.







\subsection{Pair SI in Bulgarian and Polish}\label{section:pair-SI-in-BG-and-PL}


Both in Bulgarian and in Polish, SI forms derived from lexically prefixed verbs are ambiguous between an ongoing reading, as in \REF{ongoing:pair:BG+PL:past}, and a habitual reading, as in \REF{habitual:pair:BG+PL:past}. This is not idiosyncratic of a specific verb, but holds across the morphological paradigm represented in Table \ref{table:lexical:PL+BG}. 


\ea \label{ongoing:pair:BG+PL:past} \textsc{Ongoing context}
\ea 	\gll Kogato 	vljazoh v 		ofisa 	  na Ivan, 	 toj  \minsp{(} točno)\hspace{2.4cm}	\minsp{\{} podpis-va-še	dokumenti / poprav-ja-še koleleta / otgovar-ja-še na imejli\}.	\\
%
when 		enter.\textsc{pfv.aor.1sg} 	in office  of Ivan 	
he {} just {} sign-\textsc{si}-\textsc{impf.3sg}	documents {} repair-\textsc{si}-\textsc{impf.3sg} bikes {} reply-\textsc{si}-\textsc{impf.3sg} to emails\\\hfill BG
%
%
\ex  \gll	Kiedy weszłam do gabinetu Jana, \minsp{(} właśnie)\hspace{2.4cm} \minsp{\{} podpis-yw-ał dokumenty / naprawi-a-ł rower / odpowiad-a-ł na maila\}.\\
%
when 	entered.\textsc{pfv.pst.1sg} 	to office 	John {} just {} sign-\textsc{si}-\textsc{pst.3sg}	documents {} repair-\textsc{si}-\textsc{pst.3sg} bike {} reply-\textsc{si}-\textsc{pst.3sg} to email\\ \hfill PL
%
\glt `When I entered John’s office, he was (in the middle of) \{signing documents / repairing bikes / replying to e-mails\}.'
\z
\z



\ea \label{habitual:pair:BG+PL:past} \textsc{Habitual context}\\
%
\ea \gll Predi obiknoveno	\minsp{\{} podpis-va-še dokumentite / poprav-ja-še koleletata / otgovar-ja-še na imejli\}	po-bărzo 	ot 	men,  no  veče 	ne.  \\
before usually {}  sign-\textsc{si}-\textsc{impf.3sg} documents {} repair-\textsc{si}-\textsc{impf.3sg} bikes.{\DEF} {} reply-\textsc{si}-\textsc{impf.3sg} to emails faster 	    	than 	me,
but already not\\\hfill BG
%
\ex \gll	Kiedyś zwykle	\minsp{\{} podpis-yw-ał dokumenty / na-prawi-a-ł rowery / odpowiad-a-ł na maile\} 	szybciej niż 	ja, ale 	teraz 	już 		nie. \\
before usually	{}	sign-\textsc{si}-\textsc{pst.3sg} documents {} repair-\textsc{si}-\textsc{pst.3sg} bikes {} respond-\textsc{si}-\textsc{pst.3sg} to emails faster 	  than 	I but 	now 	already 	not\\\hfill PL
\glt `In the past, usually he (used to) \{sign (the) documents / repair bikes / respond to e-mails\} faster than me but not anymore.'
%
%
\z
\z



% \ea\label{sim:ex:german-verbs}\gll Hans \minsp{\{} schläft / schlief / \minsp{*} schlafen\}.\\
% Hans {} sleeps {} slept {} {} sleep.\textsc{inf}\\
% \glt `Hans \{sleeps / slept\}.'
% \z

 
\noindent All the verbs in Table \ref{table:lexical:PL+BG} behave in a way analogous to the pattern shown in examples \REF{ongoing:pair:BG+PL:past} and \REF{habitual:pair:BG+PL:past}, allowing both habitual and ongoing readings. We were unable to find any counterexamples. 
%---------------------------------------------------------------



\subsection{Triplet SI in Bulgarian}

The triplet SI in Bulgarian cannot be used with ongoing actions, cf. \REF{ongoing:bg:triplet:past} and only has habitual readings, cf. \REF{habitual:bg:triplet:past}. 
The examples also show that the bare imperfective is grammatical in both environments.\footnote{While we assume that it is always available in ongoing contexts, we do not claim that it is always possible in every habitual context. }


% \ea\label{sim:ex:german-verbs}\gll Hans \minsp{\{} schläft / schlief / \minsp{*} schlafen\}.\\
% Hans {} sleeps {} slept {} {} sleep.\textsc{inf}\\
% \glt `Hans \{sleeps / slept\}.'
% \z


\ea \label{ongoing:bg:triplet:past} \textsc{Ongoing context}\\
%
\ea \gll Kogato telefonăt zvănna, točno \minsp{\{}  păr\v{z}eh /\hspace{1.7cm} \minsp{*} izpăr\v{z}vah\} kjufteta.\\
when phone.{\DEF} rang.\textsc{aor.3sg} just {} fry.\textsc{impf.1sg.ipfv} {} {} fry.\textsc{impf.1sg.si} meatballs \\ \hfill BG 
\glt `When the phone rang, I was (right in the middle of) frying meatballs.'
%
\ex \gll Kogato telefonăt zvănna, točno \minsp{\{} gladeh /\hspace{1.8cm} \minsp{*} izgla\v{z}dah\} drehi.\\
when phone rang.\textsc{aor.3sg} just
{} iron.\textsc{impf.1sg.ipfv} {} {} iron.\textsc{impf.1sg.si} clothes\\ \hfill BG
\glt `When the phone rang, I was (right in the middle of) ironing clothes.'
%
%
\ex \gll Kogato telefonăt zvănna, točno si \minsp{\{} pišeh /\hspace{0.15cm} \minsp{*} napisvah\} domašnoto. \\
when phone rang.\textsc{aor.3sg} just \textsc{refl.gen}
{} write.\textsc{impf.1sg.ipfv} {} {} write.\textsc{impf.1sg.si} homework.\textsc{def}\\\hfill BG
\glt `When the phone rang, I was (right in the middle of)  writing my homework.'
\z
\z



\ea \label{habitual:bg:triplet:past} \textsc{Habitual context}\\
%
\ea \gll Kogato praveh zakuska, obiknoveno \minsp{\{} păr\v{z}eh / izpăr\v{z}vah\} po 3 kjufteta na čovek.\\
when make.\textsc{impf.ipfv.1sg} breakfast usually {} fry.\textsc{impf.ipfv.1sg} {} fry.\textsc{impf.si.1sg} \textsc{distr} 3 meatballs per person\\ \hfill BG
\glt `When I made breakfast, I used to fry 3 meatballs per person.'
%
\ex \gll Predi vinagi \minsp{\{} gladeh / izgla\v{z}dah\} drehite vednaga sled prane.\\
before always {} iron.\textsc{impf.ipfv.1sg} {} iron.\textsc{impf.si.1sg} clothes.\textsc{def} immediately after washing\\ \hfill BG
\glt `Before I always ironed the clothes immediately after washing.'
%
%
\ex \gll Predi obiknoveno \minsp{\{} pišeh / napisvah\} po njakolko  knigi na godina, no sega samo po edna.\\
%
before usually {} wrote.\textsc{impf.ipfv.1sg} {} wrote.\textsc{impf.si.1sg} \textsc{distr} several books per  year, but now only \textsc{distr} one \\ \hfill BG
%
\glt `In the past, I used to write several books per year, but now only one.'
\z
\z




% wrócili się - wracali się - to return -  

\noindent This semantic pattern observed with triplet SI above is valid across the paradigm of triplet SI, a sample of which was presented in Tables \ref{table:pure:longBG} and \ref{table:pure:longPL}. Since there is no triplet SI in Polish, this part of the data is not directly comparable between the two languages. In both single ongoing and habitual scenarios presented for Bulgarian in \REF{ongoing:bg:triplet:past} and \REF{habitual:bg:triplet:past}, Polish uses primary imperfective verbs only. 


To summarize, while PL
uses primary \textsc{ipfv} verbs to render both habitual and ongoing readings in purely perfectivized verbs and the SI forms of such verbs are blocked, BG productively uses the SI forms of those verbs to exclusively ``mark'' the special kind of habitual reading (consisting of a series of temporally non-overlapping bounded events happening on separate occasions).

% This suggests that, thanks to the vast
% productivity of SI markers, BG has repurposed those that  into a specific linguistic means to mark these contexts. Put still
% differently, BG makes use of SI verbs that are „useless” in other Slavic languages due to their
% being synonymous with the respective simplex verbs.


Our  novel observation is that the properties of SI morphology are not semantically uniform across the two languages. In the next section, we propose that the two types of SI morphemes merge at different syntactic positions in BG but not in Polish. 

%---------------------------------------------------------------



\section{The syntax of SI}\label{section:cinque}



In order to formally capture the observations presented in the previous section, we propose that in BG, the two types of SI morphemes merge at two syntactic layers -- one higher and one lower -- while in PL, the low SI morpheme merges only in the lower one. 

\ea \begin{tabular}[t]{l l c c}
BG: & SI\textsubscript{high} &  $>>$  & SI\textsubscript{low}\\
PL: &           &          & SI\textsubscript{low}\\
\end{tabular}
\z

\noindent This proposal allows us to syntactically distinguish between the properties of lexical prefixes and purely perfectivizing ones
(see \S \ref{section:background-SI}). We argue that SI is not a uniform category within Bulgarian because it merges in two different syntactic positions with different properties each. Moreover, pair SI is equivalent across Bulgarian and Polish because it is merged in the same projection SI\textsubscript{low} with the same properties. PL does not have aspectual triplets because the high SI morpheme is blocked and it cannot merge in the SI\textsubscript{high} layer while Bulgarian developed a specialized habitual meaning of the SI\textsubscript{high} morpheme. Because verbs with SI\textsubscript{high} have obligatory habitual readings in BG, which are missing in PL, we propose that the SI\textsubscript{high} morpheme merges in a projection corresponding to \citeauthor{cinque1999}'s (\citeyear{cinque1999}) Asp\textsubscript{HAB}. 

We assume \citeauthor{baker85}'s (\citeyear{baker85}) Mirror Principle, according to which syntax reflects morphology and vice versa and the linearization of functional morphemes is syntactically motivated. Additionally, we follow \citet{cinque1999}, who argues that there is a fixed hierarchy of functional projections which regulates the way adverbs and functional morphemes are merged in syntax. Based on a large survey of languages, \citeauthor{cinque1999} shows that among temporal/aspectual affixes, e.g. repetitive, frequentative, terminative, continuative, retrospective, durative, progressive, completive, those that are specifically dedicated to expressing habituality scope the highest \citep[][p. 56; 70]{cinque1999}. Crucially, this means that the dedicated habitual functional head is syntactically higher than the progressive aspectual head. The complete functional hierarchy is provided below in (\Ref{cinque:hierarchy}), in which the two functional heads are highlighted with boxes.

\ea \label{cinque:hierarchy} Mood\textsubscript{speech~act} $>$ Mood\textsubscript{evaluative} $>$  
Mood\textsubscript{evidential} $>$ 
Mood\textsubscript{epistemic} $>$ 
T\textsubscript{(past)} $>$ 
T\textsubscript{(Future)} $>$
Mood\textsubscript{irrealis} $>$
\fbox{Asp\textsubscript{habitual}} $>$
T\textsubscript{(Anterior)} $>$
Asp\textsubscript{perfect} $>$
 Asp\textsubscript{retrospective} $>$
 Asp\textsubscript{durative} $>$
 \fbox{Asp\textsubscript{progressive}} $>$
 Asp\textsubscript{prospective}/Mod\textsubscript{root} $>$
 Voice $>$ 
 Asp\textsubscript{celerative} $>$ 
 Asp\textsubscript{completive} $>$
 Asp\textsubscript{(semel)repetitive} $>$
 Asp\textsubscript{iterative} \\
%
 \hfill \citet[76]{cinque1999}
\z



\noindent Our examples in the preceding sections showed that this is the case also in BG and PL. In addition, we showed that the adverbs are optional in the case of SI\textsubscript{high}, that is, it encodes the habitual reading itself rather than being merely compatible with it. SI\textsubscript{low}, on the other hand, is compatible with both frequentative and ongoing adverbs just like a null IPFV operator, for example the one proposed in \citet{kli+:ferreira2016} selecting for VPs referring to singular or plural events respectively: IPFV [VPsg / VPpl].\footnote{We assume \citeauthor{tatevosov2011}'s (\citeyear{tatevosov2011,kli+:tatevosov2015}) proposal that aspectual morphology can be lower than the actual aspectual interpretation.}

In this way the proposal offers a formalization in syntactic terms which captures the differences in the morphological productivity of SI in the two languages, the within-language split (in BG), as well as the fact that the lower SI is equivalent in the two languages. % even though as a whole they differ

\section{Discussion and conclusion}
\label{section:conclusion}

This study presented a formal description of a systematic difference in the productivity of SI in PL and BG: the Polish perfectives that do not allow subsequent secondary imperfectivization are precisely those cases where Bulgarian SI forms only have a habitual reading and the single ongoing reading is unavailable. These are the perfective forms which in Polish contain purely perfectivizing prefixes. By contrast, Polish perfectives that allow subsequent secondary imperfectivization are those cases where Bulgarian SI forms are ambiguous between a single ongoing and a habitual reading. These are the perfective forms which both in Polish and Bulgarian contain lexical prefixes. 

Based on these novel observations we proposed that there are two distinct types of SI in BG: SI\textsubscript{high} >> SI\textsubscript{low} and only one in Polish, SI\textsubscript{low}. Crucially, we showed that SI\textsubscript{low} is uniform in the two languages -- it is ambiguous between a single ongoing and habitual reading. SI\textsubscript{high} is merged in a projection corresponding to \citeauthor{cinque1999}'s (\citeyear{cinque1999}) Asp\textsubscript{HAB} and Bulgarian syntax generates this position, while Polish does not.

Previous works on SI cannot capture the novel observations we present here. For example, \citet{rivero2014} assume that all SIs in BG have both habitual and ongoing readings. This is reflected in their formal account, attributing this duality to context. 
As we have shown, this is accurate for SI with lexical prefixes in BG and in Polish, but it overgenerates for SI with purely perfectivizing prefixes. 

Conversely, \citet{markova2011} assumes (but provides no evidence) that the SI morphology in BG is in \citeauthor{cinque1999}'s (\citeyear{cinque1999}) Asp habitual projection. This has the opposite problem: it undergenerates the available ongoing interpretations of SI with lexical prefixes.

One limitation of this study is that the syntactic proposal put forth here still does not explain why it may be that one language is able to generate the higher SI layer, while the other language is not. We leave such a comprehensive explanatory account for future work.  

Additionally, one may reasonably ask why SI\textsubscript{high} is not blocked by primary imperfective forms which can also express a habitual reading. In order to address this issue, we show below that in temporal \textit{after}-clauses imposing sequential ordering between two events only SI\textsubscript{high} is possible and simple imperfective is not, as shown in \REF{bg:when-clause}. By contrast in temporal \textit{while}-clauses with two events temporally overlapping only simple imperfective is possible and  SI\textsubscript{high} is not, as shown in \REF{bg:while-clause}.


\ea\label{bg:when-clause} \gll Vseki păt sled kato \minsp{\{*?} stroeše / postrojavaše\} kăšta, tja se srutvaše. \\
%
every time after when {} built.\textsc{ipfv.3sg} {} built.\textsc{si}.\textsc{3sg} house it.\textsc{f} \textsc{refl} collapsed.\textsc{si}.\textsc{3sg}\\ \hfill BG
%
\glt `Every time ‘after’ he built (=finished building) a house, it collapsed.’
\z


\ea\label{bg:while-clause} \gll  Vseki păt dokato \{ stroeše / \minsp{*} postrojavaše\} kăšta, imaše incidenti.\\
%
every time while {} built.{\IPFV} {} {} built.\textsc{si}.\textsc{3sg} house have.\textsc{ipfv.3sg} incidents\\ \hfill BG
%
\glt ‘Every time while he was building a house, there were incidents.’
\z


\noindent One possible answer could be that there are two homophonous SI morphemes in Bulgarian, the one applying higher in syntax being specialized in expressing habituality consisting of a series of temporally non-overlapping bounded events (instead of serving the more general task of ``undoing the perfectivizing contribution of the prefix"). Panini's Principle (also referred to as Elsewhere Principle), according to which the application of a specific rule or operation overrides the application of a more general rule, would then link the function of expressing habituality to the specialized SI\textsubscript{high} form.

What remains to be studied in more detail is the interaction of SI morphology with different classes of superlexical prefixes. It is also necessary to extend the empirical scope to other Slavic languages in order to identify which languages pattern with Bulgarian and which ones pattern parametrically with Polish, and whether there are other possibilities. It also remains to be tested whether the observations we report in this study for past tense contexts can be extended to non-past tense as well.\footnote{We note that both present tense and imperfective aspect can have a multitude of meanings, e.g. non-actual readings, see  \citet{rivero2014}, \citet{nicolova2017}: 364; due to this, a study of the interaction of SI and present tense deserves a longer paper that we leave to future work. } 

Finally,  Bulgarian and Macedonian have two properties that other Slavic languages lack: they have definite articles and they have preserved the Imperfectum and Aorist tenses. To that end, it would be relevant to test whether the SI morphology interacts with Imperfectum and Aorist and behaves differently in the present and in the past, as well as whether number and referentiality of nominal complements impact the interpretation of either types of SI.  




%------------------------------------------------------------------------------

% \section*{Basics of LaTeX for Dorota}

% \begin{itemize}
% \item
% \textbf{commenting out:}
% Latex has two views: the source where we edit the \textbf{source} text, and the final \textbf{pdf}. All text that follows the sign \% in the source  is invisible in the pdf. We can use that to add comments to each other in the source on things to pay attention to or resolve, eg like this: %vcomm this is a comment by Vesela 
% more text % dcomm this is a comment by Dorota
% or something like this: %!vcomm this is a very important comment by Vesela etc....

% \item
% \textbf{quick commands:}
% Latex has a learning curve, but at least the quick commands like ctrl+B for bold, ctrl+I for italics, ctrl+C and V for copy-paste etc work. Also special Polish alphabet symbols work, which is great. And glossed examples align themselves, which is life-saving. 

% \item
% \textbf{quotation marks}
% in latex, proper quotes don't go like this: 'text' but like this: `text' - see the text in the source and the result in the pdf

% \item
% \textbf{special environments and symbols:}
% All other things require special coding that is not too hard, for example, to make something in caps lock \textsc{text}. We also have some special symbols that Latex uses as its own special environments. These are the symbols \% for commenting, \$ for math environment (formulas and some types of text formatting, like superscript and subscript),  backslashes - a single backslash starts a command, a double backslash gives a forced new line (used in tables), and the \{ brackets also have special use in that they give the scope of commands, such as the command for \textbf{bold} - you can see how the \{ brackets determine the beginning and the end of the bold. If we want to use them simply as brackets or percent sign, we use a backslash (as in this text). The underscore also has special uses, please don't use it in body text (it is used in math environment with curly brackets to give subscript text like this: regular text$_{subscript text}$).

% the math envrionment,  to make something 

%     \item 
% \textbf{new paragraphs:}
% If you want a new paragraph in the file, the text in the code has to be separated by an empty line, for example: 

% these two lines of text are not separated by an empty line
% and therefore they appear on the same line in the pdf

% but now this text is separated by an empty line and that is why it appears properly

% \item
% \textbf{labels:}
% The best part about Latex is that it keeps track of all examples and references, so we can move them around without having to change them individually. This is done by using unique descriptive \textbf{labels} that we choose to our liking and only we see. To label an example, we use the command \label{label-text-here}. For example, I can call an example with habitual in Bg \label{bg:SI:habitual}. To refer to the example bearing that label, we use the command \REF{label-text-here}, for example: "as shown in \REF{bg:SI:habitual}, BG high SI have only habitual readings"...

% \item
% \textbf{examples:} here is how to code examples in this document (it's not equivalent in all latex)

% \ea\label{sim:ex:german-verbs}\gll Hans \minsp{\{} schläft / schlief / \minsp{*} schlafen\}.\\
% Hans {} sleeps {} slept {} {} sleep.\textsc{inf}\\
% \glt `Hans \{sleeps / slept\}.'
% \z

% template:

% Single glossed example: 

% \ea\label{labeltext}\gll here are some source words.\\
% gloss gloss gloss gloss gloss\\ % notice that you don't need to do anything for the glosses to align themselves, the environment already does that for you, you just separate the words with a regular space in both the source line and the gloss line
% \glt `translation.'
% \z % this ends the example

% Example with subexamples:


% \ea\label{labeltext}
% \ea  first subexample \label{also-can-be-labeled} 
% \ex second subexample \label{labeled-differently} %note that the second subexample is with the code \ex not \ea 
% \z % this ends the (sub)example
% \z %in this case we need two - one to end the subexample and one to end the example

% Example with glossed subexamples
% \ea\label{labeltext}
% \ea \gll source words.\\
% gloss gloss\\
% \glt `translation.'
% \ex \gll source words of second subexample.\\
% gloss gloss gloss gloss gloss\\
% \glt `translation.'
% \z % the first z ends the subexample
% \z % this ends the example


% \item 
% \textbf{references:} all references will be replaced by labels. I will do this as we go along. Once there is a label for a reference, please just use the label when using the reference. Here is how to cite:
% \citet{labelname} for refs in the main text
% \citealp{label-name} for refs in parentheses


% \item
% Dorota, add your questions here 



% \end{itemize}



% % % Just comment out the input below when you're ready to go.
  % \input{example-osl.tex}

\section*{Abbreviations}

\begin{tabularx}{.5\textwidth}{@{}lQ}
\textsc{1}&first person\\
\textsc{3}&third person\\
\textsc{aor}&aorist\\
\textsc{def}&definite\\
\textsc{distr}&distributive\\
\textsc{ipfv}&imperfective\\
\textsc{gen}&genitive\\
\textsc{f}&feminine gender\\
%\textsc{n/a}&not applicable\\
\end{tabularx}%
\begin{tabularx}{.5\textwidth}{lQ@{}}
\textsc{ipfv}&imperfective\\
\textsc{pfv}&perfective\\
\textsc{pp}&past participle\\
\textsc{prs}&present tense\\
\textsc{pst}&past\\
\textsc{refl}&reflexive\\
\textsc{sg}&singular\\
\textsc{si}&secondary imperfective\\

%&\\ % this dummy row achieves correct vertical alignment of both tables
\end{tabularx}

\section*{Acknowledgments}
This research was supported by the Polish National Science Center (NCN) grant SONATA BIS-11 HS2 (2021/42/E/HS2/00143). We are very grateful to the audiences of FDSL in Berlin, to the anonymous reviewers, and to the editors and especially Berit Gehrke for their insightful comments.

\printbibliography[heading=subbibliography,notkeyword=this]

\end{document}
