\addchap{\lsAcknowledgementTitle} 
I am very thankful for having so many people in my life who have supported me in the process of writing this book and beyond. Most importantly, I am deeply grateful to my doctoral supervisor, Lieselotte Anderwald, for introducing me to the concept of enregisterment and giving me the opportunity to follow my research interests while working as her assistant. Her expertise and guidance were invaluable to me and I would like to thank her for her constant and reliable support. I also thank Elmar Eggert for his willingness to act as my second supervisor and Edgar Schneider for reviewing my thesis as a third supervisor. I would also like to express special thanks to Susanne Heinz who has always been a constant source of motivation and encouragement for me.

I am also very lucky to be surrounded by an amazing group of colleagues and friends at Kiel University who have supported me in many ways. I say thank you to Beke Hansen, Berit Johannsen, Mila Lütjohann, Johanna Gerwin, Robert Daugs and Jana Fischer for their many helpful comments on my work and for sharing my fascination with language with me. I thank Bärbel Schwarz and Tom Schoroth for their administrative, technical and moral support. I would like to thank Victoria Allen very much for the many productive writing sessions that we shared and Benjamin Peter for all the inspiring discussions about language and enregisterment that gave rise to many ideas and insights expressed in this book. I am also very grateful to Stefanie Weymann-Teschke for her good advice, her great humor and her invaluable support in the final stages of writing my thesis.

I would like to express my thanks to the current and former editors of this series, Alexandra D’Arcy, John Nerbonne and Martijn Wieling, as well as Sebastian Nordhoff for making this publication possible. I also thank all the volunteers that have supported the publication process by proofreading one of the chapters. Special thanks are due to Astrid Richardsen for helping me edit the final manuscript. I am also indebted to all the people involved in creating the newspaper databases \textit{America’s Historical Newspapers} and \textit{Nineteenth-Century U.S. Newspapers}.

Finally, I would like to thank all of my lifelong friends for their constant encouragement: Mira, Jule, Uwe, Birte and Tine. I thank my partner Christian for his love and patience and my brothers Lars and Pay Sönke as well as my sisters Irina and Berit for being the best siblings one could hope for. Above all, I thank my parents Rita and Volkert for always believing in me and for their unconditional love and support. I owe you everything.


