\addchap{\lsAcknowledgementTitle} 

\hypertarget{Toc124860605}{}
This book is a revised version of my PhD dissertation, discussed in  {July 2021} somewhere on the internet, with the title \textit{Pragmatic markers in Italian: Four case studies on illocutive functions of adverbs and sociolinguistic variation}. My PhD research has been conducted for about four years in a joint doctoral program (so-called \textit{cotutelle de thèse}) at the University of Turin and at the Humboldt University of Berlin. Many people contributed positively to it, and I want to mention them in a round of acknowledgments. I want to thank my PhD supervisors~– Massimo Cerruti (University of Turin) and Richard Waltereit (Humboldt University of Berlin) – who followed the progress of my work with interest, care, and accuracy. I am grateful for both their excellent scientific expertise and their sensibility, which supported my work in many ways. Grazie/Danke.

I want to thank the professors who made the effort of reading my PhD dissertation and gave valuable feedback to improve it and to give it its present shape: Maj-Britt Mosegaard Hansen (The University of Manchester), Caterina Mauri (University of Bologna), Mario Squartini (University of Turin), Malte Rosemeyer (Free University of Berlin) and Chiara Fedriani (University of Genoa).  I also want to thank my friends and colleagues Eugenio Goria (University of Turin) and Silvia Ballarè (University of Bologna) for their collaboration, discussions, and advice – and my former professors Gaetano Berruto, Cecilia Andorno and Davide Ricca (University of Turin), on whose work I relied very much while developing this research. Thus far for the academic side. On the other side, that of multiple realities and infinite sets, I want to deeply thank the lovely people, critters and apparitions in Turin, Berlin, Vienna and Lisbon (as well as in many other places) for accompanying me through these last years.

Thanks to you, most of this time has been light-hearted and ecstatic – and it was not too difficult to overcome the bad times. (\textit{Andiamo a fare il mondo bello}). As an extravagant (but truly honest) finale, a warm shout-out to all the musicians, poets, and radio hosts who made my working sessions much more forgiving and filled my mind with sounds, words, and inferences – inspiring further sounds, words, and inferences.\bigskip

\hfill Lisbon, summer 2023



